### 2015-12-18 loren/SULI/finalReport.tex
adding final report

\documentclass[11pt]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsthm}
 \usepackage{verbatimbox}
\usepackage{amsfonts}
\usepackage{longtable}
\usepackage{amssymb}
\usepackage{colortbl}
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{amscd,amsmath,mathrsfs}
\usepackage{setspace}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  mathescape
}
\parskip=6pt
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Fact}
\newtheorem{pourism}{Pourism}
\newtheorem{problem}{Problem}
\setlength\parindent{0pt}
%-------------------------------------------------------BEGIN DOCUMENT-----------------------------------------------------------------------------------------
\begin{document}
\title{Functional Programming in Computer Science}
\author{Loren James Anderson}
\date{\today}
\maketitle
\begin{abstract}
\noindent We detail the topics studied and work accomplished by the author during his 16-week internship in functional programming during the Fall 2015 Student Undergraduate Laboratory Internship program at Los Alamos National Laboratory, funded by the Department of Energy. Topics include fundamentals of Haskell programming, combinator parsing, and lambda calculus. We highlight the author's additions to a variable-strictness, autoparallelizing, pure functional compiler and runtime system in STG.
\end{abstract}
%-------------------------------------------------------BEGIN INTRODUCTION--------------------------------------------------------------------------------------
\section{Introduction}
Functional programming has exploded over the past decade due to its advantages over imperative languages. One clear advantage is the ease of parallelization using functional languages. With ever-increasing computing power, the overheads of parallelization are becoming insignificant compared to the computational efficiency offered by parallelization. Also, the blocks of code in functional languages are individual functions that are connected through composition. This often results in more straightforward and concise code that is easier to comprehend and modify than code of imperative languages.

During the internship, the first task was to learn the elements of functional programming through the standard language Haskell. I read the textbook \textit{Programming in Haskell} by Graham Hutton and completed all exercises.$^4$ Each chapter was devoted to a feature of Haskell; some were unique to functional programming, and others were unique to Haskell. Second, I supplemented my knowledge of functional programming by learning the fundamentals of combinator parsing through the paper ``Higher-order functions for parsing" by Graham Hutton.$^3$ The idea of combinator parsing is to construct large parsers from smaller parsers, and in doing so, build a library of increasingly powerful parsers. Combinator parsing is effective in lazy functional languages such as Haskell due to its ability to handle ambiguous grammars and ease of implementation. To conclude this section, I implemented both a simple arithmetic expression parser and evaluator and also a reverse Polish notation expression parser and evaluator. The third task was to learn the foundations of all functional programming languages: lambda calculus. Lambda calculus was invented in the 1930s by Alonzo Church to formalize the concept of effective computability. It is a Turing-complete language that has advanced the theory of functional programming and proof theory of computer science in general. I used instructions from the book \textit{Formal Syntax and Semantics of Programming Languages} by  Kurtz and Slonneger to code a lambda calculus evaluator and interpreter.$^5$ Lambda expressions that I implemented include nonnegative integers, basic mathematics functions, logical operators, list operations, and the simulation of recursion through the fixed-point combinator. The final task was to make additions to a variable-strictness, autoparallelizing, pure functional compiler and runtime system in STG. I coded a compiler pass checking for illegal instances of duplicate names between individual files and the prelude, and I coded tests for debugging purposes using exercises from the book \emph{Algorithms} by Sedgewick and Wayne.$^6$ 
%-------------------------------------------------------BEGIN HASKELL PROGRAMMING----------------------------------------------------------------------------------
\section{Haskell Programming}
Haskell is a pure functional language, as the basic blocks of code are functions and there are no side effects. Haskell was conceived during the 1987 conference on Functional Programming Languages and Computer Architecture. A committee was formed to create an open standard for functional languages, employing the best ideas of a few dozen distinct functional programming languages at the time. The standard implementation of Haskell is the Glasgow Haskell Compiler, which is an interpreter and native-code compiler. We now explain the features of Haskell that distinguish it as a standard programming language. Some are unique to functional programming, and others are unique to Haskell.

\subsection{Features}

\hspace{0.5cm}\underline{Functional}

Haskell is classified as a functional language. Functional languages are exploding in popularity recently due the ease of implementing parallelization. The functional programming paradigm involves functions being the main blocks of code in a program. These functions are similar to mathematical functions, as they can be expressed as a list of input/output pairs. For example, the addThree function that adds three integers is defined below.

\hspace{2cm}\begin{verbbox}
addThree:: Int -> Int -> Int -> Int
addThree x y z = x + y + z
\end{verbbox}
\theverbbox

The first line of code is called the type signature. The type signature of a function denotes both the type of arguments that it accepts as inputs and also the type of its output, the last type in the type signature. As expected, the addThree function above takes three variables of type Int as inputs and returns a value of type Int as an output (where the set of all values of type Int is a finite subset of the integers, due to space constraints).

The second line of code is the function itself. The infix operator $+$ is already defined in the prelude, so there is no need define it again. Regardless of its inputs, the addThree function only contains one case: it adds $x$, $y$, and $z$ together for all Int values of $x$, $y$, and $z$. Although this function had a single case, we will display functions having multiple cases and are therefore multiple lines long.

\hspace{0.5cm}\underline{Pure}

Pure languages do not have side effects. A side effect is an interaction using IO or any change to the state of the computer. Values are immutable in Haskell, as their states are unchangeable. Under the functional programming paradigm, any input to a function must be the output of a previous function or an immutable value. Therefore, no function can change the state of the computer by assigning its output to a global variable, a standard side effect. A particularly disastrous consequence of a global variable is when two functions manipulate a common global variable. Consequently, one must take extra precautions when reviewing the logic of such a program, as logical errors easily persist. Although Haskell is intended for use as a pure functional program, IO is supported for purposes of practicality.

\hspace{0.5cm}\underline{Concise}

Functional languages are considered high-level, as the syntax is more similar to standard human dialogue as opposed to technical machine-level code. In Haskell, few keywords are used and indentation is exploited to alter the program's overall structure. Hutton states,$^4$ ``Haskell programs are often between two and ten times shorter than programs written in other current languages.'' Below, we showcase the concision of Haskell through its ability to solve the classic Tower of Hanoi problem of any level in a few lines of code.

\hspace{0.2cm}\begin{verbbox}
solve :: Int -> ([Int],[Int],[Int]) -> ([Int],[Int],[Int])
solve 1 ((x:xs),int,dst) = (xs,int,x:dst)
solve n (src,int,dst) = let (src1,int1,dst1) = solve (n-1) (src,dst,int)
                            (src2,int2,dst2) = solve 1 (src1,dst1,int1)
                            (src3,int3,dst3) = solve (n-1) (int2,src2,dst2)
                        in (int3,src3,dst3)
Input:   solve 10 ([1..10],[],[]) 
Output: ([],[],[1,2,3,4,5,6,7,8,9,10])
\end{verbbox}
\theverbbox

\hspace{0.5cm}\underline{Strongly-typed}

The type system in Haskell is strongly-typed, meaning that a program will not compile if an input type of one function does not match the output type of a second composed function. A notable advantage of a strong type system is that runtime errors are an impossibility, a problem common to imperative languages. Furthermore, Haskell supports type inference, a process where function types signatures are determined by the compiler that maximizes type abstraction. For example consider the function fst that takes the first element of a pair.

\hspace{2cm}\begin{verbbox}
fst :: (a,b) -> a
fst (x,y) = x
\end{verbbox}
\theverbbox

The symbols a and b are type variables, or placeholders for types determined by the inputs. Note that types a and b may or may not be distinct, and both type variables a must designate the same type. A type variable is said to be polymorphic, or of many forms, and a type signature containing such a type is called a polymorphic type. The function is subsequently called a polymorphic function. This attention to type abstraction eliminates repetitive code.

\hspace{0.5cm}\underline{Recursion}

Many programming languages are capable of iterating a block of code, and nearly every program implements such a process. The corresponding technique in imperative languages is to invoke the for or do...while statement. These statements allow the programmer to specify either the number of times or the conditions when, respectively, to execute the desired block of code. In contrast, Haskell implements recursive functions cleverly to achieve the same process. Below, we display Haskell code that simulates the mathematical operation of exponentiation.

\hspace{2cm}\begin{verbbox}
pow :: Int -> Int -> Int                
pow x 0 = x                               
pow x n = x * (pow x (n-1))
\end{verbbox}
\theverbbox

In the second and third lines of code, pattern matching is invoked for brevity. The second line of code accepts any value for the first argument and only the value $0$ for the second argument. If those conditions are satisfied the output of the function is x. The elegance of pattern matching is that any remaining patterns (e.g. the third line of code above) are only evaluated if necessary. Therefore, the patterns need not be disjoint. The third line of code is the default case, as it accepts any two values as inputs. If the default case is evoked, x is multiplied to the value produced by the expression pow x (n-1). This corresponds to the recursive case of the exponentiation function.

\hspace{0.5cm}\underline{List Comprehensions}

The list is the fundamental data structure of Haskell, so it is necessary to possess an efficient means of creating one. List comprehension notation is the syntactic sugar used to construct new sets from existing sets. We can express the first ten positive even integers by doubling the first ten positive integers as shown below.

\hspace{2cm}\begin{verbbox}
Input: [2 * x | x <- [1..10]]  
Output: [2,4,6,8,10,12,14,16,18,20]
\end{verbbox}
\theverbbox

The expression \texttt{x <- [1..10]} is called a generator, stating that we take all possibilities of x from 1 to 10. This is quite similar to mathematical set notation: the vertical bar $|$ stands for "such that", and \texttt{<-} denotes the standard element inclusion symbol $\in$. The equivalent mathematical expression is $\{2x | x \in \{1,\ldots,10\}\}$. The beauty of list comprehension is that one can combine generators as shown.

\hspace{2cm}\begin{verbbox}
[(x,y) | x <- [1..4], y <- [x..4]]. 
\end{verbbox}
\theverbbox

This set is all pairs (x,y) such that x and y range from 1 to 4 and x is less than or equal to y. Mathematically, it is $\{(x,y) | x,y \in \{1,...,4\}, x \le y\}$.

\hspace{0.5cm}\underline{Higher-Order Functions}
One of the most powerful features of functional programming is the use of higher-order functions. When declaring the inputs and outputs of functions through a type signature, inputs and outputs are allowed to be functions themselves. Any function possessing a function input or a function output is called a higher-order function. The main benefit of higher-order functions is the reduction of code length, improving reliability and facilitating the reader's understanding of the program's logic. For example, consider mapping one function over a list of elements. One would desire to code a higher-order function that takes the original function and the list as inputs. The standard function \texttt{map} is defined below.

\hspace{2cm}\begin{verbbox}
map :: (a -> b) -> [a] -> [b]
map f xs = [f x | x <- xs]
\end{verbbox}
\theverbbox

Here, we take a function f with type signature \texttt{(a -> b)} and a list \texttt{xs} as inputs. Note that the type of element in the list, \texttt{a}, is the domain of the \texttt{f}. The output of \texttt{map} is a list of elements of type \texttt{b}, corresponding to the output type of function \texttt{f}. We display the effect of mapping \texttt{eqZero} to the list \texttt{[0,1,2]}, where \texttt{eqZero} is a predicate that determines if a number is zero.

\hspace{2cm}\begin{verbbox}
eqZero:: Int -> Bool                                                                                       
eqZero 0 = True
eqZero x = False

Input:   map eqZero [0,1,2] 
Output: [True,False,False]      
\end{verbbox}
\theverbbox

\hspace{0.5cm}\underline{Lazy Evaluation}

Haskell has lazy evaluation, meaning that it only evaluates functions when necessary. The possibility of infinite data structures is an immediate consequence of lazy evaluation. For example, consider the infinite list comprised of the positive integers $\{1,2,3,\ldots\}$. In Haskell, this list is represented as \texttt{[1..]}. The expression \texttt{head [1..]} evaluates to 1 because head only needs the first element of the list: lazy evaluation obviates the need to construct the remainder of the list. Also, Haskell evaluates either the alternative or consequent in if statements, but not both.

\hspace{2cm}\begin{verbbox}
if (even x) then (x/2) else (x+1)
\end{verbbox}
\theverbbox

In the above expression, \texttt{even x} will be evaluated to a boolean value. Then, \texttt{x/2} is evaluated if \texttt{(even x)} evaluates to true, and evaluates \texttt{(x+1)} is evaluated if \texttt{(even x)} evaluates to false. Also, lazy evaluation allows Haskell to evaluate an expression once and not evaluate the expression again. For example, consider the functions \texttt{addOne} and \texttt{addTwo} below.

\hspace{2cm}\begin{verbbox}
addOne :: Int -> Int
addOne x = x+1
addTwo :: Int -> Int
addTwo x = addOne x + addOne x                                                                                                                              
Input:  addTwo 3
Output: addOne 3 + addOne 3                
\end{verbbox}
\theverbbox

Haskell will only evaluate the subexpression \texttt{addOne 3} once in the expression \texttt{addTwo 3}.

\subsection{Tasks}

The first task of the internship was solely devoted to reading the textbook \textit{Programming in Haskell} by Graham Hutton and completing all exercises.$^4$ Hutton explained many of the above concepts in detail including concepts of equational reasoning, parsing, and IO. Most chapters concluded with extended examples such as the cracking the Caesar Cipher, a tautology checker, and Conway's Game of Life. One entire chapter was devoted to solving the classic Countdown Problem efficiently by exploiting the advantages of Haskell. The thorough treatment of topics with examples and exercises provided a solid foundation for functional programming and strengthened the my computer science background in general.

%-------------------------------------------------------BEGIN COMBINATOR PARSING--------------------------------------------------------------------------------
\section{Combinator Parsing}

Parsing is the act of analyzing text and producing a corresponding parse tree that makes the structure of the text explicit. Combinator parsing is the process of using higher-order functions to build progressively larger parsers from smaller parsers. We explore combinator parsing by building up from the fundamental parsers \texttt{success} and \texttt{failed} and ending with parsing full arithmetic expressions.

\subsection{Fundamental Parsers}

Parsers take a string of symbols and return a list of pairs containing the result and the remaining string that was unused. The \texttt{Parser} type is shown below.

\hspace{2cm}\begin{verbbox}
type Parser a b = [a] -> [(b,[a])]
\end{verbbox}
\theverbbox

As their names suggest, the parser \texttt{succeed} is always successful, and \texttt{failed} always fails. The parser \texttt{succeed} does not consume the input string, so it is necessary to specify its output.

\hspace{2cm}\begin{verbbox}
succeed :: b -> Parser a b
succeed v inp = [(v,inp)]
Input:  succeed "success" "1234" 
Output: [("success","1234")]
\end{verbbox}
\theverbbox

As shown, the \texttt{succeed} parser succeeded and returned a singleton list with its result ``succeed" and its unconsumed input string ``1234". The \texttt{failed} parser takes any input and returns an empty list.

\hspace{2cm}\begin{verbbox}
failed :: Parser a b
failed inp = []
Input:  failed "1234"
Output: []
\end{verbbox}
\theverbbox

The empty list signifies failure. The next step after defining \texttt{succeed} and \texttt{failed} is to construct a parser of a single symbol named \texttt{satisfy}. True to its name, an input predicate must be satisfied for the parser \texttt{satisfy} to succeed.

\hspace{2cm}\begin{verbbox}
satisfy :: (Char -> Bool) -> Parser Char Char
satisfy p [] = fail []
satisfy p (x:xs) | p x = succeed x xs
                 | otherwise = fail xs
Input:  satisfy isDigit "12345"
Output: [('1',"2345")]
\end{verbbox}
\theverbbox

If the input string is empty, there is failure. Otherwise, there is success if and only if the predicate is satisfied. The string \texttt{"12345"} is parsed using the \texttt{isDigit} predicate from the Data.Char module (returns true if a character is a digit, false otherwise). The predicate is satisfied because the first character is a digit. The output is \texttt{'1'}, and the unconsumed input is \texttt{"2345"}. This success is represented by the singleton list \texttt{[('1',"2345")]}, the list of successful parses. A common predicate to use as input for the \texttt{satisfy} parser is equals. The \texttt{literal} parser succeeds if the first character in the input string is equal to the input of \texttt{literal}.

\hspace{2cm}\begin{verbbox}
literal :: Char -> Parser Char Char
literal x = satisfy (==x)
Input:  literal '3' "1234" 
Output: []
\end{verbbox}
\theverbbox

The motivation for combinator parsing originates from these basic parsers. The parsers above only parse the first character of an input string. Combinator parsing involves creating larger parsers that combine these fundamental parsers to parse the entire input string.

\subsection{Combinators}

The first combinator parser implemented is \texttt{alt}. It takes two parsers as inputs, evaluates each parsing expression independently, and concatenates the results of each parse. The \texttt{alt} parser is analogous to the mathematical (inclusive) \texttt{OR}. A variety of combinator parsers are most logically expressed in infix notation, similar to logical operators.

\hspace{2cm}\begin{verbbox}
alt :: Parser a b -> Parser a b -> Parser a b
(p1 `alt` p2) inp = p1 inp ++ p2 inp
Input:  (literal '2' `alt` literal '5') "2345" 
Output: [('2',"345")]
\end{verbbox}
\theverbbox

The intermediate evaluations above are as follows: \texttt{literal '2' "2345" = [('2',"345")]} and \texttt{literal '5' = []}. The individual results are simply concatenated by \texttt{alt}. The \texttt{then} parser takes two parsers as inputs. The first parser returns a list of pairs. Following, the second parser takes the remaining parse string from each pair, parses the string, and combines its result to the first result in the pair. We convey this procedure through an example.

\hspace{2cm}\begin{verbbox}
then :: Parser a b -> Parser a c -> Parser a (b,c) (p1 `then` p2) inp 
then = [((v1, v2), out2) | (v1, out1) <- p1 inp, (v2,out2) <- p2 out1]
digit = satisfy isDigit
Input:  (digit `then` digit) "1234"
Output: [(('1','2'),"34")]
\end{verbbox}
\theverbbox

Next we apply functions to the results of successful parses through the \texttt{using} parser. The \texttt{using} parser takes a function and a parser as inputs, and it applies the input function to the output of the parse.

\hspace{2cm}\begin{verbbox}
using :: Parser a b -> (b -> c) -> Parser a c 
(p `using` f) inp = [(f v, out) | (v,out) <- p inp] 
Input:  ((succeed 'e') `using` toUpper) "anything" 
Output: [('E',"anything")]
\end{verbbox}
\theverbbox

Two parsers that exemplify combinator parsing are \texttt{some} and \texttt{many}, as they combine parsers we have previously defined. The \texttt{many} parser takes a parser as input and applies it to the input string as many times as possible. A notable use of the \texttt{using} parser is its ability to concatenate the results after applying the \texttt{then} parser. The \texttt{many} parser is always guaranteed to succeed.

\hspace{2cm}\begin{verbbox}
many :: Parser a b -> Parser a [b] 
many p = ((p `then` many p) `using` cons)`alt` (succeed []) 
cons :: (a,[a]) -> [a] 
cons (x,xs) = x:xs
Input:  many digit "12345" 
Output: [("12345",""),("1234","5"),("123","45"),
         ("12","345"),("1","2345"),("","12345")]
\end{verbbox}
\theverbbox

The \texttt{some} parser does the same, except that it may fail.

\hspace{2cm}\begin{verbbox}
some :: Parser a b -> Parser a [b] 
some p = (p `then` many p) `using` cons
Input:  some digit "a12345" 
Output: []
Input:  some digit "12345"
Output: [("12345",""),("1234","5"),("123","45"),
                      ("12","345"),("1","2345")]
\end{verbbox}
\theverbbox

The parsers previously defined above supply all that is necessary to parse arbitrary integers. Utilizing the \texttt{some} parser is necessary, as there should not be success unless an integer is parsed.

\hspace{2cm}\begin{verbbox}
number :: Parser Char [Char] 
number = some (satisfy isDigit)
Input:  number "12345" 
Output: [("12345",""),("1234","5"),("123","45"),
                      ("12","345"),("1","2345")]
Input:  number "a1234" 
Output: []
\end{verbbox}
\theverbbox

The parser \texttt{negNumber} is similar to \texttt{number}, but it parses negative numbers.

\hspace{2cm}\begin{verbbox}
negNumber :: Parser Char [Char] 
negNumber (x:xs) | x == '-' = (x:xs,""): (appendMinus(number xs)) 
                 | otherwise = failed xs 
appendMinus :: [([Char],[Char])] -> [([Char],[Char])]
appendMinus [] = []
appendMinus (x:xs) = (fst(x),'-':snd(x)):(appendMinus xs)
Input:  negNumber "12345" 
Output: [] 
Intput: negNumber "-12345" 
Output: [("-12345",""),("12345","-"),("1234","-5"),
         ("123","-45"),("12","-345"),("1","-2345")]
\end{verbbox}
\theverbbox

All that is neccessary to create a parser for general integers is to combine the \texttt{number} and \texttt{negNumber} parsers.

\hspace{2cm}\begin{verbbox}
int :: Parser Char [Char] 
int = number `alt` negNumber 
Input:  int "1234" 
Output: [("1234",""),("123","4"),
         ("12","34"),("1","234")] 
Input:  int "-1234" 
Output: [("-1234",""),("1234","-"),("123","-4"),
                      ("12","-34"),("1","-234")]
\end{verbbox}
\theverbbox

\subsection{Tasks}

To learn combinator parsing, I read the paper ``Higher-order functions for parsing'' by Graham Hutton and implemented all parsing code.$^3$ Using combinator parsing, I first coded a parser and evaluator for simple arithmetic expressions (appendix 7.1). The BNF grammar is as follows:

\hspace{2cm}\begin{verbbox}
data Expr = Add Term Term | Sub Term Term | Ter Term   
data Term = Mul Power Power | Div Power Power | Pow Power 
data Power = Po Factor Factor | Fact Factor 
data Factor = Num Int | Exp Expr
\end{verbbox}
\theverbbox

First, we create an expression from a parsing \verb|"(1+((2^3)/4))-(2*2)"| as follows:

\hspace{2cm}\begin{verbbox}
Sub (Pow (Fact (Exp (Add (Pow (Fact (Num 1))) (Pow (Fact (Exp (Ter 
(Div (Fact (Exp (Ter (Pow (Po (Num 2) (Num 3)))))) (Fact (Num 4)))
)))))))) (Pow (Fact (Exp (Ter (Mul (Fact (Num 2)) (Fact (Num 2))))))) 
\end{verbbox}
\theverbbox

Below is the corresponding abstract syntax tree.

\hspace{5cm}\begin{verbbox}
         Sub 
       /     \
    Pow       Pow
     |         |
    Fact      Fact
     |         |
    Exp       Exp
     |         |
    Add       Ter
   /   \       |
 Pow   Pow    Mul
  |     |     / \
 Fact  Fact Fact Fact
  |     |    |    |
 Num   Exp  Num  Num
  |     |    |    |
  1    Ter   2    2
        |
       Div
      /   \
    Fact Fact
     |    |
    Exp  Num
     |    |
    Ter   4
     |
    Pow
     |
     Po
    /  \ 
  Num  Num
  |     |
  2     3
\end{verbbox}
\theverbbox

Notice how the syntax of the parse gives explicit structure for evaluation. The idea is to assign addition and subtraction to expressions, multiplication and division to terms, exponentiation to power, and numbers and subexpressions to factor. The evaluation of the parse tree is straightforward, and details can be found in the appendix.

\hspace{2cm}\begin{verbbox}
Input:  evaluator "(1+((2^3)/4))-(2*2)" 
Output: -1
\end{verbbox}
\theverbbox

I also modified the parser and evaluator to simulate a reverse Polish notation (RPN) calculator (appendix 7.2). 
%-------------------------------------------------------BEGIN LAMBDA CALCULUS-----------------------------------------------------------------------------------
\section{Lambda Calculus}
Lambda calculus was invented in the 1930s by Alonzo Church to formalize the concept of effective computability. The Church-Turing thesis states that lambda calculus is a Turing-complete lanauge. All functional programming langugages are essentially some implementation of lambda calculus, and lambda calculus has advanced the theory of functional programming and proof theory of computer science in general. We describe the basics of lambda calculus, including the syntax of lambda calculus, $\alpha$-conversion, $\beta$-reduction, and variable capture. Then, we display lambda calculus expressions including nonnegative integers, basic mathematics functions, logical operators, list operations, and the simulation of recursion through the fixed-point combinator. The use of definitions in the following pages highlights the rigor in lambda calculus. All definitions are taken from the book \textit{Formal Syntax and Semantics of Programming Languages} by Kurtz and Slonneger.$^5$

\subsection{Basics}

There are four expressions in lambda calculus.

\begin{itemize}
\item Variables
\item Constants (only in impure lambda calculus)
\item Function applications
\item Lambda Abstractions
\end{itemize}

The BNF grammar is as follows:

\hspace{2cm}\begin{verbbox}
<expression> :: = <variable> : lowercase identifiers
                = <constant> : predefined objects
                = (<expression> <expression>) : combinations
                = (lambda <variable> . <expression>) : abstractions
\end{verbbox}
\theverbbox

A simple lambda expression is shown below.

\hspace{2cm}\texttt{$\lambda$x -> add x x}

This lambda expression mimics the double function. In impure lambda calculus where predefined constants and functions are allowed, we can evaluate the expression using the constant 3.

\hspace{2cm}\texttt{($\lambda$x -> add x x) 3 = add 3 3 = 6}

Here, the expression \texttt{$\lambda$x ->} signifies to substitute \texttt{3} in for \texttt{x}. Although this substitution seemed easy, the general process is much more difficult in pure lambda calculus where predefined constants and functions are not allowed. In pure lambda calculus, lambda expressions are evaluated through a process called reduction. This is primarily through substituting a free variable for an expression in a lambda expression.  In general, an expression \texttt{$(\lambda$ x -> E) E1} is reduced by substituting \texttt{E1} in for all variables \texttt{x} in \texttt{E}.

\begin{definition}
An occurrence of a variable $v$ in a lambda expression is called bound if it is within the scope of a "$\lambda v$"; otherwise it is called free.
\end{definition}

We use the notation \texttt{E[v -> E1]} to denote the replacing of every free occurrence of \texttt{v} in \texttt{E} by \texttt{E1}. This creates a new lambda expression, syntactically.

\begin{definition}
A substitution is called valid or safe if no free variable in \emph{\texttt{E1}} becomes bound as a result of the substitution \emph{\texttt{E[v -> E1]}}.
\end{definition}

If in fact a free variable becomes bound as a result of the substitution, the reduction involved \emph{variable capture}. Variable capture is one of the main difficulties of reducing lambda expressions. The state of variable capture is shown below with the generic addition function.

\hspace{2cm}\texttt{($\lambda$x -> ($\lambda$y -> add x y)) [x->y] = ($\lambda$y -> add y y)}

This is a classic case of variable capture, as the new function is the doubling function, while the original function was a generic addition function. The motivation for the rules of variable substitution comes from the valid substitution below.

\hspace{2cm}\texttt{($\lambda$x -> ($\lambda y$ -> add x y))[x -> y]}

\parskip 0pt
\hspace{2.5cm}\texttt{= ($\lambda$x -> ($\lambda y$ -> add x y))[x -> v]}

\parskip 0pt
\hspace{2.5cm}\texttt{= ($\lambda$v -> ($\lambda$y -> add v y))}

\parskip 6pt
This correct reduction does not change the original function.

\begin{definition}
The set of free variables in an expresion \emph{\texttt{E}}, denoted by \emph{\texttt{FV(E)}} is defined as follows:
\end{definition}

\hspace{1cm}a) \texttt{FV(c) = null} for any constant \texttt{c}

\hspace{1cm}b) \texttt{FV(x) = {x}} for any variable \texttt{x}

\hspace{1cm}c) \texttt{FV(E1 E2) = FV(E1) $\cup$ FV(E2)}

\hspace{1cm}d) \texttt{FV($\lambda$x -> E) = FV(E) - {x}}

A lambda expression \texttt{E} with no free variables \texttt{(FV(E) = null)} is called closed.

\begin{definition}
The substitution of an expression for a (free) variable in a lambda expression is denoted by \emph{\texttt{E[v -> E1]}} and is defined as follows:
\end{definition}

\hspace{1cm}a) \texttt{v[v -> E1] = E1} for any variable \texttt{v}

\hspace{1cm}b) \texttt{x[v -> E1] = x} for any variable \texttt{x $\neq$ v}

\hspace{1cm}c) \texttt{c[v -> E1] = c} for any constant \texttt{c}

\hspace{1cm}d) \texttt{(E1 E2)[v -> E3] = ((E1 [v -> E3])(E2 [v-> E3]))}

\hspace{1cm}e) \texttt{($\lambda$v -> E)[v -> E1] = ($\lambda$v -> E)}

\hspace{1cm}f) \texttt{($\lambda$x -> E)[v -> E1] = $\lambda$x -> (E[v -> E1])} when \texttt{x $\neq$ v} and \texttt{x $\not\in$ FV(E1)}

\hspace{1cm}g) \texttt{($\lambda$x -> E)[v -> E1] = $\lambda$z -> (E[x -> z][v -> E1])} when \texttt{x $\neq$ v} and \texttt{x $\in$ FV(E1)},

\parskip 0pt
\hspace{10cm} where \texttt{z $\neq$ v} and \texttt{z $\not\in$ FV(E E1)}

\parskip 6pt
An example evaluation using these rules is displayed below.

\hspace{1cm}\texttt{($\lambda$x -> ($\lambda$y -> f x y))[y -> x]}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> f x y)[x -> z][y -> x])\hfill(g)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> ((f x y)[x -> z]))[y -> x])\hfill(f)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> (((f x)[x -> z])(y[x -> z])))[y -> x])\hfill(d)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> (((f[x -> z])(x[x -> z]))(y[x -> z])))[y -> x])\hfill(d)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> ((f (x[x -> z])) y))[y -> x])\hfill(b)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> (($\lambda$y -> (f z) y)[y -> x])\hfill(a)}\parskip 0pt

\hspace{2cm}\texttt{= $\lambda$z -> ($\lambda$y -> f z y)\hfill(e)}\parskip 6pt

Notice how variable capture is avoided in the second line.

\subsection{Lambda Reduction}

\begin{definition}{Alpha Reduction}

If \emph{\texttt{v}} and \emph{\texttt{w}} are variables and \emph{\texttt{E}} is a lambda expression,

\emph{\hspace{2cm}\texttt{$\lambda$v -> E $=>_{\alpha}$ $\lambda$w -> E[v -> w]}}

\hspace{1cm}provided that \emph{\texttt{w}} does not occur at all in \emph{\texttt{E}}.
\end{definition}

Alpha reduction allows one to change the bound variables of a lambda expression barring variable capture. Note that conversion does not change the lambda expression; both the original and new lambda expressions are equivalent. An example alpha conversion from the previous example is shown below.

\hspace{2cm}\texttt{($\lambda$x -> ($\lambda$y -> f x y))[y -> x]}\parskip 0pt

\hspace{2.5cm}\texttt{= $\lambda$z -> (($\lambda$y -> f x y)[x -> z][y -> x])}\parskip 6pt

\begin{definition}{Beta Reduction}

If \emph{\texttt{v}} is a variable and \emph{\texttt{E}} and \emph{\texttt{E1}} are lambda expressions, \emph{\texttt{($\lambda$v -> E) E1 $=>_{\beta}$ E[v -> E1]}} provided that the substitution \emph{\texttt{E[v -> E1]}} is carried out according to the rules for a safe substitution.
\end{definition}

Due to its frequent appearance, \texttt{($\lambda$v -> E) E1} is called a $\beta$-redex or ``reduction expression". Beta reduction is the most involved rule of evaluation in lambda calculus.

\begin{definition}{Normal Form}

A lambda expression is in normal form if it contains no $\beta$-redexes so that it cannot be further reduced using the beta rule. An expression in normal form has no more function applications to evaluate.
\end{definition}

We want a methodical approach to choose which beta redexes to reduce first to get to normal form. First, not every lambda expression can be reduced to a normal form due to the classic example illustrated below.

\hspace{2cm}\texttt{($\lambda$x -> x x)($\lambda$x -> x x)}\parskip 0pt

\hspace{2.5cm}\texttt{= ($\lambda$x -> x x)($\lambda$x -> x x)}\parskip 0pt

\hspace{2.5cm}\texttt{= ($\lambda$x -> x x)($\lambda$x -> x x)\ldots}\parskip 6pt

\begin{definition}

For any lambda expression of the form \emph{\texttt{E = (($\lambda$x -> B) A)}}, we say that beta redex \emph{\texttt{E}} is outside any beta redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex in a lambda expression is outermost if there is no $\beta$-redex outside of it, and it is innermost if there is no $\beta$-redex inside of it.
\end{definition}

\begin{definition}{Normal Order and Applicative Order Reduction}

A normal order reduction always reduces the leftmost outermost $\beta$-redex first. An applicative order reduction always reduces the leftmost innermost $\beta$-redex first.

\end{definition}

The example  below distinguish the differences between outermost and innermost reductions.

\begin{center}
\texttt{($\lambda$x -> ($\lambda$y -> (add x (($\lambda$x -> (sub x 3)) y)))) 5 6}\parskip 0pt
\end{center}

\begin{tabular}{l | l}
\multicolumn{1}{c}{ \underline{Normal Order Reduction}} & \multicolumn{1}{c}{ \underline{Applcative Order Reduction}}\\
\texttt{= ($\lambda$y -> (add 5 (($\lambda$x -> (sub x 3)) y))) 6}&\texttt{= ($\lambda$x -> ($\lambda$y -> (add x ((sub y 3))))) 5 6}\\
\texttt{= (add 5 (($\lambda$x -> (sub x 3)) 6))}&\texttt{= ($\lambda$y -> (add 5 ((sub y 3)))) 6}\\
\texttt{= (add 5 ((sub 6 3)))}&\texttt{= (add 5 ((sub 6 3)))}\\
\texttt{= (add 5 3)}&\texttt{= (add 5 3)}\\
\texttt{= 8}&\texttt{= 8}\\
\end{tabular}

Note how the substitution of the constant \texttt{5} is made first in normal order reduction, while the innermost redex \texttt{(($\lambda$x -> (sub x 3)) y))} is reduced first in applicative order.

\subsection{Parser and Evaluator}

\textbf{Church-Turing Thesis}: The effectively computable functions of the positive integers are precisely those functions definable in the pure lambda calculus (and computable by Turing machines).

Church's Thesis states that lambda calculus is Turing complete. Motivated by this fact, we demonstrate how to define nonnegative integers, basic mathematical functions, logical operators, list constructors, and how to simulate recursion.

The first set of lambda expressions to define is the set of nonnegative integers. There is no `correct' way to do this, so we use the universally accepted method.

\hspace{2cm}\texttt{zero = $\lambda$f -> $\lambda$x -> x}

\hspace{2cm}\texttt{one = $\lambda$f -> $\lambda$x -> f (x)}

\hspace{2cm}\texttt{two = $\lambda$f -> $\lambda$x -> f (f (x))}

\hspace{2cm}\texttt{three = $\lambda$f -> $\lambda$x -> f (f (f (x)))}

\hspace{2cm}\texttt{n = $\lambda$f -> $\lambda$x -> $\underbrace{\texttt{f (f (\ldots f (}}_\textrm{n} \texttt{x} \underbrace{\texttt{)) \ldots)}}_\textrm{n}$}

Above, the nonnegative integers are all assigned names. This is only possible in the impure lambda calculus; in the pure lambda calculus, all functions are anonymous. The first arithmetic function is the \texttt{succ} function or successor function.

\hspace{2cm}\texttt{succ = ($\lambda$n -> $\lambda$f -> $\lambda$x f (n f x))}

This function adds one to any integer.

\hspace{2cm}\texttt{succ zero}

\hspace{2.5cm}\texttt{= ($\lambda$n -> ($\lambda$f -> ($\lambda$x -> f n f x))) ($\lambda$f -> ($\lambda$x -> x))}

\hspace{2.5cm}\texttt{= ($\lambda$f  -> ($\lambda$x -> f ($\lambda$f ->($\lambda$x -> x)) f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f ($\lambda$x -> x) x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f x)) = one}
\medskip

\hspace{2cm}\texttt{succ one}

\hspace{2.5cm}\texttt{= ($\lambda$n -> ($\lambda$f -> ($\lambda$x -> f n f x))) ($\lambda$f -> ($\lambda$x -> f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f ($\lambda$f -> ($\lambda$x -> f x)) f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f ($\lambda$x -> f x) x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f f x)) = two}

The add function is defined in a similar manner.

\hspace{2cm}\texttt{add = ($\lambda$m -> ($\lambda$n -> ($\lambda$f -> ($\lambda$x -> m f (n f x)))))}

\hspace{2cm}\texttt{add two two}

\hspace{2.5cm}\texttt{= ($\lambda$m -> ($\lambda$n -> ($\lambda$f -> ($\lambda$x -> m f n f x)))) ($\lambda$f -> ($\lambda$x -> f f x))}

\hspace{4cm}\texttt{($\lambda$f -> ($\lambda$x -> f f x))}

\hspace{2.5cm}\texttt{= ($\lambda$n -> ($\lambda$f -> ($\lambda$x -> ($\lambda$f -> ($\lambda$x -> f f x)) f n f x)))}

\hspace{4cm}\texttt{($\lambda$f -> ($\lambda$x -> f f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> ($\lambda$f -> ($\lambda$x -> f f x)) f ($\lambda$f -> ($\lambda$x -> f f x)) f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> ($\lambda$x -> f f x) ($\lambda$f -> ($\lambda$x -> f f x)) f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f f ($\lambda$f -> ($\lambda$x -> f f x)) f x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f f ($\lambda$x -> f f x) x))}

\hspace{2.5cm}\texttt{= ($\lambda$f -> ($\lambda$x -> f f f f x)) = four}

Other arithmetic operations such as subtract, multiply, exponentiate. The next set of lambda expressions is the set of arithmetic operators. The boolean values true and false are the foundations of the logical operators and many other functions in general. As in the case of defining the nonnegative integers, the stated definitions below of true and false are not the only way to define true, but they may be the most efficient. The operators and, or, and not are also defined below, and they are built from the definitions of true and false.

\hspace{2cm}\texttt{true = ($\lambda$x -> $\lambda$y x)}

\hspace{2cm}\texttt{false = ($\lambda$x -> $\lambda$y -> y)}

\hspace{2cm}\texttt{and = ($\lambda$p -> $\lambda$q -> p q p)}

\hspace{2cm}\texttt{or = ($\lambda$p -> $\lambda$q -> p p q)}

\hspace{2cm}\texttt{not = ($\lambda$p -> $\lambda$a -> $\lambda$b -> p b a)}

\hspace{2cm}\texttt{and true false}

\hspace{2.5cm}\texttt{= ($\lambda$p -> ($\lambda$q -> p q p)) ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$q -> ($\lambda$x -> ($\lambda$y -> x)) q ($\lambda$x -> ($\lambda$y -> x))) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> y)) ($\lambda$x -> ($\lambda$y -> x))}

\hspace{2.5cm}\texttt{= ($\lambda$y -> ($\lambda$x -> ($\lambda$y -> y))) ($\lambda$x -> ($\lambda$y -> x))}

\hspace{2.5cm}\texttt{= ($\lambda$x -> ($\lambda$y -> y)) = false}
\medskip

\hspace{2cm}\texttt{or true false}

\hspace{2.5cm}\texttt{= ($\lambda$p -> ($\lambda$q -> p p q)) ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$q -> ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> x)) q) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> x)) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$y -> ($\lambda$x -> ($\lambda$y -> x))) ($\lambda$x -> ($\lambda$y -> y))}

\hspace{2.5cm}\texttt{= ($\lambda$x -> ($\lambda$y -> x)) = true}

Conditional statements are left to the appendix. It is possible to contruct lists of arbitrary sizes by first constructing a \texttt{pair} function and then continually nesting \texttt{pair} functions in the second slot to the desired length (see appendix). Finally, the most efficient way to simulate recursion is through the fixed point combinator \texttt{Y}.

\hspace{2cm}\texttt{Y = ($\lambda$g -> ($\lambda$x -> g (x x)) ($\lambda$x -> g (x x)))}

As functions do not have names in pure lambda calculus, it is impossible for a function to achieve recursion through calling itself. Luckily, the fixed point combinator acts in the following way:

\hspace{2cm}\texttt{Y f}

\hspace{2.5cm}\texttt{= $\lambda$g -> (($\lambda$x -> g (x x)) ($\lambda$x -> g (x x))) f}

\hspace{2.5cm}\texttt{= ($\lambda$x -> f (x x)) ($\lambda$x -> f (x x))}

\hspace{2.5cm}\texttt{= f (($\lambda$x -> f (x x)) ($\lambda$x -> f (x x)))}

\hspace{2.5cm}\texttt{= f (Y f)}

True to its name, \texttt{f} is a fixed point of \texttt{Y}. The key to exploiting the property of the fixed-point combinator is to clevely set up a conditional statement within the function. The pseudocode of the \texttt{factorial} function is displayed below with a small evaluation in impure lambda calculus.

\small
\hspace{2cm}\texttt{fac = ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))}

\hspace{2cm}\texttt{Y fac 3}

\hspace{2.5cm}\texttt{= fac (Y fac) 3}

\hspace{2.5cm}\texttt{= ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 3}

\hspace{2.5cm}\texttt{= (if 3 = 0 then 1 else 3 $\times$ ((Y fac) (3 - 1)))}

\hspace{2.5cm}\texttt{= 3 $\times$ ((Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ (fac (Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ ((if 2 = 0 then 1 else 2 $\times$ ((Y fac) (2 - 1))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ ((Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (fac (Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (if 1 = 0 then 1 else 1 $\times$ ((Y fac) (1 - 1))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ ((Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (fac (Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))(Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (if 0 = 0 then 1 else 0 $\times$ ((Y fac) (0 - 1)))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ 1)) = 6}
\normalsize

The first equality followed from the fixed point property defined in the previous block of code

\subsection{Tasks}

My ultimate task was to create a lambda parser and evaluator in Haskell. To
do this, I first read """""""""""' and completed all exercises.  The
content of the material was similar to the content presented here in the
previous sections. Many exercises were detailed and required $\beta$
reduction to be completed by hand. Following, I created a lambda parser and
evaluator using both normal order and applicative order evaluation. In
order to complete the variable substitution with new variables, it was
necessary to provide a fresh variable supply. The first implementation
involved creating an infinite list of variables, exploiting Haskell's lazy
evaluation. This list could be split in half when a substitution needed to
be made in a combined expression (case d) in definition 4. The easiest way
to split the list in half was to create two new lists by appending every
other element to the end of each list. Since only finitely many fresh
variables were needed, starting with an infinite list of fresh variables
was sufficient. The second implementation was to thread the variables
through the program. After a expression was reduced, the fresh variable
supply was passed to inner expressions to complete any reductions. Once all
inner expressions were reduced, the remaining fresh variable supply was
passed to another expression until the entire expression was in normal
form. This method is displayed in appendix .... After creating the parser
and evaluator, I constructed the nonnegative integers, arithmetic
expressions, logical operators, and list operations. The final exercise was
to code enough functions to parse and evaluate the factorial function in
lambda calculus, as shown above.

\section{Compiler}

The final chapter in my internship was to make contributions to a
variable-strictness, pure functional, autoparallelizing programming language
compiler and runtime system in the language STG. Variable-strictness means
evaluating a parse tree from leaves to root. The opposite, non-strict evaluation or
evaluating child nodes as needed, is often too costly to provide any benefit. As
stated previously, being pure allows one to neglect side-effects and any unspecified behavior. The motivation for this new language is due
to the fact that there have been continual advances in computing power but
less so in software development. Davis$^2$  states that the new compiler will ``enable both explicit
parallelism and automatic paralleliztion, including speculative execution, and
the handling of run-time faults in a more efficient way than whole application
checkpoint/restart.'' During my last few weeks, I coded both compiler tests and also a compiler pass that checks for illegal instances of duplicate names.

\subsection{Tasks}

The compiler is much too large to check its logic after each change in
code. Consequently, it is necessary to implement a method to check the
validity of updates to the compiler that is fast and reliable. The method employed was to create a test suite for the compiler; however, the test suite was deficient
in number and quality of tests at the time I began the internship. My first task was to write tests in STG using
exercises from the book \textit{Algorithms} by Sedgewick and Wayne.$^6$ This was a shrewed plan by my advisor, Kei Davis, as I learned some of the most important
concepts of computer science while making improvements to the project. The
tests were formatted such that their main expressions would evaluate to
true. Therefore, the compiler passed a test if it evaluates the main
expression in the test to be true. An example is of the test \texttt{or} is shown below.

\hspace{3cm}\begin{verbbox}
--or :: Bool -> Bool -> Bool
or = FUN(x y -> case x of {
                         True -> case y of {
	                	      True -> true;
				      False -> true};
                         False -> case y of {
			      	      True -> true;
				      False -> false}});

--Test
main = THUNK(or true false);
\end{verbbox}
\theverbbox

The new tests suite indeed boasts speed and reliability. The sheer number of
tests, currently over 1000, increases the likelihood that a bug will be caught
if it exists. The complexity of my tests, such as implementations of mergesort
and heapsort, test a variety of cases for errors. Despite the number and
complexity of tests, the compiler currently runs all tests within a modest 30
seconds through parallelization. The new tests have indeed caught a number of
bugs. Some have been small oversights in the logic of the code that were quick
fixes, while others were more deeply nested and required a more thorough
investigation of the logic to correct.

While creating tests, I sometimes duplicated the names of functions between a
test file and the prelude. These errors are easy to determine with a correct
error message, but such a message had not been implemented yet. To combat this, I undertook
the task of creating error messages that checked for illegal instances of
duplicate names. The names can be of functions, function arguments, variables,
data types, and data constructors. The error message below indicates a duplication of function argument names.

\hspace{1cm}\begin{verbbox}
add = FUN(x x -> plusInt x x);

*** Exception:  variable x duplicated 2 times in location: toplevel.add. -- 
\end{verbbox}
\theverbbox

These error messages have expedited the debugging process in a number
of situations, normlly for duplicate names but occasionally for missing
brackets and semicolons. This is a lasting contribution to the compiler,
incorporating a substantial amount of previous knowledge gained from the
earlier months of the internship.

\section{Extra: Grundy Graph Colorings}

Although the majority of my waking hours in Los Alamos, New Mexico were spent
at Los Alamos National Laboratory, I had enough spare time to embark on
projects of my own. One such project was to submit my paper, ``Edge Grundy numbers of $P_3 \Box P_n$ and $P_3 \Box C_n$,'' for
publication.$^1$ This paper filled a difficult gap in one of my previous submitted papers,
and time was of the essence: my graduate school applications were due
soon. The problem was to determine the edge Grundy number of $P_3 \Box P_n$ for $n \in \{5,6,7\}$ and $P_3 \Box C_n$ for $n \in \{4,5,6\}$. A graph coloring is \emph{Grundy} if it is a proper coloring with the positive integers such that if $v \in V(G)$ is colored with $c>1$, then all colors $1, \ldots, c-1$ appear on neighbors of $v$.

In spring 2015, I coded a Java program that answered the question; however, the program was much too verbose to display in a
publication. Under the knowledge that Haskell is concise, I decided to recode
the program in Haskell to test this assertion. The results were remarkable: the
text of the program fit in three pages under the standard font size. A second and
more important advantage to rewriting the code in Haskell was the ease of
citing the program's logic. Since functions are the building blocks of
functional languages, I could just cite functions that perform checks and
iterate through properties of data. Necessary checks include the proper and
Grundy conditions, and iterations include coloring a vertex all possible
colors. Due to the high-order nature of Haskell, I assembled a few specific
functions that completed each of these tasks. I proved that the program was correct by using mathematical induction and simply citing these
functions from my code - a far cry from the capabilities of imperative
languages! This paper was accepted to the \emph{International Journal of Mathematics and Computer Science} in November 2015.$^1$
\begin{center}
ACKNOWLEDGMENTS
\end{center}

This work was supported in part by the U.S. Department of Energy, Office of
Science, Office of Workforce Development for Teachers and Scientists (WDTS)
under the Science Undergraduate Laboratory Internships Program (SULI). The
author first thanks Dr. Kei Davis for his outstanding mentorship throughout
the course of his internship and also Dean Prichard for helpful advice. He thanks
both Los Alamos National Laboratory for hosting the SULI program and also the SULI
program itself for its funding. He thanks Los Alamos National Laboratory for the additional funding provided through the Laboratory Directed Research and Development LDRD 20150485ER. 
%-------------------------------------------------------BEGIN APPENDIX------------------------------------------------------------------------------------------
\section{Appendix}

\subsection{Arithmetic Parser and Evaluator}

The following is Haskell code for an arithmetic expression parser and evaluator.

\begin{verbatim}
import Data.Char
type Parser a b = [a] -> [(b,[a])]

succeed :: b -> Parser a b
succeed v inp = [(v, inp)]

failed :: Parser a b
failed inp = []

satisfy :: (Char -> Bool) -> Parser Char Char
satisfy p []     = fail []
satisfy p (x:xs) | p x       = succeed x xs
                 | otherwise = fail xs

literal :: Char -> Parser Char Char
literal x = satisfy (==x)

cons :: (a,[a]) -> [a]
cons (x,xs) = x:xs

alt :: Parser a b -> Parser a b -> Parser a b
(p1 `alt` p2) inp = p1 inp ++ p2 inp

thens :: Parser a b -> Parser a c -> Parser a (b,c)
(p1 `thens` p2) inp = [((v1, v2), out2) | (v1, out1) <- 
    p1 inp, (v2,out2) <- p2 out1]

xthen :: Parser a b -> Parser a c -> Parser a c
(p1 `xthen` p2) inp = ((p1 `thens` p2) `using` snd) inp

thenx :: Parser a b -> Parser a c -> Parser a b
(p1 `thenx` p2) inp = ((p1 `thens` p2) `using` fst) inp

using :: Parser a b -> (b -> c) -> Parser a c
(p `using` f) inp = [(f v, out) | (v,out) <- p inp]

many :: Parser a b -> Parser a [b]
many p = ((p `thens` many p) `using` cons)`alt` (succeed [])

some :: Parser a b -> Parser a [b]
some p = (p `thens` many p) `using` cons

number :: Parser Char [Char]
number = some (satisfy isDigit)

negNumber :: Parser Char [Char]
negNumber (x:xs) | x == '-' = (x:xs,""): (appendMinus(number xs))
                 | otherwise  = failed xs

appendMinus :: [([Char],[Char])] -> [([Char],[Char])]
appendMinus [] = []
appendMinus (x:xs) = (fst(x),'-':snd(x)):(appendMinus xs)

digit :: Parser Char Char
digit = satisfy isDigit

int :: Parser Char [Char]
int = number `alt` negNumber

optMinus :: Parser Char [Char]
optMinus (x:xs) | x == '-' = [("-",xs)]
           | otherwise = [("",(x:xs))]

concatPair :: [(([Char],[Char]),[Char])] -> [([Char],[Char])]
concatPair [] = []
concatPair (x:xs) = (fst(fst(x)) ++ snd(fst(x)),snd(x)):(concatPair xs)

opt :: Char -> Parser Char [Char]
opt c (x:xs) | x == c = [([c],xs)]
             | otherwise = [("", (x:xs))]

int2 :: Parser Char [Char]
int2 (x:xs) = concatPair (((opt '-') `thens` int) (x:xs))

data Expr = Add Term Term | Sub Term Term | Ter Term deriving (Show, Eq, Ord)
data Term = Mul Power Power | Div Power Power | Pow Power deriving (Show, Eq, Ord)
data Power = Po Factor Factor | Fact Factor deriving (Show, Eq, Ord)
data Factor = Num Int | Exp Expr deriving (Show, Eq, Ord)

expr :: Parser Char Expr
expr = ((term `thenx` (literal '+') `thens` term) `using` plus) 
    `alt` ((term `thenx` (literal '-') `thens` term) `using` minus) 
    `alt` (term `using` Ter)

term :: Parser Char Term
term = ((power  `thenx` (literal '*') `thens` power) `using` times) 
    `alt` ((power  `thenx` (literal '/') `thens` power) `using` divide) 
    `alt` (power `using` Pow)

power :: Parser Char Power
power = ((factor `thenx` (literal '^') `thens` factor) `using` exponentiate)
    `alt` (factor `using` Fact)

factor :: Parser Char Factor
factor = factorNum `alt` parenExpr

parse :: [Char] -> Expr
parse = fst . head . expr

factorNum :: Parser Char Factor
factorNum = number `using` value

parenExpr :: Parser Char Factor
parenExpr = ((literal '(') `xthen` expr `thenx` (literal ')')) `using` Exp

value :: String -> Factor
value = Num . read

plus :: (Term,Term) -> Expr
plus (x,y) = Add x y

minus :: (Term,Term) -> Expr
minus (x,y) = Sub x y

times :: (Power,Power) -> Term
times (x,y) = Mul x y

divide :: (Power,Power) -> Term
divide (x,y) = Div x y

exponentiate :: (Factor,Factor) -> Power
exponentiate (x,y) = Po x y

evaluate :: Expr -> Int
evaluate (Add x y) = (evaluate1 x) + (evaluate1 y)
evaluate (Sub x y) = (evaluate1 x) - (evaluate1 y)
evaluate (Ter x) = (evaluate1 x)

evaluate1 :: Term -> Int
evaluate1 (Mul x y) = (evaluate2 x) * (evaluate2 y)
evaluate1 (Div x y) = (evaluate2 x) `div` (evaluate2 y)
evaluate1 (Pow x) = (evaluate2 x)

evaluate2 :: Power -> Int
evaluate2 (Po x y) = (evaluate3 x) ^ (evaluate3 y)
evaluate2 (Fact x) = (evaluate3 x)

evaluate3 :: Factor -> Int
evaluate3 (Num x) = x
evaluate3 (Exp x) = evaluate x

evaluator :: [Char] -> Int
evaluator xs = evaluate (parse xs)

\end{verbatim}

\subsection{Reverse Polish Notation Parser and Evaluator}

The following is Haskell code for a reverse Polish notation expression parser and evaluator. It is necessary to implement the code from appendix 7.1 along with the below code.

\begin{verbatim}
data RPN = SPush Int | SAdd | SSub | SMul | SDiv | SPow deriving (Show)

toRPN :: [Char] -> [RPN]
toRPN xs = toRPN1 (parse xs)

toRPN1 :: Expr -> [RPN]
toRPN1 (Add x y) = (toRPN2 x) ++ (toRPN2 y) ++ [SAdd]
toRPN1 (Sub x y) = (toRPN2 x) ++ (toRPN2 y) ++ [SSub]
toRPN1 (Ter x) = (toRPN2 x)

toRPN2 :: Term -> [RPN]
toRPN2 (Mul x y) = (toRPN3 x) ++ (toRPN3 y) ++ [SMul]
toRPN2 (Div x y) = (toRPN3 x) ++ (toRPN3 y) ++ [SDiv]
toRPN2 (Pow x) = (toRPN3 x)

toRPN3 :: Power -> [RPN]
toRPN3 (Po x y) = (toRPN4 x) ++ (toRPN4 y) ++ [SPow]
toRPN3 (Fact x) = (toRPN4 x)

toRPN4 :: Factor -> [RPN]
toRPN4 (Num x) = [SPush x]
toRPN4 (Exp x) = toRPN1 x

terp1 :: [Int]-> [RPN] -> Int
terp1 [x] [] = x
terp1 xs ((SPush z):ys) = terp1 (z:xs) ys
terp1 (x1:x2:xs) ((SAdd):ys) = terp1 ((x2+x1):xs) ys
terp1 (x1:x2:xs) ((SSub):ys) = terp1 ((x2-x1):xs) ys
terp1 (x1:x2:xs) ((SMul):ys) = terp1 ((x2*x1):xs) ys
terp1 (x1:x2:xs) ((SDiv):ys) = terp1 ((x2`div`x1):xs) ys
terp1 (x1:x2:xs) ((SPow):ys) = terp1 (((x2)^(x1)):xs) ys

final :: [Char] -> Int
final xs = terp1 [] (toRPN xs)
\end{verbatim}

\subsection{Lambda Calculus Parser and Evaluator}

The following is Haskell code for a lambda calculus lexer, parser, and evaluator using variable threading under both applicative and also normal evaluation.

\begin{verbatim}

import Data.List
import Data.Char

type Parser a b = a -> [(b,a)]

succeed :: b -> Parser a b
succeed v inp = [(v, inp)]

failed :: Parser a b
failed inp = []

satisfy :: (Char -> Bool) -> Parser [Char] Char
satisfy p []     = fail []
satisfy p (x:xs) | p x       = succeed x xs
                 | otherwise = fail xs

literal :: Char -> Parser [Char] Char
literal x = satisfy (==x)

alt :: Parser a b -> Parser a b -> Parser a b
(p1 `alt` p2) inp = p1 inp ++ p2 inp

thens :: Parser a b -> Parser a c -> Parser a (b,c)
(p1 `thens` p2) inp = [((v1, v2), out2) | (v1, out1) <-
     p1 inp, (v2,out2) <- p2 out1]

xthen :: Parser a b -> Parser a c -> Parser a c
(p1 `xthen` p2) inp = ((p1 `thens` p2) `using` snd) inp

thenx :: Parser a b -> Parser a c -> Parser a b
(p1 `thenx` p2) inp = ((p1 `thens` p2) `using` fst) inp

using :: Parser a b -> (b -> c) -> Parser a c
(p `using` f) inp = [(f v, out) | (v,out) <- p inp]

cons :: (a,[a]) -> [a]
cons (x,xs) = x:xs

many :: Parser a b -> Parser a [b]
many p = ((p `thens` many p) `using` cons)`alt` (succeed [])

some :: Parser a b -> Parser a [b]
some p = (p `thens` many p) `using` cons

number :: Parser [Char] [Char]
number = some (satisfy isDigit)

word :: Parser [Char] [Char]
word = some (satisfy isAlpha)

space :: Parser [Char] Char
space = literal ' '

tab :: Parser [Char] Char
tab = literal '\t'

newLine :: Parser [Char] Char
newLine = literal '\n'

whitespace :: Parser [Char] Char
whitespace = space `alt` tab `alt` newLine

whitespaces :: Parser [Char] [Char]
whitespaces = many whitespace

satisfy1 :: (String -> Bool) -> Parser [String] String
satisfy1 p []     = failed []
satisfy1 p (x:xs) | p x       = succeed x xs
                  | otherwise = failed xs

literal1 :: String -> Parser [String] String
literal1 x = satisfy1 (==x)

lexer :: String -> [String]
lexer [] = []
lexer (x:xs) | x == '\n' = lexer(snd(head(whitespace (x:xs))))
             | x == '\t' = lexer(snd(head(whitespace (x:xs))))
             | x == ' ' = lexer(snd(head(whitespace (x:xs))))
             | x == '\\' = "\\":(lexer xs)
             | x == '(' = "(":(lexer xs)
             | x == ')' = ")":(lexer xs)
             | isDigit x = (fst(ys)):lexer(snd(ys))
             | isAlpha x = (fst(zs)):lexer(snd(zs))
               where ys = head(number(x:xs))
                     zs = head(word(x:xs))

data Expr = Var String 
          | Abs Expr Expr 
          | Comb Expr Expr  deriving (Eq, Read)

instance Show Expr where
    show (Var v) = v
    show (Abs e1 e2) = "(\\" ++ show e1 ++ "." ++ show e2 ++ ")"
    show (Comb e1 e2) = show e1 ++ " " ++ show e2

parse :: String -> Expr
parse = fst.head.lexpr.lexer

combing :: Expr -> Expr -> Expr
combing x y = Comb x y

lexpr :: Parser [String] Expr
lexpr = (some expr) `using` (foldl1 combing)

expr :: Parser [String] Expr
expr = var `alt` lambda `alt` parenExpr

var :: Parser [String] Expr
var = alphaNums `using` Var

lambda :: Parser [String] Expr
lambda = ((literal1 "\\") `thens` var `thens` lexpr) `using` lambdaing

lambdaing :: ((String,Expr),Expr) -> Expr
lambdaing ((x,y),z) = Abs y z

parenExpr :: Parser [String] Expr
parenExpr = ((literal1 "(") `xthen` lexpr  `thenx` (literal1 ")"))

alphaNums :: Parser [String] String
alphaNums [] = []
alphaNums (x:xs) | (isAlpha (head x) && wordt x)= succeed x xs
                 | otherwise = failed xs

wordt :: String -> Bool
wordt [] = True
wordt (x:xs) = if ((isAlpha x) || (isDigit x)) then wordt xs else False

freeVars :: Expr -> [String]
freeVars (Var x) = [x]
freeVars (Comb x y) = (freeVars x)++(freeVars y)
freeVars (Abs x y) = setDiff (freeVars x) (freeVars y) []

setDiff :: [String] -> [String] -> [String] -> [String]
setDiff [] ys _ = ys
setDiff (x:xs) [] zs = setDiff xs zs []
setDiff (x:xs) (y:ys) zs = if x == y then setDiff (x:xs) ys zs else 
    setDiff (x:xs) ys (y:zs)

betaReduce1 :: [String] -> Expr -> Expr -> Expr -> ([String],Expr)
betaReduce1 xs (Var a) (Var b) c = if a == b then (xs,c) else (xs,(Var a))
betaReduce1 xs (Comb a b) c d = reduction1 xs (Comb a b) c d
betaReduce1 xs (Abs (Var a) b) (Var c) d | a == c = (xs,(Abs (Var a) b))
                                         | (a /= c) && (not (a `elem` 
    (freeVars d))) = reduction2 xs (Abs (Var a) b) (Var c) d
                                         | otherwise = reduction3 xs 
    (Abs (Var a) b) (Var c) d

reduction1 :: [String] -> Expr -> Expr -> Expr -> ([String],Expr)
reduction1 xs (Comb a b) c d = (fst z,(Comb (snd y) (snd z)))
                               where y = betaReduce1 xs a c d
                                     z = (betaReduce1 (fst y) b c d)
reduction1 _ _ _ _ = error "Here 1"

reduction2 :: [String] -> Expr -> Expr -> Expr -> ([String],Expr)
reduction2 xs (Abs (Var a) b) (Var c) d = (fst y, Abs (Var a) (snd y))
                                          where y = (betaReduce1 xs
     b (Var c) d)
reduction2 _ _ _ _ = error "Here 2"

reduction3 :: [String] -> Expr -> Expr -> Expr -> ([String],Expr)
reduction3 xs (Abs (Var a) b) (Var c) d = 
    (fst v, Abs (Var (fst r)) (snd v))
     where r = nextFree xs (Abs (Var a) b) d
           t = (betaReduce1 (snd r) b  (Var a) (Var (fst r)))
           v = (betaReduce1 (fst t) (snd t) (Var c) d)
reduction3 _ _ _ _ = error "Here 3"

nextFree :: [String] -> Expr -> Expr -> (String,[String])
nextFree [] _ _ = error "ran out"
nextFree (x:xs) y z | x `elem` (freeVars y)++(freeVars z) =
     nextFree xs y z
                    | otherwise = (x,xs)

fvs :: [String]
fvs = ['v' : show i| i<- [0..]]

oneSplit :: [String] -> [String]
oneSplit (x:y:xs) = x:(oneSplit(xs))

twoSplit :: [String] -> [String]
twoSplit (x:y:xs) = y:(twoSplit(xs))

nor e = putStrLn $ intercalate "\n" $ map show $ startEvalNormOrder e
nor2 e = last (startEvalNormOrder e)
nor3 e = startLambToInt (nor2 e)

startEvalNormOrder :: Expr -> [Expr]
startEvalNormOrder expression = expression:(evaluateNormalOrder 
    (fvs,expression))

evaluateNormalOrder :: ([String],Expr) -> [Expr] 
evaluateNormalOrder (xs,expression) = case theEval (xs,expression) of
                                              Just a -> (snd a):
    (evaluateNormalOrder (fst a, snd a))
                                              Nothing -> []

normEval :: ([String],Expr) -> ([String],Expr)
normEval (ws,(Var x)) = (ws,(Var x))
normEval (ws,(Comb (Abs (Var x) y) z)) = (betaReduce1 ws y (Var x) z)
normEval (ws,(Abs x y)) = (t1,(Abs x t2))
                          where (t1,t2) = (normEval (ws,y))
normEval (ws,(Comb x y)) | (checkEval x) = (s1, Comb s2 y)
                         | (checkEval y) = (t1, Comb x t2)             
                         | otherwise = (ws,(Comb x y))
                                       where (s1,s2) = normEval (ws, x)
                                             (t1,t2) = normEval (ws, y)
theEval :: ([String],Expr) -> Maybe ([String],Expr) 
theEval (xs,expression) | checkEval expression = Just (normEval 
    (xs,expression))
                        | otherwise = Nothing

checkEval :: Expr -> Bool
checkEval (Var x) = False
checkEval (Comb (Abs (Var x) y) z) = True
checkEval (Abs x y) = checkEval y
checkEval (Comb x y) = (checkEval x) || (checkEval y)

nor1 e = putStrLn $ intercalate "\n" $ map show $ startEvalAppOrder e

startEvalAppOrder :: Expr -> [Expr]
startEvalAppOrder expression = expression:(evaluateApplicativeOrder 
    (fvs,expression))

evaluateApplicativeOrder :: ([String],Expr) -> [Expr]
evaluateApplicativeOrder (xs,expression) = 
    case anEval (xs,expression) of
                                             Just a -> 
    (snd a):(evaluateApplicativeOrder (fst a, snd a))
                                             Nothing -> []

anEval :: ([String],Expr) -> Maybe ([String],Expr) 
anEval (xs,expression) | checkEval expression = 
    Just (appEval (xs,expression))
                       | otherwise = Nothing

appEval :: ([String],Expr) -> ([String],Expr) 
appEval (ws,(Var x)) = (ws,(Var x)) 
appEval (ws,(Comb (Abs (Var x) y) z)) | y /= (snd (appEval (ws,y))) 
    = (s1, Comb (Abs (Var x) s2) z)
                                      | z /= (snd (appEval (ws,z))) 
    = (t1, Comb (Abs (Var x) y) t2)
                                      | otherwise =
    (betaReduce1 ws y (Var x) z)    
     where (s1,s2) = appEval (ws, y)
     (t1,t2) = appEval (ws, z)
appEval (ws,(Abs x y)) = (t1,(Abs x t2))
                         where (t1,t2) = (appEval (ws,y))
appEval (ws,(Comb x y)) | Comb (snd (appEval (ws, x))) y /=  (Comb x y) 
    = (s1, Comb s2 y)
                        | Comb x (snd (appEval (ws, y))) /= (Comb x y) 
    = (t1, Comb x t2)             
                        | otherwise = (ws,(Comb x y))
                                      where (s1,s2) = appEval (ws, x)
                                            (t1,t2) = appEval (ws, y)

identity = "(\\x x)"

zero = "(\\f \\x x)"
ones = "(\\f \\x f x)"
twos = "(\\f \\x f(f x))"
threes = "(\\f \\x f(f(f x)))"
fours = "(\\f \\x f(f(f(f x))))"
fives = "(\\f \\x f(f(f(f(f x)))))"

true1 = "(\\x \\y x)"
false1 = "(\\x \\y y)"
and1 = "(\\p \\q p q p)"
or1 = "(\\p \\q p p q)"
not1 = "(\\p \\a \\b p b a)"
ifThenElse = "(\\p \\a \\b p a b)"
isZero = "(\\n n (\\x (\\x \\y y)) (\\x \\y x))"

succ1  = "(\\n \\f \\x f (n f x))"
add1 = "(\\m (\\n (\\f (\\x m f (n f x)))))"
mult = "(\\g (\\h (\\i (g (h i)))))"
pow = "(\\m (\\e e m))"
pred1 = "(\\n (\\f (\\x n (\\g  (\\h h (g f))) (\\u x) (\\u u))))"
sub1 = "(\\m (\\n n " ++ pred1 ++ " m ) )"

startIntToLamb :: (Eq a, Num a) => a -> [Char]
startIntToLamb n = "(\\f (\\ x (" ++ (intToLamb n) ++ ")))"
intToLamb 0 = " x"
intToLamb n = " ( f " ++ (intToLamb (n-1)) ++ " ) "

startLambToInt :: Expr -> Int
startLambToInt (Abs x (Abs y z)) = lambToInt z
lambToInt (Var x) = 0
lambToInt (Comb x (Comb y z)) = 1  + lambToInt (Comb y z)
lambToInt (Comb (Var x) (Var y)) = 1

safeSub = "(\\t ( " ++ ifThenElse ++ "( " ++ isZero ++ " t ))" ++
     "( "++ identity ++ " t ) " ++ "( "++ sub1 ++ " t " ++ ones ++ " )) "

fix = "(\\g (\\x g (x x)) (\\x g (x x)))"

g = "(\\r (\\n (" ++ ifThenElse ++ "( " ++ isZero ++ " n ))" ++
     "( " ++ ones ++ ")" ++ "( " ++ mult ++ "(n)" ++ "( r (" ++ 
     sub1 ++ "(n)" ++ ones ++ ")))))"

fac0 = nor $ parse (fix ++ g ++ zero)
fac1 = nor $ parse (fix ++ g ++ ones)
fac2 = nor $ parse (fix ++ g ++ twos)
fac3 = nor $ parse (fix ++ g ++ threes)
fac4 = nor $ parse (fix ++ g ++ fours)
fac5 = nor $ parse (fix ++ g ++ fives)

pair = "(\\x (\\y (\\f f x y)))"
first = "((\\p p)" ++ true1 ++ ")"
second = "((\\p p)" ++ false1 ++ ")" 
nil1 = "(\\x " ++ true1 ++ ")"
null1 = "(\\p p (\\x (\\y " ++ false1 ++ ")))"

single = "(" ++ pair ++ " (first)" ++ nil1 ++ ")"
double = "(" ++ pair ++ " (first) " ++ "(" ++ pair ++ "(second )" 
    ++  nil1 ++ "))"
triple = "(" ++ pair ++ " (first) " ++ "(" ++ pair ++ "(second )" 
    ++ "( " ++ pair ++ "(third)" ++ nil1 ++ ")))"
singles = "(" ++ pair ++ twos ++ nil1 ++ ")"
singles1 = "(" ++ pair ++ zero ++ nil1 ++ ")"
doubles = "(" ++ pair ++ ones ++ "(" ++ pair ++ twos ++ nil1 ++ "))" 
triples = "(" ++ pair ++ twos ++ "(" ++ pair ++ threes ++ "(" ++ pair
     ++ ones ++ nil1 ++ ")))"

\end{verbatim}

%-------------------------------------------------------BEGIN BIBLIOGRAPHY---------------------------------------------------------------------------------------
\begin{thebibliography}{99}
\bibitem{Z} L. Anderson, ``Edge Grundy numbers of $P_3 \Box P_n$ and $P_3 \Box C_n$,'' Accepted to Int.\ J.\ Math.\ Comput.\ Sci.\
\bibitem{Y} K. Davis, ``Enabling automatic parallelization and transparent fault tolerance,'' LANL Laboratory Directed Research and Development proposal LDRD 20150485ER.
\bibitem{X} G. Hutton, ``Higher-order functions for parsing,'' J.\ Funct.\ Program.\ 2.3, 323-343 (1992).
\bibitem{W} G. Hutton, \textit{Programming in Haskell}. (Cambridge University Press, Cambridge, ENG, 2011). 
\bibitem{V} B. Kurtz and K. Slonneger, \textit{Formal Syntax and Semantics of Programming Languages}. (Addison-Wesley, Reading, MA, 1995).
\bibitem{U} R. Sedgewick and K. Wayne, \textit{Algorithms}, 4th ed. (Addison-Wesley, Upper Saddle River, NJ, 2011).

\end{thebibliography}

\end{document}
%-------------------------------------------------------END DOCUMENT---------------------------------------------------------------------------------------------

### 2015-12-18 keistuff/codegen/Lexer.hs
debugging \!USE_OBJTYPE

@@ -1,93 +1,0 @@
-module Lexer (
-  Token(..),
-  Keyword(..),
-  Object(..),
-  Primop(..),
-  Symbol(..),
-  lexer
-) where
-
-import Data.Char(isLower,isUpper)
-import Scanner
-
-data Keyword = KWlet | KWin | KWcase | KWof deriving(Eq,Show)
-
-data Object = OFUN | OPAP | OCON | OTHUNK | OBLACKHOLE deriving(Eq,Show)
-
-data Primop = Pplus 
-            | Psub 
-            | Pmult
-            | Peq
-            | PintToBool
-              deriving(Eq,Show)
-
-data Symbol = SymArrow 
-            | SymLParen 
-            | SymRParen 
-            | SymBind 
-            | SymLBrace 
-            | SymRBrace 
-            | SymSemi
-              deriving (Eq,Show)
-
-data Token = Number Int 
-           | Ident String
-           | KW Keyword
-           | Ctor String
-           | Obj Object
-           | Sym Symbol
-           | PO Primop
-           deriving(Show, Eq)
-
-bigtab = 
-    [("->",        Sym SymArrow),
-     ("(",         Sym SymLParen),
-     (")",         Sym SymRParen),
-     ("=",         Sym SymBind),
-     ("{",         Sym SymLBrace),
-     ("}",         Sym SymRBrace),
-     (";",         Sym SymSemi),
-     ("FUN",       Obj OFUN),
-     ("CON",       Obj OCON),
-     ("PAP",       Obj OPAP),
-     ("THUNK",     Obj OTHUNK),
-     ("ERROR",     Obj OBLACKHOLE),  -- for compatibility with BJP's ministg
-     ("let",       KW KWlet), 
-     ("in",        KW KWin), 
-     ("case",      KW KWcase), 
-     ("of",        KW KWof),
-     ("plus#",     PO Pplus), 
-     ("sub#",      PO Psub),
-     ("mult#",     PO Pmult),
-     ("eq#",       PO Peq),
-     ("intToBool#",PO PintToBool)
-    ]
-             
-
-lexer :: [Char] -> [Token]
-lexer = map trans . scanner
-
-trans :: Lexeme -> Token
-trans (ScanNum, str) = Number (read str)
-
-trans (ScanSym, str) = 
-    case lookupassoc bigtab str of
-      Nothing -> error "trans error"
-      Just s -> s
-
-trans (ScanIdent, str) = 
-    case lookupassoc bigtab str of
-      Just o -> o
[…]

### 2015-12-18 keistuff/codegen/Lexer2.hs

@@ -1,91 +1,0 @@
-module Lexer2 (
-
-) where
-
-import ParseComb
-import Scanner
-
-data KeyWord = KWlet 
-             | KWin 
-             | KWcase 
-             | KWof 
-               deriving(Eq,Show)
-
-keywordtab = [ ("let",  KWlet),
-               ("in",   KWin),
-               ("case", KWcase),
-               ("of",   KWof)
-             ]
-
-data Object = OFUN 
-            | OPAP 
-            | OCON 
-            | OTHUNK 
-            | OBLACKHOLE 
-              deriving(Eq,Show)
-
-objecttab = [("FUN",       OFUN),
-             ("PAP",       OPAP),
-             ("CON",       OCON),
-             ("THUNK",     OTHUNK),
-             ("BLACKHOLE", OBLACKHOLE)
-            ]
-
-data Symbol = SymArrow 
-            | SymLParen 
-            | SymRParen 
-            | SymBind 
-            | SymLBrace 
-            | SymRBrace 
-            | SymSemi
-              deriving (Eq,Show)
-
-symboltab = [("->", SymArrow),
-             ("(",  SymLParen),
-             (")",  SymRParen),
-             ("=",  SymBind),
-             ("{",  SymLBrace),
-             ("}",  SymRBrace),
-             (";",  SymSemi)
-            ]
-
-data Primop = Pplus 
-            | Pminus 
-            | Psub
-            | Pmult
-            | Peq
-            | PintToBool
-              deriving(Eq,Show)
-
-primoptab = [("plus#",     Pplus),
-             ("minus#",    Pminus),
-             ("sub#",      Psub),
-             ("mult
-
-data Token = Number Int 
-           | Ident String
-           | KW Keyword
-           | Ctor String
-           | Obj Object
-           | Sym Symbol
-           | PO Primop
-           deriving(Show, Eq)
-
-lexer :: [Lexeme] -> [Token]
-lexer = map trans
-
-trans :: Lexeme -> Token
-trans (ScanNum, str) = Number (read str)
-trans (ScanSym, str) = Sym $ lookupassoc symboltab str
-
-trans (ScanIdent, str) =
-    if elem str ["FUN","PAP","CON","THUNK","BLACKHOLE"] then (Obj, str)
-    else if elem str ["let", "in", "case", "of"] then (Keyword, str)
-    else if isupper (head str) then (Ctor, str)
-    else if (last str) == '#' then (Primop, str)
-    else (Ident, str)
-
-lookupassoc [] 
-
-
-

### 2015-12-18 keistuff/codegen/ParseComb.hs

@@ -1,142 +1,0 @@
-module ParseComb (
-  Parser,
-  succeedp,
-  failp,
-  satisfyp,
-  literalp,
-  altp,
-  thenp,
-  cutp,
-  usingp,
-  manyp,
-  somep,
-  maybep,
-  returnp,
-  listp,
-  xthenp,
-  thenxp,
-  anyp,
-  sepbyp,
-  anymapp,
-  optp,
-  exactp,
-
--- string/char specific
-
-  alphap,
-  numeralp,
-  alphaornumeralp,
-  whitep,
-) where
-
-import Data.Char
-
-type Parser input result = [input] -> [(result,[input])]
-
-succeedp :: b -> Parser a b -- b -> [a] -> [(b,[a])]
-succeedp v inp = [(v, inp)]
-
-failp :: Parser a b
-failp inp = []
-
--- predicate first element of input list
-satisfyp :: (a -> Bool) -> Parser a a
-satisfyp p [] = failp []
-satisfyp p (x:xs) | p x		= succeedp x xs
-	  	  | otherwise	= failp xs
-
--- match first element of list
-literalp :: Eq b => b -> Parser b b
-literalp x = satisfyp (==x)
-
--- alternative
-altp :: Parser a b -> Parser a b -> Parser a b
-(p1 `altp` p2) inp = p1 inp ++ p2 inp
-
--- sequence
-thenp :: Parser t1 t2 -> Parser t1 t3 -> Parser t1 (t2, t3)
-(p1 `thenp` p2) inp = [((v1, v2),out2) | (v1, out1) <- p1 inp, (v2, out2) <- p2 out1]
-
--- cut
-cutp msg p inp =
-    case p inp of
-      [] -> error ("cut error \"" ++ msg ++ "\" at \"" ++ concatMap show (take 15 inp))
-      r -> r
-
--- just transform output
-usingp :: Parser t1 t2 -> (t2 -> t3) -> Parser t1 t3
-(p `usingp` f) inp = [(f v,out) | (v,out) <- p inp]
-
--- uncurrieed cons
-ucons (x,xs) = x:xs
-
--- match list
-listp :: Eq a => [a] -> Parser a [a]
-listp [] = succeedp []
-listp (x:xs) = (literalp x `thenp` listp xs) `usingp` ucons
-
--- zero or more
-manyp :: Parser a b -> Parser a [b]
-manyp p = ((p `thenp` manyp p) `usingp` ucons) `altp` (succeedp [])
-
--- zero or one
-
[…]

### 2015-12-18 keistuff/codegen/Scanner.hs

@@ -1,69 +1,0 @@
-
-module Scanner (
-  ScanTag(..),
-  Lexeme,
-  scannerp,
-  scanner,
-  stgSyms,
-  uncomments, -- for debugging
-) where
-
-import ParseComb
-import Data.Char(isAlpha,isDigit)
-
-
--- the unary minus problem must be handled at the parsing level
-data ScanTag = ScanNum
-             | ScanSym
-             | ScanIdent
-             | ScanJunk  -- white space, could include comments
-             deriving(Eq, Show)
-
-type Lexeme = (ScanTag, [Char])
-
-identp = alphap `thenp` manyp alphaornumeralp `usingp` uncurry (:)
-
-identdecorp = 
-    identp `thenp` (       (listp "#")
-                    `altp` (manyp (literalp '\'')))       
-    `usingp` uncurry (++)
-
-natnump = somep numeralp
-
-whitespacep = somep whitep
-
-
-scp :: Parser Char [Char] -> ScanTag -> Parser Char Lexeme
-(p `scp` t) inp = [((t,xs),out) | (xs,out) <- p inp]
-
-scanp :: [(Parser Char [Char], ScanTag)] -> Parser Char [Lexeme]
-scanp = manyp . (foldr op failp)
-       where (p,t) `op` xs = (p `scp` t) `altp` xs
-
-stgSyms = ["->", "{", "}", ";", "\\", "(", ")", "="]
-
-scannerp :: Parser Char [Lexeme]
-scannerp = 
-    scanp [ (natnump, ScanNum),
-            (whitespacep,  ScanJunk),
-            (identdecorp,  ScanIdent),
-            (anymapp listp stgSyms, ScanSym)
-          ]
-
-scanner :: [Char] -> [Lexeme]
-scanner = filter ((/=ScanJunk) . fst) . fst . head . scannerp . uncomments
-
--- hack
-uncomments = unlines . (map uncomment) . lines
-
-uncomment [] = []
-uncomment ('#':xs) = []
-uncomment (x:xs) | isAlpha x    = x : uncommenta xs
-                 | otherwise    = x : uncomment xs
-
--- any character is allowed to follow an isAlpha
-uncommenta [] = []
-uncommenta (x:xs) | isAlpha x = x : uncommenta xs
-                  | otherwise = x : uncomment xs
-
-

### 2015-12-18 keistuff/codegen/callbyname-simp.hs

@@ -1,460 +1,0 @@
-
-
-type Parser a b = [a] -> [(b,[a])]
-
--- Always succeeds
-succeedp :: t1 -> Parser t t1
-succeedp v inp = [(v, inp)]
-
--- Always fails
-failp :: Parser b c
-failp inp = []
-
-
--- If list is not empty, and p of the first item is true, then succeed. Otherwise fail.
-satisfyp :: (b -> Bool) -> Parser b b
-satisfyp p [] = failp []
-satisfyp p (x:xs) | p x		= succeedp x xs
-	  	  | otherwise	= failp xs
--- satisfyp (>2) [4,5,6]     returns    [(4,[5,6])]
-
-
--- If first item of list ==x then succeed. Otherwise fail.
-literalp :: Eq b => b -> Parser b b
-literalp x = satisfyp (==x)
--- literalp 2 [2,3,4]    returns   [(2,[3,4])]
-
-
--- applies p1 and p2 to inp and returns the list of results
-altp :: Parser a b -> Parser a b -> Parser a b
-(p1 `altp` p2) inp = p1 inp ++ p2 inp
--- ((succeedp 5) `altp` (succeedp 4)) (succeedp 3 [])    returns    [(5,[(3,[])]),(4,[(3,[])])]
-
-
--- returns tuple of results to p1 and p2 inside tuple with inp
-thenp :: Parser t1 t2 -> Parser t1 t3 -> Parser t1 (t2, t3)
-(p1 `thenp` p2) inp = [((v1, v2),out2) | (v1, out1) <- p1 inp, (v2, out2) <- p2 out1]
--- ((succeedp 5) `thenp` (succeedp 4)) (succeedp 3 [])     returns     [((5,4),[(3,[])])]
-
-
--- returns results of p inp with f of p inp at the begining
-usingp :: Parser t1 t2 -> (t2 -> t3) -> Parser t1 t3
-(p `usingp` f) inp = [(f v,out) | (v,out) <- p inp]
--- ((failp) `usingp` cons) (succeedp 3 [3,4,5])		   returns    []
--- ((succeedp 2) `usingp` (+6)) (succeedp 3 [3,4,5])       returns    [(8,[(3,[3,4,5])])]
-
-
--- functions similar to BNF*  returns list of results inside tuple with remaining input inside list. Each other item in the list removes the end from the results list
-manyp :: Parser a b -> Parser a [b]
[…]

### 2015-12-18 keistuff/codegen/test.hs

@@ -1,6 +1,0 @@
-
-data T = T {name :: String} deriving(Show)
-
-f x name = x {name = name}
-
-y = T {}

### 2015-12-18 runtime/cmm.h

@@ -58,37 +58,6 @@ extern void initCmm();
     _POPVALS0();				\
   } while (0)
 
-#define CALL0_1(R1,F)				\
-  do {						\
-    _PUSHVALS0();				\
-    _CALL(F);					\
-    _POPVALS1(R1);				\
-  } while (0)
-
-#define CALL2_0(F,P1,P2)			\
-  do {						\
-    _PUSHVALS2(P1,P2);				\
-    _CALL(F);					\
-    _POPVALS0();				\
-  } while (0)
-
-
-#define CALL3_0(F,P1,P2,P3)			\
-  do {						\
-    _PUSHVALS3(P1,P2,P3);			\
-    _CALL(F);					\
-    _POPVALS0();				\
-  } while (0)
-
-
-#define CALL4_0(F,P1,P2,P3,P4)			\
-  do {						\
-    _PUSHVALS4(P1,P2,P3,P4);			\
-    _CALL(F);					\
-    _POPVALS0();				\
-  } while (0)
-
-
 #define RETURN0()				\
   do {						\
     _RETURN();					\

### 2015-12-18 runtime/gc.c

@@ -10,7 +10,7 @@
 #include "options.h"
 
 const bool DEBUG = true;
-const bool EXTRA = false;  // run extra checks
+const bool EXTRA = true;  // run extra checks
 
 #define EXTRASTART() fprintf(stderr, "EXTRA check file %s line %d\n", __FILE__, __LINE__)
 #define EXTRAEND() fprintf(stderr, "EXTRA check succeeded %s %d\n", __FILE__, __LINE__)
@@ -47,9 +47,13 @@ Obj *deref1(Obj *op) {
 
 Obj *deref2(Obj *op) {
   while (getObjType(op) == INDIRECT) {
-    if (isLSBset(op->infoPtr))
+    if (isLSBset(op->infoPtr)) {
       // it's not clear to me (yet) whether this can happen
+      // apparently it can...serious weird, only when USE_ARGTYPE and
+      // USE_OBJTYPE are both turned off
       assert(false && "INDIRECT/forward branch");
+      return op;
+    }
     op = op->payload[0].op;
   }
   return op;

### 2015-12-18 runtime/stg.c

@@ -77,7 +77,7 @@ Cont *stgAllocCont(CInfoTab *citp) {
   stgSP = (char *)stgSP - contSize;
   assert(stgSP >= stgStack);
   Cont *contp = (Cont *)stgSP;
-  contp->cinfoPtr = citp;
+  contp->cInfoPtr = citp;
   contp->_contSize = contSize;  // to go away
   contp->layout = layoutInfoToBitmap64(&citp->layoutInfo); // temp hack
   contp->entryCode = citp->entryCode;
@@ -98,7 +98,7 @@ Cont *stgAllocCallCont(CInfoTab *citp, int argc) {
   stgSP = (char *)stgSP - contSize;
   assert(stgSP >= stgStack);
   Cont *contp = (Cont *)stgSP;
-  contp->cinfoPtr = citp;
+  contp->cInfoPtr = citp;
   contp->_contSize = contSize;
   contp->entryCode = citp->entryCode;
   contp->contType = CALLCONT;
@@ -156,10 +156,12 @@ Obj* stgNewHeapObj(InfoTab *itp) {
   stgHP = (char *)stgHP + objSize;
   memset(objp, 0, objSize); //zero out anything left by previous gc passes
   objp->infoPtr = itp;
+  strcpy(objp->ident, itp->name);  // may be overwritten
 #if USE_OBJTYPE
+  fprintf(stderr, "stgNewHeapObj setting %s objType %s\n", 
+	  objp->ident, objTypeNames[itp->objType]);
   objp->objType = itp->objType;
 #endif
-  strcpy(objp->ident, itp->name);  // may be overwritten
   fprintf(stderr, "stgNewHeapObj: "); showStgObj(objp);
   return objp;
 }
@@ -176,15 +178,16 @@ Obj* stgNewHeapPAPmask(InfoTab *itp, Bitmap64 bm) {
   Obj *objp = (Obj *)stgHP;
   stgHP = (char *)stgHP + objSize;
   memset(objp, 0, objSize); //zero out anything left by previous gc passes
-  objp->infoPtr = setLSB2(itp); // set InfoPtr bit to say this is a PAP
-#if USE_OBJTYPE
-  objp->objType = PAP;
-#endif
 #if USE_ARGTYPE
   objp->payload[fvs].argType = LONG;
 #endif
   objp->payload[fvs].b = bm;
   strcpy(objp->ident, itp->name);  // may be overwritten
[…]

### 2015-12-18 runtime/stghs.c

@@ -242,7 +242,6 @@ void checkStgObjRec(Obj *p) {
   }
 
   InfoTab *itp = getInfoPtr(p);
-  assert((uintptr_t)itp % 8 == 0 && "hc: bad infoPtr alignment");
   InfoTab it = *itp;
 
   for (i = 0; i != depth; i++) if (p == stack[i]) return;

### 2015-12-18 runtime/stgutils.c

@@ -132,18 +132,14 @@ DEFUN0(stgUpdateCont) {
 
   // the order of the following two operations is important for concurrency
   p.op->payload[0] = stgCurVal;
-  p.op->infoPtr = &it_stgIndirect; // this will supersede the # if below
-
+  p.op->infoPtr = &it_stgIndirect;
+#if USE_OBJTYPE
+  p.op->objType = INDIRECT;
+#endif
   strcpy( p.op->ident, it_stgIndirect.name );
   int newObjSize = getObjSize(p.op);
   assert(newObjSize <= oldObjSize);
   memset((char*)p.op+newObjSize, 0, oldObjSize-newObjSize);
-
-#if USE_OBJTYPE
-  p.op->objType = INDIRECT;
-#else
-  p.op->infoPtr = setLSB3(p.op->infoPtr);
-#endif
   fprintf(stderr, "stgUpdateCont leaving...\n  ");
   STGRETURN0();
   ENDFUN
@@ -187,12 +183,12 @@ void stgThunk(PtrOrLiteral self) {
   strcpy(contp->ident, self.op->ident); //override default
   // can't do this until we capture the variables in a stack frame
   // self.op->infoPtr = &it_stgBlackHole;
+  fprintf(stderr, "BLACKHOLING %s\n", self.op->ident);
 #if USE_OBJTYPE
   self.op->objType = BLACKHOLE;
-#else
-  self.op->infoPtr = setLSB2(self.op->infoPtr); // update bit to say this is a Blackhole
 #endif
-
+  //  self.op->infoPtr = setLSB2(self.op->infoPtr); // this is a Blackhole
+  assert(getObjType(self.op) == BLACKHOLE);
 }
 
 DEFUN0(stgCallCont) {

### 2015-12-18 loren/SULI/finalReport.tex
final drafts

@@
 In pure lambda calculus, lambda expressions are evaluated through a process called reduction.
-This is primarily through substituting a free variable for an expression in a lambda expression.
-In general, an expression \texttt{$(\lambda$ x -> E) E1} is reduced by substituting \texttt{E1} in for all variables \texttt{x} in \texttt{E}.
+In general, an expression \texttt{$(\lambda$x -> E) E1} is reduced by substituting \texttt{E1} in for all free variables \texttt{x} in \texttt{E}.
 
@@
 
-If in fact a free variable becomes bound as a result of the substitution, the reduction involved \emph{variable capture}.
-Variable capture is one of the main difficulties of reducing lambda expressions.
+If in fact a free variable becomes bound as a result of the substitution, the reduction involved variable capture; variable capture is one of the main difficulties of reducing lambda expressions.
 The state of variable capture is shown below with the generic addition function.
@@
 Alpha reduction allows one to change the bound variables of a lambda expression barring variable capture.
-Note that conversion does not change the lambda expression; both the original and new lambda expressions are equivalent.
+Note that conversion does not change the lambda expression: both the original and new lambda expressions are equivalent.
 An example alpha conversion from the previous example is shown below.
@@
 
-Due to its frequent appearance, \texttt{($\lambda$v -> E) E1} is called a $\beta$-redex or ``reduction expression". Beta reduction is the most involved rule of evaluation in lambda calculus.
+Due to its frequent appearance, \texttt{($\lambda$v -> E) E1} is called a $\beta$-redex or ``reduction expression." Beta reduction is the most involved rule of evaluation in lambda calculus.
 
@@
 
-We want a methodical approach to choose which beta redexes to reduce first to get to normal form.
+We want a methodical approach to choose which $\beta$-redexes to reduce first to get to normal form.
 First, not every lambda expression can be reduced to a normal form due to the classic example illustrated below.
@@
 
-For any lambda expression of the form \emph{\texttt{E = (($\lambda$x -> B) A)}}, we say that beta redex \emph{\texttt{E}} is outside any beta redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}.
+For any lambda expression of the form \emph{\texttt{E = (($\lambda$x -> B) A)}}, we say that $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}.
 A $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}.
@@
 
-The example  below distinguish the differences between outermost and innermost reductions.
+The example  below distinguishes the differences between outermost and innermost reductions.
 
 \begin{center}
-\texttt{($\lambda$x -> ($\lambda$y -> (add x (($\lambda$x -> (sub x 3)) y)))) 5 6}\parskip 0pt
+\texttt{($\lambda$x -> ($\lambda$y -> (add x (($\lambda$x -> (sub x 3)) y)))) 5 6}
 \end{center}
@@
 \multicolumn{1}{c}{ \underline{Normal Order Reduction}} & \multicolumn{1}{c}{ \underline{Applcative Order Reduction}}\\
-\texttt{= ($\lambda$y -> (add 5 (($\lambda$x -> (sub x 3)) y))) 6}&\texttt{= ($\lambda$x -> ($\lambda$y -> (add x ((sub y 3))))) 5 6}\\
-\texttt{= (add 5 (($\lambda$x -> (sub x 3)) 6))}&\texttt{= ($\lambda$y -> (add 5 ((sub y 3)))) 6}\\
-\texttt{= (add 5 ((sub 6 3)))}&\texttt{= (add 5 ((sub 6 3)))}\\
-\texttt{= (add 5 3)}&\texttt{= (add 5 3)}\\
-\texttt{= 8}&\texttt{= 8}\\
+\texttt{=($\lambda$y -> (add 5 (($\lambda$x ->(sub x 3))y))) 6}&\texttt{=($\lambda$x -> ($\lambda$y -> (add x((sub y 3))))) 5 6}\\
+\texttt{=(add 5 (($\lambda$x -> (sub x 3)) 6))}&\texttt{=($\lambda$y -> (add 5 ((sub y 3)))) 6}\\
+\texttt{=(add 5 ((sub 6 3)))}&\texttt{=(add 5 ((sub 6 3)))}\\
+\texttt{=(add 5 3)}&\texttt{=(add 5 3)}\\
+\texttt{=8}&\texttt{=8}\\
 \end{tabular}
@@
 
-Church's Thesis states that lambda calculus is Turing complete.
+Church's Thesis states that lambda calculus is Turing-complete.
 Motivated by this fact, we demonstrate how to define nonnegative integers, basic mathematical functions, logical operators, list constructors, and how to simulate recursion.
@@
 
-Other arithmetic operations such as subtract, multiply, exponentiate.
-The next set of lambda expressions is the set of arithmetic operators.
-The boolean values true and false are the foundations of the logical operators and many other functions in general.
-As in the case of defining the nonnegative integers, the stated definitions below of true and false are not the only way to define true, but they may be the most efficient.
-The operators and, or, and not are also defined below, and they are built from the definitions of true and false.
+Other arithmetic operations implemented include \texttt{subtract}, \texttt{multiply}, and \texttt{exponentiate}.
+The next set of lambda expressions is the set of logical operators.
+The boolean values \texttt{true} and \texttt{false} are the foundations of the logical operators and many other functions in general.
+As in the case of defining the nonnegative integers, the stated definitions below of \texttt{true} and \texttt{false} are not the only definitions, but they may be the most efficient.
+The operators \texttt{and}, \texttt{or}, and \texttt{not} are also defined below, and they are built from the definitions of \texttt{true} and \texttt{false}.
 
@@
 Conditional statements are left to the appendix.
-It is possible to contruct lists of arbitrary sizes by first constructing a \texttt{pair} function and then continually nesting \texttt{pair} functions in the second slot to the desired length (see appendix).
+It is possible to contruct lists of arbitrary sizes by first constructing a \texttt{pair} function and then continually nesting \texttt{pair} functions in the second slot to the desired length (see appendix 7.3).
 Finally, the most efficient way to simulate recursion is through the fixed point combinator \texttt{Y}.
@@
 
-As functions do not have names in pure lambda calculus, it is impossible for a function to achieve recursion through calling itself. Luckily, the fixed point combinator acts in the following way:
+As functions do not have names in pure lambda calculus, it is impossible for a function to achieve recursion through calling itself. Luckily, the fixed point combinator acts as follows:
 
@@
 True to its name, \texttt{f} is a fixed point of \texttt{Y}.
-The key to exploiting the property of the fixed-point combinator is to clevely set up a conditional statement within the function.
+The key to exploiting the property of the fixed-point combinator is to clevely set up a conditional statement within the input function.
 The pseudocode of the \texttt{factorial} function is displayed below with a small evaluation in impure lambda calculus.
@@
 \small
-\hspace{2cm}\texttt{fac = ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))}
-
-\hspace{2cm}\texttt{Y fac 3}
-
-\hspace{2.5cm}\texttt{= fac (Y fac) 3}
-
-\hspace{2.5cm}\texttt{= ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 3}
-
-\hspace{2.5cm}\texttt{= (if 3 = 0 then 1 else 3 $\times$ ((Y fac) (3 - 1)))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ ((Y fac) 2)}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (fac (Y fac) 2)}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 2)}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ ((if 2 = 0 then 1 else 2 $\times$ ((Y fac) (2 - 1))))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ ((Y fac) 1))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (fac (Y fac) 1))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 1))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (if 1 = 0 then 1 else 1 $\times$ ((Y fac) (1 - 1))))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ ((Y fac) 0))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (fac (Y fac) 0))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))(Y fac) 0))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (if 0 = 0 then 1 else 0 $\times$ ((Y fac) (0 - 1)))))}
-
-\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ 1)) = 6}
+\hspace{0.5cm}\texttt{fac = ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))}
+
+\hspace{0.5cm}\texttt{Y fac 3}
+
+\hspace{1cm}\texttt{= fac (Y fac) 3}
+
+\hspace{1cm}\texttt{= ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 3}
+
+\hspace{1cm}\texttt{= (if 3 = 0 then 1 else 3 $\times$ ((Y fac) (3 - 1)))}
+
+\hspace{1cm}\texttt{= 3 $\times$ ((Y fac) 2)}
+
+\hspace{1cm}\texttt{= 3 $\times$ (fac (Y fac) 2)}
+
+\hspace{1cm}\texttt{= 3 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 2)}
+
+\hspace{1cm}\texttt{= 3 $\times$ ((if 2 = 0 then 1 else 2 $\times$ ((Y fac) (2 - 1))))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ ((Y fac) 1))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (fac (Y fac) 1))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 1))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (if 1 = 0 then 1 else 1 $\times$ ((Y fac) (1 - 1))))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ ((Y fac) 0))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (fac (Y fac) 0))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))(Y fac) 0))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (if 0 = 0 then 1 else 0 $\times$ ((Y fac) (0 - 1)))))}
+
+\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ 1)) = 6}
 \normalsize
@@
 
-My ultimate task was to create a lambda parser and evaluator in Haskell. To
-do this, I first read """""""""""' and completed all exercises.  The
+My overall task was to create a lambda parser and evaluator in Haskell. To
+do this, I first read \textit{Formal Syntax and Semantics of Programming Languages} by Kurtz and Slonneger and completed all exercises.$^5$  The
 content of the material was similar to the content presented here in the
-previous sections. Many exercises were detailed and required $\beta$
+previous sections. Many exercises were detailed and required beta
 reduction to be completed by hand. Following, I created a lambda parser and
@@
 evaluation. This list could be split in half when a substitution needed to
-be made in a combined expression (case d) in definition 4. The easiest way
+be made in a combined expression (case d in Definition 4). The easiest way
 to split the list in half was to create two new lists by appending every
@@
 was sufficient. The second implementation was to thread the variables
-through the program. After a expression was reduced, the fresh variable
+through the expression. After a expression was reduced, the fresh variable
 supply was passed to inner expressions to complete any reductions. Once all
@@
 passed to another expression until the entire expression was in normal
-form. This method is displayed in appendix .... After creating the parser
+form. This method is displayed in appendix 7.3. After creating the parser
 and evaluator, I constructed the nonnegative integers, arithmetic
 expressions, logical operators, and list operations. The final exercise was
-to code enough functions to parse and evaluate the factorial function in
-lambda calculus, as shown above.
+to code enough functions to parse and evaluate the \texttt{factorial} function in pure
+lambda calculus.
 

### 2016-02-10 pubs/tfp16/paper.tex
last night insomnia

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and Dean Prichard\inst{1} \and David Ringo\inst{1,2}}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA}

\maketitle              % typeset the title of the contribution

\begin{abstract}
The trend is functional programming in scientific computing, and in particular
strict (by default) functional programming, in various guises.  The project is
the demonstration of a light-weight, higher-order, polymorphic,
strict-by-default purely functional language implementation in which we can
experiment with automatic parallelization strategies and varying degrees of
default strictess.

\keywords{strict pure functional programming, automatic parallelization}

\end{abstract}

\section{Background and Motivation}

\section{Goals}

\section{Design and Implementation}

\section{Current Status and Expected Results (by the time of the symposium)}

%
% ---- Bibliography ----
%
\begin{thebibliography}{5}

\end{thebibliography}

\end{document}

### 2016-02-10 codegen/ADT.hs
just whitespace changes

@@ -10,10 +10,10 @@ module ADT (
   Polytype(..),
   Monotype(..),
   Con,
-  biIntMCon,    
-  biLongMCon,   
-  biFloatMCon,  
-  biDoubleMCon, 
+  biIntMCon,
+  biLongMCon,
+  biFloatMCon,
+  biDoubleMCon,
   dataConName,
   tyConName,
   getDataCons,
@@ -49,7 +49,7 @@ import PPrint
   data unboxed \Chi# \alpha_1 .. \alpha_t =
   c_1# \tau_11 .. \tau_1a_1 | ... | c_n# \tau_n1 .. \tau_na_1
   
-  Polytype      \sigma ::=  \forall \alpha . \sigma 
+  Polytype      \sigma ::=  \forall \alpha . \sigma
                         |   \tau
 
   Monotype      \tau   ::=  \pi | \nu
@@ -72,14 +72,14 @@ data Def a = ObjDef (Obj a)
              deriving(Eq, Show)
 
 -- Boxed: data \Chi \alpha_1 .. \alpha_t =
--- c_1 \tau_11 .. \tau_1a_1 | ... | c_n \tau_n1 .. \tau_na_1  
+-- c_1 \tau_11 .. \tau_1a_1 | ... | c_n \tau_n1 .. \tau_na_1
 -- Unboxed: data unboxed \Chi# \alpha_1 .. \alpha_t =
--- c_1# \tau_11 .. \tau_1a_1 | ... | c_n# \tau_n1 .. \tau_na_1   
+-- c_1# \tau_11 .. \tau_1a_1 | ... | c_n# \tau_n1 .. \tau_na_1
 data TyCon = TyCon Bool Con [TyVar] [DataCon]
              deriving(Eq,Show)
              
--- Boxed True: c_x \tau_x1 .. \tau_xa_1 
--- Boxed False: c_x# \tau_x1 .. \tau_xa_1  
+-- Boxed True: c_x \tau_x1 .. \tau_xa_1
+-- Boxed False: c_x# \tau_x1 .. \tau_xa_1
 data DataCon = DataCon Con [Monotype]
                deriving(Eq,Show)
 
@@ -116,10 +116,10 @@ isBoxed m = case m of
 boxMTypes :: [TyCon] -> [TyCon]
 boxMTypes tycons =
   let -- create assoc list for TyCon names -> TyCons
-      tycons' = makeIntTyCon "0" : 
-                makeLongTyCon "0" : 
[…]

### 2016-02-10 codegen/AST.hs

@@ -31,8 +31,8 @@ import Data.Int as Int (Int64)
 -- not really the place for this, maybe need to factor
 -- the common types into a module
 --data BuiltinType = UBInt
---                 | UBDouble             
---                   deriving (Eq,Show,Ord)           
+--                 | UBDouble
+--                   deriving (Eq,Show,Ord)
 
 type Var = String
 type Con = String
@@ -87,7 +87,7 @@ projectAtoms (a:as) = error $ "InfoTab.projectAtoms: non-EAtom"
 -- code is generated for them.
 
 data Primop = Piadd -- Int -> Int -> Int
-            | Pisub 
+            | Pisub
             | Pimul
             | Pidiv
             | Pimod
@@ -112,8 +112,8 @@ primArity op = case op of
   _ -> 2
 
 -- these are the C names, not STG names
-primopTab = 
-    [("iplus_h",  Piadd), 
+primopTab =
+    [("iplus_h",  Piadd),
      ("isub_h",   Pisub),
      ("imul_h",   Pimul),
      ("idiv_h",   Pidiv),
@@ -171,7 +171,7 @@ preludeObjNames =
    "take"]
 
 
--- functions for removing some or all standard Prelude objects from the 
+-- functions for removing some or all standard Prelude objects from the
 -- list of objects in a parsed STG program.  Useful for debugging.
 rmPrelude :: [Obj a] -> [Obj a]
 rmPrelude = rmPreludeLess []
@@ -194,7 +194,7 @@ instance Unparse Atom where
   unparse (LitC c) = text c
 
 instance Unparse Primop where
-  unparse = text.stgPrimName 
+  unparse = text.stgPrimName
 
 instance Unparse a => Unparse (Alt a) where
   unparse ACon{amd, ac, avs, ae} =
@@ -263,7 +263,7 @@ instance Unparse a => Unparse (Obj a) where
 
 
 instance Unparse a => Unparse [Obj a] where
-  unparse objs = vcat $ postpunctuate semi $ map unparse objs  
+  unparse objs = vcat $ postpunctuate semi $ map unparse objs
          
 
 
@@ -290,14 +290,14 @@ instance (PPrint a) => PPrint (Obj a) where
                (text "FUN:" $+$
                 nest 2
                 (text "name:" <+> text oname $+$
-                 text "vars:" <+> listText vs $+$                        
+                 text "vars:" <+> listText vs $+$
                  text "metadata:" $+$
                  nest 2 (pprint omd) $+$
[…]

### 2016-02-10 codegen/Analysis.hs

@@ -108,7 +108,7 @@ propCallsAlt env scrut a = case a of
     let env' = deleteAll avs env
     in a{ ae = propCallsExpr env' ae }
   
-  ADef{amd, av, ae}  -> 
+  ADef{amd, av, ae}  ->
     let env' = case knownFunExprIT env scrut of
           Just it -> Map.insert av it env -- if scrut is known function, bind var in env
           Nothing -> Map.delete av env -- else honor shadowing and delete it
@@ -188,7 +188,7 @@ instance SetHA (Expr InfoTab) where
             then False
             else
               case Map.lookup (name it) fmp of
-               Just b -> b 
+               Just b -> b
                Nothing -> False -- unknown function may grow heap
           Nothing ->
             -- this works independently of knownCall analysis (for simple tests)
@@ -241,9 +241,9 @@ instance SetHA (Alt InfoTab) where
 
 addDefsToMap defs funmap =
   let
-    fmp  = deleteAll (map oname defs) funmap -- remove shadowed bindings 
+    fmp  = deleteAll (map oname defs) funmap -- remove shadowed bindings
     toAdd = Map.fromList $ map ((,True) . oname) $ filter isKFun defs
-    fmp' = Map.union toAdd fmp 
+    fmp' = Map.union toAdd fmp
     isKFun o = case o of
       FUN{} -> True
       PAP{} -> False -- Don't want PAPs for now
@@ -370,7 +370,7 @@ knownFunExprIT env e =
       [its] -> head its
       _ -> Nothing
 
-    -- check single alt  
+    -- check single alt
     checkAlt env e ADef{av, ae} = case knownFunExprIT env e of
       Just it -> knownFunExprIT (Map.insert av it env) ae
       Nothing -> knownFunExprIT env ae

### 2016-02-10 codegen/BU.hs

@@ -25,7 +25,7 @@ import Control.Monad.State
 -- import AST(Var) -- shouldn't depend on AST
 type Var = String
 
-type Subst = Map.Map TyVar Monotype 
+type Subst = Map.Map TyVar Monotype
 idSubst = Map.empty
 
 compose :: Subst -> Subst -> Subst
@@ -115,7 +115,7 @@ unifys (x:xs) (y:ys) =
 -- need a fresh variable supply
 
 freshTyVar :: State Int TyVar
-freshTyVar = 
+freshTyVar =
     do i <- get
        put $ i+1
        return $ 't':show i
@@ -127,7 +127,7 @@ freshPolyVar =
        return $ MPVar v
 
 freshPolyVars 0 = return []
-freshPolyVars n = 
+freshPolyVars n =
     do
       v <- freshPolyVar
       vs <- freshPolyVars (n-1)
@@ -139,7 +139,7 @@ freshMonoVar =
        return $ MVar v
 
 freshMonoVars 0 = return []
-freshMonoVars n = 
+freshMonoVars n =
     do
       v <- freshMonoVar
       vs <- freshMonoVars (n-1)
@@ -162,7 +162,7 @@ monoToPoly as m = error $
                   "HMStg.monoToPoly: m=" ++ show m
 
 generalize :: Substitutable a => a -> Monotype -> Polytype
-generalize ms t = 
+generalize ms t =
     let as = Set.difference (freevars t) (freevars ms)
     -- necessary?  as in t become MPVar--not necessary, just skating on thin ice
     --    t' = monoToPoly as t
@@ -179,7 +179,7 @@ data Constraint = EqC  Monotype Monotype
 instance Show Constraint where
     show (EqC m1 m2) = "EqC " ++ show m1 ++ " , " ++ show m2
     show (ExpC m p) = "ExpC " ++ show m ++ " , " ++ show p
-    show (ImpC m1 ms m2) = "ImpC " ++ show m1 ++ " " ++ 
+    show (ImpC m1 ms m2) = "ImpC " ++ show m1 ++ " " ++
                            show (Set.toList ms) ++ " " ++ show m2
 
 type Constraints = Set.Set Constraint
@@ -234,18 +234,18 @@ solve1 :: [Constraint] -> [Constraint] -> Bool -> State Int Subst
 solve1 [] [] _ = return idSubst
[…]

### 2016-02-10 codegen/CAST.hs

@@ -224,7 +224,7 @@ cInitCall :: String -> [CExpr] -> Maybe CInit
 cInitCall name args =  Just (CInitExpr (cCallExpr name args) undefNode)
 
 cNewHeapObj :: String -> String -> CBlockItem
-cNewHeapObj name val = cUserPtrDecl "Obj" name (cInitCall "stgNewHeapObj" 
+cNewHeapObj name val = cUserPtrDecl "Obj" name (cInitCall "stgNewHeapObj"
                        [cAddrvarE ("it_" ++ val)])
 
 cNewCont :: String -> String -> CBlockItem

### 2016-02-10 codegen/CMap.hs

@@ -1,11 +1,11 @@
 {-# LANGUAGE FlexibleInstances    #-}
 {-# LANGUAGE TypeSynonymInstances #-}
 {-# LANGUAGE CPP #-}
+
 #if __GLASGOW_HASKELL__ < 710
 {-# LANGUAGE OverlappingInstances #-}
 #endif
 
-{-# LANGUAGE CPP #-}
 #include "options.h"
 
 ----------------------------  Alternative interface to ConMap idea ----------------------

### 2016-02-10 codegen/CodeGen.hs

@@ -77,9 +77,9 @@ useArgType = False
 data RVal = SO              -- static object
           | HO String       -- Heap Obj, payload size, TO GO?
           | FP String Int   -- stack Formal Param, pointer to stack payload
-          | FV String Int   -- Free Variable, payload in heap via pointer to 
+          | FV String Int   -- Free Variable, payload in heap via pointer to
                             --   pointer in stack, e.g. fvpp->op->payload[i]
--- because we don't have fresh names for the Let Blocks we use existing 
+-- because we don't have fresh names for the Let Blocks we use existing
 -- names for let-bound variables for now.  We could have done the same
 -- dereference scheme for FP above
             deriving(Eq,Show)
@@ -192,7 +192,7 @@ listLookup k ((k',v):xs) | k == k' = Just v
                          | otherwise = listLookup k xs
 
 getEnvRef :: String -> Env -> String
-getEnvRef v kvs = 
+getEnvRef v kvs =
     case listLookup v kvs of
       Nothing -> error $ "getEnvRef " ++ v ++ " failed"
       Just k ->
@@ -285,7 +285,7 @@ cgo env o@(FUN it vs e name) =
             "// " ++ name ++ "(self, " ++ intercalate ", " vs ++ ")\n" ++
             "FnPtr fun_" ++ name ++ "() {\n" ++
             "  fprintf(stderr, \"" ++ name ++ " here\\n\");\n" ++
-            "  PtrOrLiteral *" ++ argp ++ 
+            "  PtrOrLiteral *" ++ argp ++
                  " = &(stgGetStackArgp()->payload[0]);\n" ++
                indent 2 inline ++
                optStr (ypn /= Yes) "  STGRETURN0();\n" ++
@@ -337,7 +337,7 @@ stgApplyGeneric env f eas direct =
              else f
         inline =
             -- new STACKFRAME
-            "{ Cont *cp = stgAllocCallOrStackCont( &it_stgStackCont, " ++ 
+            "{ Cont *cp = stgAllocCallOrStackCont( &it_stgStackCont, " ++
                  show (length pnstring + 1) ++ ");\n" ++
             "  cp->layout = " ++ npStrToBMStr ('P' : pnstring ) ++ ";\n" ++
             "  cp->payload[ 0 ] = " ++ cgv env f' ++ ";\n" ++
[…]

### 2016-02-10 codegen/Driver.hs

@@ -1,4 +1,5 @@
 {-# LANGUAGE CPP #-}
+
 #include "options.h"
 
 module Driver (

### 2016-02-10 codegen/DupCheck.hs

@@ -2,7 +2,7 @@
 
 module DupCheck (
  dupCheck
-) where 
+) where
 
 -- [X] Need to check for duplicates in
 --  [X] Objects/expressions
@@ -24,18 +24,18 @@ import Data.List
 dupCheck :: ([TyCon], [Obj ()]) -> ([TyCon], [Obj ()])
 dupCheck (ts,os) = let odups = dupCheckObjs os ("toplevel":[])
                        tdups = dupCheckTyCons ts ++ dupCheckDataCons ts
-                   in case (odups ++ tdups) of 
+                   in case (odups ++ tdups) of
                       [] -> (ts,os)
-                      ys -> error ys 
+                      ys -> error ys
 
 --               --------------Begin OBJ/EXPR/ALTS/ALT Recursive Traversal------------
 -- Checks the list of Objects
 dupCheckObjs :: [Obj ()] -> [String] -> String
-dupCheckObjs os ys = dupCheckTopLevelObjects os ys ++ concatMapGen dupCheckObj os ys 
+dupCheckObjs os ys = dupCheckTopLevelObjects os ys ++ concatMapGen dupCheckObj os ys
 
 -- checks OBJ
 dupCheckObj :: Obj () -> [String] -> String
-dupCheckObj x ys = case x of 
+dupCheckObj x ys = case x of
                      PAP{} -> []
                      BLACKHOLE{} -> []
                      CON{} -> []
@@ -44,7 +44,7 @@ dupCheckObj x ys = case x of
 
 -- checks EXPR
 dupCheckExpr :: Expr () -> [String] -> String
-dupCheckExpr expr ys = case expr of 
+dupCheckExpr expr ys = case expr of
                          EAtom{} -> []
                          EFCall{} -> []
                          EPrimop{} -> []
@@ -53,12 +53,12 @@ dupCheckExpr expr ys = case expr of
 
 -- checks ALTS
 dupCheckAlts :: Alts () -> [String] -> String
-dupCheckAlts a ys = case a of 
+dupCheckAlts a ys = case a of
                       Alts{alts,aname} -> dupCheckCons alts (ys ++ (aname:[])) ++ concatMapGen dupCheckAlt alts (ys ++ (aname:[]))
 
 -- Checks ALT
 dupCheckAlt :: Alt () -> [String] -> String
-dupCheckAlt a ys = case a of 
+dupCheckAlt a ys = case a of
[…]

### 2016-02-10 codegen/HM.hs

@@ -8,7 +8,7 @@ import qualified Data.Set as Set
 import qualified Data.Map as Map
 import Control.Monad.State
 
-type Subst = Map.Map TyVar Monotype 
+type Subst = Map.Map TyVar Monotype
 idSubst = Map.empty
 
 compose :: Subst -> Subst -> Subst
@@ -101,7 +101,7 @@ data LExpr = LLitI   Monotype Int
 -- need a fresh variable supply
 
 freshTyVar :: State Int Monotype
-freshTyVar = 
+freshTyVar =
     do i <- get
        put $ i+1
        return $ MVar $ 't':show i
@@ -123,12 +123,12 @@ monoInt  = MCon True "I" []
 -- newtype State s a = State (s -> (a, s))
 algW :: TypeEnv -> LExpr -> State Int (LExpr, Subst, Monotype)
 
-algW env (LLitI _ i) = 
+algW env (LLitI _ i) =
     return (LLitI monoInt i,
                   idSubst,
                   monoInt)
 
-algW env (LLitB _ b) = 
+algW env (LLitB _ b) =
     return (LLitB monoBool b,
             idSubst,
             monoBool)
@@ -151,12 +151,12 @@ algW env (LApp _ e1 e2) =
               s4,
               m3)
 
-algW env (LLam _ x e) = 
+algW env (LLam _ x e) =
     do tv <- freshTyVar
        let env' = extendEnv (x, PPoly [] tv) env
        (e1, s1, m1) <- algW env' e
-       return (LLam m1 x e1, 
-               s1, 
+       return (LLam m1 x e1,
+               s1,
                MFun (apply s1 tv) m1)
 
 algW env (LLet _ x e f) =
@@ -194,6 +194,3 @@ algW env (LIf _ e1 e2 e3) =
 
 algW env (LOp _ o e1 e2) =
     error "not implemented"
-
-
-

### 2016-02-10 codegen/HMBU.hs

@@ -10,7 +10,7 @@ import qualified Data.Set as Set
 import qualified Data.Map as Map
 import Control.Monad.State
 
-type Subst = Map.Map TyVar Monotype 
+type Subst = Map.Map TyVar Monotype
 idSubst = Map.empty
 
 compose :: Subst -> Subst -> Subst
@@ -104,7 +104,7 @@ instance Show LExpr where
     show LLitB{typ,mtvs,b} = show b ++ "::" ++ showj mtvs typ
     show LVar{typ,mtvs,ftv,v} = v ++ "::" ++ showj mtvs typ
     show LApp{typ,mtvs,ftv,e1,e2} = show e1 ++ " (" ++ show e2 ++ ")" ++ "::" ++ showj mtvs typ
-    show LLam{typ,mtvs,ftv,v,e1} = "\\" ++ v ++ " ->\n" ++ 
+    show LLam{typ,mtvs,ftv,v,e1} = "\\" ++ v ++ " ->\n" ++
                                indent 2 (show e1) ++ "::" ++ showj mtvs typ
     show LLet{typ,mtvs,v,e1,e2} = "let " ++ v ++ " = \n" ++
                                   indent 2 (show e1) ++ "\n" ++
@@ -186,7 +186,7 @@ doit e0 =
 -- need a fresh variable supply
 
 freshTyVar :: State Int TyVar
-freshTyVar = 
+freshTyVar =
     do i <- get
        put $ i+1
        return $ 't':show i
@@ -197,7 +197,7 @@ freshMonoVar =
        return $ MVar v
 
 freshMonoVars 0 = return []
-freshMonoVars n = 
+freshMonoVars n =
     do
       v <- freshMonoVar
       vs <- freshMonoVars (n-1)
@@ -209,7 +209,7 @@ instantiate (PPoly as m) =
        let s = Map.fromList $ zip as ms
        return $ apply s m
 
-generalize ms t = 
+generalize ms t =
     let as = Set.toList $ Set.difference (freevars t) (freevars ms)
     in PPoly as t
 
@@ -227,7 +227,7 @@ data Constraint = EqC  Monotype Monotype
 instance Show Constraint where
     show (EqC m1 m2) = "EqC " ++ show m1 ++ " " ++ show m2
     show (ExpC m p) = "ExpC " ++ show m ++ " " ++ show p
[…]

### 2016-02-10 codegen/HMStg.hs

@@ -64,7 +64,7 @@ hmstgdebug :: [Obj InfoTab] -> IO ()
 hmstgdebug os = hmstgAssumsdebug os Set.empty
 
 hmstgAssumsdebug :: [Obj InfoTab] -> Assumptions -> IO ()
-hmstgAssumsdebug os0 assums = 
+hmstgAssumsdebug os0 assums =
     let (os1,i1) = runState (dv os0) 0
         (as,cs,os2) = buNest Set.empty os1 assums :: (Set.Set Assumption,
                                                       Set.Set Constraint,
@@ -76,12 +76,12 @@ hmstgAssumsdebug os0 assums =
        putStrLn $ showObjs (co Set.empty $ backSub subst os2)
 ---}
 {-
-hmstgdebug os0 = 
+hmstgdebug os0 =
     let (os1,i1) = runState (dv os0) 0 :: ([Obj InfoTab], Int)
     in putStrLn $ showObjs (co Set.empty os1)
 -}
 {-
-hmstgdebug os0 = 
+hmstgdebug os0 =
     let (os1,i1) = runState (dv os0) 0
         (as,cs,os2) = buNest Set.empty os1 :: (Set.Set Assumption,
                                                Set.Set Constraint,
@@ -175,15 +175,15 @@ instance DV (Expr InfoTab) (Expr InfoTab) where
                   LitC c -> case maybeCMap emd of
                               Nothing -> error "dv LitC maybeCMap is Nothing"
                               Just cmap -> return $
-                                           let (TyCon boxed tname _ _) = 
-                                                  luTCon c cmap 
+                                           let (TyCon boxed tname _ _) =
+                                                  luTCon c cmap
                                            in case boxed of
                                                 True -> error $ "dv:  boxed constructor " ++
                                                                 c ++ " cannot be literal"
                                                 False -> MCon False tname []
            return $ setTyp m e
 
-    dv e@EFCall{eas} = 
+    dv e@EFCall{eas} =
         do retMono <- freshMonoVar -- return type
            eas' <- mapM dv eas -- setTyp(s) of args
[…]

### 2016-02-10 codegen/HeapObj.hs

@@ -1,4 +1,5 @@
 {-# LANGUAGE CPP #-}
+
 #include "options.h"
 
 module HeapObj (

### 2016-02-10 codegen/InfoTab.hs

@@ -676,8 +676,8 @@ showIT it@(ITAlts{}) =
     "  .cLayoutInfo.unboxedCount = " ++ show (ufvc it) ++ ",\n" ++
     -- initially indicate scrutinee as unboxed for GC, could also indicate
     -- correctly and make NULL in CodeGen
-    "  .cLayoutInfo.bm           = " ++ 
-          npStrToBMStr ( 'N' : 
+    "  .cLayoutInfo.bm           = " ++
+          npStrToBMStr ( 'N' :
                         replicate (bfvc it) 'P' ++
                         replicate (ufvc it) 'N') ++ ",\n" ++
     "};\n"

### 2016-02-10 codegen/OrderFVsArgs.hs

@@ -20,16 +20,16 @@ orderFVsArgs = typeFVs Map.empty
 -- replace bogus type with correct type, sort by boxedness, Var flavor
 repv m vts = boxedFirst [(k, f k) | (k,_) <- vts]
     where f k = case Map.lookup k m of
-                  Nothing -> error $ "HMStg.repv:  unbound variable " ++ show k ++ 
+                  Nothing -> error $ "HMStg.repv:  unbound variable " ++ show k ++
                                      " " ++ show m
                   Just m -> m
 
 -- replace bogus type with correct type, sort by boxedness, Atom flavor
 repa m ats = boxedFirst [(k, f k) | (k,_) <- ats]
     where f a = case a of
-                  Var v -> 
+                  Var v ->
                       case Map.lookup v m of
-                        Nothing -> error $ "HMStg.repa:  unbound variable " ++ 
+                        Nothing -> error $ "HMStg.repa:  unbound variable " ++
                                            show v ++ " " ++ show m
                         Just m -> m
                   LitI{} -> biIntMCon
@@ -68,8 +68,8 @@ instance TypeFVs (Obj InfoTab) where
             m' = foldr (uncurry Map.insert) m (zip vs vstyps)
         in o{e = typeFVsCommon m' e}
 
-    typeFVs m o@PAP{omd,as} = 
-        let (bc, uc, args', perm) = repa m $ zzip (map fst $ args omd) 
+    typeFVs m o@PAP{omd,as} =
+        let (bc, uc, args', perm) = repa m $ zzip (map fst $ args omd)
                                                   (map (typ . emd) as)
         in o{omd = omd{args = args',
                        bargc = bc,
@@ -77,8 +77,8 @@ instance TypeFVs (Obj InfoTab) where
                        argPerm = perm},
              as = map (typeFVsCommon m) as}
 
-    typeFVs m o@CON{omd,as} = 
-        let (bc, uc, args', perm) = repa m $ zzip (map fst $ args omd) 
+    typeFVs m o@CON{omd,as} =
+        let (bc, uc, args', perm) = repa m $ zzip (map fst $ args omd)
                                                   (map (typ . emd) as)
         in o{omd = omd{args = args',
                        bargc = bc,
[…]

### 2016-02-10 codegen/PPrint.hs

@@ -2,6 +2,7 @@
 {-# LANGUAGE TypeSynonymInstances #-}
 {-# LANGUAGE NamedFieldPuns #-}
 {-# LANGUAGE CPP #-}
+
 #if __GLASGOW_HASKELL__ < 710
 {-# LANGUAGE OverlappingInstances #-}
 #endif

### 2016-02-10 codegen/Parser.hs

@@ -139,7 +139,7 @@ groupParsed =
 
     
 
--- uncurried cons is used on several occasions to combine the results of 
+-- uncurried cons is used on several occasions to combine the results of
 -- ordered parsers
 cons = uncurry (:)
       
@@ -231,7 +231,7 @@ tokcutP msg p inp = cutP (show $
 ---------------------------- Object Parsing -------------------------------
 
 -- Parse an Object definition (no semicolon) as Obj ()
-objDefP :: Parser Token (Obj ())              
+objDefP :: Parser Token (Obj ())
 objDefP =
   varNameP >>> \name ->
   eqP >>> \_ ->
@@ -366,10 +366,10 @@ eCaseP =
 
 
 -- parse an Alts section in a case expression, accept an Alts object
-altsP :: Parser Token (Alts ())               
+altsP :: Parser Token (Alts ())
 altsP =
   let
-    name = "alts" --error "this alts not given a name!" 
+    name = "alts" --error "this alts not given a name!"
   in
    tokcutP "Expected one or more alts separated by semicolons" $
    sepByP' altP semiP >>> \alts ->

### 2016-02-10 codegen/ParserComb.hs

@@ -184,7 +184,7 @@ inSeq f sd = foldr aux sd
 
 -- version of inSeq that appends the results of Parsers in the list
 inSeq' :: [Parser i v] -> Parser i [v]
-inSeq' = inSeq cons (accept [])              
+inSeq' = inSeq cons (accept [])
 
 
 -- versions of ordP that ignore part of the input
@@ -247,7 +247,7 @@ p >>> f = \i -> case p i of
 cutP :: Show i => String -> Parser i v -> Parser i v
 cutP m p inp = case p inp of
   [] -> error m
-  x  -> x                 
+  x  -> x
 
 
 -- try to parse with given parser, but don't consume input
@@ -263,4 +263,3 @@ isNextP :: Parser i v -> Parser i Bool
 isNextP p inp =
   let b = not $ null $ p inp
   in accept b inp
-

### 2016-02-10 codegen/Rename.hs

@@ -14,7 +14,7 @@ import Data.Char (isDigit)
 
 withsuff :: String -> [String] -> (String, String)
 withsuff _ [] = error "this should never happen"
-withsuff s (x:xs) = 
+withsuff s (x:xs) =
     let (h,t) = splitAt (length s) x in
     if h == s && t /= [] && head t == '_' && all isDigit (tail t) then
         (h, tail t)
@@ -34,13 +34,13 @@ suffixname v = State $ \used -> let nv = nextv v used in (nv, nv:used)
 suffixnames :: [String] -> State [String] [String]
 suffixnames = mapM suffixname
 
--- uniquify obj names, systematically renaming 
+-- uniquify obj names, systematically renaming
 
 -- no need to flail in the monad at Atom level
 nameVar v tt = condrepl v tt
 
 nameAtom :: Atom -> [(Var, Var)] -> Atom
-nameAtom (Var v) tt = Var $ nameVar v tt 
+nameAtom (Var v) tt = Var $ nameVar v tt
 nameAtom x _ = x
 
 nameAtoms :: [Atom] -> [(Var, Var)] -> [Atom]
@@ -91,7 +91,7 @@ nameObj (CON md c as name) tt =
     do as' <- mapM ((flip nameExpr) tt) as
        return (CON md c as' name)
 
-nameObj (BLACKHOLE md name) tt = 
+nameObj (BLACKHOLE md name) tt =
     return (BLACKHOLE md name)
 
 nameExpr :: Expr a -> [(Var, String)] -> State [String] (Expr a)
@@ -115,7 +115,7 @@ nameExpr e@EAtom{ea} tt =
 nameExpr e@EFCall{ev, eas} tt =
    do
      let ev' = nameVar ev tt
-     eas' <- mapM (flip nameExpr tt) eas 
+     eas' <- mapM (flip nameExpr tt) eas
      return e{ev = ev', eas = eas'}
 
 nameExpr e@EPrimop{eas} tt =
@@ -125,17 +125,17 @@ nameExpr e@EPrimop{eas} tt =
 
 nameAlts :: Alts a -> [(Var, String)] -> State [String] (Alts a)
 nameAlts (Alts md alts name) tt =
-    do 
+    do
       name' <- suffixname name
       alts' <- mapM ((flip nameAlt) tt) alts
       return (Alts md alts' name')
 
-nameAlt (ACon md c vs e) tt = 
+nameAlt (ACon md c vs e) tt =
     do
[…]

### 2016-02-10 codegen/SCC.hs

@@ -14,7 +14,7 @@ module SCC (
 
 import AST
 import qualified Data.Map as Map
-import qualified Data.Set as Set 
+import qualified Data.Set as Set
 
 newtype Graph v = Graph (Map.Map v [v])
 
@@ -48,7 +48,7 @@ transpose g@(Graph m)
 
 --  Depth first search and forests
 
-data Tree v   = Node v (Forest v) 
+data Tree v   = Node v (Forest v)
 type Forest v = [Tree v]
 
 dff :: Ord v => Graph v -> Forest v
@@ -84,7 +84,7 @@ preorderF fs = concatMap preorderT fs
 preorderT (Node v fs) = v:preorderF fs
 
 postorder :: Ord v => Graph v -> [v]
-postorder g = postorderF (dff g) 
+postorder g = postorderF (dff g)
 
 postorderT t = postorderF [t]
 
@@ -121,14 +121,14 @@ instance Show v => Show (Graph v) where
   showsPrec d (Graph m) = shows m
   
 instance Show v => Show (Tree v) where
-  showsPrec d (Node v []) = shows v 
+  showsPrec d (Node v []) = shows v
   showsPrec d (Node v fs) = shows v . showList fs
 
 
 --  Quick Test
 
 tgraph0 :: Graph Int
-tgraph0 = graph 
+tgraph0 = graph
           [(0,[1])
           ,(1,[2,1,3])
           ,(2,[1])
@@ -136,7 +136,7 @@ tgraph0 = graph
           ]
 
 tgraph1 = graph
-          [  ('a',"jg") 
+          [  ('a',"jg")
           ,  ('b',"ia")
           ,  ('c',"he")
           ,  ('d',"")
@@ -187,7 +187,7 @@ instance SCCDepTrans (Obj [Var]) where
 
 instance SCCDepTrans (Expr [Var]) where
     dt e@ECase{ee, ealts} = e{ee = dt ee, ealts = dt ealts}
-    dt ELet{edefs, ee} = 
+    dt ELet{edefs, ee} =
         let ee' = dt ee
             edefs' = dt edefs
             allfvsl = map getFvsObj edefs'
@@ -198,19 +198,19 @@ instance SCCDepTrans (Expr [Var]) where
             -- translate list of lists of onames to list of lists of Objs
             nameobjmap = Map.fromList $ zip namel edefs'
             sccsobj = (map . map) (findItOrElse nameobjmap) sccs
-            fl = map (\edefs -> 
[…]

### 2016-02-10 codegen/STGbits.hs

@@ -19,7 +19,7 @@ abet = ['0'..'9'] ++ ['A'..'F'] -- no we are not exceeding base 16
 
 indexOf :: (Eq a, Num a1, Show a) => a -> [a] -> a1
 indexOf k xs = f xs 0
-    where f []      _             = error $ "indexOf " ++ 
+    where f []      _             = error $ "indexOf " ++
                                             show k ++ " not found"
           f (x:_)  !n | k == x    = n
           f (_:xs) !n | otherwise = f xs (n+1)
@@ -37,17 +37,17 @@ npStrToBMStr s = _npStrToBMStr s True
 
 npStrToBMInt s = read (_npStrToBMStr s False) :: Int
 
-_npStrToBMStr s f = 
+_npStrToBMStr s f =
     let offset = length $ takeWhile (=='-') s
         bitstr = dropWhile (=='-') s
         len = length bitstr
         nptobin c = case c of 'N' -> '0'
-                              'P' -> '1' 
+                              'P' -> '1'
                               _ -> error "not N or P"
         bin = map nptobin bitstr
         w1 = shiftL (strBaseNToWord64 bin 2) offset
         w2 = shiftL (fromIntegral len :: Word64) 58
-    in if f then (word64ToCULHex (w1 .|. w2)) 
+    in if f then (word64ToCULHex (w1 .|. w2))
             else (word64ToHex (w1 .|. w2))
 
 -- unneeded
@@ -58,7 +58,7 @@ word64ToBaseNStr n b = reverse $ f n
     where
       f :: Word64 -> String
       f 0 = ""
-      f n = (abet !! fromIntegral (n `mod` (fromIntegral b))) : 
+      f n = (abet !! fromIntegral (n `mod` (fromIntegral b))) :
              f (n `div` (fromIntegral b))
 
 strBaseNToM :: String -> Int -> Int -> String
@@ -71,5 +71,4 @@ word64ToCULHex w64 =
 
 word64ToHex w64 =
     let hex = word64ToBaseNStr w64 16
-    in "0x" ++ take (16 - length hex) (repeat '0') ++ hex 
-
+    in "0x" ++ take (16 - length hex) (repeat '0') ++ hex

### 2016-02-10 codegen/SetFVs.hs

@@ -25,23 +25,23 @@ dropkey k = filter (\(k',_)->k==k')
 -- transitive closure
 
 tc :: (Show a, Ord a) => Set.Set a
-                         -> [(a, Set.Set a)] 
+                         -> [(a, Set.Set a)]
                            -> Set.Set a
 
 tc lhs rhs = tc' lhs lhs rhs
-    where 
-      tc' acc newroots rhs = 
-          let fvs = Set.unions [ s | (v, s) <- rhs, Set.member v newroots ] 
+    where
+      tc' acc newroots rhs =
+          let fvs = Set.unions [ s | (v, s) <- rhs, Set.member v newroots ]
                     `Set.difference` acc
           in case Set.null fvs of
                True -> Set.union acc newroots -- done
                _ -> let bvs = Set.fromList $ map fst rhs
-                        undefs = Set.difference fvs bvs 
+                        undefs = Set.difference fvs bvs
                     in case Set.null undefs of
                          False -> error $ "SetFVs.tc:  free vars " ++ show undefs
                          _ -> let (rhsl, rhsr) = List.partition (\(v,_) -> Set.member v fvs) rhs
                               in tc' (Set.unions [acc, newroots, fvs])
-                                     (Set.unions $ map snd rhsl) 
+                                     (Set.unions $ map snd rhsl)
                                      rhsr
 
 a = tcList [1] [(1,[1,2]), (1, [3]), (2, []), (3,[1]), (4,[])]
@@ -106,7 +106,7 @@ instance {-# OVERLAPPING #-} Show ([Var],[Var]) where
     show (a,b) = "(" ++ showFVs a ++ "," ++ showFVs b ++ ")"
 
 toplevel :: (Set.Set Var) -> [Obj a] -> [Obj (Set.Set Var, Set.Set Var)]
-toplevel rtgs defs = 
+toplevel rtgs defs =
     -- tlds shadow runtime globals
     let tlds = Set.fromList $ map oname defs
         defs' = setfvs (Set.union tlds rtgs) defs
@@ -124,29 +124,29 @@ toplevel rtgs defs =
 class FVs a where fvsof :: Set.Set Var -> a -> (Set.Set Var, Set.Set Var)
 
[…]

### 2016-02-10 codegen/State.hs

@@ -45,5 +45,3 @@ concatMapM f xs = liftM concat (mapM f xs)
 
 ap :: (Monad m) => m (a -> b) -> m a -> m b
 ap mf m = do { f <- mf; x <- m; return (f x) }
-
-

### 2016-02-10 codegen/Tokenizer.hs

@@ -123,14 +123,14 @@ primopTable =
     ("imul#",    Pimul),
     ("idiv#",    Pidiv),
     ("imod#",    Pimod),
-    ("imax#",    Pimax), 
+    ("imax#",    Pimax),
     ("ieq#",     Pieq),
     ("ine#",     Pine),
     ("ilt#",     Pilt),
     ("ile#",     Pile),
     ("igt#",     Pigt),
     ("ige#",     Pige),
-    ("ineg#",    Pineg),           
+    ("ineg#",    Pineg),
     ("imin#",    Pimin),
     ("imax#",    Pimax)
   ]
@@ -195,7 +195,7 @@ tokenize ls = aux (None (0,0)) (0,0) (stripComments ls) []
     -- special case for '.':
          -- must only appear in a NumTok once, not at the start or end
          -- cannot start a token (yet)
-         -- currently 
+         -- currently
       | x == '.' =
         case st of
         NumTok {tS} | elem '.' tS
@@ -218,7 +218,7 @@ tokenize ls = aux (None (0,0)) (0,0) (stripComments ls) []
          None {} -> aux (NumTok [x] (l,c)) (l,c+1) xs toks
 
          -- accept digits mid-token, whether numeric or stringy
-         _       -> aux st{tS = x:(tS st)} (l,c+1) xs toks 
+         _       -> aux st{tS = x:(tS st)} (l,c+1) xs toks
 
     -- only valid as a non-beginning part of an identifier or constructor
     -- if found outside of such, should it be read as a comment in the comment stripper?

### 2016-02-10 codegen/Util.hs

@@ -38,12 +38,12 @@ indent i xs = space ++ indent' i xs
       indent' i (x:xs)      = x : indent' i xs
       indent' i ""          = ""
 
-lookupOrElse :: (Ord k) => k -> Map.Map k v -> v      
+lookupOrElse :: (Ord k) => k -> Map.Map k v -> v
 lookupOrElse k map = case Map.lookup k map of
                       Just k -> k
                       Nothing -> error "lookupOrElse failed!"
       
-maxPayload :: Int     
+maxPayload :: Int
 maxPayload = 32
 
 -- sanity-checking zip
@@ -56,7 +56,7 @@ precalate s ss = concatMap (s++) ss
 
 -- given a predicate and a list
 --   stably permute list according to the predicate, satisfies first
--- return 
+-- return
 --   new - permuted list
 --   permutation lists such that
 --     old[i] = new[perm[i]]
@@ -82,7 +82,7 @@ groupAllBy f xs =
   let nubbed = nubBy f xs
   in if (length nubbed == length xs)
 -- if list is full of uniques, package them in lists
-     then map (:[]) xs 
+     then map (:[]) xs
 -- else fold on the uniques, accumulating the groups (hence reverse) and the unsorted
      else reverse $ fst $ foldl ffun ([],xs) nubbed
   where ffun (groups, xs) it = let (matches, rest) = partition (f it) xs

### 2016-02-10 codegen/stgc.hs

@@ -164,8 +164,8 @@ compile  (Options {optVerbose, optDumpParse, optNoPrelude, optInput,
                  cflagsenv <- lookupEnv "CFLAGS"
                  let cflags = fromMaybe "" cflagsenv
                  let coutput = input ++ ".c"
-                 let flags = " -Wno-return-type " 
-                               ++ cflags ++ " -std=gnu99 -Wl,-rpath " ++ rtIncDir 
+                 let flags = " -Wno-return-type "
+                               ++ cflags ++ " -std=gnu99 -Wl,-rpath " ++ rtIncDir
                                ++ " -L" ++ rtLibDir ++ " -I" ++ rtIncDir
                                ++ " -lruntime"
                  writeFile coutput (pprinter $ codegener source optVerbose minihs)

### 2016-02-14 pubs/tfp16/paper.tex
more words

@@
 %
-\title{Strict Pure Functional Programming and Automatic Parallelization}
+\title{Strict Pure Functional Programming and Automatic Parallelization\\
+       (Project-in-progess Paper)}
 %
@@
 %
-\author{Kei Davis\inst{1} \and Dean Prichard\inst{1} \and David Ringo\inst{1,2}}
+\author{Kei Davis\inst{1} \and 
+        Dean Prichard\inst{1} \and 
+        David Ringo\inst{1,2} \and
+        Loren Anderson\inst{1,3} \and
+        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
+}
 %
@@
 University of New Mexico,
-Albuquerque, NM, USA}
+Albuquerque, NM, USA
+\and
+North Dakota State University,
+Fargo, North Dakota, USA
+\and 
+New Mexico Institute of Mining and Technology,
+Socorro, New Mexico, USA
+}
 
@@
 
+% moronic author listing code increments footnote counter
+\setcounter{footnote}{0}
+
 \begin{abstract}
 The trend is functional programming in scientific computing, and in particular
-strict (by default) functional programming, in various guises.  The project is
-the demonstration of a light-weight, higher-order, polymorphic,
-strict-by-default purely functional language implementation in which we can
-experiment with automatic parallelization strategies and varying degrees of
-default strictess.
+strict (by default) functional programming, in various guises.  Our project is
+the demonstration of a light-weight, higher-order, polymorphic pure functional
+language implementation in which we can experiment with automatic
+parallelization strategies and varying degrees of default strictness.
+Strictness prescribes the degree of potential parallelism, that is, we are not
+exploring speculative evaluation.
+
+\emph{Preliminary paper note: We are reporting work in progress.  At the time
+  of writing we have a complete serial implementation and are currently
+  developing the parallel runtime.  We expect to have preliminary results from
+  our parallel implementation by the time of the symposium.}
 
@@
 \section{Background and Motivation}
+The practice of highest-performance scientific computing (HPC) is, overall,
+highly conservative with respect to change.  Even as the US government commits
+to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
+highly respected member of the HPC community recently stated ``Fortran is
+essential for exascale programming.''~\cite{Heroux16}.  Existing HPC codes run
+to hundreds of thousands of lines and complete rewrites are infeasible.  C++
+has become more common for new code starts, but their nature remains the same:
+huge codes with poor (or non-existent) encapsulation of side effects, and
+multiple levels of parallelism also poorly abstracted, especially at the
+thread (shared-memory) level.  Thread-level parallel efficiency can be poor
+and is explained by Amdahl's law because the parallel model is bulk synchronous.
 
-\section{Goals}
+\subsection{The trend: strict but still pure functional programming}
+There is a small but growing understanding that \emph{pure functional
+  semantics}, in some form, is essential to reining in the complexity of these
+ever-evolving scientific codes.  Such semantics are expressed in a number
+of ways as highlighted following.
+
+\subsubsection{Rely on programmer discipline.}
+Legacy Fortran scientific codes likely represent worst practices in state
+encapsulation, with most data global and accessible to all.  It is almost
+ironic, then, that of the mainstream high-performance scientific programming
+languages (Fortran, C, and C++), Fortran 95 was the first to introduce the
+\texttt{pure} function qualifier, and with enforcement by the compiler.
+Otherwise it is entirely up to the programmer to make functions composable
+and thread safe.  Here at LANL this approach is being used to good effect
+to clean up some legacy code bases.
+
+\subsubsection{Obey constraints prescribed by a parallel runtime system.}
+There are a number of parallel runtime systems that encourage (but cannot
+enforce) a pure functional style.  We will highlight Stanford's Legion
+runtime as an example~\cite{Bauer12}.
+
+In Legion, non-function-local data, or \emph{regions}, are requested from the
+runtime system.  Regions may be shared among \emph{tasks} (essentially C++
+functions), and each task registers its access (read, write, read/write, etc.)
+and coherency requirements for each region to which it will have access.
+Other than registered access to regions tasks must not access non-constant
+global data.  In serial execution these access requirements are notionally
+superfluous, but in parallel execution the runtime can dynamically calculate
+the dependency graph and relax the programmatically defined strict ordering of
+task execution.  Very high parallel efficiency has been demonstrated on extremely
+large computing systems~\cite{Bauer14}.  Here at LANL the Legion system is
+being strongly considered for a new large-scale C++ code effort.\footnote{
+It is interesting to note that the PI of the Legion project was previously a
+co-developer of a strict, arguably pure functional language
+implementation~\cite{AikenFL}.}
+
+\subsubsection{Use a pure functional language.}
+Some twenty years after the US Department of Energy laboratory complex was
+presented with the most performant and highly parallel implementation of a
+pure functional language to date~\cite{Davis96}, and collectively had no
+conceptual understanding of what it was, much less why it might be useful or
+important, there is a small number of laboratory HPC practioners who are
+developing an appreciation for pure functional programming in Haskell.  Still,
+they see their needs as different than other Haskell users: performance is
+paramount, unboxed arrays are a (perhaps \emph{the}) primary data structure,
+and laziness is more of a nuisance--even a serious hindrance--than a
+convenience, much less algorithmically essential in the sense of Bird's
+\emph{repmin}~\cite{Bird84} or Johnsson's general attribute
+grammars~\cite{Johnsson87}.
+
+A common complaint among these budding high-performance functional programmers
+is the need to get strictness just right: to remove space leaks,
+improve performance, get communication (e.g.\ MPI) and IO to work as they
+expect, and so on.  More specifically, they expect to be able to reason about
+space and time usage using their existing mental models. While they are
+happy to write code that is strongly reminiscent of the numerical examples in
+Hughes' classic piece (whether they've read it or not)~\cite{Hughes89},
+they would also be happy to write in a slightly modified style that does
+not require non-strictness---after all, that's how they've been thinking their
+entire careers.
+
+Much anecdotal evidence suggests that many others have similar difficulties
+with lazy (or non-strict) by default semantics, with the most telling recent
+concrete evidence being the push for, and realization of, the \texttt{Strict}
+and \texttt{StrictData} language extensions newly available in GHC 8.0.
+Quoting the documentation:
+\begin{quote}
+High-performance Haskell code (e.g. numeric code) can sometimes be littered
+with bang patterns, making it harder to read. The reason is that laziness
+isn't the right default in this particular code, but the programmer has no way
+to say that except by repeatedly adding bang patterns~\cite{strict-strictdata}.
+\end{quote}
+
+We claim, then, that there is a trend towards \emph{strict-by-default pure
+  functional programming}, whether as imposed by programmer discipline, a parallel
+runtime system (such as the aforementioned Legion, wherein the pure semantics
+at function granularity allows the runtime system to relax the nominal serial
+ordering), or the language itself (e.g.\ strict Haskell).
+
+Intel's TBB?  Cilk++?
+
+\subsection{Strictness and Parallelism}
+
+It is a trivial observation that given strict functional semantics, wherein
+function arguments may be safely evaluated before function evaluation, it is
+safe to evaluate the arguments, and the function, in parallel, and that this
+principle applies recursively in the (dynamic) expression tree (or graph).
+
+\section{Project and Goals}
+
+Our project is the light-weight implementation of a pure, higher-order,
+polymorphic, functional language and runtime system with which we can
+experiment with automatic parallelization strategies with degrees of language
+strictness ranging from lazy to fully strict.  Light-weight has several
+meanings or implications.  First and foremost it means that \emph{we are not
+  attempting to compete with ghc in any respect}.\footnote{We estimate that we
+  will have, to within an order of magnitude, one person-hour of available
+  development time for each person-year that ghc currently represents, or put
+  another way, one second for each hour, respectively.}
+
+\begin{itemize}
+\item The implementation must be feasible in terms of effort;
+\item The language must be strongly typed, implicitly or explicitly.  More
+  specifically, the language should explicitly support both Hindley-Milner
+  polymorphism and unboxed types~\cite{firstclass};
+\item Straighforward direct linkage with C, preferably without a sophisticated
+  foreign function interface, is essential;
+\item The language should be ``currying-friendly'' as described in the ``fast
+  curry'' paper~\cite{fastcurry};
+\item Choice of degree of strictness is easily selected;
+\item Sharing, in the sense of what ``lazy'' means vs.\ merely non-strict,
+  should be preserved for all degrees of strictness;
+\item Tail calling should be properly implemented, i.e., not as a trampoline;
+\item etc.
+\end{itemize}
+
+For feasibility, we restrict ourselves to shared-memory implicit parallelism,
+but distributed-memory parallelism should be straightforwardly realizable
+using e.g.\ MPI or POSIX sockets.
+
+\subsection{Varying degrees of strictness}
+
+First some words about terminology.  For all degrees of strictness sharing (as
+implied by lazy but not by non-strict) of heap-allocated objects is preserved
+we will not use the term \emph{lazy} and simply refer to strictness
+properties.  As is common we will use strict in an operational sense, meaning
+that argument(s) are evaluated before being passed in a function call.
+Similarly, when referring to language semantics we refer to the operational
+rather than denotational semantics.
+
+Greater strictness implies more opportunities for parallelism, possible losses
+or gains in space efficiency, and less expressiveness.  We seek to explore
+the interplay of all of these.
+
+ghc 8.0 gives three possibilities: conventional non-strict, constructor
+strict, and function- and constructor strict, but there are possible
+variations.  The obvious variation is function, but not constructor, strict.
+In Haskell the distinction between constructors and functions is somewhat
+blurred because constructors behave like functions (outside of pattern
+matching), but this need not be the case.  In STG---ghc's intermediate
+language---the distinction is clear: heap objects are explicitly constructed
+by \texttt{let} constructs in which constructor names are explicitly mentioned
+(not aliased); constructor names are not functions.
+
+Other possible variations concern the notional arity of the underlying
+function being applied.  This is readily explained in terms of the operation of the
+STG machine described next.
+
+In STG the arity of a function is syntactically defined as the number of
+manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
+regardless of what type it might be dynamically ``instantiated'' at.  We distinguish
+\emph{application}, e.g.\ \texttt{f x y}, which is a syntactic notion, from
+actually calling a function. Thus \texttt{id id x} is an application of \texttt{id}
+to two arguments, but the underlying \emph{FUN} (user- or system- defined function)
+when actually called is given exactly one argument.
+
+In the eval-apply model, given application \texttt{f x y}, in the standard
+(non-strict) semantics \texttt{f} is first evaluated and the arity of the
+underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
+(partial application of a \emph{FUN}) is determined---in the case of
+\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
+arguments already provided.  If the arity is greater than two then a new
+\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
+the new \emph{PAP} returned.  If the arity is equal to two then the underlying
+\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
+and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
+(it must be at least one, so one in this case) the underlying \emph{FUN} is
+called-with-return with any previous argument (in case of \emph{PAP}) and new
+argument \texttt{x}, then the object returned applied to remaining argument
+\texttt{y}.
+
+Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
+evaluated?  In the non-strict semantics evaluation is on demand, after
+\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
+(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one of
+those arguments forced in subsequent evaluation (assuming that they weren't
+forced in the evaluation of \texttt{f} itself).
+
+In a strict semantics one can come up with various evaluation orders, but
+excluding the absurd, as far as distinguishing termination properties goes
+these can be distilled down to two possibilities, namely whether on
+construction of a \emph{PAP} the (new) arguments are forced or not.  (An
+absurd strategy might force every second argument.)  These are in fact
+distinguishable because (in Haskell) the first argument to \texttt{seq} might
+evaluate to a \emph{PAP}.
+
+This then gives three reasonable operational modes that are strictly ordered
+in terms of termination properties: non-strict, fully strict---all arguments
+in an application are evaluated---and the intermediate in which only those
+arguments needed by a function call are evaluated.
+
+-- spell out STG syntax?
 
@@
 
-\section{Current Status and Expected Results (by the time of the symposium)}
+Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
+is part of our master plan.  In short, our plan is to use ghc as our front-end
+to generate its intermediate form, STG, then escape to our system from there.
+There are various possible ways to accomplish the former: using ghc's option
+to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
+and, optionally, using ghc as a library.  We note that this approach has 
+a precedent in the aforementioned Nvidia research compiler.
+
+\subsection{STG to C}
+
+The back-end is written in Haskell and generates proper tail-calling C code;
+the runtime system is written in C/C++.  The system fairly faithfully
+implements the STG machine as described in Peyton-Jones and Marlow's ``fast
+curry'' paper~\cite{fastcurry} and dynamically illustrated by Pope's
+miniSTG~\cite{ministg} with the exception of the state transition rules which,
+of course, are different than those for lazy semantics.  Sharing is preserved.
+
+\subsection{Strictness}
+
+The STG machinery makes the choice of evaluation strategy easy to change, both
+in terms of function and constructor application.
+
+\section{Current Status, and Expected Results by the time of the symposium}
+
+No bridge from ghc STG to our STG yet.
+
+At the time of writing we have a complete serial implementation:  STG parser, type inferencer, code
+generator, and runtime system including a garbage collector.
+
+C stack, our stack.
+
+Mutual exclusion for garbage collector--heap check with every heap allocation.  In fact garbage collector
+can be invoked at any point in time except when populating a heap object immediately after its allocation.
+
+\subsection{Mini-Haskell}
+
+\subsection{Type system}
+
+We perform standard Hindley-Milner type inference and enforcement, augmented
+with built-in and user-defined unboxed types as described by Peyton-Jones and
+Launchbury's seminal work~\cite{unboxed}, but for unboxed types code
+generation is currently limited to built-in single-word (64-bit) or smaller
+types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
+and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
+  True\#}.
+
+\subsection{Test suite and testing}
+
+Extensive prelude.
+
+Test suite of several hundred small programs.
+
+Students learning Haskell and STG, writing Mini-Haskell and STG programs that
+serendipitously go into our automated testing suite which is partitioned
+according to our lattice of evaluation strategies (3 degrees of function
+strictness by two degrees of constructor strictness), which tells us how much
+non-strictness each program needs and something about the style in which
+programs are written.
+
+Continuous build system (Jenkins)
+
+\subsection{Parallel evaluation}
+
+Parallel runtime is being implemented following the ideas of~\cite{SPJs-many-papers}, thread pool,
+stack and heap for each thread, blah, blah.
+
+\section{Other Related Work}
+Nvidia research compiler.
+
+\section{Summary, Future Work, Conclusions?}
 
@@
 
+\bibitem{Heroux}
+January 14, 2016
+Exascale Programming: Adapting What We Have Can (and Must) Work
+Michael A. Heroux, Sandia National Laboratories
+http://www.hpcwire.com/2016/01/14/24151/
+
+\bibitem{Davis96}
+MPP Parallel Haskell. 
+K. Davis. 
+Draft Proceedings Implementation of Functional Languages 1996 (IFL '96) LNCS
+1268, pp49-54, Bonn/Bad-Godesberg, Germany, LNCS 1268, Springer-Verlag.
+
+\bibitem{Bauer12} 
+Legion: Expressing Locality and Independence with Logical
+  Regions. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings
+  of the Conference on Supercomputing, pages 1-11, November 2012.
+
+\bibitem{Bauer14} 
+Structure Slicing: Extending Logical Regions with Fields. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings of the Conference on Supercomputing, pages 845-856, November 2014.
+
+\bibitem{AikenFL}
+A. Aiken, J.H. Williams, and E.L. Wimmers
+The FL Project:  The Design of a Functional Language
+(technical report not published elsewhere), September 1993.
+http://theory.stanford.edu/~aiken/publications/trs/FLProject.pdf
+
+\bibitem{Bird84}
+R. S. Bird, 
+Using circular programs to eliminate multiple traversals of data, 
+Acta Informatica, 21(3), 239-250 (1984).
+
+\bibitem{Johnsson87}
+Thomas Johnsson
+Attribute Grammars as a Functional Programming Paradigm
+Proc. of a conference on Functional programming languages and computer architecture
+Pages 154-173 
+Springer-Verlag London, UK ©1987 
+table of contents ISBN:0-387-18317-5
+
+\bibitem{Hughes89}
+J. Hughes,
+Why Functional Programming Matters,
+Computer Journal,
+32,
+2,
+98--107,
+1989.
+
+\bibitem{strict-strictdata}
+http://ghc.haskell.org/trac/ghc/wiki/StrictPragma
+
+[ghclib] GHC/As a library http://wiki.haskell.org/GHC/As\_a\_library
+
 \end{thebibliography}

### 2016-02-17 pubs/tfp16/paper.tex
synching

@@
 
+Proper tail calling is achieved.  Because we maintain our own data and control
+(continuation) stack all generated C functions take no arguments and have return
+type void, and with a single exception are invoked in tail call position, both
+gcc and Clang/LLVM generate jumps instead of calls.  The exception is a convenience.
+As described in the fast curry paper we have a generic stgApply function that
+is used for applications of non-known functions, and known functions with
+a deficit or excess number of arguments.  The stgApply function takes as arguments
+the object (FUN, PAP, or THUNK) and its arguments, and if a THUNK, evaluates it
+with a call (rather than jump) to yield a FUN or PAP.  This allows us to implement
+a single stgApply function...can we split this?
+
 \subsection{Strictness}
@@
 generator, and runtime system including a garbage collector.
-
-C stack, our stack.
 

### 2016-02-23 pubs/tfp16/paper.tex
syncing on this fine closure day

--- 2016-02-14 pubs/tfp16/paper.tex
@@
 \title{Strict Pure Functional Programming and Automatic Parallelization\\
-       (Project-in-progess Paper)}
+       (New Project-in-progess Paper)}
 %
@@
 
-\emph{Preliminary paper note: We are reporting work in progress.  At the time
-  of writing we have a complete serial implementation and are currently
-  developing the parallel runtime.  We expect to have preliminary results from
-  our parallel implementation by the time of the symposium.}
+\emph{Preliminary paper note 1: We are reporting work in progress.  At the time
+  of writing we have a complete serial implementation, a primitive
+  proof-of-concept parallel implementation, and are currently developing a
+  more realistic parallel runtime.  We expect to have preliminary results from
+  our new parallel implementation by the time of the symposium.}
+
+\emph{Note 2:  While this does not qualify as a student paper, students (undergraduate!)
+have contributed considerably to this effort, and if accepted we plan to have one of these students
+present this paper.}
 
@@
 highly respected member of the HPC community recently stated ``Fortran is
-essential for exascale programming.''~\cite{Heroux16}.  Existing HPC codes run
-to hundreds of thousands of lines and complete rewrites are infeasible.  C++
-has become more common for new code starts, but their nature remains the same:
-huge codes with poor (or non-existent) encapsulation of side effects, and
+essential for exascale programming.''~\cite{Heroux16} Existing US government
+laboratory HPC codes run to hundreds of thousands of lines and complete
+rewrites are infeasible.\footnote{A European consortium is anticipating
+  exascale computing as well~\cite{EUexascale}, but the authors are not
+  knowledgeable about the state of large-scale, non-USA HPC codes.}  C++ has
+become more common for new code starts, but their nature remains the same:
+huge codes with fair to non-existent encapsulation of side effects, and
 multiple levels of parallelism also poorly abstracted, especially at the
 thread (shared-memory) level.  Thread-level parallel efficiency can be poor
-and is explained by Amdahl's law because the parallel model is bulk synchronous.
+and is explained by Amdahl's law because the parallel model is bulk
+synchronous.
 
@@
 
+\subsubsection{Proper tail calls}
+
+Proper tail calling is achieved.  Because we maintain our own data and control
+(continuation) stack all generated C functions take no arguments and have
+return type void, and with a single exception in the non-strict case are
+invoked in tail call positions, both gcc and Clang/LLVM generate jumps instead
+of calls for C calls in tail call positions.
+
+The exceptions to tail calling are for convenience.  Similar to what is
+described in the \emph{fast curry} paper we have a generic variadic stgApply
+function that is used for applications of non-known functions, and known
+functions with a deficit or excess number of arguments.  The stgApply function
+takes as arguments the first element of the application (FUN, PAP, THUNK, or
+BLACKHOLE) and its arguments, and if a THUNK, evaluates it with a call (rather
+than jump) to yield a FUN or PAP.  This allows us to implement a single
+stgApply function as a single C function.  The implementation of stgApply as
+described in \emph{fast curry} makes the same concession.
+
+For stgApply in the non-strict semantics the alternative to call-with-return
+would be a pair of functions stgApply1 and stgApply2.  In one possible variation, 
+stgApply1 would push a continuation with stgApply2 as the return address, then
+invoke the first element of the application.  Call-with-return becomes more
+convenient when considering the various forms of strictness:  when evaluating
+function or constructor arguments it is simpler, and arguably more efficient
+in terms of time (but not space in general) to ``evaluate in place'' than
+implement the continuation style to the logical limit.
+
 \subsection{Strictness}
@@
 
+\section{Acknowledgments}
+
+DOE/NNSA/LANL LDRD, DOE/NNSA SULI.
+
 %
@@
 http://www.hpcwire.com/2016/01/14/24151/
+
+\bititem{EUexascale}
+Toward Exascale Computing.
+European Exascale Software Initiative.
+http://www.eesi-project.eu/
 

### 2016-02-24 pubs/tfp16/paper.tex
paper additions

@@
 
--- spell out STG syntax?
-
 \section{Design and Implementation}
@@
 a precedent in the aforementioned Nvidia research compiler.
+
+\subsection{STG Language}
+\setlength{\tabcolsep}{5pt}
+
+The STG language is ...
+
+Our incarnation of STG is very similar to that given in EA, as is
+our presentation which is meant to evoke both an abstract and concrete syntax.
+
+\begin{table}
+\footnotesize % tiny scriptsize footnotesize small
+\centering
+\begin{tabular}{r r c l l}
+
+Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
+Constructor  & $C$            &     &                                              & Initial upper-case letter \\
+Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point \\
+Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
+\\
+Expression   & $e$            & ::= & $a$                                          & Atom \\
+             &                & $|$ & $f\ a_1\dots a_n$                            & Function application, $n\ge 1$ \\
+             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
+             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$ 
+                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
+             &                & $|$ & \texttt{case} $e$ \texttt{as} 
+                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n$\\
+\\
+Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
+             &                & $|$ & \_                \texttt{->} $e$            & Default \\
+\\
+Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
+             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
+             &                & $|$ &\texttt{THUNK} $e$                            & Thunk \\
+             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
+             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole
+\end{tabular}
+\end{table}
+
+While our interpretation of STG is a higher-order, polymorphic, pure functional
+language, some syntactic restrictions make direct programming rather tedious.
+This is discussed further in a later section.
 
@@
 
-At the time of writing we have a complete serial implementation:  STG parser, type inferencer, code
-generator, and runtime system including a garbage collector.
+At the time of writing we have a complete serial implementation: STG parser,
+type inferencer, code generator, and runtime system including a garbage
+collector.
 
-Mutual exclusion for garbage collector--heap check with every heap allocation.  In fact garbage collector
-can be invoked at any point in time except when populating a heap object immediately after its allocation.
+Mutual exclusion for garbage collector--heap check with every heap allocation.
+In fact garbage collector can be invoked at any point in time except when
+populating a heap object immediately after its allocation.
 
 \subsection{Mini-Haskell}
+
+While our interpretation of STG is a higher-order, polymorphic, pure
+functional language, some syntactic restrictions make direct programming
+rather tedious.  While a near-term goal is to bridge the gap\ldots
+
+First while the scrutinee of a case expression may be an arbitrary expression,
+arguments in function applications, constructor allocation, and primitive
+operations are constrained to be atoms.  For unboxed types a case
+a trivial case expression to alleviates
+the difficulty.
+\begin{verbatim}
+  f ... e ... -> case e as x of _ -> f ... x ...
+\end{verbatim}
 
@@
 
-\bititem{EUexascale}
+\bibitem{EUexascale}
 Toward Exascale Computing.

### 2016-02-24 talks/stg/dump-stg.sh
add stg slides and stuff

@@ -1,0 +1,2 @@
+#!/bin/bash
+ghc -c Fac.hs -ddump-stg

### 2016-02-24 talks/stg/stg-run.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+../../build/bin/stgc -p ./fac.stg
+./a.out

### 2016-02-24 pubs/tfp16/paper.tex
stg syntax slide

@@
 Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
-             &                & $|$ & \_                \texttt{->} $e$            & Default \\
+             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
 \\
@@
              &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
-             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole
+             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
+\\
+Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $n\ge 1$, distinguished $\mathit{main}$ \\
+
 \end{tabular}

### 2016-02-26 pubs/tfp16/paper.tex
sync before branching

@@
 Legacy Fortran scientific codes likely represent worst practices in state
-encapsulation, with most data global and accessible to all.  It is almost
+encapsulation, with most data global and accessible to all; indeed, Fortran
+was designed to make this easy via \texttt{COMMON} blocks---aggregates of
+global variables---which further obfuscate program meaning by allowing
+arbitrary naming of a block's variables at a subroutine/function granularity,
+and requiring that the entire block be brought into scope.  It is almost
 ironic, then, that of the mainstream high-performance scientific programming
-languages (Fortran, C, and C++), Fortran 95 was the first to introduce the
+languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
 \texttt{pure} function qualifier, and with enforcement by the compiler.
-Otherwise it is entirely up to the programmer to make functions composable
-and thread safe.  Here at LANL this approach is being used to good effect
-to clean up some legacy code bases.
+Otherwise it is entirely up to the programmer to make functions composable and
+thread safe.  Here at LANL this approach is being used to good effect to clean
+up some legacy code bases.
 
@@
 runtime system.  Regions may be shared among \emph{tasks} (essentially C++
-functions), and each task registers its access (read, write, read/write, etc.)
-and coherency requirements for each region to which it will have access.
-Other than registered access to regions tasks must not access non-constant
-global data.  In serial execution these access requirements are notionally
-superfluous, but in parallel execution the runtime can dynamically calculate
-the dependency graph and relax the programmatically defined strict ordering of
-task execution.  Very high parallel efficiency has been demonstrated on extremely
-large computing systems~\cite{Bauer14}.  Here at LANL the Legion system is
-being strongly considered for a new large-scale C++ code effort.\footnote{
-It is interesting to note that the PI of the Legion project was previously a
-co-developer of a strict, arguably pure functional language
-implementation~\cite{AikenFL}.}
+functions), and each task is prescribed access privileges (read, write,
+read/write, etc.)  and coherency requirements for each region to which it will
+have access.  Other than registered access to regions tasks must not access
+non-constant global data.  In serial execution these access requirements are
+notionally superfluous, but in parallel execution the runtime can dynamically
+calculate the data dependency graph and relax the programmatically defined
+serial ordering of task execution.  Very high parallel efficiency has been
+demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
+LANL the Legion system is being strongly considered for a new large-scale C++
+code effort.\footnote{It seems significant that the PI of the Legion
+  project was previously a co-developer of a strict, arguably pure functional
+  language implementation~\cite{AikenFL}.}
 
@@
 Some twenty years after the US Department of Energy laboratory complex was
-presented with the most performant and highly parallel implementation of a
-pure functional language to date~\cite{Davis96}, and collectively had no
-conceptual understanding of what it was, much less why it might be useful or
-important, there is a small number of laboratory HPC practioners who are
-developing an appreciation for pure functional programming in Haskell.  Still,
-they see their needs as different than other Haskell users: performance is
-paramount, unboxed arrays are a (perhaps \emph{the}) primary data structure,
-and laziness is more of a nuisance--even a serious hindrance--than a
-convenience, much less algorithmically essential in the sense of Bird's
-\emph{repmin}~\cite{Bird84} or Johnsson's general attribute
+presented with the most highly parallel (and thereby performant)
+implementation of a pure functional language to date~\cite{Davis96}, and
+collectively had no conceptual understanding of what it was, much less why it
+might be useful or important, there is now a small number of laboratory HPC
+practioners who are developing an appreciation for pure functional programming
+in Haskell.  Still, they see their needs as different than other Haskell
+users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
+primary data structure, and laziness is more of a nuisance--even a serious
+hindrance--than a convenience, much less ever algorithmically essential in the
+sense of Bird's \emph{repmin}~\cite{Bird84} or Johnsson's general attribute
 grammars~\cite{Johnsson87}.
@@
 A common complaint among these budding high-performance functional programmers
-is the need to get strictness just right: to remove space leaks,
-improve performance, get communication (e.g.\ MPI) and IO to work as they
+is the need to get strictness just right: to remove space leaks, achieve
+acceptable performance, get communication (e.g.\ MPI) and IO to work as they
 expect, and so on.  More specifically, they expect to be able to reason about
-space and time usage using their existing mental models. While they are
-happy to write code that is strongly reminiscent of the numerical examples in
-Hughes' classic piece (whether they've read it or not)~\cite{Hughes89},
-they would also be happy to write in a slightly modified style that does
-not require non-strictness---after all, that's how they've been thinking their
+space and time usage using their existing mental models. While they are happy
+to write code that is strongly reminiscent of the numerical examples in
+Hughes' classic piece (whether they've read it or not)~\cite{Hughes89}, they
+would also be happy to write in a slightly modified style that does not
+require non-strictness---after all, that's how they've been thinking their
 entire careers.
@@
   foreign function interface, is essential;
-\item The language should be ``currying-friendly'' as described in the ``fast
+\item The language should be \emph{currying-friendly} as described in the ``fast
   curry'' paper~\cite{fastcurry};
 \item Choice of degree of strictness is easily selected;
-\item Sharing, in the sense of what ``lazy'' means vs.\ merely non-strict,
+\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
   should be preserved for all degrees of strictness;
@@
 
-First some words about terminology.  For all degrees of strictness sharing (as
-implied by lazy but not by non-strict) of heap-allocated objects is preserved
-we will not use the term \emph{lazy} and simply refer to strictness
-properties.  As is common we will use strict in an operational sense, meaning
-that argument(s) are evaluated before being passed in a function call.
-Similarly, when referring to language semantics we refer to the operational
-rather than denotational semantics.
+First some words about terminology.  For all degrees of strictness, sharing
+(as implied by lazy but not by non-strict) of heap-allocated objects is
+preserved so we will not use the term \emph{lazy} and simply refer to
+strictness properties.  As is common we will use strict in an operational
+sense, meaning that argument(s) are evaluated before being passed in a
+function call, rather than the denotational sense.  Similarly, when referring
+to language semantics we refer to the operational rather than denotational
+semantics.
 
@@
 strict, and function- and constructor strict, but there are possible
-variations.  The obvious variation is function, but not constructor, strict.
+variations.  The obvious variation is function, but not constructor, strictness.
 In Haskell the distinction between constructors and functions is somewhat
 blurred because constructors behave like functions (outside of pattern
-matching), but this need not be the case.  In STG---ghc's intermediate
-language---the distinction is clear: heap objects are explicitly constructed
-by \texttt{let} constructs in which constructor names are explicitly mentioned
+matching), but this need not be the case.  In STG---one of ghc's intermediate
+representations---the distinction is clear: heap objects are explicitly constructed
+by \texttt{let} constructs in which constructor names are given explicitly
 (not aliased); constructor names are not functions.
 
+\section{Design and Implementation}
+
+Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
+is part of our master plan.  In short, our plan is to use ghc as our front-end
+to generate its intermediate form, STG, then escape to our system from there.
+There are various possible ways to accomplish the former: using ghc's option
+to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
+and, optionally, using ghc as a library.  We note that this approach has 
+a precedent in the aforementioned Nvidia research compiler.
+
+\subsection{STG Language}
+\setlength{\tabcolsep}{5pt}
+
+The STG language is ...
+
+Our incarnation of STG is very similar to that given in EA, as is
+our presentation which is meant to evoke both an abstract and concrete syntax.
+
+\begin{table}
+\footnotesize % tiny scriptsize footnotesize small
+\centering
+\begin{tabular}{r r c l l}
+
+Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
+Constructor  & $C$            &     &                                              & Initial upper-case letter \\
+Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
+Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
+\\
+Expression   & $e$            & ::= & $a$                                          & Atom \\
+             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
+             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
+             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$ 
+                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
+             &                & $|$ & \texttt{case} $e$ \texttt{as} 
+                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n \ge 1$\\
+\\
+Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
+             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
+\\
+Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
+             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
+             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
+             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
+             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
+\\
+Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $1\le i \le n \ge 1$, distinguished $\mathit{main}$ \\
+
+\end{tabular}
+\end{table}
+
+While our interpretation of STG is as a simple higher-order, polymorphic, pure
+functional language, some syntactic restrictions make direct programming
+rather tedious.  This is discussed further in a later section.
+
+\subsection{Degrees of strictness, take two}
+
 Other possible variations concern the notional arity of the underlying
-function being applied.  This is readily explained in terms of the operation of the
-STG machine described next.
+function being applied.  This is readily explained in terms of the operation
+of the STG machine described later.  Our description is based on that given in
+Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
+refer so many times that we subsequenty refer to it as
+\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely has another
+  connotation to the reader.}
 
@@
 in an application are evaluated---and the intermediate in which only those
-arguments needed by a function call are evaluated.
-
-\section{Design and Implementation}
-
-Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
-is part of our master plan.  In short, our plan is to use ghc as our front-end
-to generate its intermediate form, STG, then escape to our system from there.
-There are various possible ways to accomplish the former: using ghc's option
-to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
-and, optionally, using ghc as a library.  We note that this approach has 
-a precedent in the aforementioned Nvidia research compiler.
-
-\subsection{STG Language}
-\setlength{\tabcolsep}{5pt}
-
-The STG language is ...
-
-Our incarnation of STG is very similar to that given in EA, as is
-our presentation which is meant to evoke both an abstract and concrete syntax.
-
-\begin{table}
-\footnotesize % tiny scriptsize footnotesize small
-\centering
-\begin{tabular}{r r c l l}
-
-Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
-Constructor  & $C$            &     &                                              & Initial upper-case letter \\
-Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point \\
-Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
-\\
-Expression   & $e$            & ::= & $a$                                          & Atom \\
-             &                & $|$ & $f\ a_1\dots a_n$                            & Function application, $n\ge 1$ \\
-             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
-             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$ 
-                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
-             &                & $|$ & \texttt{case} $e$ \texttt{as} 
-                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n$\\
-\\
-Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
-             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
-\\
-Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
-             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
-             &                & $|$ &\texttt{THUNK} $e$                            & Thunk \\
-             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
-             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
-\\
-Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $n\ge 1$, distinguished $\mathit{main}$ \\
-
-\end{tabular}
-\end{table}
-
-While our interpretation of STG is a higher-order, polymorphic, pure functional
-language, some syntactic restrictions make direct programming rather tedious.
-This is discussed further in a later section.
+arguments needed by a function call are evaluated.  The STG machinery makes
+the choice of evaluation strategy easy to change, both in terms of function
+and constructor application.
 
@@
 
-\subsection{Strictness}
-
-The STG machinery makes the choice of evaluation strategy easy to change, both
-in terms of function and constructor application.
-
-\section{Current Status, and Expected Results by the time of the symposium}
-
-No bridge from ghc STG to our STG yet.
-
-At the time of writing we have a complete serial implementation: STG parser,
-type inferencer, code generator, and runtime system including a garbage
-collector.
-
-Mutual exclusion for garbage collector--heap check with every heap allocation.
-In fact garbage collector can be invoked at any point in time except when
-populating a heap object immediately after its allocation.
-
 \subsection{Mini-Haskell}
@@
 \end{verbatim}
+
+\section{Current Status, and Expected Results by the time of the symposium}
+
+No bridge from ghc STG to our STG yet.
+
+At the time of writing we have a complete serial implementation: STG parser,
+type inferencer, code generator, and runtime system including a garbage
+collector.
+
+Mutual exclusion for garbage collector--heap check with every heap allocation.
+In fact garbage collector can be invoked at any point in time except when
+populating a heap object immediately after its allocation.
 

### 2016-03-05 pubs/tfp16/paper.tex
sync multithread

@@
 \title{Strict Pure Functional Programming and Automatic Parallelization\\
-       (New Project-in-progess Paper)}
+       (New-project-in-progess Paper)}
 %
@@
 
-\emph{Preliminary paper note 1: We are reporting work in progress.  At the time
-  of writing we have a complete serial implementation, a primitive
+\emph{Preliminary paper note 1: We are reporting work in progress.  At the
+  time of writing we have a complete serial implementation, a primitive
   proof-of-concept parallel implementation, and are currently developing a
@@
 
-\emph{Note 2:  While this does not qualify as a student paper, students (undergraduate!)
-have contributed considerably to this effort, and if accepted we plan to have one of these students
-present this paper.}
+\emph{Note 2: While this does not qualify as a student paper, students
+  (undergraduate!)  have contributed considerably to this effort, and if
+  accepted we plan to have one of these students present this paper.}
+
+\emph{Note 3: Hopefully in the spirit of a student-friendly workshop, in this
+  draft we have chosen a more conversational style of presentation than
+  necessarily appropriate for a final paper, and have also highlighted the
+  major student contributions, and perhaps gratuitous detail about process.}
 
@@
   project was previously a co-developer of a strict, arguably pure functional
-  language implementation~\cite{AikenFL}.}
+  language implementation~\cite{AikenFL,FLreport89}, in conjunction with
+John Backus as a refinement of his FP language~\cite{Backus:1978}.}
 
@@
   specifically, the language should explicitly support both Hindley-Milner
-  polymorphism and unboxed types~\cite{firstclass};
+  polymorphism and unboxed types~\cite{Jones:1991};
 \item Straighforward direct linkage with C, preferably without a sophisticated
   foreign function interface, is essential;
-\item The language should be \emph{currying-friendly} as described in the ``fast
-  curry'' paper~\cite{fastcurry};
+\item The language should be \emph{currying-friendly} as described by Marlow
+and Peyton Jones~\cite{Marlow2004};
 \item Choice of degree of strictness is easily selected;
@@
 strict, and function- and constructor strict, but there are possible
-variations.  The obvious variation is function, but not constructor, strictness.
-In Haskell the distinction between constructors and functions is somewhat
-blurred because constructors behave like functions (outside of pattern
-matching), but this need not be the case.  In STG---one of ghc's intermediate
-representations---the distinction is clear: heap objects are explicitly constructed
-by \texttt{let} constructs in which constructor names are given explicitly
-(not aliased); constructor names are not functions.
+variations.  The obvious variation is function, but not constructor,
+strictness.  In Haskell the distinction between constructors and functions is
+somewhat blurred because constructors behave like functions (outside of
+pattern matching), but this need not be the case.  In STG---one of ghc's
+intermediate representations---the distinction is clear: heap objects are
+explicitly constructed by \texttt{let} constructs in which constructor names
+are given explicitly (not aliased); constructor names are not functions.
 
@@
 
+Operationally, when an application such as
+
 \subsection{Degrees of strictness, take two}
@@
 Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
-refer so many times that we subsequenty refer to it as
-\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely has another
-  connotation to the reader.}
+refer so many times that we will subsequenty refer to it as
+\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
+  has another connotation to the reader.}
 
@@
 manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
-regardless of what type it might be dynamically ``instantiated'' at.  We distinguish
-\emph{application}, e.g.\ \texttt{f x y}, which is a syntactic notion, from
-actually calling a function. Thus \texttt{id id x} is an application of \texttt{id}
-to two arguments, but the underlying \emph{FUN} (user- or system- defined function)
-when actually called is given exactly one argument.
+regardless of what argument type it might be dynamically applied to.  We
+distinguish \emph{application}, e.g.\ \texttt{f x y}, which is a syntactic
+notion, from actually calling a function. Thus \texttt{id id x} is an
+application of \texttt{id} to two arguments, but the underlying \emph{FUN}
+(user- or system- defined function) when actually called is given exactly one
+argument.
 
@@
 return type void, and with a single exception in the non-strict case are
-invoked in tail call positions, both gcc and Clang/LLVM generate jumps instead
-of calls for C calls in tail call positions.
+invoked in tail call positions, at non-zerop optimization levels both gcc and
+Clang/LLVM generate jumps instead of calls for C calls in tail call positions.
 
@@
 
-At the time of writing we have a complete serial implementation: STG parser,
-type inferencer, code generator, and runtime system including a garbage
-collector.
+At the time of writing we have a complete serial implementation, including
+\begin{itemize}
+\item STG (inc. ADT w/unboxed) parser
+\item type inferencer
+\item code generator
+\item runtime system including a garbage collector
+\end{itemize}
 
@@
 
-Parallel runtime is being implemented following the ideas of~\cite{SPJs-many-papers}, thread pool,
-stack and heap for each thread, blah, blah.
+Parallel runtime is being implemented following the ideas
+of~\cite{SPJs-many-papers}, thread pool, stack and heap for each thread, blah,
+blah.
 
@@
 
-DOE/NNSA/LANL LDRD, DOE/NNSA SULI.
+Funding for this project and its participants has been provided by the DOE
+NNSA LANL Laboratory Directed Research and Development program award
+20150845ER, the National Science Foundation Science, Technology, Engineering,
+and Mathematics Talent Expansion Program (STEP) program for undergraduate
+students, and the DOE Science Undergraduate Laboratory Internship program for
+undergraduate students.
+Los Alamos National Laboratory is managed and operated by Los Alamos National
+Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
+Department of Energy’s National Nuclear Security Administration (NNSA).
+
+% just get these into the bib
+\cite{ghc-as-library}
+\cite{Heren02}
 
@@
 %
-\begin{thebibliography}{5}
 
-\bibitem{Heroux}
-January 14, 2016
-Exascale Programming: Adapting What We Have Can (and Must) Work
-Michael A. Heroux, Sandia National Laboratories
-http://www.hpcwire.com/2016/01/14/24151/
+\bibliographystyle{splncs03}
 
-\bibitem{EUexascale}
-Toward Exascale Computing.
-European Exascale Software Initiative.
-http://www.eesi-project.eu/
-
-\bibitem{Davis96}
-MPP Parallel Haskell. 
-K. Davis. 
-Draft Proceedings Implementation of Functional Languages 1996 (IFL '96) LNCS
-1268, pp49-54, Bonn/Bad-Godesberg, Germany, LNCS 1268, Springer-Verlag.
-
-\bibitem{Bauer12} 
-Legion: Expressing Locality and Independence with Logical
-  Regions. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings
-  of the Conference on Supercomputing, pages 1-11, November 2012.
-
-\bibitem{Bauer14} 
-Structure Slicing: Extending Logical Regions with Fields. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings of the Conference on Supercomputing, pages 845-856, November 2014.
-
-\bibitem{AikenFL}
-A. Aiken, J.H. Williams, and E.L. Wimmers
-The FL Project:  The Design of a Functional Language
-(technical report not published elsewhere), September 1993.
-http://theory.stanford.edu/~aiken/publications/trs/FLProject.pdf
-
-\bibitem{Bird84}
-R. S. Bird, 
-Using circular programs to eliminate multiple traversals of data, 
-Acta Informatica, 21(3), 239-250 (1984).
-
-\bibitem{Johnsson87}
-Thomas Johnsson
-Attribute Grammars as a Functional Programming Paradigm
-Proc. of a conference on Functional programming languages and computer architecture
-Pages 154-173 
-Springer-Verlag London, UK ©1987 
-table of contents ISBN:0-387-18317-5
-
-\bibitem{Hughes89}
-J. Hughes,
-Why Functional Programming Matters,
-Computer Journal,
-32,
-2,
-98--107,
-1989.
-
-\bibitem{strict-strictdata}
-http://ghc.haskell.org/trac/ghc/wiki/StrictPragma
-
-[ghclib] GHC/As a library http://wiki.haskell.org/GHC/As\_a\_library
-
-\end{thebibliography}
+\bibliography{tfp}
 

### 2016-03-05 options.h

@@ -30,4 +30,8 @@
 /* if 1 initiate GC with every object allocation */
 #define ALLOC_GC 1
 
+/* number of threads, temporary solution */
+
+#define NTHREADS 2
+
 #endif

### 2016-03-05 runtime/stg.c

@@ -68,28 +68,6 @@ void startCheck() {
   }
 }
 
-// we can't be certain a value is boxed or unboxed without enabling ARG_TYPE
-// but we can do some sanity checking.  mayBeBoxed(v) means that v is not
-// definitely unboxed
-bool mayBeBoxed(PtrOrLiteral f) {
-  #if USE_ARGTYPE
-  return f.argType == HEAPOBJ &&
-         (f.op == NULL || isHeap(f.op) || isSHO(f.op)) &&
-         ((f.u & (OBJ_ALIGN - 1)) == 0);
-#else
-  return (f.op == NULL || isHeap(f.op) || isSHO(f.op)) &&
-         ((f.u & (OBJ_ALIGN - 1)) == 0);
-#endif
-}
-
-bool mayBeUnboxed(PtrOrLiteral f) {
-  #if USE_ARGTYPE
-  return f.argType != HEAPOBJ;
-#else
-  return true;
-#endif
-}
-
 void copyargs(PtrOrLiteral *dest, PtrOrLiteral *src, int count) {
   memcpy(dest, src, count * sizeof(PtrOrLiteral));
 }

### 2016-03-12 pubs/tfp16/paper.tex
synching

@@
 \begin{abstract}
-The trend is functional programming in scientific computing, and in particular
-strict (by default) functional programming, in various guises.  Our project is
-the demonstration of a light-weight, higher-order, polymorphic pure functional
-language implementation in which we can experiment with automatic
-parallelization strategies and varying degrees of default strictness.
-Strictness prescribes the degree of potential parallelism, that is, we are not
-exploring speculative evaluation.
+The claimed trend is an increasing appreciation and practice of functional
+programming in scientific computing, and in particular strict (by default)
+functional programming, in various guises.  Our project is the demonstration
+of a light-weight, higher-order, polymorphic pure functional language
+implementation in which we can experiment with automatic parallelization
+strategies and varying degrees of default strictness.  Strictness constrains
+the degree of potential parallelism, that is, we are not exploring speculative
+evaluation.
 
@@
 \section{Background and Motivation}
-The practice of highest-performance scientific computing (HPC) is, overall,
-highly conservative with respect to change.  Even as the US government commits
-to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
-highly respected member of the HPC community recently stated ``Fortran is
-essential for exascale programming.''~\cite{Heroux16} Existing US government
-laboratory HPC codes run to hundreds of thousands of lines and complete
-rewrites are infeasible.\footnote{A European consortium is anticipating
-  exascale computing as well~\cite{EUexascale}, but the authors are not
-  knowledgeable about the state of large-scale, non-USA HPC codes.}  C++ has
-become more common for new code starts, but their nature remains the same:
-huge codes with fair to non-existent encapsulation of side effects, and
-multiple levels of parallelism also poorly abstracted, especially at the
-thread (shared-memory) level.  Thread-level parallel efficiency can be poor
-and is explained by Amdahl's law because the parallel model is bulk
-synchronous.
+The practice of so-called high-performance scientific computing (HPC) is,
+overall, highly conservative with respect to change.  Even as the US
+government commits to a push for exascale computing ($10^{18}$ FLOPS in a
+single system) within the next decade, a highly respected member of the HPC
+community recently stated publicly that ``Fortran is essential for exascale
+programming.''~\cite{Heroux16}\footnote{Various EU, Japanese, and
+  international groups are also anticipating exascale
+  computing.~\cite{EUexascale,Exascale-org}} Existing US government laboratory
+HPC codes run to hundreds of thousands of lines of Fortran and complete
+rewrites are infeasible.~\footnote{The codes of our lab's British counterpart
+  are reportedly in the same state.}  C++ has become more common for new code
+starts, but their nature remains the same: huge codes with fair to
+non-existent encapsulation of side effects, with multiple levels of
+parallelism also poorly abstracted, especially at the thread (shared-memory)
+level.  Thread-level parallel efficiency can be poor and is explained by
+Amdahl's law because the parallel model \emph{as typically implemented} is
+bulk synchronous.
 
@@
 
-While our interpretation of STG is a higher-order, polymorphic, pure
-functional language, some syntactic restrictions make direct programming
-rather tedious.  While a near-term goal is to bridge the gap\ldots
+As mentioned, a current near-term goal is to use ghc as our front-end
+to generate STG as input to our system.  In the meantime, as a stop-gap
+measure to make the generation of test cases easier, we have implemented
+a \emph{mini-Haskell} front-end, a strict subset of standard Haskell syntax.
+
+???give a proper grammar and summarize any tricky translations???
 
@@
 undergraduate students.
+%
 Los Alamos National Laboratory is managed and operated by Los Alamos National

### 2016-03-16 pubs/tfp16/paper.tex
Start of the testing section

@@
 
+We have developed a extensive test suite using the CMake \cite{cmake-book}
+CTest tool. The full test suite is automatically run and any failures reported
+using a Jenkins continuous integration server \cite{jenkins-book}. There are
+test programs for both STG and mini-Haskell. We also run each test with various
+levels of garbage collection enabled.
+
+The test suite is partitioned into a number of directories which control what
+testing is done for a specific program. Given our goal of exploring various
+levels of strictness, most tests are run with all evaluation strategies (3
+degrees of function strictness by two degrees of constructor strictness).
+One interesting class of test programs are those which require non-strict
+semantics. For these programs we test not only that the test leads to the expected
+result with non-strict semantics, but also that the program returns a
+"blackhole" in the case of strict semantics. Another important class of test
+programs are those which should fail with a known error (e.g. a parsing error)
+for these programs the test suite checks the output against an expected error
+regular expression.
+
 Extensive prelude.

### 2016-03-18 pubs/tfp16/paper.tex
TFP paper: Some more on testing, a bit on GC. WIP...

@@
 
+We have implemented a simple copying garbage collector \cite{Cheney:1970}
+with small "nurseries" in the multithreaded case \cite{Marlow:2008}. We have not
+focused on the development of a moire sophisticated garbage collector, as the goal
+of this project is to explore levels of strictness.
+
 Mutual exclusion for garbage collector--heap check with every heap allocation.
@@
 
-We have developed a extensive test suite using the CMake \cite{cmake-book}
-CTest tool. The full test suite is automatically run and any failures reported
-using a Jenkins continuous integration server \cite{jenkins-book}. There are
-test programs for both STG and mini-Haskell. We also run each test with various
-levels of garbage collection enabled.
+We have developed a extensive test suite which uses the CMake \cite{cmake-book}
+CTest tool. The full test suite is automatically run, and any failures reported,
+after every comment to the version control system. We use a Jenkins continuous
+integration server \cite{jenkins-book} to automate this build and testing process.
+The (git) version control repository is hosted externally so our student
+collaborators can contribute at their leisure.
 
-The test suite is partitioned into a number of directories which control what
+There are several hundred small STG and mini-Haskell programs in the test suite.
+It is partitioned into a number of directories which control what
 testing is done for a specific program. Given our goal of exploring various
@@
 result with non-strict semantics, but also that the program returns a
-"blackhole" in the case of strict semantics. Another important class of test
-programs are those which should fail with a known error (e.g. a parsing error)
-for these programs the test suite checks the output against an expected error
-regular expression.
+"blackhole" in the case of strict semantics. Students who were Haskell/STG
+beginners have contributed substantially to the test suite. We have also found
+that the test suite has be of use as a learning tool for these students to
+discover  what level of non-strictness was required for their programs,
+a benefit that we did not initially envision. Over time we have developed
+extensive STG and mini-haskell preludes of common programming patterns, to
+aid in the writing of test programs.
 
-Extensive prelude.
+Another important class of test programs are those which should fail with a
+known error (e.g. a parsing error) for these programs the test suite checks
+the output against an expected error regular expression. This class of test
+programs has proved valuable in development and debugging of both the STG
+and mini-haskell frontends.
 
-Test suite of several hundred small programs.
+We also run each test with various levels of garbage collection enabled, this
+has helped us to find bugs not only in the garbage collector, but also in the
+code generator.
 
-Students learning Haskell and STG, writing Mini-Haskell and STG programs that
-serendipitously go into our automated testing suite which is partitioned
-according to our lattice of evaluation strategies (3 degrees of function
-strictness by two degrees of constructor strictness), which tells us how much
-non-strictness each program needs and something about the style in which
-programs are written.
-
-Continuous build system (Jenkins)
+debug level (runtime/compile time)
 

### 2016-03-18 pubs/tfp16/paper.tex
TFP paper: more tweaks to testing section

@@
 with small "nurseries" in the multithreaded case \cite{Marlow:2008}. We have not
-focused on the development of a moire sophisticated garbage collector, as the goal
+focused on the development of a more sophisticated garbage collector, as the goal
 of this project is to explore levels of strictness.
@@
 a benefit that we did not initially envision. Over time we have developed
-extensive STG and mini-haskell preludes of common programming patterns, to
+extensive STG and mini-Haskell preludes of common programming patterns, to
 aid in the writing of test programs.
@@
 programs has proved valuable in development and debugging of both the STG
-and mini-haskell frontends.
+and mini-Haskell front-ends.
 
@@
 code generator.
+
+The various enumerations of strictness and garbage collection levels mean that
+we run a quite large number of tests. However, this is fully automated via the
+testing and continuous integration infrastructure.
 

### 2016-03-23 pubs/tfp16/paper.tex
add Makefile

@@
         Loren Anderson\inst{1,3} \and
-        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
+        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
 }

### 2016-03-24 pubs/tfp16/paper.tex
adds (partial) mini-Haskell summary

--- 2016-03-18 pubs/tfp16/paper.tex
@@
 \usepackage{makeidx}  % allows for indexgeneration
+\usepackage{color} % for parenthetical notes while paper is in progress
 %
@@
 \title{Strict Pure Functional Programming and Automatic Parallelization\\
-       (New-project-in-progess Paper)}
+       (New-project-in-progress Paper)}
 %
@@
   polymorphism and unboxed types~\cite{Jones:1991};
-\item Straighforward direct linkage with C, preferably without a sophisticated
+\item Straightforward direct linkage with C, preferably without a sophisticated
   foreign function interface, is essential;
@@
 
-???give a proper grammar and summarize any tricky translations???
-
 First while the scrutinee of a case expression may be an arbitrary expression,
@@
 \end{verbatim}
+
+{\color{red} I didn't want to touch the above text, but there's obviously some 
+redundancy in the content here...}
+
+Writing STG in the concrete syntax conceived for our project is tedious and
+prone to errors.  Compared to Haskell, the eventual source language for the
+compiler, STG is relatively verbose, requiring explicit block and expression
+delimiters, and classification of heap objects. Though its general structure is
+similar to primitive Haskell, it lacks many of the conveniences that make
+Haskell a pleasure to work with.  The motivation for a more expressive language
+for testing our compiler in its nascent stages is then clear.  This language,
+mini-Haskell, was progressively "grown" from the STG syntax towards a subset of
+Haskell itself*, removing much of the frustrating syntactic cruft and adding
+some of the core features that are often taken for granted in Haskell (and other
+languages).
+
+{\color{red}
+* Exceptions include the "unboxed" keyword allowed in datatype definitions and
+the ability to specify unboxed ints and doubles with a postfix "\#" character.
+The former feature is not in any extension of Haskell, while the latter can be
+included using the MagicHash extension.}
+
+An informal specification of the mini-Haskell grammar is given in
+{\color{red}numbered table?}.  There is significant overlap with STG, resulting
+from the structural similarities, but the redundancies are included for
+completeness. {\color{red} should they be?}
+
+% Haskell Report:
+%   https://www.haskell.org/onlinereport/haskell2010/haskell.html
+% (PDF):
+%   https://www.haskell.org/definition/haskell2010.pdf
+
+% Chapter 10:
+%   https://www.haskell.org/onlinereport/haskell2010/haskellch10.html
+% (No standalone PDF)
+
+% because # looks strangely large next to math terms
+\newcommand{\hash}{{\scriptsize\#}}
+
+\begin{table}
+\centering
+\begin{tabular}{r r c l l}
+Variable         & $f,\ x$             &     &
+& Initial lower-case letter \\
+
+Constructor        & $C$                 &     & 
+& Initial upper-case letter \\
+
+Literal            & $\mathit{lit}$      & ::= & $i\ |\ d\ 
+                                                 |\ i$\hash\ $|\ d$\hash
+& Integral or floating point literal (boxed) \\
+
+                   &                     &     & 
+& optional \# suffix denotes unboxed value \\
+
+Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$
+& Constructors are first class \\ 
+% not sure if that's the right terminology 
+\\
+
+Expression         & $e$                 & ::= & $a$
+& Atomic expression \\
+
+                   &                     & $|$ & $e\ e$
+& Expression application \\
+
+                   &                     & $|$ & \texttt{\textbackslash} $\mathit{pat}_i$ 
+                                               \texttt{->} $e$
+& Lambda expression, $i > 0$ \\
+
+                   &                     & $|$ & \texttt{case} $e$ \texttt{of} $\mathit{alt}_i$
+& Case expression, $i > 0$\\
+
+                   &                     & $|$ & \texttt{let} $\mathit{odecl}_i$ \texttt{in} $e$
+& Recursive let expression, $i > 0$ \\
+
+\\
+
+Pattern            & $\mathit{pat}$      & ::= & $C\ \mathit{pat}_i$
+& Nested constructor matching \\
+                   &                     & $|$ & $\mathit{lit}$
+& Match numeric literals \\
+                   &                     & $|$ & $x$
+& Bind to variable \\
+
+\\
+
+Alternative        & $\mathit{alt}$      & ::= & $\mathit{pat}_i$ \texttt{->} $e$
+& $i > 0$ \\
+
+\\
+
+Type               & $\mathit{type}$     & $|$ & $C$
+& Concrete type (constructor naming) \\
+                   &                     & $|$ & $x$
+& polymorphic type variable \\
+                   &                     & $|$ & $\mathit{type}$ \texttt{->} $\mathit{type}$
+& Function type \\
+
+\\
+
+Object Declaration & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$
+& Type signature \\
+                   &                     & $|$ & $x = e$
+& Simple binding \\
+                   &                     & $|$ & $f\ \mathit{pat}_i = e$
+& Function declaration \\
+                   &                     &     &
+& $e$ cannot be an unboxed literal \\
+
+\\
+
+Constructor Definition & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$
+& $i \ge 0$ \\
+
+Datatype Declaration &  $\mathit{ddecl}$ & ::= & \texttt{data} 
+                                                 [\texttt{unboxed}]
+                                                 $C\ x_i =$ 
+& User-defined data type, $i \ge 1, n > 0$ \\
+                     &                   &     & $\mathit{con}_1 | \dots |
+                                                 \mathit{con}_n$
+& \\
+
+\\
+
+Program                & $\mathit{prog}$ & ::= & $(o|d)\mathit{decl}_i$
+& Objects and datatypes, no order necessary
+
+\end{tabular}
+\end{table}
+
+Because mini-Haskell was designed to replace STG, it makes sense to talk about
+it in terms of the differences between the two and how mini-Haskell is
+translated into STG for the backend.  Here, the differences are given in order
+of perceived utility, or importance to someone transition from STG to
+mini-Haskell.
+
+\paragraph{Removal of explicit heap objects} -\\
+
+In STG, all object definitions require explicit classification with
+\texttt{CON}, \texttt{PAP}, \texttt{THUNK}, \texttt{FUN}, or
+\texttt{BLACKHOLE}. This classification can be inferred from context by the
+compiler, so these keywords are no longer necessary.  Functions can then be
+defined in Haskell-style, with the parameters following the function's
+identifier on the left side of the equality. With these simple changes, the
+syntax becomes a subset of Haskell's*.
+
+Heap objects are still only created at the top level or in \texttt{let}
+expressions, so the translation is straightforward: \texttt{FUN} objects are
+identifiable by the inclusion of these parameters.  \texttt{PAP} objects are
+created when a known function is partially applied.  \texttt{CON} objects are
+created when constructors are fully applied.  Every other object becomes a
+\texttt{THUNK}.
+
+{\color{red}
+* With only these changes, the semantics are a little different; numeric
+  literals would still be treated as unboxed values, for example.  I don't know
+  if that's worth noting though.}
+
+\paragraph{The Layout Rule} -\\
+
+Mini-Haskell implements Haskell's Layout Rule in an almost direct translation of
+the specification given in section 10.3 of the Haskell 2010 Report.  The notable
+exception is the omission of any attempt to define the \texttt{parse-error}
+function.  This is apparently a tricky spot for Haskell compilers (even GHC has
+not always had it right) so, in the name of tradition, it remains a tricky spot
+for our mini-Haskell front end.  The edge cases that it will incorrectly handle
+as a result of this are exactly that: edge cases.  As such, they are easy to
+avoid in practical usage.
+
+The inclusion of the layout rule may seem small and insignificant, but it allows
+its users to produce code that is visually similar to Haskell code.  This makes
+it easier to read and write for any Haskell programmer who relies on layout to
+delineate context (I assume this is ~99.9% of human Haskell users).  It also
+allows for the use of Haskell programming environments that support "smart"
+indentation.
+
+% Layout Rule Spec:
+%   https://www.haskell.org/onlinereport/haskell2010/haskellch10.html\#x17-17800010.3
+% GHC's syntax "infelicities":
+%   https://downloads.haskell.org/~ghc/latest/docs/html/users\_guide/bugs-and-infelicities.html\#infelicities-syntax
+% More on parse-error:
+%   http://www.dcs.gla.ac.uk/mail-www/haskell/msg01711.html
+
+\paragraph{Expression to Expression application} -\\
+
+In STG, function application must be to atomic values, i.e. either literals or
+variables.  This often requires introducing bindings with either \texttt{let} or
+\texttt{case} expressions.  Consider the \texttt{append} function that
+concatenates its two list parameters.  In STG, it is written as:
+
+\begin{verbatim}
+append = FUN(l1 l2 ->
+            case l1 of
+              { Nil -> l2;
+                Cons hd tl -> 
+                  let { rec = THUNK(append tl l2);
+                        result = CON(Cons hd rec) }
+                  in result });
+\end{verbatim}
+
+In mini-Haskell, expressions may be applied to other expressions, so the
+recursive call to \texttt{append} can be parenthesized. There is no need for the
+\texttt{let} expression:
+
+\begin{verbatim}
+append l1 l2 = case l1 of
+                Nil -> l2
+                Cons hd tl -> Cons hd (append tl l2)
+\end{verbatim}
+
+This also serves as a good example of expression to expression application is
+transformed into STG.  Any non-atomic expression in a expression application is
+bound to a heap object in a generated \texttt{let} expression.  These
+expressions are thus properly "atomized" for STG.  The above mini-Haskell code
+is transformed in STG that is structurally identical to the STG code that
+precedes it.
+
+\paragraph{Fancy Pattern Matching} -\\
+
+{\color{red}\dots}
+
+\paragraph{Less significant features} -\\
+
+\begin{itemize}
+\item Constructors as functions \\
+  (partial application uses generated functions, full creates CONS)
+
+\item Primitive operations as functions \\
+  (partial application uses generated functions, full creates EPrimop)
+
+\item Lambda Expressions \\
+  (desugar to let-bound FUN objects)
+
+\item Type Annotations \\
+  (passed on to type checker)
+
+\item Haskell Block \\
+  (ignored by MHS frontend, not other Haskell compilers;
+  allows the same file to be tested on our compiler vs GHC)
+\end{itemize}
 

### 2016-03-24 pubs/tfp16/paper.tex
a -> an
synching

--- 2016-03-23 pubs/tfp16/paper.tex
@@
 \usepackage{makeidx}  % allows for indexgeneration
+
+% mkd for preliminary paper notes
+\usepackage{enumitem}
+
 %
@@
 functional programming, in various guises.  Our project is the demonstration
-of a light-weight, higher-order, polymorphic pure functional language
+of a light-weight, higher-order, polymorphic, pure functional language
 implementation in which we can experiment with automatic parallelization
-strategies and varying degrees of default strictness.  Strictness constrains
-the degree of potential parallelism, that is, we are not exploring speculative
-evaluation.
+strategies and varying degrees of default strictness.  Non-strictness
+constrains the degree of potential parallelism, that is, we are not exploring
+speculative evaluation.
+\\\ \\
+\emph{Preliminatry paper notes:}
+\begin{enumerate}[before=\itshape]
+  \item We are reporting work in progress.  At the time of writing we have a
+    complete serial implementation, a primitive proof-of-concept parallel
+    implementation, and are currently developing a more realistic parallel
+    runtime.  We expect to have preliminary results from our new parallel
+    implementation by the time of the workshop.
 
-\emph{Preliminary paper note 1: We are reporting work in progress.  At the
-  time of writing we have a complete serial implementation, a primitive
-  proof-of-concept parallel implementation, and are currently developing a
-  more realistic parallel runtime.  We expect to have preliminary results from
-  our new parallel implementation by the time of the symposium.}
+  \item While this does not qualify as a student paper, students
+    (undergraduate!)  have contributed considerably to this effort, and if
+    accepted we plan to have one of these students present this paper.
 
-\emph{Note 2: While this does not qualify as a student paper, students
-  (undergraduate!)  have contributed considerably to this effort, and if
-  accepted we plan to have one of these students present this paper.}
+  \item Hopefully in the spirit of a student-friendly workshop, for this draft
+    version we have chosen a more informal, conversational style of
+    presentation than necessarily appropriate for a final paper, have
+    highlighted the major student contributions and involvement, and included
+    perhaps gratuitous detail about process.
+  \end{enumerate}
 
-\emph{Note 3: Hopefully in the spirit of a student-friendly workshop, in this
-  draft we have chosen a more conversational style of presentation than
-  necessarily appropriate for a final paper, and have also highlighted the
-  major student contributions, and perhaps gratuitous detail about process.}
-
-\keywords{strict pure functional programming, automatic parallelization}
+\keywords{strict pure functional programming, automatic parallelization, Haskell, STG}
 
@@
 read/write, etc.)  and coherency requirements for each region to which it will
-have access.  Other than registered access to regions tasks must not access
+have access.  Other than registered access to regions, tasks must not access
 non-constant global data.  In serial execution these access requirements are
@@
   project was previously a co-developer of a strict, arguably pure functional
-  language implementation~\cite{AikenFL,FLreport89}, in conjunction with
+  language implementation~\cite{AikenFL,FLreport89}, in collaboration with
 John Backus as a refinement of his FP language~\cite{Backus:1978}.}
@@
 users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
-primary data structure, and laziness is more of a nuisance--even a serious
-hindrance--than a convenience, much less ever algorithmically essential in the
-sense of Bird's \emph{repmin}~\cite{Bird84} or Johnsson's general attribute
-grammars~\cite{Johnsson87}.
+primary data structure, and non-strictness or laziness is more of a
+nuisance--even a serious hindrance--than a convenience, much less ever
+algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84}
+or Johnsson's general attribute grammars~\cite{Johnsson87}.
 
@@
 using e.g.\ MPI or POSIX sockets.
+
+A secondary, informal goal is to gain a sense of how often relatively
+inexperienced functional programmers make essential use of
+non-strictness.\footnote{The starting experience of our student interns to
+  date has been normalized by having them complete Hutton's \emph{Programming in Haskell} textbook
+as baseline preparation.~\cite{Hutton-book}}
 

### 2016-03-25 pubs/tfp16/paper.tex

--- 2016-03-24 pubs/tfp16/paper.tex
@@
 \usepackage{makeidx}  % allows for indexgeneration
+% mkd for preliminary paper notes
+\usepackage{enumitem} 
 \usepackage{color} % for parenthetical notes while paper is in progress
@@
 
+{\color{red}
+* With only these changes, the semantics are a little different; numeric
+  literals would still be treated as unboxed values, for example.  I don't know
+  if that's worth noting though.}
+
 Heap objects are still only created at the top level or in \texttt{let}
@@
 \texttt{THUNK}.
-
-{\color{red}
-* With only these changes, the semantics are a little different; numeric
-  literals would still be treated as unboxed values, for example.  I don't know
-  if that's worth noting though.}
 
@@
 This also serves as a good example of expression to expression application is
-transformed into STG.  Any non-atomic expression in a expression application is
+transformed into STG.  Any non-atomic expression in an application expression is
 bound to a heap object in a generated \texttt{let} expression.  These

### 2016-03-25 pubs/tfp16/paper.tex
mac->dos->unix
minor edits

@@
 %
-\title{Strict Pure Functional Programming and Automatic Parallelization\\
+\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
        (New-project-in-progress Paper)}
@@
 constrains the degree of potential parallelism, that is, we are not exploring
-speculative evaluation.
-\\\ \\
+speculative evaluation.  A secondary goal is to experiment with mechanisms
+for transparent fault tolerance.
+\\
 \emph{Preliminatry paper notes:}
@@
     runtime.  We expect to have preliminary results from our new parallel
-    implementation by the time of the workshop.
+    implementation by the time of the workshop.  The fault-tolerance mechanisms
+    remain future work.
 
   \item While this does not qualify as a student paper, students
-    (undergraduate!)  have contributed considerably to this effort, and if
-    accepted we plan to have one of these students present this paper.
+    (undergraduate!)  have contributed considerably to this effort, and to the
+    writing of this paper, and will continue to do so.  If this paper is
+    accepted we plan to have one of these students present it at TiFP.
 
@@
 government commits to a push for exascale computing ($10^{18}$ FLOPS in a
-single system) within the next decade, a highly respected member of the HPC
-community recently stated publicly that ``Fortran is essential for exascale
-programming.''~\cite{Heroux16}\footnote{Various EU, Japanese, and
-  international groups are also anticipating exascale
-  computing.~\cite{EUexascale,Exascale-org}} Existing US government laboratory
-HPC codes run to hundreds of thousands of lines of Fortran and complete
-rewrites are infeasible.~\footnote{The codes of our lab's British counterpart
-  are reportedly in the same state.}  C++ has become more common for new code
-starts, but their nature remains the same: huge codes with fair to
-non-existent encapsulation of side effects, with multiple levels of
+single system) within the next decade,~\cite{Exascale-roadmap} a highly
+respected member of the HPC community recently stated publicly that ``Fortran
+is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
+  Japanese, and international groups are also anticipating the realization
+  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
+laboratory HPC codes run to hundreds of thousands of lines of Fortran and
+complete rewrites are infeasible.~\footnote{The codes of our lab's British
+  counterpart are reportedly in the same state.}  C++ has become more common
+for new code starts, but their nature remains the same: huge codes with fair
+to non-existent encapsulation of side effects, with multiple levels of
 parallelism also poorly abstracted, especially at the thread (shared-memory)
@@
 \begin{quote}
-High-performance Haskell code (e.g. numeric code) can sometimes be littered
+  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
   with bang patterns, making it harder to read. The reason is that laziness
-isn't the right default in this particular code, but the programmer has no way
-to say that except by repeatedly adding bang patterns~\cite{strict-strictdata}.
+  isn't the right default in this particular code, but the programmer has no
+  way to say that except by repeatedly adding bang
+  patterns~\cite{strict-strictdata}.
 \end{quote}
+It is a fact that \emph{HPC practioners are by and large not computer scientists}.
+The concept of a space leak, for example, caused by other than a missing explicit
+deallocation statement (C \emph{free}, etc.) is difficult to understand and not
+something they particularly want to think about.  Hence, we eschew explicit
+strictness annotations for varying uniform degrees of strictness.
 
@@
 
-Intel's TBB?  Cilk++?
+More generally, \emph{task-based} parallelism is becoming increasingly popular
+for HPC, often in conjunction with a higher-level (typically inter-node) bulk
+synchronous model typically implemented with the Message Passing Interface
+(MPI)~\cite{MPI}.  In the imperative world, example systems include the
+aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007}, and the
+HPX C++11/14 extension~\cite{HPX}.  While the concept of task is somewhat
+vague, the general idea is a unit of computation with no or
+minimal/constrained side effects.  Function call in a pure functional language
+could be regarded as the essence of the concept of task.
+
+In the quest for computing platforms approaching exascale, it is widely
+recognized that new mechanisms for fault tolerance will need to be built into
+the software stack---the scalability of checkpoint/restart may be reaching its
+end.  In the imperative world side-effect-free tasks could be safely restarted
+after failure, and the Legion system is being augmented with this capability.
+Ericsson's quasi-functional language Erlang and its runtime system was
+designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been recognized
+that the pure parts of a functional program (excluding, e.g., the I/O monad
+in Haskell) could be safely restarted.
 
@@
 meanings or implications.  First and foremost it means that \emph{we are not
-  attempting to compete with ghc in any respect}.\footnote{We estimate that we
-  will have, to within an order of magnitude, one person-hour of available
-  development time for each person-year that ghc currently represents, or put
-  another way, one second for each hour, respectively.}
+  attempting to compete with ghc in any respect}---succumbing to the
+temptation of recreating ghc language, type system, and runtime features, or
+significant code optimization, would simply divert us from our central
+goals.\footnote{We estimate that we will have, to within an order of
+  magnitude, one person-hour of total available development time for each
+  person-year that ghc currently represents, or put another way, one second
+  for each hour, respectively.}
 
@@
   \item The language should be \emph{currying-friendly} as described by Marlow
-and Peyton Jones~\cite{Marlow2004};
+and Peyton Jones~\cite{Marlow:2004};
 \item Choice of degree of strictness is easily selected;
@@
 non-strictness.\footnote{The starting experience of our student interns to
-  date has been normalized by having them complete Hutton's \emph{Programming in Haskell} textbook
-as baseline preparation.~\cite{Hutton-book}}
+  date has been normalized by having them complete Hutton's \emph{Programming
+    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
+  substituting his article \emph{Higher-order Functions for
+    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
+  book.}
 
@@
 There are various possible ways to accomplish the former: using ghc's option
-to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
+to dump STG as text, hacking ghc itself
 and, optionally, using ghc as a library.  We note that this approach has
@@
 of the STG machine described later.  Our description is based on that given in
-Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
+Marlow and Peyton Jones' \emph{fast curry} paper~\cite{Marlow:2004}, to which we
 refer so many times that we will subsequenty refer to it as
@@
 the runtime system is written in C/C++.  The system fairly faithfully
-implements the STG machine as described in Peyton-Jones and Marlow's ``fast
-curry'' paper~\cite{fastcurry} and dynamically illustrated by Pope's
-miniSTG~\cite{ministg} with the exception of the state transition rules which,
-of course, are different than those for lazy semantics.  Sharing is preserved.
+implements the STG machine as described in \emph{EA} and dynamically
+illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
+transition rules which, of course, may be different than those for non-strict
+semantics.  Sharing is preserved.
 
@@
 
-Proper tail calling is achieved.  Because we maintain our own data and control
-(continuation) stack all generated C functions take no arguments and have
-return type void, and with a single exception in the non-strict case are
-invoked in tail call positions, at non-zerop optimization levels both gcc and
-Clang/LLVM generate jumps instead of calls for C calls in tail call positions.
+While performance is not a high priority, we attain proper tail calling instead of
+using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
+Because we maintain our own data and control (continuation) stack all
+generated C functions take no explicit arguments and have return type void,
+and with a single exception in the non-strict case are invoked in tail call
+positions, at non-zero optimization levels both gcc and Clang/LLVM generate
+jumps instead of calls for C calls in tail call positions.  When using gcc
+we also use a fixed set of machine registers for passing/returning values.
 
 The exceptions to tail calling are for convenience.  Similar to what is
-described in the \emph{fast curry} paper we have a generic variadic stgApply
-function that is used for applications of non-known functions, and known
-functions with a deficit or excess number of arguments.  The stgApply function
-takes as arguments the first element of the application (FUN, PAP, THUNK, or
-BLACKHOLE) and its arguments, and if a THUNK, evaluates it with a call (rather
-than jump) to yield a FUN or PAP.  This allows us to implement a single
-stgApply function as a single C function.  The implementation of stgApply as
-described in \emph{fast curry} makes the same concession.
+described in the \emph{EA} we have a generic variadic stgApply function that
+is used for applications of non-known functions, and known functions with a
+deficit or excess number of arguments.  The stgApply function takes as
+arguments the first element of the application (FUN, PAP, THUNK, or BLACKHOLE)
+and its arguments, and if a THUNK, evaluates it with a call (rather than jump)
+to yield a FUN or PAP.  This allows us to implement a single stgApply function
+as a single C function taking a contiguous block of machine words (unboxed
+values or pointers) on the explicitly stack.  The implementation of stgApply
+as described in \emph{EA} makes the same concession regarding
+call-with-return, but the scheme described there uses an auto-generated family
+of stgApply functions parameterized by number and type (unboxed or pointer) of
+arguments.  We used this scheme initially as well, but because the parameter
+block contains a bitmap describing which arguments are boxed/unboxed for the
+benefit of the garbage collector, this elaboration is unnecessary.
 
-For stgApply in the non-strict semantics the alternative to call-with-return
+For stgApply in the non-strict semantics an alternative to call-with-return
 would be a pair of functions stgApply1 and stgApply2.  In one possible variation,
@@
 
-We have implemented a simple copying garbage collector \cite{Cheney:1970}
-with small "nurseries" in the multithreaded case \cite{Marlow:2008}. We have not
-focused on the development of a more sophisticated garbage collector, as the goal
-of this project is to explore levels of strictness.
+We have implemented a simple copying garbage collector \cite{Cheney:1970} with
+small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
+not focused on the development of a more sophisticated garbage collector
+because the goal of this project is to explore automatic parallelization.
 
@@
 with built-in and user-defined unboxed types as described by Peyton-Jones and
-Launchbury's seminal work~\cite{unboxed}, but for unboxed types code
+Launchbury's seminal work~\cite{Jones91unboxedvalues}, but for unboxed types code
 generation is currently limited to built-in single-word (64-bit) or smaller
@@
 \section{Other Related Work}
+
+In his BSc thesis, Don Stewart described an implementation wherein ghc was used as a Haskell
+compiler front-end, dumping STG code as text, parsing that text, then translating to code for
+execution on the Java Virtual Machine.~\cite{Stewart-BSc}.
+
 Nvidia research compiler.
@@
 NNSA LANL Laboratory Directed Research and Development program award
-20150845ER, the National Science Foundation Science, Technology, Engineering,
+20150845ER; the National Science Foundation Science, Technology, Engineering,
 and Mathematics Talent Expansion Program (STEP) program for undergraduate
-students, and the DOE Science Undergraduate Laboratory Internship program for
-undergraduate students.
+students; and the DOE Science Undergraduate Laboratory Internship (SULI) program.
 %

### 2016-03-28 pubs/tfp16/paper.tex
incremental work on MHS writeup

--- 2016-03-25 pubs/tfp16/paper.tex
@@
 \usepackage{makeidx}  % allows for indexgeneration
-% mkd for preliminary paper notes
-\usepackage{enumitem} 
 \usepackage{color} % for parenthetical notes while paper is in progress
@@
 functional programming, in various guises.  Our project is the demonstration
-of a light-weight, higher-order, polymorphic, pure functional language
+of a light-weight, higher-order, polymorphic pure functional language
 implementation in which we can experiment with automatic parallelization
-strategies and varying degrees of default strictness.  Non-strictness
-constrains the degree of potential parallelism, that is, we are not exploring
-speculative evaluation.
-\\\ \\
-\emph{Preliminatry paper notes:}
-\begin{enumerate}[before=\itshape]
-  \item We are reporting work in progress.  At the time of writing we have a
-    complete serial implementation, a primitive proof-of-concept parallel
-    implementation, and are currently developing a more realistic parallel
-    runtime.  We expect to have preliminary results from our new parallel
-    implementation by the time of the workshop.
+strategies and varying degrees of default strictness.  Strictness constrains
+the degree of potential parallelism, that is, we are not exploring speculative
+evaluation.
 
-  \item While this does not qualify as a student paper, students
+\emph{Preliminary paper note 1: We are reporting work in progress.  At the
+  time of writing we have a complete serial implementation, a primitive
+  proof-of-concept parallel implementation, and are currently developing a
+  more realistic parallel runtime.  We expect to have preliminary results from
+  our new parallel implementation by the time of the symposium.}
+
+\emph{Note 2: While this does not qualify as a student paper, students
   (undergraduate!)  have contributed considerably to this effort, and if
-    accepted we plan to have one of these students present this paper.
+  accepted we plan to have one of these students present this paper.}
 
-  \item Hopefully in the spirit of a student-friendly workshop, for this draft
-    version we have chosen a more informal, conversational style of
-    presentation than necessarily appropriate for a final paper, have
-    highlighted the major student contributions and involvement, and included
-    perhaps gratuitous detail about process.
-  \end{enumerate}
+\emph{Note 3: Hopefully in the spirit of a student-friendly workshop, in this
+  draft we have chosen a more conversational style of presentation than
+  necessarily appropriate for a final paper, and have also highlighted the
+  major student contributions, and perhaps gratuitous detail about process.}
 
-\keywords{strict pure functional programming, automatic parallelization, Haskell, STG}
+\keywords{strict pure functional programming, automatic parallelization}
 
@@
 read/write, etc.)  and coherency requirements for each region to which it will
-have access.  Other than registered access to regions, tasks must not access
+have access.  Other than registered access to regions tasks must not access
 non-constant global data.  In serial execution these access requirements are
@@
   project was previously a co-developer of a strict, arguably pure functional
-  language implementation~\cite{AikenFL,FLreport89}, in collaboration with
+  language implementation~\cite{AikenFL,FLreport89}, in conjunction with
 John Backus as a refinement of his FP language~\cite{Backus:1978}.}
@@
 users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
-primary data structure, and non-strictness or laziness is more of a
-nuisance--even a serious hindrance--than a convenience, much less ever
-algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84}
-or Johnsson's general attribute grammars~\cite{Johnsson87}.
+primary data structure, and laziness is more of a nuisance--even a serious
+hindrance--than a convenience, much less ever algorithmically essential in the
+sense of Bird's \emph{repmin}~\cite{Bird84} or Johnsson's general attribute
+grammars~\cite{Johnsson87}.
 
@@
 using e.g.\ MPI or POSIX sockets.
-
-A secondary, informal goal is to gain a sense of how often relatively
-inexperienced functional programmers make essential use of
-non-strictness.\footnote{The starting experience of our student interns to
-  date has been normalized by having them complete Hutton's \emph{Programming in Haskell} textbook
-as baseline preparation.~\cite{Hutton-book}}
 
@@
 
-\paragraph{Fancy Pattern Matching} -\\
+\paragraph{Nested Pattern Matching} -\\
 
-{\color{red}\dots}
+STG supports pattern matching but only in a simple form and only in
+\texttt{case} expressions.  mini-Haskell extends this to something approaching
+what is found in Haskell.  Functions, including lambda expressions, may match on
+all their arguments (with multiple non-anonymous definitions allowed to exhaust
+the pattern combinations).  Patterns in all contexts (functions and case
+expressions) may be nested arbitrarily deep.  These two features greatly
+simplify the programmer's job.
 

### 2016-03-28 codegen/MHS/Parser.hs
some comment cleanup

@@ -199,11 +199,12 @@ exprP =
                                              accept $ foldl1 EAp es  
 -- if only one Exp is parsed, foldl1 leaves it unchanged,
 -- otherwise left-associative application is parsed, so
--- e1 e2 e3 e4 == ((((e1) e2) e3) e4)
--- let, case and lambda expressions are all treated slightly differently
--- to conform with haskell parsing of function application
--- case (+) of f -> f; 1 2 produces a compile error (even with braces in alts)
--- but surrounding it with parens makes the application parse correctly
+--- e1 e2 e3 e4 == ((((e1) e2) e3) e4)
+-- let, case and lambda expressions are all treated slightly
+-- differently to conform with haskell parsing of function application
+-- case (+) of f -> f; 1 2 produces a compile error (even with braces
+-- in alts) but surrounding it with parens allows the case expression
+-- to be applied to 1 and 2.
   in orExList
      [eLetP, eCaseP, eFunP, mults]
 

### 2016-03-28 codegen/MHS/Transform.hs

@@ -554,11 +554,21 @@ instance SimpleExp Clause where
 
 {---------------------------------------------------------------------------------
 Desugar numeric literals to Boxed Int/Doubles
-This transformation adds its own datatype definitions for boxed Ints and Doubles
-as well as functions that call to them in the same convention as that used when
-generating functions for constructors used by a program.  This keeps the transform
-from being reliant on any previous transform while still compliant with the naming
-convention used to generate the abstract STG
+
+This transformation adds its own datatype definitions for boxed Ints
+and Doubles as well as functions that call to them in the same
+convention as that used when generating functions for constructors
+used by a program.  This keeps the transform from being reliant on any
+previous transform while still compliant with the naming convention
+used to generate the abstract STG
+
+Note to future David (from past David) so you don't panic:
+
+This does not modify unboxed literals. It does create top level boxed
+objects for each unique literal number.  Unboxed literals do not need
+to be processed specially, or, at least not here.  Their behavior in
+MHS should be identical to that found in MagicHash programs (i.e. no
+recursive bindings, including at the top level).
 -----------------------------------------------------------------------------------}
 
 

### 2016-03-28 pubs/tfp16/paper.tex

--- 2016-03-25 pubs/tfp16/paper.tex
@@
 
-{\color{red}\dots}
+STG supports pattern matching but only in a simple form and only in
+\texttt{case} expressions.  mini-Haskell extends this to something approaching
+what is found in Haskell.  Functions, including lambda expressions, may match on
+all their arguments (with multiple non-anonymous definitions allowed to exhaust
+the pattern combinations).  Patterns in all contexts (functions and case
+expressions) may be nested arbitrarily deep.  These two features greatly
+simplify the programmer's job.
 

### 2016-03-29 pubs/tfp16/paper.tex
pm edits

--- 2016-03-25 pubs/tfp16/paper.tex
@@
 \title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
-       (New-project-in-progress Paper)}
+       (New project paper)}
 %
@@
 \\
-\emph{Preliminatry paper notes:}
+\emph{Draft paper notes:}
 \begin{enumerate}[before=\itshape]
@@
 government commits to a push for exascale computing ($10^{18}$ FLOPS in a
-single system) within the next decade,~\cite{Exascale-roadmap} a highly
+single system) within the next decade~\cite{Exascale-roadmap} a highly
 respected member of the HPC community recently stated publicly that ``Fortran
@@
 laboratory HPC codes run to hundreds of thousands of lines of Fortran and
-complete rewrites are infeasible.~\footnote{The codes of our lab's British
+complete rewrites are infeasible.\footnote{The codes of our lab's British
   counterpart are reportedly in the same state.}  C++ has become more common
@@
 
-A common complaint among these budding high-performance functional programmers
-is the need to get strictness just right: to remove space leaks, achieve
-acceptable performance, get communication (e.g.\ MPI) and IO to work as they
-expect, and so on.  More specifically, they expect to be able to reason about
-space and time usage using their existing mental models. While they are happy
-to write code that is strongly reminiscent of the numerical examples in
-Hughes' classic piece (whether they've read it or not)~\cite{Hughes89}, they
-would also be happy to write in a slightly modified style that does not
-require non-strictness---after all, that's how they've been thinking their
-entire careers.
+A common complaint among these aspiring high-performance functional
+programmers is the need to get strictness just right: to remove space leaks,
+achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
+as they expect, and so on.  More specifically, they expect to be able to
+reason about space and time usage, and order of execution, using their
+existing mental models. While they are happy to write code that is strongly
+reminiscent of the numerical examples in Hughes' classic piece~\cite{Hughes89}
+(whether they've read it or not), they would also be happy to write in a
+slightly modified style that does not require non-strictness---after all,
+that's how they've been thinking their entire careers.
 
@@
 deallocation statement (C \emph{free}, etc.) is difficult to understand and not
-something they particularly want to think about.  Hence, we eschew explicit
+something they particularly want to try to understand.  Hence, we eschew explicit
 strictness annotations for varying uniform degrees of strictness.
-
-We claim, then, that there is a trend towards \emph{strict-by-default pure
-  functional programming}, whether as imposed by programmer discipline, a parallel
-runtime system (such as the aforementioned Legion, wherein the pure semantics
-at function granularity allows the runtime system to relax the nominal serial
-ordering), or the language itself (e.g.\ strict Haskell).
 
@@
 recognized that new mechanisms for fault tolerance will need to be built into
-the software stack---the scalability of checkpoint/restart may be reaching its
-end.  In the imperative world side-effect-free tasks could be safely restarted
-after failure, and the Legion system is being augmented with this capability.
+the software stack.  Current practice in HPC is checkpoint/restart, wherein at
+various point in time sufficient program state is dumped to backing store such
+that the program, in case of a fault, can restart from the latest stored
+state.  Most implementations are explicit, though there has been some success
+with transparent (invisible to the programmer) systems.  Wiring
+checkpoint/restart into a large-scale application is not only burdensome and
+error-prone, it can severely warp the engineering of a large-scale application
+because it must be designed such that essential state can be readily captured.
+And now, it is widely believed that the scalability of checkpoint/restart may
+be reaching its end.
+
+In the imperative world side-effect-free tasks could be safely restarted after
+failure, needing only their well-defined inputs and not arbitrary access to
+mutable global state.  The Legion system is being augmented with this
+capability, and in such a way that it will be largely transparent to the programmer.
 Ericsson's quasi-functional language Erlang and its runtime system was
-designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been recognized
-that the pure parts of a functional program (excluding, e.g., the I/O monad
-in Haskell) could be safely restarted.
+designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
+recognized that the pure parts of a functional program (excluding, e.g., the
+I/O monad in Haskell) could be safely restarted.  The recent work of
+[Trinder's student] impressively demonstrates this for an augmented ghc/runtime
+on a large-scale distributed-memory system.
+
+We claim, then, that there is a trend, in HPC at least, towards
+\emph{strict-by-default pure functional programming}, whether as imposed by
+programmer discipline, a parallel runtime system, or the language itself
+(e.g.\ strict Haskell).  Second, we claim that there is a trend in
+high-performance computing, and by transitivity functional programming,
+towards transparent fault tolerance.
 
@@
 experiment with automatic parallelization strategies with degrees of language
-strictness ranging from lazy to fully strict.  Light-weight has several
-meanings or implications.  First and foremost it means that \emph{we are not
-  attempting to compete with ghc in any respect}---succumbing to the
-temptation of recreating ghc language, type system, and runtime features, or
-significant code optimization, would simply divert us from our central
-goals.\footnote{We estimate that we will have, to within an order of
-  magnitude, one person-hour of total available development time for each
-  person-year that ghc currently represents, or put another way, one second
-  for each hour, respectively.}
+strictness ranging from lazy to fully strict, and secondarily, with mechanisms
+for transparent fault tolerance.  Light-weight has several meanings or
+implications.  First and foremost it means that \emph{we are not attempting to
+  compete with ghc in any respect}---succumbing to the temptation of
+recreating ghc language, type system, and runtime features, or significant
+code optimization, would simply divert us from our central goals.\footnote{We
+  estimate that we will have, to within an order of magnitude, one person-hour
+  of total available development time for each person-year that ghc currently
+  represents, or put another way, one second for each hour, respectively.}
 
@@
 and Peyton Jones~\cite{Marlow:2004};
-\item Choice of degree of strictness is easily selected;
+\item Choice of degree of strictness should be easily selected;
 \item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
@@
 \item Tail calling should be properly implemented, i.e., not as a trampoline;
-\item etc.
 \end{itemize}
@@
 
-A secondary, informal goal is to gain a sense of how often relatively
+A minor, informal goal is to gain a sense of how often relatively
 inexperienced functional programmers make essential use of
@@
 sense, meaning that argument(s) are evaluated before being passed in a
-function call, rather than the denotational sense.  Similarly, when referring
-to language semantics we refer to the operational rather than denotational
-semantics.
+function call, rather than the denotational sense, unless stated otherwise.
+Similarly, when referring to language semantics we refer by default to the
+operational rather than denotational semantics.
 
@@
 somewhat blurred because constructors behave like functions (outside of
-pattern matching), but this need not be the case.  In STG---one of ghc's
-intermediate representations---the distinction is clear: heap objects are
-explicitly constructed by \texttt{let} constructs in which constructor names
-are given explicitly (not aliased); constructor names are not functions.
+pattern matching), but this need not be the case.
 
@@
 is part of our master plan.  In short, our plan is to use ghc as our front-end
-to generate its intermediate form, STG, then escape to our system from there.
-There are various possible ways to accomplish the former: using ghc's option
-to dump STG as text, hacking ghc itself
-and, optionally, using ghc as a library.  We note that this approach has
-a precedent in the aforementioned Nvidia research compiler.
+to generate its intermediate form, STG~\cite{PJ-stockhardware}, then escape to
+our system from there.  There are various possible ways to accomplish the
+former: using ghc's option to dump STG as text, hacking ghc itself and,
+optionally, using ghc as a library.
 
@@
 
-The STG language is ...
+Our incarnation of STG, as is our presentation in Figure~\ref{STGsyntax} which
+is meant to evoke both an abstract and concrete syntax, are very similar to
+that given in Marlow and Peyton Jones' \emph{fast curry}
+paper~\cite{Marlow:2004}, to which we refer so many times that we will
+subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
+  the title because \emph{FC} likely has another connotation to the reader.}
 
-Our incarnation of STG is very similar to that given in EA, as is
-our presentation which is meant to evoke both an abstract and concrete syntax.
-
-\begin{table}
+\begin{figure}
+%\centering
+\begin{centering}
 \footnotesize % tiny scriptsize footnotesize small
-\centering
 \begin{tabular}{r r c l l}
@@
              &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
-             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$
-                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
-             &                & $|$ & \texttt{case} $e$ \texttt{as}
-                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n \ge 1$\\
+
+             &                & $|$ & \texttt{let \{}                              & Recursive let, $n\ge 1$ \\
+             &                &     & \texttt{ } $x_1$ \texttt{=} $\mathit{obj}_1$ \texttt{;} \\
+             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
+             &                &     & \texttt{ } $x_n$ \texttt{=} $\mathit{obj}_n$ \\
+             &                &     & \texttt{\} in} $e$  \\
+
+             &                & $|$ & \texttt{case} $e$ \texttt{of} $x$ \texttt{\{}  & Case expression, $n \ge 1$\\
+             &                &     & \texttt{ } $\mathit{alt}_1$ \texttt{;} \\
+             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
+             &                &     & \texttt{ } $\mathit{alt}_n$ \\
+             &                &     & \texttt{\}} \\
 \\
@@
 \\
-Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $1\le i \le n \ge 1$, distinguished $\mathit{main}$ \\
+Program      & $\mathit{prog}$& ::= & $f_1\ =\ \mathit{obj}_1$ \texttt{;}          & $n \ge 1$, distinguished \texttt{main}\\
+             &                &     & \texttt{ } $\dots$ \texttt{;} \\
+             &                &     & $f_n\ =\ \mathit{obj}_n$
 
 \end{tabular}
-\end{table}
+\end{centering}
+\caption{STG syntax}
+\label{STGsyntax}
+\end{figure}
 
-While our interpretation of STG is as a simple higher-order, polymorphic, pure
-functional language, some syntactic restrictions make direct programming
-rather tedious.  This is discussed further in a later section.
+Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}), i.e., a thunk being
+evaluated, are only runtime entities and not actually part of our STG language.
 
-Operationally, when an application such as
+The semantics of STG is what one would expect for a higher-order, polymorphic,
+pure functional language, with one exception for a Haskell programmer familiar
+with unboxed types:  \texttt{case} is strict in the scrutinee, so
+\texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell,
+that is, the scrutinee is evaluated first.
+
+In STG---one of ghc's
+intermediate representations---the distinction is clear: heap objects are
+explicitly constructed by \texttt{let} constructs in which constructor names
+are given explicitly (not aliased); constructor names are not functions.
+
+A distinctive feature of STG is that heap allocation is explicit; this is exactly
+what \texttt{let} bindings do.  As such unboxed values cannot be let-bound; instead
+unboxed values can be bound with \texttt{case}:
+\begin{verbatim}
+    case e of x# {
+      _ -> e
+    }
+\end{verbatim}
+where the \texttt{\#} suffix is merely a convention to signify an unboxed type.
+Deferred evaluation is made explicit:  a deferred expression evaluation is
+encoded as a thunk.
 
@@
 function being applied.  This is readily explained in terms of the operation
-of the STG machine described later.  Our description is based on that given in
-Marlow and Peyton Jones' \emph{fast curry} paper~\cite{Marlow:2004}, to which we
-refer so many times that we will subsequenty refer to it as
-\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
-  has another connotation to the reader.}
+of the STG.  Our partial description is based on that given in \emph{EA}.
 
@@
 manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
-regardless of what argument type it might be dynamically applied to.  We
-distinguish \emph{application}, e.g.\ \texttt{f x y}, which is a syntactic
-notion, from actually calling a function. Thus \texttt{id id x} is an
-application of \texttt{id} to two arguments, but the underlying \emph{FUN}
+regardless of the type of the argument (e.g., some function type) it might be
+applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
+a syntactic notion, from actually calling a function. Thus \texttt{id id x} is
+an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
 (user- or system- defined function) when actually called is given exactly one
-argument.
+argument (the first one).
 
@@
 \emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
-arguments already provided.  If the arity is greater than two then a new
+arguments previously provided.  If the arity is greater than two then a new
 \emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
@@
 (it must be at least one, so one in this case) the underlying \emph{FUN} is
-called-with-return with any previous argument (in case of \emph{PAP}) and new
+called-with-return with any previous arguments (in case of \emph{PAP}) and new
 argument \texttt{x}, then the object returned applied to remaining argument
@@
 the choice of evaluation strategy easy to change, both in terms of function
-and constructor application.
+application and building a constructor in the heap.
 
@@
 
-\subsubsection{Proper tail calls}
-
+\subsubsection{Proper tail calls.}
 While performance is not a high priority, we attain proper tail calling instead of
@@
 The exceptions to tail calling are for convenience.  Similar to what is
-described in the \emph{EA} we have a generic variadic stgApply function that
-is used for applications of non-known functions, and known functions with a
-deficit or excess number of arguments.  The stgApply function takes as
-arguments the first element of the application (FUN, PAP, THUNK, or BLACKHOLE)
-and its arguments, and if a THUNK, evaluates it with a call (rather than jump)
-to yield a FUN or PAP.  This allows us to implement a single stgApply function
-as a single C function taking a contiguous block of machine words (unboxed
-values or pointers) on the explicitly stack.  The implementation of stgApply
+described in \emph{EA} we have a generic variadic \emph{stgApply} function
+that is used for applications of non-known functions, and known functions with
+a deficit or excess number of arguments.  The \emph{stgApply} function takes
+as arguments the first element of the application (\emph{FUN}, \emph{PAP},
+\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
+evaluates it with a call (rather than jump) to yield a \emph{FUN} or
+\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
+a C function taking a contiguous block of machine words (unboxed values
+or pointers) on the explicit stack.  The implementation of \emph{stgApply}
 as described in \emph{EA} makes the same concession regarding
 call-with-return, but the scheme described there uses an auto-generated family
-of stgApply functions parameterized by number and type (unboxed or pointer) of
-arguments.  We used this scheme initially as well, but because the parameter
-block contains a bitmap describing which arguments are boxed/unboxed for the
-benefit of the garbage collector, this elaboration is unnecessary.
+of \emph{stgApply} functions parameterized by number and type (unboxed or
+pointer) of arguments.  We used this scheme initially as well, but because the
+parameter block contains a bitmap describing which arguments are boxed/unboxed
+for the benefit of the garbage collector, this elaboration is unnecessary.
 
-For stgApply in the non-strict semantics an alternative to call-with-return
-would be a pair of functions stgApply1 and stgApply2.  In one possible variation,
-stgApply1 would push a continuation with stgApply2 as the return address, then
-invoke the first element of the application.  Call-with-return becomes more
-convenient when considering the various forms of strictness:  when evaluating
-function or constructor arguments it is simpler, and arguably more efficient
-in terms of time (but not space in general) to ``evaluate in place'' than
-implement the continuation style to the logical limit.
+For \emph{stgApply} in the non-strict semantics an alternative to
+call-with-return would be a pair of functions \emph{stgApply1} and
+\emph{stgApply2}.  In one possible variation, \emph{stgApply1} would push a
+continuation with \emph{stgApply2} as the return address, then invoke the
+first element of the application.  Call-with-return becomes more convenient
+when considering the various forms of strictness: when evaluating function or
+constructor arguments it is simpler, and arguably more efficient in terms of
+time (but not space in general) to ``evaluate in place'' than implement the
+continuation style to the logical limit.
 
 \subsection{Mini-Haskell}
-
-As mentioned, a current near-term goal is to use ghc as our front-end
-to generate STG as input to our system.  In the meantime, as a stop-gap
-measure to make the generation of test cases easier, we have implemented
-a \emph{mini-Haskell} front-end, a strict subset of standard Haskell syntax.
-
-First while the scrutinee of a case expression may be an arbitrary expression,
-arguments in function applications, constructor allocation, and primitive
-operations are constrained to be atoms.  For unboxed types a case
-a trivial case expression to alleviates
-the difficulty.
+As mentioned, a current near-term goal is to use ghc as our front-end to
+generate STG as input to our system.  In the meantime, as a stop-gap measure
+to make the generation of test cases easier, we have implemented a
+\emph{mini-Haskell} front-end that supports a subset of standard Haskell
+syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
+\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
+unboxed literal value.  Our extension (to STG as well)
+is the provision of the optional \texttt{unboxed} keyword in datatype definitions
+to allow user-defined unboxed data types, e.g.,
 \begin{verbatim}
-  f ... e ... -> case e as x of _ -> f ... x ...
+  data unboxed Bool# = False# | True# .
 \end{verbatim}
-
-{\color{red} I didn't want to touch the above text, but there's obviously some 
-redundancy in the content here...}
 
@@
 similar to primitive Haskell, it lacks many of the conveniences that make
-Haskell a pleasure to work with.  The motivation for a more expressive language
-for testing our compiler in its nascent stages is then clear.  This language,
-mini-Haskell, was progressively "grown" from the STG syntax towards a subset of
+Haskell a pleasure to program in.  The motivation for a more expressive language
+for testing our compiler and runtime system in their nascent stages is then clear.  This language,
+mini-Haskell, was progressively ``grown'' from the STG syntax towards a subset of
 Haskell itself*, removing much of the frustrating syntactic cruft and adding
@@
 
-{\color{red}
-* Exceptions include the "unboxed" keyword allowed in datatype definitions and
-the ability to specify unboxed ints and doubles with a postfix "\#" character.
-The former feature is not in any extension of Haskell, while the latter can be
-included using the MagicHash extension.}
-
 An informal specification of the mini-Haskell grammar is given in
-{\color{red}numbered table?}.  There is significant overlap with STG, resulting
+Figure~\ref{miniHaskell}.  There is significant overlap with STG resulting
 from the structural similarities, but the redundancies are included for
-completeness. {\color{red} should they be?}
+completeness.
 
@@
 
-\begin{table}
+\begin{figure}
 \centering
+\footnotesize % tiny scriptsize footnotesize small
 \begin{tabular}{r r c l l}
@@
 
-Object Declaration & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$
+Object decl. & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$
 & Type signature \\
@@
 
-Constructor Definition & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$
+Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$
 & $i \ge 0$ \\
 
-Datatype Declaration &  $\mathit{ddecl}$ & ::= & \texttt{data} 
+Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} 
                                                  [\texttt{unboxed}]
@@
 Program                & $\mathit{prog}$ & ::= & $(o|d)\mathit{decl}_i$
-& Objects and datatypes, no order necessary
+& Objects and datatypes
 
 \end{tabular}
-\end{table}
+\label{miniHaskell}
+\caption{mini-Haskell syntax}
+\end{figure}
 
@@
 
-\paragraph{Removal of explicit heap objects} -\\
-
+\subsubsection{Removal of explicit heap objects.}
 In STG, all object definitions require explicit classification with
-\texttt{CON}, \texttt{PAP}, \texttt{THUNK}, \texttt{FUN}, or
-\texttt{BLACKHOLE}. This classification can be inferred from context by the
-compiler, so these keywords are no longer necessary.  Functions can then be
-defined in Haskell-style, with the parameters following the function's
-identifier on the left side of the equality. With these simple changes, the
-syntax becomes a subset of Haskell's*.
-
-{\color{red}
-* With only these changes, the semantics are a little different; numeric
-  literals would still be treated as unboxed values, for example.  I don't know
-  if that's worth noting though.}
+\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
+inferred from context by the compiler so these keywords are not necessary.
+Functions can then be defined in Haskell style with the parameters following
+the function's identifier on the left side of the equality. With these simple
+changes the syntax becomes a subset of Haskell.
 
@@
 
-\paragraph{The Layout Rule} -\\
+\subsubsection{Avoiding building thunks.}
+As \emph{EA} points out, there is no need to build a thunk for the scrutinee
+of a \texttt{case} expression because if the \texttt{case} is evaluated the
+scrutinee is certain to be evaluated.  Similarly, if a function is strict
+there is no need to build a thunk for each of its arguments (assuming boxed
+argument type).  Because STG allows only \emph{atoms} (variables or literal
+values) in argument positions, naive translation of Haskell \texttt{f e} to
+\texttt{let \{x = e\} in f x}, followed by naive code generation, will result in
+an unnecessary heap allocation.  An alternative is to translate to
+\texttt{case e of x \{\_ -> f x\}}.
 
+\subsubsection{The Layout Rule}
 Mini-Haskell implements Haskell's Layout Rule in an almost direct translation of
@@
 
-The inclusion of the layout rule may seem small and insignificant, but it allows
-its users to produce code that is visually similar to Haskell code.  This makes
-it easier to read and write for any Haskell programmer who relies on layout to
-delineate context (I assume this is ~99.9% of human Haskell users).  It also
-allows for the use of Haskell programming environments that support "smart"
-indentation.
+The inclusion of the layout rule may seem small and insignificant, but it
+allows its users to produce code that is visually similar to Haskell code.
+This makes it easier to read and write for any Haskell programmer who relies
+on layout to delineate context, presumably the vast majority.  It also allows
+for the use of Haskell programming environments that support ``smart''
+indentation, e.g., \emph{haskell-mode} for Emacs.
 
@@
 
-\paragraph{Expression to Expression application} -\\
+\subsubsection{Expression to Expression application}
 
-In STG, function application must be to atomic values, i.e. either literals or
+In STG, function application must be to atomic values, i.e., either literals or
 variables.  This often requires introducing bindings with either \texttt{let} or
 \texttt{case} expressions.  Consider the \texttt{append} function that
-concatenates its two list parameters.  In STG, it is written as:
+concatenates its two list parameters.  In STG, it is written as
 
@@
 
-This also serves as a good example of expression to expression application is
+This also serves as a good example of how expression to expression application is
 transformed into STG.  Any non-atomic expression in an application expression is
 bound to a heap object in a generated \texttt{let} expression.  These
-expressions are thus properly "atomized" for STG.  The above mini-Haskell code
+expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
 is transformed in STG that is structurally identical to the STG code that
@@
 
-\paragraph{Fancy Pattern Matching} -\\
+\subsubsection{Fancy Pattern Matching.}
+{\color{red}Implemented, not documented\dots}
 
-{\color{red}\dots}
-
-\paragraph{Less significant features} -\\
+\subsubsection{Less significant features}
 
@@
 
-\section{Current Status, and Expected Results by the time of the symposium}
+\section{Current Status, and Expected Results by the time of TiFP'16}
 
@@
 
-\section{Other Related Work}
+\section{Closely Related Work}
+
+Clearly various papers describing some of the inner workings of ghc at various point of its
+evolution have greatly informed our approach and these have all be previously cited.
 

### 2016-03-29 pubs/tfp16/paper.tex
some related work

@@
 I/O monad in Haskell) could be safely restarted.  The recent work of
-[Trinder's student] impressively demonstrates this for an augmented ghc/runtime
+[Trinder's student] impressively demonstrates this for an augmented GHC/runtime
 on a large-scale distributed-memory system.
@@
 implications.  First and foremost it means that \emph{we are not attempting to
-  compete with ghc in any respect}---succumbing to the temptation of
-recreating ghc language, type system, and runtime features, or significant
+  compete with GHC in any respect}---succumbing to the temptation of
+recreating GHC language, type system, and runtime features, or significant
 code optimization, would simply divert us from our central goals.\footnote{We
   estimate that we will have, to within an order of magnitude, one person-hour
-  of total available development time for each person-year that ghc currently
+  of total available development time for each person-year that GHC currently
   represents, or put another way, one second for each hour, respectively.}
@@
 
-ghc 8.0 gives three possibilities: conventional non-strict, constructor
+GHC 8.0 gives three possibilities: conventional non-strict, constructor
 strict, and function- and constructor strict, but there are possible
@@
 
-Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
-is part of our master plan.  In short, our plan is to use ghc as our front-end
-to generate its intermediate form, STG~\cite{PJ-stockhardware}, then escape to
-our system from there.  There are various possible ways to accomplish the
-former: using ghc's option to dump STG as text, hacking ghc itself and,
-optionally, using ghc as a library.
+Our wish list has much in common with the GHC Haskell implementation, and
+indeed GHC is part of our master plan.  In brief, the phases of the GHC
+compiler are 1) parsing and type checking; 2) transformation to the GHC
+\emph{Core} intermediate representation; 3) analysis and transformation on
+Core; 4) transformation of Core to the \emph{STG} intermediate representation;
+5) transformation of STG to machine code via one of various imperative
+representations (Cmm, LLVM, C).
+
+In short, our plan is to use GHC as our front-end to generate its intermediate
+form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
+There are various possible ways to accomplish the former: using GHC's option
+to dump STG as text, hacking GHC itself and, optionally, using GHC as a
+library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core.
 
@@
 
-Our incarnation of STG, as is our presentation in Figure~\ref{STGsyntax} which
+Our incarnation of STG, as is our presentation in Figure~\ref{fig:STGsyntax} which
 is meant to evoke both an abstract and concrete syntax, are very similar to
@@
 \caption{STG syntax}
-\label{STGsyntax}
+\label{fig:STGsyntax}
 \end{figure}
@@
 
-In STG---one of ghc's
+In STG---one of GHC's
 intermediate representations---the distinction is clear: heap objects are
@@
 While performance is not a high priority, we attain proper tail calling instead of
-using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
+using the trampoline used in earlier versions of GHC~\cite{PJ-stockhardware}.
 Because we maintain our own data and control (continuation) stack all
@@
 \subsection{Mini-Haskell}
-As mentioned, a current near-term goal is to use ghc as our front-end to
+As mentioned, a current near-term goal is to use GHC as our front-end to
 generate STG as input to our system.  In the meantime, as a stop-gap measure
@@
 compiler, STG is relatively verbose, requiring explicit block and expression
-delimiters, and classification of heap objects. Though its general structure is
-similar to primitive Haskell, it lacks many of the conveniences that make
-Haskell a pleasure to program in.  The motivation for a more expressive language
-for testing our compiler and runtime system in their nascent stages is then clear.  This language,
-mini-Haskell, was progressively ``grown'' from the STG syntax towards a subset of
-Haskell itself*, removing much of the frustrating syntactic cruft and adding
-some of the core features that are often taken for granted in Haskell (and other
-languages).
+delimiters, and classification of heap objects. Though its general structure
+is similar to primitive Haskell, it lacks many of the conveniences that make
+Haskell a pleasure to program in.  The motivation for a more expressive
+language for testing our compiler and runtime system in their nascent stages
+is then clear.  This language, mini-Haskell, was progressively ``grown'' from
+the STG syntax towards a subset of Haskell itself*, removing much of the
+frustrating syntactic cruft and adding some of the core features that are
+often taken for granted in Haskell (and other languages).
 
 An informal specification of the mini-Haskell grammar is given in
-Figure~\ref{miniHaskell}.  There is significant overlap with STG resulting
+Figure~\ref{fig:miniHaskell}.  There is significant overlap with STG resulting
 from the structural similarities, but the redundancies are included for
@@
 \end{tabular}
-\label{miniHaskell}
+\label{fig:miniHaskell}
 \caption{mini-Haskell syntax}
@@
 \subsubsection{The Layout Rule}
-Mini-Haskell implements Haskell's Layout Rule in an almost direct translation of
-the specification given in section 10.3 of the Haskell 2010 Report.  The notable
-exception is the omission of any attempt to define the \texttt{parse-error}
-function.  This is apparently a tricky spot for Haskell compilers (even GHC has
-not always had it right) so, in the name of tradition, it remains a tricky spot
-for our mini-Haskell front end.  The edge cases that it will incorrectly handle
-as a result of this are exactly that: edge cases.  As such, they are easy to
-avoid in practical usage.
+Mini-Haskell implements Haskell's layout rule in an almost direct translation
+of the specification given in section 10.3 of the Haskell 2010
+Report.~\cite{haskell2010report} The notable exception is the omission of any
+attempt to define the \texttt{parse-error} function.  This is apparently a
+tricky spot for Haskell compilers (even GHC has not always had it right) so,
+in the name of tradition, it remains a tricky spot for our mini-Haskell front
+end.  The edge cases that it will incorrectly handle as a result of this are
+exactly that: edge cases.  As such, they are easy to avoid in practical usage.
 
 The inclusion of the layout rule may seem small and insignificant, but it
-allows its users to produce code that is visually similar to Haskell code.
+allows users to produce code that is visually similar to Haskell code.
 This makes it easier to read and write for any Haskell programmer who relies
@@
 for the use of Haskell programming environments that support ``smart''
-indentation, e.g., \emph{haskell-mode} for Emacs.
+indentation, e.g., \emph{haskell-mode} for Emacs.~\cite{haskellmode}.
 
@@
 
-No bridge from ghc STG to our STG yet.
+No bridge from GHC STG to our STG yet.
 
@@
 \item type inferencer
+\item various analysis and transformation passes
 \item code generator
@@
 
-We have implemented a simple copying garbage collector \cite{Cheney:1970} with
+We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
 small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
@@
 
-We perform standard Hindley-Milner type inference and enforcement, augmented
-with built-in and user-defined unboxed types as described by Peyton-Jones and
-Launchbury's seminal work~\cite{Jones91unboxedvalues}, but for unboxed types code
-generation is currently limited to built-in single-word (64-bit) or smaller
-types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
-and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
-  True\#}.
+We perform standard Hindley-Milner type inference and
+enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
+types as described by Peyton-Jones and Launchbury~\cite{Jones91unboxedvalues},
+but for unboxed types code generation and runtime support is currently limited
+to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
+\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
+enumerations, e.g., \texttt{data unboxed Bool\# = False\# | True\#}.
 
@@
 
-We have developed a extensive test suite which uses the CMake \cite{cmake-book}
-CTest tool. The full test suite is automatically run, and any failures reported,
-after every comment to the version control system. We use a Jenkins continuous
-integration server \cite{jenkins-book} to automate this build and testing process.
-The (git) version control repository is hosted externally so our student
-collaborators can contribute at their leisure.
+We have developed a extensive test suite which uses the CMake CTest
+tool.~\cite{cmake-book} The full test suite is automatically run, and any
+failures reported, after every commit to the Git revision control system. We
+use a Jenkins continuous integration server to automate this build and testing
+process.~\cite{jenkins-book} The Git repository is hosted externally so our
+student collaborators can contribute at their leisure.
 
@@
 testing is done for a specific program. Given our goal of exploring various
-levels of strictness, most tests are run with all evaluation strategies (3
+levels of strictness, most tests are run with all evaluation strategies (three
 degrees of function strictness by two degrees of constructor strictness).
-One interesting class of test programs are those which require non-strict
+One interesting class of test programs are those which require lazy
 semantics. For these programs we test not only that the test leads to the expected
 result with non-strict semantics, but also that the program returns a
-"blackhole" in the case of strict semantics. Students who were Haskell/STG
+\emph{BLACKHOLE} in the case of strict semantics. Students who were Haskell/STG
 beginners have contributed substantially to the test suite. We have also found
@@
 a benefit that we did not initially envision. Over time we have developed
-extensive STG and mini-Haskell preludes of common programming patterns, to
+extensive STG and mini-Haskell preludes of common programming patterns to
 aid in the writing of test programs.
@@
 
-We also run each test with various levels of garbage collection enabled, this
+We also run each test with various levels of garbage collection enabled. This
 has helped us to find bugs not only in the garbage collector, but also in the
-code generator.
+code generator.  Interestingly, in practice garbage collection can both mask
+and expose bugs.
 
 The various enumerations of strictness and garbage collection levels mean that
-we run a quite large number of tests. However, this is fully automated via the
-testing and continuous integration infrastructure.
+we run a quite large number of tests (currently order thousands). However,
+this is fully automated via the testing and continuous integration
+infrastructure.
 
@@
 
-\section{Closely Related Work}
+\section{Related Work}
 
-Clearly various papers describing some of the inner workings of ghc at various point of its
-evolution have greatly informed our approach and these have all be previously cited.
+Various papers describing some of the inner workings of GHC at various point
+of its evolution have greatly informed our approach and these have all been
+previously cited.
 
-In his BSc thesis, Don Stewart described an implementation wherein ghc was used as a Haskell
-compiler front-end, dumping STG code as text, parsing that text, then translating to code for
-execution on the Java Virtual Machine.~\cite{Stewart-BSc}.
+\subsubsection{\emph{Strict but still pure} is not new.}
+Very simple implementations have been prototyped~\cite{explicit-lazy}.  
+%
+A plug-in for GHC (prior to 8.0) exists for making functions, but not data
+structures, strict.~\cite{strict-ghc-plugin}.
+%
+The Disciplined Disciple Compiler is a full-blown implementation that has been
+underway for some time, but the emphasis appears to be on a much more complex
+type system for handling stateful computation, and the prospects for its
+eventual completion are unknown~\cite{disciple}.
+%
+NVIDIA's NOVA language, and corresponding compiler, implements a strict
+(``call by value''), pure, polymorphic, higher-order functional language with
+built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
+\emph{scan}).~\cite{NVIDIA:2013} Three back-ends were implemented: for
+sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
+targeting NVIDIA GPUs.
 
-Nvidia research compiler.
+\subsubsection{Non-strictness as antithetical to parallel evaluation.}  While
+compile-time analysis can conservatively estimate when it is safe to strictify
+and parallelize, in general the problem is undecidable and the compiler must
+default to purely data-dependency-driven evaluation on demand.  This has been
+shown to be problematic for three reasons~\cite{Mattson94}.  First, it is
+often desired to force parallel evaluation beyond simply to WHNF, requiring
+elaborate strictification as previously discussed.  Second is a space
+management and runtime issue: because non-strict evaluation is generally
+space-intensive, the runtime may be forced to prematurely abort parallel
+evaluation when garbage collection is triggered by space constraints.  Third,
+when aborting a speculative evaluation, to be able to reclaim space sufficient
+bookkeeping needs to be done to be able to roll back the reduction graph.  The
+conclusion was that it is hard to amortize the runtime overhead over gains in
+parallelism, and thus justify either the runtime or application implementation
+efforts in the non-strict context.  In the strict context the runtime can
+simply parallelize until resources are saturated, and much less bookkeeping is
+required to recover space when speculative evaluations are to be terminated.
 
-\section{Summary, Future Work, Conclusions?}
+\subsubsection{GHC as a front-end.}
+The Intel Labs Haskell Research Compiler (HRC) is the most closely related
+work to ours of which we are aware.~\cite{Liu:2013,Petersen:2013} As in our
+(planned) approach they use GHC as the front end, but exit the GHC toolchain
+at the final GHC Core stage, just before transformation to STG.  The essence
+of their effort is to target an existing, highly optimizing, Intel functional
+language compiler that is ``largely language agnostic'' and performs (among
+many other optimization) SIMD vectorization.
+
+In his BSc thesis, Don Stewart described an implementation wherein GHC was
+used as a Haskell compiler front-end, dumping STG code as text, parsing that
+text, then translating to code for execution on the Java Virtual
+Machine.~\cite{Stewart-BSc}.
+
+\subsubsection{Fault tolerance for functional programming.}
+Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
+distributed Haskell based on distributed exception
+handling.~\cite{Pointon-GdH-00}.  However, it appears that their scheme for
+handling failed processing elements correctly~\cite{} may have never been
+implemented.
+
+Recently, Robert Stewart extended Haskell Distributed Parallel Haskell
+(HdpH)~\cite{hdph} to be fault resilient at an impressive scale---1400 cores
+on a distributed memory system.  Here the key recovery strategy is task
+replication.~\cite{Stewart:2013}.
+
+Trinder--speculative parallelism
+
+Trinder's paper (or just web site?) on fault tolerance.
+
+Trinder's student's thesis on fault tolerance.
+
+\section{Future Work}
+
+An almost unbounded amount of interesting work suggests itself; one such path
+would lead to the recreation of GHC.  However, given the miniscule budget (and
+thus the emphasis on student labor) we must maintain sharp focus on the main
+goals:  implementing and evaluating mechanisms for automatic parallelism and
+fault tolerance.
+
+Currently we are developing the parallel runtime.  This summer (2016) a
+returning student will (partially) bridge the gap between GHC STG and our STG,
+and augment our runtime as needed in this effort.  Just as in the Intel Haskell
+Research Compiler effort, we will not
 
@@
 
-Funding for this project and its participants has been provided by the DOE
-NNSA LANL Laboratory Directed Research and Development program award
-20150845ER; the National Science Foundation Science, Technology, Engineering,
-and Mathematics Talent Expansion Program (STEP) program for undergraduate
-students; and the DOE Science Undergraduate Laboratory Internship (SULI) program.
+Funding for this project has been provided by the DOE NNSA LANL Laboratory
+Directed Research and Development program award 20150845ER; the National
+Science Foundation Science, Technology, Engineering, and Mathematics Talent
+Expansion Program (STEP) program for undergraduate students; and the DOE
+Science Undergraduate Laboratory Internship (SULI) program.
 %
@@
 Department of Energy’s National Nuclear Security Administration (NNSA).
-
-% just get these into the bib
-\cite{ghc-as-library}
-\cite{Heren02}
 

### 2016-03-31 pubs/tfp16/paper.tex
some edits and thoughts

@@
     runtime.  We expect to have preliminary results from our new parallel
-    implementation by the time of the workshop.  The fault-tolerance mechanisms
+    implementation by the time of the symposium.  The fault-tolerance mechanisms
     remain future work.
@@
 
-  \item Hopefully in the spirit of a student-friendly workshop, for this draft
-    version we have chosen a more informal, conversational style of
+  \item Hopefully in the spirit of a student-friendly symposium, for this
+    draft version we have chosen a more informal, conversational style of
     presentation than necessarily appropriate for a final paper, have
-    highlighted the major student contributions and involvement, and included
-    perhaps gratuitous detail about process.
+    highlighted student involvement and its influence on our approach, and
+    included perhaps gratuitous detail about process.
   \end{enumerate}
@@
 
-\subsection{The trend: strict but still pure functional programming}
-There is a small but growing understanding that \emph{pure functional
-  semantics}, in some form, is essential to reining in the complexity of these
-ever-evolving scientific codes.  Such semantics are expressed in a number
-of ways as highlighted following.
+\subsection{The trend: strict but still pure functional programming in HPC}
+Interest in functional programming for HPC in general is not new, as evidenced
+by the Workshop on Functional High Performance Computing, held in conjuction
+with the International Conference on Functional Programming, in its fifth
+consecutive year.  A more precise characterization of the trend is that there
+is a small but growing understanding by practicing computational scientists
+that \emph{pure functional semantics}, in some form, is essential to reining
+in the complexity of these ever-evolving scientific codes and informing the
+designs of new ones.  Such semantics are expressed in a number of ways as
+highlighted following.
 
@@
 Otherwise it is entirely up to the programmer to make functions composable and
-thread safe.  Here at LANL this approach is being used to good effect to clean
+thread safe.  Here at LANL the use of pure functions (among other techniques
+for reducing global data dependencies) is being used to good effect to clean
 up some legacy code bases.
@@
 There are a number of parallel runtime systems that encourage (but cannot
-enforce) a pure functional style.  We will highlight Stanford's Legion
+enforce) a pure functional style.  We highlight Stanford's Legion
 runtime as an example~\cite{Bauer12}.
 
-In Legion, non-function-local data, or \emph{regions}, are requested from the
-runtime system.  Regions may be shared among \emph{tasks} (essentially C++
-functions), and each task is prescribed access privileges (read, write,
-read/write, etc.)  and coherency requirements for each region to which it will
-have access.  Other than registered access to regions, tasks must not access
-non-constant global data.  In serial execution these access requirements are
-notionally superfluous, but in parallel execution the runtime can dynamically
-calculate the data dependency graph and relax the programmatically defined
-serial ordering of task execution.  Very high parallel efficiency has been
-demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
-LANL the Legion system is being strongly considered for a new large-scale C++
-code effort.\footnote{It seems significant that the PI of the Legion
-  project was previously a co-developer of a strict, arguably pure functional
-  language implementation~\cite{AikenFL,FLreport89}, in collaboration with
-John Backus as a refinement of his FP language~\cite{Backus:1978}.}
+Using Legion, all non-function-local variables, or \emph{regions}, must
+requested from the runtime system.  Regions may be shared among \emph{tasks}
+(essentially C++ functions), and each task is prescribed access privileges
+(read, write, read/write, etc.) and coherency requirements for each region to
+which it will have access.  Other than registered access to regions, tasks
+must not access non-constant global data.  In serial execution these access
+requirements are notionally superfluous, but in parallel execution the runtime
+can dynamically calculate the data dependency graph and relax the
+programmatically defined serial ordering of task execution.  Very high
+parallel efficiency has been demonstrated on extremely large computing
+systems~\cite{Bauer14}.  Here at LANL the Legion system is being strongly
+considered for a new large-scale C++ code effort.\footnote{It seems
+  significant that the PI of the Legion project was previously a co-developer
+  of a strict, arguably pure functional language
+  implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
+  as a refinement of his FP language~\cite{Backus:1978}.}
 
@@
 nuisance--even a serious hindrance--than a convenience, much less ever
-algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84}
-or Johnsson's general attribute grammars~\cite{Johnsson87}.
+algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
+Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
+of \emph{tying the knot.}~\cite{tying-the-knot}.
 
@@
 Much anecdotal evidence suggests that many others have similar difficulties
-with lazy (or non-strict) by default semantics, with the most telling recent
-concrete evidence being the push for, and realization of, the \texttt{Strict}
-and \texttt{StrictData} language extensions newly available in GHC 8.0.
-Quoting the documentation:
+with semantics that are lazy (or non-strict) by default, with the most telling
+recent concrete evidence being the push for, and realization of, the
+\texttt{Strict} and \texttt{StrictData} language extensions newly available in
+GHC 8.0.  Quoting the documentation:
 \begin{quote}
@@
 \end{quote}
-It is a fact that \emph{HPC practioners are by and large not computer scientists}.
-The concept of a space leak, for example, caused by other than a missing explicit
-deallocation statement (C \emph{free}, etc.) is difficult to understand and not
-something they particularly want to try to understand.  Hence, we eschew explicit
-strictness annotations for varying uniform degrees of strictness.
+It is a fact that \emph{HPC practioners are by and large not computer
+  scientists}, they are domain scientists (physicists, chemists, etc.) or
+computational scientists specialized in numerical or numerically-oriented
+algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
+The concept of a space leak, for example, caused by other than a missing
+explicit deallocation statement (C \emph{free}, etc.) is difficult to
+understand and not something they particularly want to try to understand.
+Hence, we eschew explicit strictness annotations for varying uniform degrees
+of strictness.
 
@@
 (MPI)~\cite{MPI}.  In the imperative world, example systems include the
-aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007}, and the
-HPX C++11/14 extension~\cite{HPX}.  While the concept of task is somewhat
-vague, the general idea is a unit of computation with no or
+aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
+and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
+somewhat vague, the general idea is a unit of computation with no or
 minimal/constrained side effects.  Function call in a pure functional language
@@
 various point in time sufficient program state is dumped to backing store such
-that the program, in case of a fault, can restart from the latest stored
-state.  Most implementations are explicit, though there has been some success
-with transparent (invisible to the programmer) systems.  Wiring
+that the program, in case of a fault, can be (manually) restarted from the
+latest stored state.  Most implementations are explicit, though there has been
+some success with transparent (invisible to the programmer) systems.  Wiring
 checkpoint/restart into a large-scale application is not only burdensome and
@@
 mutable global state.  The Legion system is being augmented with this
-capability, and in such a way that it will be largely transparent to the programmer.
-Ericsson's quasi-functional language Erlang and its runtime system was
-designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
-recognized that the pure parts of a functional program (excluding, e.g., the
-I/O monad in Haskell) could be safely restarted.  The recent work of
-[Trinder's student] impressively demonstrates this for an augmented GHC/runtime
-on a large-scale distributed-memory system.
+capability, and in such a way that it will be largely transparent to the
+programmer.  Ericsson's quasi-functional language Erlang and its runtime
+system was designed for fault tolerance.~\cite{Cesarini:2009}.  It has long
+been recognized that the pure parts of a functional program (excluding, e.g.,
+the I/O monad in Haskell) could be safely restarted.  The recent work of
+Steward impressively demonstrates this on a large-scale distributed-memory
+system~\cite{Stewart:2013}.
 
@@
 (e.g.\ strict Haskell).  Second, we claim that there is a trend in
-high-performance computing, and by transitivity functional programming,
+high-performance computing, and by transitivity in functional programming,
 towards transparent fault tolerance.
@@
 \subsection{Strictness and Parallelism}
-
 It is a trivial observation that given strict functional semantics, wherein
@@
 \section{Project and Goals}
-
 Our project is the light-weight implementation of a pure, higher-order,
@@
   represents, or put another way, one second for each hour, respectively.}
+For another, it means that we are not undertaking significant morphing of GHC
+itself:  because this is in large part a student project it must be
+kept as simple as reasonably possible in various respects.
 
@@
   should be preserved for all degrees of strictness;
-\item Tail calling should be properly implemented, i.e., not as a trampoline;
+\item Tail calling should be properly implemented, i.e., not as a trampoline.
 \end{itemize}
@@
 \subsection{Varying degrees of strictness}
-
 First some words about terminology.  For all degrees of strictness, sharing
@@
 \section{Design and Implementation}
-
 Our wish list has much in common with the GHC Haskell implementation, and
@@
 form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
-There are various possible ways to accomplish the former: using GHC's option
+There are various possible ways to accomplish the this: using GHC's option
 to dump STG as text, hacking GHC itself and, optionally, using GHC as a
-library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core.
+library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core
+which we may transition to.
 
 \subsection{STG Language}
+
+Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
+evoke both an abstract and concrete syntax.  This presentation is adapted from
+Marlow and Peyton Jones' \emph{Eval/Apply} paper~\cite{Marlow:2004}, to which
+we refer so many times that we will subsequently refer to it as simply
+\emph{EA}. Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}),
+i.e., a thunk being evaluated, are only runtime entities and not actually part
+of our STG language.
+
 \setlength{\tabcolsep}{5pt}
-
-Our incarnation of STG, as is our presentation in Figure~\ref{fig:STGsyntax} which
-is meant to evoke both an abstract and concrete syntax, are very similar to
-that given in Marlow and Peyton Jones' \emph{fast curry}
-paper~\cite{Marlow:2004}, to which we refer so many times that we will
-subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
-  the title because \emph{FC} likely has another connotation to the reader.}
-
 \begin{figure}
@@
 
-Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}), i.e., a thunk being
-evaluated, are only runtime entities and not actually part of our STG language.
-
 The semantics of STG is what one would expect for a higher-order, polymorphic,
-pure functional language, with one exception for a Haskell programmer familiar
-with unboxed types:  \texttt{case} is strict in the scrutinee, so
-\texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell,
-that is, the scrutinee is evaluated first.
-
-In STG---one of GHC's
-intermediate representations---the distinction is clear: heap objects are
-explicitly constructed by \texttt{let} constructs in which constructor names
-are given explicitly (not aliased); constructor names are not functions.
-
-A distinctive feature of STG is that heap allocation is explicit; this is exactly
-what \texttt{let} bindings do.  As such unboxed values cannot be let-bound; instead
-unboxed values can be bound with \texttt{case}:
-\begin{verbatim}
-    case e of x# {
-      _ -> e
-    }
-\end{verbatim}
-where the \texttt{\#} suffix is merely a convention to signify an unboxed type.
-Deferred evaluation is made explicit:  a deferred expression evaluation is
-encoded as a thunk.
+pure functional language, with a few exceptions for a Haskell programmer
+familiar with unboxed types.  First, \texttt{case} is strict in the scrutinee,
+so \texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in
+Haskell, that is, the scrutinee is evaluated first.  Second, constructors are
+not functions, rather they are named in the construction (heap allocation) of
+a data object.  More generally, the operational notion of allocating a heap
+object is made explicit: this is the ``side effect'' of \texttt{let} bindings.
+As such a variable of unboxed type cannot be let-bound; instead it can be
+bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
+%\begin{verbatim}
+%    case e of x# {
+%      _ -> e
+%    }
+%\end{verbatim}
+where the \texttt{\#} suffix is merely a convention to signify that
+\texttt{x\#} has unboxed type.  Finally, deferred evaluation is made explicit:
+a deferred expression evaluation is encoded by a thunk.
 
 \subsection{Degrees of strictness, take two}
-
 Other possible variations concern the notional arity of the underlying
 function being applied.  This is readily explained in terms of the operation
-of the STG.  Our partial description is based on that given in \emph{EA}.
+of the STG abstract machine.  Our partial description is based on that given
+in \emph{EA}.
 
@@
 the new \emph{PAP} returned.  If the arity is equal to two then the underlying
-\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
-and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
-(it must be at least one, so one in this case) the underlying \emph{FUN} is
-called-with-return with any previous arguments (in case of \emph{PAP}) and new
-argument \texttt{x}, then the object returned applied to remaining argument
-\texttt{y}.
+\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
+\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
+than two (it must be at least one, so one in this case) the underlying
+\emph{FUN} is called-with-return with any previous arguments (in case of
+\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
+remaining argument \texttt{y}.
 
@@
 \texttt{f} has been evaluated, the underlying \emph{FUN} actually called
-(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one of
-those arguments forced in subsequent evaluation (assuming that they weren't
-forced in the evaluation of \texttt{f} itself).
+(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
+both of those arguments forced in subsequent evaluation (assuming that they
+weren't forced in the evaluation of \texttt{f} itself).
 
-In a strict semantics one can come up with various evaluation orders, but
-excluding the absurd, as far as distinguishing termination properties goes
-these can be distilled down to two possibilities, namely whether on
-construction of a \emph{PAP} the (new) arguments are forced or not.  (An
-absurd strategy might force every second argument.)  These are in fact
+In a strict semantics one could devise various evaluation orders, but as far
+as distinguishing termination properties these can be distilled down to two
+reasonable possibilities, namely whether on construction of a \emph{PAP} the
+(new) arguments are forced or not.  In other words, the question is whether
+application is strict, or actual function call is strict.  These are in fact
 distinguishable because (in Haskell) the first argument to \texttt{seq} might
@@
 
+The other dimension we vary is whether or not constructors are strict.
+Other dimensions are possible, for example whether \texttt{let} bindings
+(heap allocations) are constructed on demand, but we restrict ourselves
+to function/application and constructor strictness.
+
 \subsection{STG to C}
-
 The back-end is written in Haskell and generates proper tail-calling C code;
@@
 \subsubsection{Proper tail calls.}
-While performance is not a high priority, we attain proper tail calling instead of
-using the trampoline used in earlier versions of GHC~\cite{PJ-stockhardware}.
-Because we maintain our own data and control (continuation) stack all
-generated C functions take no explicit arguments and have return type void,
-and with a single exception in the non-strict case are invoked in tail call
-positions, at non-zero optimization levels both gcc and Clang/LLVM generate
-jumps instead of calls for C calls in tail call positions.  When using gcc
-we also use a fixed set of machine registers for passing/returning values.
+While performance is not a high priority, we attain proper tail calling
+instead of using the trampoline used in earlier versions of
+GHC~\cite{PJ-stockhardware}.  Because we maintain our own data and control
+(continuation) stack all generated C functions take no explicit arguments and
+have return type \emph{void}, and with a single exception in the non-strict
+case are invoked in tail call positions, at non-zero optimization levels both
+gcc and Clang/LLVM generate jumps instead of calls for C calls.  When using
+gcc we also use a fixed set of machine registers for passing/returning values.
 
@@
 that is used for applications of non-known functions, and known functions with
-a deficit or excess number of arguments.  The \emph{stgApply} function takes
+a deficit or excess number of the arguments.  The \emph{stgApply} function takes
 as arguments the first element of the application (\emph{FUN}, \emph{PAP},
@@
 for the benefit of the garbage collector, this elaboration is unnecessary.
+For \emph{direct calls}, in which a known function is applied to a number
+of arguments equal to its arity, \emph{stgApply} need not be invoked:
+the \emph{FUN} is tail-called directly.
 
-For \emph{stgApply} in the non-strict semantics an alternative to
-call-with-return would be a pair of functions \emph{stgApply1} and
-\emph{stgApply2}.  In one possible variation, \emph{stgApply1} would push a
-continuation with \emph{stgApply2} as the return address, then invoke the
-first element of the application.  Call-with-return becomes more convenient
-when considering the various forms of strictness: when evaluating function or
-constructor arguments it is simpler, and arguably more efficient in terms of
-time (but not space in general) to ``evaluate in place'' than implement the
-continuation style to the logical limit.
+Call-with-return becomes even more convenient when considering the various
+forms of strictness: when evaluating function or constructor arguments it is
+simpler, and arguably more efficient in terms of time (but not space in
+general beca) to ``evaluate in place'' than implement the continuation style
+to the logical limit.  We are aware, however, that we may need to eliminate
+this use of the native C stack so that we can properly unwind the stack in a
+more sophisticated parallel runtime when pruning multiple evaluations of the
+same thunk, and we have determined that this is a relatively minor exercise:
+one possibility is that \emph{stgApply} takes an additional argument
+indicating which of its subsequent arguments have already been forced, and at
+each invocation if unforced arguments remain, evaluates the next (possibly)
+unevaluated argument and tail calls itself, re-using the stack frame with just
+this new argument updated.
+
+\emph{It seems that we can dispose with call-with-return of STG functions,
+  including stgApply (as opposed to runtime functions) without too much
+  effort.  It also seems like this is needed if we want to be able to unwind
+  the stack as described in the ``grey hole'' paper.  I'll try to do this this
+  week (3/31/16-4/1/16) to verify the feasibility before changing the
+  preceding description (which isn't quite right)}
 
@@
 As mentioned, a current near-term goal is to use GHC as our front-end to
-generate STG as input to our system.  In the meantime, as a stop-gap measure
-to make the generation of test cases easier, we have implemented a
+generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
+measure to make the generation of test cases easier, we have implemented a
 \emph{mini-Haskell} front-end that supports a subset of standard Haskell
 syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
-\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
-unboxed literal value.  Our extension (to STG as well)
-is the provision of the optional \texttt{unboxed} keyword in datatype definitions
-to allow user-defined unboxed data types, e.g.,
+\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
+an unboxed literal value.  Our extension (to STG as well) is the provision of
+the optional \texttt{unboxed} keyword in datatype definitions to allow
+user-defined unboxed data types, e.g.,
 \begin{verbatim}
@@
 is then clear.  This language, mini-Haskell, was progressively ``grown'' from
-the STG syntax towards a subset of Haskell itself*, removing much of the
+the STG syntax towards a subset of Haskell itself, removing much of the
 frustrating syntactic cruft and adding some of the core features that are
@@
 \begin{itemize}
-\item STG (inc. ADT w/unboxed) parser
+\item STG and mini-Haskell (inc. ADT w/unboxed) parsers \emph{(both by students)}
+\item mini-Haskell to STG transformer \emph{(by student)}
 \item type inferencer
-\item various analysis and transformation passes
-\item code generator
+\item various analysis and transformation passes \emph{(all by students)}
+\item code generator \emph{(some student contribution)}
 \item runtime system including a garbage collector
+\item extensive test suite \emph{(mostly by students)}
 \end{itemize}
@@
 \subsubsection{\emph{Strict but still pure} is not new.}
-Very simple implementations have been prototyped~\cite{explicit-lazy}.  
+Very simple implementations have been prototyped, for example
+Sheard's.~\cite{Sheard:2003}
 %
 A plug-in for GHC (prior to 8.0) exists for making functions, but not data
-structures, strict.~\cite{strict-ghc-plugin}.
+structures, strict.~\cite{strict-ghc-plugin}.  The transformation is on GHC
+Core, recursively expand non-recursive let bindings into case bindings.  Thus
+it seems that while it makes functions denotationally strict, function arguments
+are still evaluated after function call.
 %
-The Disciplined Disciple Compiler is a full-blown implementation that has been
-underway for some time, but the emphasis appears to be on a much more complex
-type system for handling stateful computation, and the prospects for its
-eventual completion are unknown~\cite{disciple}.
+The Disciplined Disciple Compiler project appears to intend implementation of
+a strict, higher order functional language (or family of languages) compiler
+that has been underway for some time, but the emphasis appears to be on a much
+more complex type system for handling stateful computation, and the prospects
+for its eventual completion are unknown~\cite{disciple}.
 %
@@
 targeting NVIDIA GPUs.
-
-\subsubsection{Non-strictness as antithetical to parallel evaluation.}  While
-compile-time analysis can conservatively estimate when it is safe to strictify
-and parallelize, in general the problem is undecidable and the compiler must
-default to purely data-dependency-driven evaluation on demand.  This has been
-shown to be problematic for three reasons~\cite{Mattson94}.  First, it is
-often desired to force parallel evaluation beyond simply to WHNF, requiring
-elaborate strictification as previously discussed.  Second is a space
-management and runtime issue: because non-strict evaluation is generally
-space-intensive, the runtime may be forced to prematurely abort parallel
-evaluation when garbage collection is triggered by space constraints.  Third,
-when aborting a speculative evaluation, to be able to reclaim space sufficient
-bookkeeping needs to be done to be able to roll back the reduction graph.  The
-conclusion was that it is hard to amortize the runtime overhead over gains in
-parallelism, and thus justify either the runtime or application implementation
-efforts in the non-strict context.  In the strict context the runtime can
-simply parallelize until resources are saturated, and much less bookkeeping is
-required to recover space when speculative evaluations are to be terminated.
 
@@
 distributed Haskell based on distributed exception
-handling.~\cite{Pointon-GdH-00}.  However, it appears that their scheme for
+handling.~\cite{Pointon:2001}.  However, it appears that their scheme for
 handling failed processing elements correctly~\cite{} may have never been
@@
 replication.~\cite{Stewart:2013}.
-
-Trinder--speculative parallelism
-
-Trinder's paper (or just web site?) on fault tolerance.
-
-Trinder's student's thesis on fault tolerance.
 

### 2016-03-31 pubs/tfp16/paper.tex
do emacs save first

@@
 more complex type system for handling stateful computation, and the prospects
-for its eventual completion are unknown~\cite{disciple}.
+for its eventual completion are unknown~\cite{disciplined-disciple}.
 %

### 2016-04-01 pubs/tfp16/paper.tex
random ramblings

@@
   should be preserved for all degrees of strictness;
-\item Tail calling should be properly implemented, i.e., not as a trampoline.
+\item Tail calling must be properly and consistently implemented;
+\item The generated code is C.
 \end{itemize}
+
+The requirement for tail calling is not primarily for efficiency, rather it
+is for a reasonable parallel implementation.  Achieving this, and thereby
+sufficiently limiting the use of the native C stack, makes this an interesting
+exercise that we discuss at some length.
 
@@
 \subsection{STG to C}
-The back-end is written in Haskell and generates proper tail-calling C code;
+The back-end is written in Haskell and generates C code;
 the runtime system is written in C/C++.  The system fairly faithfully
@@
 \subsubsection{Proper tail calls.}
-While performance is not a high priority, we attain proper tail calling
-instead of using the trampoline used in earlier versions of
-GHC~\cite{PJ-stockhardware}.  Because we maintain our own data and control
-(continuation) stack all generated C functions take no explicit arguments and
-have return type \emph{void}, and with a single exception in the non-strict
-case are invoked in tail call positions, at non-zero optimization levels both
-gcc and Clang/LLVM generate jumps instead of calls for C calls.  When using
-gcc we also use a fixed set of machine registers for passing/returning values.
+Parallelism is achieved by multi-threading; our model is based on Harris et
+al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
+aspirations~\cite{Marlow:2009}.  To achieve efficient execution they
+identified the need for the ability to examine and unwind the stack.  This
+requires that use of the C stack be highly constrained if we are to 
+avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
+non-portable code to directly manipulate the C stack, complicating an
+already non-trivial implementation effort.  We therefore must limit the
+use of the C stack to calls to the runtime, in other words, when executing
+code generated from the user program (and not in a runtime call), there
+must be no growing of (pushing return addresses or data onto) the C stack.
 
-The exceptions to tail calling are for convenience.  Similar to what is
-described in \emph{EA} we have a generic variadic \emph{stgApply} function
-that is used for applications of non-known functions, and known functions with
-a deficit or excess number of the arguments.  The \emph{stgApply} function takes
+First we establish that this is possible.  By maintaining our own data
+and control (continuation) stack all generated C functions can take no
+explicit arguments and have return type \emph{void}. If we arrange that
+generated C code has only calls (other than runtime calls, which do not call
+back to user-program code) in tail call positions, that is, immediately before
+a return in the control flow graph, then at non-zero optimization levels both
+gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When using
+gcc we can also use a fixed set of machine registers for passing/returning
+values, much like GHC.
+
+\subsubsection{A taste of the continuation style.}
+Now to implementation, which we discuss for a simple case, then the most complex
+case.  To clearly distinguish between a notional C call (a syntactic concept)
+and a machine-level call (pushing a return address onto the native C stack) we
+will refer to the latter as \emph{call-with-return}, and to a machine-level
+tail call as \emph{jump}.
+
+First a detail.  Our implementation of the STG abstract machine has an
+abstract register called the \emph{current value register} or \emph{CVR}.
+Roughly speaking, it can hold the pointer to what is about to be evaluated or
+what was just evaluated.  Now consider the evaluation of \texttt{case e of
+  <alts>} if call-with-return is disallowed.  If the compiler cannot determine
+that evaluating \texttt{e} does not jump somewhere then there must be some
+mechanism to get to the \texttt{alts}.  Our solution is to have a separate C
+function for the \texttt{alts}, and before executing the code to evaluate
+\texttt{e}, to push a continuation---the address of this function and the free
+variables of the \texttt{alts}---onto the stack.  Whatever the code for
+\texttt{e} might do, ultimately (in the absence of error or non-termination)
+the value of \texttt{e} is in the CVR, and some code somewhere reaches its end
+and ``returns''---jumps to the address
+in the continuation on the top of the stack, thus entering the code for
+\texttt{alts}.  Thus continuations direct where to go to, not how to return
+to caller as in a stack for an imperative language.
+
+\subsubsection{Is this too much for this paper?}
+Similar to what is described in \emph{EA}, in an earlier
+implementation we have a generic variadic \emph{stgApply} C function that is
+used for applications of non-known functions, and known functions with a
+deficit or excess number of the arguments.  The \emph{stgApply} function takes
 as arguments the first element of the application (\emph{FUN}, \emph{PAP},
 \emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
-evaluates it with a call (rather than jump) to yield a \emph{FUN} or
-\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
-a C function taking a contiguous block of machine words (unboxed values
-or pointers) on the explicit stack.  The implementation of \emph{stgApply}
-as described in \emph{EA} makes the same concession regarding
-call-with-return, but the scheme described there uses an auto-generated family
-of \emph{stgApply} functions parameterized by number and type (unboxed or
-pointer) of arguments.  We used this scheme initially as well, but because the
-parameter block contains a bitmap describing which arguments are boxed/unboxed
-for the benefit of the garbage collector, this elaboration is unnecessary.
-For \emph{direct calls}, in which a known function is applied to a number
-of arguments equal to its arity, \emph{stgApply} need not be invoked:
-the \emph{FUN} is tail-called directly.
+evaluates it with a call-with-return to yield a \emph{FUN} or \emph{PAP}.
+This allows us to implement a single \emph{stgApply} function as a C function
+taking a contiguous block of machine words (unboxed values or pointers) on the
+explicit stack.  The implementation of \emph{stgApply} as described in
+\emph{EA} performs the same call-with-return, but the scheme described there
+uses an auto-generated family of \emph{stgApply} functions parameterized by
+number and type (unboxed or pointer) of arguments.  (We used this scheme
+initially as well, but because the stack is now fully explicit and the stack
+frame is self-describing (e.g., contains a bitmap describing which arguments
+are boxed/unboxed for the benefit of the garbage collector), this elaboration
+is unnecessary.) Call-with-return was also used in the case that a \emph{FUN}
+or \emph{PAP} is provided more than its arity number of arguments, in which
+case \emph{FUN} or \emph{PAP} is called-with-return with its arity number
+of objects, and the returned object and remaining argument passed (via jump)
+to \emph{stgApply}.  Finally, for strict evaluation orders it is straightforward
+to evaluate function or constructor arguments one at a time with call-with-return.
 
-Call-with-return becomes even more convenient when considering the various
-forms of strictness: when evaluating function or constructor arguments it is
-simpler, and arguably more efficient in terms of time (but not space in
-general beca) to ``evaluate in place'' than implement the continuation style
-to the logical limit.  We are aware, however, that we may need to eliminate
-this use of the native C stack so that we can properly unwind the stack in a
-more sophisticated parallel runtime when pruning multiple evaluations of the
-same thunk, and we have determined that this is a relatively minor exercise:
-one possibility is that \emph{stgApply} takes an additional argument
-indicating which of its subsequent arguments have already been forced, and at
-each invocation if unforced arguments remain, evaluates the next (possibly)
-unevaluated argument and tail calls itself, re-using the stack frame with just
-this new argument updated.
+At that point of development these calls-with-return were clearly a hack
+in an otherwise continuation-passing (via the explicit stack) control flow.
+Because call-with-return could be of unbounded call depth this would be
+problematic for stack unwinding.
 
-\emph{It seems that we can dispose with call-with-return of STG functions,
-  including stgApply (as opposed to runtime functions) without too much
-  effort.  It also seems like this is needed if we want to be able to unwind
-  the stack as described in the ``grey hole'' paper.  I'll try to do this this
-  week (3/31/16-4/1/16) to verify the feasibility before changing the
-  preceding description (which isn't quite right)}
+The solution, slightly simplified, entailed adding a new parameter to \emph{stgApply}
+and make the first parameter, the object to be applied, implicit.  The implicit
+parameter is in the CVR---a pointer to the object to be applied when jumping to \emph{stgApply}.
+The new argument simply indicates which of the boxed arguments
+
+For \emph{direct calls}, in which a known function is applied
+to a number of arguments equal to its arity, \emph{stgApply} need not be
+invoked: the \emph{FUN} is tail-called directly.
+
+In addition to being in a tail call position, the gcc constraints for
+generating a tail call (jump) are (approximately) that the total size of the
+caller and callee arguments be equal, and the return type sizes be equal.  For
+Clang the requirement is slightly more strict: caller and callee type
+signatures must be the same.  These are a consequence of the C calling
+convention that the caller, not callee, cleans up the stack (removes the
+callee's stack frame).  However, because we maintain our own stack we can
+define our own calling convention(s), so tail calls can be made to functions
+of notionally differing type.  In particular, top-of-stack frames can be
+adjusted in size and content without undue copying.
 

### 2016-04-01 pubs/tfp16/paper.tex
minor edits

@@
 \usepackage{color} % for parenthetical notes while paper is in progress
+\usepackage{comment}
 %
@@
 %
-\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
+\title{Automatic Parallelization and Transparent Fault Tolerance\\
        (New project paper)}
@@
 speculative evaluation.  A secondary goal is to experiment with mechanisms
-for transparent fault tolerance.
+for transparent fault tolerance, another trend in scientific computing, and
+by transitivity functional programming.
 \\
@@
     highlighted student involvement and its influence on our approach, and
-    included perhaps gratuitous detail about process.
+    included perhaps gratuitous detail about student contributions in particular.
   \end{enumerate}
 
-\keywords{strict pure functional programming, automatic parallelization, Haskell, STG}
+\keywords{strict pure functional programming, automatic parallelization, Haskell, STG,
+transparent fault tolerance.}
 
@@
 by the Workshop on Functional High Performance Computing, held in conjuction
-with the International Conference on Functional Programming, in its fifth
+with the International Conference on Functional Programming, being now in its fifth
 consecutive year.  A more precise characterization of the trend is that there
@@
 Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
-of \emph{tying the knot.}~\cite{tying-the-knot}.
+of \emph{tying the knot}~\cite{tying-the-knot}.
 
@@
 existing mental models. While they are happy to write code that is strongly
-reminiscent of the numerical examples in Hughes' classic piece~\cite{Hughes89}
-(whether they've read it or not), they would also be happy to write in a
-slightly modified style that does not require non-strictness---after all,
+reminiscent of the numerical examples in Hughes' classic piece (whether
+they've read it or not)~\cite{Hughes89}, they would also be happy to write in
+a slightly modified style that does not require non-strictness---after all,
 that's how they've been thinking their entire careers.
@@
 understand and not something they particularly want to try to understand.
-Hence, we eschew explicit strictness annotations for varying uniform degrees
-of strictness.
+%Hence, we eschew explicit strictness annotations for varying uniform degrees
+%of strictness.
 
@@
 somewhat vague, the general idea is a unit of computation with no or
-minimal/constrained side effects.  Function call in a pure functional language
-could be regarded as the essence of the concept of task.
+minimal/constrained side effects.  A function in a pure functional language
+could be regarded as the essence of the concept of task.  All of these
+systems are of course in languages with strict function-call semantics.
 
@@
 the software stack.  Current practice in HPC is checkpoint/restart, wherein at
-various point in time sufficient program state is dumped to backing store such
+various points in time sufficient program state is dumped to backing store such
 that the program, in case of a fault, can be (manually) restarted from the
@@
 programmer.  Ericsson's quasi-functional language Erlang and its runtime
-system was designed for fault tolerance.~\cite{Cesarini:2009}.  It has long
+system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
 been recognized that the pure parts of a functional program (excluding, e.g.,
 the I/O monad in Haskell) could be safely restarted.  The recent work of
-Steward impressively demonstrates this on a large-scale distributed-memory
+Stewart impressively demonstrates this on a large-scale distributed-memory
 system~\cite{Stewart:2013}.
@@
 (e.g.\ strict Haskell).  Second, we claim that there is a trend in
-high-performance computing, and by transitivity in functional programming,
-towards transparent fault tolerance.
+high-performance computing, and by transitivity (in addition to Stewart's
+work) in functional programming, towards transparent fault tolerance.
 
@@
 polymorphic, functional language and runtime system with which we can
-experiment with automatic parallelization strategies with degrees of language
-strictness ranging from lazy to fully strict, and secondarily, with mechanisms
-for transparent fault tolerance.  Light-weight has several meanings or
-implications.  First and foremost it means that \emph{we are not attempting to
-  compete with GHC in any respect}---succumbing to the temptation of
-recreating GHC language, type system, and runtime features, or significant
-code optimization, would simply divert us from our central goals.\footnote{We
-  estimate that we will have, to within an order of magnitude, one person-hour
-  of total available development time for each person-year that GHC currently
-  represents, or put another way, one second for each hour, respectively.}
-For another, it means that we are not undertaking significant morphing of GHC
-itself:  because this is in large part a student project it must be
-kept as simple as reasonably possible in various respects.
-
+experiment with automatic parallelization strategies with varying degrees of
+language strictness, and secondarily, with mechanisms for transparent fault
+tolerance.  Light-weight has several implications.  First and foremost it
+means that \emph{we are not attempting to compete with GHC in any
+  respect}---succumbing to the temptation of recreating GHC language, type
+system, and runtime features, or significant code optimization, would simply
+divert us from our central goals.\footnote{We estimate that we will have, to
+  within an order of magnitude, one person-hour of total available development
+  time for each person-year that GHC currently represents, or put another way,
+  one second for each hour, respectively.}  For another, it means that we are
+not undertaking significant morphing of GHC itself: because this is in large
+part a student project it must be kept as simple as reasonably possible in
+various respects.  Our initial requirements are that
 \begin{itemize}
@@
 
+For feasibility, we restrict ourselves to shared-memory implicit parallelism.
 The requirement for tail calling is not primarily for efficiency, rather it
@@
 exercise that we discuss at some length.
-
-For feasibility, we restrict ourselves to shared-memory implicit parallelism,
-but distributed-memory parallelism should be straightforwardly realizable
-using e.g.\ MPI or POSIX sockets.
 
@@
 
-\subsection{Varying degrees of strictness}
-First some words about terminology.  For all degrees of strictness, sharing
-(as implied by lazy but not by non-strict) of heap-allocated objects is
-preserved so we will not use the term \emph{lazy} and simply refer to
-strictness properties.  As is common we will use strict in an operational
-sense, meaning that argument(s) are evaluated before being passed in a
-function call, rather than the denotational sense, unless stated otherwise.
-Similarly, when referring to language semantics we refer by default to the
-operational rather than denotational semantics.
-
-Greater strictness implies more opportunities for parallelism, possible losses
-or gains in space efficiency, and less expressiveness.  We seek to explore
-the interplay of all of these.
-
-GHC 8.0 gives three possibilities: conventional non-strict, constructor
-strict, and function- and constructor strict, but there are possible
-variations.  The obvious variation is function, but not constructor,
-strictness.  In Haskell the distinction between constructors and functions is
-somewhat blurred because constructors behave like functions (outside of
-pattern matching), but this need not be the case.
-
 \section{Design and Implementation}
@@
 compiler are 1) parsing and type checking; 2) transformation to the GHC
-\emph{Core} intermediate representation; 3) analysis and transformation on
-Core; 4) transformation of Core to the \emph{STG} intermediate representation;
-5) transformation of STG to machine code via one of various imperative
+\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; 3)
+analysis and transformation on Core; 4) transformation of Core to the
+\emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
+transformation of STG to machine code via one of various imperative
 representations (Cmm, LLVM, C).
 
-In short, our plan is to use GHC as our front-end to generate its intermediate
-form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
-There are various possible ways to accomplish the this: using GHC's option
-to dump STG as text, hacking GHC itself and, optionally, using GHC as a
-library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core
-which we may transition to.
+In short, our plan is to use GHC as our front-end to generate one of its
+intermediate forms, STG or Core, then escape to our system from there.  There
+are various possible ways to accomplish the this: using GHC's option to dump
+STG or Core as text, hacking GHC itself and, optionally, using GHC as a
+library~\cite{ghc-as-library}.  For now we start with our own STG, with
+bridging the gap with GHC scheduled to commence soon.
 
@@
 
-\subsection{Degrees of strictness, take two}
+\subsection{Degrees of strictness}
+First some words about terminology.  For all degrees of strictness, sharing
+(as implied by lazy but not by non-strict) of heap-allocated objects is
+preserved so we will not use the term \emph{lazy} and simply refer to
+strictness properties.  As is common we will use strict in an operational
+sense, meaning that argument(s) are evaluated before being passed in a
+function call, rather than the denotational sense, unless stated otherwise.
+Similarly, when referring to language semantics we refer by default to the
+operational rather than denotational semantics.
+
+Greater strictness implies more opportunities for parallelism, possible losses
+or gains in space efficiency, and less expressiveness.  We seek to explore
+the interplay of all of these.
+
+GHC 8.0 gives three possibilities: conventional non-strict, constructor
+strict, and function- and constructor strict, but there are possible
+variations.  The obvious variation is function, but not constructor,
+strictness.  In Haskell the distinction between constructors and functions is
+somewhat blurred because constructors behave like functions (outside of
+pattern matching), but in STG the distinction is clear.
+
 Other possible variations concern the notional arity of the underlying
@@
 \emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
-arguments previously provided.  If the arity is greater than two then a new
+arguments previously provided.  If the arity is greater than two a new
 \emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
-the new \emph{PAP} returned.  If the arity is equal to two then the underlying
+the new \emph{PAP} returned.  If the arity is equal to two the underlying
 \emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
@@
 
-The other dimension we vary is whether or not constructors are strict.
-Other dimensions are possible, for example whether \texttt{let} bindings
-(heap allocations) are constructed on demand, but we restrict ourselves
-to function/application and constructor strictness.
+The other dimension we vary is whether or not constructors are strict.  Other
+dimensions are possible, for example whether \texttt{let} bindings are strict,
+but we restrict ourselves to function/application and constructor strictness.
 
@@
 al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
-aspirations~\cite{Marlow:2009}.  To achieve efficient execution they
+aspirations~\cite{Marlow:2009}.  To achieve efficient parallel execution they
 identified the need for the ability to examine and unwind the stack.  This
@@
 
-First we establish that this is possible.  By maintaining our own data
-and control (continuation) stack all generated C functions can take no
+First we establish that this is possible in principle.  By maintaining our own
+data and control (continuation) stack all generated C functions can take no
 explicit arguments and have return type \emph{void}. If we arrange that
@@
 a return in the control flow graph, then at non-zero optimization levels both
-gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When using
-gcc we can also use a fixed set of machine registers for passing/returning
-values, much like GHC.
+gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When
+using gcc we can also use a fixed set of machine registers for
+passing/returning values, much like GHC.
 
 \subsubsection{A taste of the continuation style.}
-Now to implementation, which we discuss for a simple case, then the most complex
-case.  To clearly distinguish between a notional C call (a syntactic concept)
-and a machine-level call (pushing a return address onto the native C stack) we
-will refer to the latter as \emph{call-with-return}, and to a machine-level
-tail call as \emph{jump}.
-
-First a detail.  Our implementation of the STG abstract machine has an
-abstract register called the \emph{current value register} or \emph{CVR}.
-Roughly speaking, it can hold the pointer to what is about to be evaluated or
-what was just evaluated.  Now consider the evaluation of \texttt{case e of
-  <alts>} if call-with-return is disallowed.  If the compiler cannot determine
-that evaluating \texttt{e} does not jump somewhere then there must be some
-mechanism to get to the \texttt{alts}.  Our solution is to have a separate C
-function for the \texttt{alts}, and before executing the code to evaluate
-\texttt{e}, to push a continuation---the address of this function and the free
-variables of the \texttt{alts}---onto the stack.  Whatever the code for
-\texttt{e} might do, ultimately (in the absence of error or non-termination)
-the value of \texttt{e} is in the CVR, and some code somewhere reaches its end
-and ``returns''---jumps to the address
-in the continuation on the top of the stack, thus entering the code for
-\texttt{alts}.  Thus continuations direct where to go to, not how to return
-to caller as in a stack for an imperative language.
-
+Now to implementation.  To clearly distinguish between a notional C call (a
+syntactic concept) and a machine-level call (pushing a return address onto the
+native C stack) we will refer to the latter as \emph{call-with-return}, and to
+a machine-level tail call as \emph{jump}.
+
+First some details.  Our implementation of the STG abstract machine has an
+abstract register called the \emph{current value} register or
+\emph{CV}.\footnote{There are actually two registers, one for pointers and
+  one for unboxed values, but for simplicity we minimize discussion about
+  unboxed values.}  Roughly speaking, it can hold the pointer to what is about
+to be evaluated or what was just evaluated.  For example, when initiating
+evaluation of a \emph{THUNK} object, \emph{CV} contains a pointer to that
+object; it is much like the \emph{self} pointer in C++.  On completion of
+evaluation \emph{CV} will contain a pointer to the object to which the
+\emph{THUNK} evaluated.  A \emph{THUNK} will not return another
+\emph{THUNK}---an evaluation is always ``big step'' to (at least) WHNF, that
+is, to a \emph{FUN}, \emph{PAP}, or \emph{CON}.
+% These latter, if evaluated, simply return through the continuation stack,
+%effectively returning \emph{self}.
+
+All continuations contain a code pointer, optionally some data, and layout
+information.  These continuations are maintained on a stack.  When the
+evaluation of an object is completed its return is by jumping through the code
+pointer of the continuation on the top of the stack (``returning through the
+continuation stack'').  For example, the argument frame for a function is one
+type of continuation and its associated code simply pops the continuation off
+the stack and returns through the next continuation on the stack.
+
+Before evaluating a user program an initial continuation is pushed onto
+the stack that points to a C function that performs whatever post-user-program
+processing might be wanted, such as displaying the result of the program in
+\emph{CV}\@.  The driver then jumps to the entry point of the user program,
+which will ultimately return through this continuation.  This makes clear
+that continuations direct \emph{where to go to}, not \emph{how to return
+to the caller} as in a stack for an imperative language, despite the use of
+the term ``return.''
+
+Now consider the evaluation of \texttt{case e of x \{ alts \}} if call-with-return
+is disallowed.  If the compiler cannot determine that evaluating \texttt{e}
+does not jump somewhere then there must be some mechanism to get to the
+\texttt{alts}.  Our solution follows \emph{EA}, using a separate C function
+for the \texttt{alts}, and before executing the code to evaluate \texttt{e},
+to push a continuation---the address of this function and the free variables
+of the \texttt{alts}---onto the stack.  Whatever the code for \texttt{e} might
+do, ultimately (in the absence of error or non-termination) the value of
+\texttt{e} is in \emph{CV}, and some code somewhere reaches its end and returns
+through the continuation stack, thus entering the code for \texttt{alts},
+which immediately binds \emph{x} to the value in \emph{CV}.
+
+\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
+functions used for applications of non-known (not statically identifiable)
+functions, and applications of known functions to other than their arity
+number of arguments.  This family is parameterized by the number of actual
+arguments in the application and whether each is boxed or unboxed.  Because
+we have an explicit stack we are able to use a single C \emph{stgApply} function
+taking a contiguous block of machine words (unboxed values or
+pointers) on the explicit stack.  Some modifications are required, however,
+because \emph{EA} uses call-with-return.  Specifically, when a \emph{FUN} or is
+\emph{PAP} applied to more than its arity number of arguments
+a call-with-return to the (underlying) \emph{FUN} is made with the arity
+number of arguments, then the appropriate \emph{stgApply} function is jumped
+to, passing the newly returned object and the remaining arguments.  Our
+solution to this entails passing \emph{stgApply} the object to be applied in
+\emph{CV}, together with an extra argument indicating which of the object
+to be applied and its boxed arguments have been already been evaluated (by
+\emph{stgApply}).  If the object to be applied, or some argument, remains to
+be evaluated (because of the strictness level), it is jumped to, with
+return as ever through the continuation stack.  This is actually fairly
+efficient because we can arrange that the stack frame for \emph{stgApply}
+is reused, so except for the indirect jumps on return this is essentially
+a loop.  For the cases of arguments in excess of the arity of the object
+being applied, in effect \emph{stgApply}'s argument frame is popped and
+two continuations are pushed, first one to jump to \emph{stgApply} with
+the excess arguments, and the next an argument frame for the object to
+applied, then the code for the object to be applied is jumped to.
+
+\begin{comment}
 \subsubsection{Is this too much for this paper?}
@@
 and make the first parameter, the object to be applied, implicit.  The implicit
-parameter is in the CVR---a pointer to the object to be applied when jumping to \emph{stgApply}.
+parameter is in \emph{CV}---a pointer to the object to be applied when jumping to \emph{stgApply}.
 The new argument simply indicates which of the boxed arguments
@@
 adjusted in size and content without undue copying.
+\end{comment}
 
@@
 
-No bridge from GHC STG to our STG yet.
+Developing the bridge between GHC Core or STG to our system is scheduled
+to start immediately after TiFP'16.
 
@@
 \begin{itemize}
-\item STG and mini-Haskell (inc. ADT w/unboxed) parsers \emph{(both by students)}
-\item mini-Haskell to STG transformer \emph{(by student)}
-\item type inferencer
-\item various analysis and transformation passes \emph{(all by students)}
-\item code generator \emph{(some student contribution)}
-\item runtime system including a garbage collector
-\item extensive test suite \emph{(mostly by students)}
+\item STG and mini-Haskell (inc. ADT w/unboxed) parsers \emph{(both by students)},
+\item mini-Haskell to STG transformer \emph{(by student)},
+\item type inferencer,
+\item various analysis and transformation passes \emph{(most by students)},
+\item code generator \emph{(some student contribution)},
+\item runtime system including a garbage collector,
+\item extensive test suite \emph{(mostly by students)}.
 \end{itemize}
@@
 We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
-small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
-not focused on the development of a more sophisticated garbage collector
-because the goal of this project is to explore automatic parallelization.
+small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. We have
+not worked on the development of a more sophisticated garbage collector
+because this would be a distraction from our main goals.
 

### 2016-04-05 pubs/tfp16/paper.tex
sync just paper
fix a few typos

@@
 might be useful or important, there is now a small number of laboratory HPC
-practioners who are developing an appreciation for pure functional programming
+practitioners who are developing an appreciation for pure functional programming
 in Haskell.  Still, they see their needs as different than other Haskell
@@
 \end{quote}
-It is a fact that \emph{HPC practioners are by and large not computer
+It is a fact that \emph{HPC practitioners are by and large not computer
   scientists}, they are domain scientists (physicists, chemists, etc.) or
@@
 \item Haskell Block \\
-  (ignored by MHS frontend, not other Haskell compilers;
+  (ignored by MHS front-end, not other Haskell compilers;
   allows the same file to be tested on our compiler vs GHC)
@@
 An almost unbounded amount of interesting work suggests itself; one such path
-would lead to the recreation of GHC.  However, given the miniscule budget (and
+would lead to the recreation of GHC.  However, given the minuscule budget (and
 thus the emphasis on student labor) we must maintain sharp focus on the main

### 2016-04-05 pubs/tfp16/paper.tex
more small changes

@@
 \usepackage{makeidx}  % allows for indexgeneration
+%
 % mkd for preliminary paper notes
@@
 \usepackage{color} % for parenthetical notes while paper is in progress
-\usepackage{comment}
+\usepackage{comment} % comment environment
+\usepackage{hyperref} 
 %
@@
 Interest in functional programming for HPC in general is not new, as evidenced
-by the Workshop on Functional High Performance Computing, held in conjuction
+by the Workshop on Functional High Performance Computing, held in conjunction
 with the International Conference on Functional Programming, being now in its fifth
@@
 al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
-aspirations~\cite{Marlow:2009}.  To achieve efficient parallel execution they
-identified the need for the ability to examine and unwind the stack.  This
-requires that use of the C stack be highly constrained if we are to 
-avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
+aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel 
+execution they identified the need for the ability to examine and unwind the 
+stack.  This requires that use of the C stack be highly constrained if we are 
+to avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
 non-portable code to directly manipulate the C stack, complicating an
@@
 
-There are several hundred small STG and mini-Haskell programs in the test suite.
+There are several hundred small STG and mini-Haskell programs in the test suite 
+(Most of which were written by students).
 It is partitioned into a number of directories which control what
@@
 
-debug level (runtime/compile time)
+We have also implemented a simple logging facility that can be controlled at 
+both compile and runtime which helps in debugging why a test is failing.
 
@@
 Parallel runtime is being implemented following the ideas
-of~\cite{SPJs-many-papers}, thread pool, stack and heap for each thread, blah,
-blah.
+of~\cite{Harris:2005,Marlow:2008,Marlow:2009}, thread pool, stack and heap 
+for each thread, blah, blah.
 

### 2016-04-05 pubs/tfp16/paper.tex
sync with Dean

--- 2016-04-01 pubs/tfp16/paper.tex
@@
 \begin{abstract}
-The claimed trend is an increasing appreciation and practice of functional
+A claimed trend is an increasing appreciation and practice of functional
 programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  Our project is the demonstration
-of a light-weight, higher-order, polymorphic, pure functional language
-implementation in which we can experiment with automatic parallelization
-strategies and varying degrees of default strictness.  Non-strictness
-constrains the degree of potential parallelism, that is, we are not exploring
-speculative evaluation.  A secondary goal is to experiment with mechanisms
-for transparent fault tolerance, another trend in scientific computing, and
-by transitivity functional programming.
+functional programming, in various guises.  A second claimed trend in both
+scientific computing and functional programming is transparent fault
+tolerance.  Our project goal is the demonstration of a light-weight,
+higher-order, polymorphic, pure functional language implementation in which we
+can experiment with automatic parallelization strategies, varying degrees of
+strictness, and mechanisms for transparent fault tolerance.
+%We do not consider speculative evaluation, or semantic strictness inferred by
+%program analysis, so potential parallelism is dictated by the degree of
+%strictness.
 \\
@@
 
-  \item Hopefully in the spirit of a student-friendly symposium, for this
-    draft version we have chosen a more informal, conversational style of
-    presentation than necessarily appropriate for a final paper, have
-    highlighted student involvement and its influence on our approach, and
-    included perhaps gratuitous detail about student contributions in particular.
+\item Hopefully in the spirit of a student-friendly symposium, for this draft
+  version we have chosen a more informal, conversational style of presentation
+  than necessarily appropriate for a final paper, have highlighted student
+  involvement and its influence on our approach, and pointed out
+  student contributions in particular.
 \end{enumerate}

### 2016-04-05 runtime/args.h

@@ -6,8 +6,9 @@
 extern float GCThreshold; // fraction of total heap used before gc runs.
 
 typedef enum {
-  LAZY, // default
-  STRICT1,
+  LAZY,    // default
+  STRICT1, // just function's arity args
+  STRICT2, // all args in application
 } EvalStrategy;
 
 extern int evalStrategy;

### 2016-04-05 runtime/cmm.h

@@ -18,7 +18,7 @@
 #include <stdio.h>
 #include <stdlib.h>
 
-#define TRAMPOLINE 1
+#define TRAMPOLINE 0
 
 #if TRAMPOLINE
 //------ fake typedef fp (*fp)() -- no recursive types in C

### 2016-04-05 runtime/stgapply.c

@@ -3,6 +3,54 @@
 #include "stgutils.h"
 #include "stgapply.h"
 
+// split current stack cont into two
+
+FnPtr stgApplyCurVal();
+
+Cont *stgContSplit(int excess) {
+  Cont *cont1 = (Cont *)stgSP;
+  size_t size1 = getContSize(cont1);
+  int pls1 = cont1->layout.bitmap.size;
+  int arity1 = pls1 - excess - 1;
+
+  int pls2 = 1 + excess; // funoid + excess
+  Cont *cont2 = stgAllocCallOrStackCont( &it_stgStackCont, pls2 );
+  size_t size2 = sizeof(Cont) + pls2 * sizeof(PtrOrLiteral);
+
+  int pls3 = pls1 - excess;
+  Cont *cont3 = stgAllocCallOrStackCont( &it_stgStackCont, pls3 );
+  size_t size3 = sizeof(Cont) + pls3 * sizeof(PtrOrLiteral);
+
+  Bitmap64 bm = cont1->layout;
+  bm.bitmap.size = pls2;
+  bm.bitmap.mask <<= (pls3 - 1); // room for funoid
+  bm.bitmap.mask |= 0x1UL;
+  cont2->layout = bm;
+  cont2->entryCode = &stgApplyCurVal;
+  cont2->payload[0].op = NULL;
+  #if USE_ARGTYPE
+  cont2->payload[0].argType = HEAPOBJ;
+  #endif
+  memcpy(&cont2->payload[1],
+         &cont1->payload[1 + arity1],
+	 excess * sizeof(PtrOrLiteral));
+
+  bm = cont1->layout;
+  bm.bitmap.size = pls3;  // mask can stay the same
+  cont3->layout = bm;
+  memcpy(&cont3->payload[0],
+         &cont1->payload[0],
+	 pls3 * sizeof(PtrOrLiteral));
+  //  cont3->entryCode = getInfoPtr(cont3->payload[0].op)->funFields.trueEntryCode;
+
+  // quash cont1
+  memmove(cont1, cont2, size2 + size3);
+  stgSP += size1;
+  return (Cont *)stgSP;
+}
+
+
+
 // might want to pass in bitmap and argv instead
 void stgEvalStackFrameArgs(Cont *cp) {
   // don't evaluate the funoid
@@ -20,11 +68,100 @@ void stgEvalStackFrameArgs(Cont *cp) {
 }
 
 
[…]

### 2016-04-05 runtime/stgutils.h

@@ -54,15 +54,15 @@ void stgPopContIfPopMe();
   STGCALL0(getInfoPtr(stgCurVal.op)->entryCode);	     \
   if (getObjType(stgCurVal.op) == BLACKHOLE) {		     \
     LOG(LOG_ERROR, "STGEVAL terminating on BLACKHOLE\n");   \
-    showStgVal(LOG_ERROR, stgCurVal);				     \
+    showStgVal(LOG_ERROR, stgCurVal);			     \
     exit(0);						     \
   }							     \
   if (getObjType(stgCurVal.op) == THUNK) {		     \
     LOG(LOG_ERROR, "THUNK at end of STGEVAL!\n");	     \
-    showStgVal(LOG_ERROR, stgCurVal);				     \
+    showStgVal(LOG_ERROR, stgCurVal);			     \
     assert(false);					     \
   }							     \
-  GC();					\
+  GC();							     \
 } while (0)
 
 // bye bye!
@@ -72,7 +72,7 @@ void stgPopContIfPopMe();
   derefStgCurVal();						     \
   if (getObjType(stgCurVal.op) == BLACKHOLE) {			     \
     LOG(LOG_ERROR, "STGJUMP terminating on BLACKHOLE\n");	     \
-    showStgVal(LOG_ERROR, stgCurVal);					     \
+    showStgVal(LOG_ERROR, stgCurVal);				     \
     exit(0);							     \
   }								     \
   STGJUMP0(getInfoPtr(stgCurVal.op)->entryCode);		     \

### 2016-04-05 pubs/tfp16/paper.tex
a few more changes

@@
 
-Using Legion, all non-function-local variables, or \emph{regions}, must
+Using Legion, all non-function-local variables, or \emph{regions}, must be
 requested from the runtime system.  Regions may be shared among \emph{tasks}
@@
 minimal/constrained side effects.  A function in a pure functional language
-could be regarded as the essence of the concept of task.  All of these
+could be regarded as the essence of the concept of a task.  All of these
 systems are of course in languages with strict function-call semantics.
@@
 The back-end is written in Haskell and generates C code;
-the runtime system is written in C/C++.  The system fairly faithfully
-implements the STG machine as described in \emph{EA} and dynamically
+the runtime system is written in C/C++. This allows us to leverage the extensive
+optimizers in modern C compilers such as gcc and Clang/LLVM. The system fairly 
+faithfully implements the STG machine as described in \emph{EA} and dynamically
 illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
@@
 enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jones91unboxedvalues},
+types as described by Peyton-Jones and Launchbury~\cite{Jone:1991},
 but for unboxed types code generation and runtime support is currently limited

### 2016-04-05 pubs/tfp16/paper.tex
more small tweaks

@@
 Mutual exclusion for garbage collector--heap check with every heap allocation.
-In fact garbage collector can be invoked at any point in time except when
+In fact the garbage collector can be invoked at any point in time except when
 populating a heap object immediately after its allocation.
@@
 enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jone:1991},
+types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
 but for unboxed types code generation and runtime support is currently limited
@@
 structures, strict.~\cite{strict-ghc-plugin}.  The transformation is on GHC
-Core, recursively expand non-recursive let bindings into case bindings.  Thus
+Core, recursively expanding non-recursive let bindings into case bindings.  Thus
 it seems that while it makes functions denotationally strict, function arguments
-are still evaluated after function call.
+are still evaluated after function calls.
 %
@@
 and augment our runtime as needed in this effort.  Just as in the Intel Haskell
-Research Compiler effort, we will not
+Research Compiler effort, we will use GHC as the front-end.
 

### 2016-04-05 pubs/tfp16/paper.tex
sync just paper

--- 2016-04-05 pubs/tfp16/paper.tex
@@
   than necessarily appropriate for a final paper, have highlighted student
-  involvement and its influence on our approach, and pointed out
-  student contributions in particular.
+  involvement and its influence on our approach, and pointed out student
+  contributions in particular.
 \end{enumerate}
 
-\keywords{strict pure functional programming, automatic parallelization, Haskell, STG,
-transparent fault tolerance.}
+\keywords{Strict pure functional programming, automatic parallelization,
+  Haskell, STG, GHC, GHC Core, transparent fault tolerance.}
 
@@
 parallelism also poorly abstracted, especially at the thread (shared-memory)
-level.  Thread-level parallel efficiency can be poor and is explained by
-Amdahl's law because the parallel model \emph{as typically implemented} is
-bulk synchronous.
-
-\subsection{The trend: strict but still pure functional programming in HPC}
+level.  Thread-level parallel efficiency can be good when the parallel model
+is bulk synchronous; tasking models are generally avoided unless abstracted
+by, e.g., a C++ template library.
+
+\subsection{A trend: strict but still pure functional programming in HPC}
 Interest in functional programming for HPC in general is not new, as evidenced
@@
 that \emph{pure functional semantics}, in some form, is essential to reining
-in the complexity of these ever-evolving scientific codes and informing the
+in the complexity of ever-evolving scientific codes and informing the
 designs of new ones.  Such semantics are expressed in a number of ways as
@@
 arbitrary naming of a block's variables at a subroutine/function granularity,
-and requiring that the entire block be brought into scope.  It is almost
-ironic, then, that of the mainstream high-performance scientific programming
-languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
-\texttt{pure} function qualifier, and with enforcement by the compiler.
-Otherwise it is entirely up to the programmer to make functions composable and
-thread safe.  Here at LANL the use of pure functions (among other techniques
-for reducing global data dependencies) is being used to good effect to clean
-up some legacy code bases.
+and requiring that entire blocks be brought into scope.  It is almost ironic,
+then, that of the mainstream high-performance scientific programming languages
+(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
+function qualifier, and with enforcement by the compiler.  Otherwise it is
+entirely up to the programmer to make functions composable and thread safe.
+Here at LANL the use of pure functions (among other techniques for reducing
+global data dependencies) is being used to good effect to clean up some legacy
+code bases.
 
@@
 might be useful or important, there is now a small number of laboratory HPC
-practitioners who are developing an appreciation for pure functional programming
-in Haskell.  Still, they see their needs as different than other Haskell
-users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
-primary data structure, and non-strictness or laziness is more of a
-nuisance--even a serious hindrance--than a convenience, much less ever
-algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
-Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
-of \emph{tying the knot}~\cite{tying-the-knot}.
+practitioners who are developing an appreciation for pure functional
+programming in Haskell.  Still, they see their needs as different than other
+Haskell users: performance is paramount, vectors (``Efficient
+Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
+non-strictness or laziness is more of a nuisance---even a serious
+hindrance---than a convenience, much less ever algorithmically essential in
+the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
+grammars~\cite{Johnsson87}, or other instances of \emph{tying the
+  knot}~\cite{tying-the-knot}.
 
@@
 programmers is the need to get strictness just right: to remove space leaks,
-achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
+achieve acceptable performance, get I/O and inter-process communication to work
 as they expect, and so on.  More specifically, they expect to be able to
@@
 synchronous model typically implemented with the Message Passing Interface
-(MPI)~\cite{MPI}.  In the imperative world, example systems include the
+(MPI) \cite{MPI}.  In the imperative world, example systems include the
 aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
@@
 that the program, in case of a fault, can be (manually) restarted from the
-latest stored state.  Most implementations are explicit, though there has been
+last stored state.  Most implementations are explicit, though there has been
 some success with transparent (invisible to the programmer) systems.  Wiring
@@
 the I/O monad in Haskell) could be safely restarted.  The recent work of
-Stewart impressively demonstrates this on a large-scale distributed-memory
-system~\cite{Stewart:2013}.
+Stewart impressively demonstrates this, in a pure functional setting, on a
+large-scale distributed-memory system~\cite{Stewart:2013}.
 
@@
 \begin{itemize}
-\item The implementation must be feasible in terms of effort;
+\item The implementation must be feasible in terms of available effort;
 \item The language must be strongly typed, implicitly or explicitly.  More
@@
 sufficiently limiting the use of the native C stack, makes this an interesting
-exercise that we discuss at some length.
+exercise that we will discuss at some length.
 
@@
 \emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
-transformation of STG to machine code via one of various imperative
-representations (Cmm, LLVM, C).
+transformation of STG to machine code via various imperative
+representations (Cmm, LLVM, C, native code generation).
 
@@
 library~\cite{ghc-as-library}.  For now we start with our own STG, with
-bridging the gap with GHC scheduled to commence soon.
+bridging the gap with GHC scheduled to commence quite soon.
 
@@
 In STG the arity of a function is syntactically defined as the number of
-manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
-regardless of the type of the argument (e.g., some function type) it might be
-applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
-a syntactic notion, from actually calling a function. Thus \texttt{id id x} is
-an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
-(user- or system- defined function) when actually called is given exactly one
-argument (the first one).
+manifest formal parameters in a function definition, thus \texttt{id x = x}
+has by definition arity one regardless of the type of the argument (e.g., some
+function type) it might be applied to.  We distinguish \emph{application},
+e.g., \texttt{f x y}, which is a syntactic notion, from actually calling a
+function. Thus \texttt{id id x} is an application of \texttt{id} to two
+arguments, but the underlying \emph{FUN} (user- or system- defined function)
+when actually called is given exactly one argument (the first one).
 
@@
 (non-strict) semantics \texttt{f} is first evaluated and the arity of the
-underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
-(partial application of a \emph{FUN}) is determined---in the case of
-\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
-arguments previously provided.  If the arity is greater than two a new
-\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
-the new \emph{PAP} returned.  If the arity is equal to two the underlying
-\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
-\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
-than two (it must be at least one, so one in this case) the underlying
-\emph{FUN} is called-with-return with any previous arguments (in case of
+underlying \emph{FUN} or \emph{PAP} (partial application of a \emph{FUN}) is
+determined---in the case of \emph{PAP} the arity is that of the underlying
+\emph{FUN} less the number of arguments previously provided.  If the arity is
+greater than two a new \emph{PAP} is created with arguments \texttt{x} and
+\texttt{y} inserted and the new \emph{PAP} returned.  If the arity is equal to
+two the underlying \emph{FUN} is tail-called (jumped to) with any previous
+arguments (in case of \emph{PAP}) and new arguments \texttt{x} and \texttt{y}.
+If the arity is less than two (it must be at least one, so one in this case)
+the underlying \emph{FUN} is called with any previous arguments (in case of
 \emph{PAP}) and new argument \texttt{x}, then the object returned applied to
@@
 (i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
-both of those arguments forced in subsequent evaluation (assuming that they
-weren't forced in the evaluation of \texttt{f} itself).
+both of those arguments possibly forced in subsequent evaluation (assuming
+that they weren't forced in the evaluation of \texttt{f} itself).
 
@@
 This then gives three reasonable operational modes that are strictly ordered
-in terms of termination properties: non-strict, fully strict---all arguments
-in an application are evaluated---and the intermediate in which only those
-arguments needed by a function call are evaluated.  The STG machinery makes
-the choice of evaluation strategy easy to change, both in terms of function
-application and building a constructor in the heap.
-
-The other dimension we vary is whether or not constructors are strict.  Other
-dimensions are possible, for example whether \texttt{let} bindings are strict,
-but we restrict ourselves to function/application and constructor strictness.
+in terms of termination properties: non-strict; function strict, in which only
+those arguments needed by a function call are evaluated; and application
+strict, in which all arguments of an application are evaluated.  The other
+dimension we vary is whether or not constructors are strict.  Other dimensions
+are possible, for example whether \texttt{let} bindings are strict, but we
+restrict ourselves to function/application and constructor strictness.  The
+STG machinery makes the choice of evaluation strategy easy to change, both in
+terms of function application and building a constructor in the heap.
 
@@
 
-All continuations contain a code pointer, optionally some data, and layout
+Continuations contain a code pointer, optionally some data, and layout
 information.  These continuations are maintained on a stack.  When the
@@
 
-Now consider the evaluation of \texttt{case e of x \{ alts \}} if call-with-return
-is disallowed.  If the compiler cannot determine that evaluating \texttt{e}
-does not jump somewhere then there must be some mechanism to get to the
-\texttt{alts}.  Our solution follows \emph{EA}, using a separate C function
-for the \texttt{alts}, and before executing the code to evaluate \texttt{e},
-to push a continuation---the address of this function and the free variables
-of the \texttt{alts}---onto the stack.  Whatever the code for \texttt{e} might
-do, ultimately (in the absence of error or non-termination) the value of
-\texttt{e} is in \emph{CV}, and some code somewhere reaches its end and returns
-through the continuation stack, thus entering the code for \texttt{alts},
-which immediately binds \emph{x} to the value in \emph{CV}.
+Now consider the evaluation of \texttt{case e of x \{ alts \}} if
+call-with-return is disallowed.  If the compiler cannot determine that
+evaluating \texttt{e} does not jump somewhere then there must be some
+mechanism to get to the \texttt{alts}.  Our solution follows \emph{EA}, using
+a separate C function for the \texttt{alts}, and before executing the code to
+evaluate \texttt{e}, to push a continuation---the address of this function and
+the free variables of the \texttt{alts}---onto the stack.  Whatever the code
+for \texttt{e} might do, ultimately (in the absence of error or
+non-termination) the value of \texttt{e} is in \emph{CV}, and some code
+somewhere has reached its end and returns through the continuation stack, thus
+entering the code for \texttt{alts}, which immediately associates \texttt{x}
+with the value in \emph{CV}.
 
@@
 number of arguments.  This family is parameterized by the number of actual
-arguments in the application and whether each is boxed or unboxed.  Because
-we have an explicit stack we are able to use a single C \emph{stgApply} function
-taking a contiguous block of machine words (unboxed values or
-pointers) on the explicit stack.  Some modifications are required, however,
-because \emph{EA} uses call-with-return.  Specifically, when a \emph{FUN} or is
-\emph{PAP} applied to more than its arity number of arguments
-a call-with-return to the (underlying) \emph{FUN} is made with the arity
-number of arguments, then the appropriate \emph{stgApply} function is jumped
-to, passing the newly returned object and the remaining arguments.  Our
-solution to this entails passing \emph{stgApply} the object to be applied in
-\emph{CV}, together with an extra argument indicating which of the object
-to be applied and its boxed arguments have been already been evaluated (by
-\emph{stgApply}).  If the object to be applied, or some argument, remains to
-be evaluated (because of the strictness level), it is jumped to, with
-return as ever through the continuation stack.  This is actually fairly
-efficient because we can arrange that the stack frame for \emph{stgApply}
-is reused, so except for the indirect jumps on return this is essentially
-a loop.  For the cases of arguments in excess of the arity of the object
-being applied, in effect \emph{stgApply}'s argument frame is popped and
-two continuations are pushed, first one to jump to \emph{stgApply} with
-the excess arguments, and the next an argument frame for the object to
-applied, then the code for the object to be applied is jumped to.
+arguments in the application and whether each is boxed or unboxed.  Because we
+have an explicit stack we are able to use a single C \emph{stgApply} function
+taking a contiguous block of machine words (unboxed values or pointers) on the
+explicit stack.  Some modifications are required, however, because \emph{EA}
+uses call-with-return.  Specifically, when a \emph{FUN} or is \emph{PAP}
+applied to more than its arity number of arguments a call-with-return to the
+(underlying) \emph{FUN} is made with the arity number of arguments, then the
+appropriate \emph{stgApply} function is jumped to, passing the newly returned
+object and the remaining arguments.  Call-with-return becomes even more
+convenient for strict evaluation:  arguments can be evaluated in a tight
+loop.
+
+Our solution to this entails passing \emph{stgApply} the object to be applied
+in \emph{CV}, together with an extra argument---call it
+\emph{index}---indicating which of the object to be applied and its boxed
+arguments have been already been evaluated (previously by \emph{stgApply}).
+If the object to be applied, or some argument (because of the strictness
+level), remains to be evaluated, it is jumped to, with return as ever through
+the continuation stack.  This is actually fairly efficient because we can
+arrange that the stack frame for \emph{stgApply} is reused, updating only
+\emph{index}, so except for the indirect jumps on return this is essentially a
+loop.  For the cases of arguments in excess of the arity of the object being
+applied, in effect \emph{stgApply}'s argument frame is popped and two
+continuations are pushed, the first one to jump to \emph{stgApply} with the
+excess arguments, the next an argument frame for the object to applied, then
+the code for the object to be applied is jumped to.
 
@@
 
-Mutual exclusion for garbage collector--heap check with every heap allocation.
-In fact garbage collector can be invoked at any point in time except when
-populating a heap object immediately after its allocation.
+%Mutual exclusion for garbage collector--heap check with every heap allocation.
+%In fact garbage collector can be invoked at any point in time except when
+%populating a heap object immediately after its allocation.
 
@@
 
-Various papers describing some of the inner workings of GHC at various point
-of its evolution have greatly informed our approach and these have all been
+Various papers describing some of the inner workings of GHC at various points
+in its evolution have greatly informed our approach and these have all been
 previously cited.
@@
 A plug-in for GHC (prior to 8.0) exists for making functions, but not data
-structures, strict.~\cite{strict-ghc-plugin}.  The transformation is on GHC
-Core, recursively expand non-recursive let bindings into case bindings.  Thus
-it seems that while it makes functions denotationally strict, function arguments
-are still evaluated after function call.
+structures, strict using a transformation on GHC Core that recursively expands
+non-recursive let bindings into case bindings.~\cite{strict-ghc-plugin}.  Thus
+it seems that while it makes functions denotationally strict, function
+arguments are still evaluated after function call.
 %
@@
 built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
-\emph{scan}).~\cite{NVIDIA:2013} Three back-ends were implemented: for
+\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
 sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
@@
 The Intel Labs Haskell Research Compiler (HRC) is the most closely related
-work to ours of which we are aware.~\cite{Liu:2013,Petersen:2013} As in our
-(planned) approach they use GHC as the front end, but exit the GHC toolchain
+work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
+planned approach they use GHC as the front end, but exit the GHC toolchain
 at the final GHC Core stage, just before transformation to STG.  The essence
@@
 
-In his BSc thesis, Don Stewart described an implementation wherein GHC was
-used as a Haskell compiler front-end, dumping STG code as text, parsing that
-text, then translating to code for execution on the Java Virtual
-Machine.~\cite{Stewart-BSc}.
+Don Stewart described an implementation wherein GHC was used as a Haskell compiler
+front-end, dumping STG code as text, parsing that text, then translating to
+code for execution on the Java Virtual Machine.~\cite{Stewart-BSc}.
 
@@
 distributed Haskell based on distributed exception
-handling.~\cite{Pointon:2001}.  However, it appears that their scheme for
+handling~\cite{Pointon:2001}.  However, it appears that their scheme for
 handling failed processing elements correctly~\cite{} may have never been
@@
 Currently we are developing the parallel runtime.  This summer (2016) a
-returning student will (partially) bridge the gap between GHC STG and our STG,
-and augment our runtime as needed in this effort.  Just as in the Intel Haskell
-Research Compiler effort, we will not
+returning student will (partially) bridge the gap between GHC STG (or Core)
+and our STG, and augment our runtime as needed in this effort.  Just as in the
+Intel Haskell Research Compiler effort, we will not attempt to reuse the GHC
+runtime.
 

### 2016-04-05 pubs/tfp16/paper.tex
one more small fix

--- 2016-04-05 pubs/tfp16/paper.tex
@@
 Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
-Department of Energy’s National Nuclear Security Administration (NNSA).
+Department of Energy's National Nuclear Security Administration (NNSA).
 

### 2016-04-05 pubs/tfp16/paper.tex

--- 2016-04-05 pubs/tfp16/paper.tex
@@
 enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jones91unboxedvalues},
+types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
 but for unboxed types code generation and runtime support is currently limited
@@
 Currently we are developing the parallel runtime.  This summer (2016) a
-returning student will (partially) bridge the gap between GHC STG (or Core)
-and our STG, and augment our runtime as needed in this effort.  Just as in the
-Intel Haskell Research Compiler effort, we will not attempt to reuse the GHC
+returning student will bridge the gap between GHC STG (or Core) and our STG,
+and augment our runtime as needed in this effort.  Just as in the Intel
+Haskell Research Compiler effort, we will not attempt to reuse the GHC
 runtime.

### 2016-04-05 pubs/tfp16/paper.tex
edits

@@
 The practice of so-called high-performance scientific computing (HPC) is,
-overall, highly conservative with respect to change.  Even as the US
-government commits to a push for exascale computing ($10^{18}$ FLOPS in a
-single system) within the next decade~\cite{Exascale-roadmap} a highly
-respected member of the HPC community recently stated publicly that ``Fortran
-is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
-  Japanese, and international groups are also anticipating the realization
-  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
-laboratory HPC codes run to hundreds of thousands of lines of Fortran and
+overall, highly conservative with respect to change.  Even as the USA and
+other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
+a single system) within the next
+decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a highly respected
+member of the HPC community recently stated publicly that ``Fortran is
+essential for exascale programming.''~\cite{Heroux16} Existing US government
+laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
 complete rewrites are infeasible.\footnote{The codes of our lab's British
-  counterpart are reportedly in the same state.}  C++ has become more common
-for new code starts, but their nature remains the same: huge codes with fair
-to non-existent encapsulation of side effects, with multiple levels of
-parallelism also poorly abstracted, especially at the thread (shared-memory)
-level.  Thread-level parallel efficiency can be good when the parallel model
-is bulk synchronous; tasking models are generally avoided unless abstracted
-by, e.g., a C++ template library.
+  counterpart are similar.}  C++ has become more common for new code starts,
+but their nature remains the same: huge codes with fair to non-existent
+encapsulation of side effects, with multiple levels of parallelism sometimes
+poorly abstracted, especially at the thread (shared-memory) and computational
+accelerator (e.g., GPU) levels.  Thread-level parallel efficiency can be good
+when the parallel model is bulk synchronous; tasking models are generally
+avoided unless abstracted by, e.g., a C++ template or runtime library, because
+of the programming complexity.
 
 \subsection{A trend: strict but still pure functional programming in HPC}
-Interest in functional programming for HPC in general is not new, as evidenced
+Interest in functional programming for HPC in general is not new as evidenced
 by the Workshop on Functional High Performance Computing, held in conjunction
@@
 entirely up to the programmer to make functions composable and thread safe.
-Here at LANL the use of pure functions (among other techniques for reducing
-global data dependencies) is being used to good effect to clean up some legacy
-code bases.
+%Here at LANL the use of pure functions (among other techniques for reducing
+%global data dependencies) is being used to good effect to clean up some legacy
+%code bases.
 
@@
 which it will have access.  Other than registered access to regions, tasks
-must not access non-constant global data.  In serial execution these access
+must not access non-constant global data.  Thus the inputs and outputs of
+tasks are exactly known to the runtime system.  In serial execution these access
 requirements are notionally superfluous, but in parallel execution the runtime
@@
 parallel efficiency has been demonstrated on extremely large computing
-systems~\cite{Bauer14}.  Here at LANL the Legion system is being strongly
-considered for a new large-scale C++ code effort.\footnote{It seems
+systems~\cite{Bauer14}.\footnote{It seems
   significant that the PI of the Legion project was previously a co-developer
@@
 collectively had no conceptual understanding of what it was, much less why it
-might be useful or important, there is now a small number of laboratory HPC
+might be of some import, there is now a small number of laboratory HPC
 practitioners who are developing an appreciation for pure functional
@@
 as they expect, and so on.  More specifically, they expect to be able to
-reason about space and time usage, and order of execution, using their
+reason about space and time usage, and order of evaluation, using their
 existing mental models. While they are happy to write code that is strongly
@@
 \end{quote}
-It is a fact that \emph{HPC practitioners are by and large not computer
+It is a fact that \emph{HPC practitioners are generally not computer
   scientists}, they are domain scientists (physicists, chemists, etc.) or
@@
 for HPC, often in conjunction with a higher-level (typically inter-node) bulk
-synchronous model typically implemented with the Message Passing Interface
-(MPI) \cite{MPI}.  In the imperative world, example systems include the
+synchronous model usually implemented with the Message Passing Interface
+(MPI)~\cite{MPI}\@.  In the imperative world, example systems include the
 aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
@@
 
+\subsection{Another trend: transparent fault tolerance}
 In the quest for computing platforms approaching exascale, it is widely
@@
 the software stack.  Current practice in HPC is checkpoint/restart, wherein at
-various points in time sufficient program state is dumped to backing store such
-that the program, in case of a fault, can be (manually) restarted from the
-last stored state.  Most implementations are explicit, though there has been
-some success with transparent (invisible to the programmer) systems.  Wiring
-checkpoint/restart into a large-scale application is not only burdensome and
-error-prone, it can severely warp the engineering of a large-scale application
+various points in time sufficient program state is dumped to backing store
+such that the program, in case of a fault, can be (manually) restarted from
+the last stored state.  Most implementations are explicit, though there has
+been some success with transparent (invisible to the programmer) systems.
+Wiring checkpoint/restart into a large-scale application is not only burdensome and
+error-prone, it can severely warp the engineering of the application
 because it must be designed such that essential state can be readily captured.
-And now, it is widely believed that the scalability of checkpoint/restart may
-be reaching its end.
+Furthermore, there is a concern that the scalability of checkpoint/restart is
+reaching its end.
 
@@
 various respects.  Our initial requirements are that
+%
 \begin{itemize}
@@
 \end{itemize}
-
+%
 For feasibility, we restrict ourselves to shared-memory implicit parallelism.
@@
 In a strict semantics one could devise various evaluation orders, but as far
-as distinguishing termination properties these can be distilled down to two
-reasonable possibilities, namely whether on construction of a \emph{PAP} the
-(new) arguments are forced or not.  In other words, the question is whether
-application is strict, or actual function call is strict.  These are in fact
-distinguishable because (in Haskell) the first argument to \texttt{seq} might
-evaluate to a \emph{PAP}.
+as distinguishing termination properties it seems these can be distilled down
+to two reasonable possibilities, namely whether on construction of a
+\emph{PAP} the (new) arguments are forced or not.  In other words, the
+question is whether application is strict, or actual function call is strict.
+These are in fact distinguishable because (in Haskell) the first argument to
+\texttt{seq} might evaluate to a \emph{PAP}.
 
@@
 restrict ourselves to function/application and constructor strictness.  The
-STG machinery makes the choice of evaluation strategy easy to change, both in
-terms of function application and building a constructor in the heap.
+STG machinery makes the choice of evaluation strategy easy to change.
+%, both in terms of function application and building a constructor in the
+%heap.
 
 \subsection{STG to C}
-The back-end is written in Haskell and generates C code;
-the runtime system is written in C/C++. This allows us to leverage the extensive
-optimizers in modern C compilers such as gcc and Clang/LLVM. The system fairly 
-faithfully implements the STG machine as described in \emph{EA} and dynamically
+The back-end is written in Haskell and generates C code; the runtime system is
+written in C/C++. This gives us the extensive optimization performed by modern
+C compilers such as gcc and Clang/LLVM. The system fairly faithfully
+implements the STG machine as described in \emph{EA} and dynamically
 illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
-transition rules which, of course, may be different than those for non-strict
-semantics.  Sharing is preserved.
-
-\subsubsection{Proper tail calls.}
+transition rules which may be different than those for non-strict semantics.
+Sharing is preserved.
+
+\subsubsection{Proper tail calls in C}
 Parallelism is achieved by multi-threading; our model is based on Harris et
@@
 execution they identified the need for the ability to examine and unwind the
-stack.  This requires that use of the C stack be highly constrained if we are 
-to avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
-non-portable code to directly manipulate the C stack, complicating an
-already non-trivial implementation effort.  We therefore must limit the
-use of the C stack to calls to the runtime, in other words, when executing
-code generated from the user program (and not in a runtime call), there
-must be no growing of (pushing return addresses or data onto) the C stack.
+stack in quashing concurrent evaluation of the same \emph{THUNK}.  This
+requires that use of the C stack be highly constrained if we are to avoid
+complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
+code to directly manipulate the C stack, complicating an already non-trivial
+implementation effort.  We therefore must limit the use of the C stack to
+calls to the runtime, in other words, when executing code generated from the
+user program (and not in a runtime call), there must be no growing of (pushing
+return addresses or data onto) the C stack.
+%
+To clearly distinguish between a notional C call (a syntactic concept) and a
+machine-level call (pushing a return address onto the native C stack) we will
+refer to the latter as \emph{call-with-return}, and to a machine-level tail
+call as \emph{jump}.
 
@@
 a return in the control flow graph, then at non-zero optimization levels both
-gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When
-using gcc we can also use a fixed set of machine registers for
-passing/returning values, much like GHC.
-
+gcc and Clang/LLVM will generate jumps instead of calls-with-return for C
+calls.  When using gcc we can also use a fixed set of machine registers for
+passing/returning values, much like GHC using the LLVM back-end with its
+custom GHC calling convention.
+
+The STG abstract machine is implemented using a continuation stack.
+Continuations contain a code pointer, optionally some data, and layout
+information for the benefit of the garbage collector.  For example, the
+argument frame for a function call is one type of continuation.  When the
+evaluation of a C function (corresponding to some STG code) is completed its
+return is by jumping through the code pointer of the continuation on the top
+of the stack (``returning through the continuation stack'').  In the case of
+an argument frame for a function, the associated code simply pops the
+continuation off the stack and returns through the next continuation on the
+stack.
+
+\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
+functions used for applications of non-known (not statically identifiable)
+functions, and applications of known functions to other than their arity
+number of arguments.  Here the object to be applied is first evaluated with
+call-with-return.  When a resulting \emph{FUN} is applied to more than its
+arity number of arguments, it is called-with-return with its arity number of
+arguments, and the resulting value applied to the remaining arguments.  When
+considering more strict semantics call-with-return is even more convenient
+because arguments can be evaluated in a tight loop.  Unfortunately these uses
+of call-with-return can result in the C stack growing unboundedly.  Our
+solution uses a single C \emph{stgApply} function with no calls-with-return.
+This is straightforwardly realizable because the stack is explicit, and with
+the C compiler call optimization, the continuation style provides a proper
+\emph{goto} mechanism where the \emph{code labels}---entities that name blocks
+of code, can be treated as first class (e.g.\ stored), and serve as targets of
+a jump---are C function addresses (c.f.\ \S 6.1~\cite{PJ-stockhardware}).
+
+\begin{comment}
 \subsubsection{A taste of the continuation style.}
@@
 the code for the object to be applied is jumped to.
+\end{comment}
 

### 2016-04-06 pubs/tfp16/paper.tex
write up of fancy patterns, smaller features of MHS

--- 2016-04-05 pubs/tfp16/paper.tex
@@
 \begin{abstract}
-A claimed trend is an increasing appreciation and practice of functional
+The claimed trend is an increasing appreciation and practice of functional
 programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  A second claimed trend in both
-scientific computing and functional programming is transparent fault
-tolerance.  Our project goal is the demonstration of a light-weight,
-higher-order, polymorphic, pure functional language implementation in which we
-can experiment with automatic parallelization strategies, varying degrees of
-strictness, and mechanisms for transparent fault tolerance.
-%We do not consider speculative evaluation, or semantic strictness inferred by
-%program analysis, so potential parallelism is dictated by the degree of
-%strictness.
+functional programming, in various guises.  Our project is the demonstration
+of a light-weight, higher-order, polymorphic, pure functional language
+implementation in which we can experiment with automatic parallelization
+strategies and varying degrees of default strictness.  Non-strictness
+constrains the degree of potential parallelism, that is, we are not exploring
+speculative evaluation.  A secondary goal is to experiment with mechanisms
+for transparent fault tolerance, another trend in scientific computing, and
+by transitivity functional programming.
 \\
@@
 
-\item Hopefully in the spirit of a student-friendly symposium, for this draft
-  version we have chosen a more informal, conversational style of presentation
-  than necessarily appropriate for a final paper, have highlighted student
-  involvement and its influence on our approach, and pointed out student
-  contributions in particular.
+  \item Hopefully in the spirit of a student-friendly symposium, for this
+    draft version we have chosen a more informal, conversational style of
+    presentation than necessarily appropriate for a final paper, have
+    highlighted student involvement and its influence on our approach, and
+    included perhaps gratuitous detail about student contributions in particular.
   \end{enumerate}
 
-\keywords{Strict pure functional programming, automatic parallelization,
-  Haskell, STG, GHC, GHC Core, transparent fault tolerance.}
+\keywords{strict pure functional programming, automatic parallelization, Haskell, STG,
+transparent fault tolerance.}
 
@@
 parallelism also poorly abstracted, especially at the thread (shared-memory)
-level.  Thread-level parallel efficiency can be good when the parallel model
-is bulk synchronous; tasking models are generally avoided unless abstracted
-by, e.g., a C++ template library.
-
-\subsection{A trend: strict but still pure functional programming in HPC}
+level.  Thread-level parallel efficiency can be poor and is explained by
+Amdahl's law because the parallel model \emph{as typically implemented} is
+bulk synchronous.
+
+\subsection{The trend: strict but still pure functional programming in HPC}
 Interest in functional programming for HPC in general is not new, as evidenced
@@
 that \emph{pure functional semantics}, in some form, is essential to reining
-in the complexity of ever-evolving scientific codes and informing the
+in the complexity of these ever-evolving scientific codes and informing the
 designs of new ones.  Such semantics are expressed in a number of ways as
@@
 arbitrary naming of a block's variables at a subroutine/function granularity,
-and requiring that entire blocks be brought into scope.  It is almost ironic,
-then, that of the mainstream high-performance scientific programming languages
-(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
-function qualifier, and with enforcement by the compiler.  Otherwise it is
-entirely up to the programmer to make functions composable and thread safe.
-Here at LANL the use of pure functions (among other techniques for reducing
-global data dependencies) is being used to good effect to clean up some legacy
-code bases.
+and requiring that the entire block be brought into scope.  It is almost
+ironic, then, that of the mainstream high-performance scientific programming
+languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
+\texttt{pure} function qualifier, and with enforcement by the compiler.
+Otherwise it is entirely up to the programmer to make functions composable and
+thread safe.  Here at LANL the use of pure functions (among other techniques
+for reducing global data dependencies) is being used to good effect to clean
+up some legacy code bases.
 
@@
 
-Using Legion, all non-function-local variables, or \emph{regions}, must be
+Using Legion, all non-function-local variables, or \emph{regions}, must
 requested from the runtime system.  Regions may be shared among \emph{tasks}
@@
 might be useful or important, there is now a small number of laboratory HPC
-practitioners who are developing an appreciation for pure functional
-programming in Haskell.  Still, they see their needs as different than other
-Haskell users: performance is paramount, vectors (``Efficient
-Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
-non-strictness or laziness is more of a nuisance---even a serious
-hindrance---than a convenience, much less ever algorithmically essential in
-the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
-grammars~\cite{Johnsson87}, or other instances of \emph{tying the
-  knot}~\cite{tying-the-knot}.
+practitioners who are developing an appreciation for pure functional programming
+in Haskell.  Still, they see their needs as different than other Haskell
+users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
+primary data structure, and non-strictness or laziness is more of a
+nuisance--even a serious hindrance--than a convenience, much less ever
+algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
+Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
+of \emph{tying the knot}~\cite{tying-the-knot}.
 
@@
 programmers is the need to get strictness just right: to remove space leaks,
-achieve acceptable performance, get I/O and inter-process communication to work
+achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
 as they expect, and so on.  More specifically, they expect to be able to
@@
 synchronous model typically implemented with the Message Passing Interface
-(MPI) \cite{MPI}.  In the imperative world, example systems include the
+(MPI)~\cite{MPI}.  In the imperative world, example systems include the
 aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
@@
 minimal/constrained side effects.  A function in a pure functional language
-could be regarded as the essence of the concept of a task.  All of these
+could be regarded as the essence of the concept of task.  All of these
 systems are of course in languages with strict function-call semantics.
@@
 that the program, in case of a fault, can be (manually) restarted from the
-last stored state.  Most implementations are explicit, though there has been
+latest stored state.  Most implementations are explicit, though there has been
 some success with transparent (invisible to the programmer) systems.  Wiring
@@
 the I/O monad in Haskell) could be safely restarted.  The recent work of
-Stewart impressively demonstrates this, in a pure functional setting, on a
-large-scale distributed-memory system~\cite{Stewart:2013}.
+Stewart impressively demonstrates this on a large-scale distributed-memory
+system~\cite{Stewart:2013}.
 
@@
 \begin{itemize}
-\item The implementation must be feasible in terms of available effort;
+\item The implementation must be feasible in terms of effort;
 \item The language must be strongly typed, implicitly or explicitly.  More
@@
 sufficiently limiting the use of the native C stack, makes this an interesting
-exercise that we will discuss at some length.
+exercise that we discuss at some length.
 
@@
 \emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
-transformation of STG to machine code via various imperative
-representations (Cmm, LLVM, C, native code generation).
+transformation of STG to machine code via one of various imperative
+representations (Cmm, LLVM, C).
 
@@
 library~\cite{ghc-as-library}.  For now we start with our own STG, with
-bridging the gap with GHC scheduled to commence quite soon.
+bridging the gap with GHC scheduled to commence soon.
 
@@
 In STG the arity of a function is syntactically defined as the number of
-manifest formal parameters in a function definition, thus \texttt{id x = x}
-has by definition arity one regardless of the type of the argument (e.g., some
-function type) it might be applied to.  We distinguish \emph{application},
-e.g., \texttt{f x y}, which is a syntactic notion, from actually calling a
-function. Thus \texttt{id id x} is an application of \texttt{id} to two
-arguments, but the underlying \emph{FUN} (user- or system- defined function)
-when actually called is given exactly one argument (the first one).
+manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
+regardless of the type of the argument (e.g., some function type) it might be
+applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
+a syntactic notion, from actually calling a function. Thus \texttt{id id x} is
+an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
+(user- or system- defined function) when actually called is given exactly one
+argument (the first one).
 
@@
 (non-strict) semantics \texttt{f} is first evaluated and the arity of the
-underlying \emph{FUN} or \emph{PAP} (partial application of a \emph{FUN}) is
-determined---in the case of \emph{PAP} the arity is that of the underlying
-\emph{FUN} less the number of arguments previously provided.  If the arity is
-greater than two a new \emph{PAP} is created with arguments \texttt{x} and
-\texttt{y} inserted and the new \emph{PAP} returned.  If the arity is equal to
-two the underlying \emph{FUN} is tail-called (jumped to) with any previous
-arguments (in case of \emph{PAP}) and new arguments \texttt{x} and \texttt{y}.
-If the arity is less than two (it must be at least one, so one in this case)
-the underlying \emph{FUN} is called with any previous arguments (in case of
+underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
+(partial application of a \emph{FUN}) is determined---in the case of
+\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
+arguments previously provided.  If the arity is greater than two a new
+\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
+the new \emph{PAP} returned.  If the arity is equal to two the underlying
+\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
+\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
+than two (it must be at least one, so one in this case) the underlying
+\emph{FUN} is called-with-return with any previous arguments (in case of
 \emph{PAP}) and new argument \texttt{x}, then the object returned applied to
@@
 (i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
-both of those arguments possibly forced in subsequent evaluation (assuming
-that they weren't forced in the evaluation of \texttt{f} itself).
+both of those arguments forced in subsequent evaluation (assuming that they
+weren't forced in the evaluation of \texttt{f} itself).
 
@@
 This then gives three reasonable operational modes that are strictly ordered
-in terms of termination properties: non-strict; function strict, in which only
-those arguments needed by a function call are evaluated; and application
-strict, in which all arguments of an application are evaluated.  The other
-dimension we vary is whether or not constructors are strict.  Other dimensions
-are possible, for example whether \texttt{let} bindings are strict, but we
-restrict ourselves to function/application and constructor strictness.  The
-STG machinery makes the choice of evaluation strategy easy to change, both in
-terms of function application and building a constructor in the heap.
+in terms of termination properties: non-strict, fully strict---all arguments
+in an application are evaluated---and the intermediate in which only those
+arguments needed by a function call are evaluated.  The STG machinery makes
+the choice of evaluation strategy easy to change, both in terms of function
+application and building a constructor in the heap.
+
+The other dimension we vary is whether or not constructors are strict.  Other
+dimensions are possible, for example whether \texttt{let} bindings are strict,
+but we restrict ourselves to function/application and constructor strictness.
 
@@
 The back-end is written in Haskell and generates C code;
-the runtime system is written in C/C++. This allows us to leverage the extensive
-optimizers in modern C compilers such as gcc and Clang/LLVM. The system fairly 
-faithfully implements the STG machine as described in \emph{EA} and dynamically
+the runtime system is written in C/C++.  The system fairly faithfully
+implements the STG machine as described in \emph{EA} and dynamically
 illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
@@
 
-Continuations contain a code pointer, optionally some data, and layout
+All continuations contain a code pointer, optionally some data, and layout
 information.  These continuations are maintained on a stack.  When the
@@
 
-Now consider the evaluation of \texttt{case e of x \{ alts \}} if
-call-with-return is disallowed.  If the compiler cannot determine that
-evaluating \texttt{e} does not jump somewhere then there must be some
-mechanism to get to the \texttt{alts}.  Our solution follows \emph{EA}, using
-a separate C function for the \texttt{alts}, and before executing the code to
-evaluate \texttt{e}, to push a continuation---the address of this function and
-the free variables of the \texttt{alts}---onto the stack.  Whatever the code
-for \texttt{e} might do, ultimately (in the absence of error or
-non-termination) the value of \texttt{e} is in \emph{CV}, and some code
-somewhere has reached its end and returns through the continuation stack, thus
-entering the code for \texttt{alts}, which immediately associates \texttt{x}
-with the value in \emph{CV}.
+Now consider the evaluation of \texttt{case e of x \{ alts \}} if call-with-return
+is disallowed.  If the compiler cannot determine that evaluating \texttt{e}
+does not jump somewhere then there must be some mechanism to get to the
+\texttt{alts}.  Our solution follows \emph{EA}, using a separate C function
+for the \texttt{alts}, and before executing the code to evaluate \texttt{e},
+to push a continuation---the address of this function and the free variables
+of the \texttt{alts}---onto the stack.  Whatever the code for \texttt{e} might
+do, ultimately (in the absence of error or non-termination) the value of
+\texttt{e} is in \emph{CV}, and some code somewhere reaches its end and returns
+through the continuation stack, thus entering the code for \texttt{alts},
+which immediately binds \emph{x} to the value in \emph{CV}.
 
@@
 number of arguments.  This family is parameterized by the number of actual
-arguments in the application and whether each is boxed or unboxed.  Because we
-have an explicit stack we are able to use a single C \emph{stgApply} function
-taking a contiguous block of machine words (unboxed values or pointers) on the
-explicit stack.  Some modifications are required, however, because \emph{EA}
-uses call-with-return.  Specifically, when a \emph{FUN} or is \emph{PAP}
-applied to more than its arity number of arguments a call-with-return to the
-(underlying) \emph{FUN} is made with the arity number of arguments, then the
-appropriate \emph{stgApply} function is jumped to, passing the newly returned
-object and the remaining arguments.  Call-with-return becomes even more
-convenient for strict evaluation:  arguments can be evaluated in a tight
-loop.
-
-Our solution to this entails passing \emph{stgApply} the object to be applied
-in \emph{CV}, together with an extra argument---call it
-\emph{index}---indicating which of the object to be applied and its boxed
-arguments have been already been evaluated (previously by \emph{stgApply}).
-If the object to be applied, or some argument (because of the strictness
-level), remains to be evaluated, it is jumped to, with return as ever through
-the continuation stack.  This is actually fairly efficient because we can
-arrange that the stack frame for \emph{stgApply} is reused, updating only
-\emph{index}, so except for the indirect jumps on return this is essentially a
-loop.  For the cases of arguments in excess of the arity of the object being
-applied, in effect \emph{stgApply}'s argument frame is popped and two
-continuations are pushed, the first one to jump to \emph{stgApply} with the
-excess arguments, the next an argument frame for the object to applied, then
-the code for the object to be applied is jumped to.
+arguments in the application and whether each is boxed or unboxed.  Because
+we have an explicit stack we are able to use a single C \emph{stgApply} function
+taking a contiguous block of machine words (unboxed values or
+pointers) on the explicit stack.  Some modifications are required, however,
+because \emph{EA} uses call-with-return.  Specifically, when a \emph{FUN} or is
+\emph{PAP} applied to more than its arity number of arguments
+a call-with-return to the (underlying) \emph{FUN} is made with the arity
+number of arguments, then the appropriate \emph{stgApply} function is jumped
+to, passing the newly returned object and the remaining arguments.  Our
+solution to this entails passing \emph{stgApply} the object to be applied in
+\emph{CV}, together with an extra argument indicating which of the object
+to be applied and its boxed arguments have been already been evaluated (by
+\emph{stgApply}).  If the object to be applied, or some argument, remains to
+be evaluated (because of the strictness level), it is jumped to, with
+return as ever through the continuation stack.  This is actually fairly
+efficient because we can arrange that the stack frame for \emph{stgApply}
+is reused, so except for the indirect jumps on return this is essentially
+a loop.  For the cases of arguments in excess of the arity of the object
+being applied, in effect \emph{stgApply}'s argument frame is popped and
+two continuations are pushed, first one to jump to \emph{stgApply} with
+the excess arguments, and the next an argument frame for the object to
+applied, then the code for the object to be applied is jumped to.
 
@@
 translated into STG for the backend.  Here, the differences are given in order
-of perceived utility, or importance to someone transition from STG to
+of perceived utility, or importance to someone transitioning from STG to
 mini-Haskell.
@@
 expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
-is transformed in STG that is structurally identical to the STG code that
+is transformed into STG that is structurally identical to the STG code that
 precedes it.
@@
 \subsubsection{Fancy Pattern Matching.}
-{\color{red}Implemented, not documented\dots}
+
+STG supports pattern matching but only in a simple form and only in
+\texttt{case} expressions.  mini-Haskell extends this to something approaching
+what is found in Haskell.  Functions, including lambda expressions, may match on
+all their arguments. Named functions may have multiple definitions to exhaust
+the combinations of patterns in their arguments and patterns in all contexts
+(functions and case expressions) may be nested arbitrarily deep.  This feature
+greatly simplifies the task of definining more complex behavior.  Without it,
+each matched parameter of a function requires an additional level of nested
+\texttt{case} expresssions. The same is true of each level of nesting within a
+single pattern.
+
+Even relatively simple functions can be painful to write, such as \texttt{unzip}.
+
+\begin{verbatim}
+unzip = FUN(list ->
+  case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
+                 Cons x xs ->
+                   case x of {
+                     TP2 a b ->     --TP2 is the Tuple constructor
+                       case unzip xs of {
+                         TP2 as bs ->
+                           let {aas = CON(Cons a as);
+                                bbs = CON(Cons b bs);
+                                res = CON(TP2 aas bbs)}
+                           in res; }}});
+\end{verbatim}
+
+In mini-Haskell, this becomes
+
+\begin{verbatim}
+unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
+unzip Nil = TP2 Nil Nil
+unzip (Cons (TP2 a b) xs) =
+  case unzip xs of
+    TP2 as bs -> TP2 (Cons a as) (Cons b bs)
+\end{verbatim}
+
+Of course, this complex pattern matching must eventually be transformed into the
+primitive, nested case expressions. Luckily, there are well-documented processes
+for achieving this, and Philip Wadler's chapter in \emph{The Implementation of
+  Functional Programming Languages} \cite{Jones:1987} % doing this right?
+on the compilation of this type of pattern matching proved invaluable in
+bringing it to mini-Haskell.  Wadler describes the \texttt{match} function that
+that, in its simplest form, produces a correct, but potentially large
+translation with some redundancy.  This function was implemented with only minor
+modifications to fit the shape of mini-Haskell's abstract form.  The
+implementation is augmented by some light flow-analysis to eliminate some of the
+redundancy and make the resulting code easier to read.  There are further
+optimizations suggested by Wadler to address the size of the generated code from
+this process (for example, when multiple branches lead to the same large
+expression), but the additional complexity outweighs the benefit for a testing
+language.
 
 \subsubsection{Less significant features}
+
+mini-Haskell provides some smaller conveniences that don't warrant their own
+section, but are worth mentioning.
 
@@
 \item Constructors as functions \\
-  (partial application uses generated functions, full creates CONS)
+  As in Haskell, constructors are first class, acting as true functions.  This
+  is implemented by simply generating a function that wraps the appropriate
+  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
+  constructor generates a call to this function.  Fully applied constructors
+  generate local \texttt{let}-bound objects to avoid an unnecessary function
+  call.
 
 \item Primitive operations as functions \\
-  (partial application uses generated functions, full creates EPrimop)
+  Similar to constructors, primitive operations are given first-class status as
+  functions, with roughly the same rules.  Partially applied calls are
+  translated to generated functions that wrap the actual primitive
+  operations. Fully applied calls generate true primitive operation calls.
 
 \item Lambda Expressions \\
-  (desugar to let-bound FUN objects)
+  mini-Haskell supports anonymous lambda expressions for the simple, one-off
+  functions that often find a place in Haskell code.  This was certainly the
+  simplest feature to add, requiring only that the function be given a unique
+  identifier to bind it to a \texttt{FUN} object in STG.
 
 \item Type Annotations \\
-  (passed on to type checker)
-
+  The ability to assert the type of some variable is often valuable as a
+  sanity-check and provides documentation that the compiler can verify.
+  mini-Haskell's type annotations (or signatures) behave similarly to those in
+  Haskell, being passed to the typechecker as initial assumptions.
+
+% I don't think I've described this very well at all... - david
 \item Haskell Block \\
-  (ignored by MHS front-end, not other Haskell compilers;
-  allows the same file to be tested on our compiler vs GHC)
+  Being a subset of Haskell, it may be desireable to test a mini-Haskell program
+  compiled through our toolchain against the same program (and same source file)
+  compiled with another Haskell compiler.  Because mini-Haskell exposes a
+  different set of primitives, the Haskell Block was introduced, to allow the
+  inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
+  primitives (e.g. \texttt{(+)} and \texttt{(-)}) to the names given in
+  mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored
+  by the mini-Haskell compiler, and arbitrary ``real'' Haskell code can be
+  placed therein to create the same mini-Haskell environment for full Haskell
+  compilers.
+
 \end{itemize}
@@
 
-%Mutual exclusion for garbage collector--heap check with every heap allocation.
-%In fact garbage collector can be invoked at any point in time except when
-%populating a heap object immediately after its allocation.
+Mutual exclusion for garbage collector--heap check with every heap allocation.
+In fact garbage collector can be invoked at any point in time except when
+populating a heap object immediately after its allocation.
 
@@
 enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
+types as described by Peyton-Jones and Launchbury~\cite{Jones91unboxedvalues},
 but for unboxed types code generation and runtime support is currently limited
@@
 
-Various papers describing some of the inner workings of GHC at various points
-in its evolution have greatly informed our approach and these have all been
+Various papers describing some of the inner workings of GHC at various point
+of its evolution have greatly informed our approach and these have all been
 previously cited.
@@
 A plug-in for GHC (prior to 8.0) exists for making functions, but not data
-structures, strict using a transformation on GHC Core that recursively expands
-non-recursive let bindings into case bindings.~\cite{strict-ghc-plugin}.  Thus
-it seems that while it makes functions denotationally strict, function
-arguments are still evaluated after function call.
+structures, strict.~\cite{strict-ghc-plugin}.  The transformation is on GHC
+Core, recursively expand non-recursive let bindings into case bindings.  Thus
+it seems that while it makes functions denotationally strict, function arguments
+are still evaluated after function call.
 %
@@
 built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
-\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
+\emph{scan}).~\cite{NVIDIA:2013} Three back-ends were implemented: for
 sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
@@
 The Intel Labs Haskell Research Compiler (HRC) is the most closely related
-work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
-planned approach they use GHC as the front end, but exit the GHC toolchain
+work to ours of which we are aware.~\cite{Liu:2013,Petersen:2013} As in our
+(planned) approach they use GHC as the front end, but exit the GHC toolchain
 at the final GHC Core stage, just before transformation to STG.  The essence
@@
 
-Don Stewart described an implementation wherein GHC was used as a Haskell compiler
-front-end, dumping STG code as text, parsing that text, then translating to
-code for execution on the Java Virtual Machine.~\cite{Stewart-BSc}.
+In his BSc thesis, Don Stewart described an implementation wherein GHC was
+used as a Haskell compiler front-end, dumping STG code as text, parsing that
+text, then translating to code for execution on the Java Virtual
+Machine.~\cite{Stewart-BSc}.
 
@@
 distributed Haskell based on distributed exception
-handling~\cite{Pointon:2001}.  However, it appears that their scheme for
+handling.~\cite{Pointon:2001}.  However, it appears that their scheme for
 handling failed processing elements correctly~\cite{} may have never been
@@
 Currently we are developing the parallel runtime.  This summer (2016) a
-returning student will bridge the gap between GHC STG (or Core) and our STG,
-and augment our runtime as needed in this effort.  Just as in the Intel
-Haskell Research Compiler effort, we will not attempt to reuse the GHC
-runtime.
+returning student will (partially) bridge the gap between GHC STG and our STG,
+and augment our runtime as needed in this effort.  Just as in the Intel Haskell
+Research Compiler effort, we will not
 

### 2016-04-06 pubs/tfp16/paper.tex

--- 2016-04-05 pubs/tfp16/paper.tex
@@
 \item Constructors as functions \\
-  (partial application uses generated functions, full creates CONS)
+  As in Haskell, constructors are first class, acting as true functions.  This
+  is implemented by simply generating a function that wraps the appropriate
+  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
+  constructor generates a call to this function.  Fully applied constructors
+  generate local \texttt{let}-bound objects to avoid an unnecessary function
+  call.
 
 \item Primitive operations as functions \\
-  (partial application uses generated functions, full creates EPrimop)
+  Similar to constructors, primitive operations are given first-class status as
+  functions, with roughly the same rules.  Partially applied calls are
+  translated to generated functions that wrap the actual primitive
+  operations. Fully applied calls generate true primitive operation calls.
 
 \item Lambda Expressions \\
-  (desugar to let-bound FUN objects)
+  mini-Haskell supports anonymous lambda expressions for the simple, one-off
+  functions that often find a place in Haskell code.  This was certainly the
+  simplest feature to add, requiring only that the function be given a unique
+  identifier to bind it to a \texttt{FUN} object in STG.
 
 \item Type Annotations \\
-  (passed on to type checker)
-
+  The ability to assert the type of some variable is often valuable as a
+  sanity-check and provides documentation that the compiler can verify.
+  mini-Haskell's type annotations (or signatures) behave like those in
+  Haskell, being passed to the typechecker as initial assumptions.
+
+% I don't think I've described this very well at all... - david
 \item Haskell Block \\
-  (ignored by MHS front-end, not other Haskell compilers;
-  allows the same file to be tested on our compiler vs GHC)
+  Being a subset of Haskell, it may be desireable to test a mini-Haskell program
+  compiled through our toolchain against the same program (and same source file)
+  compiled with another Haskell compiler.  Because mini-Haskell exposes a
+  different set of primitives, the Haskell Block was introduced, to allow the
+  inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
+  primitives (e.g. \texttt{(+)} and \texttt{(-)}) to the names given in
+  mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored
+  by the mini-Haskell compiler, and arbitrary ``real'' Haskell code can be
+  placed therein to create the same mini-Haskell environment for full Haskell
+  compilers.
+
 \end{itemize}

### 2016-04-06 pubs/tfp16/paper.tex
formatting tweaks

@@
 
-\subsubsection{Proper tail calls in C}
+\subsubsection{Proper tail calls in C.}
 Parallelism is achieved by multi-threading; our model is based on Harris et
@@
 
-\subsubsection{The Layout Rule}
+\subsubsection{The layout rule.}
 Mini-Haskell implements Haskell's layout rule in an almost direct translation
@@
 
-\subsubsection{Expression to Expression application}
+\subsubsection{Expression to expression application.}
 
@@
 
-\subsubsection{Fancy Pattern Matching.}
+\subsubsection{Fancy pattern matching.}
 
@@
 
-\subsubsection{Less significant features}
-
-mini-Haskell provides some smaller conveniences that don't warrant their own
+\subsubsection{Less significant features.}
+
+Mini-Haskell provides some smaller conveniences that don't warrant their own
 section, but are worth mentioning.

### 2016-04-06 pubs/tfp16/paper.tex
reorganize

@@
 a data object.  More generally, the operational notion of allocating a heap
-object is made explicit: this is the ``side effect'' of \texttt{let} bindings.
+object is explicit: this is the ``side effect'' of \texttt{let} bindings.
 As such a variable of unboxed type cannot be let-bound; instead it can be
@@
 where the \texttt{\#} suffix is merely a convention to signify that
-\texttt{x\#} has unboxed type.  Finally, deferred evaluation is made explicit:
+\texttt{x\#} has unboxed type.  Finally, deferred evaluation is explicit:
 a deferred expression evaluation is encoded by a thunk.
@@
 This is straightforwardly realizable because the stack is explicit, and with
-the C compiler call optimization, the continuation style provides a proper
+modern C compiler call optimization, the continuation style enables a proper
 \emph{goto} mechanism where the \emph{code labels}---entities that name blocks
@@
 arrange that the stack frame for \emph{stgApply} is reused, updating only
-\emph{index}, so except for the indirect jumps on return this is essentially a
-loop.  For the cases of arguments in excess of the arity of the object being
+\emph{index}, so this is realized as a machine-level loop using an indirect
+jump.  For the cases of arguments in excess of the arity of the object being
 applied, in effect \emph{stgApply}'s argument frame is popped and two
@@
 
-\subsection{Mini-Haskell}
+\subsection{Type system}
+
+We perform standard Hindley-Milner type inference and
+enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
+types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
+but for unboxed types code generation and runtime support is currently limited
+to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
+\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
+enumerations, e.g., \texttt{data unboxed Bool\# = False\# | True\#}.
+
+\subsection{Parallel evaluation and other developments}
+
+Parallel runtime is being implemented following the ideas
+of~\cite{Harris:2005,Marlow:2008,Marlow:2009}, thread pool, stack and heap 
+for each thread, blah, blah.
+
+For ease of developing the test suite we have implemented a \emph{mini-Haskell}
+front end as detailed in Appendix~A.
+
+We use a continuous integration system augmented with logic to categorize
+programs in a large and growing test suite according to which degrees of non-strictness they
+require to terminate correctly, and to provide other logging and debugging
+information.  This is detailed in Appendix~B.
+
+\section{Current Status, Future Work}
+
+Developing the bridge between GHC Core or STG to our system is scheduled
+to start immediately after TiFP'16.
+
+At the time of writing we have a complete serial implementation, including
+\begin{itemize}
+\item STG parsers \emph{(by student)},
+\item mini-Haskell parser, mini-Haskell to STG transformer \emph{(by student)},
+\item type inferencer,
+\item various analysis and transformation passes \emph{(some by students)},
+\item code generator \emph{(some student contribution)},
+\item runtime system including a garbage collector,
+\item continuous integration and testing system,
+\item extensive test suite \emph{(mostly by students)}.
+\end{itemize}
+
+We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
+small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. We have
+not worked on the development of a more sophisticated garbage collector
+because this would be a distraction from our main goals.
+
+%Mutual exclusion for garbage collector--heap check with every heap allocation.
+%In fact garbage collector can be invoked at any point in time except when
+%populating a heap object immediately after its allocation.
+
+An almost unbounded amount of interesting work suggests itself; one such path
+would lead to the recreation of GHC.  However, given the minuscule budget (and
+thus the emphasis on student labor) we must maintain sharp focus on the main
+goals:  implementing and evaluating mechanisms for automatic parallelism and
+fault tolerance.
+
+Currently we are developing the parallel runtime.  This summer (2016) a
+returning student will bridge the gap between GHC STG (or Core) and our STG,
+and augment our runtime as needed in this effort.  Just as in the Intel
+Haskell Research Compiler effort, we will not attempt to reuse the GHC
+runtime.
+
+\section{Related Work}
+
+Various papers describing some of the inner workings of GHC at various points
+in its evolution have greatly informed our approach and these have all been
+previously cited.
+
+\subsubsection{Strict but still pure.}
+The concept of strict but still pure dates back to early studies of the lambda
+calculus.  In terms of a notionally concrete programming language the first
+thorough explications may be of Backus' FP~\cite{Backus:1978} and follow-on
+FL implementation~\cite{AikenFL,FLreport89}.
+%
+Perhaps the first explication of the potential merits of a strict-by-default
+pure functional language over a non-strict one is by Sheard, who also provided
+a very simple prototype implementation~\cite{Sheard:2003}.
+%
+A plug-in for GHC (prior to 8.0) exists for making functions, but not data
+structures, strict using a transformation on GHC Core that recursively expands
+non-recursive let bindings into case bindings~\cite{strict-ghc-plugin}.  Thus
+it seems that while it makes functions denotationally strict, function
+arguments are still evaluated after function call.
+%
+The Disciplined Disciple Compiler project appears to intend implementation of
+a strict, higher order functional language (or family of languages) compiler
+that has been underway for some time, but the emphasis appears to be on a much
+more complex type system for handling stateful computation, and the prospects
+for its eventual completion are unknown~\cite{disciplined-disciple}.
+%
+NVIDIA's NOVA language, and corresponding compiler, implements a strict
+(``call by value''), pure, polymorphic, higher-order functional language with
+built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
+\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
+sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
+targeting NVIDIA GPUs.
+
+\subsubsection{GHC as a front-end.}
+The Intel Labs Haskell Research Compiler (HRC) is the most closely related
+work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
+planned approach they use GHC as the front end, but exit the GHC toolchain
+at the final GHC Core stage, just before transformation to STG.  The essence
+of their effort is to target an existing, highly optimizing, Intel functional
+language compiler that is ``largely language agnostic'' and performs (among
+many other optimization) SIMD vectorization.
+
+Don Stewart described an implementation wherein GHC was used as a Haskell compiler
+front-end, dumping STG code as text, parsing that text, then translating to
+code for execution on the Java Virtual Machine~\cite{Stewart-BSc}.
+
+\subsubsection{Fault tolerance for functional programming.}
+Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
+distributed Haskell based on distributed exception
+handling~\cite{Pointon:2001}.  However, it appears that their scheme for
+handling failed processing elements correctly~\cite{} may have never been
+implemented.
+
+Recently, Robert Stewart extended Haskell Distributed Parallel Haskell
+(HdpH)~\cite{hdph} to be fault resilient at an impressive scale---1400 cores
+on a distributed memory system.  Here the key recovery strategy is task
+replication~\cite{Stewart:2013}.
+
+\section{Acknowledgments}
+
+Funding for this project has been provided by the DOE NNSA LANL Laboratory
+Directed Research and Development program award 20150845ER; the National
+Science Foundation Science, Technology, Engineering, and Mathematics Talent
+Expansion Program (STEP) program for undergraduate students; and the DOE
+Science Undergraduate Laboratory Internship (SULI) program.
+%
+Los Alamos National Laboratory is managed and operated by Los Alamos National
+Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
+Department of Energy's National Nuclear Security Administration (NNSA).
+
+%
+% ---- Appendix ----
+%
+
+\appendix
+
+\section{Mini-Haskell}
+
 As mentioned, a current near-term goal is to use GHC as our front-end to
@@
 it in terms of the differences between the two and how mini-Haskell is
-translated into STG for the backend.  Here, the differences are given in order
+translated into STG for the back-end.  Here, the differences are given in order
 of perceived utility, or importance to someone transitioning from STG to
@@
 of the specification given in section 10.3 of the Haskell 2010
-Report.~\cite{haskell2010report} The notable exception is the omission of any
+Report~\cite{haskell2010report}. The notable exception is the omission of any
 attempt to define the \texttt{parse-error} function.  This is apparently a
@@
 for the use of Haskell programming environments that support ``smart''
-indentation, e.g., \emph{haskell-mode} for Emacs.~\cite{haskellmode}.
+indentation, e.g., \emph{haskell-mode} for Emacs~\cite{haskellmode}.
 
@@
 (functions and case expressions) may be nested arbitrarily deep.  This feature
-greatly simplifies the task of definining more complex behavior.  Without it,
+greatly simplifies the task of defining more complex behavior.  Without it,
 each matched parameter of a function requires an additional level of nested
-\texttt{case} expresssions. The same is true of each level of nesting within a
+\texttt{case} expressions. The same is true of each level of nesting within a
 single pattern.
@@
   mini-Haskell's type annotations (or signatures) behave like those in
-  Haskell, being passed to the typechecker as initial assumptions.
+  Haskell, being passed to the type checker as initial assumptions.
 
@@
 \item Haskell Block \\
-  Being a subset of Haskell, it may be desireable to test a mini-Haskell program
+  Being a subset of Haskell, it may be desirable to test a mini-Haskell program
   compiled through our toolchain against the same program (and same source file)
@@
 
-\section{Current Status, and Expected Results by the time of TiFP'16}
-
-Developing the bridge between GHC Core or STG to our system is scheduled
-to start immediately after TiFP'16.
-
-At the time of writing we have a complete serial implementation, including
-\begin{itemize}
-\item STG and mini-Haskell (inc. ADT w/unboxed) parsers \emph{(both by students)},
-\item mini-Haskell to STG transformer \emph{(by student)},
-\item type inferencer,
-\item various analysis and transformation passes \emph{(most by students)},
-\item code generator \emph{(some student contribution)},
-\item runtime system including a garbage collector,
-\item extensive test suite \emph{(mostly by students)}.
-\end{itemize}
-
-We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
-small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. We have
-not worked on the development of a more sophisticated garbage collector
-because this would be a distraction from our main goals.
-
-%Mutual exclusion for garbage collector--heap check with every heap allocation.
-%In fact garbage collector can be invoked at any point in time except when
-%populating a heap object immediately after its allocation.
-
-\subsection{Type system}
-
-We perform standard Hindley-Milner type inference and
-enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
-but for unboxed types code generation and runtime support is currently limited
-to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
-\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
-enumerations, e.g., \texttt{data unboxed Bool\# = False\# | True\#}.
-
-\subsection{Test suite and testing}
+\section{Test suite and testing}
 
 We have developed a extensive test suite which uses the CMake CTest
-tool.~\cite{cmake-book} The full test suite is automatically run, and any
+tool~\cite{cmake-book}. The full test suite is automatically run, and any
 failures reported, after every commit to the Git revision control system. We
 use a Jenkins continuous integration server to automate this build and testing
-process.~\cite{jenkins-book} The Git repository is hosted externally so our
+process~\cite{jenkins-book}. The Git repository is hosted externally so our
 student collaborators can contribute at their leisure.
@@
 
-\subsection{Parallel evaluation}
-
-Parallel runtime is being implemented following the ideas
-of~\cite{Harris:2005,Marlow:2008,Marlow:2009}, thread pool, stack and heap 
-for each thread, blah, blah.
-
-\section{Related Work}
-
-Various papers describing some of the inner workings of GHC at various points
-in its evolution have greatly informed our approach and these have all been
-previously cited.
-
-\subsubsection{\emph{Strict but still pure} is not new.}
-Very simple implementations have been prototyped, for example
-Sheard's.~\cite{Sheard:2003}
-%
-A plug-in for GHC (prior to 8.0) exists for making functions, but not data
-structures, strict using a transformation on GHC Core that recursively expands
-non-recursive let bindings into case bindings.~\cite{strict-ghc-plugin}.  Thus
-it seems that while it makes functions denotationally strict, function
-arguments are still evaluated after function call.
-%
-The Disciplined Disciple Compiler project appears to intend implementation of
-a strict, higher order functional language (or family of languages) compiler
-that has been underway for some time, but the emphasis appears to be on a much
-more complex type system for handling stateful computation, and the prospects
-for its eventual completion are unknown~\cite{disciplined-disciple}.
-%
-NVIDIA's NOVA language, and corresponding compiler, implements a strict
-(``call by value''), pure, polymorphic, higher-order functional language with
-built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
-\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
-sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
-targeting NVIDIA GPUs.
-
-\subsubsection{GHC as a front-end.}
-The Intel Labs Haskell Research Compiler (HRC) is the most closely related
-work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
-planned approach they use GHC as the front end, but exit the GHC toolchain
-at the final GHC Core stage, just before transformation to STG.  The essence
-of their effort is to target an existing, highly optimizing, Intel functional
-language compiler that is ``largely language agnostic'' and performs (among
-many other optimization) SIMD vectorization.
-
-Don Stewart described an implementation wherein GHC was used as a Haskell compiler
-front-end, dumping STG code as text, parsing that text, then translating to
-code for execution on the Java Virtual Machine.~\cite{Stewart-BSc}.
-
-\subsubsection{Fault tolerance for functional programming.}
-Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
-distributed Haskell based on distributed exception
-handling~\cite{Pointon:2001}.  However, it appears that their scheme for
-handling failed processing elements correctly~\cite{} may have never been
-implemented.
-
-Recently, Robert Stewart extended Haskell Distributed Parallel Haskell
-(HdpH)~\cite{hdph} to be fault resilient at an impressive scale---1400 cores
-on a distributed memory system.  Here the key recovery strategy is task
-replication.~\cite{Stewart:2013}.
-
-\section{Future Work}
-
-An almost unbounded amount of interesting work suggests itself; one such path
-would lead to the recreation of GHC.  However, given the minuscule budget (and
-thus the emphasis on student labor) we must maintain sharp focus on the main
-goals:  implementing and evaluating mechanisms for automatic parallelism and
-fault tolerance.
-
-Currently we are developing the parallel runtime.  This summer (2016) a
-returning student will bridge the gap between GHC STG (or Core) and our STG,
-and augment our runtime as needed in this effort.  Just as in the Intel
-Haskell Research Compiler effort, we will not attempt to reuse the GHC
-runtime.
-
-\section{Acknowledgments}
-
-Funding for this project has been provided by the DOE NNSA LANL Laboratory
-Directed Research and Development program award 20150845ER; the National
-Science Foundation Science, Technology, Engineering, and Mathematics Talent
-Expansion Program (STEP) program for undergraduate students; and the DOE
-Science Undergraduate Laboratory Internship (SULI) program.
-%
-Los Alamos National Laboratory is managed and operated by Los Alamos National
-Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
-Department of Energy's National Nuclear Security Administration (NNSA).
-
 %

### 2016-04-06 pubs/tfp16/paper.tex
safely under the page limit without really losing anything

@@
 member of the HPC community recently stated publicly that ``Fortran is
-essential for exascale programming.''~\cite{Heroux16} Existing US government
+essential for exascale programming''~\cite{Heroux16}.  Existing US government
 laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
@@
 poorly abstracted, especially at the thread (shared-memory) and computational
-accelerator (e.g., GPU) levels.  Thread-level parallel efficiency can be good
+accelerator (e.g.\ GPU) levels.  Thread-level parallel efficiency can be good
 when the parallel model is bulk synchronous; tasking models are generally
@@
 by the Workshop on Functional High Performance Computing, held in conjunction
-with the International Conference on Functional Programming, being now in its fifth
+with the International Conference on Functional Programming, now in its fifth
 consecutive year.  A more precise characterization of the trend is that there
@@
 must not access non-constant global data.  Thus the inputs and outputs of
-tasks are exactly known to the runtime system.  In serial execution these access
-requirements are notionally superfluous, but in parallel execution the runtime
-can dynamically calculate the data dependency graph and relax the
+tasks are exactly known to the runtime system.  In serial execution these
+access requirements are notionally superfluous, but in parallel execution the
+runtime can dynamically calculate the data dependency graph and relax the
 programmatically defined serial ordering of task execution.  Very high
 parallel efficiency has been demonstrated on extremely large computing
-systems~\cite{Bauer14}.\footnote{It seems
-  significant that the PI of the Legion project was previously a co-developer
-  of a strict, arguably pure functional language
+systems~\cite{Bauer14}.\footnote{We note that the PI of the Legion project was
+  previously a co-developer of a strict, arguably pure functional language
   implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
@@
 presented with the most highly parallel (and thereby performant)
-implementation of a pure functional language to date~\cite{Davis96}, and
-collectively had no conceptual understanding of what it was, much less why it
-might be of some import, there is now a small number of laboratory HPC
-practitioners who are developing an appreciation for pure functional
-programming in Haskell.  Still, they see their needs as different than other
-Haskell users:  performance is paramount, vectors (``Efficient
-Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
-non-strictness or laziness is more of a nuisance---even a serious
-hindrance---than a convenience, much less ever algorithmically essential in
-the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
-grammars~\cite{Johnsson87}, or other instances of \emph{tying the
-  knot}~\cite{tying-the-knot}.
+implementation of a pure functional language to date~\cite{Davis96}, to no
+effect whatsoever, there is now a small number of laboratory HPC practitioners
+who are developing an appreciation for pure functional programming in Haskell.
+However, they see their needs as different than other Haskell users: performance
+is paramount, vectors~\cite{vector} are a (perhaps \emph{the}) primary data
+structure, and non-strictness or laziness is more of a nuisance---even a
+serious hindrance---than a convenience, much less ever algorithmically
+essential in the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's
+general attribute grammars~\cite{Johnsson87}, or other instances of
+\emph{tying the knot}~\cite{tying-the-knot}.
 
@@
 with semantics that are lazy (or non-strict) by default, with the most telling
-recent concrete evidence being the push for, and realization of, the
-\texttt{Strict} and \texttt{StrictData} language extensions newly available in
-GHC 8.0.  Quoting the documentation:
+recent concrete evidence being the push for the \texttt{Strict} and
+\texttt{StrictData} language extensions newly available in GHC 8.0.  Quoting
+the documentation:
 \begin{quote}
@@
 explicit deallocation statement (C \emph{free}, etc.) is difficult to
-understand and not something they particularly want to try to understand.
+understand and not something they care to need to understand.
 %Hence, we eschew explicit strictness annotations for varying uniform degrees
@@
 been some success with transparent (invisible to the programmer) systems.
-Wiring checkpoint/restart into a large-scale application is not only burdensome and
-error-prone, it can severely warp the engineering of the application
-because it must be designed such that essential state can be readily captured.
-Furthermore, there is a concern that the scalability of checkpoint/restart is
-reaching its end.
+Wiring checkpoint/restart into a large-scale application is not only
+burdensome and error-prone, it can severely warp the engineering of the
+application because it must be designed such that consistent, essential state
+can be readily captured.  Furthermore, there is a concern that the scalability
+of checkpoint/restart is reaching its end as data transfer to backing store
+becomes an increasing bottleneck.
 
@@
 system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
-been recognized that the pure parts of a functional program (excluding, e.g.,
-the I/O monad in Haskell) could be safely restarted.  The recent work of
-Stewart impressively demonstrates this, in a pure functional setting, on a
-large-scale distributed-memory system~\cite{Stewart:2013}.
+been recognized that the side-effect-free parts of a functional program
+(excluding, e.g., the I/O monad in Haskell) could be safely replicated or
+restarted.  The recent work of Stewart impressively demonstrates this, in a
+pure functional setting, on a large-scale distributed-memory
+system~\cite{Stewart:2013}.
 
@@
 For feasibility, we restrict ourselves to shared-memory implicit parallelism.
-The requirement for tail calling is not primarily for efficiency, rather it
-is for a reasonable parallel implementation.  Achieving this, and thereby
-sufficiently limiting the use of the native C stack, makes this an interesting
-exercise that we will discuss at some length.
+The requirement for tail calling is not primarily for efficiency, rather it is
+for a reasonable parallel implementation.  Achieving this, and thereby
+sufficiently limiting the use of the native C stack, will be discussed at some
+length.
 
@@
 indeed GHC is part of our master plan.  In brief, the phases of the GHC
-compiler are 1) parsing and type checking; 2) transformation to the GHC
-\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; 3)
-analysis and transformation on Core; 4) transformation of Core to the
-\emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
+compiler are (1) parsing and type checking; (2) transformation to the GHC
+\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; (3)
+analysis and transformation on Core; (4) transformation of Core to the
+\emph{STG} intermediate representation~\cite{PJ-stockhardware}; (5)
 transformation of STG to machine code via various imperative
@@
 are various possible ways to accomplish the this: using GHC's option to dump
-STG or Core as text, hacking GHC itself and, optionally, using GHC as a
-library~\cite{ghc-as-library}.  For now we start with our own STG, with
-bridging the gap with GHC scheduled to commence quite soon.
+STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
+library~\cite{ghc-as-library}.  Initially we start with our own STG.
+%, with bridging the gap with GHC scheduled to commence quite soon.
 
@@
 \emph{EA}. Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}),
-i.e., a thunk being evaluated, are only runtime entities and not actually part
-of our STG language.
+signifying a \emph{THUNK} being evaluated, are only runtime entities and not
+actually part of our STG language.
 
@@
 a data object.  More generally, the operational notion of allocating a heap
-object is explicit: this is the ``side effect'' of \texttt{let} bindings.
-As such a variable of unboxed type cannot be let-bound; instead it can be
-bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
-%\begin{verbatim}
-%    case e of x# {
-%      _ -> e
-%    }
-%\end{verbatim}
-where the \texttt{\#} suffix is merely a convention to signify that
-\texttt{x\#} has unboxed type.  Finally, deferred evaluation is explicit:
-a deferred expression evaluation is encoded by a thunk.
+object is explicit: this is the operational semantics of \texttt{let} bindings.
+%As such a variable of unboxed type cannot be let-bound; instead it can be
+%bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
+%where the \texttt{\#} suffix is merely a convention to signify that
+%\texttt{x\#} has unboxed type.  
+Finally, deferred evaluation is explicit:
+a deferred expression evaluation is encoded by a \emph{THUNK}.
 
@@
 
-Greater strictness implies more opportunities for parallelism, possible losses
-or gains in space efficiency, and less expressiveness.  We seek to explore
-the interplay of all of these.
-
-GHC 8.0 gives three possibilities: conventional non-strict, constructor
-strict, and function- and constructor strict, but there are possible
-variations.  The obvious variation is function, but not constructor,
-strictness.  In Haskell the distinction between constructors and functions is
-somewhat blurred because constructors behave like functions (outside of
-pattern matching), but in STG the distinction is clear.
+Greater strictness implies more opportunities for automatic parallelism,
+possible losses or gains in space efficiency, and less expressiveness.  We
+seek to explore the interplay of all of these.
+
+GHC 8.0 gives three defaults: conventional non-strict, constructor strict, and
+function- and constructor strict, but there are possible variations.  The
+obvious variation is function, but not constructor, strictness.  In Haskell
+the distinction between constructors and functions is somewhat blurred
+%because constructors behave like functions 
+(outside of data definitions and pattern matching), but in STG the distinction
+is clear.
 

### 2016-04-07 pubs/tfp16/paper.tex
I think we're getting close

@@
 
+\setlength{\tabcolsep}{5pt}
+\begin{figure}
+\centering
+%\begin{centering}
+\footnotesize % tiny scriptsize footnotesize small
+\begin{tabular}{r r c l l}
+Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
+Constructor  & $C$            &     &                                              & Initial upper-case letter \\
+Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
+Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
+\\
+Expression   & $e$            & ::= & $a$                                          & Atom \\
+             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
+             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
+
+             &                & $|$ & \texttt{let \{}                              & Recursive let, $n\ge 1$ \\
+             &                &     & \texttt{ } $\mathit{odecl}_1$ \texttt{;} \\
+             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
+             &                &     & \texttt{ } $\mathit{odecl}_n$ \texttt{;} \\
+%             &                &     & \texttt{ } $x_1$ \texttt{=} $\mathit{obj}_1$ \texttt{;} \\
+%             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
+%             &                &     & \texttt{ } $x_n$ \texttt{=} $\mathit{obj}_n$ \\
+             &                &     & \texttt{\} in} $e$  \\
+
+             &                & $|$ & \texttt{case} $e$ \texttt{of} $x$ \texttt{\{}  & Case expression, $n \ge 1$\\
+             &                &     & \texttt{ } $\mathit{alt}_1$ \texttt{;} \\
+             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
+             &                &     & \texttt{ } $\mathit{alt}_n$ \texttt{\}} \\
+%            &                &     & \texttt{\}} \\
+\\
+Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
+             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
+\\
+Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
+             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
+             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
+             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
+             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
+\\
+Object decl. & $\mathit{odecl}$ &     & $x = \mathit{obj}$                           & Simple binding \\
+\\
+Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$ & $i \ge 0$ \\
+\\
+Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} [\texttt{unboxed}] & User-defined data type  \\
+               &                   &     & $C\ x_i =$                       & $i \ge 0, n > 0$         \\
+               &                   &     & $\mathit{con}_1 | \dots |  \mathit{con}_n$ \\
+
+\\
+Program      & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_1$ & Object and data defns \\
+             &                 &     & \texttt{ } $\dots$ \texttt{;} \\
+             &                 &     & $\mathit{(o|d)decl}_n$ & 
+%Program      & $\mathit{prog}$& ::= & $f_1\ =\ \mathit{obj}_1$ \texttt{;}          & $n \ge 1$, distinguished \texttt{main}\\
+%             &                &     & \texttt{ } $\dots$ \texttt{;} \\
+%             &                &     & $f_n\ =\ \mathit{obj}_n$
+
+\end{tabular}
+%\end{centering}
+\caption{STG syntax}
+\label{fig:STGsyntax}
+\end{figure}
+
 Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
@@
 actually part of our STG language.
-
-\setlength{\tabcolsep}{5pt}
-\begin{figure}
-%\centering
-\begin{centering}
-\footnotesize % tiny scriptsize footnotesize small
-\begin{tabular}{r r c l l}
-
-Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
-Constructor  & $C$            &     &                                              & Initial upper-case letter \\
-Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
-Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
-\\
-Expression   & $e$            & ::= & $a$                                          & Atom \\
-             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
-             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
-
-             &                & $|$ & \texttt{let \{}                              & Recursive let, $n\ge 1$ \\
-             &                &     & \texttt{ } $x_1$ \texttt{=} $\mathit{obj}_1$ \texttt{;} \\
-             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
-             &                &     & \texttt{ } $x_n$ \texttt{=} $\mathit{obj}_n$ \\
-             &                &     & \texttt{\} in} $e$  \\
-
-             &                & $|$ & \texttt{case} $e$ \texttt{of} $x$ \texttt{\{}  & Case expression, $n \ge 1$\\
-             &                &     & \texttt{ } $\mathit{alt}_1$ \texttt{;} \\
-             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
-             &                &     & \texttt{ } $\mathit{alt}_n$ \\
-             &                &     & \texttt{\}} \\
-\\
-Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
-             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
-\\
-Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
-             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
-             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
-             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
-             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
-\\
-Program      & $\mathit{prog}$& ::= & $f_1\ =\ \mathit{obj}_1$ \texttt{;}          & $n \ge 1$, distinguished \texttt{main}\\
-             &                &     & \texttt{ } $\dots$ \texttt{;} \\
-             &                &     & $f_n\ =\ \mathit{obj}_n$
-
-\end{tabular}
-\end{centering}
-\caption{STG syntax}
-\label{fig:STGsyntax}
-\end{figure}
 
@@
 evaluation of a C function (corresponding to some STG code) is completed its
-return is by jumping through the code pointer of the continuation on the top
-of the stack (``returning through the continuation stack'').  In the case of
-an argument frame for a function, the associated code simply pops the
+return is by an indirect jump using the code pointer in the continuation on
+the top of the stack (``returning through the continuation stack'').  In the
+case of an argument frame for a function, the associated code simply pops the
 continuation off the stack and returns through the next continuation on the
@@
 because arguments can be evaluated in a tight loop.  Unfortunately these uses
-of call-with-return can result in the C stack growing unboundedly.  Our
+of call-with-return can result in the C stack growing without bound.  Our
 solution uses a single C \emph{stgApply} function with no calls-with-return.
@@
 information.  These continuations are maintained on a stack.  When the
-evaluation of an object is completed its return is by jumping through the code
-pointer of the continuation on the top of the stack (``returning through the
-continuation stack'').  For example, the argument frame for a function is one
-type of continuation and its associated code simply pops the continuation off
-the stack and returns through the next continuation on the stack.
+evaluation of an object is completed its return is by an indirect jump using
+the code pointer in the continuation on the top of the stack (``returning
+through the continuation stack'').  For example, the argument frame for a
+function is one type of continuation and its associated code simply pops the
+continuation off the stack and returns through the next continuation on the
+stack.
 
@@
 
-We perform standard Hindley-Milner type inference and
-enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
-types as described by Peyton-Jones and Launchbury~\cite{Jones:1991},
-but for unboxed types code generation and runtime support is currently limited
-to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
-\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
-enumerations, e.g., \texttt{data unboxed Bool\# = False\# | True\#}.
-
-\subsection{Parallel evaluation and other developments}
-
-Parallel runtime is being implemented following the ideas
-of~\cite{Harris:2005,Marlow:2008,Marlow:2009}, thread pool, stack and heap 
-for each thread, blah, blah.
-
-For ease of developing the test suite we have implemented a \emph{mini-Haskell}
-front end as detailed in Appendix~A.
-
-We use a continuous integration system augmented with logic to categorize
-programs in a large and growing test suite according to which degrees of non-strictness they
-require to terminate correctly, and to provide other logging and debugging
-information.  This is detailed in Appendix~B.
-
-\section{Current Status, Future Work}
-
-Developing the bridge between GHC Core or STG to our system is scheduled
-to start immediately after TiFP'16.
-
-At the time of writing we have a complete serial implementation, including
+Type information is required for code generation.  For example, at the
+coarsest level it is necessary to know whether values are boxed, and so
+represented by pointers, or unboxed, to avoid expensive tagging.  Because we
+do not yet get type information directly from GHC we use a subset of Haskell
+syntax (e.g., no named fields) for defining algebraic data types, extended
+with built-in and user-defined unboxed types.  We perform standard
+Hindley-Milner type inference and enforcement following Heren et
+al.~\cite{Heren02}, again extended with unboxed types.  The syntax,
+restrictions, and extensions to the type inference algorithm for unboxed types
+follow Peyton Jones and Launchbury~\cite{Jones:1991}.
+
+\section{Current Status (April 8, 2016), Future Work}
+
+An almost unbounded amount of interesting development suggests itself; one
+such path would lead to the recreation of GHC.  However, given the minuscule
+budget (and therefore the employment of student interns) we must maintain
+sharp focus on the main goals: implementing and evaluating mechanisms for
+automatic parallelism and fault tolerance.  At the time of writing we have a
+fully functioning serial implementation including
 \begin{itemize}
-\item STG parsers \emph{(by student)},
+\item STG parser \emph{(by student)},
 \item mini-Haskell parser, mini-Haskell to STG transformer \emph{(by student)},
@@
 
-We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
-small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. We have
-not worked on the development of a more sophisticated garbage collector
-because this would be a distraction from our main goals.
-
-%Mutual exclusion for garbage collector--heap check with every heap allocation.
-%In fact garbage collector can be invoked at any point in time except when
-%populating a heap object immediately after its allocation.
-
-An almost unbounded amount of interesting work suggests itself; one such path
-would lead to the recreation of GHC.  However, given the minuscule budget (and
-thus the emphasis on student labor) we must maintain sharp focus on the main
-goals:  implementing and evaluating mechanisms for automatic parallelism and
-fault tolerance.
-
-Currently we are developing the parallel runtime.  This summer (2016) a
-returning student will bridge the gap between GHC STG (or Core) and our STG,
-and augment our runtime as needed in this effort.  Just as in the Intel
-Haskell Research Compiler effort, we will not attempt to reuse the GHC
-runtime.
+The \emph{mini-Haskell} front-end is detailed in Appendix~A.
+
+This summer (2016) a returning student will bridge the gap between GHC STG (or
+Core) and our STG, augmenting the runtime as needed.  Just as in the Intel
+Haskell Research Compiler effort we will not attempt to reuse the GHC runtime.
+
+For unboxed types code generation and runtime support is currently limited
+to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
+\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
+enumerations, e.g.,
+\begin{verbatim}
+     data unboxed Bool# = False# | True# .
+\end{verbatim}
+
+We use a continuous integration system augmented with logic to categorize
+programs in a large and growing test suite according to which degrees of
+non-strictness they require to terminate correctly, and to provide other
+logging and debugging information.  This is detailed in Appendix~B.
+
+% For various reasons we have chosen to make our implementation 64-bit only.
+
+We are currently developing the parallel runtime.  Parallel garbage collection
+is likely out of scope.
+
+%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
+%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}.
+
+%\cite{Harris:2005,Marlow:2008,Marlow:2009}
 
@@
 \subsubsection{Strict but still pure.}
-The concept of strict but still pure dates back to early studies of the lambda
-calculus.  In terms of a notionally concrete programming language the first
-thorough explications may be of Backus' FP~\cite{Backus:1978} and follow-on
-FL implementation~\cite{AikenFL,FLreport89}.
-%
-Perhaps the first explication of the potential merits of a strict-by-default
+The concept of strict but still pure in some sense dates back to the concept
+of applicative order reduction in the lambda calculus.  In terms of a
+notionally concrete programming language the first thorough explications may
+be Backus' FP~\cite{Backus:1978} and follow-on FL
+implementation~\cite{AikenFL,FLreport89}.
+%
+A notable argument for the potential merits of a strict-by-default
 pure functional language over a non-strict one is by Sheard, who also provided
@@
 handling~\cite{Pointon:2001}.  However, it appears that their scheme for
-handling failed processing elements correctly~\cite{} may have never been
-implemented.
+handling failed processing elements correctly may have
+never been implemented~\cite{Trinder:2000}.
 
@@
 \begin{tabular}{r r c l l}
-Variable         & $f,\ x$             &     &
-& Initial lower-case letter \\
-
-Constructor        & $C$                 &     & 
-& Initial upper-case letter \\
+Variable         & $f,\ x$             &     &  & Initial lower-case letter \\
+
+Constructor        & $C$                 &     & & Initial upper-case letter \\
 
 Literal            & $\mathit{lit}$      & ::= & $i\ |\ d\ 
-                                                 |\ i$\hash\ $|\ d$\hash
-& Integral or floating point literal (boxed) \\
-
-                   &                     &     & 
-& optional \# suffix denotes unboxed value \\
-
-Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$
-& Constructors are first class \\ 
-% not sure if that's the right terminology 
-\\
-
-Expression         & $e$                 & ::= & $a$
-& Atomic expression \\
-
-                   &                     & $|$ & $e\ e$
-& Expression application \\
+                                                 |\ i$\hash\ $|\ d$\hash & Integral or floating point literal \\
+
+                   &                     &     &  & \# suffix denotes unboxed value \\
+
+Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$ & Constructors are first class \\% not sure if that's the right terminology 
+\\
+
+Expression         & $e$                 & ::= & $a$ & Atomic expression \\
+
+                   &                     & $|$ & $e\ e$ & Expression application \\
 
                    &                     & $|$ & \texttt{\textbackslash} $\mathit{pat}_i$ 
-                                               \texttt{->} $e$
-& Lambda expression, $i > 0$ \\
-
-                   &                     & $|$ & \texttt{case} $e$ \texttt{of} $\mathit{alt}_i$
-& Case expression, $i > 0$\\
-
-                   &                     & $|$ & \texttt{let} $\mathit{odecl}_i$ \texttt{in} $e$
-& Recursive let expression, $i > 0$ \\
-
-\\
-
-Pattern            & $\mathit{pat}$      & ::= & $C\ \mathit{pat}_i$
-& Nested constructor matching \\
-                   &                     & $|$ & $\mathit{lit}$
-& Match numeric literals \\
-                   &                     & $|$ & $x$
-& Bind to variable \\
-
-\\
-
-Alternative        & $\mathit{alt}$      & ::= & $\mathit{pat}_i$ \texttt{->} $e$
-& $i > 0$ \\
-
-\\
-
-Type               & $\mathit{type}$     & $|$ & $C$
-& Concrete type (constructor naming) \\
-                   &                     & $|$ & $x$
-& polymorphic type variable \\
-                   &                     & $|$ & $\mathit{type}$ \texttt{->} $\mathit{type}$
-& Function type \\
-
-\\
-
-Object decl. & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$
-& Type signature \\
-                   &                     & $|$ & $x = e$
-& Simple binding \\
-                   &                     & $|$ & $f\ \mathit{pat}_i = e$
-& Function declaration \\
-                   &                     &     &
-& $e$ cannot be an unboxed literal \\
-
-\\
-
-Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$
-& $i \ge 0$ \\
-
-Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} 
-                                                 [\texttt{unboxed}]
-                                                 $C\ x_i =$ 
-& User-defined data type, $i \ge 1, n > 0$ \\
-                     &                   &     & $\mathit{con}_1 | \dots |
-                                                 \mathit{con}_n$
-& \\
-
-\\
-
-Program                & $\mathit{prog}$ & ::= & $(o|d)\mathit{decl}_i$
-& Objects and datatypes
-
+                                                 \texttt{->} $e$ & Lambda expression, $i > 0$ \\
+
+                   &                     & $|$ & \texttt{case} $e$ \texttt{of} $\mathit{alt}_i$ & Case expression, $i > 0$\\
+
+                   &                     & $|$ & \texttt{let} $\mathit{odecl}_i$ \texttt{in} $e$ & Recursive let expression, $i > 0$ \\
+
+\\
+Pattern            & $\mathit{pat}$      & ::= & $C\ \mathit{pat}_i$ & Nested constructor matching \\
+                   &                     & $|$ & $\mathit{lit}$ & Match numeric literals \\
+                   &                     & $|$ & $x$ & Bind to variable \\
+
+\\
+Alternative        & $\mathit{alt}$      & ::= & $\mathit{pat}_i$ \texttt{->} $e$ & $i > 0$ \\
+\\
+Type               & $\mathit{type}$     & $|$ & $C$ & Concrete type (constructor naming) \\
+                   &                     & $|$ & $x$ & polymorphic type variable \\
+                   &                     & $|$ & $\mathit{type}$ \texttt{->} $\mathit{type}$ & Function type \\
+\\
+Object decl. & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$ & Type signature \\
+                   &                     & $|$ & $x = e$ & Simple binding \\
+                   &                     & $|$ & $f\ \mathit{pat}_i = e$ & Function declaration \\
+                   &                     &     & & $e$ cannot be an unboxed literal \\
+\\
+Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$ & $i \ge 0$ \\
+\\
+Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} [\texttt{unboxed}] & User-defined data type  \\
+               &                   &     & $C\ x_i =$                       & $i \ge 0, n > 0$         \\
+               &                   &     & $\mathit{con}_1 | \dots |  \mathit{con}_n$ \\
+
+\\
+Program                & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_i$ & Object and data defns
 \end{tabular}
@@
 There are several hundred small STG and mini-Haskell programs in the test suite 
-(Most of which were written by students).
 It is partitioned into a number of directories which control what
@@
 We also run each test with various levels of garbage collection enabled. This
-has helped us to find bugs not only in the garbage collector, but also in the
-code generator.  Interestingly, in practice garbage collection can both mask
-and expose bugs.
+has exposed bugs not only in the garbage collector, but also in the code
+generator and runtime.  Interestingly, in practice garbage collection both
+masks and exposes bugs.
 
@@
 We have also implemented a simple logging facility that can be controlled at 
-both compile and runtime which helps in debugging why a test is failing.
+both compile and run time which helps in determining why a test is failing.
 

### 2016-04-07 pubs/tfp16/paper.tex
fix label/ref for Fig 2
sync

@@
 An informal specification of the mini-Haskell grammar is given in
-Figure~\ref{fig:miniHaskell}.  There is significant overlap with STG resulting
+%something really broken
+%Figure~\ref{fig:miniHaskell}.  
+Figure~2.  
+There is significant overlap with STG resulting
 from the structural similarities, but the redundancies are included for
@@
 \end{tabular}
+\caption{mini-Haskell syntax}
 \label{fig:miniHaskell}
-\caption{mini-Haskell syntax}
 \end{figure}
 
-Because mini-Haskell was designed to replace STG, it makes sense to talk about
-it in terms of the differences between the two and how mini-Haskell is
-translated into STG for the back-end.  Here, the differences are given in order
-of perceived utility, or importance to someone transitioning from STG to
-mini-Haskell.
+Because mini-Haskell was designed to supplement STG for test generation, it
+makes sense to talk about it in terms of the differences between the two and
+how mini-Haskell is translated into STG for the back-end.  Here, the
+differences are given in order of perceived utility, or importance to someone
+transitioning from STG to mini-Haskell.
 

### 2016-04-07 pubs/tfp16/paper.tex
sync

@@
              &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
-             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
+             &                & $|$ & $\texttt{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
              &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
@@
 \\
-Program      & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_1$ & Object and data defns \\
+Program      & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_1$ \texttt{;}            & Object and data defns \\
              &                 &     & \texttt{ } $\dots$ \texttt{;} \\
@@
 we refer so many times that we will subsequently refer to it as simply
-\emph{EA}. Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}),
-signifying a \emph{THUNK} being evaluated, are only runtime entities and not
-actually part of our STG language.
+\emph{EA}.  Black hole (\emph{BLACKHOLE}), signifying a \emph{THUNK} being
+evaluated, is only a runtime entity and not actually part of our STG language.
 
@@
 
-This then gives three reasonable operational modes that are strictly ordered
+This then gives three reasonable operational modes which are strictly ordered
 in terms of termination properties: non-strict; function strict, in which only
@@
 restrict ourselves to function/application and constructor strictness.  The
-STG machinery makes the choice of evaluation strategy easy to change.
-%, both in terms of function application and building a constructor in the
-%heap.
+STG machinery makes the choice of evaluation strategy easy to change, though
+changing code generation can be more efficient by avoiding the unnecessary
+creation of \emph{THUNKS} as discussed in Appendix~A.
 
@@
 
-For unboxed types code generation and runtime support is currently limited
-to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
-\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
-enumerations, e.g.,
-\begin{verbatim}
-     data unboxed Bool# = False# | True# .
-\end{verbatim}
+While we can parse and type-check more general unboxed types, code generation
+and runtime support is currently limited to built-in single-word (64-bit) or
+smaller types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#},
+\texttt{Double\#}) and user-defined enumerations (simple sums), e.g.,
+\texttt{data unboxed Bool\# = False\# | True\#}.
+%\begin{verbatim}
+%     data unboxed Bool# = False# | True# .
+%\end{verbatim}
 

### 2016-04-08 pubs/tfp16/paper.tex
figure number didn't stick
this is what's getting submitted

@@
 \title{Automatic Parallelization and Transparent Fault Tolerance\\
-       (New project paper)}
+       (\emph{New project} paper)}
 %
@@
         Loren Anderson\inst{1,3} \and
-        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
+        Jacob Marks\inst{1,4}\thanks{Ringo, Anderson, and Marks are undergraduate students.}
 }
@@
 programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  A second claimed trend in both
-scientific computing and functional programming is transparent fault
-tolerance.  Our project goal is the demonstration of a light-weight,
-higher-order, polymorphic, pure functional language implementation in which we
-can experiment with automatic parallelization strategies, varying degrees of
+functional programming, in various guises.  
+%
+%A second claimed trend is, given a strict functional semantics, in the
+%development of automatically parallelizing runtime systems.
+Besides making programs easier to reason about, this enables automatic
+parallelization to various degrees.
+%
+A second claimed trend in both scientific computing and functional programming
+is transparent fault tolerance.  
+%
+Our project goal is the demonstration of a light-weight, higher-order,
+polymorphic, pure functional language implementation in which we can
+experiment with automatic parallelization strategies, varying degrees of
 strictness, and mechanisms for transparent fault tolerance.
@@
 
-\item While this does not qualify as a student paper, students
+\item Hopefully in the spirit of a student-friendly symposium, for this draft
+  version we have chosen a more informal, conversational style of presentation
+  than necessarily appropriate for a final paper, have highlighted student
+  preparation, involvement, and its influence on our approach, and pointed out
+  some specific student contributions.
+
+\item While this \emph{does not qualify} as a student paper, students
   (undergraduate!) have contributed considerably to this effort, and to the
@@
   accepted we plan to have one of these students present it at TiFP.
-
-\item Hopefully in the spirit of a student-friendly symposium, for this draft
-  version we have chosen a more informal, conversational style of presentation
-  than necessarily appropriate for a final paper, have highlighted student
-  involvement and its influence on our approach, and pointed out student
-  contributions in particular.
 \end{enumerate}
@@
 
-\subsection{A trend: strict but still pure functional programming in HPC}
+\subsection{A trend: strict, pure functional programming in HPC and FP}
 Interest in functional programming for HPC in general is not new as evidenced
@@
 that \emph{pure functional semantics}, in some form, is essential to reining
-in the complexity of ever-evolving scientific codes and informing the
-designs of new ones.  Such semantics are expressed in a number of ways as
-highlighted following.
+in the complexity of ever-evolving scientific codes and informing the designs
+of new ones, and that strict semantics tend to be preferred.  Such semantics
+are expressed in a number of ways as highlighted following.
 
@@
 serious hindrance---than a convenience, much less ever algorithmically
-essential in the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's
-general attribute grammars~\cite{Johnsson87}, or other instances of
-\emph{tying the knot}~\cite{tying-the-knot}.
+essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
+Johnsson's general attribute grammars~\cite{Johnsson87}, 
+or other instances of \emph{tying the knot}~\cite{tying-the-knot}.
 
@@
 aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
-and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
+and the HPX C++11/14 extension~\cite{HPX}, all of which also seek to
+make parallelism automatic and transparent.  While the concept of task is
 somewhat vague, the general idea is a unit of computation with no or
@@
 
-\subsection{Another trend: transparent fault tolerance}
+\subsection{Another trend: transparent fault tolerance in HPC and FP}
 In the quest for computing platforms approaching exascale, it is widely
@@
 We claim, then, that there is a trend, in HPC at least, towards
-\emph{strict-by-default pure functional programming}, whether as imposed by
-programmer discipline, a parallel runtime system, or the language itself
-(e.g.\ strict Haskell).  Second, we claim that there is a trend in
+\emph{strict-by-default pure functional programming} and automatic
+parallelism.
+%whether as imposed by programmer discipline, a parallel runtime system, or the
+%language itself (e.g.\ strict Haskell).
+Second, we claim that there is a trend in
 high-performance computing, and by transitivity (in addition to Stewart's
@@
 not undertaking significant morphing of GHC itself: because this is in large
-part a student project it must be kept as simple as reasonably possible in
-various respects.  Our initial requirements are that
+part an undergraduate student project it must be kept as simple as reasonably
+possible.
+
+Our initial requirements are that
 %
@@
 %
-For feasibility, we restrict ourselves to shared-memory implicit parallelism.
+For feasibility we restrict ourselves to shared-memory implicit parallelism.
 The requirement for tail calling is not primarily for efficiency, rather it is
@@
 Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
-evoke both an abstract and concrete syntax.  This presentation is adapted from
-Marlow and Peyton Jones' \emph{Eval/Apply} paper~\cite{Marlow:2004}, to which
-we refer so many times that we will subsequently refer to it as simply
-\emph{EA}.  Black hole (\emph{BLACKHOLE}), signifying a \emph{THUNK} being
-evaluated, is only a runtime entity and not actually part of our STG language.
-
-The semantics of STG is what one would expect for a higher-order, polymorphic,
-pure functional language, with a few exceptions for a Haskell programmer
-familiar with unboxed types.  First, \texttt{case} is strict in the scrutinee,
-so \texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in
-Haskell, that is, the scrutinee is evaluated first.  Second, constructors are
-not functions, rather they are named in the construction (heap allocation) of
-a data object.  More generally, the operational notion of allocating a heap
-object is explicit: this is the operational semantics of \texttt{let} bindings.
+evoke both an abstract and concrete syntax.  This formulation and presentation
+is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
+paper~\cite{Marlow:2004}, to which we refer so many times that we will
+subsequently refer to it as simply \emph{EA}.  A (\emph{BLACKHOLE}),
+signifying a \emph{THUNK} being evaluated, is only a runtime entity and
+not part of the language.
+
+With a few exceptions the semantics of STG is what a Haskell programmer
+familiar with unboxed types would expect for a higher-order, polymorphic, pure
+functional language.  Points to note are, first, that \texttt{case} is strict
+in the scrutinee (the \emph{e} in \texttt{case} \emph{e} \texttt{of}), so \texttt{case} is
+similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell, that is, the
+scrutinee is evaluated first.  Second, constructors are not functions, rather
+they are named in the construction (heap allocation) of a data object.  More
+generally, the operational notion of allocating a heap object is explicit:
+this is the operational semantics of \texttt{let} bindings.
 %As such a variable of unboxed type cannot be let-bound; instead it can be
@@
 Finally, deferred evaluation is explicit:
-a deferred expression evaluation is encoded by a \emph{THUNK}.
+a deferred expression evaluation is encoded as a \emph{THUNK}.
 
@@
 (as implied by lazy but not by non-strict) of heap-allocated objects is
-preserved so we will not use the term \emph{lazy} and simply refer to
+preserved so we will not use the term lazy and simply refer to
 strictness properties.  As is common we will use strict in an operational
@@
 
-In STG the arity of a function is syntactically defined as the number of
-manifest formal parameters in a function definition, thus \texttt{id x = x}
-has by definition arity one regardless of the type of the argument (e.g., some
-function type) it might be applied to.  We distinguish \emph{application},
-e.g., \texttt{f x y}, which is a syntactic notion, from actually calling a
-function. Thus \texttt{id id x} is an application of \texttt{id} to two
-arguments, but the underlying \emph{FUN} (user- or system- defined function)
-when actually called is given exactly one argument (the first one).
+In STG the arity of a function is defined as the number of formal parameters
+in the function definition, thus \texttt{id x = x} has by definition arity one
+regardless of the type of the argument (e.g., some function type) it might be
+applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
+a syntactic notion, from actual function \emph{call}. Thus \texttt{id id x} is
+an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
+(user- or system- defined function) when called is given exactly one
+argument (the first one).
 
@@
 
-Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
+Given application \texttt{f x y}, when might \texttt{x} and \texttt{y} be
 evaluated?  In the non-strict semantics evaluation is on demand, after
@@
 In a strict semantics one could devise various evaluation orders, but as far
-as distinguishing termination properties it seems these can be distilled down
-to two reasonable possibilities, namely whether on construction of a
+as distinguishing termination properties it seems that these can be distilled
+down to two reasonable possibilities, namely whether on construction of a
 \emph{PAP} the (new) arguments are forced or not.  In other words, the
-question is whether application is strict, or actual function call is strict.
-These are in fact distinguishable because (in Haskell) the first argument to
+question is whether application is strict, or function call is strict.  These
+are in fact distinguishable because (in Haskell) the first argument to
 \texttt{seq} might evaluate to a \emph{PAP}.
@@
 This then gives three reasonable operational modes which are strictly ordered
-in terms of termination properties: non-strict; function strict, in which only
-those arguments needed by a function call are evaluated; and application
-strict, in which all arguments of an application are evaluated.  The other
-dimension we vary is whether or not constructors are strict.  Other dimensions
-are possible, for example whether \texttt{let} bindings are strict, but we
-restrict ourselves to function/application and constructor strictness.  The
-STG machinery makes the choice of evaluation strategy easy to change, though
-changing code generation can be more efficient by avoiding the unnecessary
-creation of \emph{THUNKS} as discussed in Appendix~A.
+in terms of termination properties: 
+non-strict, function strict, and application strict.
+% non-strict; function strict, in which only
+%those arguments needed by a function call are evaluated; and application
+%strict, in which all arguments of an application are evaluated.  
+The other dimension we vary is whether or not constructors are strict.  Other
+dimensions are possible, for example whether \texttt{let} bindings are strict,
+but we restrict ourselves to function/application and constructor strictness.
+The STG machinery makes the choice of evaluation strategy easy to change,
+though changing code generation can be more efficient by avoiding the
+unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.
 
@@
 execution they identified the need for the ability to examine and unwind the
-stack in quashing concurrent evaluation of the same \emph{THUNK}.  This
+stack in quashing concurrent evaluation of the same \emph{THUNK}\@.  This
 requires that use of the C stack be highly constrained if we are to avoid
@@
 functions, and applications of known functions to other than their arity
-number of arguments.  Here the object to be applied is first evaluated with
+number of arguments.  There the object to be applied is first evaluated with
 call-with-return.  When a resulting \emph{FUN} is applied to more than its
@@
 This is straightforwardly realizable because the stack is explicit, and with
-modern C compiler call optimization, the continuation style enables a proper
-\emph{goto} mechanism where the \emph{code labels}---entities that name blocks
-of code, can be treated as first class (e.g.\ stored), and serve as targets of
-a jump---are C function addresses (c.f.\ \S 6.1~\cite{PJ-stockhardware}).
+modern C compiler call optimization, the continuation style enables a jump
+(\emph{go to}) mechanism where the \emph{code labels}---entities that name
+blocks of code, can be treated as first class (e.g., stored for later use),
+and serve as targets of a jump---are C function addresses (c.f.\ \S
+6.1~\cite{PJ-stockhardware}).
 
@@
 is likely out of scope.
+%
+The fault-tolerance mechanisms remain future work.
 
@@
 %
-A plug-in for GHC (prior to 8.0) exists for making functions, but not data
-structures, strict using a transformation on GHC Core that recursively expands
-non-recursive let bindings into case bindings~\cite{strict-ghc-plugin}.  Thus
-it seems that while it makes functions denotationally strict, function
-arguments are still evaluated after function call.
+A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
+their non-recursive \texttt{let} bindings using a transformation on GHC Core
+that recursively expands those \texttt{let} bindings into \texttt{case}
+bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
 %
@@
 for its eventual completion are unknown~\cite{disciplined-disciple}.
-%
+
+\subsubsection{Strict, pure, auto-parallelizing.}
 NVIDIA's NOVA language, and corresponding compiler, implements a strict
@@
 language compiler that is ``largely language agnostic'' and performs (among
-many other optimization) SIMD vectorization.
+many other optimizations) SIMD vectorization.
 
@@
 \emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
-an unboxed literal value.  Our extension (to STG as well) is the provision of
+an unboxed literal value.  Our extension (as for STG) is the provision of
 the optional \texttt{unboxed} keyword in datatype definitions to allow
@@
 An informal specification of the mini-Haskell grammar is given in
-%something really broken
-%Figure~\ref{fig:miniHaskell}.  
-Figure~2.  
-There is significant overlap with STG resulting
+Figure~\ref{fig:miniHaskell}.  There is significant overlap with STG resulting
 from the structural similarities, but the redundancies are included for
@@
 how mini-Haskell is translated into STG for the back-end.  Here, the
-differences are given in order of perceived utility, or importance to someone
+differences are given in order of perceived utility, or importance in
 transitioning from STG to mini-Haskell.
@@
 \end{verbatim}
-
 In mini-Haskell, expressions may be applied to other expressions, so the
@@
 \texttt{let} expression:
-
 \begin{verbatim}
@@
 \end{verbatim}
-
 This also serves as a good example of how expression to expression application is
@@
 STG supports pattern matching but only in a simple form and only in
-\texttt{case} expressions.  mini-Haskell extends this to something approaching
+\texttt{case} expressions.  Mini-Haskell extends this to something approaching
 what is found in Haskell.  Functions, including lambda expressions, may match on
@@
 Even relatively simple functions can be painful to write, such as \texttt{unzip}.
-
 \begin{verbatim}
@@
 \end{verbatim}
-
 In mini-Haskell, this becomes
-
 \begin{verbatim}
@@
 
-Of course, this complex pattern matching must eventually be transformed into the
-primitive, nested case expressions. Luckily, there are well-documented processes
-for achieving this, and Philip Wadler's chapter in \emph{The Implementation of
-  Functional Programming Languages} \cite{Jones:1987} % doing this right?
-on the compilation of this type of pattern matching proved invaluable in
-bringing it to mini-Haskell.  Wadler describes the \texttt{match} function that
-that, in its simplest form, produces a correct, but potentially large
-translation with some redundancy.  This function was implemented with only minor
-modifications to fit the shape of mini-Haskell's abstract form.  The
-implementation is augmented by some light flow-analysis to eliminate some of the
-redundancy and make the resulting code easier to read.  There are further
-optimizations suggested by Wadler to address the size of the generated code from
-this process (for example, when multiple branches lead to the same large
-expression), but the additional complexity outweighs the benefit for a testing
-language.
+Of course, this complex pattern matching must eventually be transformed into
+the primitive, nested case expressions. Fortunately there are well-documented
+algorithms for achieving this, and Philip Wadler's chapter in \emph{The
+  Implementation of Functional Programming Languages}~\cite{Jones:1987} %
+doing this right?  on the compilation of this type of pattern matching proved
+invaluable.  Wadler describes the \emph{match} function that which, in its
+simplest form, produces a correct, but potentially large translation with some
+redundancy.  This function was implemented with only minor modifications to
+fit the shape of mini-Haskell's abstract form.  The implementation is
+augmented by some light flow-analysis to eliminate some of the redundancy and
+make the resulting code easier to read.  There are further optimizations
+suggested by Wadler to address the size of the generated code ,for example,
+when multiple branches lead to the same large expression, but the additional
+complexity outweighs the benefit for a testing language.
 
@@
 
-Mini-Haskell provides some smaller conveniences that don't warrant their own
-section, but are worth mentioning.
-
+Mini-Haskell provides some smaller conveniences that don't warrant individual
+sections but are worth mentioning.
 \begin{itemize}
-\item Constructors as functions \\
+\item \emph{Constructors as functions.}
   As in Haskell, constructors are first class, acting as true functions.  This
@@
 
-\item Primitive operations as functions \\
+\item \emph{Primitive operations as functions.}
   Similar to constructors, primitive operations are given first-class status as
-  functions, with roughly the same rules.  Partially applied calls are
-  translated to generated functions that wrap the actual primitive
+  functions, with roughly the same rules.  Partial applications are
+  translated to calls to generated functions that wrap the actual primitive
   operations. Fully applied calls generate true primitive operation calls.
 
-\item Lambda Expressions \\
+\item \emph{Lambda expressions.}
   mini-Haskell supports anonymous lambda expressions for the simple, one-off
@@
 
-\item Type Annotations \\
+\item \emph{Type annotations.}
   The ability to assert the type of some variable is often valuable as a
@@
 % I don't think I've described this very well at all... - david
-\item Haskell Block \\
+\item \emph{Haskell Block.}
   Being a subset of Haskell, it may be desirable to test a mini-Haskell program
@@
 
-We have developed a extensive test suite which uses the CMake CTest
+We have developed an extensive test suite which uses the CMake CTest
 tool~\cite{cmake-book}. The full test suite is automatically run, and any
@@
 
-There are several hundred small STG and mini-Haskell programs in the test suite 
-It is partitioned into a number of directories which control what
-testing is done for a specific program. Given our goal of exploring various
+There are several hundred small STG and mini-Haskell programs in the test suite, 
+which is partitioned into a number of directories which control what
+testing is done. Given our goal of exploring various
 levels of strictness, most tests are run with all evaluation strategies (three
 degrees of function strictness by two degrees of constructor strictness).
-One interesting class of test programs are those which require lazy
+One interesting class of test programs are those which require lazy or non-strict
 semantics. For these programs we test not only that the test leads to the expected
@@
 beginners have contributed substantially to the test suite. We have also found
-that the test suite has be of use as a learning tool for these students to
+that the test suite has been of use as a learning tool for these students to
 discover what level of non-strictness was required for their programs,
@@
 Another important class of test programs are those which should fail with a
-known error (e.g. a parsing error) for these programs the test suite checks
-the output against an expected error regular expression. This class of test
-programs has proved valuable in development and debugging of both the STG
-and mini-Haskell front-ends.
+known error, e.g., a parsing or typing error.  For these programs the test
+suite checks the output against an expected error regular expression. This
+class of test programs has proved valuable in development and debugging of
+both the STG and mini-Haskell front-ends.
 

### 2016-04-08 pubs/tfp16/paper.tex
it only takes submitting to find typos

@@
 \begin{abstract}
+
 A claimed trend is an increasing appreciation and practice of functional
 programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  
-%
-%A second claimed trend is, given a strict functional semantics, in the
-%development of automatically parallelizing runtime systems.
-Besides making programs easier to reason about, this enables automatic
-parallelization to various degrees.
-%
-A second claimed trend in both scientific computing and functional programming
-is transparent fault tolerance.  
-%
-Our project goal is the demonstration of a light-weight, higher-order,
-polymorphic, pure functional language implementation in which we can
-experiment with automatic parallelization strategies, varying degrees of
-strictness, and mechanisms for transparent fault tolerance.
+functional programming, in various guises.  Besides making programs easier to
+reason about, this enables automatic parallelization to various degrees.  A
+second claimed trend in both scientific computing and functional programming
+is transparent fault tolerance.  Our project goal is the demonstration of a
+light-weight, higher-order, polymorphic, pure functional language
+implementation in which we can experiment with automatic parallelization
+strategies, varying degrees of strictness, and mechanisms for transparent
+fault tolerance.\\
 %We do not consider speculative evaluation, or semantic strictness inferred by
@@
 %strictness.
-\\
 \emph{Draft paper notes:}
@@
 
-\keywords{Strict pure functional programming, automatic parallelization,
-  Haskell, STG, GHC, GHC Core, transparent fault tolerance.}
+\keywords{ Strict pure functional programming, automatic parallelization,
+  Haskell, STG, GHC, GHC Core, transparent fault tolerance.  }
 
@@
 end.  The edge cases that it will incorrectly handle as a result of this are
-exactly that: edge cases.  As such, they are easy to avoid in practical usage.
-
-The inclusion of the layout rule may seem small and insignificant, but it
+exactly that: edge cases.  As such, they are easy to avoid in practice.
+
+The inclusion of the layout rule may seem insignificant, but it
 allows users to produce code that is visually similar to Haskell code.
@@
 the primitive, nested case expressions. Fortunately there are well-documented
-algorithms for achieving this, and Philip Wadler's chapter in \emph{The
-  Implementation of Functional Programming Languages}~\cite{Jones:1987} %
-doing this right?  on the compilation of this type of pattern matching proved
-invaluable.  Wadler describes the \emph{match} function that which, in its
-simplest form, produces a correct, but potentially large translation with some
-redundancy.  This function was implemented with only minor modifications to
-fit the shape of mini-Haskell's abstract form.  The implementation is
-augmented by some light flow-analysis to eliminate some of the redundancy and
-make the resulting code easier to read.  There are further optimizations
-suggested by Wadler to address the size of the generated code ,for example,
-when multiple branches lead to the same large expression, but the additional
-complexity outweighs the benefit for a testing language.
+algorithms for achieving this, and Philip Wadler's chapter
+% in \emph{The Implementation of Functional Programming Languages}
+on the compilation of this type of pattern matching proved
+invaluable~\cite{Jones:1987}.  Wadler describes the \emph{match} function
+which, in its simplest form, produces a correct, but potentially large
+translation with some redundancy.  This function was implemented with only
+minor modifications to fit the shape of mini-Haskell's abstract form.  The
+implementation is augmented by some light flow-analysis to eliminate some of
+the redundancy and make the resulting code easier to read.  There are further
+optimizations suggested by Wadler to address the size of the generated code,
+for example, when multiple branches lead to the same large expression, but the
+additional complexity outweighs the benefit for a testing language.
 

### 2016-04-12 pubs/tfp16/paper.tex
and yet more
still mid-hack stgApply

@@
 However, they see their needs as different than other Haskell users: performance
-is paramount, vectors~\cite{vector} are a (perhaps \emph{the}) primary data
+is paramount, array/vector~\cite{vector} is a (perhaps \emph{the}) primary data
 structure, and non-strictness or laziness is more of a nuisance---even a
@@
 %
-For feasibility we restrict ourselves to shared-memory implicit parallelism.
-The requirement for tail calling is not primarily for efficiency, rather it is
-for a reasonable parallel implementation.  Achieving this, and thereby
-sufficiently limiting the use of the native C stack, will be discussed at some
-length.
+For feasibility we restrict ourselves to implicit parallelism on shared-memory
+multiprocessors.  The requirement for tail calling is not primarily for
+efficiency, rather for an efficient parallel implementation.  Achieving
+this, and thereby sufficiently limiting the use of the native C stack, will be
+discussed at some length.
 
@@
              &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
-             &                & $|$ & $\texttt{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
+             &                & $|$ & $\texttt{PAP}\ f\ a_1\dots a_n$               & Partial application \\
              &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
@@
 Hindley-Milner type inference and enforcement following Heren et
-al.~\cite{Heren02}, again extended with unboxed types.  The syntax,
-restrictions, and extensions to the type inference algorithm for unboxed types
-follow Peyton Jones and Launchbury~\cite{Jones:1991}.
+al.~\cite{Heren02}, again extended with unboxed types.  The extensions to the
+type inference algorithm, syntax, and restrictions for unboxed types follow
+Peyton Jones and Launchbury~\cite{Jones:1991}.
+\begin{comment}
+% breaks the page limit
+For convenience general recursive
+\texttt{let}s are implicitly factored in the usual way to maximize let-polymorphism;
+this would normally already be performed by a front-end.
+\end{comment}
+\begin{comment}
+% breaks the page limit
+One would expect an automated STG generator such as GHC to factor recursive
+\texttt{let} expressions into nested recursive and non-recursive \texttt{let}s
+in the usual way to maximize let-polymorphism.  Because our STG is intended
+to be somewhat convenient to write by hand we relax this constraint and perform
+this transformation implicitly using a strongly-connected components algorithm
+inspired by King and Launchbury~\cite{King:1995}.
+\end{comment}
 
@@
 Core) and our STG, augmenting the runtime as needed.  Just as in the Intel
-Haskell Research Compiler effort we will not attempt to reuse the GHC runtime.
+Haskell Research Compiler effort we will not use the GHC runtime.
 
@@
 frustrating syntactic cruft and adding some of the core features that are
-often taken for granted in Haskell (and other languages).
+often taken for granted in Haskell (and other languages).  An alternative
+approach would translate from mini-Haskell to GHC Core, then either to STG or,
+in the case that the back end eventually take, compiling Core directly.
 

### 2016-04-12 codegen/SCC.hs

@@ -8,7 +8,9 @@ module SCC (
 ) where
 
 {-
-   David King and John Launchbury, Lazy Depth-First Search and Linear Graph Algorithms in Haskell,
+   David King and John Launchbury, Lazy Depth-First Search and Linear Graph Algorithms in Haskell
+
+    Structuring depth-first search algorithms in Haskell,
     ACM Principles of Programming Languages, San Francisco, 1995.
 -}
 

### 2016-04-12 runtime/stg.c

@@ -130,7 +130,8 @@ Cont *stgAllocCallOrStackCont(CInfoTab *citp, int argc) {
   contp->cInfoPtr = citp;
   contp->_contSize = contSize;  // to go away
 
-  contp->layout.bits = 0x0UL;  // for GC
+  contp->layout.bitmap.mask = 0x0UL;  // for GC
+  contp->layout.bitmap.size = argc;   // make consistent
 
   contp->entryCode = citp->entryCode;
   contp->contType = citp->contType;

### 2016-04-12 runtime/stgapply.c

@@ -7,27 +7,24 @@
 
 FnPtr stgApplyCurVal();
 
-Cont *stgContSplit(int excess) {
+Cont *stgContSplit(int arity1, int excess) {
+  // arity1 is arity of funoid
+  // cont1 is cont to split
   Cont *cont1 = (Cont *)stgSP;
-  size_t size1 = getContSize(cont1);
-  int pls1 = cont1->layout.bitmap.size;
-  int arity1 = pls1 - excess - 1;
-
-  int pls2 = 1 + excess; // funoid + excess
+  // funoid + excess, subsequent application
+  int pls2 = 1 + excess; 
   Cont *cont2 = stgAllocCallOrStackCont( &it_stgStackCont, pls2 );
-  size_t size2 = sizeof(Cont) + pls2 * sizeof(PtrOrLiteral);
-
-  int pls3 = pls1 - excess;
+  // funoid + arity1
+  int pls3 = arity1 + 1;
   Cont *cont3 = stgAllocCallOrStackCont( &it_stgStackCont, pls3 );
-  size_t size3 = sizeof(Cont) + pls3 * sizeof(PtrOrLiteral);
 
   Bitmap64 bm = cont1->layout;
-  bm.bitmap.size = pls2;
-  bm.bitmap.mask <<= (pls3 - 1); // room for funoid
-  bm.bitmap.mask |= 0x1UL;
+  // bm.bitmap.size is set by alloc
+  bm.bitmap.mask >>= arity1; // room for funoid, eliminate just right
+  bm.bitmap.mask |= 0x1UL;   // funoid is boxed
   cont2->layout = bm;
-  cont2->entryCode = &stgApplyCurVal;
-  cont2->payload[0].op = NULL;
+  cont2->entryCode = &stgApplyCurVal;  // goto stgApplyCurVal
+  cont2->payload[0].op = NULL;  // no funoid yet
   #if USE_ARGTYPE
   cont2->payload[0].argType = HEAPOBJ;
   #endif
@@ -44,8 +41,8 @@ Cont *stgContSplit(int excess) {
   //  cont3->entryCode = getInfoPtr(cont3->payload[0].op)->funFields.trueEntryCode;
 
   // quash cont1
-  memmove(cont1, cont2, size2 + size3);
-  stgSP += size1;
+  memmove(cont1, cont2, getContSize(cont2) + getContSize(cont3));
+  stgSP += getContSize(cont1);
[…]

### 2016-04-15 pubs/tfp16/paper.tex
CodeGen.hs adds extra arg to stgApply, fix strictness in stgapply.c

@@
 \item continuous integration and testing system,
-\item extensive test suite \emph{(mostly by students)}.
+\item extensive test suite \emph{(mostly by students\footnote{Many are
+    students' solutions to exercises in learning algorithms and data
+    structures.})}.
 \end{itemize}
@@
 
-Various papers describing some of the inner workings of GHC at various points
-in its evolution have greatly informed our approach and these have all been
-previously cited.
+Previously cited papers describing some of the inner workings of GHC at
+various points in its evolution have greatly informed our approach.
 
@@
 
-Recently, Robert Stewart extended Haskell Distributed Parallel Haskell
+More recently, Robert Stewart extended Haskell Distributed Parallel Haskell
 (HdpH)~\cite{hdph} to be fault resilient at an impressive scale---1400 cores

### 2016-04-15 codegen/CodeGen.hs

@@ -334,11 +334,12 @@ stgApplyGeneric env f eas direct =
         inline =
             -- new STACKFRAME
             "{ Cont *cp = stgAllocCallOrStackCont( &it_stgStackCont, " ++
-                 show (length pnstring + 1) ++ ");\n" ++
-            "  cp->layout = " ++ npStrToBMStr ('P' : pnstring ) ++ ";\n" ++
-            "  cp->payload[ 0 ] = " ++ cgv env f' ++ ";\n" ++
+                 show (length pnstring + 2) ++ ");\n" ++
+            "  cp->layout = " ++ npStrToBMStr ('N' : 'P' : pnstring ) ++ ";\n" ++
+            "  cp->payload[ 0 ] = " ++ cga [] (LitI 0) ++ ";\n" ++
+            "  cp->payload[ 1 ] = " ++ cgv env f' ++ ";\n" ++
             concat ["  cp->payload[ " ++ show i ++ " ] = " ++ cga env a ++ ";\n"
-                    | (i,a) <- zip [1..] as ] ++
+                    | (i,a) <- zip [2..] as ] ++
 
             (if direct then
                 "  // DIRECT TAIL CALL " ++ f ++ " " ++ showas as ++ "\n" ++
@@ -348,7 +349,7 @@ stgApplyGeneric env f eas direct =
                 "  STGJUMP0(fun_" ++ f ++ ");\n"
             else
                 "  // INDIRECT TAIL CALL " ++ f' ++ " " ++ showas as ++ "\n" ++
-                "  STGJUMP0(stgApply);\n") ++
+                "  STGJUMP0(stgApplyNew);\n") ++
             "}\n"
     in return ((inline, Yes), [])
 
@@ -379,7 +380,7 @@ cge env e@(EFCall it f eas) =
     case (knownCall it) of
       Nothing -> stgApplyGeneric env f eas False
       Just kit -> if arity kit == length eas
-                  then stgApplyGeneric env f eas True
+                  then stgApplyGeneric env f eas False
                   else stgApplyGeneric env f eas False
 
 cge env (EPrimop it op eas) =

### 2016-04-15 runtime/stgapply.h

@@ -1,7 +1,8 @@
 #ifndef stgapply_h
 #define stgapply_h
 
+FnPtr stgApplyNew();
 FnPtr stgApply();
-void stgEvalStackFrameArgs(Cont *);
+//void stgEvalStackFrameArgs(Cont *);
 
 #endif

### 2016-04-15 runtime/stgapply_tc.c

@@ -150,6 +150,13 @@ FnPtr stgApply() {
   STGJUMP();  // bye bye
 }
 
+FnPtr stgApplyNew() {
+  Cont *stackframe = stgGetStackArgp();
+  stackframe->entryCode = &stgApply2;
+  stgCurVal = stackframe->payload[1];  // the funoid
+  STGJUMP();  // bye bye
+}
+
 FnPtr stgApply2() {
   Cont *stackframe = stgGetStackArgp();
   derefStgCurVal();
@@ -177,10 +184,10 @@ FnPtr stgApply2() {
       Bitmap64 papargmap = argv[0].op->payload[fvCount].b;
       int papargc = papargmap.bitmap.size;
       arity -= papargc;
-      int excess = fargc - arity;
-      argsToEval = excess >= 0 ? arity : 0;
-      break;
     } // if PAP
+    int excess = fargc - arity;
+    argsToEval = excess >= 0 ? arity : 0;
+    break;
   } // STRICT1
   } // switch
 

### 2016-04-19 pubs/tfp16/paper.tex
random edits

@@
 written in C/C++. This gives us the extensive optimization performed by modern
-C compilers such as gcc and Clang/LLVM. The system fairly faithfully
+C compilers such as GCC and Clang/LLVM. The system fairly faithfully
 implements the STG machine as described in \emph{EA} and dynamically
@@
 %
-To clearly distinguish between a notional C call (a syntactic concept) and a
-machine-level call (pushing a return address onto the native C stack) we will
-refer to the latter as \emph{call-with-return}, and to a machine-level tail
-call as \emph{jump}.
-
-First we establish that this is possible in principle.  By maintaining our own
-data and control (continuation) stack all generated C functions can take no
-explicit arguments and have return type \emph{void}. If we arrange that
-generated C code has only calls (other than runtime calls, which do not call
-back to user-program code) in tail call positions, that is, immediately before
-a return in the control flow graph, then at non-zero optimization levels both
-gcc and Clang/LLVM will generate jumps instead of calls-with-return for C
-calls.  When using gcc we can also use a fixed set of machine registers for
-passing/returning values, much like GHC using the LLVM back-end with its
+To clearly distinguish what a C call (a syntactic concept) can compile to, we
+will refer to a machine-level call (pushing a return address onto the native C stack)
+as \emph{call-with-return}, and to a machine-level jump as \emph{jump}.
+
+First we establish that this is possible in principle on x86-64/Linux.  By
+maintaining our own data and control (continuation) stack all generated C
+functions can take a uniform (including empty) set of explicit arguments and
+have uniform return type). If we arrange that generated C code has only calls
+(other than runtime or other external calls, which do not call back to
+user-program code) in tail call positions, that is, immediately before a
+return in the control flow graph, then at non-zero optimization levels both
+GCC and Clang/LLVM will generate jumps instead of calls-with-return for C
+calls.  When using GCC we can also use an explicit set of machine registers
+for passing/returning values, much like GHC using the LLVM back-end with its
 custom GHC calling convention.
 
-The STG abstract machine is implemented using a continuation stack.
+The STG abstract machine is implemented using a combined data and continuation stack.
 Continuations contain a code pointer, optionally some data, and layout
-information for the benefit of the garbage collector.  For example, the
-argument frame for a function call is one type of continuation.  When the
-evaluation of a C function (corresponding to some STG code) is completed its
-return is by an indirect jump using the code pointer in the continuation on
-the top of the stack (``returning through the continuation stack'').  In the
-case of an argument frame for a function, the associated code simply pops the
-continuation off the stack and returns through the next continuation on the
-stack.
+information for the benefit of the garbage collector.  
+% For example, the argument frame for a function call is one type of
+% continuation.
+When the evaluation of a C function (corresponding to some STG code) is
+completed its return is by an indirect jump using the code pointer in the
+continuation on the top of the stack (``returning through the continuation
+stack'').
+%In the case of an argument frame for a function, the associated code simply
+%pops the continuation off the stack and returns through the next continuation
+%on the stack.
 
@@
 
-In addition to being in a tail call position, the gcc constraints for
+In addition to being in a tail call position, the GCC constraints for
 generating a tail call (jump) are (approximately) that the total size of the
@@
 
-We are currently developing the parallel runtime.  Parallel garbage collection
-is likely out of scope.
+For simplicity the implementation is 64-bit only.
+%
+The current prototype parallel runtime is based on Pthreads and a non-blocking
+work queue~\cite{Michael:1996}.  We are currently developing a more realistic
+parallel runtime, experimenting with using the Argobots user-level thread library~\cite{Seo:2015}.
 %
 The fault-tolerance mechanisms remain future work.
+%
+Parallel garbage collection is likely out of scope.
 
@@
 Science Foundation Science, Technology, Engineering, and Mathematics Talent
-Expansion Program (STEP) program for undergraduate students; and the DOE
-Science Undergraduate Laboratory Internship (SULI) program.
+Expansion Program (NSF STEP) program for undergraduate students; and the Department of Energy
+Science Undergraduate Laboratory Internship (DOE SULI) program.
 %
@@
 an unnecessary heap allocation.  An alternative is to translate to
-\texttt{case e of x \{\_ -> f x\}}.
+\texttt{case e of x \{\_ -> f x\}}, which we regard as a role of a front-end
+such as mini-Haskell.
 
@@
                         result = CON(Cons hd rec) }
-                  in result });
+                  in result })
 \end{verbatim}
@@
                                 res = CON(TP2 aas bbs)}
-                           in res; }}});
+                           in res }}})
 \end{verbatim}

### 2016-04-19 pubs/tfp16/paper.tex
more minor edits

@@
 \usepackage{enumitem} 
-\usepackage{color} % for parenthetical notes while paper is in progress
+\usepackage[dvipsnames]{xcolor} % for parenthetical notes while paper is in progress
 \usepackage{comment} % comment environment
@@
 %strictness.
+{\color{OliveGreen}
 \emph{Draft paper notes:}
@@
 \item We are reporting work in progress.  At the time of writing we have a
-  complete serial implementation, a primitive proof-of-concept parallel
+  serial implementation, a primitive proof-of-concept parallel
   implementation, and are currently developing a more realistic parallel
@@
 
-\item Hopefully in the spirit of a student-friendly symposium, for this draft
-  version we have chosen a more informal, conversational style of presentation
-  than necessarily appropriate for a final paper, have highlighted student
-  preparation, involvement, and its influence on our approach, and pointed out
-  some specific student contributions.
-
-\item While this \emph{does not qualify} as a student paper, students
+\item Hopefully in the spirit of a student-friendly symposium, this draft
+  version is liberally annotated with indicators of student contributions, 
+  student preparation, and the influence of their involvement on our approach.
+
+\item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
   (undergraduate!) have contributed considerably to this effort, and to the
@@
 \end{enumerate}
+}
 
@@
 %
-The current prototype parallel runtime is based on Pthreads and a non-blocking
+The current, primitive parallel runtime is based on Pthreads and a non-blocking
 work queue~\cite{Michael:1996}.  We are currently developing a more realistic

### 2016-04-19 pubs/tfp16/paper.tex
more edits

@@
 strategies, varying degrees of strictness, and mechanisms for transparent
-fault tolerance.\\
-%We do not consider speculative evaluation, or semantic strictness inferred by
-%program analysis, so potential parallelism is dictated by the degree of
-%strictness.
+fault tolerance.
+%
+We do not consider speculative evaluation, or semantic strictness inferred by
+program analysis, so potential parallelism is dictated by the degree of
+strictness.\\
 {\color{OliveGreen}
@@
 \item Hopefully in the spirit of a student-friendly symposium, this draft
-  version is liberally annotated with indicators of student contributions, 
-  student preparation, and the influence of their involvement on our approach.
+  is liberally annotated with indicators of student contributions, 
+  student preparation, the influence of their involvement on our approach,
+  and other ancillary comments.
+  A final paper would of course be cleaned up.
 
@@
   writing of this paper, and will continue to do so.  If this paper is
-  accepted we plan to have one of these students present it at TiFP.
+  accepted we plan to have one of these students present it at TiFP'16.
 \end{enumerate}
@@
 
-\keywords{ Strict pure functional programming, automatic parallelization,
-  Haskell, STG, GHC, GHC Core, transparent fault tolerance.  }
+\keywords{Strict pure functional programming, automatic parallelization,
+  transparent fault tolerance, Haskell, STG, GHC, Core.}
 
@@
 the software stack.  Current practice in HPC is checkpoint/restart, wherein at
-various points in time sufficient program state is dumped to backing store
+various points in time sufficient global program state is dumped to backing store
 such that the program, in case of a fault, can be (manually) restarted from
@@
 application because it must be designed such that consistent, essential state
-can be readily captured.  Furthermore, there is a concern that the scalability
-of checkpoint/restart is reaching its end as data transfer to backing store
-becomes an increasing bottleneck.
+can be readily captured.  Finally, the scalability of global
+checkpoint/restart is reaching its end as data transfer to backing store
+becomes an increasing bottleneck~\cite{DOE:2009}.
 
@@
 We claim, then, that there is a trend, in HPC at least, towards
-\emph{strict-by-default pure functional programming} and automatic
-parallelism.
+\emph{strict-by-default pure functional programming} and \emph{automatic
+parallelism}.
 %whether as imposed by programmer discipline, a parallel runtime system, or the
 %language itself (e.g.\ strict Haskell).
-Second, we claim that there is a trend in
-high-performance computing, and by transitivity (in addition to Stewart's
-work) in functional programming, towards transparent fault tolerance.
+Second, we claim that there is a trend in high-performance computing, and
+functional programming, towards \emph{transparent fault tolerance}.
 
@@
 non-strictness.\footnote{The starting experience of our student interns to
-  date has been normalized by having them complete Hutton's \emph{Programming
+  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
     in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
@@
 is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
-paper~\cite{Marlow:2004}, to which we refer so many times that we will
-subsequently refer to it as simply \emph{EA}.  A (\emph{BLACKHOLE}),
-signifying a \emph{THUNK} being evaluated, is only a runtime entity and
-not part of the language.
+paper~\cite{Marlow:2004}, to which we will subsequently refer as \emph{EA}\@.  A
+\emph{BLACKHOLE}, signifying a \emph{THUNK} being evaluated, is only a
+runtime entity and not part of the language.
 
@@
 To clearly distinguish what a C call (a syntactic concept) can compile to, we
-will refer to a machine-level call (pushing a return address onto the native C stack)
-as \emph{call-with-return}, and to a machine-level jump as \emph{jump}.
+will refer to a machine-level call---pushing a return address onto the native
+C stack---as \emph{call-with-return}, and to a machine-level jump as
+\emph{jump}.
 
@@
 functions can take a uniform (including empty) set of explicit arguments and
-have uniform return type). If we arrange that generated C code has only calls
+have uniform return type. If we arrange that generated C code has only calls
 (other than runtime or other external calls, which do not call back to
@@
 Peyton Jones and Launchbury~\cite{Jones:1991}.
+%\begin{comment}
+% breaks the page limit
+For convenience we allow general recursive
+\texttt{let}s that are implicitly factored in the usual way to maximize let-polymorphism;
+this would normally already be performed by a front-end.
+%\end{comment}
 \begin{comment}
-% breaks the page limit
-For convenience general recursive
-\texttt{let}s are implicitly factored in the usual way to maximize let-polymorphism;
-this would normally already be performed by a front-end.
-\end{comment}
-\begin{comment}
-% breaks the page limit
+ breaks the page limit
 One would expect an automated STG generator such as GHC to factor recursive
@@
 An almost unbounded amount of interesting development suggests itself; one
-such path would lead to the recreation of GHC.  However, given the minuscule
+such path could in principle lead to the recreation of GHC\@.  However, given the minuscule
 budget (and therefore the employment of student interns) we must maintain
@@
 degrees of function strictness by two degrees of constructor strictness).
-One interesting class of test programs are those which require lazy or non-strict
+One interesting class of test programs are those that require lazy or non-strict
 semantics. For these programs we test not only that the test leads to the expected
 result with non-strict semantics, but also that the program returns a
-\emph{BLACKHOLE} in the case of strict semantics. Students who were Haskell/STG
+\emph{BLACKHOLE} in the case of a stricter semantics. Students who were Haskell/STG
 beginners have contributed substantially to the test suite. We have also found
@@
 
-Another important class of test programs are those which should fail with a
+Another important class of test programs are those that should fail with a
 known error, e.g., a parsing or typing error.  For these programs the test
@@
 We also run each test with various levels of garbage collection enabled. This
-has exposed bugs not only in the garbage collector, but also in the code
-generator and runtime.  Interestingly, in practice garbage collection both
-masks and exposes bugs.
-
-The various enumerations of strictness and garbage collection levels mean that
+has exposed bugs in both the code generator and runtime system.
+Interestingly, in practice garbage collection can both mask and expose bugs.
+
+The various combinations of strictness and garbage collection levels mean that
 we run a quite large number of tests (currently order thousands). However,

### 2016-04-19 pubs/tfp16/paper.tex
random edits

@@
 strictness.\\
-{\color{OliveGreen}
+{\color{OliveGreen} % \bf
 \emph{Draft paper notes:}
@@
 
-\item Hopefully in the spirit of a student-friendly symposium, this draft
-  is liberally annotated with indicators of student contributions, 
-  student preparation, the influence of their involvement on our approach,
-  and other ancillary comments.
-  A final paper would of course be cleaned up.
+\item Hopefully in the spirit of a student-friendly symposium, this draft is
+  liberally annotated with indicators of student contributions, student
+  preparation, the influence of their involvement on our approach, and other
+  incidental and ancillary comments.  A final paper would of course be cleaned
+  up.
 
@@
 checkpoint/restart is reaching its end as data transfer to backing store
-becomes an increasing bottleneck~\cite{DOE:2009}.
+becomes an increasing bottleneck~\cite{Daly:2006,DOE:2009}.
 
@@
 %We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
-%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}.
+%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
 
@@
                    case x of {
-                     TP2 a b ->     --TP2 is the Tuple constructor
+                     TP2 a b ->     --TP2 is the pair constructor
                        case unzip xs of {

### 2016-04-20 pubs/tfp16/paper.tex
random cleanup and edits

@@
 \usepackage{hyperref}
+
+\renewcommand{\floatpagefraction}{0.8}
 %
@@
 \centering
-%\begin{centering}
 \footnotesize % tiny scriptsize footnotesize small
@@
              &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
-             &                &     & \texttt{ } $\mathit{odecl}_n$ \texttt{;} \\
+             &                &     & \texttt{ } $\mathit{odecl}_n$ \\
 %             &                &     & \texttt{ } $x_1$ \texttt{=} $\mathit{obj}_1$ \texttt{;} \\
@@
              &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
-             &                & $|$ & $\texttt{PAP}\ f\ a_1\dots a_n$               & Partial application \\
+             &                & $|$ &\texttt{PAP} $f\ a_1\dots a_n$               & Partial application \\
              &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
 \\
-Object decl. & $\mathit{odecl}$ &     & $x = \mathit{obj}$                           & Simple binding \\
+Object decl. & $\mathit{odecl}$ & ::=  & $x = \mathit{obj}$                           & Simple binding \\
 \\
-Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$ & $i \ge 0$ \\
-\\
-Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} [\texttt{unboxed}] & User-defined data type  \\
-               &                   &     & $C\ x_i =$                       & $i \ge 0, n > 0$         \\
-               &                   &     & $\mathit{con}_1 | \dots |  \mathit{con}_n$ \\
-
-\\
-Program      & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_1$ \texttt{;}            & Object and data defns \\
-             &                 &     & \texttt{ } $\dots$ \texttt{;} \\
-             &                 &     & $\mathit{(o|d)decl}_n$ & 
+%%%Constructor defn. & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$ & $i \ge 0$ \\
+%%%\\
+%%%Datatype defn. &  $\mathit{ddecl}$ & ::= & \texttt{data} [\texttt{unboxed}] & User-defined data type  \\
+%%%               &                   &     & $C\ x_i =$                       & $i \ge 0, n > 0$         \\
+%%%               &                   &     & $\mathit{con}_1 | \dots |  \mathit{con}_n$ \\
+%%%\\
+%%%Program      & $\mathit{prog}$ & ::= & $\mathit{(o|d)decl}_1$ \texttt{;}            & Object and data defns, \\
+%%%             &                 &     & \texttt{ } $\dots$ \texttt{;}                & distinguished \texttt{main}\\
+%%%             &                 &     & $\mathit{(o|d)decl}_n$ & 
+Program      & $\mathit{prog}$ & ::= & $\mathit{odecl}_1$ \texttt{;}            & Object and data defns, \\
+             &                 &     & \texttt{ } $\dots$ \texttt{;}                & distinguished \texttt{main}\\
+             &                 &     & $\mathit{odecl}_n$ & 
 %Program      & $\mathit{prog}$& ::= & $f_1\ =\ \mathit{obj}_1$ \texttt{;}          & $n \ge 1$, distinguished \texttt{main}\\
@@
 \end{tabular}
-%\end{centering}
 \caption{STG syntax}
@@
 is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
-paper~\cite{Marlow:2004}, to which we will subsequently refer as \emph{EA}\@.  A
-\emph{BLACKHOLE}, signifying a \emph{THUNK} being evaluated, is only a
-runtime entity and not part of the language.
+paper~\cite{Marlow:2004}, to which we will subsequently refer as \emph{EA}\@.
+A \emph{BLACKHOLE} object, signifying a \emph{THUNK} being evaluated, is only
+a run-time entity and not part of the language.  Omitted is the syntax for
+algebraic data type definitions which is given in
+Appendix~\ref{app:minihaskell}.
 
@@
 question is whether application is strict, or function call is strict.  These
-are in fact distinguishable because (in Haskell) the first argument to
-\texttt{seq} might evaluate to a \emph{PAP}.
+are in fact distinguishable because (in Haskell, for example) the first
+argument to \texttt{seq} could evaluate to a \emph{PAP}.
 
@@
 arity number of arguments, it is called-with-return with its arity number of
-arguments, and the resulting value applied to the remaining arguments.  When
-considering more strict semantics call-with-return is even more convenient
-because arguments can be evaluated in a tight loop.  Unfortunately these uses
+arguments, and the resulting value applied to the remaining arguments.  
+%
+Call-with-return becomes even more convenient for strict evaluation: arguments
+can be evaluated in a tight loop, and similarly for strict constructors, if
+the STG generator (human or otherwise) has not arranged for them to be already
+evaluated.
+%When considering more strict semantics call-with-return is even more
+%convenient because arguments can be evaluated in a tight loop.
+
+Unfortunately these uses
 of call-with-return can result in the C stack growing without bound.  Our
@@
 6.1~\cite{PJ-stockhardware}).
+%For \emph{direct calls}, in which a known function is applied a number of
+%arguments equal to its arity, \emph{stgApply} need not be invoked: the
+%\emph{FUN} can be tail-called directly.
 
@@
 functions, and applications of known functions to other than their arity
-number of arguments.  This family is parameterized by the number of actual
-arguments in the application and whether each is boxed or unboxed.  Because we
-have an explicit stack we are able to use a single C \emph{stgApply} function
-taking a contiguous block of machine words (unboxed values or pointers) on the
-explicit stack.  Some modifications are required, however, because \emph{EA}
-uses call-with-return.  Specifically, when a \emph{FUN} or is \emph{PAP}
-applied to more than its arity number of arguments a call-with-return to the
-(underlying) \emph{FUN} is made with the arity number of arguments, then the
-appropriate \emph{stgApply} function is jumped to, passing the newly returned
-object and the remaining arguments.  Call-with-return becomes even more
-convenient for strict evaluation:  arguments can be evaluated in a tight
-loop.
+number of arguments.  This family is parameterized by the number of arguments
+in the application and whether each is boxed or unboxed.  Because we have an
+explicit stack we are able to use a single C \emph{stgApply} function taking a
+contiguous block of machine words (unboxed values or pointers) on the explicit
+stack.  Some modifications are required, however, because \emph{EA} uses
+call-with-return.  Specifically, when a \emph{FUN} or is \emph{PAP} applied to
+more than its arity number of arguments a call-with-return to the (underlying)
+\emph{FUN} is made with the arity number of arguments, then the appropriate
+\emph{stgApply} function is jumped to, passing the newly returned object and
+the remaining arguments.  Call-with-return becomes even more convenient for
+strict evaluation: arguments can be evaluated in a tight loop, and similarly
+for strict constructors, if the STG generator (human or otherwise) has not
+arranged for them to be already evaluated.
 
@@
 % breaks the page limit
-For convenience we allow general recursive
-\texttt{let}s that are implicitly factored in the usual way to maximize let-polymorphism;
-this would normally already be performed by a front-end.
+For convenience we allow general recursive \texttt{let}s that are implicitly
+factored in the usual way to maximize let-polymorphism; this would normally
+already be performed by a front-end.
 %\end{comment}
@@
 An almost unbounded amount of interesting development suggests itself; one
-such path could in principle lead to the recreation of GHC\@.  However, given the minuscule
-budget (and therefore the employment of student interns) we must maintain
-sharp focus on the main goals: implementing and evaluating mechanisms for
-automatic parallelism and fault tolerance.  At the time of writing we have a
-fully functioning serial implementation including
+such path could in principle lead to the recreation of GHC\@.  However, given
+the minuscule budget (and therefore the employment of student interns) we must
+maintain sharp focus on the main goals: implementing and evaluating mechanisms
+for automatic parallelism and fault tolerance.  At the time of writing we have
+a fully functioning serial implementation including
 \begin{itemize}
@@
 
-The \emph{mini-Haskell} front-end is detailed in Appendix~A.
+The \emph{mini-Haskell} front-end is detailed in Appendix~\ref{app:minihaskell}.
 
 This summer (2016) a returning student will bridge the gap between GHC STG (or
-Core) and our STG, augmenting the runtime as needed.  Just as in the Intel
-Haskell Research Compiler effort we will not use the GHC runtime.
+Core) and our STG, augmenting the runtime as needed.  Another student will
+develop an instrumentation infrastructure
+concurrently with getting up to speed in functional programming.
+%Just as in the Intel
+%Haskell Research Compiler effort we will not use the GHC runtime.
 
@@
 non-strictness they require to terminate correctly, and to provide other
-logging and debugging information.  This is detailed in Appendix~B.
+logging and debugging information.  This is detailed in Appendix~\ref{app:testing}.
 
@@
 The Disciplined Disciple Compiler project appears to intend implementation of
-a strict, higher order functional language (or family of languages) compiler
-that has been underway for some time, but the emphasis appears to be on a much
+a strict, higher order, functional language (or family of languages) compiler
+and has been underway for some time, but the emphasis appears to be on a much
 more complex type system for handling stateful computation, and the prospects
@@
 \section{Mini-Haskell}
-
+\label{app:minihaskell}
 As mentioned, a current near-term goal is to use GHC as our front-end to
@@
 user-defined unboxed data types, e.g.,
+{\footnotesize
 \begin{verbatim}
   data unboxed Bool# = False# | True# .
-\end{verbatim}
+\end{verbatim}}
 
@@
 
+{\footnotesize
 \begin{verbatim}
@@
                   in result })
-\end{verbatim}
+\end{verbatim}}
 In mini-Haskell, expressions may be applied to other expressions, so the
@@
 \texttt{let} expression:
+{\footnotesize
 \begin{verbatim}
@@
                  Cons hd tl -> Cons hd (append tl l2)
-\end{verbatim}
+\end{verbatim}}
 This also serves as a good example of how expression to expression application is
@@
 Even relatively simple functions can be painful to write, such as \texttt{unzip}.
+{\footnotesize
 \begin{verbatim}
@@
                            in res }}})
-\end{verbatim}
-In mini-Haskell, this becomes
+\end{verbatim}}
+\noindent In mini-Haskell, this becomes
+{\footnotesize
 \begin{verbatim}
@@
     TP2 as bs -> TP2 (Cons a as) (Cons b bs)
-\end{verbatim}
+\end{verbatim}}
 
@@
 
-\item \emph{Type annotations.}
-  The ability to assert the type of some variable is often valuable as a
-  sanity-check and provides documentation that the compiler can verify.
-  mini-Haskell's type annotations (or signatures) behave like those in
-  Haskell, being passed to the type checker as initial assumptions.
+\item \emph{Type annotations.}  The ability to assert the type of some
+  variable is often valuable as a sanity-check and provides documentation that
+  the compiler can verify.  Mini-Haskell's type annotations (or signatures)
+  behave like those in Haskell, being passed to the type checker as initial
+  assumptions.
 
@@
   compiled with another Haskell compiler.  Because mini-Haskell exposes a
-  different set of primitives, the Haskell Block was introduced, to allow the
+  different set of primitives the Haskell Block was introduced to allow the
   inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
@@
 \section{Test suite and testing}
-
+\label{app:testing}
 We have developed an extensive test suite which uses the CMake CTest
@@
 
-We have also implemented a simple logging facility that can be controlled at 
+We have also implemented a simple logging facility which can be controlled at 
 both compile and run time which helps in determining why a test is failing.

### 2016-04-21 pubs/tfp16/paper.tex
random edits

@@
 However, they see their needs as different than other Haskell users: performance
-is paramount, array/vector~\cite{vector} is a (perhaps \emph{the}) primary data
+is paramount, array/vector
+%~\cite{vector} 
+is a (perhaps \emph{the}) primary data
 structure, and non-strictness or laziness is more of a nuisance---even a
@@
 inexperienced functional programmers make essential use of
-non-strictness.\footnote{The starting experience of our student interns to
-  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
-    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
-  substituting his article \emph{Higher-order Functions for
-    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
-  book.}
+non-strictness.%\footnote{The starting experience of our student interns to
+%  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
+%    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
+%  substituting his article \emph{Higher-order Functions for
+%    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
+%  book.}
 
@@
 generally, the operational notion of allocating a heap object is explicit:
-this is the operational semantics of \texttt{let} bindings.
+this is the operational semantics of \texttt{let} bindings.  Numeric literals
+are considered constructors in pattern matching.
 %As such a variable of unboxed type cannot be let-bound; instead it can be
@@
 Greater strictness implies more opportunities for automatic parallelism,
-possible losses or gains in space efficiency, and less expressiveness.  We
+possible losses or gains in space and time efficiency, and less expressiveness.  We
 seek to explore the interplay of all of these.
@@
 
-This summer (2016) a returning student will bridge the gap between GHC STG (or
-Core) and our STG, augmenting the runtime as needed.  Another student will
-develop an instrumentation infrastructure
-concurrently with getting up to speed in functional programming.
-%Just as in the Intel
-%Haskell Research Compiler effort we will not use the GHC runtime.
+We use a continuous integration system augmented with logic to categorize
+programs in a large and growing test suite according to which degrees of
+non-strictness they require to terminate correctly, and to provide other
+logging and debugging information.  This is detailed in Appendix~\ref{app:testing}.
 
@@
 
-We use a continuous integration system augmented with logic to categorize
-programs in a large and growing test suite according to which degrees of
-non-strictness they require to terminate correctly, and to provide other
-logging and debugging information.  This is detailed in Appendix~\ref{app:testing}.
-
-% For various reasons we have chosen to make our implementation 64-bit only.
-
 For simplicity the implementation is 64-bit only.
@@
 
+This summer (2016) a returning student will bridge the gap between GHC STG (or
+Core) and our STG, augmenting the runtime as needed.  Another student will
+develop an instrumentation infrastructure integrated with the testing
+framework to record and present runtime statistics such as number of heap and
+stack allocations and maximum stack depth.
+
 %We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
 %small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
-
 %\cite{Harris:2005,Marlow:2008,Marlow:2009}
@@
 work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
-planned approach they use GHC as the front end, but exit the GHC toolchain
-at the final GHC Core stage, just before transformation to STG.  The essence
-of their effort is to target an existing, highly optimizing, Intel functional
-language compiler that is ``largely language agnostic'' and performs (among
-many other optimizations) SIMD vectorization.
+planned approach they use GHC as the front end, but exit GHC at the final GHC
+Core stage, just before transformation to STG.  The essence of their effort is
+to target an existing, highly optimizing, Intel functional language compiler
+that is ``largely language agnostic'' and performs (among many other
+optimizations) SIMD vectorization.
 
@@
 \end{verbatim}}
-In mini-Haskell, expressions may be applied to other expressions, so the
+In mini-Haskell expressions may be applied to other expressions, so the
 recursive call to \texttt{append} can be parenthesized. There is no need for the
@@
 \end{verbatim}}
-\noindent In mini-Haskell, this becomes
+\noindent In mini-Haskell this becomes
 {\footnotesize
@@
 
+\section{Student Development}
+
+Student interns' backgrounds in computer programming has ranged from minimal
+to some exposure to one or more mainstream imperative languages, Lisp, and
+Haskell.  To bring them to a normalized level of proficiency their first
+assignment has been to work through the entirety of Hutton's \emph{Programming
+  in Haskell} textbook~\cite{Hutton-book}, substituting one of his articles
+\emph{Higher-order Functions for Parsing}~\cite{Hutton-parsing:1992} or
+\emph{Monadic Parser Combinators}~\cite{Hutton:1996} for the corresponding
+material in the book.  This may be followed by writing a parser and ADT for a
+progressively enriched lambda calculus, together with call-by-value,
+call-by-name, and (in one case) call-by-need interpreters.  The concept of
+compilation can be illustrated with an exercise in parsing infix arithmetic
+expressions to an AST, and transforming that to a sequence of operations for a
+simple stack machine.
+% breaks page limit
+%  This learning phase can take the
+%better part of a summer or one-semester internship, so we particularly
+%seek return interns.
+
 %

### 2016-04-21 runtime/cmm.h

@@ -41,7 +41,7 @@ typedef FnPtr (*CmmFnPtr)();
     return ((FnPtr)f);				\
   } while (0)
 
-#else
+#else  // not trampoline
 
 typedef void FnPtr;
 typedef FnPtr (*CmmFnPtr)();

### 2016-04-22 pubs/tfp16/paper.tex
interim or final paper submission, deadline is Monday

@@
         Loren Anderson\inst{1,3} \and
-        Jacob Marks\inst{1,4}\thanks{Ringo, Anderson, and Marks are undergraduate students.}
+        Jacob Marks\inst{1,4}%\thanks{Ringo, Anderson, and Marks are undergraduate students.}
 }
@@
 We do not consider speculative evaluation, or semantic strictness inferred by
-program analysis, so potential parallelism is dictated by the degree of
-strictness.\\
+program analysis, so potential parallelism is dictated by the specified degree of
+strictness.
+
+\hspace{0.2in} Our development has been greatly informed by the extant
+literature and we indicate both where we follow, and deviate from,
+these sources.
+%
+\\
 {\color{OliveGreen} % \bf
@@
   incidental and ancillary comments.  A final paper would of course be cleaned
-  up.
+  up and potentially considerably shortened.
 
 \item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
-  (undergraduate!) have contributed considerably to this effort, and to the
+  (all undergraduate) have contributed considerably to this effort, and to the
   writing of this paper, and will continue to do so.  If this paper is
@@
 \keywords{Strict pure functional programming, automatic parallelization,
-  transparent fault tolerance, Haskell, STG, GHC, Core.}
+  transparent fault tolerance, Haskell, STG, GHC, GHC Core.}
 
@@
 a single system) within the next
-decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a highly respected
+decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a respected
 member of the HPC community recently stated publicly that ``Fortran is
@@
 programmers is the need to get strictness just right: to remove space leaks,
-achieve acceptable performance, get I/O and inter-process communication to work
+achieve competitive performance, get I/O and inter-process communication to work
 as they expect, and so on.  More specifically, they expect to be able to
@@
 Our initial requirements are that
-%
 \begin{itemize}
@@
 analysis and transformation on Core; (4) transformation of Core to the
-\emph{STG} intermediate representation~\cite{PJ-stockhardware}; (5)
+\emph{STG} intermediate representation~\cite{PJ-stockhardware}; and (5)
 transformation of STG to machine code via various imperative
@@
 possible losses or gains in space and time efficiency, and less expressiveness.  We
-seek to explore the interplay of all of these.
+seek to explore the interplay of all of these in various ways.
 
@@
 the distinction between constructors and functions is somewhat blurred
-%because constructors behave like functions 
-(outside of data definitions and pattern matching), but in STG the distinction
+(outside of data definitions and pattern matching) but in STG the distinction
 is clear.
@@
 in the function definition, thus \texttt{id x = x} has by definition arity one
-regardless of the type of the argument (e.g., some function type) it might be
-applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
-a syntactic notion, from actual function \emph{call}. Thus \texttt{id id x} is
-an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
-(user- or system- defined function) when called is given exactly one
-argument (the first one).
-
-In the eval-apply model, given application \texttt{f x y}, in the standard
-(non-strict) semantics \texttt{f} is first evaluated and the arity of the
-underlying \emph{FUN} or \emph{PAP} (partial application of a \emph{FUN}) is
-determined---in the case of \emph{PAP} the arity is that of the underlying
-\emph{FUN} less the number of arguments previously provided.  If the arity is
-greater than two a new \emph{PAP} is created with arguments \texttt{x} and
-\texttt{y} inserted and the new \emph{PAP} returned.  If the arity is equal to
-two the underlying \emph{FUN} is tail-called (jumped to) with any previous
-arguments (in case of \emph{PAP}) and new arguments \texttt{x} and \texttt{y}.
-If the arity is less than two (it must be at least one, so one in this case)
-the underlying \emph{FUN} is called with any previous arguments (in case of
-\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
-remaining argument \texttt{y}.
+regardless of the type of the argument it might be applied to (e.g., some
+function type).  We distinguish \emph{application}, e.g., \texttt{f x y},
+which is a syntactic notion, from actual function \emph{call}. Thus \texttt{id
+  id x} is an application of \texttt{id} to two arguments, but the underlying
+\emph{FUN} (user- or system- defined function) when called is given exactly
+one argument (the first one).
+
+In the eval-apply model, given application \texttt{f x y}, in the general case, with standard
+(non-strict) semantics, \texttt{f} is first evaluated and the
+arity of the underlying \emph{FUN} or \emph{PAP} (partial application of a
+\emph{FUN}) is determined---in the case of \emph{PAP} the arity is that of the
+underlying \emph{FUN} less the number of arguments previously provided.  If
+the arity is greater than two a new \emph{PAP} is created with arguments
+\texttt{x} and \texttt{y} inserted and the new \emph{PAP} returned.  If the
+arity is equal to two the underlying \emph{FUN} is tail-called (jumped to)
+with any previous arguments (in case of \emph{PAP}) and new arguments
+\texttt{x} and \texttt{y}.  If the arity is less than two (it must be at least
+one, so one in this case) the underlying \emph{FUN} is called with any
+previous arguments (in case of \emph{PAP}) and new argument \texttt{x}, then
+the object returned applied to remaining argument \texttt{y}.
 
@@
 (i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
-both of those arguments possibly forced in subsequent evaluation (assuming
-that they weren't forced in the evaluation of \texttt{f} itself).
-
-In a strict semantics one could devise various evaluation orders, but as far
-as distinguishing termination properties it seems that these can be distilled
+both of those arguments possibly forced in subsequent evaluation, assuming
+that they weren't forced in the evaluation of \texttt{f} itself.
+
+In a more strict semantics one could devise various evaluation orders, but as far
+as distinguishing termination properties it seems that these come
 down to two reasonable possibilities, namely whether on construction of a
@@
 The STG machinery makes the choice of evaluation strategy easy to change,
-though changing code generation can be more efficient by avoiding the
+though changing STG generation can be more efficient by avoiding the
 unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.
@@
 number of arguments.  There the object to be applied is first evaluated with
-call-with-return.  When a resulting \emph{FUN} is applied to more than its
-arity number of arguments, it is called-with-return with its arity number of
-arguments, and the resulting value applied to the remaining arguments.  
+call-with-return.  When a resulting \emph{FUN} or \emph{PAP} is applied to
+more than its arity number of arguments, it is called-with-return with its
+arity number of arguments, and the resulting value applied to the remaining
+arguments.
 %
@@
 Type information is required for code generation.  For example, at the
-coarsest level it is necessary to know whether values are boxed, and so
-represented by pointers, or unboxed, to avoid expensive tagging.  Because we
+coarsest level it is necessary to know whether values are boxed, or otherwise
+represented by pointers, to avoid expensive tagging.  Because we
 do not yet get type information directly from GHC we use a subset of Haskell
@@
 %
-The current, primitive parallel runtime is based on Pthreads and a non-blocking
-work queue~\cite{Michael:1996}.  We are currently developing a more realistic
-parallel runtime, experimenting with using the Argobots user-level thread library~\cite{Seo:2015}.
+The current, primitive parallel runtime is based on Pthreads and a
+non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
+more realistic parallel runtime, experimenting with using the Argobots
+user-level thread library~\cite{Seo:2015}.
 %
@@
 
-Previously cited papers describing some of the inner workings of GHC at
-various points in its evolution have greatly informed our approach.
+\emph{Nihil sub sole novum}---there is nothing new under the sun (Ecclesistes 1:9).
+%every new idea has some sort of precedent or echo from the past~\cite{nihil}.
+
+%Previously cited papers describing some of the inner workings of GHC at
+%various points in its evolution have greatly informed our approach.
 
@@
 pure functional language over a non-strict one is by Sheard, who also provided
-a very simple prototype implementation~\cite{Sheard:2003}.
+a very simple embedded implementation~\cite{Sheard:2003}.
 %
@@
 
-Don Stewart described an implementation wherein GHC was used as a Haskell compiler
-front-end, dumping STG code as text, parsing that text, then translating to
-code for execution on the Java Virtual Machine~\cite{Stewart-BSc}.
+Don Stewart described an implementation wherein GHC was used as a Haskell
+compiler front-end, dumping STG code as text, parsing that text, then
+translating to Java code for compilation and execution on the Java Virtual
+Machine~\cite{Stewart-BSc}.
 
@@
 
-\section{Student Development}
+\section{Student Preparation}
 
@@
 expressions to an AST, and transforming that to a sequence of operations for a
-simple stack machine.
+simple stack machine.  From this point they have then been able to substantively
+contribute to the project.
+
 % breaks page limit

### 2016-04-25 pubs/tfp16/paper.tex
paper as submitted

@@
 \title{Automatic Parallelization and Transparent Fault Tolerance\\
-       (\emph{New project} paper)}
+       (\emph{Project article})}
 %
@@
 programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  Besides making programs easier to
-reason about, this enables automatic parallelization to various degrees.  A
-second claimed trend in both scientific computing and functional programming
-is transparent fault tolerance.  Our project goal is the demonstration of a
-light-weight, higher-order, polymorphic, pure functional language
-implementation in which we can experiment with automatic parallelization
-strategies, varying degrees of strictness, and mechanisms for transparent
-fault tolerance.
+functional programming, in various guises.  Besides making large programs
+easier to reason about, this enables automatic parallelization to various
+degrees.  A second claimed trend in both scientific computing and functional
+programming is the development of mechanisms for transparent fault tolerance.
+Our project goal is the demonstration of a light-weight, higher-order,
+polymorphic, pure functional language implementation in which we can
+experiment with automatic parallelization strategies, varying degrees of
+strictness, and mechanisms for transparent fault tolerance.
 %
@@
   implementation, and are currently developing a more realistic parallel
-  runtime.  We expect to have preliminary results from our new parallel
-  implementation by the time of the symposium.  The fault-tolerance mechanisms
-  remain future work.
+  runtime.  
+%We expect to have preliminary results from our new parallel
+%implementation by the time of the symposium.  
+The fault-tolerance mechanisms remain future work.
 
@@
 runtime can dynamically calculate the data dependency graph and relax the
-programmatically defined serial ordering of task execution.  Very high
-parallel efficiency has been demonstrated on extremely large computing
-systems~\cite{Bauer14}.\footnote{We note that the PI of the Legion project was
-  previously a co-developer of a strict, arguably pure functional language
+programmatically defined serial order of task execution to a partial, and
+therefore parallelizable, order.  Very high parallel efficiency has been
+demonstrated on extremely large computing systems~\cite{Bauer14}.\footnote{We
+  note that the PI of the Legion project was previously a co-developer of a
+  strict, arguably pure functional language
   implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
@@
 \subsubsection{Use a pure functional language.}
-Some twenty years after the US Department of Energy laboratory complex was
+Twenty years ago the US Department of Energy laboratory complex was
 presented with the most highly parallel (and thereby performant)
-implementation of a pure functional language to date~\cite{Davis96}, to no
-effect whatsoever, there is now a small number of laboratory HPC practitioners
+implementation of a pure functional language to date, to no
+effect whatsoever~\cite{Davis96}.\footnote{In 1996, on the LANL Thinking Machines
+CM-5, the Top 1 supercomputer in 1993~\cite{top1:1993}.}
+Today, however, there is a small number of laboratory HPC practitioners
 who are developing an appreciation for pure functional programming in Haskell.
@@
 they've read it or not)~\cite{Hughes89}, they would also be happy to write in
-a slightly modified style that does not require non-strictness---after all,
-that's how they've been thinking their entire careers.
+a slightly modified style that does not require non-strictness 
+%
+and so that their programs behave as they expect.
+%---after all, that's how they've been thinking their entire careers.
 
@@
 The concept of a space leak, for example, caused by other than a missing
-explicit deallocation statement (C \emph{free}, etc.) is difficult to
-understand and not something they care to need to understand.
+explicit deallocation statement (C \texttt{free}, Fortran \texttt{DEALLOCATE}, 
+etc.) is difficult to understand and reason about.
+%and not something they care to need to understand.
 %Hence, we eschew explicit strictness annotations for varying uniform degrees
@@
 been some success with transparent (invisible to the programmer) systems.
-Wiring checkpoint/restart into a large-scale application is not only
+Manually wiring checkpoint/restart into a large-scale application is not only
 burdensome and error-prone, it can severely warp the engineering of the
@@
 system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
-been recognized that the side-effect-free parts of a functional program
-(excluding, e.g., the I/O monad in Haskell) could be safely replicated or
-restarted.  The recent work of Stewart impressively demonstrates this, in a
-pure functional setting, on a large-scale distributed-memory
-system~\cite{Stewart:2013}.
+been recognized that the side-effect-free (or revocably side-effecting)
+parts of a functional program could be safely replicated or restarted.  The
+recent work of Stewart impressively demonstrates this, in a pure functional
+setting, on a large-scale distributed-memory system~\cite{Stewart:2013}.
 
@@
 \subsection{Strictness and Parallelism}
-It is a trivial observation that given strict functional semantics, wherein
-function arguments may be safely evaluated before function evaluation, it is
-safe to evaluate the arguments, and the function, in parallel, and that this
-principle applies recursively in the (dynamic) expression tree (or graph).
+%It is a trivial observation that given strict functional semantics, wherein
+%function arguments may be safely evaluated before function evaluation, it is
+%safe to evaluate the arguments, and the function, in parallel, and that this
+%principle applies recursively in the (dynamic) expression tree (or graph).
+
+Given strict functional semantics wherein function arguments may be safely
+evaluated before function evaluation or call, it is safe to evaluate the
+arguments, and the function, in parallel, and that this applies recursively in
+the (dynamic) expression tree (or graph), and similarly for data constructors
+as functions.
 
@@
 means that \emph{we are not attempting to compete with GHC in any
-  respect}---succumbing to the temptation of recreating GHC language, type
-system, and runtime features, or significant code optimization, would simply
-divert us from our central goals.\footnote{We estimate that we will have, to
-  within an order of magnitude, one person-hour of total available development
-  time for each person-year that GHC currently represents, or put another way,
-  one second for each hour, respectively.}  For another, it means that we are
-not undertaking significant morphing of GHC itself: because this is in large
-part an undergraduate student project it must be kept as simple as reasonably
-possible.
+  respect}---succumbing to the temptation of recreating a significant subset
+of GHC language, type system, or runtime features, or significant code
+optimization, would simply divert us from our central goals.\footnote{We
+  estimate that we will have, to within an order of magnitude, one person-hour
+  of total available development time for each person-year that GHC currently
+  represents, or in other units, one second for each hour.}
+For another, it means that we are not undertaking significant morphing of GHC
+itself: because this is in large part an undergraduate student project it must
+be kept as simple as reasonably possible.
 
@@
   should be preserved for all degrees of strictness;
-\item Tail calling must be properly and consistently implemented;
-\item The generated code is C.
+\item The generated code is C;
+\item A certain consistency in tail calling in C must be achieved.
 \end{itemize}
@@
 inexperienced functional programmers make essential use of
-non-strictness.%\footnote{The starting experience of our student interns to
+non-strictness and laziness.
+%\footnote{The starting experience of our student interns to
 %  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
@@
 though changing STG generation can be more efficient by avoiding the
-unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.
+unnecessary creation of \emph{THUNK}s as discussed in Appendix~A.
 
@@
 maintaining our own data and control (continuation) stack all generated C
-functions can take a uniform (including empty) set of explicit arguments and
-have uniform return type. If we arrange that generated C code has only calls
-(other than runtime or other external calls, which do not call back to
-user-program code) in tail call positions, that is, immediately before a
-return in the control flow graph, then at non-zero optimization levels both
-GCC and Clang/LLVM will generate jumps instead of calls-with-return for C
-calls.  When using GCC we can also use an explicit set of machine registers
-for passing/returning values, much like GHC using the LLVM back-end with its
-custom GHC calling convention.
+functions can take a uniform (including empty) set of explicit arguments
+(passed in registers) and have uniform return type. If we arrange that
+generated C code has only calls (other than runtime or other external calls,
+which do not call back to user-program code) in tail call positions, that is,
+immediately before a return in the control flow graph, then at non-zero
+optimization levels both GCC and Clang/LLVM will compile C calls to jumps
+instead of calls-with-return.  When using GCC we can also use an explicit set
+of machine registers for passing/returning values, much like GHC using the
+LLVM back-end with its custom GHC calling convention.
 
@@
 functions, and applications of known functions to other than their arity
-number of arguments.  There the object to be applied is first evaluated with
-call-with-return.  When a resulting \emph{FUN} or \emph{PAP} is applied to
-more than its arity number of arguments, it is called-with-return with its
-arity number of arguments, and the resulting value applied to the remaining
-arguments.
+number of arguments.  There the object to be applied, if a \emph{THUNK}, is
+first evaluated with call-with-return.  When a (resulting) \emph{FUN} or
+\emph{PAP} is applied to more than its arity number of arguments, it is
+called-with-return with its arity number of arguments, and the resulting value
+applied to the remaining arguments.
 %
@@
 
-Unfortunately these uses
-of call-with-return can result in the C stack growing without bound.  Our
-solution uses a single C \emph{stgApply} function with no calls-with-return.
-This is straightforwardly realizable because the stack is explicit, and with
-modern C compiler call optimization, the continuation style enables a jump
-(\emph{go to}) mechanism where the \emph{code labels}---entities that name
-blocks of code, can be treated as first class (e.g., stored for later use),
-and serve as targets of a jump---are C function addresses (c.f.\ \S
-6.1~\cite{PJ-stockhardware}).
+These uses of call-with-return, which in general cannot be inlined, grow the C
+stack, and without bound.  Our solution uses a single C \emph{stgApply}
+function and an auxiliary function with no calls-with-return.  This is
+straightforwardly realizable because the stack is explicit, and with modern C
+compiler call optimization the continuation style enables a jump mechanism
+where the \emph{code labels}---entities that name blocks of code, can be
+treated as first class (e.g., stored for later use), and serve as targets of a
+jump---are C function addresses (c.f.\ \S 6.1~\cite{PJ-stockhardware}).
 %For \emph{direct calls}, in which a known function is applied a number of
@@
 \item continuous integration and testing system,
-\item extensive test suite \emph{(mostly by students\footnote{Many are
+\item extensive test suite \emph{(mostly by students)}.\footnote{Many are
     students' solutions to exercises in learning algorithms and data
-    structures.})}.
+    structures.}
 \end{itemize}
@@
 develop an instrumentation infrastructure integrated with the testing
-framework to record and present runtime statistics such as number of heap and
-stack allocations and maximum stack depth.
+framework to record and present time and space run-time statistics such as
+number of heap and stack allocations and maximum stack depth.
 
@@
 
-\emph{Nihil sub sole novum}---there is nothing new under the sun (Ecclesistes 1:9).
+\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
 %every new idea has some sort of precedent or echo from the past~\cite{nihil}.
@@
 Science Undergraduate Laboratory Internship (DOE SULI) program.
-%
+
 Los Alamos National Laboratory is managed and operated by Los Alamos National
@@
 inferred from context by the compiler so these keywords are not necessary.
-Functions can then be defined in Haskell style with the parameters following
+Functions can then be defined in Haskell style with the formal parameters following
 the function's identifier on the left side of the equality. With these simple
@@
 expressions, so the translation is straightforward: \texttt{FUN} objects are
-identifiable by the inclusion of these parameters.  \texttt{PAP} objects are
+identifiable by the inclusion of these parameters.  A \texttt{PAP} object is
 created when a known function is partially applied.  \texttt{CON} objects are
@@
 
-In STG, function application must be to atomic values, i.e., either literals or
+In STG function application must be to atomic values, i.e., either literals or
 variables.  This often requires introducing bindings with either \texttt{let} or
 \texttt{case} expressions.  Consider the \texttt{append} function that
-concatenates its two list parameters.  In STG, it is written as
-
+concatenates its two list parameters, which can be expressed in STG by
 {\footnotesize
@@
 \end{verbatim}}
-In mini-Haskell expressions may be applied to other expressions, so the
+\noindent In mini-Haskell expressions may be applied to other expressions, so the
 recursive call to \texttt{append} can be parenthesized. There is no need for the
@@
 \end{verbatim}}
-This also serves as a good example of how expression to expression application is
+\noindent This also serves as a good example of how expression to expression application is
 transformed into STG.  Any non-atomic expression in an application expression is
@@
 single pattern.
-
-Even relatively simple functions can be painful to write, such as \texttt{unzip}.
+%
+Even relatively simple functions can be painful to write, such as \texttt{unzip}:
 {\footnotesize
@@
 
-There are several hundred small STG and mini-Haskell programs in the test suite, 
-which is partitioned into a number of directories which control what
-testing is done. Given our goal of exploring various
-levels of strictness, most tests are run with all evaluation strategies (three
-degrees of function strictness by two degrees of constructor strictness).
-One interesting class of test programs are those that require lazy or non-strict
-semantics. For these programs we test not only that the test leads to the expected
-result with non-strict semantics, but also that the program returns a
-\emph{BLACKHOLE} in the case of a stricter semantics. Students who were Haskell/STG
-beginners have contributed substantially to the test suite. We have also found
-that the test suite has been of use as a learning tool for these students to
-discover what level of non-strictness was required for their programs,
-a benefit that we did not initially envision. Over time we have developed
-extensive STG and mini-Haskell preludes of common programming patterns to
-aid in the writing of test programs.
+There are several hundred small STG and mini-Haskell programs in the test
+suite, which is partitioned into a number of directories which control what
+testing is done. Given our goal of exploring various levels of strictness,
+most tests are run with all evaluation strategies (three degrees of function
+strictness by two degrees of constructor strictness).  Interesting classes of
+test programs are those that require less-strict or sharing semantics.  For
+these programs we test not only that the test leads to the expected result
+with less strict semantics, but also that the program returns a
+\emph{BLACKHOLE} in the case of a stricter semantics. Students who were
+Haskell/STG beginners have contributed substantially to the test suite. We
+have also found that the test suite has been of use as a learning tool for
+these students to discover what levels of non-strictness are required for their
+programs, a benefit that we did not initially envision. Over time we have
+developed extensive STG and mini-Haskell preludes of common programming
+patterns to aid in the writing of test programs.
 
@@
 simple stack machine.  From this point they have then been able to substantively
-contribute to the project.
+contribute to this project.
 

### 2016-04-29 pubs/hcar16/hcar16.tex
submission to Haskell Community and Activities Report 2015

\documentclass[DIV16,twocolumn,10pt]{scrreprt}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage[final]{hcar}

%include polycode.fmt

\begin{document}

\begin{hcarentry}{Auto-parallelizing Pure Functional Language System}
\report{Kei Davis}
\status{active}
\participants{Dean Prichard, David Ringo, Loren Anderson, Jacob Marks}
\makeheader

%(WHAT IS IT?)

The main project goal is the demonstration of a light-weight, higher-order,
polymorphic, pure functional language implementation in which we can
experiment with automatic parallelization strategies, varying degrees of
default function and constructor strictness.  A secondary goal is to
experiment with mechanisms for transparent fault tolerance.
%
We do not consider speculative or eager evaluation, or semantic strictness
inferred by program analysis, so potential parallelism is dictated by the
specified degree of default strictness and any strictness annotations.

Our approach is similar to that of the
\href{https://dl.acm.org/citation.cfm?id=2503779}{Intel
  Labs Haskell Research Compiler}: to use GHC as a front-end to generate STG
(or Core), then exit to our own back-end compiler.  As in their case we do not
attempt to use the GHC runtime.  Our implementation is \emph{light-weight} in
that we are not attempting to support or recreate the vast functionality of
GHC and its runtime.  This approach is also similar to
\href{http://www.cse.unsw.edu.au/~pls/thesis/dons-thesis.ps.gz}{Don Stewart's}
except that we generate C instead of Java.

%(WHAT IS ITS STATUS? / WHAT HAS HAPPENED SINCE LAST TIME?)
\subsubsection*{Current Status}
Currently we have a fully functioning serial implementation and a primitive
proof-of-design parallel implementation.

%(CAN OTHERS GET IT?)

%(WHAT ARE THE IMMEDIATE PLANS?)
\subsubsection*{Immediate Plans}

We are currently developing a more realistic parallel runtime.
%
Bridging the gap between GHC STG (or Core) to our STG representation will be
undertaken starting June 2016.  An instrumentation framework will be developed
in summer 2016.

\FurtherReading
 A \href{https://github.com/losalamos/APPFL}{project web site} is under construction.

\subsubsection*{Undergraduate/post-graduate Internships}

If you are a United States citizen or permanent resident alien studying
computer science or mathematics at the undergraduate level, or are a recent
graduate, with strong interests in Haskell programming, compiler/runtime
development, and pursuing a spring, fall, or summer internship at Los
Alamos National Laboratory, this could be for you.

We don't expect applicants to necessarily already be highly accomplished
Haskell programmers---such an internship is expected to be a combination of
further developing your programming/Haskell skills and putting them to good
use.  If you're already a strong C hacker we could use that too.

\emph{\bfseries The application process requires a bit of work so don't leave
  enquiries until the last day/month.}\\

\begin{tabular}{l|l}
Term & Application Deadline \\
\hline
Summer 2016  & Closed \\
Fall 2016    & May 31, 2016 \\
Spring 2017  & Approx. July 2016 \\
Summer 2017  & Approx. January 2017 \\
Fall 2017    & Approx. May 2017
\end{tabular}\\

\noindent Email me at kei (at) lanl (dot) gov if interested in more information, and
feel free to pass this along. \\

\end{hcarentry}
\end{document}

### 2016-05-04 pubs/hcar16/hcar16.tex
some minor edit to paper

@@
 Our approach is similar to that of the
-\href{https://dl.acm.org/citation.cfm?id=2503779}{Intel
-  Labs Haskell Research Compiler}: to use GHC as a front-end to generate STG
-(or Core), then exit to our own back-end compiler.  As in their case we do not
-attempt to use the GHC runtime.  Our implementation is \emph{light-weight} in
-that we are not attempting to support or recreate the vast functionality of
-GHC and its runtime.  This approach is also similar to
+\href{https://dl.acm.org/citation.cfm?id=2503779}{Intel Labs Haskell Research
+  Compiler}: to use GHC as a front-end to generate STG (or Core), then exit to
+our own back-end compiler that is itself written in Haskell.  As in their case
+we do not attempt to use the GHC runtime.  Our implementation is
+\emph{light-weight} in that we are not attempting to support or recreate the
+vast functionality of GHC and its runtime.  This approach is also similar to
 \href{http://www.cse.unsw.edu.au/~pls/thesis/dons-thesis.ps.gz}{Don Stewart's}

### 2016-05-26 pubs/tfp16/slides/slides.tex

\documentclass{beamer}

%\usetheme[unclass]{LANL-beta}
\usetheme{LANL-beta}
\usepackage{graphicx}

\title{Automatic Parallelization and Transparent Fault Tolerance (project article)}
%\subtitle{(Project article)}
\author{Kei Davis, Dean Prichard, \\\emph{David Ringo}, Loren Anderson, \\and Jacob Marks\\ \ \\}
%\date[4/26/2016]{Some Random Meeting \\ April 26, 2016}
\date{Trends in Functional Programming, June 8-10, 2016}
\LAUR{LA-UR-16-?????}

\begin{document}

\begin{frame}
\maketitle
\end{frame}

\begin{frame}
\frametitle{Here is a simple frame}
    \begin{itemize}
        \item Here is the first bullet in a standard bulleted list,
            as is customary for a presentation like this
        \begin{itemize}
            \item It includes some sub-bullets
            \item Here they are
        \end{itemize}
        \item Here is another bullet
        \item And here is another, with some math:
            \[ e^{i\theta} = \cos \theta + i \sin \theta \]
    \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Here is a more complicated frame, including some graphics}
    \begin{columns}[t]
        \begin{column}{.48\linewidth}
            \begin{itemize}
                \item Here is a bulleted list that sits alongside a graphic
                \item With a second bullet item
                \item And a third
                \begin{itemize}
                   \item It can have sub-bullets too
                   \item Like this
                \end{itemize}
            \end{itemize}
        \end{column}
        \begin{column}[T]{.48\linewidth}
            \begin{figure}
                \begin{center}
                    \fbox{\includegraphics[width=0.9\linewidth]%
                        {LANL-logo-gray}}\\[2.0ex]
                    insert caption here
                \end{center}
            \end{figure}%
        \end{column}
    \end{columns}
\end{frame}

\end{document}

### 2016-05-27 options.h
refactoring

@@ -3,10 +3,10 @@
 
 /* options used by both C and haskell side for conditional compilation */
 
-/* If 1 use argType enum field in PtrOrLiteral */
+/* 0/1, If 1 use argType enum field in PtrOrLiteral */
 #define USE_ARGTYPE 1
 
-/* If 1 use objType enum field in Obj */
+/* 0/1, If 1 use objType enum field in Obj */
 #define USE_OBJTYPE 1
 
 /* If 1 use .pi = PI header in InfoTab */

### 2016-05-27 runtime/heap.c

@@ -1,5 +1,16 @@
 #include "stg.h"
+#include "heap.h"
 #include <string.h>  // for memcpy()
+
+// this is a temporary hack as we incorporate Bitmap64s into continuations
+/*
+Bitmap64 layoutInfoToBitmap64(LayoutInfo *lip) {
+  Bitmap64 bm;
+  bm.bitmap.mask = (0x1UL << lip->boxedCount) - 1;  // boxed vals first
+  bm.bitmap.size = lip->boxedCount + lip->unboxedCount;
+  return bm;
+}
+*/
 
 // we can't be certain a value is boxed or unboxed without enabling ARG_TYPE
 // but we can do some sanity checking.  mayBeBoxed(v) means that v is not
@@ -21,6 +32,28 @@ bool mayBeUnboxed(PtrOrLiteral f) {
 #else
   return true;
 #endif
+}
+
+Obj *derefPoL(PtrOrLiteral f) {
+  assert(mayBeBoxed(f) && "derefPoL: not a HEAPOBJ");
+  return derefHO(f.op);
+}
+
+void derefStgCurVal() {
+  assert(mayBeBoxed(stgCurVal)); // return stgCurVal
+  while (getObjType(stgCurVal.op) == INDIRECT) { // return stgCurVal
+    stgCurVal = stgCurVal.op->payload[0]; // return stgCurVal
+    assert(mayBeBoxed(stgCurVal)); // return stgCurVal
+  }
+}
+
+Obj *derefHO(Obj *op) {
+  while (getObjType(op) == INDIRECT) {
+    PtrOrLiteral v = op->payload[0];
+    assert(mayBeBoxed(v));
+    op = v.op;
+  }
+  return op;
 }
 
 Obj* stgNewHeapObj(InfoTab *itp) {

### 2016-05-27 runtime/obj.h

@@ -3,6 +3,7 @@
 
 #include "options.h"
 #include "stg.h"
+#include "heap.h"
 // need to factor this out
 #include "stack.h"
 

### 2016-05-31 pubs/tfp16/slides/slides.tex
sync for language-c-quote

@@
 \begin{frame}
+\frametitle{In Search of Automatic Parallelization\\
+\emph{or at least automatic scheduling}}
+  \begin{figure}
+    \begin{center}
+%      \fbox{
+      \includegraphics[width=0.95\linewidth,height=1in]%
+        {figures/S3Dtaskgraph.png}~\footnote{Courtesy Stanford Legion project}
+%}
+\\[2.0ex]  % vertical space
+      S3D task dependency, combustion chemistry calculation
+    \end{center}
+  \end{figure}
+Simplest interesting chemistry, task graph much larger with more complex reactants.
+\vspace{0.1in}
+Schedule by hand?
+\end{frame}
+
+\begin{frame}
+\frametitle{Scientific Computing in our Microcosm}
+Local evolution of scientific computing
+\begin{itemize}
+  \item Serial Fortran programs
+  \item MPI everywhere (inter- and intra-node)
+  \item C, C++
+  \item MPI+X, X is Pthreads, OpenMP, OpenCL, CUDA, etc.
+  \item Parallel runtimes, e.g., Cilk++, Intel Threading Building Blocks, Stanford's Legion, etc.
+\end{itemize}
+
+We have a `new' generation of scientific programmers, aka computational scientists, who
+have some understanding of meaning and virtue of \emph{pure functional}, and even dabble
+in Haskell programming.
+\vspace{0.1in}
+
+\emph{In a multi-100,000-line program, do not temporarily alter the global speed-of-light
+`constant' variable.}
+\end{frame}
+
+\begin{frame}
+\frametitle{Non-strictness/laziness Anathema to Parallelism}
+How to get around?
+\begin{itemize}
+  \item Strictness analysis
+  \item Bang patterns
+  \item Par/pseq, other specifications
+  \item Speculative evaluation
+  \item \ldots
+  \item \emph{Strict(er) default semantics}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
 \frametitle{Here is a simple frame}
@@
 \begin{frame}
-\frametitle{Here is a more complicated frame, including some graphics}
+\frametitle{In Search of Automatic Parallelization\\
+\emph{or at least automatic scheduling}}
     \begin{columns}[t]

### 2016-06-01 codegen/BU.hs
mid freevarer investigation...
The order in which input files are concatenated suddenly matters. I'm
sure I've broken something inadvertently.

@@ -32,16 +32,16 @@ compose :: Subst -> Subst -> Subst
 compose s1 s2 = Map.map (apply s1) s2 `Map.union` s1
 
 class Substitutable a where
-    apply :: Subst -> a -> a
-    freevars   :: a -> Set.Set TyVar
+    apply    :: Subst -> a -> a
+    freevars :: a -> Set.Set TyVar
 
 instance Substitutable Polytype where
     freevars (PPoly tvs mt) = (freevars mt) `Set.difference` (Set.fromList tvs)
     freevars (PMono mt)     = freevars mt
 
-    apply s (PPoly tvs mt)  = PPoly tvs $ apply s' mt
+    apply s  (PPoly tvs mt) = PPoly tvs $ apply s' mt
                               where s' = foldr Map.delete s tvs
-    apply s (PMono mt)      = PMono $ apply s mt
+    apply s  (PMono mt)     = PMono $ apply s mt
 
 instance Substitutable Monotype where
     freevars (MVar v)        = Set.singleton v
@@ -112,8 +112,7 @@ unifys (x:xs) (y:ys) =
         s2 = unifys (map (apply s1) xs) (map (apply s1) ys)
     in compose s2 s1
 
--- need a fresh
-variable supply
+-- need a fresh variable supply
 
 freshTyVar :: State Int TyVar
 freshTyVar =

### 2016-06-01 codegen/CMap.hs

@@ -6,7 +6,7 @@
 {-# LANGUAGE OverlappingInstances #-}
 #endif
 
-#include "../options.h"
+-- #include "../options.h"
 
 module CMap
 (
@@ -34,6 +34,7 @@ import Data.Maybe (fromJust)
 import Data.List ((\\), find, intercalate)
 import Data.Char (isNumber)
 import Debug.Trace
+import Options
 
 type CMap = Map.Map Con TyCon
 

### 2016-06-01 codegen/CodeGen.hs

@@ -343,7 +343,7 @@ cge env (EPrimop it op eas) =
 -- Allocating all objs in ELet before making their payloads valid is somewhat problematic
 -- if GC can be initiated by a single object allocation.  This is exacerbated by strict
 -- constructors.  This is not readily fixed by changing the order of things here because
--- of mutually recursive definitions.  
+-- of mutually recursive definitions.
 -- TODO:
 --   Make runtime more robust
 --   then for strict constructors evaluate the args
@@ -370,10 +370,12 @@ cge env (ELet it os e) =
 cge env ecase@(ECase _ e a) =
     do ((ecode, eypn), efunc) <- cge env e
        -- weird:  compiler requires parens around if, ghci does not
-       (if eypn == No then
+       -- fixed: if-then-else has special layout in do blocks
+       if eypn == No
+         then
             cgeInline env (isBoxede e) (ecode, efunc) a
-        else
-            cgeNoInline env (isBoxede e) (ecode, efunc) a)
+         else
+            cgeNoInline env (isBoxede e) (ecode, efunc) a
 
 -- TODO:  inline
 -- TODO:  YPN from Alts, Alt
@@ -400,9 +402,9 @@ cgaltsInline
      -> Bool
      -> State Int (([Char], YPN), [([Char], [Char])])
 
-cgaltsInline env a@(Alts it alts name) boxed =
+cgaltsInline env a@(Alts it alts name scrt) boxed =
     let contName = "ccont_" ++ name
-        scrutPtr = "scrutPtr_" ++ name
+        scrutPtr = "scrutPtr_" ++ name -- TODO: scrt var here?
         phonyforward = "FnPtr " ++ name ++ "();"
         phonyfun = "FnPtr "++ name ++ "() {}\n"
         switch = length alts > 1
@@ -446,7 +448,7 @@ cgeNoInline
      -> Alts InfoTab
      -> State Int (([Char], YPN), [([Char], [Char])])
 
-cgeNoInline env boxed (ecode, efunc) a@(Alts italts alts aname) =
+cgeNoInline env boxed (ecode, efunc) a@(Alts italts alts aname scrt) =
     let contName = "ccont_" ++ aname
         pre = "// scrutinee may STGJUMP or STGRETURN\n" ++
               "Cont *" ++ contName ++ " = stgAllocCont( &it_" ++ aname ++ ");\n" ++
@@ -470,9 +472,9 @@ cgalts
[…]

### 2016-06-01 codegen/ConvertSTG.hs

@@ -1,9 +1,8 @@
--- #!/usr/local/bin/runhaskell -XCPP -cpp -DREWRITE_STG=1
--- not working: Flags don't seem to be passed properly
 {-# LANGUAGE FlexibleInstances    #-}
 {-# LANGUAGE NamedFieldPuns       #-}
 {-# LANGUAGE TypeSynonymInstances #-}
-
+-- #!/usr/local/bin/runhaskell -XCPP -cpp -DREWRITE_STG=1
+-- not working: Flags don't seem to be passed properly
 
 
 
@@ -21,6 +20,8 @@ import           Parser
 import           PPrint
 import           Rename
 import           SetFVs
+import InfoTab
+import HMStg
 import           State
 import           System.Environment (getArgs)
 import           System.IO
@@ -44,7 +45,7 @@ main =
        files <- getArgs
        mapM_ (uncurry rewriteScruts) . zip files . map (++ ".new") $ files
   else
-    print "run/compile ConvertSTG with -cpp -DREWRITE_STG flags for GHC"
+    error "Only run/compile ConvertSTG.hs with -cpp -DREWRITE_STG flags for GHC"
 
 
 -- ^ Modify an old STG file to update the syntax of case expressions
@@ -61,7 +62,7 @@ rewriteScruts infile outfile =
         env = Set.fromList . map (\(ObjDef o) -> oname o) .
               filter isObj . rights $ eithers
         (modified, n) = runState (asb env eithers) 0
-    print $ show n ++ " case scrutinee bindings added"
+    putStrLn $ show n ++ " case scrutinee bindings added"
     writeFile outfile . show . vcat . map unparse $ modified
 
 
@@ -102,7 +103,7 @@ nextBind env =
 
 type Env = Set.Set Var
 
-
+-- ^ ASB  = Add Scrutinee Binding
 class ASB a where
   -- ^ Traverse STG AST and add unique scrutinee bindings to convert
   -- an old STG file to the new syntax
@@ -190,6 +191,9 @@ instance ASB (Alt a) where
 
 -- Test conversion with Driver in Progress
 --------------------------------------------------
+
+withPrelude x = "../prelude/Prelude.stg.new":[x]
+
 tester :: (String -> a) -> (a -> String) -> [FilePath] -> IO ()
 tester tfun sfun infiles =
  do
@@ -251,3 +255,18 @@ defaultcaser inp = let (tycons, objs) = renamer inp
[…]

### 2016-06-01 codegen/HMStg.hs

@@ -55,11 +55,13 @@ hmstgAssums os0 assums =
 --    in let res = typeFVs Map.empty $ co Set.empty $ backSub subst os2 :: [Obj InfoTab]
     in let res = co Set.empty $ backSub subst os2 :: [Obj InfoTab]
        in if res == res then res else error "this shouldn't happen"
+    -- Note for future reader: res == res forces evaluation and will catch
+    --   any errors that laziness hides
 
 hmstg :: [Obj InfoTab] -> [Obj InfoTab]
 hmstg os = hmstgAssums os Set.empty
 
---{-
+
 hmstgdebug :: [Obj InfoTab] -> IO ()
 hmstgdebug os = hmstgAssumsdebug os Set.empty
 
@@ -73,28 +75,7 @@ hmstgAssumsdebug os0 assums =
     in
 --       putStrLn $ show $ vcat $ map (text . show) $ Set.toList cs
        putStrLn $ showObjs (co Set.empty $ backSub subst os2)
----}
-{-
-hmstgdebug os0 =
-    let (os1,i1) = runState (dv os0) 0 :: ([Obj InfoTab], Int)
-    in putStrLn $ showObjs (co Set.empty os1)
--}
-{-
-hmstgdebug os0 =
-    let (os1,i1) = runState (dv os0) 0
-        (as,cs,os2) = buNest Set.empty os1 :: (Set.Set Assumption,
-                                               Set.Set Constraint,
-                                               [Obj InfoTab])
-        (subst, _) = runState (solve cs) i1
-    in putStrLn $ show (Set.toList as) ++ show (Set.toList cs) ++ showObjs os2
--}
 
-{-
-    in (Set.toList as,
-        Map.toList subst,
-        apply subst $ Set.toList cs,
-        "\n\n" ++ showObjs (backSub subst os2))
--}
 
 -- distribute fresh type variables, monomorphic type variables, for BU in advance
 
@@ -336,10 +317,13 @@ instance BU (Expr InfoTab) where
 
     bu mtvs e = error $ "HMStg.bu Expr: " ++ show e
 
-buNest :: Set.Set Monotype
-       -> [Obj InfoTab]
-       -> Assumptions -- assumptions generated by mhs type signatures
-       -> (Set.Set (String, Monotype), Set.Set Constraint, [Obj InfoTab])
+-- ^ Entry to bottom up generation of Constraints
+buNest
+  :: Set.Set Monotype -- ^ environment
+  -> [Obj InfoTab]    -- ^ Letrec'd Objects to process
[…]

### 2016-06-01 codegen/InfoTab.hs

@@ -7,7 +7,7 @@
 {-# LANGUAGE TupleSections         #-}
 
 {-# LANGUAGE CPP #-}
-#include "../options.h"
+-- #include "../options.h"
 
 module InfoTab(
   InfoTab(..),
@@ -195,12 +195,12 @@ data InfoTab =
       typ :: Monotype,
       ctyp :: Polytype,
       fvs :: [(Var,Monotype)],
-      bfvc :: Int,  -- boxed FV count
-      ufvc :: Int,  -- unboxed FV count
+      bfvc :: Int,          -- boxed FV count
+      ufvc :: Int,          -- unboxed FV count
       truefvs :: [Var],
-      scVar :: Var,
-      name :: String,         -- for C infotab
-      entryCode :: String }   -- for C infotab
+      scVar :: Var,         -- binding for scrutinee
+      name :: String,       -- for C infotab
+      entryCode :: String } -- for C infotab
 
   -- the following may be useful later
   -- for now case continuation is handled by Alts.ITAlts
@@ -441,7 +441,7 @@ instance MakeIT (Alts ([Var],[Var])) where
       where
         scvar = case scrt of
                   EAtom _ (Var v) -> v
-                  _ -> error "InfoTab.makeIT: scbnd should always be EAtom _ (Var _)"
+                  _ -> error "InfoTab.makeIT: scrt should always be EAtom _ (Var _)"
 
 instance MakeIT (Alt ([Var],[Var])) where
     makeIT ACon{amd = (fvs,truefvs)} =
@@ -557,7 +557,8 @@ showIT it@(ITBlackhole {}) =
     "};\n"
 
 -- CONTINUATION CInfoTab
-
+-- TODO: Do something with the scrutinee binding here?
+--   don't think so... only useful as a binding in codegen
 showIT it@(ITAlts{}) =
     "CInfoTab it_" ++ name it ++ " __attribute__((aligned(8))) = {\n" ++
     "  .name                     = " ++ show (name it) ++ ",\n" ++

### 2016-06-01 codegen/MHS/ToSTG.hs

@@ -1,25 +1,24 @@
-{-# LANGUAGE
-NamedFieldPuns,
-TypeSynonymInstances,
-FlexibleInstances,
-MultiParamTypeClasses#-}
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE NamedFieldPuns        #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
 
 module MHS.ToSTG
 (
   makeSTG
 ) where
 
-import MHS.Transform
-import MHS.AST
-import Analysis (defAlt)
-import BU (Assumption, Assumptions)
-import PPrint
-import State
-import qualified Data.Set as Set
-import qualified Data.Map as Map
-import Data.List (isPrefixOf, partition)
-import Data.Function (on)
-import Debug.Trace
+import           Analysis      (defAlt)
+import           BU            (Assumption, Assumptions)
+import           Data.Function (on)
+import           Data.List     (isPrefixOf, partition)
+import qualified Data.Map      as Map
+import qualified Data.Set      as Set
+import           Debug.Trace
+import           MHS.AST
+import           MHS.Transform
+import           PPrint
+import           State
 
 makeSTG :: [Defn] -> ([TyCon], [Obj ()], Assumptions)
 makeSTG defs =
@@ -67,23 +66,23 @@ newMtyp m dict =
       makeDict (MCon _ c1 ms) (MCon _ c2 ns) =
         Map.unions (Map.singleton c1 c2 : zipWith makeDict ms ns)
       makeDict _ _ = Map.empty
-        
+
   in case m of
       MCon b c ms ->
         do
           ms' <- foldy dict ms
           return $ MCon b c ms'
-       
+
       MVar v ->
         do
           v' <- monoVar
           return $ MVar v'
-          
+
       MFun m1 m2 ->
         do
           [m1',m2'] <- foldy dict [m1,m2]
           return $ MFun m1' m2'
-          
+
       m -> return m
 
 
@@ -101,7 +100,7 @@ instance Assume Defn where
          do
            assume oexp
            insertAS (bnd,m)
-               
+
     _ -> error $ unlines
          ["ToSTG.stgAssum given non ODefn arg:",
           show $ pprint def]
@@ -142,9 +141,9 @@ stgTycon ddef =
       _ -> error $ unlines
            ["ToSTG.stgTycon passed non DDefn argument",
             show $ pprint ddef]
[…]

### 2016-06-01 codegen/Parser.hs

@@ -381,8 +381,8 @@ altsP =
   let
     name = "alts" --error "this alts not given a name!"
     scrtP | reWriteSTG = emptyP `using` const (EAtom () $ Var "parsed_rewrite_stg")
-            | otherwise = tokcutP "Expected variable binding the case scrutinee" $
-                          eAtomP
+          | otherwise = tokcutP "Expected variable binding the case scrutinee" $
+                        eAtomP
   in
     scrtP >>> \scrt ->
     tokcutP "Expected left brace to open the alt block of a case expr"
@@ -414,10 +414,10 @@ altP =
 atomP :: Parser Token Atom
 atomP = orExList [
   varNameP `using` Var,
-  intP `using` LitI,
-  --longP `using` LitL,
-  fltP `using` LitF,
-  --dblP `using` LitD,
+  intP     `using` LitI,
+  --longP  `using` LitL,
+  fltP     `using` LitF,
+  --dblP   `using` LitD,
   conNameP `using` LitC
   ]
 

### 2016-06-01 codegen/SetFVs.hs

@@ -155,11 +155,11 @@ instance SetFVs (Expr a) (Expr (Set.Set Var, Set.Set Var)) where
 
     -- EFCall introduces a Var
     setfvs tlds (EFCall _ f as) =
-        let as' = map (setfvs tlds) as
-            (easFVs, easTFVs) = unzip $ map emd as'
+        let as'                = map (setfvs tlds) as
+            (easFVs, easTFVs)  = unzip $ map emd as'
             (asfvs, astruefvs) = (Set.unions easFVs, Set.unions easTFVs)
-            myfvs   = (Set.singleton f `Set.union` asfvs) `Set.difference` tlds
-            truefvs = Set.singleton f `Set.union` astruefvs
+            myfvs              = (Set.singleton f `Set.union` asfvs) `Set.difference` tlds
+            truefvs            = Set.singleton f  `Set.union` astruefvs
         in EFCall (myfvs,truefvs) f as'
 
     setfvs tlds (EPrimop _ p as) =
@@ -197,6 +197,7 @@ instance SetFVs (Alt a) (Alt (Set.Set Var, Set.Set Var)) where
         let vset = Set.fromList vs
             e' = setfvs (Set.difference tlds vset) e
             (efvs, etruefvs) = emd e'
+
             myfvs =   Set.difference efvs     vset
             truefvs = Set.difference etruefvs vset
         in ACon (myfvs,truefvs) c vs e'
@@ -209,14 +210,19 @@ instance SetFVs (Alt a) (Alt (Set.Set Var, Set.Set Var)) where
             truefvs = Set.difference etruefvs vset
         in  ADef (myfvs,truefvs) v e'
 
+-- alts block introduces scope with scrutinee binding
 instance SetFVs (Alts a) (Alts (Set.Set Var, Set.Set Var)) where
     setfvs tlds (Alts _ alts name scrt) =
         let vset = Set.singleton scvar
+            -- I'm not positive there's much meaning to the notion of a
+            -- free variable in the scrt (EAtom type), but it's set here
+            -- for completeness. -dmr
             scrt' = setfvs tlds scrt
             alts' = setfvs tlds alts
             (altsfvls, altstruefvls) = unzip $ map amd alts'
-            myfvs   = Set.unions altsfvls     Set.\\ vset
-            truefvs = Set.unions altstruefvls Set.\\ vset
[…]

### 2016-06-01 options.h

@@ -3,6 +3,10 @@
 
 /* options used by both C and haskell side for conditional compilation */
 
+/* Controls syntax of case expressions and allows main in ConvertSTG
+   to convert old STG files to new syntax */
+#define REWRITE_STG 0
+
 /* If 1 use argType enum field in PtrOrLiteral */
 #define USE_ARGTYPE 1
 

### 2016-06-02 pubs/tfp16/slides/slides.tex
alpha draft slides

@@
 
-\title{Automatic Parallelization and Transparent Fault Tolerance (project article)}
+\title{Automatic Parallelization and Transparent Fault Tolerance (project in progress)}
 %\subtitle{(Project article)}
@@
 \maketitle
+\end{frame}
+
+\begin{frame}
+\frametitle{Scientific Computing in our Microcosm}
+Local evolution of scientific computing
+\begin{itemize}
+  \item Serial Fortran programs
+  \item MPI everywhere (inter- and intra-node)
+  \item C, C++
+  \item MPI+X, X is Pthreads, OpenMP, OpenCL, CUDA, etc.
+  \item Parallel runtimes, e.g., Cilk++, Intel Threading Building Blocks, Stanford's Legion, etc.
+\end{itemize}
+Currently MPI+threading model dominates.
 \end{frame}
@@
 \begin{frame}
-\frametitle{Scientific Computing in our Microcosm}
-Local evolution of scientific computing
-\begin{itemize}
-  \item Serial Fortran programs
-  \item MPI everywhere (inter- and intra-node)
-  \item C, C++
-  \item MPI+X, X is Pthreads, OpenMP, OpenCL, CUDA, etc.
-  \item Parallel runtimes, e.g., Cilk++, Intel Threading Building Blocks, Stanford's Legion, etc.
-\end{itemize}
+  \frametitle{Transparent Fault Tolerance?}
+  \begin{columns}[t]
+    \begin{column}{.48\linewidth}
+      Checkpoint:
+      \begin{enumerate}
+      \item Processes synchronize at predetermined point;
+      \item Stop the world;
+      \item Dump global state;
+      \item Resume computation.
+      \end{enumerate}
+    \end{column}
+    \begin{column}[T]{.48\linewidth}
+      Restart:
+      \begin{enumerate}
+      \item Observe that application has hung/crashed;
+      \item Identify last valid (complete) checkpoint image;
+      \item Re-launch application specifying checkpoint image.
+      \end{enumerate}
+    \end{column}
+  \end{columns}
+\vspace{0.1in}
+General claim:  ``data movement costs more energy than computation.''
 
-We have a `new' generation of scientific programmers, aka computational scientists, who
+Fact 1: In C/R, idle processors may cost more energy than data movement.
+
+Fact 2:  C/R isn't scaling.
+\end{frame}
+
+\begin{frame}
+\frametitle{Pure Functional Semantics}
+Now we have a `new' generation of scientific programmers, aka computational scientists, who
 have some understanding of meaning and virtue of \emph{pure functional}, and even dabble
@@
 `constant' variable.}
+\vspace{0.1in}
+
+\emph{But}, these are not computer scientists:  
+\begin{itemize}
+\item Passing functions as arguments is familiar (since early Fortran);
+\item Implicit space leaks are confusing;
+\item Non-strict evaluation is largely irrelevant, strictness annotations are unintuitive.
+\item Unboxed primitives and arrays of the same (unboxed vector) are the primary data.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Claimed Trends}
+Trends in functional programming
+\begin{itemize}
+  \item Appreciation of the virtues of strict-by-default semantics;
+  \item Transparent fault tolerance.
+\end{itemize}
+Trends in scientific computing
+\begin{itemize}
+  \item Implementation of pure functional computational concepts/components.
+  \item Recognition that checkpoint/restart is increasingly intractable and unscalable.
+\end{itemize}
 \end{frame}
@@
 \frametitle{Non-strictness/laziness Anathema to Parallelism}
-How to get around?
+(Semantic) function strictness $\implies$ \\
+\ \ \ function arguments can be evaluated early, in parallel
+
 \begin{itemize}
@@
 \end{itemize}
+Similarly constructors.
+\\\vspace{0.1in}
+Update:  Strict \texttt{let} bindings.
+\end{frame}
+
+\begin{frame}
+  \frametitle{Musings}
+  Suppose we had a strict(-er) pure functional language implementation than Haskell?
+  \begin{itemize}
+  \item Could we keep a large number of hardware threads (e.g., up to 256) busy with
+    \emph{implicit} parallelism?
+    \begin{itemize}
+    \item With what parallel efficiency?
+      \begin{itemize}
+      \item With different scheduling strategies?
+      \end{itemize}
+    \end{itemize}
+  \item How often do new functional programmers/old scientific programmers
+    make essential use of non-strictness?  \emph{(very rarely)}
+    \begin{itemize}
+    \item Laziness? \emph{(never)}
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Project} 
+A \emph{light-weight} implementation of a pure, higher-order, polymorphic,
+functional language and runtime system with which we can experiment with
+automatic parallelization strategies with varying degrees of language
+strictness, and secondarily, with mechanisms for transparent fault tolerance.
+\begin{enumerate}
+\item Haskell to STG (or Core) via GHC \emph{(todo)};
+\item STG to C \emph{(serial done, parallel in progress)};
+\item Fault tolerance \emph{TBD}.
+\end{enumerate}
+\end{frame}
+
+\begin{frame}
+\frametitle{Reinventing the wheel?}  Why not use Manticore, MultiMLton, F\#,
+multicore OCaml, etc., or even GHC in toto?
+\begin{itemize}
+  \item Very light weight:  easy to instrument, modify;
+  \item Know exactly what is going on under the hood;
+  \item There exist conduits for getting student interns up to speed inexpensively;
+  \item We have a small cadre of scientists already interested in Haskell/GHC.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Technical hurdles:  tail calling and unwinding the stack}
+Following SPJ/STG our control structure is a stack of continuations.  
+\\\vspace{0.1in}
+Each continuation contains an address to \emph{go to next}, not \emph{go back
+  to}, so we want something like \emph{goto} or \emph{tail calling}.
+\\\vspace{0.1in}
+SPJ identifies defines the term \emph{code label} as something that can be
+\begin{itemize}
+  \item used to name an arbitrary block of code;
+  \item can be manipulated, i.e., stored for later use;
+  \item can be used as the destination of a jump (not just call).
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile] % for verbatim
+\frametitle{All the myriad ways in C}
+\begin{itemize}
+\item setjmp/longjmp;
+\item giant switch;
+\item computed goto (gcc extension);
+\item trampoline, the simplest of which parameterless C functions return the
+  address (function pointer) of where to go to next:
+\begin{verbatim}
+      while (1) f = (*f)();
+\end{verbatim}
+\item implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
+\item generate assembly code, e.g., LLVM, directly.
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Modern compilers to the rescue:  sibling call}
+Clang/LLVM and GCC implement a restricted form of tail call, the \emph{sibling}
+call, i.e, jump reusing the TOS frame.  Here
+\begin{verbatim}
+void f() {
+  ...
+  g();
+}
+\end{verbatim}
+function \texttt{g} is jumped to; there is no (implicit) return following
+\texttt{g();}.  
+\\\vspace{0.1in}
+Critically, this also works for indirect calls to previously stored addresses.
+\begin{verbatim}
+    (getInfoPtr(stgCurVal.op)->entryCode)();
+\end{verbatim}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Sibling Call}
+The restrictions are due to the x86(-64)/*nix/C calling
+conventions---caller clean-up---and the possibility of pointer aliasing.
+\\\vspace{0.1in}
+Sufficient formula for recent gcc and Clang/LLVM on Linux or MacOS X
+\begin{itemize}
+\item Call in leaf position of CFG;
+\item Caller and callee have the same type signature
+\item No addresses taken of formal params;
+\item -O3.
+\end{itemize}
+Obviously this is not portable.
+\end{frame}
+
+\begin{frame}
+\frametitle{Unwinding stacks}
+In one version of GHC, Harris et al.\ identify the need to unwind thread stacks for shared
+memory parallelism.
+\\\vspace{0.1in}
+Again, we do not want to contemplate using, e.g., setjmp/longjmp.
+\\\vspace{0.1in}
+Therefore, we do not want the C main stack or Pthreads stacks to grow.
+\\\vspace{0.1in}
+Exception:  calls to the runtime.
+\\\vspace{0.1in}
+Again, sibling call saves us.
+\end{frame}
+
+\begin{frame}
+\frametitle{Here is a simple frame}
 \end{frame}
@@
 \begin{frame}
-\frametitle{In Search of Automatic Parallelization\\
-\emph{or at least automatic scheduling}}
+\frametitle{Example frame}
     \begin{columns}[t]

### 2016-06-02 codegen/CodeGen.hs
runtime gitignore

@@ -519,7 +519,7 @@ cgalts env (Alts it alts name scrt) boxed =
       let (codeypns, funcss) = unzip codefuncs
       let (codes, ypns) = unzip codeypns
       let its = [citems|
-                  LOG(LOG_INFO, $string:(name ++ " here"));
+                  LOG(LOG_INFO, $string:(name ++ " here\n"));
                   typename Cont *$id:contName = stgGetStackArgp();
                   $comment:("// make self-popping")
                   stgCaseToPopMe($id:contName);

### 2016-06-09 runtime/gc.c
added debugging functions, ft. more segfaults and now infinite loops

@@ -231,8 +231,8 @@ void gc(void) {
   //LOG(LOG_INFO, "GARBAGE COLLECTION DISABLED in gc.c/gc(void)\n"); return;
 
   //heapCheck(); // disable untill it passes tests
-  
   //successfully made about 1400 test segfault :)
+  //more segfaults + infinite loops
   //Obj **array = mallocArrayOfAllObjects();
   //addObjects(array);
   

### 2016-06-09 runtime/sanity.c

@@ -13,8 +13,12 @@
 
 //number of heap objects. a safe overestimate is (stgHP - stgHeap) / sizeof(Obj)
 #define numHeapObjs (((char *)stgHP - (char *)stgHeap) / sizeof(Obj))
+
 //number of static heap objects
 #define numStaticHeapObjs stgStatObjCount
+
+//macro for fprintf
+#define PRINTF(...) fprintf(stderr, __VA_ARGS__)
 
 /**
  *this function creates an array of pointers to all of the objects.
@@ -38,8 +42,8 @@ void addObjects (Obj **objArray) {
   //i stays at the beginning of the array, 
   //j moves to keep track of how many objects have been added
   //k is for going through the payload boxed objects
-  //flag is for forced continuation in while loop
-  int i, j, k, flag;
+  //dup_exists is for forced continuation in while loop so as to not copy duplicate objects over
+  int i, j, k, dup_exists;
   
   //mark the end of the static heap object range
   int end = stgStatObjCount;
@@ -55,49 +59,105 @@ void addObjects (Obj **objArray) {
   //now we have to check the payloads of the static heap objects
   //and add them to the part of the array which isn't comprised
   //of static heap objects
-  //TODO add payload objects to end of array
 
-  //temp -> payload[0].op
-  //end + i
-  //boxed count, dumbass
   j = 0;
+
+  PRINTF("\nstgStatObjCount: %zd\n\n", stgStatObjCount);
   while (j != end) {
    
-    flag = 0;
+    //TODO: wtf is going on
+
+    //why you're doing this:
+    //you just wanted to make sure that the first chunk of the array
+    //was the static heap objects loaded correctly
+    //that's why the last j+1 should match stgStatObjCount
+
+
+    //current dilemma: 
+    //apparently, everything you've added is a SHO
+    //and you always get two more SHO's than stgStatObjCount
+    //?
+
+    if (isSHO(objArray[j])) {
+      PRINTF("SHO #%d", j+1);
+    }
+
+    printObjInfo(objArray[j]);
+    
+    //if there are no boxed variables to copy over, move to the next
+    //object in the array and continue
+    if (objArray[j]->_infoPtr->layoutInfo.boxedCount == 0) {
+      j++;
+      continue;
+    }
+    
+    dup_exists = 0;
[…]

### 2016-06-09 runtime/sanity.h

@@ -6,6 +6,8 @@
 
 Obj **mallocArrayOfAllObjects();
 void addObjects (Obj **objArray);
+void printObjInfo (Obj *obj);
+char *objTypeToString(Obj *obj);
 void sanityCheckSingleSHO (Obj *obj);
 void sanityCheckObj (Obj *obj);
 bool checkPtr8BitAligned (Obj *obj);

### 2016-06-10 pubs/tfp16/slides/slides.tex
figures
mention mini-Haskell
make printer-friendly build target for slides
tweak slide
update slides. More gradual exposition

@@
 %\usetheme[unclass]{LANL-beta}
-\usetheme{LANL-beta}
+
+% Use make print to create less inky slides for printing
+\ifdefined\printableSlides
+\usetheme[lessInk]{LANL-beta}
+\else
+\usetheme[blue]{LANL-beta}
+\fi
+
 \usepackage{graphicx}
 
-\title{Automatic Parallelization and Transparent Fault Tolerance (project in progress)}
+\title{Automatic Parallelization and Transparent Fault Tolerance}
 %\subtitle{(Project article)}
@@
 \frametitle{Scientific Computing in our Microcosm}
-Local evolution of scientific computing
+\only<1-9>{\textbf{Local evolution of scientific computing}}
 \begin{itemize}
-  \item Serial Fortran programs
-  \item MPI everywhere (inter- and intra-node)
-  \item C, C++
-  \item MPI+X, X is Pthreads, OpenMP, OpenCL, CUDA, etc.
-  \item Parallel runtimes, e.g., Cilk++, Intel Threading Building Blocks, Stanford's Legion, etc.
+  \item<2-9> Serial Fortran programs
+    \\\only<2>{\ \ Simple, imperative.}
+  \item<3-9> C
+    \\\only<3>{\ \ Similar. More pleasant syntax (Free form source code!)}
+  \item<4-9> Vendor communication libraries
+    \\\only<4>{\ \  Non-portable, but parallel}
+  \item<5-9> MPI everywhere (inter- and intra-node)
+    \\\only<5>{\ \ De facto standard emerges to solve portability}
+  \item<6-9> C++
+    \\\only<6>{\ \ Object oriented goodness (and more!)}
+  \item<7-9> MPI+\{Pthreads, OpenMP, OpenCL, CUDA, etc\}
+    \\\only<7>{\ \ We augment MPI now with other explicit parallel models}
+  \item<8-9> Parallel runtimes, e.g., Cilk Plus, Intel Threading Building Blocks,
+    Stanford's Legion, etc.
+    \\\only<8>{\ \ Automated scheduling for irregularly shaped problems. Task-based}
 \end{itemize}
-Currently MPI+threading model dominates.
+\only<9>{%
+\textbf{
+Currently MPI+threading model dominates.}}
+
 \end{frame}
@@
     \begin{center}
-%      \fbox{
       \includegraphics[width=0.95\linewidth,height=1in]%
         {figures/S3Dtaskgraph.png}~\footnote{Courtesy Stanford Legion project}
-%}
 \\[2.0ex]  % vertical space
@@
 Schedule by hand?
+\end{frame}
+
+\begin{frame}
+  \frametitle{Transparent Fault Tolerance?}
+
+  \onslide<1->{%
+    Processor clock speeds stopped growing a decade ago
+    \\\vspace{0.1in}}
+
+  \onslide<2->{%
+    Solution? More processors.
+    \\\vspace{0.1in}}
+
+  \onslide<3->{%
+    But, with more processors comes smaller Mean Time to Failure.
+    \\\vspace{0.1in}}
+
+  \onslide<4->{%
+    Obviously, we can't just hope to get lucky on a months-long calculation}
 \end{frame}
@@
     \end{column}
+    \onslide<2->{%
     \begin{column}[T]{.48\linewidth}
@@
       \end{enumerate}
-    \end{column}
+    \end{column}}
   \end{columns}
 \vspace{0.1in}
-General claim:  ``data movement costs more energy than computation.''
-
-Fact 1: In C/R, idle processors may cost more energy than data movement.
-
-Fact 2:  C/R isn't scaling.
+\onslide<3>{%
+Inherent problems:
+\begin{itemize}
+\item Idle processors during relatively slow writes burn tight energy budgets.
+\item MTTF is shrinking and C/R isn't scaling.
+\end{itemize}}
 \end{frame}
@@
 
+\onslide<2->{%
+  They understand:\\
   \emph{In a multi-100,000-line program, do not temporarily alter the global speed-of-light
     `constant' variable.}
-\vspace{0.1in}
+  \vspace{0.1in}}
 
+\onslide<3->{%
   \emph{But}, these are not computer scientists:
   \begin{itemize}
-\item Passing functions as arguments is familiar (since early Fortran);
-\item Implicit space leaks are confusing;
-\item Non-strict evaluation is largely irrelevant, strictness annotations are unintuitive.
+  \item Passing functions as arguments is familiar (since early Fortran).
+  \item Implicit space leaks are confusing.
+  \item Non-strict evaluation is largely irrelevant
   \item Unboxed primitives and arrays of the same (unboxed vector) are the primary data.
-\end{itemize}
+  \end{itemize}}
 \end{frame}
@@
 \frametitle{Claimed Trends}
-Trends in functional programming
+\textbf{Trends in functional programming}
 \begin{itemize}
-  \item Appreciation of the virtues of strict-by-default semantics;
-  \item Transparent fault tolerance.
+  \item<1-> Appreciation of the virtues of strict-by-default semantics.
+    \\\only<1>{\ \ GHC Strict and StrictData for HPC}
+  \item<2-> Transparent fault tolerance.
+    \\\only<2>{\ \ Stewart's HdpH-RS}
 \end{itemize}
-Trends in scientific computing
+\onslide<3->{\textbf{Trends in scientific computing}}
 \begin{itemize}
-  \item Implementation of pure functional computational concepts/components.
-  \item Recognition that checkpoint/restart is increasingly intractable and unscalable.
+  \item<4-> Implementation of pure functional computational concepts/components.    
+  \item<4-> Recognition that checkpoint/restart is increasingly intractable and unscalable.
+    \\\only<4>{%
+      \ \ Legion is ``pure with respect to $X$''\\
+      \ \ Also adding task-restarting for fault-tolerance}
 \end{itemize}
@@
 (Semantic) function strictness $\implies$ \\
-\ \ \ function arguments can be evaluated early, in parallel
-
+\ \ \ function arguments can be evaluated early, in parallel\\
+\ \ \ How can we achieve this?
 \begin{itemize}
-  \item Strictness analysis
-  \item Bang patterns
-  \item Par/pseq, other specifications
-  \item Speculative evaluation
-  \item \ldots
-  \item \emph{Strict(er) default semantics}
+  \item<2-> Strictness analysis
+    \\\only<2>{\ \ Can be inconclusive.}
+  \item<3-> Bang patterns in bindings
+    \\\only<3>{\ \ Prone to error. Hard to read.}
+  \item<4-> Par/pseq, other specifications on expressions
+    \\\only<4>{\ \ Ditto.}
+  \item<5-> Speculative evaluation
+    \\\only<5>{\ \ Difficult to get right without annotation.}
+  \item<6-> \ldots
+  \item<7-> \emph{Strict(er) default semantics}
 \end{itemize}
+\onslide<8->{%
 Similarly constructors.
@@
 Update:  Strict \texttt{let} bindings.
+}
 \end{frame}
@@
   \begin{itemize}
-  \item Could we keep a large number of hardware threads (e.g., up to 256) busy with
+  \item<1-> Could we keep a large number of hardware threads (e.g., up to 256) busy with
     \emph{implicit} parallelism?
@@
     \end{itemize}
-  \item How often do new functional programmers/old scientific programmers
+  \item<2-> How often do new functional programmers/old scientific programmers
     make essential use of non-strictness?  \emph{(very rarely)}
@@
 \begin{enumerate}
-\item Haskell to STG (or Core) via GHC \emph{(todo)};
-\item STG to C \emph{(serial done, parallel in progress)};
-\item Fault tolerance \emph{TBD}.
+\item<2-> Haskell to STG (or Core) via GHC \emph{(todo)};
+  \\\only<2>{%
+    \ \ STG and Core are higher-order polymorphic functional languages in their own right.}
+\item<3-> STG to C \emph{(serial done, parallel in progress)};
+  \\\only<3>{%
+    \ \ Fairly faithful to STG papers}
+\item<4-> Mini-Haskell front end (done).
+  \\\only<4>{%
+    \ \ Close to ML, syntatically}
+\item<5-> Fault tolerance \emph{TBD};
 \end{enumerate}
@@
 \frametitle{Reinventing the wheel?}  Why not use Manticore, MultiMLton, F\#,
-multicore OCaml, etc., or even GHC in toto?
+multicore OCaml, SAC etc., \onslide<3->{or even GHC in toto?}
 \begin{itemize}
-  \item Very light weight:  easy to instrument, modify;
-  \item Know exactly what is going on under the hood;
-  \item There exist conduits for getting student interns up to speed inexpensively;
-  \item We have a small cadre of scientists already interested in Haskell/GHC.
+  \item<2-> We have a small cadre of scientists already interested in Haskell/GHC.
+  \item<3-> Very light weight:  easy to instrument, modify;
+  \item<4-> Know exactly what is going on under the hood;
+  \item<5-> There exist conduits for getting student interns up to speed inexpensively;
 \end{itemize}
@@
 \begin{itemize}
-\item setjmp/longjmp;
-\item giant switch;
-\item computed goto (gcc extension);
-\item trampoline, the simplest of which parameterless C functions return the
+\item<1-> setjmp/longjmp;
+\only<1>{Significant overhead and bookkeeping}
+\item<2-> giant switch;
+\only<2>{Ugly hack, not really linkable}
+\item<3-> computed goto (gcc extension);
+\only<3>{Limited to gcc}
+\item<4-> trampoline, the simplest of which parameterless C functions return the
   address (function pointer) of where to go to next:
@@
 \end{verbatim}
-\item implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
-\item generate assembly code, e.g., LLVM, directly.
+\only<4>{Our intial implementation used this approach}
+\item<5-> implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
+\only<5>{Fun, but out of scope}
+\item<6-> generate assembly code, e.g., LLVM, directly.
+\only<6>{Ambitious. Haskell LLVM bindings are unreliable}
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Here is a simple frame}
-\end{frame}
+  \frametitle{Our Future}
 
-\begin{frame}
-\frametitle{Here is a simple frame}
   \begin{itemize}
-        \item Here is the first bullet in a standard bulleted list,
-            as is customary for a presentation like this
-        \begin{itemize}
-            \item It includes some sub-bullets
-            \item Here they are
+  \item<1-> GHC tie-in and parallel implementation (hopefully) this summer
+  \item<2-> Experimentation with non-trivial code
+  \item<3-> Fault tolerance sometime thereafter
   \end{itemize}
-        \item Here is another bullet
-        \item And here is another, with some math:
-            \[ e^{i\theta} = \cos \theta + i \sin \theta \]
-    \end{itemize}
-\end{frame}
 
-\begin{frame}
-\frametitle{Example frame}
-    \begin{columns}[t]
-        \begin{column}{.48\linewidth}
-            \begin{itemize}
-                \item Here is a bulleted list that sits alongside a graphic
-                \item With a second bullet item
-                \item And a third
-                \begin{itemize}
-                   \item It can have sub-bullets too
-                   \item Like this
-                \end{itemize}
-            \end{itemize}
-        \end{column}
-        \begin{column}[T]{.48\linewidth}
-            \begin{figure}
-                \begin{center}
-                    \fbox{\includegraphics[width=0.9\linewidth]%
-                        {LANL-logo-gray}}\\[2.0ex]
-                    insert caption here
-                \end{center}
-            \end{figure}%
-        \end{column}
-    \end{columns}
+  \onslide<4->\textbf{Questions?}
 \end{frame}

### 2016-06-11 pubs/tfp16/slides/slides.tex
touchup slides

@@
     Stanford's Legion, etc.
-    \\\only<8>{\ \ Automated scheduling for irregularly shaped problems. Task-based}
+    \\\only<8>{\ \ Automated scheduling for irregularly shaped problems.}
 \end{itemize}
@@
   \end{figure}
+  \onslide<2->{%
     Simplest interesting chemistry, task graph much larger with more complex reactants.
     \vspace{0.1in}
-Schedule by hand?
+    Schedule by hand?}
+  \\\onslide<3->{%
+    \textbf{This comes for free in the pure functional world}}
 \end{frame}
@@
 \vspace{0.1in}
-\onslide<3>{%
+\onslide<3->{%
 Inherent problems:
 \begin{itemize}
-\item Idle processors during relatively slow writes burn tight energy budgets.
+\item Idle processors burn tight energy budgets.
 \item MTTF is shrinking and C/R isn't scaling.
 \end{itemize}}
+\onslide<4->{%
+  \textbf{With referential transparency, this is a much smaller problem}}
 \end{frame}
@@
   \item Implicit space leaks are confusing.
-  \item Non-strict evaluation is largely irrelevant
+  \item Non-strict evaluation is largely irrelevant.
   \item Unboxed primitives and arrays of the same (unboxed vector) are the primary data.
@@
     \ \ Fairly faithful to STG papers}
-\item<4-> Mini-Haskell front end (done).
+\item<4-> Mini-Haskell front end (done);
   \\\only<4>{%
     \ \ Close to ML, syntatically}
-\item<5-> Fault tolerance \emph{TBD};
+\item<5-> Fault tolerance \emph{TBD}.
 \end{enumerate}
@@
 \begin{itemize}
-  \item<2-> We have a small cadre of scientists already interested in Haskell/GHC.
+  \item<2-> We have a small cadre of scientists already interested in Haskell/GHC;
   \item<3-> Very light weight:  easy to instrument, modify;
   \item<4-> Know exactly what is going on under the hood;
-  \item<5-> There exist conduits for getting student interns up to speed inexpensively;
+  \item<5-> There exist conduits for getting student interns up to speed inexpensively.
 \end{itemize}
@@
 \item<1-> setjmp/longjmp;
-\only<1>{Significant overhead and bookkeeping}
+\\\only<1>{\ \ Significant overhead and bookkeeping}
 \item<2-> giant switch;
-\only<2>{Ugly hack, not really linkable}
+\\\only<2>{\ \ Ugly hack, not really linkable}
 \item<3-> computed goto (gcc extension);
-\only<3>{Limited to gcc}
+\\\only<3>{\ \ Limited to gcc}
 \item<4-> trampoline, the simplest of which parameterless C functions return the
@@
 \end{verbatim}
-\only<4>{Our intial implementation used this approach}
+\only<4>{\ \ Our intial implementation used this approach}
 \item<5-> implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
-\only<5>{Fun, but out of scope}
+\\\only<5>{\ \ Fun, but out of scope}
 \item<6-> generate assembly code, e.g., LLVM, directly.
-\only<6>{Ambitious. Haskell LLVM bindings are unreliable}
+\\\only<6>{\ \ Ambitious. Haskell LLVM bindings are unreliable}
 \end{itemize}
@@
 
-  \onslide<4->\textbf{Questions?}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Thank you}
+  \textbf{Questions?}
 \end{frame}

### 2016-06-15 codegen/FromGHC.hs
Initial commit, setup for Haskell stack

@@ -1,0 +1,12 @@
+
+module FromGHC
+  ()
+where
+
+
+import           ADT
+import           AST
+import qualified GHC
+
+
+fun = GHC.mkSrcLoc        

### 2016-06-15 codegen/Driver.hs
add -Wall -Werror to stgc build wrapper and fix various warnings in generated c code

@@ -258,11 +258,11 @@ codegener inp v mhs = let (tycons, objs) = heapchecker mhs inp
                           (shoForward, shoDef) = showSHOs objs
                           (funForwards, funDefs) = cgObjs objs stgRTSGlobals
                           (stgStatObjCount, stgStatObj) = shos objs
-                          defs =  header : funForwards ++ 
-                                  typeEnums ++ 
+                          defs =  header : funForwards ++
+                                  typeEnums ++
                                   infotab ++
-                                  shoForward ++ 
-                                  shoDef ++ 
+                                  shoForward ++
+                                  shoDef ++
                                   [ stgStatObjCount, stgStatObj ] ++
                                   concat funDefs ++
                                   footer v

### 2016-06-15 pubs/tfp16/slides/slides.tex
add LAUR #, acknowledgements to slides

@@
 \date{Trends in Functional Programming, June 8-10, 2016}
-\LAUR{LA-UR-16-?????}
+\LAUR{LA-UR-16-24056}
 
@@
 \begin{frame}
+  \frametitle{Acknowledgements}
+
+  \footnotesize{Funding for this project has been provided by the DOE
+    NNSA LANL Laboratory Directed Research and Development program
+    award 20150845ER; the National Science Foundation Science,
+    Technology, Engineering, and Mathematics Talent Expansion Program
+    (NSF STEP) program for undergraduate students; and the Department
+    of Energy Science Undergraduate Laboratory Internship (DOE SULI)
+    program.}
+
+  \vspace{0.2in}
+
+  \footnotesize{Los Alamos National Laboratory is managed and operated
+    by Los Alamos National Security, LLC (LANS), under contract number
+    DE-AC52-06NA25396 for the Department of Energy’s National Nuclear
+    Security Administration (NNSA).}
+\end{frame}
+
+\begin{frame}
   \frametitle{Thank you}

### 2016-06-15 runtime/sanity.c
added objects to array with very basic sanity checking, more must be done

@@ -40,9 +40,6 @@ Obj **mallocArrayOfAllObjects() {
  */
 void addObjects (Obj *objArray[]) {
 
-  //i stays at the beginning of the array, 
-  //j moves to keep track of how many objects have been added
-  //k is for going through the payload boxed objects
 
   //int i, j = 0, k;
   //mark the end of the static heap object range
@@ -54,7 +51,7 @@ void addObjects (Obj *objArray[]) {
     // object itself is valid
     sanityCheckSingleSHO(stgStatObj[i]);
     objArray[i] = stgStatObj[i];
-    printObjInfo(objArray[i]);
+    //printObjInfo(objArray[i]);
   }
 
   //could be j = i
@@ -70,59 +67,42 @@ void addObjects (Obj *objArray[]) {
     }
     if (k == j) {
       objArray[j] = stgCurVal.op;
+      j++;
     }
   }
+
   //reset i to mark beginning of array
-  /*i = 0;
+  i = 0;
   int p;
   while (i != j) {
-    for (p = 0; p != objArray[i]->_infoPtr->layoutInfo.boxedCount; p++) {
-      if (!isInObjArray(objArray, objArray[i]->payload[p].op, i, j)) {
+ 
+  //these change based on whether the object is a thunk / blackhole or not
+  int start = 0;
+  int end = getInfoPtr(objArray[i])->layoutInfo.boxedCount;
+   
+    switch (getObjType(objArray[i])) {
+      
+      case THUNK:
+      case BLACKHOLE:
+        start = startTHUNKFVsB(objArray[i]);
+        end = endTHUNKFVsB(objArray[i]);
+        break; 
+      default:
+        break;
+    }
+
+    for (p = start; p != end; p++) {
+      //otherwise if the payload isn't in the array, add it
+      if (!isInObjArray(objArray, objArray[i]->payload[p].op, 0, j-1)) {
+          //make sure pointer is aligned, not null, and in the heap before you
+          //add it 
+          sanityCheckPtr(objArray[i]->payload[p].op);
           objArray[j] = objArray[i]->payload[p].op;
           j++;
       }
     }
     i++;
-  }*/
-
-  //outer (i) loop traverses the objarray by object
-  /*for (i = 0; i < end; i++) {    
-    //if the object doesn't have boxed variables in it's payload
-    //go to the next one
-    if (objArray[i]->_infoPtr->layoutInfo.boxedCount == 0) {
[…]

### 2016-06-15 runtime/sanity.h

@@ -8,6 +8,7 @@ Obj **mallocArrayOfAllObjects();
 void addObjects (Obj **objArray);
 void printObjInfo (Obj *obj);
 char *objTypeToString(Obj *obj);
+void sanityCheckPtr(Obj *obj);
 bool isInObjArray (Obj *objArray[], Obj *obj, int start, int end);
 void sanityCheckSingleSHO (Obj *obj);
 void sanityCheckObj (Obj *obj);

### 2016-07-12 runtime/gc.c
some more sanity checker stuff

@@ -13,7 +13,7 @@
 
 static void *scanPtr, *freePtr;
 
-__attribute__((always_inline)) void setHeapArgType(PtrOrLiteral *f) {
+__attribute__((always_inline)) inline void setHeapArgType(PtrOrLiteral *f) {
   #if USE_ARGTYPE
         f->argType = HEAPOBJ;
   #endif
@@ -99,7 +99,7 @@ PtrOrLiteral updatePtrByValue (PtrOrLiteral f) {
 }
 
 
-__attribute__((always_inline)) void updatePtr(PtrOrLiteral *f) {
+__attribute__((always_inline)) inline void updatePtr(PtrOrLiteral *f) {
   *f = updatePtrByValue(*f);
 }
 

### 2016-07-12 runtime/heap.c

@@ -1,6 +1,6 @@
 #include "stg.h"
 #include "heap.h"
-#include "gc.c"
+#include "gc.h"
 #include <string.h>  // for memcpy()
 
 // this is a temporary hack as we incorporate Bitmap64s into continuations

### 2016-07-12 runtime/sanity.c

@@ -12,7 +12,7 @@
 #include "show.h"
 #include "sanity.h"
 
-__attribute__((always_inline)) char *getIdent(Obj *obj) {
+__attribute__((always_inline)) inline char *getIdent(Obj *obj) {
   #if USE_IDENT
     return obj->ident;
   #else
@@ -20,7 +20,7 @@ __attribute__((always_inline)) char *getIdent(Obj *obj) {
   #endif
 }
 
-__attribute__((always_inline)) bool checkObjType(Obj *obj, ObjType objType) {
+__attribute__((always_inline)) inline bool checkObjType(Obj *obj, ObjType objType) {
   #if USE_OBJTYPE
     return (obj->objType == objType);
   #else
@@ -28,7 +28,7 @@ __attribute__((always_inline)) bool checkObjType(Obj *obj, ObjType objType) {
   #endif
 }
 
-__attribute__((always_inline)) bool checkInfoTabName(InfoTab *it, char *name) {
+__attribute__((always_inline)) inline bool checkInfoTabName(InfoTab *it, char *name) {
   #if USE_INFOTAB_NAME
     return (strcmp(it->name, "stgIndirect") == 0);
   #else
@@ -36,7 +36,7 @@ __attribute__((always_inline)) bool checkInfoTabName(InfoTab *it, char *name) {
   #endif
 }
 
-__attribute__((always_inline)) bool checkIdent(Obj *obj) {
+__attribute__((always_inline)) inline bool checkIdent(Obj *obj) {
   #if USE_IDENT
     for (int i = 0; obj->ident[i] != '\0'; i++) {
       assert(i < IDENT_SIZE && "sanity: bad ident size");
@@ -46,7 +46,7 @@ __attribute__((always_inline)) bool checkIdent(Obj *obj) {
   return true;
 }
 
-__attribute__((always_inline)) bool checkInfoTabHeader(InfoTab *it) {
+__attribute__((always_inline)) inline bool checkInfoTabHeader(InfoTab *it) {
   #if DEBUG_INFOTAB
     return (it->pi = PI());
   #else
@@ -169,7 +169,7 @@ int addObjects(Obj *objArray[]) {
 
   int i = 0;
   while (i < objCount) {
-    LOG(LOG_SPEW, "i %d\n",i);
+    //LOG(LOG_SPEW, "i %d\n",i);
     size_t start, end;
 
     assert(getObjType(objArray[i]) != INDIRECT && "sanity: indirect in objArray");
[…]

### 2016-07-12 runtime/show.c

@@ -281,153 +281,6 @@ void showStgValDebug(LogLevel priority, PtrOrLiteral v) {
 
 void showObjSpaceInfo();
 
-void checkStgObjRec(Obj *p) {
-  if (p == NULL) return;
-
-  size_t i;
-
-  assert(depth + 1 < showDepthLimit && "hc: checkStgObjRec depth exceeded\n");
-  assert((uintptr_t)p % OBJ_ALIGN == 0 && "hc: bad Obj alignment");
-  // following causes test symbolT10 to fail!!!
-  if (!(isHeap(p) || isSHO(p))) {
-    showObjSpaceInfo();
-    LOG(LOG_ERROR, "hc:  bad Obj location, object addr is %p, recursive depth %d\n", p, depth);
-    showStgObjRecPretty(LOG_ERROR, p);
-    assert(false);
-  }
-
-  InfoTab *itp = getInfoPtr(p);
-  InfoTab it = *itp;
-
-  for (i = 0; i != depth; i++) if (p == stack[i]) return;
-  stack[depth++] = p;
-
-  ObjType type = getObjType(p);
-  assert(type > PHONYSTARTOBJ && type < PHONYENDOBJ && "hc: bad obj type");
-
-  if (strcmp(it.name, p->ident)) {
-    if (type != PAP) {
-      LOG(LOG_ERROR, "mismatch in infotab and object names \"%s\" != \"%s\"\n",
-          it.name, p->ident);
-      assert(false);
-    }
-  }
-
-  switch (type) {
-  case FUN: {
-    assert(it.objType == FUN && "hc: FUN infotab type mismatch");
-    int FVCount = endFUNFVsU(p);
-    if (FVCount) {
-      // check that unboxed FVs really are unboxed
-      for (i = startFUNFVsU(p); i < FVCount; i++) {
-        assert(mayBeUnboxed(p->payload[i]) && "hc: unexpected boxed FV in FUN");
-      }
-      // check that boxed FVs really are boxed
-      for (i = startFUNFVsB(p); i < endFUNFVsB(p); i++) {
-        assert(mayBeBoxed(p->payload[i]) && "hc: unexpected unboxed FV in FUN");
-        checkStgObjRec(p->payload[i].op);
-      }
-    }
-    break;
-  }
-
-  case PAP: {
-    assert(it.objType == FUN && "hc: PAP infotab type mismatch");
-    int FVCount = endPAPFVsU(p);
-    if (FVCount) {
[…]

### 2016-07-12 pubs/tfp16-draft/paper.tex
rename dir

--- 2016-04-25 pubs/tfp16/paper.tex
@@
 
-\renewcommand{\floatpagefraction}{0.8}
+\renewcommand{\floatpagefraction}{0.99}
 %
@@
 %
-\title{Automatic Parallelization and Transparent Fault Tolerance\\
-       (\emph{Project article})}
+\title{Automatic Parallelization and Transparent Fault Tolerance}
 %
@@
 these sources.
-%
-\\
-{\color{OliveGreen} % \bf
-\emph{Draft paper notes:}
-\begin{enumerate}[before=\itshape]
-\item We are reporting work in progress.  At the time of writing we have a
-  serial implementation, a primitive proof-of-concept parallel
-  implementation, and are currently developing a more realistic parallel
-  runtime.  
-%We expect to have preliminary results from our new parallel
-%implementation by the time of the symposium.  
-The fault-tolerance mechanisms remain future work.
-
-\item Hopefully in the spirit of a student-friendly symposium, this draft is
-  liberally annotated with indicators of student contributions, student
-  preparation, the influence of their involvement on our approach, and other
-  incidental and ancillary comments.  A final paper would of course be cleaned
-  up and potentially considerably shortened.
-
-\item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
-  (all undergraduate) have contributed considerably to this effort, and to the
-  writing of this paper, and will continue to do so.  If this paper is
-  accepted we plan to have one of these students present it at TiFP'16.
-\end{enumerate}
-}
 
@@
 A minor, informal goal is to gain a sense of how often relatively
-inexperienced functional programmers make essential use of
-non-strictness and laziness.
-%\footnote{The starting experience of our student interns to
-%  date has been normalized by having them work through the entirety of Hutton's \emph{Programming
-%    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
-%  substituting his article \emph{Higher-order Functions for
-%    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
-%  book.}
+inexperienced functional programmers make essential use of non-strictness.
 
@@
 STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
-library~\cite{ghc-as-library}.  Initially we start with our own STG.
-%, with bridging the gap with GHC scheduled to commence quite soon.
+library~\cite{ghc-as-library}.  Initially we started with our own STG.
 
 \subsection{STG Language}
+\renewcommand{\floatpagefraction}{0.99}
+
+Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
+evoke both an abstract and concrete syntax.  This formulation and presentation
+is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
+paper~\cite{Marlow:2004}, to which we will subsequently refer as \emph{EA}\@.
+A \emph{BLACKHOLE} object, signifying a \emph{THUNK} being evaluated, is only
+a run-time entity and not part of the language.  Omitted is the syntax for
+algebraic data type definitions which is given in
+Appendix~\ref{app:minihaskell}.
 
@@
 
-Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
-evoke both an abstract and concrete syntax.  This formulation and presentation
-is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
-paper~\cite{Marlow:2004}, to which we will subsequently refer as \emph{EA}\@.
-A \emph{BLACKHOLE} object, signifying a \emph{THUNK} being evaluated, is only
-a run-time entity and not part of the language.  Omitted is the syntax for
-algebraic data type definitions which is given in
-Appendix~\ref{app:minihaskell}.
-
 With a few exceptions the semantics of STG is what a Haskell programmer
@@
 preserved so we will not use the term lazy and simply refer to
-strictness properties.  As is common we will use strict in an operational
+strictness properties.\footnote{%
+  We can readily simulate the absence of sharing (i.e, call by name instead of
+  call by need), in the serial case at least, simply by not updating \emph{THUNK}s
+after evaluation, resulting in, for example, exponential running time for
+simple mergesort on lists.}
+  As is common we will use strict in an operational
 sense, meaning that argument(s) are evaluated before being passed in a
@@
 
-\section{Current Status (April 8, 2016), Future Work}
+\section{Current Status, Future Work}
 
@@
 maintain sharp focus on the main goals: implementing and evaluating mechanisms
-for automatic parallelism and fault tolerance.  At the time of writing we have
+for automatic parallelism and fault tolerance.  Currently we have
 a fully functioning serial implementation including
 \begin{itemize}
-\item STG parser \emph{(by student)},
-\item mini-Haskell parser, mini-Haskell to STG transformer \emph{(by student)},
+\item STG parser,
+\item mini-Haskell parser, mini-Haskell to STG transformer,
 \item type inferencer,
-\item various analysis and transformation passes \emph{(some by students)},
-\item code generator \emph{(some student contribution)},
+\item various analysis and transformation passes,
+\item code generator,
 \item runtime system including a garbage collector,
 \item continuous integration and testing system,
-\item extensive test suite \emph{(mostly by students)}.\footnote{Many are
-    students' solutions to exercises in learning algorithms and data
-    structures.}
+\item extensive test suite, automated testing.
 \end{itemize}
@@
 %
+Bridging the gap between GHC STG and our STG, and the corresponding
+augmentation of the runtime is showing early signs of success.
+%
+An instrumentation infrastructure, integrated with the testing framework to
+record and present time and space run-time statistics such as number of heap
+and stack allocations and maximum stack depth, is planned for the near future.
+%
 The fault-tolerance mechanisms remain future work.
@@
 Parallel garbage collection is likely out of scope.
-
-This summer (2016) a returning student will bridge the gap between GHC STG (or
-Core) and our STG, augmenting the runtime as needed.  Another student will
-develop an instrumentation infrastructure integrated with the testing
-framework to record and present time and space run-time statistics such as
-number of heap and stack allocations and maximum stack depth.
-
+%
 %We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
@@
 % because # looks strangely large next to math terms
+
 \newcommand{\hash}{{\scriptsize\#}}
-
-\begin{figure}
+\begin{figure}[h!tb]
 \centering
@@
 
-\section{Student Preparation}
-
-Student interns' backgrounds in computer programming has ranged from minimal
-to some exposure to one or more mainstream imperative languages, Lisp, and
-Haskell.  To bring them to a normalized level of proficiency their first
-assignment has been to work through the entirety of Hutton's \emph{Programming
-  in Haskell} textbook~\cite{Hutton-book}, substituting one of his articles
-\emph{Higher-order Functions for Parsing}~\cite{Hutton-parsing:1992} or
-\emph{Monadic Parser Combinators}~\cite{Hutton:1996} for the corresponding
-material in the book.  This may be followed by writing a parser and ADT for a
-progressively enriched lambda calculus, together with call-by-value,
-call-by-name, and (in one case) call-by-need interpreters.  The concept of
-compilation can be illustrated with an exercise in parsing infix arithmetic
-expressions to an AST, and transforming that to a sequence of operations for a
-simple stack machine.  From this point they have then been able to substantively
-contribute to this project.
-
-% breaks page limit
-%  This learning phase can take the
-%better part of a summer or one-semester internship, so we particularly
-%seek return interns.
-
-%
-% ---- Bibliography ----
-%
-
 \bibliographystyle{splncs03}

### 2016-07-12 runtime/gc.c
more runtime cleanup/sanity checking stuff

@@ -207,8 +207,8 @@ void gc(void) {
 
   if (sanityChecker) {
     LOG(LOG_SPEW, "before GC\n");
-    heapCheck(true);
-    stackCheck(true);
+    heapCheck(true, LOG_DEBUG);
+    stackCheck(true, LOG_DEBUG);
   }
 
   // check if GC should run at all
@@ -246,8 +246,8 @@ void gc(void) {
 
   if (sanityChecker) {
     LOG(LOG_SPEW, "after GC\n");
-    heapCheck(true);
-    stackCheck(true);
+    heapCheck(true, LOG_DEBUG);
+    stackCheck(true, LOG_DEBUG);
   }
 
   if(LOG_LEVEL == LOG_SPEW) {

### 2016-07-12 runtime/sanity.c

@@ -323,27 +323,27 @@ void checkObjFull(Obj *obj) {
 }
 
 
-void heapCheck(bool display) {
+void heapCheck(bool display, LogLevel logLevel) {
   Obj **objArray = mallocArrayOfAllObjects();
   int objCount = addObjects(objArray);
-  if (display) LOG(LOG_INFO, "stg HEAP:\n------------\n");
+  if (display) LOG(logLevel, "stg HEAP:\n------------\n");
   for (int i = 0; i < objCount; i++) {
     checkObjFull(objArray[i]);
-    if (display) showStgObj(LOG_INFO, objArray[i]);
+    if (display) showStgObj(logLevel, objArray[i]);
   }
-  if (display) LOG(LOG_INFO, "\n");
+  if (display) LOG(logLevel, "\n");
 }
 
 
-void stackCheck(bool display) {
-  if (display) LOG(LOG_INFO, "stg STACK:\n-----------\n");
+void stackCheck(bool display, LogLevel logLevel) {
+  if (display) LOG(logLevel, "stg STACK:\n-----------\n");
   for (Cont *p = (Cont *)stgSP;
        (char *)p < (char*) stgStack + stgStackSize;
        p = (Cont *)((char*)p + getContSize(p))) {
     int contType = getContType(p);
     assert(contType > PHONYSTARTCONT &&
      contType < PHONYENDCONT && "sanity: bad cont type");
-    LOG(LOG_SPEW, "check Cont %s %s\n", contTypeNames[contType], p->ident);
+    LOG(logLevel, "check Cont %s %s\n", contTypeNames[contType], p->ident);
     Bitmap64 bm = p->layout;
     uint64_t mask = bm.bitmap.mask;
     int size = bm.bitmap.size;
@@ -362,7 +362,7 @@ void stackCheck(bool display) {
         }
       }
     }
-    if (display) showStgCont(LOG_LEVEL, (Cont *)p);
+    if (display) showStgCont(logLevel, (Cont *)p);
   }
-  if (display) LOG(LOG_INFO, "\n");
+  if (display) LOG(logLevel, "\n");
 }

### 2016-07-12 runtime/sanity.h

@@ -1,9 +1,10 @@
 #ifndef SANITY_H
 #define SANITY_H
 
+#include "log.h"
 #include <ctype.h>
 
-void heapCheck(bool display);
-void stackCheck(bool display);
+void heapCheck(bool display, LogLevel logLevel);
+void stackCheck(bool display, LogLevel logLevel);
 
 #endif

### 2016-07-12 runtime/show.c

@@ -13,22 +13,15 @@
 
 void showStgObj(LogLevel priority, Obj *p) {
   showStgObjPretty(priority, p);
-  // showStgObjDebug(Obj *p)
 }
+
 void showStgVal(LogLevel priority, PtrOrLiteral v) {
   showStgValPretty(priority, v);
-  // showStgValDebug(v);
 }
 
 static const int showDepthLimit = 1000;
 static int depth;
 static Obj *stack[1000];
-
-void showStgObjRecDebug(LogLevel priority, Obj *p);
-void showStgObjDebug(LogLevel priority, Obj *p) {
-  depth = 0;
-  showStgObjRecDebug(priority, p);
-}
 
 void showStgObjRecPretty(LogLevel priority, Obj *p);
 void showStgObjPretty(LogLevel priority,Obj *p) {
@@ -198,87 +191,6 @@ void showStgValPretty(LogLevel priority, PtrOrLiteral v) {
 #endif
 }
 
-
-void showStgObjRecDebug(LogLevel priority, Obj *p) {
-
-  // depth check first
-  if (depth+1 >= showDepthLimit) {
-    LOG(LOG_ERROR, "******showStgObjRec depth exceeded\n");
-    return;
-  }
-  for (int i=0; i != depth; i++) {
-    if (p == stack[i]) {
-      LOG(LOG_ERROR, "   ***cycle\n");
-      return;
-    }
-  }
-  stack[depth++] = p;
-
-
-
-  InfoTab it = *(getInfoPtr(p));
-  ObjType type = getObjType(p);
-  LOG(priority, "%s %s %s ", objTypeNames[type],
-	  objTypeNames[it.objType], it.name);
-  switch (type) {
-  case FUN:
-    LOG(priority, "\n");
-    break;
-
-  case PAP:
-    LOG(priority, "\n");
-    break;
-
-  case CON:
-    LOG(priority, "tag %d arity %d\n", it.conFields.tag, it.conFields.arity );
-    for (int i = 0; i != it.conFields.arity; i++)
-      showStgValDebug(priority, p->payload[i]);
-    break;
-
-  case THUNK:
-    LOG(priority, "\n");
-    break;
-
-  case BLACKHOLE:
-    LOG(priority, "\n");
-    break;
-
-  case INDIRECT:
-    LOG(priority, "INDIRECT to\n");
[…]

### 2016-07-12 pubs/tfp16-draft/paper.tex
Add GHC-Bridge section to TFP paper

@@
 
+\section{(Ab)Using GHC}
+
+The bridge to full Haskell is in progress, with GHC providing much of
+the structure.  We are using GHC as a library (as opposed to modifying
+its source to use our code generation and runtime system).  This
+choice was made primarily due to the size and complexity of a project
+as large as GHC.  Fully understanding and modifying even its build
+system would be a significant undertaking.
+
+We are using GHC's pipeline to translate Haskell source to GHC's
+representation of the STG language, which we then translate to our
+own.  Up to the STG generation stage, this usage is equivalent to
+GHC's \texttt{--make} behavior, which allows us to benefit from the
+module dependency analysis, typechecking, Core-level transformations
+and informative error messages for invalid programs.
+
+Though we follow the same rough idea as GHC in implementing the STG
+abstract machine, this bridging process is not entirely
+straightforward.  The challenges we've seen fit into two categories.
+
+\subsection{Producing Full STG Programs}
+
+When compiling a standard Haskell program with GHC, modules imported
+from other packages have already been compiled, and GHC simply links
+the user object code with them.  Our compiler has no facility for
+linking GHC-produced object code, so we need it to produce
+\textit{full} STG programs built up from primitive operations.  GHC
+does not offer object-to-STG translation, so this means we must
+produce our own Prelude and supporting Base libraries to be included
+with every program, and decoupling Haskell from GHC's wired-in types
+as much as possible.
+
+This is partially accomplished with extensions like
+\texttt{NoImplicitPrelude} and \texttt{RebindableSyntax} (which
+implies the former).  We cannot get a complete separation, however:
+Numeric literals are tied to GHC's \texttt{Integer} and
+\texttt{Rational} types, and guards are tied to GHC's \texttt{Bool}.
+Lists and tuples, being special syntactic forms are also tightly
+wired-in. To get around this, we provide placeholder definitions that
+parallel any that we cannot replace, and use them to crudely translate
+the exceptional cases.
+
+\subsection{Navigating GHC's Internals}
+
+One of the goals of using GHC as a library was to avoid being
+overwhelmed with the necessary complexity of such a large and mature
+tool, but this has not been entirely unavoidable.  Simply producing
+STG without incurring any (non-temporary) file creation proved not to
+be a difficult task, and can be accomplished using only the interface
+exposed via the primary API, the \texttt{GHC} module.  This abstract
+STG is even ostensibly very close to our own, but the auxiliary data
+structures (e.g. those encapsulating identifier information and types)
+are rich and deeply nested.  Determining what is strictly necessary
+for our backend and what we have no implementation of (e.g. the
+Foreign Function Interface) has taken considerable time and effort.
+
+It is a testament to the talents of the GHC contributors and
+maintainers that we have managed to come even this far.  Their
+documentation and exposed interfaces have made navigating and
+comprehending the project internals much more pleasant than expected.
+
 \section{Mini-Haskell}

### 2016-07-12 pubs/tfp16-draft/paper.tex
Trims Mini-Haskell appendix.  Minor edits therein.

@@
 % because # looks strangely large next to math terms
-
 \newcommand{\hash}{{\scriptsize\#}}
+
 \begin{figure}[h!tb]
@@
 
-Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$ & Constructors are first class \\% not sure if that's the right terminology 
+Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$ & Constructors are first class \\
 \\
@@
 In STG, all object definitions require explicit classification with
-\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
-inferred from context by the compiler so these keywords are not necessary.
-Functions can then be defined in Haskell style with the formal parameters following
-the function's identifier on the left side of the equality. With these simple
-changes the syntax becomes a subset of Haskell.
-
-Heap objects are still only created at the top level or in \texttt{let}
-expressions, so the translation is straightforward: \texttt{FUN} objects are
+\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This
+classification can be inferred from context by the compiler so these
+keywords are not necessary.  Functions can then be defined in Haskell
+style with the formal parameters following the function's identifier
+on the left side of the equality. With these simple changes the syntax
+becomes a subset of Haskell. Heap objects are still only created at
+the top level or in \texttt{let} expressions, so the translation is
+straightforward.
+
+\begin{comment} Trimming Mini-Haskell to fit page limit
+\texttt{FUN} objects are
 identifiable by the inclusion of these parameters.  A \texttt{PAP} object is
@@
 \texttt{THUNK}.
+\end{comment}
 
@@
 
+\begin{comment} Trimming Mini-Haskell to fit page limit
 The inclusion of the layout rule may seem insignificant, but it
@@
 indentation, e.g., \emph{haskell-mode} for Emacs~\cite{haskellmode}.
+\end{comment}
 
@@
 \begin{verbatim}
-   append = FUN(l1 l2 ->
-               case l1 of
-                 { Nil -> l2;
-                   Cons hd tl -> 
-                     let { rec = THUNK(append tl l2);
+   append = FUN(l1 l2 -> case l1 of x {
+               Nil -> l2; 
+               Cons hd tl -> let { rec = THUNK(append tl l2);
                                    result = CON(Cons hd rec) }
@@
 all their arguments. Named functions may have multiple definitions to exhaust
-the combinations of patterns in their arguments and patterns in all contexts
+the combinations of patterns in their arguments, and patterns in all contexts
 (functions and case expressions) may be nested arbitrarily deep.  This feature
@@
    unzip = FUN(list -> 
-     case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
-                    Cons x xs ->
-                      case x of {
-                        TP2 a b ->     --TP2 is the pair constructor
-                          case unzip xs of {
-                            TP2 as bs ->
-                              let {aas = CON(Cons a as);
+     case list of s1 { Nil -> let {res = CON( Pair nil nil)} in res;
+                    Cons x xs -> case x of s2 {
+                      Pair a b -> case unzip xs of s3 {
+                        Pair as bs -> let {aas = CON(Cons a as);
                                            bbs = CON(Cons b bs);
-                                   res = CON(TP2 aas bbs)}
+                                           res = CON(Pair aas bbs)}
                                       in res }}})
@@
 \begin{verbatim}
-   unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
-   unzip Nil = TP2 Nil Nil
-   unzip (Cons (TP2 a b) xs) =
+   unzip :: List (Pair a b) -> Pair (List a) (List b)
+   unzip Nil = Pair Nil Nil
+   unzip (Cons (Pair a b) xs) =
      case unzip xs of
-       TP2 as bs -> TP2 (Cons a as) (Cons b bs)
+       Pair as bs -> Pair (Cons a as) (Cons b bs)
 \end{verbatim}}
@@
 on the compilation of this type of pattern matching proved
-invaluable~\cite{Jones:1987}.  Wadler describes the \emph{match} function
+invaluable~\cite{Jones:1987}.
+
+\begin{comment} Trimming Mini-Haskell to fit page limit
+Wadler describes the \emph{match} function
 which, in its simplest form, produces a correct, but potentially large
@@
 additional complexity outweighs the benefit for a testing language.
+\end{comment}
 
@@
   is implemented by simply generating a function that wraps the appropriate
-  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
-  constructor generates a call to this function.  Fully applied constructors
-  generate local \texttt{let}-bound objects to avoid an unnecessary function
-  call.
+  \texttt{let}-bound \texttt{CON} object in STG.  %% Any partially applied
+  %% constructor generates a call to this function.  Fully applied constructors
+  %% generate local \texttt{let}-bound objects to avoid an unnecessary function
+  %% call.
 
@@
   Similar to constructors, primitive operations are given first-class status as
-  functions, with roughly the same rules.  Partial applications are
-  translated to calls to generated functions that wrap the actual primitive
-  operations. Fully applied calls generate true primitive operation calls.
+  functions, with roughly the same rules.  %% Partial applications are
+  %% translated to calls to generated functions that wrap the actual primitive
+  %% operations. Fully applied calls generate true primitive operation calls.
 
@@
   mini-Haskell supports anonymous lambda expressions for the simple, one-off
-  functions that often find a place in Haskell code.  This was certainly the
-  simplest feature to add, requiring only that the function be given a unique
-  identifier to bind it to a \texttt{FUN} object in STG.
+  functions that often find a place in Haskell code.%%   This was certainly the
+  %% simplest feature to add, requiring only that the function be given a unique
+  %% identifier to bind it to a \texttt{FUN} object in STG.
 
@@
   inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
-  primitives (e.g. \texttt{(+)} and \texttt{(-)}) to the names given in
+  primitives (e.g. \texttt{(+#)} and \texttt{(-#)}) to the names given in
   mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored

### 2016-07-12 pubs/tfp16-draft/paper.tex
Trims Mini-Haskell appendix.  Minor edits therein.

@@
      case unzip xs of
-       Pair as bs -> Pair (Cons a as) (Cons b bs)
+       TP2 as bs -> Pair (Cons a as) (Cons b bs)
 \end{verbatim}}

### 2016-07-12 runtime/args.c
first cut at performance counters

@@ -11,12 +11,14 @@ int constStrict = 0;
 
 bool sanityChecker = false;
 
+int perfCounters = 0;
+
 int LoggingLevel = LOG_LEVEL;
 
 
 void parseArgs (int argc, char **argv) {
   int c;
-  while ((c = getopt (argc, argv, "gt:e:c:l:s")) != -1)
+  while ((c = getopt (argc, argv, "gt:e:c:l:sp:")) != -1)
     switch (c) {
     case 'g':
       GCThreshold = 1.0; // never run
@@ -40,6 +42,10 @@ void parseArgs (int argc, char **argv) {
       
     case 's':
       sanityChecker = true;
+      break;
+      
+    case 'p':
+      perfCounters = atoi(optarg);
       break;
         
     }

### 2016-07-12 runtime/gc.c

@@ -10,6 +10,7 @@
 #include "obj.h"
 #include "options.h"
 #include "sanity.h"
+#include "args.h"
 
 static void *scanPtr, *freePtr;
 
@@ -68,6 +69,7 @@ PtrOrLiteral updatePtrByValue (PtrOrLiteral f) {
       int size = getObjSize(p);
       LOG(LOG_SPEW, "copy %s %s from->to size=%d\n",
                     objTypeNames[getObjType(p)], p->ident, size);
+      if (perfCounters) perfCounter.heapBytesCopied += size;
       memcpy(freePtr, p, size);
       if (EXTRA_CHECKS_GC) {
 	      assert(isLSBset((InfoTab *)freePtr) == 0 && "gc: bad alignment");
@@ -199,6 +201,9 @@ void gc(void) {
 
   size_t before = stgHP - stgHeap;
 
+  if (perfCounters > 1) {
+    if (before > perfCounter.heapMaxSize) perfCounter.heapMaxSize = before;
+  }
   if(LOG_LEVEL == LOG_SPEW) {
     if(LoggingLevel == LOG_SPEW) {
       LOG(LOG_SPEW, "start gc heap size %lx\n", before);
@@ -206,13 +211,15 @@ void gc(void) {
   }
 
   if (sanityChecker) {
-    LOG(LOG_SPEW, "before GC\n");
-    heapCheck(true, LOG_DEBUG);
-    stackCheck(true, LOG_DEBUG);
+    LOG(LOG_NONE, "before GC\n");
+    heapCheck(true, LOG_NONE);
+    stackCheck(true, LOG_NONE);
   }
 
   // check if GC should run at all
   if(stgHP-stgHeap <= GCThreshold*stgHeapSize) return;
+  
+  if (perfCounters) perfCounter.heapCollections++;
 
   // add stgCurVal
   if (stgCurVal.op != NULL)
@@ -245,9 +252,9 @@ void gc(void) {
   swapPtrs();
 
   if (sanityChecker) {
-    LOG(LOG_SPEW, "after GC\n");
-    heapCheck(true, LOG_DEBUG);
-    stackCheck(true, LOG_DEBUG);
+    LOG(LOG_NONE, "after GC\n");
+    heapCheck(true, LOG_NONE);
+    stackCheck(true, LOG_NONE);
   }
 
   if(LOG_LEVEL == LOG_SPEW) {

### 2016-07-12 runtime/heap.c

@@ -75,6 +75,10 @@ Obj* stgNewHeapObj(InfoTab *itp) {
   }
   size_t objSize = sizeof(Obj) + payloadSize * sizeof(PtrOrLiteral);
   objSize = ((objSize + OBJ_ALIGNM1)/OBJ_ALIGN)*OBJ_ALIGN;
+  if (perfCounters) {
+    perfCounter.heapBytesAllocated += objSize;
+    perfCounter.heapAllocations++;
+  }
   Obj *objp = (Obj *)stgHP;
   stgHP = (char *)stgHP + objSize;
 
@@ -113,6 +117,10 @@ Obj* stgNewHeapPAPmask(InfoTab *itp, Bitmap64 bm) {
   size_t objSize = sizeof(Obj) +
     (fvCount + bm.bitmap.size + 1) * sizeof(PtrOrLiteral);
   objSize = ((objSize + OBJ_ALIGNM1)/OBJ_ALIGN)*OBJ_ALIGN;
+  if (perfCounters) {
+    perfCounter.heapBytesAllocated += objSize;
+    perfCounter.heapAllocations++;
+  }
   Obj *objp = (Obj *)stgHP;
   stgHP = (char *)stgHP + objSize;
 #if USE_ARGTYPE

### 2016-07-12 runtime/show.h

@@ -7,5 +7,4 @@
 void showStgObj(LogLevel priority, Obj *);
 void showStgCont(LogLevel priority, Cont *c);
 
-
 #endif

### 2016-07-12 pubs/tfp16-draft/paper.tex
fitting to page limit

--- 2016-07-12 pubs/tfp16-draft/paper.tex
@@
 
-\subsection{Strictness and Parallelism}
+%\subsection{Strictness and Parallelism}
 %It is a trivial observation that given strict functional semantics, wherein
@@
 
+\section{Project and Goals}
 Given strict functional semantics wherein function arguments may be safely
@@
 as functions.
-
-\section{Project and Goals}
+%
 Our project is the light-weight implementation of a pure, higher-order,
@@
 be kept as simple as reasonably possible.
-
+%
 Our initial requirements are that
@@
 
-\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
+%\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
 %every new idea has some sort of precedent or echo from the past~\cite{nihil}.
+There is a vast body of related work; we highlight only the very most closely related.
 
@@
 \subsubsection{Strict but still pure.}
+\vspace{-0.1in}
+%\paragraph{Strict but still pure.}
 The concept of strict but still pure in some sense dates back to the concept
@@
 
+\vspace{-0.1in}
 \subsubsection{Strict, pure, auto-parallelizing.}
+%\paragraph{Strict, pure, auto-parallelizing.}
 NVIDIA's NOVA language, and corresponding compiler, implements a strict
@@
 
+\vspace{-0.1in}
 \subsubsection{GHC as a front-end.}
+%\paragraph{GHC as a front-end.}
 The Intel Labs Haskell Research Compiler (HRC) is the most closely related
@@
 
+\vspace{-0.1in}
 \subsubsection{Fault tolerance for functional programming.}
+%\paragraph{Fault tolerance for functional programming.}
 Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
@@
 
+\vspace{-0.1in}
 \subsubsection{Removal of explicit heap objects.}
@@
 
+\vspace{-0.1in}
 \subsubsection{Avoiding building thunks.}
@@
 
+\vspace{-0.1in}
 \subsubsection{The layout rule.}
@@
 
+\vspace{-0.1in}
 \subsubsection{Expression to expression application.}
-
 In STG function application must be to atomic values, i.e., either literals or
@@
                  { Nil -> l2;
-                   Cons hd tl -> 
-                     let { rec = THUNK(append tl l2);
+                   Cons hd tl -> let { rec = THUNK(append tl l2);
                                        result = CON(Cons hd rec) }
@@
 
+\vspace{-0.1in}
 \subsubsection{Fancy pattern matching.}
-
 STG supports pattern matching but only in a simple form and only in
@@
 
+\vspace{-0.1in}
 \subsubsection{Less significant features.}
-
 Mini-Haskell provides some smaller conveniences that don't warrant individual

### 2016-07-12 pubs/tfp16-draft/paper.tex

@@
 \begin{comment}
- breaks the page limit
+% breaks the page limit
 One would expect an automated STG generator such as GHC to factor recursive
@@
 In STG, all object definitions require explicit classification with
-\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
-inferred from context by the compiler so these keywords are not necessary.
-Functions can then be defined in Haskell style with the formal parameters following
-the function's identifier on the left side of the equality. With these simple
-changes the syntax becomes a subset of Haskell.
-
-Heap objects are still only created at the top level or in \texttt{let}
-expressions, so the translation is straightforward: \texttt{FUN} objects are
+\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This
+classification can be inferred from context by the compiler so these
+keywords are not necessary.  Functions can then be defined in Haskell
+style with the formal parameters following the function's identifier
+on the left side of the equality. With these simple changes the syntax
+becomes a subset of Haskell. Heap objects are still only created at
+the top level or in \texttt{let} expressions, so the translation is
+straightforward.
+%Trimming Mini-Haskell to fit page limit
+%\begin{comment}
+\texttt{FUN} objects are
 identifiable by the inclusion of these parameters.  A \texttt{PAP} object is
@@
 \texttt{THUNK}.
+%\end{comment}
 
@@
 exactly that: edge cases.  As such, they are easy to avoid in practice.
-
+%Trimming Mini-Haskell to fit page limit
+\begin{comment} 
 The inclusion of the layout rule may seem insignificant, but it
@@
 \begin{verbatim}
-   unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
-   unzip Nil = TP2 Nil Nil
-   unzip (Cons (TP2 a b) xs) =
+   unzip :: List (Pair a b) -> Pair (List a) (List b)
+   unzip Nil = Pair Nil Nil
+   unzip (Cons (Pair a b) xs) =
      case unzip xs of
-       TP2 as bs -> TP2 (Cons a as) (Cons b bs)
+       Pair as bs -> Pair (Cons a as) (Cons b bs)
+
 \end{verbatim}}
@@
 on the compilation of this type of pattern matching proved
-invaluable~\cite{Jones:1987}.  Wadler describes the \emph{match} function
+invaluable~\cite{Jones:1987}.
+%Trimming Mini-Haskell to fit page limit
+%\begin{comment} 
+Wadler describes the \emph{match} function
 which, in its simplest form, produces a correct, but potentially large
@@
 additional complexity outweighs the benefit for a testing language.
+%\end{comment}
 
@@
   inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
-  primitives (e.g. \texttt{(+)} and \texttt{(-)}) to the names given in
+  primitives (e.g. \texttt{(+\#)} and \texttt{(-\#)}) to the names given in
   mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored

### 2016-07-14 pubs/tfp16-draft/paper.tex
paper minor edits

@@
 such path could in principle lead to the recreation of GHC\@.  However, given
-the minuscule budget (and therefore the employment of student interns) we must
-maintain sharp focus on the main goals: implementing and evaluating mechanisms
-for automatic parallelism and fault tolerance.  Currently we have
-a fully functioning serial implementation including
+the time constraints we must maintain sharp focus on the main goals:
+implementing and evaluating mechanisms for automatic parallelism and fault
+tolerance.  Currently we have a fully functioning serial implementation
+including
 \begin{itemize}
 \item STG parser,
-\item mini-Haskell parser, mini-Haskell to STG transformer,
+\item mini-Haskell parser, mini-Haskell to STG AST transformer,
+\item partial coupling with the GHC front-end,
 \item type inferencer,
@@
 \item code generator,
-\item runtime system including a garbage collector,
+\item runtime system including a (serial) garbage collector and instrumentation framework,
 \item continuous integration and testing system,
@@
 \end{itemize}
-
-The \emph{mini-Haskell} front-end is detailed in Appendix~\ref{app:minihaskell}.
+Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary}
+describe the \emph{mini-Haskell} front-end and our approach to using GHC as a front end, respectively.
+%
+
+While wall-clock time can be used to simply compare the relative efficiency of
+execution under various evaluation strategies, we seek more detailed
+information.  We have in place a simple infrastructure for optionally logging
+a thorough collection of run-time statistics such as number, size, and type of heap
+and stack allocations, as well as various garbage collection statistics.  We also
+measure cumulative time spent in various operations.  Future work includes processing
+these measurement in an other than ad hoc fashion.
 
@@
 %
-Bridging the gap between GHC STG and our STG, and the corresponding
-augmentation of the runtime is showing early signs of success.
+%Bridging the gap between GHC STG and our STG, and the corresponding
+%augmentation of the runtime is showing early signs of success.
 %
@@
 notionally concrete programming language the first thorough explications may
-be Backus' FP~\cite{Backus:1978} and follow-on FL
-implementation~\cite{AikenFL,FLreport89}.
+be Backus' FP~\cite{Backus:1978} (building on Landin's ``theory''~\cite{Landin64}).
+% and follow-on FL implementation~\cite{AikenFL,FLreport89}.
 %
@@
 \appendix
-
-\section{(Ab)Using GHC}
-
-The bridge to full Haskell is in progress, with GHC providing much of
-the structure.  We are using GHC as a library (as opposed to modifying
-its source to use our code generation and runtime system).  This
-choice was made primarily due to the size and complexity of a project
-as large as GHC.  Fully understanding and modifying even its build
-system would be a significant undertaking.
-
-We are using GHC's pipeline to translate Haskell source to GHC's
-representation of the STG language, which we then translate to our
-own.  Up to the STG generation stage, this usage is equivalent to
-GHC's \texttt{--make} behavior, which allows us to benefit from the
-module dependency analysis, typechecking, Core-level transformations
-and informative error messages for invalid programs.
-
-Though we follow the same rough idea as GHC in implementing the STG
-abstract machine, this bridging process is not entirely
-straightforward.  The challenges we've seen fit into two categories.
-
-\subsection{Producing Full STG Programs}
-
-When compiling a standard Haskell program with GHC, modules imported
-from other packages have already been compiled, and GHC simply links
-the user object code with them.  Our compiler has no facility for
-linking GHC-produced object code, so we need it to produce
-\textit{full} STG programs built up from primitive operations.  GHC
-does not offer object-to-STG translation, so this means we must
-produce our own Prelude and supporting Base libraries to be included
-with every program, and decoupling Haskell from GHC's wired-in types
-as much as possible.
-
-This is partially accomplished with extensions like
-\texttt{NoImplicitPrelude} and \texttt{RebindableSyntax} (which
-implies the former).  We cannot get a complete separation, however:
-Numeric literals are tied to GHC's \texttt{Integer} and
-\texttt{Rational} types, and guards are tied to GHC's \texttt{Bool}.
-Lists and tuples, being special syntactic forms are also tightly
-wired-in. To get around this, we provide placeholder definitions that
-parallel any that we cannot replace, and use them to crudely translate
-the exceptional cases.
-
-\subsection{Navigating GHC's Internals}
-
-One of the goals of using GHC as a library was to avoid being
-overwhelmed with the necessary complexity of such a large and mature
-tool, but this has not been entirely unavoidable.  Simply producing
-STG without incurring any (non-temporary) file creation proved not to
-be a difficult task, and can be accomplished using only the interface
-exposed via the primary API, the \texttt{GHC} module.  This abstract
-STG is even ostensibly very close to our own, but the auxiliary data
-structures (e.g. those encapsulating identifier information and types)
-are rich and deeply nested.  Determining what is strictly necessary
-for our backend and what we have no implementation of (e.g. the
-Foreign Function Interface) has taken considerable time and effort.
-
-It is a testament to the talents of the GHC contributors and
-maintainers that we have managed to come even this far.  Their
-documentation and exposed interfaces have made navigating and
-comprehending the project internals much more pleasant than expected.
 
@@
 
+\section{Using GHC as a Library}
+\label{app:ghclibrary}
+
+The bridge to full Haskell is in progress, with GHC providing much of
+the structure.  We are using GHC as a library (as opposed to modifying
+its source to use our code generation and runtime system).  This
+choice was made primarily due to the size and complexity of a project
+as large as GHC.  Fully understanding and modifying even its build
+system would be a significant undertaking.
+
+We are using GHC's pipeline to translate Haskell source to GHC's
+representation of the STG language, which we then translate to our
+own.  Up to the STG generation stage, this usage is equivalent to
+GHC's \texttt{--make} behavior, which allows us to benefit from the
+module dependency analysis, typechecking, Core-level transformations
+and informative error messages for invalid programs.
+
+Though we follow the same rough idea as GHC in implementing the STG
+abstract machine, this bridging process is not entirely
+straightforward.  The challenges we've seen fit into two categories.
+
+%\subsection{Producing Full STG Programs}
+\subsubsection{Producing full STG programs.}
+When compiling a standard Haskell program with GHC, modules imported
+from other packages have already been compiled, and GHC simply links
+the user object code with them.  Our compiler has no facility for
+linking GHC-produced object code, so we need it to produce
+\textit{full}\/ STG programs built up from primitive operations.  GHC
+does not offer object-to-STG translation, so this means we must
+produce our own Prelude and supporting Base libraries to be included
+with every program, and decoupling Haskell from GHC's wired-in types
+as much as possible.
+
+This is partially accomplished with extensions like
+\texttt{NoImplicitPrelude} and \texttt{RebindableSyntax} (which
+implies the former).  We cannot get a complete separation, however:
+numeric literals are tied to GHC's \texttt{Integer} and
+\texttt{Rational} types, and guards are tied to GHC's \texttt{Bool}.
+Lists and tuples, being special syntactic forms, are also tightly
+wired-in. To get around this we provide placeholder definitions that
+parallel any that we cannot replace, and use them to crudely translate
+the exceptional cases.
+
+%\subsection{Navigating GHC's Internals}
+\subsubsection{Navigating GHC's internals.}
+One of the goals of using GHC as a library was to avoid
+the internal complexity of such a large and mature
+tool, but this has not been entirely unavoidable.  Simply producing
+STG without incurring any (non-temporary) file creation proved not to
+be a difficult task, and can be accomplished using only the interface
+exposed via the primary API, the \texttt{GHC} module.  This abstract
+STG is even ostensibly very close to our own, but the auxiliary data
+structures (e.g. those encapsulating identifier information and types)
+are rich and deeply nested.  Determining what is strictly necessary
+for our backend and what we have no implementation of (e.g. the
+Foreign Function Interface) has taken considerable time and effort.
+
+It is a testament to the talents of the GHC contributors and
+maintainers that we have managed to come even this far.  Their
+documentation and exposed interfaces have made navigating and
+comprehending the project internals much more pleasant than expected.
+
 \section{Test suite and testing}

### 2016-07-15 pubs/tfp16-draft/paper.tex
paper as submitted

@@
 \begin{abstract}
-
-A claimed trend is an increasing appreciation and practice of functional
-programming in scientific computing, and in particular strict (by default)
-functional programming, in various guises.  Besides making large programs
-easier to reason about, this enables automatic parallelization to various
-degrees.  A second claimed trend in both scientific computing and functional
-programming is the development of mechanisms for transparent fault tolerance.
-Our project goal is the demonstration of a light-weight, higher-order,
-polymorphic, pure functional language implementation in which we can
-experiment with automatic parallelization strategies, varying degrees of
-strictness, and mechanisms for transparent fault tolerance.
+{\color{OliveGreen} % \bf
+\emph{Draft paper notes, per the CfP:  Ringo, Anderson, and Marks are undergraduate
+students.  This is a ``project'' paper.}
+}
+
+There is an increasing appreciation and practice of functional programming in
+scientific computing, and in particular strict (by default) functional
+programming, in various guises.  Besides making large programs easier to
+reason about, this enables automatic parallelization to various degrees.  A
+second trend in both scientific computing and functional programming is the
+development of mechanisms for transparent fault tolerance.  Our project goal
+is the demonstration of a light-weight, higher-order, polymorphic, pure
+functional language implementation in which we can experiment with automatic
+parallelization strategies, varying degrees of strictness, and mechanisms for
+transparent fault tolerance.
 %
@@
 The practice of so-called high-performance scientific computing (HPC) is,
-overall, highly conservative with respect to change.  Even as the USA and
-other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
+overall, highly conservative with respect to change.  Even as 
+various governments commit to a push for exascale computing ($10^{18}$ FLOPS in
 a single system) within the next
@@
 member of the HPC community recently stated publicly that ``Fortran is
-essential for exascale programming''~\cite{Heroux16}.  Existing US government
+essential for exascale programming,'' meaning that existing codes
+will be adapted, not rewritten~\cite{Heroux16}.  Existing US government
 laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
@@
 entirely up to the programmer to make functions composable and thread safe.
+Here at LANL the concept of \emph{pure functional} is being used to good
+effect to clean up---restructure, modularize, reduce to composable
+components---some legacy code bases.
 %Here at LANL the use of pure functions (among other techniques for reducing
@@
 implementation of a pure functional language to date, to no
-effect whatsoever~\cite{Davis96}.\footnote{In 1996, on the LANL Thinking Machines
+effect whatsoever~\cite{Davis96}.\footnote{In 1996 on the LANL Thinking Machines
 CM-5, the Top 1 supercomputer in 1993~\cite{top1:1993}.}
@@
   represents, or in other units, one second for each hour.}
-For another, it means that we are not undertaking significant morphing of GHC
-itself: because this is in large part an undergraduate student project it must
-be kept as simple as reasonably possible.
+For another, it means that we are not undertaking morphing of GHC
+itself: the system is huge and complex.
 %
@@
 
-In short, our plan is to use GHC as our front-end to generate one of its
+In short, we will use GHC as the front-end to generate one of its
 intermediate forms, STG or Core, then escape to our system from there.  There
 are various possible ways to accomplish the this: using GHC's option to dump
-STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
+STG or Core as text, hacking GHC itself, or as we have chosen, using GHC as a
 library~\cite{ghc-as-library}.  Initially we started with our own STG.
@@
 in \emph{EA}.
-
+%
 In STG the arity of a function is defined as the number of formal parameters
@@
 transition rules which may be different than those for non-strict semantics.
-Sharing is preserved.
 
@@
 \item code generator,
-\item runtime system including a (serial) garbage collector and instrumentation framework,
-\item continuous integration and testing system,
-\item extensive test suite, automated testing.
+\item runtime system including a (serial) garbage collector and instrumentation framework, and
+\item continuous integration and testing system, extensive test suite.
 \end{itemize}
-Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary}
-describe the \emph{mini-Haskell} front-end and our approach to using GHC as a front end, respectively.
+Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary} describe the
+mini-Haskell front-end and our approach to using GHC as a front end,
+respectively.
 %
@@
 information.  We have in place a simple infrastructure for optionally logging
-a thorough collection of run-time statistics such as number, size, and type of heap
-and stack allocations, as well as various garbage collection statistics.  We also
-measure cumulative time spent in various operations.  Future work includes processing
-these measurement in an other than ad hoc fashion.
+a thorough collection of run-time statistics such as number, size, and type of
+heap and stack allocations, as well as various garbage collection statistics.
+We also measure cumulative time spent in various operations.  Future work
+includes integrating this with the testing system, processing these statistics
+in an other than ad hoc fashion.
 
@@
 %
-An instrumentation infrastructure, integrated with the testing framework to
-record and present time and space run-time statistics such as number of heap
-and stack allocations and maximum stack depth, is planned for the near future.
-%
 The fault-tolerance mechanisms remain future work.
@@
 %every new idea has some sort of precedent or echo from the past~\cite{nihil}.
-There is a vast body of related work; we highlight only the very most closely related.
+There is a vast body of related work; we highlight only the most similar to ours.
 
@@
 be Backus' FP~\cite{Backus:1978} (building on Landin's ``theory''~\cite{Landin64}).
-% and follow-on FL implementation~\cite{AikenFL,FLreport89}.
-%
+
 A notable argument for the potential merits of a strict-by-default
@@
 a very simple embedded implementation~\cite{Sheard:2003}.
-%
+
 A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
@@
 bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
-%
+
+Manticore is a derivative of Standard ML with no mutable state and both
+implicit and explicit (threaded) mechanisms for parallelism.  Threads
+communicate and synchronize via synchronous channels~\cite{Manticore}.
+Numerous other ML derivatives, e.g.\ MultiMLton~\cite{MultiMLton},
+OCaml~\cite{OCaml}, Poly/ML~\cite{PolyML}, and more distantly
+NESL~\cite{NESL}, support various forms of parallelism.
+
 The Disciplined Disciple Compiler project appears to intend implementation of
@@
 and has been underway for some time, but the emphasis appears to be on a much
-more complex type system for handling stateful computation, and the prospects
-for its eventual completion are unknown~\cite{disciplined-disciple}.
+more complex type system for handling stateful computation~\cite{disciplined-disciple}.
 
@@
 
+Futhark is a statically-typed, data-parallel,
+pure functional language targeting GPUs via OpenCL~\cite{Futhark}.  Notably, it
+supports regular nested data-parallelism, and a purity-preserving
+update-in-place mechanism via a uniqueness type system.
+
 \vspace{-0.1in}
@@
 %\paragraph{GHC as a front-end.}
-The Intel Labs Haskell Research Compiler (HRC) is the most closely related
-work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
-planned approach they use GHC as the front end, but exit GHC at the final GHC
-Core stage, just before transformation to STG.  The essence of their effort is
-to target an existing, highly optimizing, Intel functional language compiler
-that is ``largely language agnostic'' and performs (among many other
-optimizations) SIMD vectorization.
+The Intel Labs Haskell Research Compiler is the most closely related work to ours
+of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our approach they
+use GHC as the front end, but exit GHC at the final GHC Core stage, just
+before transformation to STG.  The essence of their effort is to target an
+existing, highly optimizing, Intel functional language compiler that is
+``largely language agnostic'' and performs (among many other optimizations)
+SIMD vectorization.
 
@@
 \label{app:minihaskell}
-As mentioned, a current near-term goal is to use GHC as our front-end to
-generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
-measure to make the generation of test cases easier, we have implemented a
+As mentioned, we are transitioning to GHC as our front-end to
+generate STG (or Core) as input to our system.  Previously, as a stop-gap
+measure to make the generation of test cases easier, we implemented a
 \emph{mini-Haskell} front-end that supports a subset of standard Haskell
@@
 is similar to primitive Haskell, it lacks many of the conveniences that make
-Haskell a pleasure to program in.  The motivation for a more expressive
+Haskell convenient to program in.  The motivation for a more expressive
 language for testing our compiler and runtime system in their nascent stages
@@
 approach would translate from mini-Haskell to GHC Core, then either to STG or,
-in the case that the back end eventually take, compiling Core directly.
+in the case that the back end may eventually take, compiling Core directly.
 
@@
 \vspace{-0.1in}
-\subsubsection{Fancy pattern matching.}
+\subsubsection{Nested pattern matching.}
 STG supports pattern matching but only in a simple form and only in
@@
 \end{verbatim}}
-
-Of course, this complex pattern matching must eventually be transformed into
-the primitive, nested case expressions. Fortunately there are well-documented
-algorithms for achieving this, and Philip Wadler's chapter
+This nested pattern matching must eventually be transformed into
+primitive, nested case expressions.  There are well-documented
+algorithms for achieving this, and Wadler's description
 % in \emph{The Implementation of Functional Programming Languages}
-on the compilation of this type of pattern matching proved
+of the compilation of this type of pattern matching proved
 invaluable~\cite{Jones:1987}.
@@
 %\begin{comment} 
-Wadler describes the \emph{match} function
-which, in its simplest form, produces a correct, but potentially large
-translation with some redundancy.  This function was implemented with only
-minor modifications to fit the shape of mini-Haskell's abstract form.  The
-implementation is augmented by some light flow-analysis to eliminate some of
-the redundancy and make the resulting code easier to read.  There are further
-optimizations suggested by Wadler to address the size of the generated code,
-for example, when multiple branches lead to the same large expression, but the
-additional complexity outweighs the benefit for a testing language.
+Wadler describes the \emph{match} function which, in its simplest form,
+produces a correct but potentially large translation with some redundancy.
+This function was implemented with only minor modifications to fit the shape
+of mini-Haskell's abstract form.  The implementation is augmented by some
+light flow-analysis to eliminate some of the redundancy and make the resulting
+code easier to read.  Wadler suggests further optimizations to address the
+size of the generated code, for example, when multiple branches lead to the
+same large expression, but the additional complexity outweighs the benefit for
+a testing language.
 %\end{comment}
@@
   is implemented by simply generating a function that wraps the appropriate
-  \texttt{let}-bound \texttt{CON} object in STG.  %% Any partially applied
-  %% constructor generates a call to this function.  Fully applied constructors
-  %% generate local \texttt{let}-bound objects to avoid an unnecessary function
-  %% call.
+  \texttt{let}-bound \texttt{CON} object in STG.  
+%
+  Any partially applied
+  constructor generates a call to this function.  Fully applied constructors
+  generate local \texttt{let}-bound objects to avoid an unnecessary function
+  call.
 
@@
   the compiler can verify.  Mini-Haskell's type annotations (or signatures)
-  behave like those in Haskell, being passed to the type checker as initial
+  behave like those in Haskell, being passed to the type checker as % initial
   assumptions.
@@
 documentation and exposed interfaces have made navigating and
-comprehending the project internals much more pleasant than expected.
+comprehending the internals much more pleasant than expected.
 
@@
 strictness by two degrees of constructor strictness).  Interesting classes of
-test programs are those that require less-strict or sharing semantics.  For
+test programs are those that require less-strict semantics.  For
 these programs we test not only that the test leads to the expected result
@@
 Another important class of test programs are those that should fail with a
-known error, e.g., a parsing or typing error.  For these programs the test
-suite checks the output against an expected error regular expression. This
-class of test programs has proved valuable in development and debugging of
-both the STG and mini-Haskell front-ends.
+known error, e.g., a syntax or typing error.  For these programs the test
+suite checks the output against an expected error regular expression. 
+%This
+%class of test programs has proved valuable in development and debugging of
+%both the STG and mini-Haskell front-ends.
 
@@
 
-We have also implemented a simple logging facility which can be controlled at 
-both compile and run time which helps in determining why a test is failing.
+%We have also implemented a simple logging facility which can be controlled at 
+%both compile and run time which helps in determining why a test is failing.
 

### 2017-02-22 pubs/demand_analysis/new.verb.tex
demand analysis manuscript

\documentclass[11pt]{article}

\usepackage{geometry}
\geometry{textwidth=7.5in,
  textheight=9in}

%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkScale{6}
%\SetWatermarkColor{yellow}

\begin{document}

\title{PERs from Partial Projections for Higher-order Polymorphic Demand Analysis}
\author{Kei Davis\\
  Los Alamos National Laboratory\\
  kei at lanl.gov}
\maketitle

\setcounter{page}{1} %% This command is optional. 
                       %% May set page number only for first page in
                       %% issue, if desired.

\newcommand{\com}{\newcommand}
\com{\rcom}{\renewcommand}

\rcom{\ddagger}{*}

% -------- A figure with a line at top and bottom, a caption and a label
% -------- Syntax: \fig{figoptions}{caption}{label}{ ..... }
\com{\fig}[4]{
\begin{figure}[#1]
\begin{center}
\begin{minipage}{\columnwidth}
\rule[1ex]{\textwidth}{1pt}
%\vspace{1ex}
#4
\vspace{-1em}
\caption{#2 \label{#3}}
\rule[1ex]{\textwidth}{1pt}
\end{minipage}
\end{center}
\vspace{-3ex}
\end{figure}
}

\com{\progsize}{}                       % displayed prog
\com{\subtypesize}{\footnotesize}       % sub type
\com{\subsubtypesize}{\scriptsize}      % footnote-sub or sub-sub type--\sstt
\com{\subtypesubstyle}{\scriptstyle}    % scriptscriptstyle/scriptstyle
\com{\subtypespace}{\hspace{0.0pt}}     % initial spacing before type subscript

%\com{\backup}{\hspace{-0.12cm}}
\com{\espace}{\vspace{6pt}}
\com{\itt}[1]{\mbox{\it #1}}
%\com{\Ss}{\sf{S}}
%\com{\Ps}{\sf{P}}

\com{\OUTC}{\itt{OUTC}}
\com{\SEL}{\itt{SEL}}
\com{\SPINE}{\itt{SPINE}}
\com{\LBR}{\itt{LBR}}

\com{\newenv}{\newenvironment}
\com{\subsubsect}[1]{\paragraph{{#1}}}

\com{\leqn}[1]{\lefteqn{\it{}#1}}

\com{\BSA}{BSA}
\com{\FSA}{FSA}
\com{\BLA}{BTA}
\com{\FLA}{FTA}
\com{\lba}[1]{\proj{#1}}
\com{\gfa}[1]{\proj{#1}}
\com{\lfa}[1]{\proj{#1}}

\com{\data}{{\it data}_{\stt}}
\com{\for}{{\it for}_{\stt}}
\com{\fac}{{\it fac}_{\stt}}
\com{\unfac}{{\it unfac}_{\stt}}

\com{\xxx}{\L{}x.(x,\ldots,x)}

\com{\proj}[1]{{|\,#1\,|}}
\com{\projl}[1]{\proj{\lift{#1}}}
\com{\projll}[1]{\proj{\liftt{#1}}}

\com{\bG}{\mbox{\boldmath$\Gamma$}}
\com{\ith}{i^{\it{}th}}
\com{\im}[1]{\it{}Im(#1)}

\com{\commentout}[1]{}
\com{\comment}[2]{\typeout{DON'T FORGET: #1}}
\com{\uncomment}[2]{#2}

\com{\inftyd}{{\bf\infty}}
\com{\iwe}{\we_i}
\com{\dual}[1]{#1^\partial}

\com{\psafety}{\G\o\f\we\f\o\D}
\com{\pliveness}{\G\o\f\se\f\o\D}
\com{\ptsafety}{\G\o\f\we\f\o(\T~\G)}

% predicate stuff

\com{\uF}{\u{F}}
\com{\dF}{\d{F}}
\com{\uphi}{\u{\phi}}
\com{\dphi}{\d{\phi}}
\com{\upsi}{\u{\psi}}
\com{\dpsi}{\d{\psi}}
\com{\uU}{\u{U}}
\com{\dU}{\d{U}}
\com{\uD}{\u{D}}
\com{\dD}{\d{D}}
\com{\uV}{\u{V}}
\com{\dV}{\d{V}}
\com{\udU}{\ud{U}}
\com{\uf}{\u{f}}
\com{\df}{\d{f}}
\com{\udf}{\ud{f}}
\com{\dx}{\d{x}}
\com{\ux}{\u{x}}
\com{\udx}{\ud{x}}

\com{\OK}{\G}

% semantics names

%\com{\tfil}[2]{#1{\lefteqn{{}_{\rm#2}}}\raisebox{-0.3ex}{${}^{{}^{\hspace{-0.06em}\bot}}$}}
%\com{\tfils}[3]{#1{\lefteqn{{}_{\rm#2}}}\raisebox{-0.3ex}{${}^{{}^{\hspace{-0.06em}\bot#3}}$}}

%\com{\tfil}[2]{#1{\lefteqn{{}_{\rm#2}}}_{}}
%\com{\tfils}[3]{#1{\lefteqn{{}_{\rm#2}}}\raisebox{-0.3ex}{${}^{{}^{\hspace{-0.06em}\bot#3}}$}}

\com{\tfil}[2]{#1_{\ubot\rm#2}}
\com{\tfils}[3]{#1_{{\ubot\rm#2}}^{#3}}

\com{\Dl}{\tfil{\Ds}{}}
\com{\Dlz}{\tfil{\Ds}{0}}
\com{\Sl}{\tfil{\Ss}{}}
\com{\Slz}{\tfil{\Ss}{0}}
\com{\Slo}{\tfil{\Ss}{1}}
\com{\Nl}{\tfil{\Ns}{}}
\com{\Nlz}{\tfil{\Ns}{0}}
\com{\Nlo}{\tfil{\Ns}{1}}
\com{\Nlt}{\tfil{\Ns}{2}}
\com{\Nlzs}{\tfils{\Ns}{0}{s}}
\com{\Nlts}{\tfils{\Ns}{2}{s}}
\com{\Nls}{\tfils{\Ns}{}{s}}

\com{\semfont}{\rm}		% sans serif
\com{\Ps}{\semfont P}
\com{\Pz}{\Ps_{\rm0}}

\com{\Gs}{\semfont G}
\com{\Gz}{\Gs_{\rm0}}
\com{\Go}{\Gs_{\rm1}}

\com{\Hs}{\semfont H}
\com{\Hz}{\Hs_{\rm0}}
\com{\Ho}{\Hs_{\rm1}}

\com{\Ds}{\semfont D}
\com{\Dz}{\Ds_{\rm0}}

\com{\Ss}{\semfont S}
\com{\Sz}{\Ss_{\rm0}}
\com{\So}{\Ss_{\rm1}}

\com{\Ms}{\semfont M}

\com{\Bs}{\semfont B}
\com{\Bz}{\Bs_{\rm0}}
\com{\Bo}{\Bs_{\rm1}}
\com{\Bt}{\Bs_{\rm2}}
\com{\Boh}{\Bs_{\rm1}^\#}
\com{\Bzh}{\Bs_{\rm0}^\#}
\com{\Bth}{\Bs_{\rm2}^\#}
\com{\Bd}{\Bs_{\rm{}d}}
\com{\Bdd}{\Bs_{\rm{}dd}}

\com{\Ls}{\semfont L}
\com{\Lz}{\Ls_{\rm0}}
\com{\Lo}{\Ls_{\rm1}}
\com{\Lt}{\Ls_{\rm2}}
\com{\Loh}{\Ls_{\rm1}^\#}
\com{\Lzh}{\Ls_{\rm0}^\#}
\com{\Lth}{\Ls_{\rm2}^\#}

\com{\Fs}{\semfont F}
\com{\Fz}{\Fs_{\rm0}}
\com{\Fo}{\Fs_{\rm1}}
\com{\Ft}{\Fs_{\rm2}}
\com{\Fzh}{\Fs_{\rm0}^\#}
\com{\Foh}{\Fs_{\rm1}^\#}
\com{\Fth}{\Fs_{\rm2}^\#}
\com{\Fd}{\Fs_{\rm{}d}}

\com{\Ns}{\semfont N}
\com{\Nz}{\Ns_{\rm0}}
\com{\No}{\Ns_{\rm1}}
\com{\Nt}{\Ns_{\rm2}}

\com{\Th}{\T^\#}

% data proj operators

\com{\up}[1]{#1\!\!\uparrow}
\com{\dn}[1]{#1\!\!\downarrow}
\com{\uG}{\up{\G}}
\com{\dG}{\dn{\G}}

% binary operators

\com{\amp}{\:\&\:}
\com{\lub}{\sqcup}
\com{\Lub}{\bigsqcup}
\com{\Amp}{\mbox{\large\&}}
\com{\glb}{\sqcap}
\com{\Glb}{\,{\overline{\raisebox{0ex}[1.4ex][.8ex]{\makebox[.65em]{$|\ \/|$}}}}\,}
\com{\guard}{\rhd}
\com{\comp}{\circ}

\com{\we}{\sqsubseteq}
\com{\se}{\sqsupseteq}
\com{\st}{\sqsupset}
\com{\wt}{\sqsubset}
\com{\eq}{\ =\ }
\com{\swe}{\sqsubseteq_{\rm{}s}} 
\com{\spwe}{\sqsubseteq_{\rm{}s+}}
\com{\swt}{\sqsubset_{\rm{}s}} 
\com{\gt}{>}
\com{\lt}{<}

% boolean values

\rcom{\land}{\wedge}
\rcom{\lor}{\vee}
\com{\liff}{\Leftrightarrow}
\com{\limp}{\Rightarrow}
\com{\iso}{\cong}
\com{\True}{{\it True}}
\com{\False}{{\it False}}

\com{\itlift}{{\it lift}}
\com{\itdrop}{{\it drop}}
\com{\itmkunit}{{\it mkunit}}
\com{\itmkinti}{{\it mkint_i}}
\com{\itmkfun}{{\it mkfun}}
\com{\itplus}{{\it plus}}
\com{\itcon}{{\it con}}
\com{\itseli}{{\it sel_i}}
\com{\ittuple}{{\it tuple}}
\com{\itchoose}{{\it choose}}
\com{\itapply}{{\it apply}}
\com{\itlfp}{{\it lfp}}
\com{\itfix}{{\it fix}}

% domains

\com{\ints}{{\bf{}Z}}
\com{\nullset}{\{\}}
\com{\Int}{{\it Int}}
\com{\Bool}{{\it Bool}}
\com{\ubot}{{\!\bot}}
\com{\one}{{\bf 1}}
\com{\unit}{{\mit()}}
\com{\two}{{\bf 2}}
\com{\three}{{\bf 3}}
\com{\lift}[1]{{{#1}_\ubot}}
\com{\liftt}[1]{{#1}_{\ubot_{\scriptstyle\ubot}}}
\com{\lifttp}[1]{({#1})_{\ubot_{\scriptstyle\ubot}}}
\com{\liftp}[1]{({#1})_\ubot}
\com{\llift}[1]{{{#1}_{\ubot'}}}
\com{\lliftp}[1]{{{(#1)}_{\ubot'}}}
\com{\slift}[1]{{{#1}_{\!\underline{\bot}}}}
\com{\sliftp}[1]{{({#1})_{\!\underline{\bot}}}}
\com{\lone}{\lift{\one}}
\com{\llone}{\liftt{\one}}
\com{\sone}{\slift{\one}}
\com{\lbot}{{\it lift~\bot}}
\com{\llbot}{{\it lift^{\rm2}~\bot}}
\com{\lunit}{{\it lift}~\unit}
\com{\llunit}{{\it lift}^{\rm2}~\unit}
\com{\botl}{{\rm1}}
\com{\botll}{{\rm2}}
\com{\U}{U}
\com{\id}{{\it id}}
\com{\Truth}{{\it Truth}}
\com{\lU}{\lift{U}}
\com{\V}{V}
\com{\lV}{\lift{V}}
\com{\W}{W}
\com{\ox}{\otimes}
\com{\bo}{\o^{B}}
\com{\fto}{\rightarrow} 
\com{\dto}{\stackrel{\cup}{\rightarrow}}
\com{\sto}{\stackrel{\rm s}{\rightarrow}}
\com{\sbto}{\stackrel{\rm sb}{\rightarrow}}
\com{\ssbto}{\stackrel{\rm ssb}{\rightarrow}}
\com{\sbbto}{\stackrel{\rm sbb}{\rightarrow}}
\com{\ito}{\stackrel{\rm i}{\rightarrow}}
\com{\gto}{\stackrel{\mit B}{\rightarrow}}
\com{\gpto}{\stackrel{\rm d}{\rightarrow}}
\com{\gdto}{\stackrel{\mit F}{\rightarrow}}
\com{\gtpo}{\stackrel{\rm gd}{\rightarrow}}
\com{\sgto}{\stackrel{\rm sg}{\rightarrow}}
\com{\rto}{\leftrightarrow}
\com{\plus}{+}

\com{\tmn}{\theta_{mn}}
\com{\tnm}{\theta_{nm}}
\com{\tmy}{\theta_{m\infty}}
\com{\tyy}{\theta_{\infty\infty}}
\com{\tiy}{\theta_{i\infty}}
\com{\tym}{\theta_{\infty{}m}}
\com{\tyz}{\theta_{\infty{\rm0}}}
\com{\tyo}{\theta_{\infty{\rm1}}}
\com{\tyi}{\theta_{\infty{}i}}
\com{\tyip}{\theta_{\infty{}i+{\rm1}}}
%\com{\th}{\theta}

% Greek letters

\com{\T}{\tau}
\com{\K}{\kappa}
\com{\R}{\rho}
\rcom{\S}{\sigma}
\rcom{\L}{\lambda}
\com{\Lx}{\lambda{}x}
\com{\sLx}{\underline{\lambda}x}
\com{\sL}{\underline{\lambda}}
\com{\Z}{\zeta}

% functions

\com{\lf}{\lift{f}}
\com{\f}{f}
\com{\g}{g}
\rcom{\lg}{\lift{g}}

% projections and projection transformers

\com{\A}{\alpha}
\com{\lA}{\lift{\A}}
\com{\sA}{\slift{\A}}
\com{\B}{\beta}
\com{\lB}{\lift{\B}}
%\com{\sB}{\slift{\B}}
\com{\G}{\gamma}
\com{\lG}{\lift{\G}}
%\com{\sG}{\slift{\G}}
\com{\D}{\delta}
\com{\lD}{\lift{\D}}
\com{\sD}{\slift{\D}}
\com{\BOT}{{\it BOT}}
\com{\lBOT}{\lift{\BOT}}
\com{\sBOT}{\slift{\BOT}}
\com{\ID}{{\it ID}}
\com{\LAM}{{\it LAM}}
\com{\lID}{\lift{\ID}}
\com{\sID}{\slift{\ID}}
\com{\ABS}{{\it ABS}}
\com{\STR}{{\it STR}}
\com{\Ab}{\alpha_{@@@b@@@}}
\com{\Ax}{\alpha_{@@@x@@@}}
\com{\Ay}{\alpha_{@@@y@@@}}
\com{\Az}{\alpha_{@@@z@@@}}
\com{\Axs}{\alpha_{@@@xs@@@}}
\com{\Ays}{\alpha_{@@@ys@@@}}
\com{\Azs}{\alpha_{@@@zs@@@}}

\com{\FNIL}{{\it FNIL}}
\com{\NIL}{{\it NIL}}
\com{\CONS}{{\it CONS}}
\com{\FCONS}{{\it FCONS}}
\com{\TRUE}{{\it TRUE}}
\com{\FALSE}{{\it FALSE}}
\com{\FAIL}{{\it FAIL}}
\com{\HEAD}{{\it HEAD}}
\com{\TAIL}{{\it TAIL}}
\com{\INF}{{\it INF}}
\com{\FINF}{{\it FINF}}
\com{\FIN}{{\it FIN}}
\com{\PAIR}{{\it PAIR}}
\com{\FST}{{\it FST}}
\com{\SND}{{\it SND}}
\com{\II}{{\it II}}
\com{\IF}{{\it IF}}
\com{\FI}{{\it FI}}
\com{\FF}{{\it FF}}

\com{\LA}{\lambda\alpha}
\com{\sLA}{\underline{\lambda}\alpha}
\com{\gLA}{\L\slift{\A}}
%\com{\gL}{\overline{\lambda}}
\com{\laal}{\L\A.\A}
\com{\laab}{\sL\A.\ABS}
\com{\laid}{\sL\A.\ID}
\com{\labo}{\gLA.\sBOT}

% odds & ends

\com{\dd}{\ddagger}
\com{\chain}[1]{\{#1_{\rm0},#1_{\rm1},#1_{\rm2},\ldots\}}

% typewriter font

\com{\oy}{\overline{@y@}}

\com{\xnot}{{\rm@x@_0}}
\com{\xone}{{\rm@x@_1}}
\com{\xtwo}{{\rm@x@_2}}
\com{\xthree}{{\rm@x@_3}}
\com{\xxi}{{@x@_i}}
\com{\xn}{{@x@_n}}

\com{\ee}{@e@}
\com{\ex}{@x@}
\com{\evar}{@x@}
\com{\enum}{@n@_i}
\com{\esum}{\eone\ @+@\ \etwo}
\com{\elam}{@\x.e@}
\com{\ehlam}{@\#x.e@}
\com{\elamt}{@\x:T.e@}
\com{\ehlamt}{@\#x:T.e@}
\com{\eapp}{\eone~\etwo}
\com{\ehapp}{@app# @\eone~\etwo}
\com{\efapp}{@f e@}
\com{\eunit}{@()@}
\com{\etup}{@(@\eone@,@\ldots@,@\en@)@}
\com{\esel}{@let (@\xone@,@\ldots@,@\xn@) =@\ \enot\ @in@\ \eone}
\com{\econ}{@c@_i\ @e@}
\com{\ecase}{@case@\ \enot\ @of @\cone\ \xone\ @->@\ \eone@;@\ \ldots@;@\ \cn\ \xn\ @->@\ \en}
\com{\efix}{@fix e@}
\com{\ehfix}{@fix# e@}
\com{\estr}{@seq@~\eone~\etwo}

\com{\enot}{{\rm@e@_0}}
\com{\eone}{{\rm@e@_1}}
\com{\etwo}{{\rm@e@_2}}
\com{\ei}{{@e@_i}}
\com{\en}{{@e@_n}}

\com{\senot}{{\rm@@@e@@@_{\subtypesubstyle0}}}
\com{\seone}{{\rm@@@e@@@_{\subtypesubstyle1}}}
\com{\setwo}{{\rm@@@e@@@_{\subtypesubstyle2}}}
\com{\sei}{{@@@e@@@_{\subtypesubstyle{}i}}}
\com{\sen}{{@@@e@@@_{\subtypesubstyle{}n}}}
\com{\setup}{{@@@(@@@\seone@@@,@@@\ldots@@@,@@@\sen@@@)@@@}}

\com{\fnot}{{\rm@f@_0}}
\com{\fone}{{\rm@f@_1}}
\com{\ftwo}{{\rm@f@_2}}
\com{\fii}{{@f@_i}}
\com{\fn}{{@f@_n}}

\com{\cnot}{{\rm{\tt{}c}_0}}
\com{\cone}{{\rm{\tt{}c}_1}}
\com{\ctwo}{{\rm{\tt{}c}_2}}
\com{\ci}{{{\tt{}c}_i}}
\com{\cn}{{{\tt{}c}_n}}
\com{\anot}{{\rm@A@_0}}
\com{\aone}{{\rm@A@_1}}
\com{\atwo}{{\rm@A@_2}}
\com{\ai}{{@A@_i}}
\com{\aj}{{@A@_j}}
\com{\an}{{@A@_n}}
\com{\ta}{@A@}
\com{\te}{@E@}
\com{\tenot}{{@E@_{\it gl}}}
\com{\teone}{{\rm@E@_1}}
\com{\tei}{{@E@_i}}
\com{\ten}{{@E@_n}}
\com{\tnot}{{\rm@T@_0}}
\com{\tone}{{\rm@T@_1}}
\com{\ttwo}{{\rm@T@_2}}
\com{\tthree}{{\rm@T@_3}}
\com{\ti}{{@T@_i}}
\com{\tj}{{@T@_j}}
\com{\tn}{{@T@_n}}
\com{\ttt}{@T@}
\com{\unot}{{\rm@U@_0}}
\com{\uone}{{\rm@U@_1}}
\com{\utwo}{{\rm@U@_2}}
\com{\uthree}{{\rm@U@_3}}
\com{\ui}{{@U@_i}}
\com{\un}{{@U@_n}}
\com{\tunit}{@()@}
\com{\tint}{@Int@}
\com{\tfun}{\tone\ @->@\ \ttwo}
\com{\tfunp}{@(@\tone\ @->@\ \ttwo@)@}
\com{\thfun}{\tone\ \harrow\ \ttwo}
\com{\thfunp}{@(@\tone\ \harrow\ \ttwo@)@}
\com{\harrow}{@#>@}
\com{\tprod}{@(@\tone@,@\ldots@,@\tn@)@}
\com{\teprod}{@(@\teone@,@\ldots@,@\ten@)@}
\com{\tsum}{\cone\ \tone\ @+@\ \ldots\ @+@\ \cn\ \tn}
\com{\tdefn}{\aone\ @=@\ \tone@;@\ \ldots@;@\ \an\ @=@\ \tn}

\com{\bproj}{{\rm\bf{}fproj}}
\com{\bfabsf}{{\rm\bf{}fabsf}}

\com{\scone}{{\rm@@@c@@@_1}}
\com{\scn}{{@@@c@@@_n}}
\com{\saone}{{\rm@@@A@@@_1}}
\com{\san}{{@@@A@@@_n}}

\com{\stenot}{{@@@E@@@_{\it\subtypesubstyle{}gl}}}
\com{\sten}{{@@@E@@@_{\subtypesubstyle{}n}}}
\com{\stnot}{{@@@T@@@_{\rm\subtypesubstyle0}}}
\com{\stone}{{@@@T@@@_{\rm\subtypesubstyle1}}}
\com{\sttwo}{{@@@T@@@_{\rm\subtypesubstyle2}}}
\com{\sti}{{@@@T@@@_{\subtypesubstyle{}i}}}
\com{\stn}{{@@@T@@@_{\subtypesubstyle{}n}}}

\com{\sta}{{@@@A@@@}}
\com{\ste}{{@@@E@@@}}
\com{\stt}{{@@@T@@@}}
\com{\stu}{{@@@U@@@}}
\com{\std}{{@@@D@@@}}
\com{\stunit}{{@@@()@@@}}
\com{\stint}{{@@@Int@@@}}
\com{\stbool}{{@@@Bool@@@}}
\com{\stintlist}{{@@@IntList@@@}}
\com{\stfun}{{\stone@@@->@@@\sttwo}}
\com{\sthfun}{{\stone@@@#>@@@\sttwo}}
\com{\stprod}{{@@@(@@@\stone@@@,@@@\ldots@@@,@@@\stn@@@)@@@}}
\com{\steprod}{{@@@(@@@\steone@@@,@@@\ldots@@@,@@@\sten@@@)@@@}}
\com{\stsum}{{\scone\ \stone\ @@@+@@@\ \ldots\ @@@+@@@\ \scn\ \stn}}
\com{\stdefn}{{\saone\ @@@=@@@\ \stone@@@;@@@\ \ldots@@@;@@@\ \san\ @@@=@@@\ \stn}}

\com{\sstt}{{\subtypespace\makebox{\subsubtypesize\tt T}}}

% Proj and SProj

\com{\ProjT}{{\it Proj}_{\stt}}
\com{\ProjU}{{\it Proj}_{\stu}}
\com{\SProjT}{{\it SProj}_{\stt}}
\com{\SProjU}{{\it SProj}_{\stu}}
\com{\SProjE}{{\it SProj}_{\ste}}
\com{\SAbsT}{{\it SAbs}_{\stt}}
\com{\FProjT}{{\it FProj}_{\stt}}
\com{\FProjU}{{\it FProj}_{\stu}}
\com{\FProjE}{{\it FProj}_{\ste}}
\com{\FAbsT}{{\it FAbs}_{\stt}}
\com{\FAbsFT}{{\it FAbsF_{\stt}}}
\com{\FTranT}{{\it FTran}_{\stt}}

\com{\Eprod}{@(@\Eone@,@\ldots@,@\En@)@}
\com{\Ei}{{@E@_i}}
\com{\Eone}{{\rm@E@_1}}
\com{\En}{{@E@_n}}

% misc symbols

\com{\oin}{{\rm1}\leq{}i\leq{}n}
\com{\ojn}{{\rm1}\leq{}j\leq{}n}
\com{\lubio}{\lub_{i\geq\rm0}}
\com{\glbio}{\glb_{i\geq\rm0}}
\com{\Lubio}{\Lub_{i\geq\rm0}}
\com{\Glbio}{\Glb_{i\geq\rm0}}
\com{\mkint}[1]{{\it{}mkint}^{#1}_{{\cal{}N}\br{@@@n@@@}}}
\rcom{\u}[1]{\acute{#1}}
\rcom{\d}[1]{\grave{#1}}
\com{\ud}[1]{\widehat{#1}}

% semantic functions

\com{\Epe}[2]{\mbox{${\cal E}_p^{#1}\lb#2\rb$}}
\com{\cEp}[1]{\mbox{${\cal E}_p^{#1}$}}
\com{\Rpe}[2]{\mbox{${\cal E}_p^{#1}\lb#2\rb$}}
\com{\cRp}[1]{\mbox{${\cal R}_p^{#1}$}}
\com{\Tpe}[2]{\mbox{${\cal E}_p^{#1}\lb#2\rb$}}
\com{\cTp}[1]{\mbox{${\cal T}_p^{#1}$}}

\com{\Ee}[2]{\mbox{${\cal E}^{#1}\lb#2\rb$}}
\com{\Edefne}[2]{\mbox{${\cal E}^{#1}_{{\it{}defns}}\lb#2\rb$}}
\com{\cE}[1]{\mbox{$\cal E$}^{#1}}
\com{\cEdefn}[1]{\mbox{${\cal E}^{#1}_{\it{}defns}$}}

\com{\cX}[2]  {\mbox{$\cal #1$}^{\it{}#2}}
\com{\Xe}[3]  {\mbox{${\cal #1}^{\it{}#2}\br{#3}$}}
\com{\cXd}[2] {\mbox{${\cal #1}^{\it{}#2}_{\dd}$}}
\com{\Xde}[3] {\mbox{${\cal #1}^{\it{}#2}_{\dd}\br{#3}$}}
\com{\cXs}[3] {\mbox{${\cal #1}^{\it{}#2}_{#3}$}}
\com{\Xse}[4] {\mbox{${\cal #1}^{\it{}#2}_{#3}\br{#4}$}}
\com{\cXds}[3] {\mbox{${\cal #1}^{\it{}#2}_{\dd,#3}$}}
\com{\Xdse}[4] {\mbox{${\cal #1}^{\it{}#2}_{#3}\br{\dd,#4}$}}
\com{\cXdefn}[2] {\mbox{${\cal #1}^{\it{}#2}_{\it{}defns}$}}
\com{\Xdefne}[3] {\mbox{${\cal #1}^{\it{}#2}_{\it{}defns}\br{#3}$}}

\com{\cR}[1]{\cX{R}{#1}}
\rcom{\Re}[2]{\Xe{R}{#1}{#2}}
\com{\cRd}[1]{\cXd{R}{#1}}
\com{\Rde}[2]{\Xde{R}{#1}{#2}}
\com{\cRs}[2]{\cXs{R}{#1}{#2}}
\com{\Rse}[3]{\Xse{R}{#1}{#2}{#3}}
\com{\cRds}[2]{\cXds{R}{#1}{#2}}
\com{\Rdse}[3]{\Xdse{R}{#1}{#2}{#3}}
\com{\cRdefn}[1]{\cXdefn{R}{#1}}
\com{\Rdefne}[2]{\Xdefne{R}{#1}{#2}}

\com{\cV}[1]{\cX{V}{#1}}
\com{\Ve}[2]{\Xe{V}{#1}{#2}}
\com{\cVd}[1]{\cXd{V}{#1}}
\com{\Vde}[2]{\Xde{V}{#1}{#2}}
\com{\cVs}[2]{\cXs{V}{#1}{#2}}
\com{\Vse}[3]{\Xse{V}{#1}{#2}{#3}}
\com{\cVds}[2]{\cXds{V}{#1}{#2}}
\com{\Vdse}[3]{\Xdse{V}{#1}{#2}{#3}}
\com{\cVdefn}[1]{\cXdefn{V}{#1}}
\com{\Vdefne}[2]{\Xdefne{V}{#1}{#2}}

\com{\cB}[1]{\cX{B}{#1}}
\com{\Be}[2]{\Xe{B}{#1}{#2}}
\com{\cBd}[1]{\cXd{B}{#1}}
\com{\Bde}[2]{\Xde{B}{#1}{#2}}
\com{\cBs}[2]{\cXs{B}{#1}{#2}}
\com{\Bse}[3]{\Xse{B}{#1}{#2}{#3}}
\com{\cBds}[2]{\cXds{B}{#1}{#2}}
\com{\Bdse}[3]{\Xdse{B}{#1}{#2}{#3}}
\com{\cBdefn}[1]{\cXdefn{B}{#1}}
\com{\Bdefne}[2]{\Xdefne{B}{#1}{#2}}

\com{\Te}[2]{\mbox{${\cal T}^{#1}\lb#2\rb$}}
\com{\Tde}[2]{\mbox{${\cal T}^{#1}_{\ddagger}\lb#2\rb$}}
\com{\Tdefne}[2]{\mbox{${\cal T}^{#1}_{{\it{}defns}}\lb#2\rb$}}
\com{\cT}[1]{\mbox{$\cal T$}^{#1}}
\com{\cTd}[1]{\mbox{${\cal T}^{#1}_{\ddagger}$}}
\com{\cTdefn}[1]{\mbox{${\cal T}^{#1}_{\it{}defns}$}}

\com{\Ie}[2]{\mbox{${\cal P}^{#1}\lb#2\rb$}}
\com{\Ide}[2]{\mbox{${\cal P}^{#1}_{\ddagger}\lb#2\rb$}}
\com{\Idefne}[2]{\mbox{${\cal P}^{#1}_{{\it{}defns}}\lb#2\rb$}}
\com{\cI}[1]{\mbox{$\cal P$}^{#1}}
\com{\cId}[1]{\mbox{${\cal P}^{#1}_{\ddagger}$}}
\com{\cIdefn}[1]{\mbox{${\cal P}^{#1}_{\it{}defns}$}}

\com{\cTt}[1]{\cXt{T}{#1}{\cal F}}
\com{\Tte}[2]{\Xte{T}{#1}{\cal F}{#2}}

\com{\cDATA}[1]{\mbox{$\it{}DATA$}^{#1}}
\com{\cDATAd}[1]{\mbox{$\it{}DATA$}_{\dd}^{#1}}
\com{\cDATAdefn}[1]{\mbox{${\it{}DATA}_{\it{}defns}^{#1}$}}
\com{\DATAe}[2]{\mbox{${\it{}DATA}^{#1}\lb#2\rb$}}
\com{\DATAdefne}[2]{\mbox{${\it{}DATA}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cFUN}[1]{\mbox{$\it{}FUN$}^{#1}}
\com{\cFUNd}[1]{\mbox{${\it{}FUN}_{\dd}^{#1}$}}
\com{\cFUNdefn}[1]{\mbox{${\it{}FUN}_{\it{}defns}^{#1}$}}
\com{\FUNe}[2]{\mbox{${\it{}FUN}^{#1}\lb#2\rb$}}
\com{\FUNde}[2]{\mbox{${\it{}FUN}_{\dd}^{#1}\lb#2\rb$}}
\com{\FUNdefne}[2]{\mbox{${\it{}FUN}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cCHOOSE}[1]{\mbox{$\it{}CHOOSE$}^{#1}}
\com{\cCHOOSEd}[1]{\mbox{${\it{}CHOOSE}_{\dd}^{#1}$}}
\com{\cCHOOSEdefn}[1]{\mbox{${\it{}CHOOSE}_{\it{}defns}^{#1}$}}
\com{\CHOOSEe}[2]{\mbox{${\it{}CHOOSE}^{#1}\lb#2\rb$}}
\com{\CHOOSEde}[2]{\mbox{${\it{}CHOOSE}_{\dd}^{#1}\lb#2\rb$}}
\com{\CHOOSEdefne}[2]{\mbox{${\it{}CHOOSE}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cSTR}[1]{\mbox{$\it{}STR$}^{#1}}
\com{\cSTRd}[1]{\mbox{${\it{}STR}_{\dd}^{#1}$}}
\com{\cSTRdefn}[1]{\mbox{${\it{}STR}_{\it{}defns}^{#1}$}}
\com{\STRe}[2]{\mbox{${\it{}STR}^{#1}\lb#2\rb$}}
\com{\STRde}[2]{\mbox{${\it{}STR}_{\dd}^{#1}\lb#2\rb$}}
\com{\STRdefne}[2]{\mbox{${\it{}STR}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cUNFAC}[1]{\mbox{$\it{}UNFAC$}^{#1}}
\com{\cUNFACd}[1]{\mbox{${\it{}UNFAC}_{\dd}^{#1}$}}
\com{\cUNFACdefn}[1]{\mbox{${\it{}UNFAC}_{\it{}defns}^{#1}$}}
\com{\UNFACe}[2]{\mbox{${\it{}UNFAC}^{#1}\lb#2\rb$}}
\com{\UNFACde}[2]{\mbox{${\it{}UNFAC}_{\dd}^{#1}\lb#2\rb$}}
\com{\UNFACdefne}[2]{\mbox{${\it{}UNFAC}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cFAC}[1]{\mbox{$\it{}FAC$}^{#1}}
\com{\cFACd}[1]{\mbox{${\it{}FAC}_{\dd}^{#1}$}}
\com{\cFACdefn}[1]{\mbox{${\it{}FAC}_{\it{}defns}^{#1}$}}
\com{\FACe}[2]{\mbox{${\it{}FAC}^{#1}\lb#2\rb$}}
\com{\FACde}[2]{\mbox{${\it{}FAC}_{\dd}^{#1}\lb#2\rb$}}
\com{\FACdefne}[2]{\mbox{${\it{}FAC}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cPR}[1]{\mbox{$\it{}PR$}^{#1}}
\com{\cPRd}[1]{\mbox{${\it{}PR_\dd}^{#1}$}}
\com{\PRe}[2]{\mbox{${\it{}PR}^{#1}\lb#2\rb$}}
\com{\PRde}[2]{\mbox{${\it{}PR_\dd}^{#1}\lb#2\rb$}}

\com{\cEM}[1]{\mbox{$\it{}EM$}^{#1}}
\com{\cEMd}[1]{\mbox{${\it{}EM_\dd}^{#1}$}}
\com{\EMe}[2]{\mbox{${\it{}EM}^{#1}\lb#2\rb$}}
\com{\EMde}[2]{\mbox{${\it{}EM_\dd}^{#1}\lb#2\rb$}}

\com{\cBOT}[1]{\mbox{$\it{}BOT$}^{#1}}
\com{\cBOTd}[1]{\mbox{${\it{}BOT}_{\dd}^{#1}$}}
\com{\cBOTdefn}[1]{\mbox{${\it{}BOT}_{\it{}defns}^{#1}$}}
\com{\BOTe}[2]{\mbox{${\it{}BOT}^{#1}\lb#2\rb$}}
\com{\BOTde}[2]{\mbox{${\it{}BOT}_{\dd}^{#1}\lb#2\rb$}}
\com{\BOTdefne}[2]{\mbox{${\it{}BOT}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\ABSe}[1]{\mbox{${\it{}ABS}\lb#1\rb$}}

\com{\cCON}[1]{\mbox{$\it{}CON$}^{#1}}
\com{\cCONd}[1]{\mbox{${\it{}CON}_{\dd}^{#1}$}}
\com{\cCONdefn}[1]{\mbox{${\it{}CON}_{\it{}defns}^{#1}$}}
\com{\CONe}[2]{\mbox{${\it{}CON}^{#1}\lb#2\rb$}}
\com{\CONde}[2]{\mbox{${\it{}CON}_{\dd}^{#1}\lb#2\rb$}}
\com{\CONdefne}[2]{\mbox{${\it{}CON}_{\it{}defns}^{#1}\lb#2\rb$}}

\com{\cD}[1]{\mbox{$\cal D$}^{#1}}
\com{\cDd}[1]{\mbox{${\cal D}^{#1}_{\it{}defns}$}}
\com{\De}[2]{\mbox{${\cal D}^{#1}\lb#2\rb$}}
\com{\Dde}[2]{\mbox{${\cal D}^{#1}_{\it{}defns}\lb#2\rb$}}

\com{\cF}[1]{\cX{F}{#1}}
\com{\cFd}[1]{\cXd{F}{#1}}
\com{\Fe}[2]{\Xe{F}{#1}{#2}}
\com{\Fde}[2]{\Xde{F}{#1}{#2}}
\com{\Fdefne}[2]{\Xdefne{F}{#1}{#2}}
\com{\cFdefn}[1]{\cXdefn{F}{#1}}

\com{\sq}[1]{[\,#1\,]}
\com{\lb}{[\![\,}                       % semantic brackets
\com{\rb}{\,]\!]}
\com{\br}[1]{\lb#1\rb}

% NEWCOMMANDS

\com{\chapt}[1]{
	\chapter{{#1}}
%	\markboth{
%	  {\bf\underline{\arabic{page}\hspace*{\fill}
%	   Chapter \thechapter. \ {#1}}}}{}
  	\thispagestyle{plain}}
\com{\sect}[1]{%\clearpage
	\section{{#1}}
%	\markright{{\bf\underline{\thesection.
%		    \ {#1}\hspace*{\fill}*\arabic{page}}}}
  	}
\com{\subsect}[1]{\subsection{{#1}}}
\com{\preface}{
	\chapter*{Preface}
	\addcontentsline{toc}{chapter}{Preface}
	\thispagestyle{plain} }
\rcom{\ni}{} %{\noindent}
\com{\be}{\begin{enumerate}}
%\com{\ee}{\end{enumerate}}

% math formatting
\com{\ba}{\begin{array}}
\com{\ea}{\end{array}}
\com{\mc}{\multicolumn}
\com{\zero}[1]{\lefteqn{\mbox{$#1$}}}

\com{\sa}{\[\renewcommand{\arraystretch}{1.15}\setlength{\arraycolsep}{0in}\begin{array}{lllllllllll}}
\com{\fa}{\end{array}\]\posv}

\com{\posv}{\vspace{2pt}}

\newenv{eqns}{\[\renewcommand{\arraystretch}{1.15}\setlength{\arraycolsep}{0.5ex}\begin{array}{lllllll}}{\end{array}\]\posv}
\com{\reason}[1]{\mbox{\rm \ \ [#1]}}
\com{\beqs}{\begin{eqns}}
\com{\eeqs}{\end{eqns}}

\newenv{eqqns}{\[\renewcommand{\arraystretch}{1.15}\setlength{\arraycolsep}{0in}\begin{array}{lllllll}}{\end{array}\]\posv}
\com{\sreason}[1]{\mbox{\rm\ \ \ \ [#1]}}
\com{\beqqs}{\begin{eqqns}}
\com{\eeqqs}{\end{eqqns}}

%       Theorems, Lemmata and Corollaries are numbered consecutively,
%       Definitions and Proofs are unnumbered. All take an additional
%       argument (often {}) to appear in the title.

%\newtheorem{theorem}{Theorem}[chapter]
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{Fact}[theorem]{Fact}
%\newtheorem{definition}{Definition}
%\newtheorem{proof}{Proof}
%\newtheorem{skproof}{Sketch Proof}
%\newtheorem{conjecture}{Conjecture}
%\newtheorem{exemplar}{Example.}
%\newtheorem{acase}{Case}
%\rcom{\thedefinition}{\mbox{}}
%\rcom{\theproof}{\mbox{}}
%\rcom{\theskproof}{\mbox{}}
%\rcom{\theexemplar}{\mbox{}}
%\rcom{\theacase}{\mbox{}}

\newenv{thm}[1]{\begin{theorem} \mbox{\bf #1}\\ \rm}{\end{theorem}}
\newenv{lem}[1]{\begin{lemma}   \mbox{\bf #1}\\ \rm}{\end{lemma}}
\newenv{cor}[1]{\begin{corollary}   \mbox{\bf #1}\\ \rm}{\end{corollary}}
\newenv{npcor}[1]{\begin{corollary}   \mbox{\bf #1}\\ \rm}{\ \ \mbox{$\Box$}\end{corollary}}
\newenv{fact}[1]{\begin{Fact}   \mbox{\bf #1}\\ \rm}{\end{Fact}}
\newenv{prop}[1]{\begin{proposition}   \mbox{\bf #1}\\ \rm}{\end{proposition}}
\newenv{npprop}[1]{\begin{proposition}   \mbox{\bf #1}\\ \rm}{\ \ \mbox{$\Box$}\end{proposition}}
\newenv{defn}[1]{\begin{definition} \mbox{\bf #1}\\ \rm}{\end{definition}}
\newenv{pf}[1]{\begin{proof}\mbox{\bf#1}\\ \rm}{\ \ \mbox{$\Box$}\end{proof}}
\newenv{sketch}[1]{\begin{skproof}    \mbox{\bf #1}\\ \rm}{\ \ \mbox{$\Box$}\end{skproof}}
%\newenv{example}{\begin{exemplar}\rm}{\end{exemplar}}
%\newenv{case}[1]{\begin{acase}   \mbox{\bf #1}\\ \rm}{\end{acase}}
\newenv{conj}[1]{\begin{conjecture}\mbox{\bf #1}\\ \rm}{\ \ \mbox{$\Box$}\end{conjecture}}

\com{\Case}[1]{\noindent\mbox{{\bf Case:} \ #1}\\ }
\com{\Casep}[1]{\noindent\mbox{{\bf Case:} \ #1}}

% common symbols

\com{\la}{\leftarrow}
\com{\ra}{\rightarrow}
\com{\implies}{\Rightarrow}
\com{\x}{\times}
\rcom{\approx}{\sqsubseteq}
\rcom{\o}{\circ}                % compose
\com{\entails}{\vdash}
\rcom{\vec}{}
\com{\LUB}{\bigsqcup}
\com{\bij}{\stackrel{bij}{\rightarrow}}

\com{\range}[1]{(#1)}
\com{\curly}[1]{\{#1\}}
\com{\sqb}[1]{\lb {#1} \rb}
\rcom{\angle}[1]{\langle{#1}\rangle}
\com{\dangle}[1]{\dlangle{#1}\drangle}
\com{\dlangle}{\langle\!\langle}
\com{\drangle}{\rangle\!\rangle}

\com{\inv}[1]{#1^{-\rm1}}
\com{\shrp}{SHARP}
\com{\flt}{FLAT}

\com{\E}[3]{\mbox{${\cal E}_{#1}\lb#2\rb$}_{#3}}
\com{\M}[1]{\mbox{${\cal M}\lb#1\rb$}}
%\com{\D}[1]{\mbox{${\cal D}\lb#1\rb$}}
\com{\Ep}[2]{\E{\phi}{#1}{#2}}
\com{\Epn}[2]{\E{\phi_n}{#1}{#2}}
\com{\Epr}[1]{\Ep{#1}{\rho}}

\com{\Dpn}[1]{\D{#1}{\phi_n}}

\com{\Ea}[3]{\mbox{${\cal E}^\#_{#1}\lb#2\rb$}_{#3}}
\com{\Ma}[1]{\mbox{${\cal M}^\#\lb#1\rb$}}
\com{\Da}[1]{\mbox{${\cal D}^\#\lb#1\rb$}}
\com{\Pa}[3]{\mbox{${\cal P}^\#_{#1}\lb#2\rb$}_{#3}}

\com{\Eap}[2]{\Ea{\phia}{#1}{#2}}
\com{\Eapn}[2]{\Ea{\phia_n}{#1}{#2}}
\com{\Eapr}[1]{\Eap{#1}{\rhoa}}

\com{\Pap}[2]{\Pa{\phia}{#1}{#2}}
\com{\Papr}[1]{\Pa{\phia}{#1}{\rhoa}}
\com{\Fapd}[1]{\mbox{$\cF^\#_{\phia}\lb#1\rb$}_{\Delta}}

\com{\rhoa}{\rho^\#}
\com{\phia}{\phi^\#}

\com{\envxv}{\curly{x\mapsto{}v}}
\com{\envxg}{\curly{x\mapsto\gamma}}
\com{\envxgv}{\curly{x\mapsto\gamma\/v}}
\com{\envxsv}{\curly{x\mapsto\sigma_f\/v}}
\com{\envxvy}{\curly{x\mapsto{}v,x_k\mapsto{}y_k}}

\com{\lamv}{\lambda{}v}

%\com{\V}{\mbox{$V$}}
\com{\Vi}{\mbox{$V_i$}}
\com{\Vis}{\mbox{$\{V_i\}$}}
\com{\Vb}{\mbox{$V_{\!\bot}$}}
\com{\fis}{\mbox{$\{f_i\}$}}
\com{\fIis}{\mbox{$\{f_i\}$}}

\com{\sigmap}{\sigma_{p}}
\com{\sigmapi}{\sigma_{p_i}}
\com{\sigmapI}{\sigma_{p_1}}
\com{\sigmapII}{\sigma_{p_2}}
\com{\deltap}{\delta_{p}}
\com{\pip}{\pi_{p}}

\com{\phD}{\varphi_D}
\com{\psD}{\psi_D}

\com{\ve}{\vec{@e@}}
\com{\vx}{\vec{@x@}}
\com{\vy}{\vec{@y@}}
\com{\vT}{\vec{@T@}}
\com{\vF}{\vec{F}}
\com{\vv}{\vec{v}}

\com{\X}{X}

\com{\fxge}{\mbox{@f x = @$\cdots$@(g e)@$\cdots$}}
\com{\C}{\mbox{$Dom$}}
\com{\Cn}{\mbox{$\C^n$}}
\com{\Cm}{\mbox{$\C^m$}}
\com{\Cs}{\mbox{$\C_s$}}
\com{\Csn}{\mbox{$\C_s^n$}}
\com{\Csm}{\mbox{$\C_s^m$}}

\com{\void}{\mbox{\bf 1}}
\com{\cat}{\mbox{$\cal C$}}

\com{\MIX}{\mbox{$mix$}}
\com{\pel}{\mbox{PEL}}

\com{\rep}[1]{\overline{\rule{0ex}{1.5ex}{#1}}}
\com{\Rep}{\rep}
\com{\subrep}[1]{\stackrel{\rule[-.5mm]{1.5ex}{.2mm}}{#1}}
\com{\mix}[2]{mix\ \rep{{#1}}\ \rep{{#2}}}
\com{\safe}[3]{#2 \o #1 = #2 \o #1 \o #3}
\com{\safes}[3]{#2 \o #1 \approx #1 \o #3}
\com{\division}{\mbox{$(\sigma,\delta,\pi)$}}
\com{\control}{\mbox{$\{(V_i,\ f_i:V_i \ra V)\}$}}
\com{\controlI}{\mbox{$\{(V_i,\ f_i:V \ra V)\}$}}
\com{\set}[2]{\{{#1}\ |\ {#2} \}}

\com{\all}[2]{\forall {#1}\ .\ {#2}}
%\com{\case}[3]{\begin{array}[t]{lcl}\zero{case\ {#1}\ in}\\
%			&:\\
%			\ \ {#2} &\Rightarrow& {#3}\\
%			&:
%		\end{array}}
\com{\casec}[3]{\begin{array}[c]{lcl}\zero{case\ {#1}\ in}\\
			&:\\
			\ \ {#2} &\Rightarrow& {#3}\\
			&:
		\end{array}}

%\com{\proj}{\mbox{\bf proj}}

\com{\forder}{\mbox{$1^{\rm st}$-order}}
\com{\adhoc}{\mbox{\it ad hoc}}
\com{\finite}{\mbox{\rm finite}}
\com{\consistent}[1]{\mbox{$\{#1\}$ \rm consistent}}

\com{\tr}[1]{\mbox{{\rm #1}}}
\com{\tb}[1]{\mbox{{\bf #1}}}

\com{\axm}[1]{\mbox{$\begin{array}{c}{#1}\end{array}$}}
\com{\infer}[2]{\mbox{$\begin{array}{c}\underline{\raisebox{1.2ex}{$#1$}}\\{#2}\end{array}$}}
\com{\infero}[2]{\mbox{$\begin{array}{c}\raisebox{.4ex}{$#1$}\\\overline{\rule{0ex}{2.2ex}{#2}}\end{array}$}}

\com{\xv}{\mbox{\footnotesize$\{ \vec{x}\mapsto\vec{v} \}$}}
\com{\xFv}{\mbox{\footnotesize$\{ \vec{x}\mapsto\vec{F\ \gamma\ v} \}$}}
\com{\xiv}{\mbox{\footnotesize$\{ \vec{x}\mapsto\vec{v},\ \vec{x_i}\mapsto\vec{v_i} \}$}}
\com{\xiFv}{\mbox{\footnotesize$\{ \vec{x}\mapsto\vec{F\ \gamma\ v},\ 
			\vec{x_i}\mapsto\vec{v_i} \}$}}
\com{\xiFHv}{\mbox{\footnotesize$\{ \vec{x}\mapsto\vec{F\ \gamma\ v},\ 
			\vec{x_i}\mapsto\vec{H_i\ \gamma\ v_i} \}$}}
\com{\xF}{\vx::\vF}

\com{\etal}{\mbox{\it et al}}
\com{\bslash}[1]{{#1}/x}

\sloppy

\begin{abstract}
  {\it First-order\/} projection-based strictness analysis (properly,
  \emph{demand analysis}) was first introduced by Wadler and Hughes in
  1987~\cite{WH87}.  This development was restricted to first-order,
  monomorphic languages and provided no proof of soundness.  Subsequent
  developments extended the theory to other analyses, approaches to higher
  order, and polymorphism.  Yet other developments were based on
  generalization of projections, specifically partial equivalence relations
  and partial projections, the latter notably simplifying a difficulty with
  projection-based strictness dating back to Wadler and Hughes.  However, to
  date no development has combined the desirable contributions of all of these
  approaches.  We present the first practical higher-order, polymorphic,
  partial-projection/partial-equivalence-relation based demand analysis,
  together with a proof of satisfaction of the fundamental safety condition
  for such analyses.  Such an analysis can be used both to improve serial code
  generation, as well as enable safe automatic parallelization, for lazy or
  non-strict functional languages.

% has proven genuinely useful in partial evaluation
% \cite{Lau91a}.  There have been three notable generalisations
% of projection-based analysis to higher order.  The first lacked a
% formal basis \cite{Mog89}; the second used structures strictly more
% general than projections, namely {\it partial equivalence relations\/}
% (PERs) \cite{HS91}; the third involved a complex construction that
% gave rise to impractically large abstract domains \cite{Dav93b}.  This
% paper presents a technique free of these shortcomings: it is simple,
% entirely projection-based, satisfies a formal correctness condition,
% and gives rise to reasonably small abstract domains.  Though the
% technique is cast in terms of projections, there is also an
% interpretation in terms of PERs.  The principal limitation of the
% technique is the restriction to {\it monomorphic\/} typing.
\end{abstract}

%\keywords{Binding-time analysis, abstract interpretation, projections, PERs.}

\section{Introduction and Background}

We take as given that demand analysis is essential for good serial code
generation, and for identifying where parallelism may be safely introduced.
Numerous such analysis techniques have been proposed and implemented; we
greatly narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the technique has
been proven to satisfy.

Analysis techniques can usually be identified as being based on either
a non-standard denotational semantics or a non-standard typing.
Examples in the latter category include those of Gomard \cite{Go92},
Jensen \cite{Jen92}, the Nielsons \cite{NN88}, Schmidt \cite{Sch88},
and Henglein and Mossin \cite{HM94}.  Our focus is on those techniques
based on non-standard interpretation, in particular, those using
projections or partial equivalence relations (PERs) as the basic
abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$
function $\BOT$, the least projection, which specifies that values are
entirely dynamic; and projections $\FST$ and $\SND$ on product
domains, defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\hspace{1.0in} \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs,
respectively.  The nominal goal of analysis is, given function $f$
denoted by some programming-language expression, and projection $\D$
encoding the staticness of the argument of $f$, to determine $\G$
satisfying the {\it safety condition\/} $\G\o\f\we\f\o\D$.  For
example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition, in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.

There have been three notable attempts to generalise Launchbury's
techniques to higher order.  The first was Mogensen's generalisation
of the polymorphic technique \cite{Mog89}.  Though successfully
implemented, there is no formal statement of what it means for the
analysis to be correct; even if such a statement were made, proving
correctness would likely be difficult because of the highly
intensional nature of the analysis: the non-standard values associated
with expressions are strongly dependent on their syntactic structure,
and projections are encoded symbolically as {\it abstract closures\/},
with approximation performed algebraically `on-the-fly' according to
time and space considerations.  Nonetheless, the experiment provided
evidence for the {\it practicality\/} of the projection-based approach
at higher order.

The second generalisation was Hunt and Sands', of the monomorphic
technique to higher order \cite{HS91}.  Their observation was that a
projection, regarded as a set of domain-range pairs, is an {\it
equivalence relation\/}: given $\G$, values $u$ and $v$ are in the
same equivalence class if $\G~u=\G~v$, and the canonical elements of
the equivalence classes are the set of fixed points (range) of $\G$.
Hunt showed that the safety condition $\G\o\f\we\f\o\D$ holds iff $f$
is related to itself by $\D\fto\G$ where $\fto$ is the standard
operation on binary relations, so 
"" 
	(\G\o\f\we\f\o\D)\
	\Leftrightarrow\ 
	(\forall u,v\ .\ \D\ u=\D\ v\ \Rightarrow\ \G\ (f\ u)=\G\ (f\ v))\ .  
""
Then, for example, $(\BOT\fto\ID)~(f,f)$ asserts
that $f$ maps dynamic arguments to static results.  In general
$\D\fto\G$ is not an equivalence relation, but it is always a {\it
partial\/} equivalence relation: it is symmetric and transitive but
not necessarily reflexive.  (Partialness is crucial: if $\D\fto\G$
were in general reflexive then $(\D\fto\G)(f,f)$ would hold for all
$f$, hence such assertions would tell nothing.)  Unlike projections
regarded as relations, PERs are closed under $\fto$; the result, as
Hunt and Sands show, is that `scaling up' to higher-order analysis is
reasonably straightforward.  One disadvantage of their method is that
PER spaces are considerably larger than the projection spaces on the
same domains, and it is not clear which PERs to choose for (finite)
abstract domains.  Here they borrowed heavily from the projection
world, using standard abstract projection domains at ground types.
Further, its practicality has not been demonstrated by implementation,
and, because of the unfamiliar territory, a promising route to a
polymorphic generalisation is obscure.

The third generalisation was ours, to an entirely projection-based,
monomorphic, higher-order technique \cite{Dav93b}.  One observation
motivating the approach is that there is no meaningful abstraction of
values to projections, only of functions $f$ to projection
transformers $\T$ (functions from projections to projections)
satisfying $(\T~\D)\o\f\we\f\o\D$ for all $\D$.  To make this
abstraction possible a semantics intermediate between the standard and
analysis semantics was introduced.  Moving from the standard to
intermediate semantics involved a translation of each ground type @T@
to a function type @E@\ @->@\ @T@ (for a fixed type @E@), the values
of which, being functions, could then be abstracted.  The result,
while proven correct with respect to a formal safety condition, is
probably not practicable because of the growth in the sizes of
(usefully rich) abstract domains induced by the type translation.

This paper presents a technique far simpler than our
previous one.  No intermediate semantics is required, and the
correctness condition and proof are much simpler.  Because the
translation of ground types to function types is avoided the abstract
domains are much smaller. Though entirely projection-based, we show
that there is a reading of the results in terms of PERs, intimating a
close relationship with the PER-based technique.

\section{Language and Standard Semantics}

The source language is a simple, strongly typed, monomorphic,
non-strict functional language.  The grammar for the language of types
and type definitions is given in Figure~\ref{fig:types}.
\fig{tb}{Types and type definitions.}{fig:types}{
""
@T@	&::=&	\ta			&\reason{Type Name}
	&|&	\tint			&\reason{Integer}
	&|&	\tprod			&\reason{Product, $n\geq0$}
	&|&	\tsum			&\reason{Sum, $n\geq1$} 
	&|&	\tfun			&\reason{Function} \espace 
@D@	&::=&	\tdefn			&\reason{Type Definitions}
""
}
Nullary product corresponds to the so-called {\it unit\/} type.  A
unary product @(T)@ will always have the same semantics as @T@.  The
types used in the examples are defined as follows.
@@@
     FunList =   nil () 
               + cons (Int -> Int, FunList) ,

     FunTree =   leaf (Int -> Int) 
	       + branch (FunTree, FunTree) .
@@@

The grammar for expressions is given in Figure~\ref{fig:exprs}.
Addition for integers is provided as typical of operations on flat data
types in this setting.  A unary tuple @(e)@ will always have the same
semantics as @e@.  The (monomorphic) typing of expressions is entirely
standard and is omitted.
\fig{tb}{Expressions.}{fig:exprs}{
""
@e@	&::=&	\evar				&\reason{Variable} 
	&|&	\enum				&\reason{Numeral}
	&|&	\esum				&\reason{Integer addition}
	&|&	\eapp				&\reason{Function application}
	&|&	\etup				&\reason{Tuple construction}
	&|&	@let@\ @(@\xone@,@\ldots@,@\xn@)@\ @=@\ \enot\ @in@\ \eone			&\reason{Tuple decomposition}
	&|&	\econ				&\reason{Sum construction}
	&|&	@case@\ \enot\ @of @\{\ci\ \xxi\ @->@\ \ei\}	&\reason{Sum decomposition}
	&|&	\elamt				&\reason{Lambda abstraction}
	&|&	\eapp				&\reason{Function application}
	&|&	\efix				&\reason{Fixed point}
""
}

\subsection{Expression semantics}
Since two different expression semantics will be given, following
Abramsky \cite{Abr90} we define a semantics $\cE{}$ parameterised by a
set of {\it defining constants}.  The semantics $\cE{}$ is defined in
Figure~\ref{fig:gensem}; the defining constants are {\it plus},
$\itt{sel}_i$, {\it tuple}, $\itt{inc}_i$, $\itt{outc}_i$, {\it
choose}, {\it mkfun}, {\it apply}, and {\it fix}.  The two instances
of $\cE{}$ are distinguished by a superscript: $\cE{\Ss}$ for the
standard semantics and $\cE{\Ps}$ for the non-standard semantics.  The
corresponding type semantics, and defining constants, have the same 
superscripts.

It is useful to regard the free-variable environment of each
expression as having some tuple type $\tprod$, and environment lookup
as indexing (as in a categorical semantics, or De~Bruijn indexing);
variables are indexed implicitly or explicitly by their index in the
free variable environment.  Then for both versions of the semantics
and all expressions @e@ of type @T@ with environment of type $\tprod$,
""
	\Ee{}{@e@}\ \in\ \Te{}{\tprod}~\fto~\Te{}{@T@}\ .
""
Noting that $\R\br{\xxi}$ is short for $\itt{sel}_i~\R$, environment
update $\R[\xxi\mapsto{}v]$ is defined by
"""
	tuple (sel_1 \R, \ldots, sel_{i-1} \R, v, sel_{i+1} \R, \ldots, sel_n \R) .
"""

\fig{tb}{Parameterised semantics.}{fig:gensem}{
""
	\Ee{}{\evar_i}\ \R = \R\br{@x@_i} = \itt{sel}_i\ \R\espace\espace
%	\Ee{}{\eunit}\ \R = \itt{mkunit}\ \R\espace\espace
%	\Ee{}{\enum}\ \R = \itt{mkint}_i\espace\espace
	\Ee{}{\esum}\ \R = \itt{plus}~(\Ee{}{\eone}\ \R,\ \Ee{}{\etwo}\ \R)\espace\espace
	\Ee{}{\etup}\ \R = \itt{tuple}~(\Ee{}{\eone}\ \R,~\ldots,~\Ee{}{\en}\ \R)\espace\espace
	\Ee{}{\esel}\ \R
	\ \ \ \ \ = \Ee{}{\eone}\ \R[\xxi\mapsto\itt{sel}_i\ (\Ee{}{\enot}\ \R)~|~\oin]\espace\espace
	\Ee{}{\econ}\ \R = \itt{inc}_i\ (\Ee{}{@e@}\ \R)\espace\espace
	\Ee{}{\ecase}\ \R
	\ \ \ \ \  = \itt{choose}~(\ba[t]{l}\Ee{}{\enot}\ \R,
	\Ee{}{\eone}\ \R[\xone\mapsto\itt{outc}_1\ (\Ee{}{\enot}\ \R)],
	\ \ \ ~\vdots
	\Ee{}{\en}\ \R[\xn\mapsto\itt{outc}_n\ (\Ee{}{\enot}\ \R)])\ea\espace\espace
	\Ee{}{\elam}\ \R = \itt{mkfun}\ (\L{}x.\Ee{}{@e@}\ \R[\evar\mapsto{}x])\espace\espace
	\Ee{}{\eapp}\ \R = \itt{apply}\ (\Ee{}{\eone}\ \R)\ (\Ee{}{\etwo}\ \R)\espace\espace
	\Ee{}{\efix}\ \R = (\itt{fix}\o\itt{apply})\ (\Ee{}{@e@}\ \R)
""
}

\subsection{Standard semantics}

The standard $\Ss$ type and expression semantics are defined in
Figure~\ref{fig:ssem}.  
Domain $\plus$ is separated (lifted) sum.  
For more generality, products are unlifted; a unary sum-of-products gives a
more usual lifted product.  
Function types give rise to lifted function
spaces as in Abramsky's lazy lambda calculus \cite{Abr89}.
The important feature of this approach is the modelling of the
fact that functions are, in effect, embedded in the simplest of
lazy data structures, which either have no WHNF---have value $\bot$---or
have a WHNF and contain an applicable function value---have value
$\itt{lift}~f$ for some $f$.  (This can be made even more explicit by
introducing a unary sum type, whence "\L" becomes a constructor \cite{Dav94}.)
Recursive type definitions give rise to recursive domain
specifications which have the usual least-fixed-point solutions.

\fig{tb}{Standard type and expression semantics.}{fig:ssem}{
"""
\Te{\Ss}{\tint} = \Int = \lift{\bf{Z}} \espace\espace
\Te{\Ss}{\tprod} = \Te{\Ss}{\tone}\ \times\ \ldots\ \times\ \Te{\Ss}{\tn}\espace\espace
\Te{\Ss}{\tsum} = \Te{\Ss}{\tone}\ \plus\ \ldots\ \plus\ \Te{\Ss}{\tn}\espace\espace
\Te{\Ss}{\tfun} = \liftp{\Te{\Ss}{\tone} \fto \Te{\Ss}{\ttwo}}

%	\itt{mkunit}^{\Ss}\ \R			 = 	()\espace\espace
%	\itt{mkint}^{\Ss}_n\ \R			 = 	\itt{lift}\ n\espace\espace
	\itt{plus}^{\Ss}\ (x,y) 		 = 	x+y\espace\espace
	\itt{tuple}^{\Ss}\ (x_1,\ldots,x_n)	 = 	(x_1,\ldots,x_n)\espace\espace
	\itt{sel}_i^{\,\Ss}\ (x_1,\ldots,x_n)	 = 	x_i\espace\espace
	\itt{inc}_i^{\Ss} 			 = 	\itt{in}_i\o\itt{lift}\espace\espace
	\itt{outc}_i^{\Ss}\ x			 = 	\itt{drop}\o\itt{out}_i\espace\espace
	\ba{lll}\itt{choose}^{\Ss}\ (\bot,&\ x_1,\ldots,x_n)		 = \bot
	\itt{choose}^{\Ss}\ (\itt{in}_i~v,&\ x_1,\ldots,x_n)	 = x_i\ea\espace\espace
	\itt{mkfun}^{\Ss}		 = 	\itt{lift}\espace\espace
	\itt{apply}^{\Ss}		 = 	\itt{drop}\espace\espace
	\itt{fix}^{\Ss}  		 = 	\itt{lfp} \reason{Least fixed point}
"""
}

\section{Domain Factorisation}

A key observation of our previous work
\cite{Dav93b} is that since there is no concept
of staticness of the body of a lambda expression, there is no point in
having projections on function spaces.  Hence values are factored into
their evaluable or {\it data\/} parts, and their unevaluable but
applicable {\it forward\/} parts.  For example, for expressions of
function type ${@T@\ @->@\ @U@}$, with
$\liftp{T\fto{}U}=\Te{\Ss}{@T@\ @->@\ @U@}$, we distinguish two
degrees of definedness---between values of the form
$\itt{lift}~f$ and $\bot$, corresponding to expressions that do and
do not have a WHNF, respectively.  These two degrees of definedness 
may be encoded by the two-point
domain $\lone=\{\bot, \itt{lift}~()\}$---the {\it data domain\/} corresponding
to type ${@T@\ @->@\ @U@}$.  There are two distinct projections on $\lone$,
namely $\ID$ and $\BOT$: denoted functions are 
either static or dynamic.  Here 
the forward domain is $T\fto{}U$, and the factorisation of $\liftp{T\fto{}U}$
is $\lone\times(T\fto{}U)$.  The factorisation function
${\it{}fac}_{@@@T->U@@@}\in\liftp{T\fto{}U}\fto(\lone\times(T\fto{}U))$
is an embedding defined by
${\it{}fac}_{@@@T->U@@@}~\bot=(\bot,\bot)$, and
${\it{}fac}_{@@@T->U@@@}~(\itt{lift}~f)=(\lunit,~f)$ for all $f$.
Every embedding {\it determines\/} a corresponding projection; 
this follows from the relations $g\o h\we\itt{id}$ and $h\o\g=\itt{id}$
that hold between an embedding $g$ and its corresponding projection $h$.
The importance of factorisation being an embedding is that no information
is lost---values may be recovered by unfactoring.  Here the
projection is given by
${\it{}unfac}_{@@@T->U@@@}~(\bot,f)=\bot$ and
${\it{}unfac}_{@@@T->U@@@}~(\lunit,~f)=\itt{lift}~f$.

Generally, the data domain corresponding to type @T@ is $\De{}{@T@}$,
where $\cD{}$ is defined exactly like $\cT{\Ss}$ except that function
spaces are replaced by the one-point domain $\one$, that is,
$\De{}{\tfun}=\lift{\one}$ for all $\tone$ and $\ttwo$.  
The function $\data$ from values in
$\Te{\Ss}{@T@}$ to their data parts in $\De{}{@T@}$ is a projection:
it is like the identity except that values from function spaces are
mapped into $\one$.  The projection $\data$ is defined in terms of the
structure of @T@ as given in Figure~\ref{fig:fact}.
\fig{tb}{Factorisation and unfactorisation functions.}{fig:fact}{
""
	{\it{}data}_{\stint} = {\it{}id}_{\stint}\ ,\espace
	{\it{}data}_{\stprod} = {\it{}data}_{\stone} \times \ldots \times {\it{}data}_{\stn} ,\espace
	{\it{}data}_{\stsum} = {\it{}data}_{\stone} \plus \ldots \plus {\it{}data}_{\stn} ,\espace
	{\it{}data}_{\stfun} = {\liftp{\Lx.\unit}} .
""

"""
	{\it{}for}_{\stint} = \Lx.\unit\ ,\espace
	{\it{}for}_{\stprod} = {\it{}for}_{\stone}\ \times\ \ldots\ \times\ {\it{}for}_{\stn}\ ,\espace
        \ba{lll}{\it{}for}_{\stsum}\ \bot	&\ = \bot
	{\it{}for}_{\stsum}\ (\itt{in}_i~v) &\ = (\bot,\ldots,\bot,v,\bot_,\ldots,\bot)	&\ \ \reason{$v$ in $\ith$ position}\ ,\ea\espace
	{\it{}for}_{\stfun} = \itt{drop}\ .
"""

"""
	\itt{unfac}_{\stint} (n,()) = n ,\espace
	\itt{unfac}_{\stprod} ((d_1,...,d_n), (f_1,...,f_n)) = (v_1,...,v_n)
	~~~{\rm{}where} 
	~~~~~~v_i=\itt{unfac}_{\sti} (d_i,f_i), \oin ,\espace
	\itt{unfac}_{\stsum} ((d_1,...,d_n), (f_1,...,f_n)) = (v_1,...,v_n)
	~~~{\rm{}where} 
	~~~~~~v_i=\itt{unfac}_{\sti} (d_i,f_i), \oin ,\espace
	\itt{unfac}_{\stfun} (\bot,f) = \bot,
	\itt{unfac}_{\stfun} (\lunit, f) = \itt{lift} f . 
"""
}
The definition of ${\it{}data}_{\stfun}$ uses function lifting, 
defined by $\lf~\bot=\bot$
and $\lf~(\itt{lift}~x)=\itt{lift}~(f~x)$.  Recursive type definitions
give rise to recursive function specifications with exactly one solution.

In the same style as  $\De{}{@T@}$ and $\data$ we define
$\Tde{\Ss}{@T@}$ to give the forward domain for @T@, and $\for$ to be
the function mapping values in $\Te{\Ss}{@T@}$ to their forward parts
in $\Tde{\Ss}{@T@}$, as follows.  Roughly, $\cTd{\Ss}$ is like
$\cT{\Ss}$ with all lifting removed and sum replaced by product, so
""
	\Tde{\Ss}{\tint} = \one ,\espace
	\Tde{\Ss}{\tprod} = \Tde{\Ss}{\tone} \times \ldots \times \Tde{\Ss}{\tn} ,\espace
	\Tde{\Ss}{\tsum} = \Tde{\Ss}{\tone} \times \ldots \times \Tde{\Ss}{\tn} ,\espace
	\Tde{\Ss}{\tfun} = \Te{\Ss}{\tone} \fto \Te{\Ss}{\ttwo} .
""
The mapping from standard values to their forward parts is defined in
Figure~\ref{fig:fact}.
We write $\fac$ for $\Lx.(\data~x,\ \for~x)$.  Then
$\De{}{@T@}\times\Tde{\Ss}{@T@}$ is a factorisation of
$\Te{\Ss}{@T@}$, and
""
	\fac \in \Te{\Ss}{@T@} \fto (\De{}{@T@} \times \Tde{\Ss}{@T@})
""
is an embedding which, as mentioned, determines the corresponding projection
$\unfac$ as defined in Figure~\ref{fig:fact}.

We give some further examples.  The factorisation of
$\Int$ is $\Int\times\one$; more generally, the factorisation of any
domain $D$ corresponding to a type not containing @->@ is just
$D\times\one$.  For @FunList@ the standard domain is
"""
	\mu{}X . \lone+(\liftp{\itt{Int}\fto\itt{Int}}\x{}X) ,
"""
that is, lists of functions from integers to integers. In factored form
the domain is
"""
	(\mu{}X . \lone+(\lone\x{}X)) \x (\mu{}Y . (\itt{Int}\fto\itt{Int})\x{}Y) ,
"""
that is, pairs consisting of a list of values from $\{\bot,\ \lbot\}$, and
an infinite tuple of functions from integers to integers.  Then the 
standard value of the list @cons (\x.x-1, cons (\x.x-2, nil ()))@ is
"""
	\itt{incons} (\itt{lift} (\L{}x.x-1), \itt{incons} (\itt{lift} (\L{}x.x-2), \itt{innil} ())) ,
"""
the data value is
"""
	\itt{incons} (\lunit, \itt{incons} (\lunit, \itt{innil} ())) ,
"""
and the forward value is 
$(\L x.x-1,~(\L x.x-2,~(\bot,~(\bot,~(\bot,~\ldots)))))$.

\section{Projection Semantics}

Standard values $v$ and $v'$ are related by nonstandard value
$(\A,\K)$ when their data parts are related by projection $\A$, and
their forward parts are {\it logically related\/} \cite{Abr90} by
$\K$.  At each type @T@ the relation is
$\Re{}{@T@}~((\A,\K),\cdot,\cdot)$, defined by
""
	{\Re{}{@T@}~((\A,\K),v,v')} =&\mit(\A~d=\A~d')\ \wedge\ \Rde{}{@T@}~(\K,f,f')
		&\mbox{\rm{}where}
		&\ \ \ \ba[t]{lll}(d,f)  &\:= \fac~v
				\mit(d',f')	&\:= \fac~v' ,\ea
""
where $\Rde{}{@T@}$ is defined by
""
	\Rde{}{\tint}~((),(),()) = \itt{True}\ ,\espace
	\Rde{}{\tprod} = \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace
	\Rde{}{\tsum} = \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace
	\Rde{}{\tfun} = \Re{}{\tone}\ \fto\ \Re{}{\ttwo}\ .
""
Here $\x$ and $\fto$ are the standard operations on {\it ternary\/}
relations:  for relations $R$ and $S$ we have
$(R\times{}S)((x,x'),(y,y'),(z,z'))$ iff $R(x,y,z)$ and
$S(x',y',z')$, and $(R\fto{}S)(f,g,h)$ iff for all $x$, $y$, and $z$
such that $R(x,y,z)$ we have $S(f~x,\ g\ y,\ h~z)$.  Here, recursive
type definitions give recursive relation specifications, which have
inclusive least-fixed-point solutions. (A relation is inclusive if,
when it holds for each element of a directed set, and in particular
an ascending chain, it also holds at
the limit.  Such relations are sometimes called {\it admissible\/} or
{\it directed complete}.  Some work is required to show inclusivity of
these recursively-defined relations.) Thus the $\Ps$ type semantics
$\cT{\Ps}$ must be
""
	\Te{\Ps}{@T@} = \proj{\De{}{@T@}}\ \times\ \Tde{\Ps}{@T@}\ ,
""
where $\proj{\De{}{@T@}}$ is the lattice of projections on domain
$\De{}{@T@}$, and $\cTd{\Ps}$ is defined exactly like $\cTd{\Ss}$ with
superscript $\Ps$ everywhere replacing superscript $\Ss$.

\fig{tb}{Projection semantics.}{fig:fsem}{
""
%	\itt{mkunit}^{\Ps}\ \R			 = 	()
%\espace\espace
%	\itt{mkint}^{\Ps}_n\ \R			 = 	\itt{lift}\ n
%\espace\espace
	\itt{plus}^{\Ps}\ ((\A,()),(\B,())) = \left\{\ba{ll}(\ID,()),	&\ \ \ \ \mbox{\rm if $\A=\ID$ and $\B=\ID$}
			(\BOT,()),	&\ \ \ \ \mbox{\rm otherwise}\ea\right.\espace
	\itt{tuple}^{\Ps}\ ((\A_1,\K_1),\ldots,(\A_n,\K_n)) = ((\A_1\x\ldots\x\A_n),(\K_1,\ldots,\K_n))\espace
	\itt{sel}_i^{\,\Ps}\ ((\A_1\x\ldots\x\A_n),(\K_1,\ldots,\K_n))  = (\A_i,\K_i)\espace
	\itt{inc}_i^{\Ps}\ (\A,\K)	 = 	(C_i\ \A,\ (\top,\ldots,\top,\K,\top,\ldots,\top))\espace
	\itt{outc}_i^{\Ps}\ (\A,(\K_1,\ldots,\K_n))	 = (\OUTC_i\ \A,\ \K_i)\espace
	\itt{choose}^{\Ps}\ ((\A,\K),x_1,\ldots,x_n) = \left\{\ba{ll}\bot,		&\ \ \ \ \mbox{\rm if}\ \A\not\se\glb_{\oin} C_i~\BOT
			x_1\glb\ldots\glb{}x_n,	&\ \ \ \ \mbox{\rm otherwise}\ea\right.\espace
	\itt{mkfun}^{\Ps}\ f			 = 	(\ID,f)\espace
	\itt{apply}^{\Ps}\ (\A,f) = \left\{\ba{lll}\bot,	&\ \ \ \ \mbox{\rm if}\ \A=\BOT
						f,	&\ \ \ \ \mbox{\rm if}\ \A=\ID \ea\right.\espace
	\itt{fix}^{\Ps} = 	\itt{gfp} \reason{Greatest fixed point}
""
}

The defining constants for the projection semantics $\cE{\Ps}$ are
given in Figure~\ref{fig:fsem}. The following notation is used for
specifying projections on the data domains.  For sum type $\tsum$ with
data domain $D_1\plus\ldots\plus D_n$, define $C_i~\A$ to be
$\ID+\cdots+\ID+\A+\ID+\cdots+\ID$ where $\A$ is the $i^{th}$ summand,
$\OUTC_i~(\G_1+\cdots+\G_n)$ to be $\G_i$, and $\OUTC_i~\BOT$ to be
$\BOT$.

\ \\ \noindent {\bf The Central Result.} For expression @e@ of type
@T@ and free-variable environment type @E@, the functions
$\Ee{\Ps}{@e@}$, $\Ee{\Ss}{@e@}$, and $\Ee{\Ss}{@e@}$ are logically
related by $\cR{}$, that is
""
	(\Re{}{@E@}\fto\Re{}{@T@})\ (\Ee{\Ps}{@e@},\ \Ee{\Ss}{@e@},\ \Ee{\Ss}{@e@})\ .
""
Further, $\Re{}{@T@}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
$\K$.\ \ %$\Box$

\ \\ \noindent The bulk of proof is omitted; it consists of showing
that the defining constants are similarly related, and a simple
induction on the structure of expressions showing that if the defining
constants are logically related, then so are the semantic functions
$\cE{\Ps}$, $\cE{\Ss}$, and $\cE{\Ss}$.
By way of example we consider the expression form $\esum$ and the
relevant defining constant $\it plus$.  To show that $\it plus^{\Ss}$
and $\it plus^{\Ps}$ are correctly related we need to show that
"""
	((\Re{}{\tint}\x\Re{}{\tint})\fto\Re{}{\tint}) (plus^{\Ps},plus^{\Ss},plus^{\Ss}) .
"""
By the definitions of $\fto$ and $\x$ on ternary relations, this is
equivalent to 
""
	(\Re{}{\tint}\x\Re{}{\tint}) (x,y,z) \limp \Re{}{\tint}\
		\ (plus^{\Ps} x, plus^{\Ss} y, plus^{\Ss} z) ,
""
for all $x$, $y$, and $z$.  In turn, this is equivalent to
"""
	&\ \Re{}{\tint} (x,y,z) \land \Re{}{\tint} (x',y',z')
\limp	&\ \Re{}{\tint} (plus^{\Ps} (x,x'), plus^{\Ss} (y,y'), plus^{\Ss} (z,z'))
"""
for all $x$, $x'$, $y$, $y'$, $z$, and $z'$; equivalently, for
all values $y$, $y'$, $z$, and $z'$ from the integer domain
$\Te{\Ss}{\tint}$ we have
""
	\mit(\ID y=\ID z) \land (\ID y'=\ID z') \limp \mit\ID (y+y')=\ID (z+z') ,
""
which trivially holds, and for $\A_1\neq\ID$ or $\A_2\neq\ID$ that
""
	\mit(\A_1 y=\A_1 z) \land (\A_2 y'=\A_2 z') \limp \mit\BOT (y+y')=\BOT (z+z') ,
""
which is also trivial.  For the inductive case $\esum$ with 
environment type @E@ we need to show that if
""
	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\eone}, \Ee{\Ss}{\eone}, \Ee{\Ss}{\eone}) 
""
and
""
	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\etwo}, \Ee{\Ss}{\etwo}, \Ee{\Ss}{\etwo}) 
""
then
"""
	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\esum},\Ee{\Ss}{\esum},\Ee{\Ss}{\esum}) .
"""
Suppose $\Re{}{@E@}~(\R^{\Ps},\R^{\Ss},\R'^{\Ss})$.  Then by 
the induction hypothesis we have that
$\Re{}{\tint}~(\A_i,v_i,v'_i)$, where 
$(\A_i,())=\Ee{\Ps}{\ei}~\R^{\Ps}$,
$(v_i,())=\Ee{\Ps}{\ei}~\R^{\Ss}$, and
$(v'_i,())=\Ee{\Ps}{\ei}~\R'^{\Ss}$, for $i=1,2$.
Now
"""
	\Re{}{\tint} (\Ee{\Ps}{\esum} \R^{\Ps}, \Ee{\Ss}{\esum}~\R^{\Ss}, \Ee{\Ss}{\esum}~\R'^{\Ss})
"""
iff
"""
	\Re{}{\tint} (&plus^{\Ps} ((\A_1,()),(\A_2,())), 
		      &plus^{\Ss} \mit((v_1,()),(v_2,())), 
		      &plus^{\Ss} \mit((v'_1,()),(v'_2,()))) ,
"""
which holds since $\it plus^{\Ps}$ and $\it plus^{\Ps}$, and their
arguments, are correctly related.

\section{Abstract Domains}

At each type @T@ we require a finite abstraction of $\Te{\Ps}{@T@}$.
This abstract domain is ${\it{}FProj}_{\stt}\times{\it{}FFor}_{\stt}$,
where ${\it{}FProj}_{\stt}$ is a finite abstraction of the lattice
$\proj{\De{}{@T@}}$, and ${\it{}FFor}_{\stt}$ is a finite abstraction
of $\Tde{\Ps}{@T@}$.  The definition of ${\it{}FProj}_{\stt}$ is based
on Launchbury's \cite{Lau91a}.  A projection $\G$ is in
${\it{}FProj}_{\stt}$ if $\G~\mbox{\bf proj}~@T@$ can be inferred from
the rules given in Figure~\ref{fig:infrules}.
%
\fig{tb}{Inference rules for finite projection domains.}{fig:infrules}{
""
\begin{array}{lll}\
\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tint}\ \ \ \ \ \ \ \ \ \ & \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tint}\
\espace\espace
\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tfun} & \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tfun}\end{array}\
\espace\espace
\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tsum}\
\espace\espace
\infero{\G_1\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{(C_1\ \G_1)\glb\cdots\glb(C_n\ \G_n)\ \ \mbox{\bf{}proj}\ \ \tsum}\
\espace\espace
\infero{\G_1\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{\G_1\x\ldots\x\G_n\ \ \mbox{\bf{}proj}\ \ \tprod}\
\espace\espace
\infer{\ba{ll}&\G_1\ \ \mbox{\bf{}proj}\ \ \aone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \an
\vdash & P(\G_1,...,\G_n)\ \ \mbox{\bf{}proj}\ \ \ti@(@\aone@,...,@\an@)@\ea}{\mu(\G_1,\ldots,\G_n).P(\G_1,\ldots,\G_n)\ \ \mbox{\bf{}proj}\ \ \ai}\
\ \ \ \ \ [\mbox{\rm{}where}\ \ai@=@\ti@(@\aone@,...,@\an@)@]
""
}
For recursively-defined types the rules yield only those projections
that are {\it uniform}---that act on each recursive instance of a 
data structure in the same
way.  Thus ${\it{}FProj}_{@@@FunList@@@}$ comprises $\BOT$ and
$\SPINE~\A$ for $\A$ ranging over $\BOT$ and $\ID$, where
""
	\SPINE\ \A = \mu\G.(\itt{NIL}\ \ID)\ \glb\ (\itt{CONS}\ (\A\ \x\ \G))\ ,
""
so $\SPINE\ \BOT$ specifies static spines and dynamic elements, and
$\SPINE\ \ID$ is the identity, thus specifying static spines and
static elements.  More generally, the abstract list
constructor is isomorphic to lifting.  Similarly,
${\it{}FProj}_{@@@FunTree@@@}$
comprises $\BOT$ and $\LBR\ \A$ for $\A$ ranging over $\BOT$ and $\ID$, where
""
	\LBR\ \A = \mu\G.(\itt{LEAF}\ \A)\ \glb\ (\itt{BRANCH}\ (\G\ \x\ \G))\ ,
""
so $\LBR\ \A$ specifies static branches and leaves, and $\A$ of the
values in the leaf nodes.  Again, the abstract tree constructor is
isomorphic to lifting.  This compares favourably with BHA strictness
analysis, for which the corresponding abstract constructors are
typically {\it double\/} lifting \cite{Wad87,Sew93}.

Value $\K$ is in ${\it{}FFor}_{\stt}$ if $\K~\bfabsf~@T@$ can be
inferred from the following.
First, there is only one forward value at type $\tint$.
""
	\axm{() {} \bfabsf {} \tint} .
""
For products and sums,
""
\infer{\K_1 {} \bfabsf {} \tone { } \cdots { } \K_n {} \bfabsf {} \tn}\
{(\K_1,\ldots,\K_n) {} \bfabsf {} \tprod} ,
""
%\espace\espace
""
\infero{\K_1 {} \bfabsf {} \tone { } \cdots { } \K_n {} \bfabsf {} \tn}\
{(\K_1,\ldots,\K_n) { } \bfabsf {} \tsum} .
""
Function spaces consist of a set of step functions closed under lub.
""
\infero{\ba{ll}\A_1\in{\it{}FProj}_{\stone} { { } } \K_1 \bfabsf \tone
\A_2\in{\it{}FProj}_{\sttwo} { { } } \K_2 \bfabsf \ttwo\ea}\
{\it{}step ((\A_1,\K_1),(\A_2,\K_2)) {} \bfabsf {} \tfunp} ,
""
where
""
	step (v_1,v_2) x = v_2,		&{\rm if} v_1\we{}x 
	step (v_1,v_2) x = \bot,	&{\rm otherwise} ,
""
and
""
\infer{\K_1 {} \bfabsf {} \tfunp { { } } \K_2 {} \bfabsf {} \tfunp}\
{(\K_1\lub\K_2) {} \bfabsf {} \tfunp} .
""
This gives the full space of monotonic functions on the abstract
domains.

For recursively-defined types, roughly speaking, we choose those
forward values that represent each component of the same type by the
same value.  Given type definitions $\tdefn$, which will be written
$@A@_i@ = T@_i@(A@_1@,...,A@_n@)@,\ \oin$, if from assumptions
$\K_i~\bfabsf~\ai$ for $\oin$ it may be deduced that
$P_i(\K_1,\ldots,\K_n)~~\bfabsf~~\ti(\aone\ldots\an)$ for $\oin$, then
""
	\mu(\K_1,...,\K_n).(P_1(\K_1,...,\K_n),...,P_n(\K_1,...,\K_n))
""
is a tuple $(\K_1,...,\K_n)$ such that $\K_i\ \bfabsf\ @A@_i$
for $\oin$.
For all @T@ the set ${\it{}FProj}_{\stt}\times{\it{}FFor}_{\stt}$ is a 
finite lattice containing the top and bottom elements of $\Te{\Ps}{@T@}$.

For @T@ not containing @->@ the domain $\Tde{\Ps}{@T@}$ is isomorphic
to $\one$, so ${\it{}FFor}_{\stint}$ is $\one$.  For @Int@\ @->@\ @Int@
we have
$\Tde{\Ps}{@Int@\ @->@\ @Int@}=(\proj{\De{}{\tint}}\x\one)\fto(\proj{\De{}{\tint}}\x\one)$,
so ${\it{}FFor}_{@@@Int->Int@@@}$ is
$({\it{}FProj}_{\stint}\x\one)\fto({\it{}FProj}_{\stint}\x\one)$.  A
data structure of recursive type @A=T(A)@ may be thought of as some
(possibly infinite) number of elements of @T(())@.  For example, the
value $\itt{cons}\ (f,\ \itt{cons}\ (g,\ \itt{nil}\ ()))$ in the
standard domain for @FunList@ decomposes into $\itt{cons}\ (f,())$,
$\itt{cons}\ (g,())$, and $\itt{nil}\ ()$.  The (implicit) abstraction
function maps such a data structure to the greatest lower bound of
these elements, giving a safe abstraction of the nonstandard values of
each element.  Thus
${\it{}FFor}_{@@@A=T(A)@@@}={\it{}FFor}_{@@@T(())@@@}$, so
${\it{}FFor}_{@@@FunList@@@}\iso{\it{}FFor}_{@@@FunTree@@@}\iso{\it{}FFor}_{@@@Int->Int@@@}$.

\section{Examples of Analysis}

For all closed expressions @e@ the abstract value
$\Ee{\Ps}{@e@}~[\,]$ of @e@ is of the form $(\ID,\K)$, showing
that closed expressions are always entirely static.  For expressions of
function type the abstract forward value $\K$ is a function from
abstract arguments of @e@ to abstract results.

First we consider functions on lists.
Let expression @length@ denote the usual length function:
@@@
     fix (\length . \xs . case xs of
                            nil u  -> 0
                            cons p -> let (z,zs) = p in
                                        1 + length zs)
@@@
The abstract forward value of @length@ is of the form
$\L(\A,\K).(\T~\A,())$, where $\T$ maps $\SPINE~\BOT$ and $\SPINE~\ID$
to $\ID$, and $\BOT$ to $\BOT$.  This reveals that the result of
@length@ is independent of the values of list elements, and gives
a static result when the argument has a static spine.  It is worth
pointing out that the result of @length@ is static---determined---for
infinite-length arguments with static spines; staticness and definedness
are independent properties.

Let @append@ stand for the expression denoting the usual function for
appending two lists:
@@@
     fix (\append . \xs . \ys . case xs of
                                  nil u  -> ys
                                  cons p -> let (z,zs) = p in
		                              cons (z, 
						    append zs ys))
@@@
The abstract value of @append@ is
""
(\ID, \L(\A_{@@@xs@@@},\K_{@@@xs@@@}).
(\ID, \L(\A_{@@@ys@@@},\K_{@@@ys@@@}).
(\A_{@@@xs@@@}\glb\A_{@@@ys@@@}, \K_{@@@xs@@@}\glb\K_{@@@ys@@@}))) .
""
This reveals that partial applications of @append@ are static up to
WHNF, and the abstract value of the result is the greatest lower bound
of the two arguments.  In general, the abstract value of a closed
expression of the form $@\x@_1@.\x@_2\ldots@.e@$ will reveal that all
partial applications of the expression are static up to WHNF.

Let @reverse1@ stand for the expression denoting the naive reverse
function:
@@@
     fix (\reverse1 . \xs . case xs of
                              nil u  -> nil ()
                              cons p -> let (z,zs) = p in
                                          append (reverse zs) 
			                         (cons (z, nil ())))
@@@
The abstract forward value of @reverse1@ is the identity, so the
abstraction of the elements of a list doesn't change by reversing the
list.

Let @compose@ stand for @\f.\g.\x.f (g x)@.  The abstract value of @compose@ is
""
(\ID, \L(\A_{@@@f@@@},\K_{@@@f@@@}).
(\ID, \L(\A_{@@@g@@@},\K_{@@@g@@@}).
(\ID, \left\{\ba{lll}\bot, &\mbox{\rm if}\ \A_{@@@f@@@}=\BOT\mbox{\rm or} \A_{@@@g@@@}=\BOT
	\K_{@@@f@@@}\o\K_{@@@g@@@},&\mbox{\rm otherwise}\ea\right\}))) .
""
Thus, the
result of the application of the composition of two functions is
dynamic, and maps all values to dynamic values, if either function is
dynamic; otherwise, the result is given by the application of the
composition of the abstract forward values to the abstract argument.

Let @listcomp@ stand for the expression denoting the function
that composes lists of functions:
@@@
     fix (\listcomp . \fs . case fs of
                              nil u  -> \x.x
                              cons p -> let (g,gs) = p in 
                                          compose g (listcomp gs))
@@@
The abstract value of @listcomp@ is
""
	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
	(\ID,\ \left\{\ba{lll}\
\bot,		&\mbox{\rm if}\ \A_{@@@fs@@@}\neq\SPINE\ \ID
\glb_{i\geq0}(\K_{@@@fs@@@})^i,	&\mbox{\rm otherwise}\ea\right\})) .
""
Since the abstract values of lists contain no information about the
length of the lists of which they are abstractions, the abstract value
of the composition of list elements is the glb of the composition over
all lengths.

Let @flatten@ stand for the expression denoting the function that
flattens trees into lists:
@@@
     fix (\flatten . \t . case t of
	                    leaf l   -> cons (l, nil ())
	                    branch p -> let (t1,t2) = p in
			                  append (flatten t1) 
					         (flatten t2))
@@@
The abstract value of @flatten@ is
""
	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
	(\T\ \A,\ \left\{\ba{lll}\
\bot,		&\mbox{\rm if}\ \T\ \A_{@@@fs@@@}\neq\LBR\ \ID
\K_{@@@fs@@@},	&\mbox{\rm otherwise}\ea\right\})) ,
""
where $\T$ maps $\BOT$ to $\BOT$, and $\LBR~\A$ to $\SPINE~\A$ for $\A$
ranging over $\ID$ and $\BOT$.

The function denoted by @compose listcomp flatten@ that composes
trees of functions has abstract value
""
	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
	(\ID,\ \left\{\ba{lll}\
\bot,		&\mbox{\rm if}\ \A_{@@@fs@@@}\neq\LBR\ \ID
\glb_{i\geq0}(\K_{@@@fs@@@})^i,	&\mbox{\rm otherwise}\ea\right\}))\ .
""
Similarly to the case for lists, the abstract values of trees contain
no information about the structure of the trees of which they are abstractions,
so the abstract value of the composition of the values of the leaves
is the glb of the composition over all tree structures.

\section{More on Abstract Domains}

The sizes of the abstract domains and the representations of the
abstract values can be considerably optimised.  The nonstandard
semantics of application---embodied by $\it apply^{\Ps}$---guarantees
that the abstract values $(\BOT,f)$ and $(\BOT,f')$ from
$\Te{\Ps}{\tfun}$, for all $f$ and $f'$, are effectively the same:
$\it apply^{\Ps}~(\BOT,f)=\bot$ for all $f$.  A practical analyser
would take advantage of this fact, identifying $(\BOT,f)$ over all
$f$.  More generally, for function types embedded within data
structures, e.g.\ $@(@\tint@,@\tfun@)@$, abstract values
$((\A,()),(\BOT,f))$ would be identified over all $f$.

In the following we restrict attention to {\it denotable\/} values: at
each type @T@ those values that can be expressed as $\Ee{}{@e@}~[\,]$
for some @e@.  For every value $v\in\Te{\Ss}{@T@}$ there is a best
abstraction---a greatest value $(\A,\K)\in\Te{\Ps}{@T@}$ such that
$\Re{}{@T@}((\A,\K),v,v)$.  For types $\tone$ and $\ttwo$ not
containing @->@ we have
$\Tde{\Ps}{\tfun}\iso(\proj{\De{}{\tone}}\x\one)\fto(\proj{\De{}{\ttwo}}\x\one)$,
and if $\K\in\Tde{\Ps}{@T@}$ is greatest such that
$\Rde{}{\tfun}~(\K,f,f)$ for denotable $f$, then $\K$ maps $(\ID,())$
to $(\ID,())$ and distributes over $\glb$.  The subset of functions
from $\Tde{\Ps}{\tfun}$ that map $(\ID,())$ to $(\ID,())$ and
distribute over $\glb$ forms a complete lattice, hence attention may
be restricted to this subset.  Not only does this reduce the number of
abstract functions, it also reduces their representation: such
functions are determined by their behaviour on the unique $\glb$-{\it
basis\/} of the lattice $\proj{\De{}{\tone}}$---the set $B$ of values
such that every element of $\proj{\De{}{\tone}}$ is the glb of some
(possibly empty) subset of $B$, and no element of $B$ can be expressed
as the glb of some subset not containing that element.  This
optimisation can be generalised to higher order: $\Tde{\Ps}{\tfun}$
may be restricted to functions that map $\top$ to $\top$ and
distribute over $\glb$, for all function types $\tfun$.  (These
results follow from those shown in earlier work \cite{Dav94}.)

\section{Related Work}

Consel \cite{Con90} describes a binding-time analysis for higher-order
untyped languages.  As in our analysis abstract values have two parts,
the first describing the static/dynamic properties of values, and the
second describing how (for function types) abstract arguments are
mapped to abstract results; there appears to be an implicit domain
factorisation similar to ours, based on implicit type information
collected by analysis semantics.  In this respect there are many
superficial similarities between the two techniques.  No formal
relation to the standard semantics is given, making a formal
comparison with our technique problematic.

Hughes \cite{Hug88} proposes a general approach to promoting related
analysis techniques to higher order; instantiation to projection-based
analysis yields a technique much more closely related to that of our
previous work \cite{Dav93b} than to the one given here: transformation
of his equations to the same general form as ours reveals that
translation of ground types to function types (as previously mentioned)
is implicit in the technique.

\section{Conclusion}

We have successfully generalised Launchbury's polymorphic
partial-projection-based demand analysis to higher-order, using abstract
domains smaller than those typically used in BHA strictness analysis.

%The next step would be to generalise to handle Hindley-Milner
%polymorphism.  This has been done with good results at first order for
%binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
%at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
%use of these theories will give a reasonably straightforward
%generalisation to polymorphism.  In addition to making the analysis
%more widely applicable, this should also greatly reduce the run-time
%cost of analysis, as it does for BHA analysis.

%Though not developed here, using the same approach it is possible to
%give a strictness analysis technique, again closely related to Hunt's
%PER-based strictness analysis technique \cite{Hun91}.

\begin{thebibliography}{123456}

\bibitem[Abr89]{Abr89}
S. Abramsky. 
``The lazy lambda calculus.'' 
{\it Research Topics in Functional Programming.}
David Turner, ed., Addison-Wesley 1989.

\bibitem[Abr90]{Abr90}
S. Abramsky. 
``Abstract interpretation, logical relations and Kan extensions.''
{\it Journal of Logic and Computation}, 1, 1990.

\bibitem[Bar91]{Bar91}
G. Baraki.
``A note on abstract interpretation of polymorphic functions.''
In \cite{Hug91}.

\bibitem[Bar93]{Bar93}
G. Baraki.
{\it Abstract Interpretation of Polymorphic Higher-Order Functions.}
Ph.D. thesis, Research report FP-1993-7, 
Department of Computing Science, University of Glasgow.

\bibitem[BEJ88]{BEJ88}
D. Bjorner, A.P. Ershov, and N.D. Jones, eds.
{\it Partial Evaluation and Mixed Computation,
Proceedings IFIP TC2 Workshop, Gammel Avern{\ae}s}, Denmark, October 1987.
North-Holland, 1988.

\bibitem[Con90]{Con90}
C. Consel. 
``Binding Time Analysis for Higher Order Untyped Functional Languages.''
Proceedings of the 1990 ACM Conference on LISP and 
Functional Programming, pp264-272.

\bibitem[Dav93]{Dav93b}
K. Davis.
``Higher-order Binding-time Analysis.'' 
{\it Proceedings of the 1993 ACM on Partial Evaluation and
Semantics-Based Program Manipulation (PEPM '93)},
ACM Press, 1993.

\bibitem[Dav94]{Dav94}
K. Davis.
{\it Projection-based Program Analysis.}
Ph.D. thesis, Computing Science Department,
The University of Glasgow, 1994.

\bibitem[Go92]{Go92}
C.K. Gomard.
``A self-applicable partial evaluator for the lambda calculus:  Correctness
and pragmatics.''
ACM TOPLAS, Vol 14, No.\ 2, April 1992.

\bibitem[HM94]{HM94}
F. Henglein and C. Mossin.
``Polymorphic binding-time analysis.''
European Symposium on Programming (ESOP '94).

\bibitem[Hug88]{Hug88}
J. Hughes.
Backwards analysis of functional programs.
In \cite{BEJ88}.

\bibitem[Hug91]{Hug91}
J. Hughes, ed.
{\it Proceedings of the 1991 Conference on Functional Programming Languages
and Computer Architecture (FPCA '91)}, Cambridge, Sept 1991. LNCS 523,
Springer Verlag, 1991.

\bibitem[Hun91]{Hun91}
S. Hunt.
``PERs generalise projections for strictness analysis (extended abstract).''
{\it Proceedings of the 1990 Glasgow Workshop
on Functional Programming}.
Simon L. Peyton Jones {\it et al.}, eds.
Springer Workshops in Computing.
Springer-Verlag, 1991.

\bibitem[HS91]{HS91}
S. Hunt and D. Sands.
``Binding time analysis:  a new PERspective.''
{\it ACM Symposium on Partial Evaluation
and Semantics-Based Program Manipulation},
SIGPLAN Notices Vol.\ 26, No.\ 9, 1991.

\bibitem[Jen92]{Jen92}
T. Jensen.
{\it Abstract Interpretation in Logical Form}.
Ph.D. thesis, Report 93/11, Department of Computer Science,
University of Copenhagen, 1992.

\bibitem[Jon88]{Jon88}
N.D. Jones.
``Automatic program specialization: A re-examination from
basic principles.'' In \cite{BEJ88}.

\bibitem[Lau88]{Lau88}
J. Launchbury.
``Projections for specialisation.''
In \cite{BEJ88}.

\bibitem[Lau91a]{Lau91a}
J. Launchbury. 
{\it Projection Factorisations in Partial Evaluation.} PhD
Thesis, Glasgow University, Nov 89. Distinguished Dissertation in
Computer Science, Vol 1, CUP, 1991.

\bibitem[Mog89]{Mog89}
T. Mogensen.
``Binding-time analysis for polymorphically typed higher order languages.''
{\it International Joint Conference on Theory and Practice of Software
Development}. LNCS 352. Springer-Verlag 1989.

\bibitem[NN88]{NN88}
H.R. Nielson and F. Nielson.
``Automatic binding-time analysis for a typed $\L$-calculus.''
{\it Science of Computer Programming 10},
North Holland, 1988.  Also in POPL '88.

\bibitem[Sch88]{Sch88}
D.A. Schmidt.
``Static properties of partial reduction.''
In \cite{BEJ88}.

\bibitem[Sew93]{Sew93}
J. Seward.
``Polymorphic strictness analysis using frontiers.''
{\it Proceedings of the 1993 ACM on Partial Evaluation and
Semantics-Based Program Manipulation (PEPM '93)},
ACM Press, 1993.

\bibitem[Wad87]{Wad87}
P. Wadler.
``Strictness analysis on non-flat domains by abstract interpretation over finite domains. '' 
S. Abramsky, C. Hankin, eds.
{\it Abstract Interpretation of Declarative Languages.}
Ellis-Horwood, 1987.

\end{thebibliography}

\end{document}

### 2017-02-22 pubs/demand_analysis/paper.tex
demand analysis manuscript

\newcount\isrr          % isrr=0 submitted version, isrr=1 dep res rep isrr=3 final PEPM '94
\def\theisrr{\the\isrr}
\isrr=3

\documentstyle[fleqn]{fpca93}
\newcommand{\progsize}{\normalsize}
\newcommand{\typescriptstyle}{\scriptscriptstyle}
\newcommand{\subtypesize}{\scriptsize}
%\newcommand{\sm}{$\mbox{\tt \ \ \ \ \ }\begin{array}{llllllll}}
%\newcommand{\fm}{\end{array}$}
\newcommand{\subtypespace}{\hspace{0.0pt}}
\newcommand{\subtypesubstyle}{\scriptstyle}             % scriptscriptstyle/scriptstyle

\newcommand{\com}{\newcommand}
\com{\rcom}{\renewcommand}

\input{../phd/commands}
\input{../phd/special}

% -------- A figure with a line at top and bottom, a caption and a label
% -------- Syntax: \fig{figoptions}{caption}{label}{ ..... }
\com{\fig}[4]{
\begin{figure}[#1]
\begin{center}
\begin{minipage}{\columnwidth}
\rule[1ex]{\textwidth}{1pt}
%\vspace{1ex}
#4
\caption{#2 \label{#3}}
\vspace{1ex}
\rule[1ex]{\textwidth}{1pt}
\end{minipage}
\end{center}
\end{figure}
}

\com{\backup}{\hspace{-0.12cm}}
\rcom{\espace}{\vspace{4pt}}
\com{\itt}[1]{\mbox{\it #1}}
%\com{\Ss}{\sf{S}}
\com{\Ps}{\sf{P}}

\com{\OUTC}{\itt{OUTC}}
\com{\SEL}{\itt{SEL}}
\com{\SPINE}{\itt{SPINE}}
\com{\LBR}{\itt{LBR}}

\sloppy

\pagestyle{empty}

\begin{document}

\thispagestyle{empty}

\title{PERs from Projections for Binding-time Analysis}

\author{Kei Davis\\Computing Science Department\\
        University of Glasgow\\
        Glasgow G12 8QQ, UK\\
        kei@dcs.glasgow.ac.uk.}

\date{}
\maketitle

\begin{abstract}

{\it First-order\/} projection-based binding-time analysis has proven
genuinely useful in partial evaluation \cite{Lau91a,Lau91c}.  There
have been three notable generalisations of projection-based analysis to
higher order.  The first lacked a formal basis \cite{Mog89}; the second
used structures strictly more general that projections, namely {\it
partial equivalence relations\/} (PERs) \cite{HS91}; the third involved
a complex construction that gave rise to impractically large abstract
domains \cite{Dav93b}.  This paper presents a technique free of
these shortcomings:  it is simple, entirely projection-based,
satisfies a formal correctness condition, and gives rise to reasonably
small abstract domains.  Though the technique is cast in terms of
projections, there is also an interpretation in terms of PERs.
The principal limitation of the technique is the restriction to
{\it monomorphic\/} typing.

\end{abstract}

\section{Introduction and Background}

We take as given that binding-time analysis is essential for good
partial evaluation, and we do not address the issue of annotating
programs according to the results of analysis.  Numerous binding-time
analysis techniques have been proposed and implemented; we greatly
narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the
technique has been proven to satisfy.

Analysis techniques can usually be identified as being based on either a
non-standard denotational semantics or a non-standard typing.  Examples
in the latter category include those of Gomard \cite{Go92}, Jensen
\cite{Jen92}, the Nielsons \cite{NN88}, Schmidt \cite{Sch88}, and
Henglein and Mossin \cite{HM94}.  Our
focus is on those techniques based on non-standard interpretation, in
particular, those using projections or partial equivalence relations
(PERs) as the basic abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$ function
$\BOT$, the least projection, which specifies that values are entirely
dynamic; and projections $\FST$ and $\SND$ on product domains,
defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\ \ \ \ \ \ \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs, 
respectively.  The nominal goal of analysis is,
given function $f$ denoted by some programming-language expression, and
projection $\D$ encoding the staticness of the argument of $f$, to
determine $\G$ satisfying the {\it safety condition\/}
$\G\o\f\we\f\o\D$.  For example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition satisfies, and in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.  His work culminated in
the first strongly-typed partial evaluator \cite{Lau91c}---strong
evidence for the value of the projection-based approach.

There have been three notable attempts to generalise Launchbury's
techniques to higher order.  The first was Mogensen's generalisation of
the polymorphic technique \cite{Mog89}.  Though successfully
implemented, there is no formal statement of what it means for the
analysis to be correct; even if such a statement were made, proving
correctness would likely be difficult because of the highly intensional
nature of the analysis:  the non-standard values associated with
expressions are strongly dependent on their syntactic structure, and
projections are encoded symbolically as {\it abstract closures\/}, with
approximation performed algebraically `on-the-fly' according to time
and space considerations.  Nonetheless, the experiment provided
evidence for the {\it practicality\/} of the projection-based approach
at higher order.

The second generalisation was Hunt and Sands', of the monomorphic
technique to higher order \cite{HS91}.  Their observation was that a
projection, regarded as a set of domain-range pairs, is an {\it
equivalence relation\/}:  given $\G$, values $u$ and $v$ are in the
same equivalence class if $\G~u=\G~v$, and the canonical elements of
the equivalence classes are the set of fixed points (range) of $\G$.
Hunt showed that the safety condition $\G\o\f\we\f\o\D$ holds iff $f$
is related to itself by $\D\fto\G$ where $\fto$ is the standard
operation on binary relations, so
\beqs
\it &\it (\G\o\f\we\f\o\D)\\
\it \Leftrightarrow&\it (\forall\ u,v\ .\ \D\ u=\D\ v\ \Rightarrow\ \G\ (f\ u)=\G\ (f\ v))\ .
\eeqs
Then, for example, $(\BOT\fto\ID)~(f,f)$ asserts that $f$ maps dynamic
arguments to static results.  In general $\D\fto\G$ is not an
equivalence relation, but it is always a {\it partial\/} equivalence
relation: it is symmetric and transitive but not necessarily
reflexive.  Unlike projections regarded as relations, PERs are closed
under $\fto$; the result, as Hunt and Sands show, is that `scaling up'
to higher-order analysis is reasonably straightforward.  One
disadvantage of their method is that PER spaces are considerably larger
than the projection spaces on the same domains, and it is not clear
which PERs to choose for (finite) abstract domains.  Here they borrowed
heavily from the projection world, using standard abstract projection
domains at ground types.  Further, its practicality has not been
demonstrated by implementation, and, because of the unfamiliar
territory, a promising route to a polymorphic generalisation is obscure.

The third generalisation was ours, to an entirely projection-based,
monomorphic, higher-order technique \cite{Dav93b}.  One observation
motivating the approach is that there is no meaningful abstraction of
values to projections, only of functions $f$ to projection transformers
$\T$ (functions from projections to projections) satisfying
$(\T~\D)\o\f\we\f\o\D$ for all $\D$.  To make this abstraction possible
a semantics intermediate between the standard and analysis semantics
was introduced.  Moving from the standard to intermediate semantics
involved a translation of each ground type \mbox{\tt T} to a function type
\mbox{\tt E}\ \mbox{\tt ->}\ \mbox{\tt T} (for a fixed type \mbox{\tt E}), the values of which, being
functions, could then be abstracted.  The result, while proven correct
with respect to a formal safety condition, is probably not practicable
because of the growth in the sizes of (usefully rich) abstract domains
induced by the type translation.

This paper presents a technique far simpler technique than our previous
one.  No intermediate semantics is required, and the correctness
condition and proof are much simpler.  Because the translation of
ground types to function types is avoided the abstract domains are much
smaller. Though entirely projection-based, we show that there is a
reading of the results in terms of PERs, intimating a close
relationship with the PER-based technique.

\section{Language and Standard Semantics}
The source language is a simple, strongly typed, monomorphic,
non-strict functional language.  The grammar for the language
of types and type definitions is given in Figure~\ref{fig:types}.
\fig{tb}{Types and type definitions.}{fig:types}{
\beqs
\it \makebox{\tt T}&\it ::=&\it \ta&\it \reason{Type\ Name}\\
\it &\it |&\it \tint&\it \reason{Integer}\\
\it &\it |&\it \tprod&\it \reason{Product,\ $n\geq{\rm0}$}\\
\it &\it |&\it \tsum&\it \reason{Sum,\ $n\geq{\rm1}$}\\
\it &\it |&\it \tfun&\it \reason{Function}\ \espace\\
\it \makebox{\tt D}&\it ::=&\it \tdefn&\it \reason{Type\ Definitions}
\eeqs
}
Nullary product corresponds to the so-called {\it unit\/} type.  A unary
product \mbox{\tt (T)} will always have the same semantics as \mbox{\tt T}.
The types used in the examples are defined as follows.
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ FunList\ =\ \ \ nil\ ()\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ +\ cons\ (Int\ ->\ Int,\ FunList)\ ,\ }\vspace{-0.0em}\\
\makebox{\progsize\tt }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ FunTree\ =\ \ \ leaf\ (Int\ ->\ Int)\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ +\ branch\ (FunTree,\ FunTree)\ .}\vspace{-0.0em}
\end{flushleft}

The grammar for expressions is given in Figure~\ref{fig:exprs}.
Addition for integers is provided as typical of operations on flat data
types in this setting.  A unary tuple \mbox{\tt (e)} will always have the same
semantics as \mbox{\tt e}.  The (monomorphic) typing of expressions is entirely
standard and is omitted.

\fig{tb}{Expressions.}{fig:exprs}{
\beqs
\it \makebox{\tt e}&\it ::=&\it \evar&\it \reason{Variable}\\
\it &\it |&\it \enum&\it \reason{Numeral}\\
\it &\it |&\it \esum&\it \reason{Integer\ addition}\\
\it &\it |&\it \eapp&\it \reason{Function\ application}\\
\it &\it |&\it \etup&\it \reason{Tuple\ construction}\\
\it &\it |&\it \makebox{\tt let}\ \makebox{\tt (}\xone\makebox{\tt ,}\ldots\makebox{\tt ,}\xn\makebox{\tt )}\ \makebox{\tt =}\ \enot\\
\it &\it &\it \makebox{\tt \ \ in}\ \eone&\it \reason{Tuple\ decomposition}\\
\it &\it |&\it \econ&\it \reason{Sum\ construction}\\
\it &\it |&\it \makebox{\tt case}\ \enot\ \makebox{\tt of}\\
\it &\it &\it \makebox{\tt \ \ }\{\ci\ \xxi\ \makebox{\tt ->}\ \ei\}&\it \reason{Sum\ decomposition}\\
\it &\it |&\it \elamt&\it \reason{Lambda\ abstraction}\\
\it &\it |&\it \eapp&\it \reason{Function\ application}\\
\it &\it |&\it \efix&\it \reason{Fixed\ point}
\eeqs
}

\subsection{Expression semantics}

Since two different expression semantics will be given, following
Abramsky \cite{Abr90} we define a semantics $\cE{}$ parameterised by a
set of {\it defining constants}.  The semantics $\cE{}$ is defined in
Figure~\ref{fig:gensem}; the defining constants are {\it
plus}, $\itt{sel}_i$, {\it tuple}, $\itt{inc}_i$, $\itt{outc}_i$, {\it
choose}, {\it mkfun}, {\it apply}, and {\it fix}.  The two instances of
$\cE{}$ are distinguished by a superscript:  $\cE{\Ss}$ for the
standard semantics and $\cE{\Ps}$ for the non-standard semantics.  The
corresponding type semantics have the same superscripts, as do the
defining constants.

It is useful to regard the free-variable environment of each expression
as having some tuple type $\tprod$, and environment lookup as indexing
(as in a categorical semantics, or De~Bruijn indexing); variables are
indexed implicitly or explicitly by their index in the free variable
environment.  Then for both versions of the semantics and all
expressions \mbox{\tt e} of type \mbox{\tt T} with environment of type $\tprod$,
\beqs
\it \Ee{}{\makebox{\tt e}}\ \in\ \Te{}{\tprod}~\fto~\Te{}{\makebox{\tt T}}\ .
\eeqs
Noting that $\R\br{\xxi}$ is short for $\itt{sel}_i~\R$, environment
update $\R[\xxi\mapsto{}v]$ is defined by
\beqqs
\it tuple\ (&\it sel_{\rm1}\ \R,\ \ldots,\ sel_{i-{\rm1}}\ \R,\\
\it &\it v,\\
\it &\it sel_{i+{\rm1}}\ \R,\ \ldots,\ sel_n\ \R)\ .
\eeqqs
It is convenient to regard \mbox{\tt Int} as being defined as the infinite sum
\beqs
\it \tint\makebox{\tt \ =\ }\ldots\makebox{\tt \ +\ n}_{-{\rm1}}\makebox{\tt \ ()\ +\ n}_{{\rm0}}\makebox{\tt \ ()\ +\ n}_{{\rm1}}\makebox{\tt \ ()\ +\ }\ldots\ ,
\eeqs
where in practice we write $\mbox{\tt n}_i$ as short for $\mbox{\tt n}_i~\mbox{\tt ()}$.

\fig{tb}{Parameterised semantics.}{fig:gensem}{
\beqs
\it \Ee{}{\evar_i}\ \R\ =\ \R\br{\makebox{\tt x}_i}\ =\ \itt{sel}_i\ \R\espace\espace\\
\it %\ \Ee{}{\eunit}\ \R\ =\ \itt{mkunit}\ \R\espace\espace\\
\it %\ \Ee{}{\enum}\ \R\ =\ \itt{mkint}_i\espace\espace\\
\it \Ee{}{\esum}\ \R\ =\ \itt{plus}~(\Ee{}{\eone}\ \R,\ \Ee{}{\etwo}\ \R)\espace\espace\\
\it \Ee{}{\etup}\ \R\\
\it \ \ \ \ \ =\ \itt{tuple}~(\Ee{}{\eone}\ \R,~\ldots,~\Ee{}{\en}\ \R)\espace\espace\\
\it \Ee{}{\esel}\ \R\\
\it \ \ \ \ \ =\ \Ee{}{\eone}\ \R[\xxi\mapsto\itt{sel}_i\ (\Ee{}{\enot}\ \R)~|~\oin]\espace\espace\\
\it \Ee{}{\econ}\ \R\ =\ \itt{inc}_i\ (\Ee{}{\makebox{\tt e}}\ \R)\espace\espace\\
\it \Ee{}{\ecase}\ \R\\
\it \ \ \ \ \ \ =\ \itt{choose}~(\ba[t]{l}\Ee{}{\enot}\ \R,\\
\it \Ee{}{\eone}\ \R[\xone\mapsto\itt{outc}_{\rm1}\ (\Ee{}{\enot}\ \R)],\\
\it \ \ \ ~\vdots\\
\it \Ee{}{\en}\ \R[\xn\mapsto\itt{outc}_n\ (\Ee{}{\enot}\ \R)])\ea\espace\espace\\
\it \Ee{}{\elam}\ \R\ =\ \itt{mkfun}\ (\L{}x.\Ee{}{\makebox{\tt e}}\ \R[\evar\mapsto{}x])\espace\espace\\
\it \Ee{}{\eapp}\ \R\ =\ \itt{apply}\ (\Ee{}{\eone}\ \R)\ (\Ee{}{\etwo}\ \R)\espace\espace\\
\it \Ee{}{\efix}\ \R\ =\ (\itt{fix}\o\itt{apply})\ (\Ee{}{\makebox{\tt e}}\ \R)
\eeqs
}

\subsection{Standard semantics}

The standard $\Ss$ type and expression semantics are defined in
Figure~\ref{fig:ssem}.  Function types give rise to lifted function
spaces as in Abramsky's lazy lambda calculus \cite{Abr89}, and the
semantics distinguishes those expressions of function type that have
WHNF (have value $\itt{lift}~f$ for some $f$) and those that do not (have
value $\bot$).  Products are unlifted; a unary sum-of-products gives a
lifted product.  Domain $\plus$ is separated sum.  Recursive type
definitions give rise to recursive domain specifications which have the
usual least-fixed-point solutions.

\fig{tb}{Standard type and expression semantics.}{fig:ssem}{
\beqs
\it \Te{\Ss}{\tint}\ =\ \Int\ =\ \lift{\bf{Z}}\ \espace\espace\\
\it \Te{\Ss}{\tprod}\ =\ \Te{\Ss}{\tone}\ \times\ \ldots\ \times\ \Te{\Ss}{\tn}\espace\espace\\
\it \Te{\Ss}{\tsum}\\
\it \ \ \ \ \ =\ \Te{\Ss}{\tone}\ \plus\ \ldots\ \plus\ \Te{\Ss}{\tn}\espace\espace\\
\it \Te{\Ss}{\tfun}\ =\ \liftp{\Te{\Ss}{\tone}\ \fto\ \Te{\Ss}{\ttwo}}\\
\it \\
\it %\ \itt{mkunit}^{\Ss}\ \R\ =\ ()\espace\espace\\
\it %\ \itt{mkint}^{\Ss}_n\ \R\ =\ \itt{lift}\ n\espace\espace\\
\it \itt{plus}^{\Ss}\ (x,y)\ =\ x+y\espace\espace\\
\it \itt{tuple}^{\Ss}\ (x_{\rm1},\ldots,x_n)\ =\ (x_{\rm1},\ldots,x_n)\espace\espace\\
\it \itt{sel}_i^{\,\Ss}\ (x_{\rm1},\ldots,x_n)\ =\ x_i\espace\espace\\
\it \itt{inc}_i^{\Ss}\ =\ \itt{in}_i\o\itt{lift}\espace\espace\\
\it \itt{outc}_i^{\Ss}\ x\ =\ \itt{drop}\o\itt{out}_i\espace\espace\\
\it \ba{lll}\itt{choose}^{\Ss}\ (\bot,&\it \ x_{\rm1},\ldots,x_n)\ =\ \bot\\
\it \itt{choose}^{\Ss}\ (\itt{in}_i~v,&\it \ x_{\rm1},\ldots,x_n)\ =\ x_i\ea\espace\espace\\
\it \itt{mkfun}^{\Ss}\ =\ \itt{lift}\espace\espace\\
\it \itt{apply}^{\Ss}\ =\ \itt{drop}\espace\espace\\
\it \itt{fix}^{\Ss}\ =\ \itt{lfp}\ \reason{Least\ fixed\ point}
\eeqs
}

\section{Domain Factorisation}

A key observation of \cite{Dav93b} was that since there is no concept
of staticness of the body of a lambda expression, there is no point in
having projections on function spaces.  Hence domains are factored into
their evaluable or {\it data\/} parts, and their unevaluable but
applicable {\it forward\/} parts.  For example, for
$\liftp{T\fto{}U}=\Te{\Ss}{\mbox{\tt T}\ \mbox{\tt ->}\ \mbox{\tt U}}$ we need only distinguish two
degrees of definedness---between $\bot$ and values of the form
$\itt{lift}~f$.  This may be encoded by $\lone$---the data part of
$\liftp{T\fto{}U}$---on which there are precisely two projections,
namely $\ID$ and $\BOT$.  Here the forward part is $T\fto{}U$.

The {\it data domain\/} corresponding to type \mbox{\tt T} is $\De{}{\mbox{\tt T}}$,
where $\cD{}$ is defined exactly like $\cT{\Ss}$ except that function
spaces are replaced by the one-point domain $\one=\{\unit\}$, that is,
$\De{}{\tfun}=\lift{\one}$.  The function $\data$ from values in
$\Te{\Ss}{\mbox{\tt T}}$ to their data parts in $\De{}{\mbox{\tt T}}$ is a projection:
it is like the identity except that values from function spaces are
mapped into $\one$.  The projection $\data$ is defined in terms of the
structure of \mbox{\tt T} as follows.
\beqs
\it {\it{}data}_{\subtypespace\stint}\ =\ {\it{}id}_{\subtypespace\stint}\ ,\espace\\
\it {\it{}data}_{\subtypespace\stprod}\ =\ {\it{}data}_{\subtypespace\stone}\ \times\ \ldots\ \times\ {\it{}data}_{\subtypespace\stn}\ ,\espace\\
\it {\it{}data}_{\subtypespace\stsum}\ =\ {\it{}data}_{\subtypespace\stone}\ \plus\ \ldots\ \plus\ {\it{}data}_{\subtypespace\stn}\ ,\espace\\
\it {\it{}data}_{\subtypespace\stfun}\ =\ {\liftp{\Lx.\unit}}\ .
\eeqs
The last definition uses function lifting, defined by $\lf~\bot=\bot$
and $\lf~(\itt{lift}~x)=\itt{lift}~(f~x)$.  Recursive type definitions
give rise to recursive function specifications with only one solution.

In the same style as  $\De{}{\mbox{\tt T}}$ and $\data$ we define
$\Tde{\Ss}{\mbox{\tt T}}$ to give the forward domain for \mbox{\tt T}, and $\fun$ to be
the function mapping values in $\Te{\Ss}{\mbox{\tt T}}$ to their forward parts
in $\Tde{\Ss}{\mbox{\tt T}}$, as follows.  Roughly, $\cTd{\Ss}$ is like
$\cT{\Ss}$ with all lifting removed and sum replaced by product.
\beqs
\it \Tde{\Ss}{\tint}\ =\ \one\ ,\espace\\
\it \Tde{\Ss}{\tprod}\ =\ \Tde{\Ss}{\tone}\ \times\ \ldots\ \times\ \Tde{\Ss}{\tn}\ ,\espace\\
\it \Tde{\Ss}{\tsum}\\
\it \ \ \ \ \ =\ \Tde{\Ss}{\tone}\ \times\ \ldots\ \times\ \Tde{\Ss}{\tn}\ ,\espace\\
\it \Tde{\Ss}{\tfun}\ =\ \Te{\Ss}{\tone}\ \fto\ \Te{\Ss}{\ttwo}\ .
\eeqs
The mapping from standard values to their forward parts is defined by
\beqs
\it {\it{}fun}_{\subtypespace\stint}\ =\ \Lx.\unit\ ,\espace\\
\it {\it{}fun}_{\subtypespace\stprod}\ =\ {\it{}fun}_{\subtypespace\stone}\ \times\ \ldots\ \times\ {\it{}fun}_{\subtypespace\stn}\ ,\espace\\
\it \ba{lll}{\it{}fun}_{\subtypespace\stsum}\ \bot&\it =\ \bot\espace\\
\it {\it{}fun}_{\subtypespace\stsum}\ (\itt{in}_i~v)\\
\it \ \ \ \ \ =\ (\bot,\ldots,\bot,v,\bot_,\ldots,\bot)&\it \reason{$v$\ in\ $\ith$\ position}\ ,\ea\espace\\
\it {\it{}fun}_{\subtypespace\stfun}\ =\ \itt{drop}\ .
\eeqs
We write $\fac$ for $\Lx.(\data~x,\ \fun~x)$.  Then
$\De{}{\mbox{\tt T}}\times\Tde{\Ss}{\mbox{\tt T}}$ is a factorisation of $\Te{\Ss}{\mbox{\tt T}}$,
and
\beqs
\it \fac\ \in\ \Te{\Ss}{\makebox{\tt T}}\ \fto\ (\De{}{\makebox{\tt T}}\ \times\ \Tde{\Ss}{\makebox{\tt T}})
\eeqs
is an embedding which {\it determines\/} the corresponding projection
$\unfac$ (since they are related by $\unfac\o\fac=\itt{id}$ and
$\fac\o\unfac\we\itt{id}$).  Rather than give an explicit definition of
$\unfac$ we give some examples.  The factorisation of $\Int$ is
$\Int\times\one$; more generally, the factorisation of any domain $D$
corresponding to a type not containing \mbox{\tt ->} is just $D\times\one$.  If
$T=\Te{\Ss}{\mbox{\tt T}}$ and $U=\Te{\Ss}{\mbox{\tt U}}$ then the factorisation of
$\lift{(T\fto{}U)}=\Te{\Ss}{\mbox{\tt T}\ \mbox{\tt ->}\ \mbox{\tt U}}$ is
$\lift{\one}\times(T\fto{}U)$---note that factorisation `stops' at
function-space constructors.  Now
${\it{}fac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~\bot=(\bot,\bot)$, and
${\it{}fac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\itt{lift}~f)=(\lunit,~f)$ for all $f$.  In
the other direction ${\it{}unfac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\bot,f)=\bot$ and
${\it{}unfac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\lunit,~f)=(\itt{lift}~f)$; this must be
since in general $\unfac\o\fac$ is the identity.

\section{Projection Semantics}

Standard values $v$ and $v'$ are related by nonstandard value $(\A,\K)$
when their data parts are related by $\A$, and their forward parts are
logically related by $\K$.  At each type \mbox{\tt T} the relation is
$\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$, defined by
\beqs
\it \lefteqn{\Re{}{\makebox{\tt T}}~((\A,\K),v,v')}\\
\it \ \ \ \ \ =&\it \mit(\A~d=\A~d')\ \wedge\ \Rde{}{\makebox{\tt T}}~(\K,f,f')\\
\it &\it \mbox{\rm{}where}\\
\it &\it \ \ \ \ba[t]{lll}(d,f)&\it \:=\ \fac~v\\
\it \mit(d',f')&\it \:=\ \fac~v'\ ,\ea
\eeqs
where $\Rde{}{\mbox{\tt T}}$ is defined by
\beqs
\it \Rde{}{\tint}~((),(),())\ =\ \itt{True}\ ,\espace\\
\it \Rde{}{\tprod}\ =\ \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace\\
\it \Rde{}{\tsum}\\
\it \ \ \ \ \ =\ \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace\\
\it \Rde{}{\tfun}\ =\ \Re{}{\tone}\ \fto\ \Re{}{\ttwo}\ .
\eeqs
Here $\x$ and $\fto$ are the standard operations on (ternary)
relations:  for relations $R$ and $S$ we have
$(R\times{}S)((x,y),(x',y'),(x'',y''))$ iff $R(x,x',x'')$ and
$S(y,y',y'')$, and $(R\fto{}S)(f,g,h)$ iff for all $x$, $y$, and $z$
such that $R(x,y,z)$ we have $S(f~x,\ g\ y,\ h~z)$.  Here, recursive
type definitions give recursive relation specifications, which have
inclusive least-fixed-point solutions. (A relation is inclusive if,
when it holds for each element of an ascending chain, it also holds at
the limit.  Such relations are sometimes called {\it admissible\/} or
{\it chain complete}.  Some work is required to show inclusivity of
these recursively-defined relations.) Thus the $\Ps$ type semantics
$\cT{\Ps}$ must be
\beqs
\it \Te{\Ps}{\makebox{\tt T}}\ =\ \proj{\De{}{\makebox{\tt T}}}\ \times\ \Tde{\Ps}{\makebox{\tt T}}\ ,
\eeqs
where $\proj{\De{}{\mbox{\tt T}}}$ is the lattice of projections on domain
$\De{}{\mbox{\tt T}}$, and $\cTd{\Ps}$ is defined exactly like $\cTd{\Ss}$ with
superscript $\Ps$ everywhere replacing superscript $\Ss$.

\fig{tb}{Projection semantics.}{fig:fsem}{
\beqs
\it %\ \itt{mkunit}^{\Ps}\ \R\ =\ ()\\
\it %\espace\espace\\
\it %\ \itt{mkint}^{\Ps}_n\ \R\ =\ \itt{lift}\ n\\
\it %\espace\espace\\
\it \itt{plus}^{\Ps}\ ((\A,()),(\B,()))\\
\it \ \ \ =\ \left\{\ba{ll}(\ID,()),&\it \ \ \ \ \mbox{\rm\ if\ $\A=\ID$\ and\ $\B=\ID$}\\
\it (\BOT,()),&\it \ \ \ \ \mbox{\rm\ otherwise}\ea\right.\espace\\
\it \itt{tuple}^{\Ps}\ ((\A_{\rm1},\K_{\rm1}),\ldots,(\A_n,\K_n))\\
\it \ \ \ =\ ((\A_{\rm1}\x\ldots\x\A_n),(\K_{\rm1},\ldots,\K_n))\espace\\
\it \itt{sel}_i^{\,\Ps}\ ((\A_{\rm1}\x\ldots\x\A_n),(\K_{\rm1},\ldots,\K_n))\ =\ (\A_i,\K_i)\espace\\
\it \itt{inc}_i^{\Ps}\ (\A,\K)\ =\ (C_i\ \A,\ (\top,\ldots,\top,\K,\top,\ldots,\top))\espace\\
\it \itt{outc}_i^{\Ps}\ (\A,(\K_{\rm1},\ldots,\K_n))\ =\ (\OUTC_i\ \A,\ \K_i)\espace\\
\it \itt{choose}^{\Ps}\ ((\A,\K),x_{\rm1},\ldots,x_n)\\
\it \ \ \ \ \ =\ \left\{\ba{ll}\bot,&\it \ \ \ \ \mbox{\rm\ if}\ \A\not\se\glb_{\oin}\ C_i~\BOT\\
\it x_{\rm1}\glb\ldots\glb{}x_n,&\it \ \ \ \ \mbox{\rm\ otherwise}\ea\right.\espace\\
\it \itt{mkfun}^{\Ps}\ f\ =\ (\ID,f)\espace\\
\it \itt{apply}^{\Ps}\ (\A,f)\ =\ \left\{\ba{lll}\bot,&\it \ \ \ \ \mbox{\rm\ if}\ \A=\BOT\\
\it f,&\it \ \ \ \ \mbox{\rm\ if}\ \A=\ID\ \ea\right.\espace\\
\it \itt{fix}^{\Ps}\ =\ \itt{gfp}\ \reason{Greatest\ fixed\ point}
\eeqs
}

The defining constants for the projection semantics $\cE{\Ps}$ are
given in Figure~\ref{fig:fsem}. The following notation is used for
specifying projections on the data domains.  For sum type $\tsum$ with
data domain $D_1\plus\ldots\plus D_n$ define $C_i~\A$ to be
$\ID+\cdots+\ID+\A+\ID+\cdots+\ID$ where $\A$ is the $i^{th}$ summand,
$\OUTC_i~(\G_1+\cdots+\G_n)$ to be $\G_i$, and $\OUTC_i~\BOT$ to be
$\BOT$.

\ \\
\noindent {\bf The Central Result.}  For expression \mbox{\tt e} of type \mbox{\tt T} with free-variable
environment of type \mbox{\tt E}, the functions $\Ee{\Ps}{\mbox{\tt e}}$,
$\Ee{\Ss}{\mbox{\tt e}}$, and $\Ee{\Ss}{\mbox{\tt e}}$ are logically related by $\cR{}$,
that is
\beqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\makebox{\tt T}})\ (\Ee{\Ps}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}})\ .
\eeqs
Further, $\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and $\K$.\ \ $\Box$

\ \\
\noindent The bulk of proof is omitted; it consists of showing that the defining
constants are similarly related, and a simple induction on the
structure of expressions showing that if the defining constants are
logically related, then so are the semantic functions.  By way of example
we consider the expression form $\esum$, and the relevant defining constant
$\it plus$.
To show that $\it plus^{\Ss}$ and $\it plus^{\Ps}$ are correctly related
we need to show that
\beqqs
\it ((\Re{}{\tint}\x\Re{}{\tint})\fto\Re{}{\tint})\ (&\it plus^{\Ps},\\
\it &\it plus^{\Ss},\\
\it &\it plus^{\Ss})\ ,
\eeqqs
that is, for all $\A_1$, $\A_2$, $d_1$, $d_2$, $d_1'$, and $d_2'$ that
\beqs
\it &\it \mit(\ID\ d_{\rm1}=\ID\ d_{\rm1}')\ \land\ (\ID\ d_{\rm2}=\ID\ d_{\rm2}')\\
\it \limp&\it \mit\ID\ (d_{\rm1}+d_{\rm2})=\ID\ (d_{\rm1}'+d_{\rm2}')
\eeqs
and for $\A_1\neq\ID$ or $\A_2\neq\ID$ that
\beqs
\it &\it \mit(\A_{\rm1}\ d_{\rm1}=\A_{\rm1}\ d_{\rm1}')\ \land\ (\A_{\rm2}\ d_{\rm2}=\A_{\rm2}\ d_{\rm2}')\\
\it \limp&\it \mit\BOT\ (d_{\rm1}+d_{\rm2})=\BOT\ (d_{\rm1}'+d_{\rm2}')\ .
\eeqs
For the inductive case $\esum$ with environment type \mbox{\tt E} we need to show that if
\beqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (\Ee{\Ps}{\eone},\ \Ee{\Ss}{\eone},\ \Ee{\Ss}{\eone})
\eeqs
and
\beqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (\Ee{\Ps}{\etwo},\ \Ee{\Ss}{\etwo},\ \Ee{\Ss}{\etwo})
\eeqs
then
\beqqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (&\it \Ee{\Ps}{\esum},\\
\it &\it \Ee{\Ss}{\esum},\\
\it &\it \Ee{\Ss}{\esum})\ .
\eeqqs
Suppose $\Re{}{\mbox{\tt E}}~(\R^{\Ps},\R^{\Ss},\R'^{\Ss})$.  By the induction
hypothesis we have
$\Re{}{\tint}~(\A_i,v_i,v'_i)$, where 
$(\A_i,())=\Ee{\Ps}{\ei}~\R^{\Ps}$,
$(v_i,())=\Ee{\Ps}{\ei}~\R^{\Ss}$, and
$(v'_i,())=\Ee{\Ps}{\ei}~\R'^{\Ss}$, for $i=1,2$.
Now
\beqqs
\it \Re{}{\tint}\ (&\it \Ee{\Ps}{\esum}\ \R^{\Ps},\\
\it &\it \Ee{\Ss}{\esum}~\R^{\Ss},\\
\it &\it \Ee{\Ss}{\esum}~\R'^{\Ss})
\eeqqs
iff
\beqqs
\it \Re{}{\tint}\ (&\it plus^{\Ps}\ ((\A_{\rm1},()),(\A_{\rm2},())),\\
\it &\it plus^{\Ss}\ \mit((v_{\rm1},()),(v_{\rm2},())),\\
\it &\it plus^{\Ss}\ \mit((v'_{\rm1},()),(v'_{\rm2},())))\ ,
\eeqqs
which holds since $\it plus^{\Ps}$ and $\it plus^{\Ps}$, and their various
arguments, are correctly related.

\section{Abstract Domains}

At each type \mbox{\tt T} we require a finite abstraction of $\Te{\Ps}{\mbox{\tt T}}$.
This abstract domain is ${\it{}FProj}_{\subtypespace\stt}\times{\it{}FFor}_{\subtypespace\stt}$,
where ${\it{}FProj}_{\subtypespace\stt}$ is a finite abstraction of the lattice
$\proj{\De{}{\mbox{\tt T}}}$, and ${\it{}FFor}_{\subtypespace\stt}$ is a finite abstraction
of $\Tde{\Ps}{\mbox{\tt T}}$.  The definition of ${\it{}FProj}_{\subtypespace\stt}$ is based
on that in \cite{Lau91a}.  A projection $\G$ is in
${\it{}FProj}_{\subtypespace\stt}$ if $\G~\mbox{\bf proj}~\mbox{\tt T}$ can be inferred from
the rules given in Figure~\ref{fig:infrules}.
%
\fig{tb}{Inference rules for finite projection domains.}{fig:infrules}{
\beqs
\it \begin{array}{lll}\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tint}\ \ \ \ \ \ \ \ \ \ &\it \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tint}\espace\espace\\
\it \axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tfun}&\it \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tfun}\end{array}\espace\espace\\
\it \axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tsum}\espace\espace\\
\it \infero{\G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{(C_{\rm1}\ \G_{\rm1})\glb\cdots\glb(C_n\ \G_n)\ \ \mbox{\bf{}proj}\ \ \tsum}\espace\espace\\
\it \infero{\G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{\G_{\rm1}\x\ldots\x\G_n\ \ \mbox{\bf{}proj}\ \ \tprod}\espace\espace\\
\it \infer{\ba{ll}&\it \G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \aone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \an\\
\it \vdash&\it P(\G_{\rm1},\ldots ,\G_n)\ \ \mbox{\bf{}proj}\ \ \ti\makebox{\tt (}\aone\makebox{\tt ,...,}\an\makebox{\tt )}\ea}{\mu(\G_{\rm1},\ldots,\G_n).P(\G_{\rm1},\ldots,\G_n)\ \ \mbox{\bf{}proj}\ \ \ai}\espace\\
\it \ \ \ \ \ [\mbox{\rm{}where}\ \ai\makebox{\tt =}\ti\makebox{\tt (}\aone\makebox{\tt ,...,}\an\makebox{\tt )}]
\eeqs
}
For recursively-defined types the rules yield only those
projections that act on each recursive instance of a data structure in
the same way.  Thus ${\it{}FProj}_{\subtypespace{\makebox{\subtypesize\tt FunList}} }$ comprises $\BOT$ and
$\SPINE~\A$ for $\A$ ranging over $\BOT$ and $\ID$, where
\beqs
\it \SPINE\ \A\ =\ \mu\G.(\itt{NIL}\ \ID)\ \glb\ (\itt{CONS}\ (\A\ \x\ \G))\ ,
\eeqs
so $\SPINE\ \BOT$ specifies static spines and dynamic elements, and
$\SPINE\ \ID$ is the identity.  More generally, the abstract list
constructor is isomorphic to lifting.  Similarly,
${\it{}FProj}_{\subtypespace{\makebox{\subtypesize\tt FunTree}} }$
comprises $\BOT$ and $\LBR\ \A$ for $\A$ ranging over $\BOT$ and $\ID$, where
\beqs
\it \LBR\ \A\ =\ \mu\G.(\itt{LEAF}\ \A)\ \glb\ (\itt{BRANCH}\ (\G\ \x\ \G))\ ,
\eeqs
so $\LBR\ \A$ specifies static branches and leaves, and $\A$ of the
leaf nodes.  Again, the abstract tree constructor is isomorphic to
lifting.  This compares favourably with
BHA strictness analysis, for which the corresponding abstract
constructors are typically {\it double\/} lifting \cite{Wad87,Sew93}.

Value $\K$ is in ${\it{}FFor}_{\subtypespace\stt}$ if $\K~\bfabsf~\mbox{\tt T}$ can be
inferred from the following.

There is only one forward value at type $\tint$.
\beqs
\it \axm{()\ {}\ \bfabsf\ {}\ \tint}\ .
\eeqs
For products and sums,
\beqs
\it \infer{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tone\ {\ }\ \cdots\ {\ }\ \K_n\ {}\ \bfabsf\ {}\ \tn}{(\K_{\rm1},\ldots,\K_n)\ {}\ \bfabsf\ {}\ \tprod}\ ,
\eeqs
%\espace\espace
\beqs
\it \infero{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tone\ {\ }\ \cdots\ {\ }\ \K_n\ {}\ \bfabsf\ {}\ \tn}{(\K_{\rm1},\ldots,\K_n)\ {\ }\ \bfabsf\ {}\ \tsum}\ .
\eeqs
Function spaces consist of a set of step functions closed under lub.
\beqs
\it \infero{\ba{ll}\T_{\rm1}\in{\it{}FTran}_{\subtypespace\stone}\ {\ {\ }\ }\ \K_{\rm1}\ \bfabsf\ \tone\\
\it \T_{\rm2}\in{\it{}FTran}_{\subtypespace\sttwo}\ {\ {\ }\ }\ \K_{\rm2}\ \bfabsf\ \ttwo\ea}{\it{}step\ ((\T_{\rm1},\K_{\rm1}),(\T_{\rm2},\K_{\rm2}))\ {}\ \bfabsf\ {}\ \tfunp}\ ,
\eeqs
where
\beqs
\it step\ (v_{\rm1},v_{\rm2})\ x\ =\ v_{\rm2},&\it {\rm\ if}\ v_{\rm1}\we{}x\\
\it step\ (v_{\rm1},v_{\rm2})\ x\ =\ \bot,&\it {\rm\ otherwise}\ ,
\eeqs
and
\beqs
\it \infer{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tfunp\ {\ {\ }\ }\ \K_{\rm2}\ {}\ \bfabsf\ {}\ \tfunp}{(\K_{\rm1}\lub\K_{\rm2})\ {}\ \bfabsf\ {}\ \tfunp}\ .
\eeqs
This gives the full space of monotonic functions on the abstract
domains.

For recursively-defined types, roughly speaking, we choose those
forward values that represent each component of the same type by the
same value.  Given type definitions $\tdefn$, which we will write
$\mbox{\tt A}_i\mbox{\tt =T}_i\mbox{\tt (A}_1\mbox{\tt ,...,A}_n\mbox{\tt )},\ \oin$, if by assuming
$\K_i~\bfabsf~\ai$ for $\oin$ we may deduce
$P_i(\K_1,\ldots,\K_n)~\bfabsf~\ti(\aone\ldots\an)$ for $\oin$, then
\beqs
\it \mu(\K_{\rm1},\ldots ,\K_n).(P_{\rm1}(\K_{\rm1},\ldots ,\K_n),\ldots ,P_n(\K_{\rm1},\ldots ,\K_n))
\eeqs
is a tuple $(\K_1,...,\K_n)$ of values such that $\K_i\ \bfabsf\ \mbox{\tt A}_i$
for $\oin$.

For all \mbox{\tt T} the set ${\it{}FProj}_{\subtypespace\stt}\times{\it{}FFor}_{\subtypespace\stt}$ is a 
finite lattice containing the top and bottom elements of $\Te{\Ps}{\mbox{\tt T}}$.

For \mbox{\tt T} not containing \mbox{\tt ->} the domain $\Tde{\Ps}{\mbox{\tt T}}$ is isomorphic
to $\one$, so ${\it{}FFor}_{\subtypespace\stint}$ is $\one$.  For \mbox{\tt Int}\ \mbox{\tt ->}\ \mbox{\tt Int}
we have
$\Tde{\Ps}{\mbox{\tt Int}\ \mbox{\tt ->}\ \mbox{\tt Int}}=(\proj{\De{}{\tint}}\x\one)\fto(\proj{\De{}{\tint}}\x\one)$,
so ${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt Int->Int}} }$ is
$({\it{}FProj}_{\subtypespace\stint}\x\one)\fto({\it{}FProj}_{\subtypespace\stint}\x\one)$.  A
data structure of recursive type \mbox{\tt A=T(A)} may be thought of as some
(possibly infinite) number of elements of \mbox{\tt T(())}.  For example, the
value $\itt{cons}\ (f,\ \itt{cons}\ (g,\ \itt{nil}\ ()))$ in the
standard domain for \mbox{\tt FunList} decomposes into $\itt{cons}\ (f,())$,
$\itt{cons}\ (g,())$, and $\itt{nil}\ ()$.  The (implicit) abstraction
function maps such a data structure to the greatest lower bound of
these elements, giving a safe abstraction of the nonstandard values of
each element.  Thus
${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt A=T(A)}} }={\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt T(())}} }$, so
${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt FunList}} }\iso{\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt FunTree}} }\iso{\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt Int->Int}} }$.

\section{Examples of Analysis}

For all closed expressions \mbox{\tt e} the abstract value
$\Ee{\Ps}{\mbox{\tt e}}~[\,]$ of \mbox{\tt e} is of the form $(\ID,\K)$, showing
that closed expressions are always entirely static.  For expressions of
function type the abstract forward value $\K$ is a function from
abstract arguments of \mbox{\tt e} to abstract results.

First we consider functions on lists.
Let expression \mbox{\tt length} denote usual length function:
\if 0\theisrr
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}length\ .\ {\char'134}xs\ .\ case\ xs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ 0}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 1\ +\ length\ zs)\ .}\vspace{-0.0em}
\end{flushleft}
\else
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}length\ .\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ {\char'134}xs\ .\ case\ xs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ 0}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 1\ +\ length\ zs)\ .}\vspace{-0.0em}
\end{flushleft}
\fi
The abstract forward value of \mbox{\tt length} is of the form
$\L(\A,\K).(\T~\A,())$, where $\T$ maps $\SPINE~\BOT$ and $\SPINE~\ID$
to $\ID$, and $\BOT$ to $\BOT$.  This reveals that the result of
\mbox{\tt length} is independent of the values of list elements, and gives
a static result when the argument has a static spine.

Let \mbox{\tt append} stand for the expression denoting the usual function for
appending two lists:
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}append\ .\ \ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ {\char'134}xs\ .\ {\char'134}ys\ .\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ case\ xs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ ys}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ (z,\ append\ zs\ ys))\ .}\vspace{-0.0em}
\end{flushleft}
The abstract value of \mbox{\tt append} is
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt xs}} },\K_{\subtypespace{\makebox{\subtypesize\tt xs}} }).\\
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt ys}} },\K_{\subtypespace{\makebox{\subtypesize\tt ys}} }).\\
\it (\A_{\subtypespace{\makebox{\subtypesize\tt xs}} }\glb\A_{\subtypespace{\makebox{\subtypesize\tt ys}} },\ \K_{\subtypespace{\makebox{\subtypesize\tt xs}} }\glb\K_{\subtypespace{\makebox{\subtypesize\tt ys}} })))\ .
\eeqs
This reveals that partial applications of \mbox{\tt append} are static up to WHNF, and the
abstract value of the result is the greatest lower bound of the two
arguments.  In general, the abstract value of a closed expression
of the form $\mbox{\tt {\char'134}x}_1\mbox{\tt .{\char'134}x}_2\ldots\mbox{\tt .e}$ will reveal that all partial
applications of the expression are static up to WHNF.

Let \mbox{\tt reverse1} stand for the expression denoting the naive reverse function:
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}reverse1\ .\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ {\char'134}xs\ .\ case\ xs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ ->\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ ()}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ append\ (reverse\ zs)\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (cons\ (z,\ nil\ ())))\ .}\vspace{-0.0em}
\end{flushleft}
The abstract forward value of \mbox{\tt reverse1} is the identity, so the abstraction of
the elements of a list doesn't change by reversing the list.

Let \mbox{\tt compose} stand for \mbox{\tt {\char'134}f.{\char'134}g.{\char'134}x.f\ (g\ x)}.  The abstract value of \mbox{\tt compose} is
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt f}} },\K_{\subtypespace{\makebox{\subtypesize\tt f}} }).\\
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt g}} },\K_{\subtypespace{\makebox{\subtypesize\tt g}} }).\\
\it (\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt f}} }=\BOT\mbox{\rm\ or}\ \A_{\subtypespace{\makebox{\subtypesize\tt g}} }=\BOT\\
\it \K_{\subtypespace{\makebox{\subtypesize\tt f}} }\o\K_{\subtypespace{\makebox{\subtypesize\tt g}} },&\it \mbox{\rm\ otherwise}\ea\right\})))\ .
\eeqs
Thus, the
result of the application of the composition of two functions is
dynamic, and maps all values to dynamic values, if either function is
dynamic; otherwise, the result is given by the application of the
composition of the abstract forward values to the abstract argument.

Let \mbox{\tt listcomp} stand for the expression denoting the function
that composes lists of functions:
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}listcomp\ .\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ {\char'134}fs\ .\ case\ fs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ ->\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\char'134}x.x}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ let\ (g,gs)\ =\ p\ in\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ compose\ g\ (listcomp\ gs))\ .}\vspace{-0.0em}
\end{flushleft}
The abstract value of \mbox{\tt listcomp} is
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).\\
\it (\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\SPINE\ \ID\\
\it \glb_{i\geq{\rm0}}(\K_{\subtypespace{\makebox{\subtypesize\tt fs}} })^i,&\it \mbox{\rm\ otherwise}\ea\right\}))\ .
\eeqs
Since the abstract values of lists contain no information about the
length of the lists of which they are abstractions, the abstract value
of the composition of list elements is the glb of the composition over
all lengths.

\commentout{
Next we consider functions on trees.
Let \mbox{\tt count} be short for
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}count\ .\ {\char'134}t\ .\ case\ t\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ leaf\ l\ \ \ ->\ 1}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ branch\ p\ ->\ let\ (t1,t2)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ count\ t1\ +\ count\ t2)\ ,}\vspace{-0.0em}
\end{flushleft}
denoting the function that counts leaves.  The abstract forward value
\mbox{\tt count} is of the form $\L(\A,\K).(\T~\A ())$, where $\T$ maps
$\LBR~\BOT$ and $\LBR~\ID$ to $\ID$, and $\BOT$ to $\BOT$.  This
reveals that the result of \mbox{\tt count} is independent of the values of leaf
elements, and maps trees with static branch and leaf structure, but
possibly dynamic values at the leaves, to static results.
}

Let \mbox{\tt flatten} stand for the expression denoting the function that flattens
trees into lists:
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}flatten\ .\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ {\char'134}t\ .\ case\ t\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ leaf\ l\ \ \ ->\ cons\ (l,\ nil\ ())}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ branch\ p\ ->\ let\ (t1,t2)\ =\ p\ in}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ append\ (flatten\ t1)\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (flatten\ t2))\ .}\vspace{-0.0em}
\end{flushleft}
The abstract value of \mbox{\tt flatten} is
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).\\
\it (\T\ \A,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \T\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\LBR\ \ID\\
\it \K_{\subtypespace{\makebox{\subtypesize\tt fs}} },&\it \mbox{\rm\ otherwise}\ea\right\}))\ ,
\eeqs
where $\T$ maps $\BOT$ to $\BOT$, and $\LBR~\A$ to $\SPINE~\A$ for $\A$
ranging over $\ID$ and $\BOT$.

The function denoted by \mbox{\tt compose\ listcomp\ flatten} that composes
trees of functions has abstract value
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).\\
\it (\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\LBR\ \ID\\
\it \glb_{i\geq{\rm0}}(\K_{\subtypespace{\makebox{\subtypesize\tt fs}} })^i,&\it \mbox{\rm\ otherwise}\ea\right\}))\ .
\eeqs
Similarly to the case for lists, the abstract values of trees contain
no information about the structure of the trees of which they are abstractions,
so the abstract value of the composition of the values of the leaves
is the glb of the composition over all tree structures.

\commentout{
Let \mbox{\tt treecomp} stand for the expression denote the function
that composes trees of functions:
\begin{flushleft}
\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}treecomp\ .\ {\char'134}t\ .\ case\ fs\ of}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ leaf\ f\ \ \ ->\ f}\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ branch\ p\ ->\ let\ (t1,t2)\ =\ p\ in\ }\vspace{-0.0em}\\
\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ compose\ (listcomp\ t1)\ (listcomp\ t2))\ ,}\vspace{-0.0em}
\end{flushleft}
The abstract value of \mbox{\tt treecomp} is simply
\beqs
\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).(\ID,\ \left\{\ba{lll}\bot&\it \ \ \ \ \ \ \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }=\BOT\\
\it \K_{\subtypespace{\makebox{\subtypesize\tt fs}} }&\it \ \ \ \ \ \ \mbox{\rm\ otherwise}\ \ea\ \right\}))\ .
\eeqs
That is ...}

\section{More on Abstract Domains}

The sizes of the abstract domains and the representations of the
abstract values can be considerably optimised.  The nonstandard
semantics of application---embodied by $\it apply^{\Ps}$---guarantees
that the abstract values $(\BOT,f)$ and $(\BOT,f')$ from
$\Te{\Ps}{\tfun}$, for all $f$ and $f'$, are effectively the same:
$\it apply^{\Ps}~(\BOT,f)=\bot$ for all $f$.  A practical analyser
would take advantage of this fact, identifying $(\BOT,f)$ over
all $f$.  More generally, for function types embedded
within data structures, e.g.\ $\mbox{\tt (}\tint\mbox{\tt ,}\tfun\mbox{\tt )}$, abstract values
$((\A,()),(\BOT,f))$ would be identified over all $f$.

In the following we restrict attention to {\it denotable\/} values:  at
each type \mbox{\tt T} those values that can be expressed as $\Ee{}{\mbox{\tt e}}~[]$ for
some \mbox{\tt e}.  For every value $v\in\Te{\Ss}{\mbox{\tt T}}$ there is a best
abstraction---a greatest value $(\A,\K)\in\Te{\Ps}{\mbox{\tt T}}$ such that
$\Re{}{\mbox{\tt T}}((\A,\K),v,v)$.  For types $\tone$ and $\ttwo$ not
containing \mbox{\tt ->} we have
$\Tde{\Ps}{\tfun}\iso(\proj{\De{}{\tone}}\x\one)\fto(\proj{\De{}{\ttwo}}\x\one)$,
and if $\K\in\Tde{\Ps}{\mbox{\tt T}}$ is greatest such that
$\Rde{}{\tfun}~(\K,f,f)$ for denotable $f$, then $\K$ maps $(\ID,())$
to $(\ID,())$ and distributes over $\glb$.  The subset of functions
from $\Tde{\Ps}{\tfun}$ that map $(\ID,())$ to $(\ID,())$ and
distribute over $\glb$ forms a complete lattice, hence attention may be
restricted to this subset.  Not only does this reduce the number of
abstract functions, it also reduces their representation:  such
functions are determined by their behaviour on the $\glb$-{\it basis\/}
of the lattice $\proj{\De{}{\tone}}$---the set $B$ of values such that
every element of $\proj{\De{}{\tone}}$ is the glb of some (possibly
empty) subset of $B$, and no element of $B$ can be expressed as the glb
of some subset not containing that element.  This optimisation can be
generalised to higher order:  $\Tde{\Ps}{\tfun}$ may be restricted to
functions that map $\top$ to $\top$ and distribute over $\glb$, for all
function types $\tfun$.  (These results follow from those shown in
\cite{Dav94}.)

\section{Related Work}

Consel \cite{Con90} describes a binding-time analysis for higher-order
untyped languages.  As in our analysis abstract values have two parts,
the first describing the static/dynamic properties of values, and the
second describing how (for function types) abstract arguments are
mapped to abstract results; there appears to be an implicit domain
factorisation similar to ours, based on implicit type information
collected by analysis semantics.  In this respect there are many
superficial similarities between the two techniques.  No formal
relation to the standard semantics is given, making a formal comparison
with our technique problematic.

\section{Conclusion}

We have successfully generalised Launchbury's monomorphic
projection-based binding-time analysis to higher-order, using
abstract domains smaller than those typically used in BHA
strictness analysis.

The next step would be to generalise to handle Hindley-Milner
polymorphism.  This has been done with good results at first order for
binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
use of these theories will give a reasonably straightforward generalisation
to polymorphism.  In addition to making the analysis more widely applicable,
this should also greatly reduce the run-time cost of analysis.

Though not developed here, using the same approach it is
possible to give a strictness analysis technique, again closely related to
Hunt's PER-based strictness analysis technique \cite{Hun91}.

\section{Acknowledgements}

Thanks to the referees for their constructive comments on the first
version of this paper.

\begin{thebibliography}{123456}

\bibitem[Abr89]{Abr89}
S. Abramsky. 
``The lazy lambda calculus.'' 
{\it Research Topics in Functional Programming.}
David Turner, ed., Addison-Wesley 1989.

\bibitem[Abr90]{Abr90}
S. Abramsky. 
``Abstract interpretation, logical relations and Kan extensions.''
{\it Journal of Logic and Computation}, 1, 1990.

\bibitem[Bar91]{Bar91}
G. Baraki.
``A note on abstract interpretation of polymorphic functions.''
In \cite{Hug91}.

\bibitem[Bar93]{Bar93}
G. Baraki.
{\it Abstract Interpretation of Polymorphic Higher-Order Functions.}
Ph.D. thesis, Research report FP-1993-7, 
Department of Computing Science, University of Glasgow.

\bibitem[BEJ88]{BEJ88}
D. Bj{\odash}rner, A.P. Ershov, and N.D. Jones, eds.
{\it Partial Evaluation and Mixed Computation,
Proceedings IFIP TC2 Workshop, Gammel Avern{\ae}s}, Denmark, October 1987.
North-Holland, 1988.

\bibitem[Con90]{Con90}
C. Consel. 
``Binding Time Analysis for Higher Order Untyped Functional Languages.''
Proceedings of the 1990 ACM Conference on LISP and 
Functional Programming, pp264-272.

\bibitem[Dav93]{Dav93b}
K. Davis.
``Higher-order Binding-time Analysis.'' 
{\it Proceedings of the 1993 ACM on Partial Evaluation and
Semantics-Based Program Manipulation (PEPM '93)},
ACM Press, 1993.

\bibitem[Dav94]{Dav94}
K. Davis.\ \ 
{\it Projection-based Program Analysis.}
Thesis submitted for degree of Ph.D., Computing Science Department,
University of Glasgow, 1994.

\bibitem[Go92]{Go92}
C.K. Gomard.
``A self-applicable partial evaluator for the lambda calculus:  Correctness
and pragmatics.''
ACM TOPLAS, Vol 14, No.\ 2, April 1992.

\bibitem[HM94]{HM94}
F. Henglein and C. Mossin.
``Polymorphic binding-time analysis.''
European Symposium on Programming (ESOP '94).

\bibitem[Hug91]{Hug91}
J. Hughes, ed.
{\it Proceedings of the 1991 Conference on Functional Programming Languages
and Computer Architecture (FPCA '91)}, Cambridge, Sept 1991. LNCS 523,
Springer Verlag, 1991.

\bibitem[Hun91]{Hun91}
S. Hunt.
``PERs generalise projections for strictness analysis (extended abstract).''
{\it Proceedings of the 1990 Glasgow Workshop
on Functional Programming}.
Simon L. Peyton Jones {\it et al.}, eds.
Springer Workshops in Computing.
Springer-Verlag, 1991.

\bibitem[HS91]{HS91}
S. Hunt and D. Sands.
``Binding time analysis:  a new PERspective.''
{\it ACM Symposium on Partial Evaluation
and Semantics-Based Program Manipulation},
SIGPLAN Notices Vol.\ 26, No.\ 9, 1991.

\bibitem[Jen92]{Jen92}
T. Jensen.
{\it Abstract Interpretation in Logical Form}.
Ph.D. thesis, Report 93/11, Department of Computer Science,
University of Copenhagen, 1992.

\bibitem[Jon88]{Jon88}
N.D. Jones.
``Automatic program specialization: A re-examination from
basic principles.'' In \cite{BEJ88}.

\bibitem[Lau88]{Lau88}
J. Launchbury.
``Projections for specialisation.''
In \cite{BEJ88}.

\bibitem[Lau91a]{Lau91a}
J. Launchbury. 
{\it Projection Factorisations in Partial Evaluation.} PhD
Thesis, Glasgow University, Nov 89. Distinguished Dissertation in
Computer Science, Vol 1, CUP, 1991.

\bibitem[Lau91c]{Lau91c}
J. Launchbury.
``A strongly-typed, self-applicable partial evaluator.''
In \cite{Hug91}.

\bibitem[Mog89]{Mog89}
T. Mogensen.
``Binding-time analysis for polymorphically typed higher order languages.''
{\it International Joint Conference on Theory and Practice of Software
Development}. LNCS 352. Springer-Verlag 1989.

\bibitem[NN88]{NN88}
H.R. Nielson and F. Nielson.
``Automatic binding-time analysis for a typed $\L$-calculus.''
{\it Science of Computer Programming 10},
North Holland, 1988.  Also in POPL '88.

\bibitem[Sch88]{Sch88}
D.A. Schmidt.
``Static properties of partial reduction.''
In \cite{BEJ88}.

\bibitem[Sew93]{Sew93}
J. Seward.
``Polymorphic strictness analysis using frontiers.''
{\it Proceedings of the 1993 ACM on Partial Evaluation and
Semantics-Based Program Manipulation (PEPM '93)},
ACM Press, 1993.

\bibitem[Wad87]{Wad87}
P. Wadler.
``Strictness analysis on non-flat domains by abstract interpretation over finite domains. '' 
S. Abramsky, C. Hankin, eds.
{\it Abstract Interpretation of Declarative Languages.}
Ellis-Horwood, 1987.

\end{thebibliography}

\end{document}

### 2017-02-22 pubs/demand_analysis/smjsamp.tex
demand analysis manuscript

%% Kluwer Small Journal Article Sample File
%% June 20, 1994

%%   Use one of these two commands: 
\documentstyle{smjrnl} % For Computer Modern Fonts, or
%\documentstyle[smjfonts]{smjrnl} % for PostScript fonts

\begin{document}
\begin{article}

%%%%% To be entered at Kluwers: =====>>

\journame{Small Journal Name}
\volnumber{9}
\issuenumber{4}
\issuemonth{October}
\volyear{1992}

%% Do not delete either of the following two commands.
%% Please supply facing curly brackets for the part you
%% are not using for this article.
%\received{May 1, 1991}\revised{}

\authorrunninghead{J.R. Anderson and M. Matessa}
\titlerunninghead{An Incremental, Bayesian Algorithm for Categorization}

\setcounter{page}{275} %% This command is optional. 
                       %% May set page number only for first page in
                       %% issue, if desired.

%% <<== End of commands to be entered at Kluwers

%%  Authors, start here ==>>

\title{Explorations of an Incremental, Bayesian Algorithm
for Categorization}

\authors{John R. Anderson and Michael Matessa}
\email{ja\o s@andrew.cmu.edu}

\affil{Department of Psychology, Carnegie Mellon University, Pittsburgh,
PA 15213}

\editor{Dennis Kibler}

\abstract{An incremental categorization algorithm is described which,
at each step, assigns the next instance to the most probable
category. Etc.}

\keywords{Bayesian inference, concept learning, human learning, 
incremental algorithms}

\section{Introduction}
We have been engaged in a project to understand human categorization which
has let us to develop a machine learning algorithm. Our research began as an
exploration of the issue of whether human categorization can be considered
optimal.

To pursue the issue of whether human cognition is optimal requires
specifying two things. First we need a definition of optimality.
second, we need a specification of the structure of the environment
so we can determine what behavior is optimal in that environment.

\subsection{Preliminary definition of optimization}
Our assumption has been that the goal of categorization is to predict
unknown features of various objects that we encounter.

\subsubsection{The structure of the environment}
Our theory of the structure of the environment has been focused the
structure of living things (arguably, the largest portion of the
objects in the world) because of the aid biology gives in objectively
specifying the organization of these objects.

Formally, this amounts to calculating:
\begin{equation}
g_i(y|f)=\sum_x P(x|F_n)f_i(y|x)
\end{equation}
where $g_i(y|F_n)$ is the function specifying the probability an object will
display a value $y$ on a dimension $i$ given $F_n$ the observed feature
structure of all the objects.
\newpage
Here is an example of a wide equation:
\begin{wideequation}
\begin{equation}
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2
\end{equation}
\end{wideequation}

This is an example of a split equation,
moving the top half to the left
and the bottom to the right.
\begin{splitmath}
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2\\
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2
\end{splitmath}

\subsection{Footnote example}
Here is some text with footnotes in it.
Here is some text with footnotes in it.
Here is some text with footnotes in it.\footnote{This is a footnote.}

More text.\footnote{This is a second footnote.
This is a second footnote.
This is a second footnote.
This is a second footnote.
This is a second footnote.
This is a second footnote.} 
More text.\footnote{This is yet another footnote.
This is yet another footnote.
This is yet another footnote.
This is yet another footnote.
This is yet another footnote.
This is yet another footnote.}

\subsection{Indented text}
In an example satisfies the seed of a clause, then it satisfies the clause
as well. In addition, seeds have the following property:

\begin{itemize}
\item[] 
If a seed of clause $c_T$, and example {\bf x} satisfies $c_T$ but
not $c$, then {\bf x} has at least one attibute in $c_T$ that
is not in $c$.\hfill({\tt*})
\end{itemize}
The procedure below...

\subsection{Bulleted List}
Here is an example of a bulleted list:
\begin{itemize}
\item
Some text here. Some text here. Some text here. Some text here. 
Some text here.
Some text here. Some text here. Some text here. Some text here.

\item
Some text here. Some text here. Some text here.
\end{itemize}

\subsection{Numbered List}
Here is an example of a numbered list:
\begin{enumerate}
\item
Some text here. Some text here. Some text here. Some text here.

Some text here. Some text here. Some text here. Some text here. Some
text here.

\item
Some text here. Some text here. Some text here.

\begin{enumerate}
\item
Some text here. Some text here. Some text here. Some text here. Some
text here.  Some text here. Some text here. Some text here. Some text
here.

\item
Some text here. Some text here. Some text here.
\begin{enumerate}
\item
Some text here. Some text here. Some text here. Some text here. Some
text here.  Some text here. Some text here. Some text here. Some text
here.

\item
Some text here. Some text here. Some text here.
\end{enumerate}
\end{enumerate}
\end{enumerate}

\subsection{To Illustrate an Algorithm}
This is the command to use when you want to illustrate an algorithm
with some pseudo code. A backslash followed with a space will
indent the line. Every line will be printed
as it is seen on the screen. Blank lines will be preserved.
Math and font changes may be used.

The command
\verb+\bit+ will produce bold italics if you are using PostScript fonts, 
boldface in Computer Modern. \verb+\note{}+ will position the
note on the right margin. A backslash followed by a space
will provide a space a bit wider than the width of 2 `M's.

\begin{algorithm}
{\bit Evaluate-Single-FOE} ({\bf x$_f$, I$_0$, I$_1$}):
\ {\bf I}+ := {\bf I}$_1$;
\ ($\phi,\theta$) := (0,0);
\ {\it repeat}\note{/*usually only 1 interation required*/}
\ \ (s$_{opt}${\bf E}$_\eta$) := {\bit Optimal-Shift} ({\bf I$_0$,I$^+$,I$_0$,x$_f$});
\ \ ($\phi^+$, $\theta^+$) := {\bit Equivalent-Rotation} ({\bf s}$_{opt}$);
\ \ ($\phi$, $\theta$) := ($\phi$, $\theta$) + ($\phi^+$, $\theta^+$);
\ \ {\bf I}$^+$:= {\bit Derotate-Image} ({\bf I}$_1$, $\phi$, $\theta$);
\ \ {\it until} ($|\phi^+|\leq\phi_{max}$ \& $|\theta^+|\leq\theta_{max}$);
\ {\it return} ({\bf I}$^+$, $\phi$, $\theta$, E$_\eta$).
End pseudo-code.
\end{algorithm}

\newpage
\subsection{Figures}
Here is an example of a figure with .5 inch space left for
the illustration:
\begin{figure}[h]
\vspace*{.5in}
\caption{This is a figure caption.
This is a figure caption.
This is a figure caption.}
\end{figure}

\section{Making Tables}
Use caption on top of the table. 
Use \verb+\hline+ at the top of the table, underneath the column headers
and at the end of the table.

You are discouraged from using vertical lines in tables, but
it you must include vertical lines, you must also use 
\verb+\savehline+ instead of \verb+\hline+ or there will be a
gap between the vertical and horizontal lines.
(\verb+\hline+ has been redefined to add some vertical space above and
below it.)

The following form will spread out to the width of the page:

\begin{table}[h]
\caption{This is an example table caption. As you can
see, it will be as wide as the table that it captions.}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr}
\hline
$\alpha\beta\Gamma\Delta$ One&Two&Three\cr
\hline
one&two&three\cr
one&two&three\cr
\hline
\end{tabular*}
\end{table}

\begin{table}[h]
\caption{This is a table caption and will fit
the width of the table that it is captioning.}
\begin{tabular}{lcr}
\hline
$\alpha\beta\Gamma\Delta$ One&Two&Three\cr
\hline
one&two&three\cr
one&two&three\cr
\hline
\end{tabular}
\end{table}

\subsection{Unusual Tables}
There are three cases in which authors have felt the need to
make a table other than the plain `tabular' table in this style.
These more unusual tables are: tables with vertical lines;
tables that do not use `tabular'; and tables that have
a number of `tabular's nested. In this section we will show
how to make these tables.

\subsubsection{Vertical Lines}
Notice in the previous examples that no vertical lines were used.
If at all possible to make your meaning clear without vertical
lines, please leave them out. However, if you really must use
vertical lines, you must use \verb+\savehline+ instead of
\verb+\hline+ and
you must add another letter to the preamble.

For instance, this is the effect we usually want; 
horizontal lines extending to
the edge of the text in the column:

\begin{table}[h]
\begin{tabular}{c}\hline
 A         \\ \hline
 B         \\ \hline
\end{tabular}
\end{table}

Since we usually do not want vertical
lines in tables and we do want the horizontal lines to extend
exactly to the left and right of text, 
we must go to some extra efforts to get the vertical lines
to extend to the top and bottom of the column and to not
have extra horizontal space to the right of the vertical line
in the last column.
Remember to
use \verb+\savehline+ and to
to add another column entry to the table preamble
and then not to use that column, as seen below:

\begin{table}[h]
\begin{tabular}{|c|c}\savehline
 A         \\ \savehline
 B         \\ \savehline
\end{tabular}
\end{table}

\newpage

\subsubsection{Table Made Without `tabular'}
Here is an example of a table made without `tabular.' Notice
the caption must be written above the table.

\begin{specialtable}[h]
\caption{Training a Thermal Linear Machine.}
\label{thermal-alg}
\hrule
\vskip3pt
\begin{enumerate}
  \item Initialize $\beta$ to $2$.

  \item If linear machine is correct for all instances or 
$emag/lmag < \alpha$ for the last $2*n$ instances, then return
($n =$ the number of features).

  \item Otherwise, pass through the training instances once, and for each 
instance ${\bf Y}$ that	would be misclassified by the linear machine and 
for which $k < \beta$, immediately
	\begin{enumerate}
	  \item Compute correction $c = \frac{\beta^{2}}{\beta+k}$, and update
${\bf W}_{i}$ and  ${\bf W}_{j}$.

	  \item If the magnitude of the linear machine decreased on this
		adjustment, but increased on the previous adjustment, then
		anneal $\beta$ to $a \beta - b$. 
	\end{enumerate}

  \item Go to step 2.
\end{enumerate}
\vskip1pt
\hrule
\end{specialtable}

\subsubsection{Nesting `Tabular's}
The table macros are made to automatically format the
caption so that it is the width of the table and the caption
appears at the top of the table. This works fine as long
as you use the normal \verb+\begin{tabular}..+ 
to make your
table, and you don't nest another \verb+\begin{tabular}..+ with it. 
However, if you want to use some other way of formatting
your table, or if you want to nest one or more tabulars within
the table, you must start with \verb+\begin{specialtable}+, as you
see in the example below.
Notice that the caption is given at the top of the table.

\begin{specialtable}[h]
\caption{Quantitative and qualitative descriptions}
\begin{tabular}{ccc}& Quantitative & Qualitative \\ \hline & & \\(a) & 
    \begin{tabular}{rr} {\it Time} &
     {\it Level}\\ \hline 3 sec. & 0.01 m\\ 4 sec. & 0.23 m\\ 5 sec. &
     0.31 m\\ \hline
    \end{tabular} & 
           \begin{tabular}{c}$ Level(between(t_1, t_2)) = $\\ \ \ \ \ $
              (between(zero,top), increasing)$\\
           \end{tabular}\\& & \\(b) & $ Amount = 2.5 * 
                    Level + 0.7 * Level ^2 $ & $ Amount =M^+(Level) $\\
                    & & \\  \hline
\end{tabular}
\label{tabqq}
\end{specialtable}

\section{Theorems, Proofs, Examples, etc.}

\begin{example}
Consider an example in which $B_S$ is the structure...

The term $P(B_X)$ is our probability---prior to observing the data
in database $D$---that the data-generating process is a belief network with
structure $B_{S1}$.
\end{example}

\begin{proclaim}{Theorem 1}LEARN-MONOTONE-K-CNF makes at most
$(n+1)^K$ mistakes on any monotone K-CNF formula. (Recall that
n is the size of the largest example seen.
\end{proclaim}

\begin{proof}
Thus on each mistake, we decrease the cost by at least 1, and since the 
cost is never negative, the algorithm makes less than $(n+1)^K$
mistakes total.

The running time of this algorithm is clearly polynomial in 
{\bf size}$(f_T)$ and the length of the longest example seen.
\end{proof}

\begin{proof}
Thus on each mistake, we decrease the cost by at least 1, and since the 
cost is never negative, the algorithm makes less than $(n+1)^K$
mistakes total.

The running time of this algorithm is clearly polynomial in 
{\bf size}$(f_T)$ and the length of the longest example seen.

\[
\alpha\beta\Gamma\Delta\inmathqed
\]
\end{proof}

\begin{proof}[Proof of Theorem A.1]
This is a proof with a particular term. Call for the
particular term after `proof', i.e., 
\verb+\begin{proof}[Proof of Theorem A.1]+.
\end{proof}

In this section, we present an efficient formula for computing $P(B_X, D)$.
We do so by first introducing four assumptions.

\begin{demo}{Assumption 1}
The database variables, which we denote as $Z$, are discrete.
\end{demo}

As this assumption states...

\acknowledgements
We would like to thank....

Trying `cite', \cite{jacobs}, \cite{francis}.

\appendix{} 
% Please supply facing curly brackets if there is only one appendix.
% Supply a letter if you are using a number of appendices;
%% i.e., \appendix{A} ...\appendix{B}.

\appendixtitle{Sample Appendix Title} % optional appendix title

This is an appendix.
\begin{equation}
\sum_k P(k) \sum_i \sum_y f_i(y|k)^2
\end{equation}

\begin{references}
\bibitem{jacobs}Jacobs, E., ``Design Method Optimizes Scanning
 Phased Array,'' Microwaves, April 1982, pp.\ 69--70.

\bibitem{francis} Francis, M., ``Out-of-band response of array 
 antennas,'' Antenna Meas.  Tech. Proc., September 28--October 2,
1987, Seattle, p.~14.
\end{references}

For alphabetical references:

Maude Francis, (Francis, 1987) showed important new results
with array antennas.

\begin{alphareferences}
Francis, M., ``Out-of-band response of array 
antennas,'' Antenna Meas.  Tech. Proc., September 28--October 2,
1987, Seattle, p.~14.

Jacobs, E., ``Design Method Optimizes Scanning
Phased Array,'' Microwaves, April 1982, pp.\ 69--70.
\end{alphareferences}

\vskip12pt
You can also use Bibtex. See smjrnl.doc for documentation on 
using Bibtex.

%% Volume Table of Contents:

\begin{volumetoc}
\tocnumberline{Number 1}

\tocnumberline{Numbers 2/3 (Special Issue on Computational Learning
Theory)}

\tocnumberline{Number 4}

\TOCarticle{Explorations of an Incremental, Bayesian Algorithm for 
Categorization}{John R. Anderson and Michael Matessa}{275}

\TOCarticle{A Bayesian Method for the Induction of Probabilistic Networks
from Data}{Gregory F. Cooper and Edward Herskovits}{309}

\TOCarticle{Learning Boolean Functions in an Infinite Attribute Space}
{Avrim Blum}{373}

\TOCarticle{Technical Note: First Nearest Neighbor Classification on Frey 
and Slate's Letter Recognition Problem}{Terence C. Fogarty}{387}
\end{volumetoc}

%%%%  Issue Table of Contents %%%%

% \international % <== This command makes `An International Journal' appear
\begin{issuetoc}
\TOCarticle{Explorations of an Incremental, Bayesian Algorithm for 
Categorization}{John R. Anderson and Michael Matessa}{275}

\TOCarticle{A Bayesian Method for the Induction of Probabilistic Networks
from Data}{Gregory F. Cooper and Edward Herskovits}{309}

\TOCarticle{Learning Boolean Functions in an Infinite Attribute Space}
{Avrim Blum}{373}

\TOCarticle{Technical Note: First Nearest Neighbor Classification on Frey 
and Slate's Letter Recognition Problem}{Terence C. Fogarty}{387}

\end{issuetoc}

%%%% End of Issue Table of Contents %%%%

\end{article}
\end{document}

### 2017-02-22 pubs/demand_analysis/lex.yy.c

@@ -1,0 +1,2351 @@
+
+#line 3 "lex.yy.c"
+
+#define  YY_INT_ALIGNED short int
+
+/* A lexical scanner generated by flex */
+
+#define FLEX_SCANNER
+#define YY_FLEX_MAJOR_VERSION 2
+#define YY_FLEX_MINOR_VERSION 6
+#define YY_FLEX_SUBMINOR_VERSION 0
+#if YY_FLEX_SUBMINOR_VERSION > 0
+#define FLEX_BETA
+#endif
+
+/* First, we deal with  platform-specific or compiler-specific issues. */
+
+/* begin standard C headers. */
+#include <stdio.h>
+#include <string.h>
+#include <errno.h>
+#include <stdlib.h>
+
+/* end standard C headers. */
+
+/* flex integer type definitions */
+
+#ifndef FLEXINT_H
+#define FLEXINT_H
+
+/* C99 systems have <inttypes.h>. Non-C99 systems may or may not. */
+
+#if defined (__STDC_VERSION__) && __STDC_VERSION__ >= 199901L
+
+/* C99 says to define __STDC_LIMIT_MACROS before including stdint.h,
+ * if you want the limit (max/min) macros for int types. 
+ */
+#ifndef __STDC_LIMIT_MACROS
+#define __STDC_LIMIT_MACROS 1
+#endif
+
+#include <inttypes.h>
+typedef int8_t flex_int8_t;
+typedef uint8_t flex_uint8_t;
+typedef int16_t flex_int16_t;
+typedef uint16_t flex_uint16_t;
+typedef int32_t flex_int32_t;
+typedef uint32_t flex_uint32_t;
+#else
+typedef signed char flex_int8_t;
+typedef short int flex_int16_t;
+typedef int flex_int32_t;
+typedef unsigned char flex_uint8_t; 
+typedef unsigned short int flex_uint16_t;
+typedef unsigned int flex_uint32_t;
+
+/* Limits of integral types. */
+#ifndef INT8_MIN
+#define INT8_MIN               (-128)
+#endif
+#ifndef INT16_MIN
+#define INT16_MIN              (-32767-1)
+#endif
+#ifndef INT32_MIN
+#define INT32_MIN              (-2147483647-1)
+#endif
+#ifndef INT8_MAX
+#define INT8_MAX               (127)
+#endif
+#ifndef INT16_MAX
+#define INT16_MAX              (32767)
+#endif
+#ifndef INT32_MAX
+#define INT32_MAX              (2147483647)
+#endif
[…]

### 2017-02-22 pubs/demand_analysis/see.c

@@ -1,0 +1,5 @@
+#include <stdlib.h>
+
+main() {
+  printf("%i",'\\');
+}

### 2017-02-22 pubs/demand_analysis/new.verb.tex
test push

@@
 \begin{abstract}
-  {\it First-order\/} projection-based strictness analysis (properly,
+  {\it First-order\/} projection-based strictness analysis (generally,
   \emph{demand analysis}) was first introduced by Wadler and Hughes in
@@
   monomorphic languages and provided no proof of soundness.  Subsequent
-  developments extended the theory to other analyses, approaches to higher
-  order, and polymorphism.  Yet other developments were based on
-  generalization of projections, specifically partial equivalence relations
-  and partial projections, the latter notably simplifying a difficulty with
-  projection-based strictness dating back to Wadler and Hughes.  However, to
-  date no development has combined the desirable contributions of all of these
-  approaches.  We present the first practical higher-order, polymorphic,
+  developments extended the theory to other analyses (most notably
+  binding-time analysis), approaches to higher order, and polymorphism.  Yet
+  other developments were based on generalization of projections, specifically
+  partial equivalence relations and partial projections, the latter notably
+  simplifying a difficulty with projection-based strictness analysis dating
+  back to Wadler and Hughes.  However, to date no development has combined the
+  desirable attributes of all of these approaches.  We present the first
+  practical, higher-order, polymorphic,
   partial-projection/partial-equivalence-relation based demand analysis,
@@
   for such analyses.  Such an analysis can be used both to improve serial code
-  generation, as well as enable safe automatic parallelization, for lazy or
+  generation as well as enable safe automatic parallelization for lazy or
   non-strict functional languages.
@@
 
-We take as given that demand analysis is essential for good serial code
-generation, and for identifying where parallelism may be safely introduced.
-Numerous such analysis techniques have been proposed and implemented; we
-greatly narrow the field of discussion by restricting attention to those for
-which there is a formally stated notion of correctness that the technique has
-been proven to satisfy.
-
-Analysis techniques can usually be identified as being based on either
-a non-standard denotational semantics or a non-standard typing.
-Examples in the latter category include those of Gomard \cite{Go92},
-Jensen \cite{Jen92}, the Nielsons \cite{NN88}, Schmidt \cite{Sch88},
-and Henglein and Mossin \cite{HM94}.  Our focus is on those techniques
-based on non-standard interpretation, in particular, those using
-projections or partial equivalence relations (PERs) as the basic
-abstract values.
+We take as given that demand analysis, aka strictness analysis, is essential
+for good serial code generation, and for identifying where parallelism may be
+safely introduced.  An example of the former is the current ghc Haskell
+compiler, where strictness analysis is used to unbox values and thereby
+registerize code.  A potential application of the latter is in the context of
+the APPFL parallelizing, which assumes strict semantics by default for the
+purpose of exploring automatic parallelization techniques, but in the presence
+of non-strict semantics could use such an analysis to enable such
+parallelization.  Numerous such analysis techniques have been proposed and
+implemented; we greatly narrow the field of discussion by restricting
+attention to those for which there is a formally stated notion of correctness
+that the technique has been proven to satisfy.
+
+Analysis techniques can usually be identified as being based on either a
+non-standard denotational semantics or a non-standard typing.  Examples in the
+latter category include those of Gomard \cite{Go92}, Jensen \cite{Jen92}, the
+Nielsons \cite{NN88}, Schmidt \cite{Sch88}, and Henglein and Mossin
+\cite{HM94}.  Our focus is on those techniques based on non-standard
+interpretation, in particular, those using projections, partial projections, or
+partial equivalence relations (PERs) as the basic abstract values.
 
 A domain {\it projection\/} is a continuous idempotent function that
-approximates the identity.  Launchbury \cite{Lau88} hit upon the idea
+approximates the identity.  Wadler and Hughes first presented the idea of
+projection-based program analysis in the context of
+
+Launchbury \cite{Lau88} hit upon the idea
 of using projections to encode degrees of staticness of data.  The

### 2017-04-07 pubs/hcar16.2/AutoparallelizingPureFunc-KA.tex
new HCAR entry w/ pointer to Loren paper and hinting at David program analysis

--- 2016-04-29 pubs/hcar16/hcar16.tex
@@
-\documentclass[DIV16,twocolumn,10pt]{scrreprt}
-\usepackage{paralist}
-\usepackage{graphicx}
-\usepackage[final]{hcar}
-
-%include polycode.fmt
-
-\begin{document}
-
-\begin{hcarentry}{Auto-parallelizing Pure Functional Language System}
-\report{Kei Davis}
+% AutoparallelizingPureFunc-KA.tex
+\begin{hcarentry}[new]{Auto-parallelizing Pure Functional Language System}
+\report{Kei Davis}%05/16
 \status{active}
-\participants{Dean Prichard, David Ringo, Loren Anderson, Jacob Marks}
+\participants{Dean Prichard, David Ringo, Loren Anderson}
 \makeheader
-
-%(WHAT IS IT?)
 
@@
 polymorphic, pure functional language implementation in which we can
-experiment with automatic parallelization strategies, varying degrees of
+experiment with automatic parallelization strategies and varying degrees of
 default function and constructor strictness.  A secondary goal is to
 experiment with mechanisms for transparent fault tolerance.
-%
+
 We do not consider speculative or eager evaluation, or semantic strictness
@@
 Our approach is similar to that of the
-\href{https://dl.acm.org/citation.cfm?id=2503779}{Intel
-  Labs Haskell Research Compiler}: to use GHC as a front-end to generate STG
-(or Core), then exit to our own back-end compiler.  As in their case we do not
-attempt to use the GHC runtime.  Our implementation is \emph{light-weight} in
-that we are not attempting to support or recreate the vast functionality of
-GHC and its runtime.  This approach is also similar to
+\href{https://dl.acm.org/citation.cfm?id=2503779}{Intel Labs Haskell
+Research Compiler}, using GHC as a front-end to generate STG, then
+exit to our own back-end compiler.  As in their case we do not attempt to use
+the GHC runtime.  Our implementation is \emph{light-weight} in that we are not
+attempting to support or recreate the vast functionality of GHC and its
+runtime.  This approach is also similar to
 \href{http://www.cse.unsw.edu.au/~pls/thesis/dons-thesis.ps.gz}{Don Stewart's}
@@
 
-%(WHAT IS ITS STATUS? / WHAT HAS HAPPENED SINCE LAST TIME?)
 \subsubsection*{Current Status}
 Currently we have a fully functioning serial implementation and a primitive
-proof-of-design parallel implementation.
+proof-of-design parallel implementation.  The most recent major development
+was the ``bridge'' between GHC and our system.  Thus we can now compile and run
+Haskell programs with simple primitive and algebraic data types.
 
-%(CAN OTHERS GET IT?)
-
-%(WHAT ARE THE IMMEDIATE PLANS?)
 \subsubsection*{Immediate Plans}
 
-We are currently developing a more realistic parallel runtime.
-%
-Bridging the gap between GHC STG (or Core) to our STG representation will be
-undertaken starting June 2016.  An instrumentation framework will be developed
-in summer 2016.
-
-\FurtherReading
- A \href{https://github.com/losalamos/APPFL}{project web site} is under construction.
+We are currently developing a more realistic parallel runtime.  Tentatively
+the fault tolerance mechanism is scheduled as a Master's thesis project starting
+summer 2017.
 
@@
 graduate, with strong interests in Haskell programming, compiler/runtime
-development, and pursuing a spring, fall, or summer internship at Los
-Alamos National Laboratory, this could be for you.
+development, and pursuing a spring, fall, or summer internship at Los Alamos
+National Laboratory, this could be for you.
 
@@
 \emph{\bfseries The application process requires a bit of work so don't leave
-  enquiries until the last day/month.}\\
+enquiries until the last day/month.}  Dates for terms beyond summer 2017 are
+best guesses based on prior years.\\
 
-\begin{tabular}{l|l}
-Term & Application Deadline \\
+\begin{tabular}{l||l|l}
+Term & Application Opening & Deadline \\
 \hline
-Summer 2016  & Closed \\
-Fall 2016    & May 31, 2016 \\
-Spring 2017  & Approx. July 2016 \\
-Summer 2017  & Approx. January 2017 \\
-Fall 2017    & Approx. May 2017
+Summer 2017  & Open & Jan 2017 \\
+Fall 2017    & Jan 2017 & May 2017 \\
+Spring 2018  & May 2017 & Jul 2017 \\
 \end{tabular}\\
 
-\noindent Email me at kei (at) lanl (dot) gov if interested in more information, and
-feel free to pass this along. \\
+\noindent Email me at kei (at) lanl (dot) gov if interested in more
+information, and feel free to pass this along. \\
+
+\FurtherReading
+Email me as above for the Trends in Functional Programming 2016 paper
+about this project.
 
 \end{hcarentry}
-\end{document}

### 2017-04-07 pubs/hcar17.1/AutoparallelizingPureFunc-KA.tex
new HCAR entry w/ pointer to Loren paper and hinting at David program analysis

--- 2017-04-07 pubs/hcar16.2/AutoparallelizingPureFunc-KA.tex
@@
 % AutoparallelizingPureFunc-KA.tex
-\begin{hcarentry}[new]{Auto-parallelizing Pure Functional Language System}
-\report{Kei Davis}%05/16
+\begin{hcarentry}[updated]{Auto-parallelizing Pure Functional Language System}
+\report{Kei Davis}%11/16
 \status{active}
-\participants{Dean Prichard, David Ringo, Loren Anderson}
+\participants{Dean Prichard, David Ringo, Loren Anderson, Jacob Marks}
 \makeheader
@@
 The main project goal is the demonstration of a light-weight, higher-order,
-polymorphic, pure functional language implementation in which we can
-experiment with automatic parallelization strategies and varying degrees of
-default function and constructor strictness.  A secondary goal is to
-experiment with mechanisms for transparent fault tolerance.
+polymorphic, pure functional language implementation with which we can
+experiment with automatic parallelization strategies, varying degrees of
+default function and constructor strictness, and lightweight instrumentation.
 
-We do not consider speculative or eager evaluation, or semantic strictness
-inferred by program analysis, so potential parallelism is dictated by the
-specified degree of default strictness and any strictness annotations.
+We do not consider speculative or eager evaluation, but do plan to infer
+strictness by program analysis, so potential parallelism is dictated by the
+specified degree of default strictness, language constructs for parallelism,
+and program analysis.
 
@@
 \href{https://dl.acm.org/citation.cfm?id=2503779}{Intel Labs Haskell
-Research Compiler}, using GHC as a front-end to generate STG, then
+Research Compiler}: we use GHC as a front-end to generate STG, then
 exit to our own back-end compiler.  As in their case we do not attempt to use
@@
 proof-of-design parallel implementation. The most recent major development
-was the ``bridge'' between GHC and our system.  Thus we can now compile and run
+was the bridge between GHC and our system. Thus we can now compile and run
 Haskell programs with simple primitive and algebraic data types.
+
+Additionally, we have developed a new strictness analysis technique that will
+be incorporated into the compiler.  We'll miss the Static Analysis Symposium
+2017 submission deadline, unfortunately.
 
@@
 
-We are currently developing a more realistic parallel runtime.  Tentatively
-the fault tolerance mechanism is scheduled as a Master's thesis project starting
-summer 2017.
+We are currently developing a more realistic parallel runtime.
+The implementation of the strictness analysis is scheduled as a Master's thesis project
+starting May 2017.
 
@@
 \emph{\bfseries The application process requires a bit of work so don't leave
-enquiries until the last day/month.}  Dates for terms beyond summer 2017 are
+enquiries until the last day/month.}  Dates for terms beyond fall 2017 are
 best guesses based on prior years.\\
 
-\begin{tabular}{l||l|l}
-Term & Application Opening & Deadline \\
+\begin{tabular}{l||l||l}
+Term & Application & Deadline \\
+     & Opening &  \\
 \hline
-Summer 2017  & Open & Jan 2017 \\
-Fall 2017    & Jan 2017 & May 2017 \\
-Spring 2018  & May 2017 & Jul 2017 \\
+Fall 2017    & Open      & May 31 2017 \\
+Spring 2018  & July 2017 & Oct 2017 \\
+Summer 2018  & Oct 2017  & Jan 2018 \\
 \end{tabular}\\
 
-\noindent Email me at kei (at) lanl (dot) gov if interested in more
+\noindent Email kei (at) lanl (dot) gov if interested in more
 information, and feel free to pass this along. \\
@@
 \FurtherReading
-Email me as above for the Trends in Functional Programming 2016 paper
-about this project.
+Email same address as above for the Trends in Functional Programming 2016
+paper about this project.
 
+Intern Loren Anderson did an interesting Haskell exercise while here for this
+\href{http://ijmcs.future-in-tech.net/11.1/R-Anderson.pdf}{mathematics paper}.
 \end{hcarentry}

### 2017-04-17 pubs/demand_analysis/jlsc.tex

--- 2017-02-22 pubs/demand_analysis/new.verb.tex
@@
-\documentclass[11pt]{article}
-
-\usepackage{geometry}
-\geometry{textwidth=7.5in,
-  textheight=9in}
-
-%\usepackage{draftwatermark}
-%\SetWatermarkText{DRAFT}
-%\SetWatermarkScale{6}
-%\SetWatermarkColor{yellow}
+%% Kluwer Small Journal Article Sample File
+%% June 20, 1994
+
+%%   Use one of these two commands: 
+\documentstyle{smjrnl} % For Computer Modern Fonts, or
+%\documentstyle[smjfonts]{smjrnl} % for PostScript fonts
 
 \begin{document}
-
-\title{PERs from Partial Projections for Higher-order Polymorphic Demand Analysis}
-\author{Kei Davis\\
-  Los Alamos National Laboratory\\
-  kei at lanl.gov}
-\maketitle
+\begin{article}
+
+%%%%% To be entered at Kluwers: =====>>
+
+\journame{Lisp and Symbolic Computation}
+\volnumber{?}
+\issuenumber{?}
+\issuemonth{?}
+\volyear{1995}
+
+%% Do not delete either of the following two commands.
+%% Please supply facing curly brackets for the part you
+%% are not using for this article.
+%\received{May 1, 1991}\revised{}
+
+\authorrunninghead{K. Davis}
+\titlerunninghead{PERs from Projections for Binding-time Analysis}
 
@@
                        %% issue, if desired.
+
+%% <<== End of commands to be entered at Kluwers
+
+%%  Authors, start here ==>>
 
@@
 
-\com{\data}{{\it data}_{\stt}}
-\com{\for}{{\it for}_{\stt}}
-\com{\fac}{{\it fac}_{\stt}}
-\com{\unfac}{{\it unfac}_{\stt}}
+\com{\data}{{\it data}_{\subtypespace\stt}}
+\com{\for}{{\it for}_{\subtypespace\stt}}
+\com{\fac}{{\it fac}_{\subtypespace\stt}}
+\com{\unfac}{{\it unfac}_{\subtypespace\stt}}
 
@@
 \com{\tyip}{\theta_{\infty{}i+{\rm1}}}
-%\com{\th}{\theta}
+\com{\th}{\theta}
 
@@
 \com{\STR}{{\it STR}}
-\com{\Ab}{\alpha_{@@@b@@@}}
-\com{\Ax}{\alpha_{@@@x@@@}}
-\com{\Ay}{\alpha_{@@@y@@@}}
-\com{\Az}{\alpha_{@@@z@@@}}
-\com{\Axs}{\alpha_{@@@xs@@@}}
-\com{\Ays}{\alpha_{@@@ys@@@}}
-\com{\Azs}{\alpha_{@@@zs@@@}}
+\com{\Ab}{\alpha_{\subtypespace{\makebox{\subtypesize\tt b}} }}
+\com{\Ax}{\alpha_{\subtypespace{\makebox{\subtypesize\tt x}} }}
+\com{\Ay}{\alpha_{\subtypespace{\makebox{\subtypesize\tt y}} }}
+\com{\Az}{\alpha_{\subtypespace{\makebox{\subtypesize\tt z}} }}
+\com{\Axs}{\alpha_{\subtypespace{\makebox{\subtypesize\tt xs}} }}
+\com{\Ays}{\alpha_{\subtypespace{\makebox{\subtypesize\tt ys}} }}
+\com{\Azs}{\alpha_{\subtypespace{\makebox{\subtypesize\tt zs}} }}
 
@@
 
-\com{\oy}{\overline{@y@}}
-
-\com{\xnot}{{\rm@x@_0}}
-\com{\xone}{{\rm@x@_1}}
-\com{\xtwo}{{\rm@x@_2}}
-\com{\xthree}{{\rm@x@_3}}
-\com{\xxi}{{@x@_i}}
-\com{\xn}{{@x@_n}}
-
-\com{\ee}{@e@}
-\com{\ex}{@x@}
-\com{\evar}{@x@}
-\com{\enum}{@n@_i}
-\com{\esum}{\eone\ @+@\ \etwo}
-\com{\elam}{@\x.e@}
-\com{\ehlam}{@\#x.e@}
-\com{\elamt}{@\x:T.e@}
-\com{\ehlamt}{@\#x:T.e@}
+\com{\oy}{\overline{\mbox{\tt y}}}
+
+\com{\xnot}{{\rm\mbox{\tt x}_0}}
+\com{\xone}{{\rm\mbox{\tt x}_1}}
+\com{\xtwo}{{\rm\mbox{\tt x}_2}}
+\com{\xthree}{{\rm\mbox{\tt x}_3}}
+\com{\xxi}{{\mbox{\tt x}_i}}
+\com{\xn}{{\mbox{\tt x}_n}}
+
+\com{\ee}{\mbox{\tt e}}
+\com{\ex}{\mbox{\tt x}}
+\com{\evar}{\mbox{\tt x}}
+\com{\enum}{\mbox{\tt n}_i}
+\com{\esum}{\eone\ \mbox{\tt +}\ \etwo}
+\com{\elam}{\mbox{\tt {\char'134}x.e}}
+\com{\ehlam}{\mbox{\tt {\char'134}{\char'43}x.e}}
+\com{\elamt}{\mbox{\tt {\char'134}x:T.e}}
+\com{\ehlamt}{\mbox{\tt {\char'134}{\char'43}x:T.e}}
 \com{\eapp}{\eone~\etwo}
-\com{\ehapp}{@app# @\eone~\etwo}
-\com{\efapp}{@f e@}
-\com{\eunit}{@()@}
-\com{\etup}{@(@\eone@,@\ldots@,@\en@)@}
-\com{\esel}{@let (@\xone@,@\ldots@,@\xn@) =@\ \enot\ @in@\ \eone}
-\com{\econ}{@c@_i\ @e@}
-\com{\ecase}{@case@\ \enot\ @of @\cone\ \xone\ @->@\ \eone@;@\ \ldots@;@\ \cn\ \xn\ @->@\ \en}
-\com{\efix}{@fix e@}
-\com{\ehfix}{@fix# e@}
-\com{\estr}{@seq@~\eone~\etwo}
-
-\com{\enot}{{\rm@e@_0}}
-\com{\eone}{{\rm@e@_1}}
-\com{\etwo}{{\rm@e@_2}}
-\com{\ei}{{@e@_i}}
-\com{\en}{{@e@_n}}
-
-\com{\senot}{{\rm@@@e@@@_{\subtypesubstyle0}}}
-\com{\seone}{{\rm@@@e@@@_{\subtypesubstyle1}}}
-\com{\setwo}{{\rm@@@e@@@_{\subtypesubstyle2}}}
-\com{\sei}{{@@@e@@@_{\subtypesubstyle{}i}}}
-\com{\sen}{{@@@e@@@_{\subtypesubstyle{}n}}}
-\com{\setup}{{@@@(@@@\seone@@@,@@@\ldots@@@,@@@\sen@@@)@@@}}
-
-\com{\fnot}{{\rm@f@_0}}
-\com{\fone}{{\rm@f@_1}}
-\com{\ftwo}{{\rm@f@_2}}
-\com{\fii}{{@f@_i}}
-\com{\fn}{{@f@_n}}
+\com{\ehapp}{\mbox{\tt app{\char'43}\ }\eone~\etwo}
+\com{\efapp}{\mbox{\tt f\ e}}
+\com{\eunit}{\mbox{\tt ()}}
+\com{\etup}{\mbox{\tt (}\eone\mbox{\tt ,}\ldots\mbox{\tt ,}\en\mbox{\tt )}}
+\com{\esel}{\mbox{\tt let\ (}\xone\mbox{\tt ,}\ldots\mbox{\tt ,}\xn\mbox{\tt )\ =}\ \enot\ \mbox{\tt in}\ \eone}
+\com{\econ}{\mbox{\tt c}_i\ \mbox{\tt e}}
+\com{\ecase}{\mbox{\tt case}\ \enot\ \mbox{\tt of\ }\cone\ \xone\ \mbox{\tt ->}\ \eone\mbox{\tt ;}\ \ldots\mbox{\tt ;}\ \cn\ \xn\ \mbox{\tt ->}\ \en}
+\com{\efix}{\mbox{\tt fix\ e}}
+\com{\ehfix}{\mbox{\tt fix{\char'43}\ e}}
+\com{\estr}{\mbox{\tt seq}~\eone~\etwo}
+
+\com{\enot}{{\rm\mbox{\tt e}_0}}
+\com{\eone}{{\rm\mbox{\tt e}_1}}
+\com{\etwo}{{\rm\mbox{\tt e}_2}}
+\com{\ei}{{\mbox{\tt e}_i}}
+\com{\en}{{\mbox{\tt e}_n}}
+
+\com{\senot}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle0}}}
+\com{\seone}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle1}}}
+\com{\setwo}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle2}}}
+\com{\sei}{{{\makebox{\subtypesize\tt e}} _{\subtypesubstyle{}i}}}
+\com{\sen}{{{\makebox{\subtypesize\tt e}} _{\subtypesubstyle{}n}}}
+\com{\setup}{{{\makebox{\subtypesize\tt (}} \seone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \sen{\makebox{\subtypesize\tt )}} }}
+
+\com{\fnot}{{\rm\mbox{\tt f}_0}}
+\com{\fone}{{\rm\mbox{\tt f}_1}}
+\com{\ftwo}{{\rm\mbox{\tt f}_2}}
+\com{\fii}{{\mbox{\tt f}_i}}
+\com{\fn}{{\mbox{\tt f}_n}}
 
@@
 \com{\cn}{{{\tt{}c}_n}}
-\com{\anot}{{\rm@A@_0}}
-\com{\aone}{{\rm@A@_1}}
-\com{\atwo}{{\rm@A@_2}}
-\com{\ai}{{@A@_i}}
-\com{\aj}{{@A@_j}}
-\com{\an}{{@A@_n}}
-\com{\ta}{@A@}
-\com{\te}{@E@}
-\com{\tenot}{{@E@_{\it gl}}}
-\com{\teone}{{\rm@E@_1}}
-\com{\tei}{{@E@_i}}
-\com{\ten}{{@E@_n}}
-\com{\tnot}{{\rm@T@_0}}
-\com{\tone}{{\rm@T@_1}}
-\com{\ttwo}{{\rm@T@_2}}
-\com{\tthree}{{\rm@T@_3}}
-\com{\ti}{{@T@_i}}
-\com{\tj}{{@T@_j}}
-\com{\tn}{{@T@_n}}
-\com{\ttt}{@T@}
-\com{\unot}{{\rm@U@_0}}
-\com{\uone}{{\rm@U@_1}}
-\com{\utwo}{{\rm@U@_2}}
-\com{\uthree}{{\rm@U@_3}}
-\com{\ui}{{@U@_i}}
-\com{\un}{{@U@_n}}
-\com{\tunit}{@()@}
-\com{\tint}{@Int@}
-\com{\tfun}{\tone\ @->@\ \ttwo}
-\com{\tfunp}{@(@\tone\ @->@\ \ttwo@)@}
+\com{\anot}{{\rm\mbox{\tt A}_0}}
+\com{\aone}{{\rm\mbox{\tt A}_1}}
+\com{\atwo}{{\rm\mbox{\tt A}_2}}
+\com{\ai}{{\mbox{\tt A}_i}}
+\com{\aj}{{\mbox{\tt A}_j}}
+\com{\an}{{\mbox{\tt A}_n}}
+\com{\ta}{\mbox{\tt A}}
+\com{\te}{\mbox{\tt E}}
+\com{\tenot}{{\mbox{\tt E}_{\it gl}}}
+\com{\teone}{{\rm\mbox{\tt E}_1}}
+\com{\tei}{{\mbox{\tt E}_i}}
+\com{\ten}{{\mbox{\tt E}_n}}
+\com{\tnot}{{\rm\mbox{\tt T}_0}}
+\com{\tone}{{\rm\mbox{\tt T}_1}}
+\com{\ttwo}{{\rm\mbox{\tt T}_2}}
+\com{\tthree}{{\rm\mbox{\tt T}_3}}
+\com{\ti}{{\mbox{\tt T}_i}}
+\com{\tj}{{\mbox{\tt T}_j}}
+\com{\tn}{{\mbox{\tt T}_n}}
+\com{\ttt}{\mbox{\tt T}}
+\com{\unot}{{\rm\mbox{\tt U}_0}}
+\com{\uone}{{\rm\mbox{\tt U}_1}}
+\com{\utwo}{{\rm\mbox{\tt U}_2}}
+\com{\uthree}{{\rm\mbox{\tt U}_3}}
+\com{\ui}{{\mbox{\tt U}_i}}
+\com{\un}{{\mbox{\tt U}_n}}
+\com{\tunit}{\mbox{\tt ()}}
+\com{\tint}{\mbox{\tt Int}}
+\com{\tfun}{\tone\ \mbox{\tt ->}\ \ttwo}
+\com{\tfunp}{\mbox{\tt (}\tone\ \mbox{\tt ->}\ \ttwo\mbox{\tt )}}
 \com{\thfun}{\tone\ \harrow\ \ttwo}
-\com{\thfunp}{@(@\tone\ \harrow\ \ttwo@)@}
-\com{\harrow}{@#>@}
-\com{\tprod}{@(@\tone@,@\ldots@,@\tn@)@}
-\com{\teprod}{@(@\teone@,@\ldots@,@\ten@)@}
-\com{\tsum}{\cone\ \tone\ @+@\ \ldots\ @+@\ \cn\ \tn}
-\com{\tdefn}{\aone\ @=@\ \tone@;@\ \ldots@;@\ \an\ @=@\ \tn}
+\com{\thfunp}{\mbox{\tt (}\tone\ \harrow\ \ttwo\mbox{\tt )}}
+\com{\harrow}{\mbox{\tt {\char'43}>}}
+\com{\tprod}{\mbox{\tt (}\tone\mbox{\tt ,}\ldots\mbox{\tt ,}\tn\mbox{\tt )}}
+\com{\teprod}{\mbox{\tt (}\teone\mbox{\tt ,}\ldots\mbox{\tt ,}\ten\mbox{\tt )}}
+\com{\tsum}{\cone\ \tone\ \mbox{\tt +}\ \ldots\ \mbox{\tt +}\ \cn\ \tn}
+\com{\tdefn}{\aone\ \mbox{\tt =}\ \tone\mbox{\tt ;}\ \ldots\mbox{\tt ;}\ \an\ \mbox{\tt =}\ \tn}
 
@@
 
-\com{\scone}{{\rm@@@c@@@_1}}
-\com{\scn}{{@@@c@@@_n}}
-\com{\saone}{{\rm@@@A@@@_1}}
-\com{\san}{{@@@A@@@_n}}
-
-\com{\stenot}{{@@@E@@@_{\it\subtypesubstyle{}gl}}}
-\com{\sten}{{@@@E@@@_{\subtypesubstyle{}n}}}
-\com{\stnot}{{@@@T@@@_{\rm\subtypesubstyle0}}}
-\com{\stone}{{@@@T@@@_{\rm\subtypesubstyle1}}}
-\com{\sttwo}{{@@@T@@@_{\rm\subtypesubstyle2}}}
-\com{\sti}{{@@@T@@@_{\subtypesubstyle{}i}}}
-\com{\stn}{{@@@T@@@_{\subtypesubstyle{}n}}}
-
-\com{\sta}{{@@@A@@@}}
-\com{\ste}{{@@@E@@@}}
-\com{\stt}{{@@@T@@@}}
-\com{\stu}{{@@@U@@@}}
-\com{\std}{{@@@D@@@}}
-\com{\stunit}{{@@@()@@@}}
-\com{\stint}{{@@@Int@@@}}
-\com{\stbool}{{@@@Bool@@@}}
-\com{\stintlist}{{@@@IntList@@@}}
-\com{\stfun}{{\stone@@@->@@@\sttwo}}
-\com{\sthfun}{{\stone@@@#>@@@\sttwo}}
-\com{\stprod}{{@@@(@@@\stone@@@,@@@\ldots@@@,@@@\stn@@@)@@@}}
-\com{\steprod}{{@@@(@@@\steone@@@,@@@\ldots@@@,@@@\sten@@@)@@@}}
-\com{\stsum}{{\scone\ \stone\ @@@+@@@\ \ldots\ @@@+@@@\ \scn\ \stn}}
-\com{\stdefn}{{\saone\ @@@=@@@\ \stone@@@;@@@\ \ldots@@@;@@@\ \san\ @@@=@@@\ \stn}}
+\com{\scone}{{\rm{\makebox{\subtypesize\tt c}} _1}}
+\com{\scn}{{{\makebox{\subtypesize\tt c}} _n}}
+\com{\saone}{{\rm{\makebox{\subtypesize\tt A}} _1}}
+\com{\san}{{{\makebox{\subtypesize\tt A}} _n}}
+
+\com{\stenot}{{{\makebox{\subtypesize\tt E}} _{\it\subtypesubstyle{}gl}}}
+\com{\sten}{{{\makebox{\subtypesize\tt E}} _{\subtypesubstyle{}n}}}
+\com{\stnot}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle0}}}
+\com{\stone}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle1}}}
+\com{\sttwo}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle2}}}
+\com{\sti}{{{\makebox{\subtypesize\tt T}} _{\subtypesubstyle{}i}}}
+\com{\stn}{{{\makebox{\subtypesize\tt T}} _{\subtypesubstyle{}n}}}
+
+\com{\sta}{{{\makebox{\subtypesize\tt A}} }}
+\com{\ste}{{{\makebox{\subtypesize\tt E}} }}
+\com{\stt}{{{\makebox{\subtypesize\tt T}} }}
+\com{\stu}{{{\makebox{\subtypesize\tt U}} }}
+\com{\std}{{{\makebox{\subtypesize\tt D}} }}
+\com{\stunit}{{{\makebox{\subtypesize\tt ()}} }}
+\com{\stint}{{{\makebox{\subtypesize\tt Int}} }}
+\com{\stbool}{{{\makebox{\subtypesize\tt Bool}} }}
+\com{\stintlist}{{{\makebox{\subtypesize\tt IntList}} }}
+\com{\stfun}{{\stone{\makebox{\subtypesize\tt ->}} \sttwo}}
+\com{\sthfun}{{\stone{\makebox{\subtypesize\tt {\char'43}>}} \sttwo}}
+\com{\stprod}{{{\makebox{\subtypesize\tt (}} \stone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \stn{\makebox{\subtypesize\tt )}} }}
+\com{\steprod}{{{\makebox{\subtypesize\tt (}} \steone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \sten{\makebox{\subtypesize\tt )}} }}
+\com{\stsum}{{\scone\ \stone\ {\makebox{\subtypesize\tt +}} \ \ldots\ {\makebox{\subtypesize\tt +}} \ \scn\ \stn}}
+\com{\stdefn}{{\saone\ {\makebox{\subtypesize\tt =}} \ \stone{\makebox{\subtypesize\tt ;}} \ \ldots{\makebox{\subtypesize\tt ;}} \ \san\ {\makebox{\subtypesize\tt =}} \ \stn}}
 
@@
 
-\com{\ProjT}{{\it Proj}_{\stt}}
-\com{\ProjU}{{\it Proj}_{\stu}}
-\com{\SProjT}{{\it SProj}_{\stt}}
-\com{\SProjU}{{\it SProj}_{\stu}}
-\com{\SProjE}{{\it SProj}_{\ste}}
-\com{\SAbsT}{{\it SAbs}_{\stt}}
-\com{\FProjT}{{\it FProj}_{\stt}}
-\com{\FProjU}{{\it FProj}_{\stu}}
-\com{\FProjE}{{\it FProj}_{\ste}}
-\com{\FAbsT}{{\it FAbs}_{\stt}}
-\com{\FAbsFT}{{\it FAbsF_{\stt}}}
-\com{\FTranT}{{\it FTran}_{\stt}}
-
-\com{\Eprod}{@(@\Eone@,@\ldots@,@\En@)@}
-\com{\Ei}{{@E@_i}}
-\com{\Eone}{{\rm@E@_1}}
-\com{\En}{{@E@_n}}
+\com{\ProjT}{{\it Proj}_{\subtypespace\stt}}
+\com{\ProjU}{{\it Proj}_{\subtypespace\stu}}
+\com{\SProjT}{{\it SProj}_{\subtypespace\stt}}
+\com{\SProjU}{{\it SProj}_{\subtypespace\stu}}
+\com{\SProjE}{{\it SProj}_{\subtypespace\ste}}
+\com{\SAbsT}{{\it SAbs}_{\subtypespace\stt}}
+\com{\FProjT}{{\it FProj}_{\subtypespace\stt}}
+\com{\FProjU}{{\it FProj}_{\subtypespace\stu}}
+\com{\FProjE}{{\it FProj}_{\subtypespace\ste}}
+\com{\FAbsT}{{\it FAbs}_{\subtypespace\stt}}
+\com{\FAbsFT}{{\it FAbsF_{\subtypespace\stt}}}
+\com{\FTranT}{{\it FTran}_{\subtypespace\stt}}
+
+\com{\Eprod}{\mbox{\tt (}\Eone\mbox{\tt ,}\ldots\mbox{\tt ,}\En\mbox{\tt )}}
+\com{\Ei}{{\mbox{\tt E}_i}}
+\com{\Eone}{{\rm\mbox{\tt E}_1}}
+\com{\En}{{\mbox{\tt E}_n}}
 
@@
 \com{\Glbio}{\Glb_{i\geq\rm0}}
-\com{\mkint}[1]{{\it{}mkint}^{#1}_{{\cal{}N}\br{@@@n@@@}}}
+\com{\mkint}[1]{{\it{}mkint}^{#1}_{{\cal{}N}\br{{\makebox{\subtypesize\tt n}} }}}
 \rcom{\u}[1]{\acute{#1}}
@@
 
-\com{\ve}{\vec{@e@}}
-\com{\vx}{\vec{@x@}}
-\com{\vy}{\vec{@y@}}
-\com{\vT}{\vec{@T@}}
+\com{\ve}{\vec{\mbox{\tt e}}}
+\com{\vx}{\vec{\mbox{\tt x}}}
+\com{\vy}{\vec{\mbox{\tt y}}}
+\com{\vT}{\vec{\mbox{\tt T}}}
 \com{\vF}{\vec{F}}
@@
 
-\com{\fxge}{\mbox{@f x = @$\cdots$@(g e)@$\cdots$}}
+\com{\fxge}{\mbox{\mbox{\tt f\ x\ =\ }$\cdots$\mbox{\tt (g\ e)}$\cdots$}}
 \com{\C}{\mbox{$Dom$}}
@@
 
-\begin{abstract}
-  {\it First-order\/} projection-based strictness analysis (properly,
-  \emph{demand analysis}) was first introduced by Wadler and Hughes in
-  1987~\cite{WH87}.  This development was restricted to first-order,
-  monomorphic languages and provided no proof of soundness.  Subsequent
-  developments extended the theory to other analyses, approaches to higher
-  order, and polymorphism.  Yet other developments were based on
-  generalization of projections, specifically partial equivalence relations
-  and partial projections, the latter notably simplifying a difficulty with
-  projection-based strictness dating back to Wadler and Hughes.  However, to
-  date no development has combined the desirable contributions of all of these
-  approaches.  We present the first practical higher-order, polymorphic,
-  partial-projection/partial-equivalence-relation based demand analysis,
-  together with a proof of satisfaction of the fundamental safety condition
-  for such analyses.  Such an analysis can be used both to improve serial code
-  generation, as well as enable safe automatic parallelization, for lazy or
-  non-strict functional languages.
-
-% has proven genuinely useful in partial evaluation
-% \cite{Lau91a}.  There have been three notable generalisations
-% of projection-based analysis to higher order.  The first lacked a
-% formal basis \cite{Mog89}; the second used structures strictly more
-% general than projections, namely {\it partial equivalence relations\/}
-% (PERs) \cite{HS91}; the third involved a complex construction that
-% gave rise to impractically large abstract domains \cite{Dav93b}.  This
-% paper presents a technique free of these shortcomings: it is simple,
-% entirely projection-based, satisfies a formal correctness condition,
-% and gives rise to reasonably small abstract domains.  Though the
-% technique is cast in terms of projections, there is also an
-% interpretation in terms of PERs.  The principal limitation of the
-% technique is the restriction to {\it monomorphic\/} typing.
-\end{abstract}
-
-%\keywords{Binding-time analysis, abstract interpretation, projections, PERs.}
+\title{PERs from Projections for Binding-time Analysis}
+
+\authors{Kei Davis}
+\email{kei@crl.nmsu.edu}
+
+\affil{Center for Non-Linear Studies \\
+Mail Stop B258 \\
+Los Alamos National Laboratory \\
+Los Alamos, NM  87545, U.S.A.}
+
+%\editor{Dennis Kibler}
+
+\abstract{{\it First-order\/} projection-based binding-time analysis
+has proven genuinely useful in partial evaluation
+\cite{Lau91a}.  There have been three notable generalisations
+of projection-based analysis to higher order.  The first lacked a
+formal basis \cite{Mog89}; the second used structures strictly more
+general than projections, namely {\it partial equivalence relations\/}
+(PERs) \cite{HS91}; the third involved a complex construction that
+gave rise to impractically large abstract domains \cite{Dav93b}.  This
+paper presents a technique free of these shortcomings: it is simple,
+entirely projection-based, satisfies a formal correctness condition,
+and gives rise to reasonably small abstract domains.  Though the
+technique is cast in terms of projections, there is also an
+interpretation in terms of PERs.  The principal limitation of the
+technique is the restriction to {\it monomorphic\/} typing.}
+
+\keywords{Binding-time analysis, abstract interpretation, projections,
+PERs.}
 
@@
 
-We take as given that demand analysis is essential for good serial code
-generation, and for identifying where parallelism may be safely introduced.
-Numerous such analysis techniques have been proposed and implemented; we
-greatly narrow the field of discussion by restricting attention to those for
-which there is a formally stated notion of correctness that the technique has
-been proven to satisfy.
+We take as given that binding-time analysis is essential for good
+partial evaluation, and we do not address the issue of annotating
+programs according to the results of analysis.  Numerous binding-time
+analysis techniques have been proposed and implemented; we greatly
+narrow the field of discussion by restricting attention to those for
+which there is a formally stated notion of correctness that the
+technique has been proven to satisfy.
 
@@
 operation on binary relations, so 
-"" 
-	(\G\o\f\we\f\o\D)\
-	\Leftrightarrow\ 
-	(\forall u,v\ .\ \D\ u=\D\ v\ \Rightarrow\ \G\ (f\ u)=\G\ (f\ v))\ .  
-""
+\beqs
+\it (\G\o\f\we\f\o\D)\Leftrightarrow(\forall\ u,v\ .\ \D\ u=\D\ v\ \Rightarrow\ \G\ (f\ u)=\G\ (f\ v))\ .
+\eeqs
 Then, for example, $(\BOT\fto\ID)~(f,f)$ asserts
@@
 analysis semantics was introduced.  Moving from the standard to
-intermediate semantics involved a translation of each ground type @T@
-to a function type @E@\ @->@\ @T@ (for a fixed type @E@), the values
+intermediate semantics involved a translation of each ground type \mbox{\tt T}
+to a function type \mbox{\tt E}\ \mbox{\tt ->}\ \mbox{\tt T} (for a fixed type \mbox{\tt E}), the values
 of which, being functions, could then be abstracted.  The result,
@@
 \fig{tb}{Types and type definitions.}{fig:types}{
-""
-@T@	&::=&	\ta			&\reason{Type Name}
-	&|&	\tint			&\reason{Integer}
-	&|&	\tprod			&\reason{Product, $n\geq0$}
-	&|&	\tsum			&\reason{Sum, $n\geq1$} 
-	&|&	\tfun			&\reason{Function} \espace 
-@D@	&::=&	\tdefn			&\reason{Type Definitions}
-""
+\beqs
+\it \makebox{\tt T}&\it ::=&\it \ta&\it \reason{Type\ Name}\\
+\it &\it |&\it \tint&\it \reason{Integer}\\
+\it &\it |&\it \tprod&\it \reason{Product,\ $n\geq{\rm0}$}\\
+\it &\it |&\it \tsum&\it \reason{Sum,\ $n\geq{\rm1}$}\\
+\it &\it |&\it \tfun&\it \reason{Function}\ \espace\\
+\it \makebox{\tt D}&\it ::=&\it \tdefn&\it \reason{Type\ Definitions}
+\eeqs
 }
 Nullary product corresponds to the so-called {\it unit\/} type.  A
-unary product @(T)@ will always have the same semantics as @T@.  The
+unary product \mbox{\tt (T)} will always have the same semantics as \mbox{\tt T}.  The
 types used in the examples are defined as follows.
-@@@
-     FunList =   nil () 
-               + cons (Int -> Int, FunList) ,
-
-     FunTree =   leaf (Int -> Int) 
-	       + branch (FunTree, FunTree) .
-@@@
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ FunList\ =\ \ \ nil\ ()\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ +\ cons\ (Int\ ->\ Int,\ FunList)\ ,\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt }\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ FunTree\ =\ \ \ leaf\ (Int\ ->\ Int)\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt 	\ \ \ \ \ \ \ +\ branch\ (FunTree,\ FunTree)\ .}\vspace{-0.0em}
+\end{flushleft}
 
@@
 Addition for integers is provided as typical of operations on flat data
-types in this setting.  A unary tuple @(e)@ will always have the same
-semantics as @e@.  The (monomorphic) typing of expressions is entirely
+types in this setting.  A unary tuple \mbox{\tt (e)} will always have the same
+semantics as \mbox{\tt e}.  The (monomorphic) typing of expressions is entirely
 standard and is omitted.
 \fig{tb}{Expressions.}{fig:exprs}{
-""
-@e@	&::=&	\evar				&\reason{Variable} 
-	&|&	\enum				&\reason{Numeral}
-	&|&	\esum				&\reason{Integer addition}
-	&|&	\eapp				&\reason{Function application}
-	&|&	\etup				&\reason{Tuple construction}
-	&|&	@let@\ @(@\xone@,@\ldots@,@\xn@)@\ @=@\ \enot\ @in@\ \eone			&\reason{Tuple decomposition}
-	&|&	\econ				&\reason{Sum construction}
-	&|&	@case@\ \enot\ @of @\{\ci\ \xxi\ @->@\ \ei\}	&\reason{Sum decomposition}
-	&|&	\elamt				&\reason{Lambda abstraction}
-	&|&	\eapp				&\reason{Function application}
-	&|&	\efix				&\reason{Fixed point}
-""
+\beqs
+\it \makebox{\tt e}&\it ::=&\it \evar&\it \reason{Variable}\\
+\it &\it |&\it \enum&\it \reason{Numeral}\\
+\it &\it |&\it \esum&\it \reason{Integer\ addition}\\
+\it &\it |&\it \eapp&\it \reason{Function\ application}\\
+\it &\it |&\it \etup&\it \reason{Tuple\ construction}\\
+\it &\it |&\it \makebox{\tt let}\ \makebox{\tt (}\xone\makebox{\tt ,}\ldots\makebox{\tt ,}\xn\makebox{\tt )}\ \makebox{\tt =}\ \enot\ \makebox{\tt in}\ \eone&\it \reason{Tuple\ decomposition}\\
+\it &\it |&\it \econ&\it \reason{Sum\ construction}\\
+\it &\it |&\it \makebox{\tt case}\ \enot\ \makebox{\tt of\ }\{\ci\ \xxi\ \makebox{\tt ->}\ \ei\}&\it \reason{Sum\ decomposition}\\
+\it &\it |&\it \elamt&\it \reason{Lambda\ abstraction}\\
+\it &\it |&\it \eapp&\it \reason{Function\ application}\\
+\it &\it |&\it \efix&\it \reason{Fixed\ point}
+\eeqs
 }
@@
 free variable environment.  Then for both versions of the semantics
-and all expressions @e@ of type @T@ with environment of type $\tprod$,
-""
-	\Ee{}{@e@}\ \in\ \Te{}{\tprod}~\fto~\Te{}{@T@}\ .
-""
+and all expressions \mbox{\tt e} of type \mbox{\tt T} with environment of type $\tprod$,
+\beqs
+\it \Ee{}{\makebox{\tt e}}\ \in\ \Te{}{\tprod}~\fto~\Te{}{\makebox{\tt T}}\ .
+\eeqs
 Noting that $\R\br{\xxi}$ is short for $\itt{sel}_i~\R$, environment
 update $\R[\xxi\mapsto{}v]$ is defined by
-"""
-	tuple (sel_1 \R, \ldots, sel_{i-1} \R, v, sel_{i+1} \R, \ldots, sel_n \R) .
-"""
+\beqqs
+\it tuple\ (sel_{\rm1}\ \R,\ \ldots,\ sel_{i-{\rm1}}\ \R,\ v,\ sel_{i+{\rm1}}\ \R,\ \ldots,\ sel_n\ \R)\ .
+\eeqqs
 
 \fig{tb}{Parameterised semantics.}{fig:gensem}{
-""
-	\Ee{}{\evar_i}\ \R = \R\br{@x@_i} = \itt{sel}_i\ \R\espace\espace
-%	\Ee{}{\eunit}\ \R = \itt{mkunit}\ \R\espace\espace
-%	\Ee{}{\enum}\ \R = \itt{mkint}_i\espace\espace
-	\Ee{}{\esum}\ \R = \itt{plus}~(\Ee{}{\eone}\ \R,\ \Ee{}{\etwo}\ \R)\espace\espace
-	\Ee{}{\etup}\ \R = \itt{tuple}~(\Ee{}{\eone}\ \R,~\ldots,~\Ee{}{\en}\ \R)\espace\espace
-	\Ee{}{\esel}\ \R
-	\ \ \ \ \ = \Ee{}{\eone}\ \R[\xxi\mapsto\itt{sel}_i\ (\Ee{}{\enot}\ \R)~|~\oin]\espace\espace
-	\Ee{}{\econ}\ \R = \itt{inc}_i\ (\Ee{}{@e@}\ \R)\espace\espace
-	\Ee{}{\ecase}\ \R
-	\ \ \ \ \  = \itt{choose}~(\ba[t]{l}\Ee{}{\enot}\ \R,
-	\Ee{}{\eone}\ \R[\xone\mapsto\itt{outc}_1\ (\Ee{}{\enot}\ \R)],
-	\ \ \ ~\vdots
-	\Ee{}{\en}\ \R[\xn\mapsto\itt{outc}_n\ (\Ee{}{\enot}\ \R)])\ea\espace\espace
-	\Ee{}{\elam}\ \R = \itt{mkfun}\ (\L{}x.\Ee{}{@e@}\ \R[\evar\mapsto{}x])\espace\espace
-	\Ee{}{\eapp}\ \R = \itt{apply}\ (\Ee{}{\eone}\ \R)\ (\Ee{}{\etwo}\ \R)\espace\espace
-	\Ee{}{\efix}\ \R = (\itt{fix}\o\itt{apply})\ (\Ee{}{@e@}\ \R)
-""
+\beqs
+\it \Ee{}{\evar_i}\ \R\ =\ \R\br{\makebox{\tt x}_i}\ =\ \itt{sel}_i\ \R\espace\espace\\
+\it %\ \Ee{}{\eunit}\ \R\ =\ \itt{mkunit}\ \R\espace\espace\\
+\it %\ \Ee{}{\enum}\ \R\ =\ \itt{mkint}_i\espace\espace\\
+\it \Ee{}{\esum}\ \R\ =\ \itt{plus}~(\Ee{}{\eone}\ \R,\ \Ee{}{\etwo}\ \R)\espace\espace\\
+\it \Ee{}{\etup}\ \R\ =\ \itt{tuple}~(\Ee{}{\eone}\ \R,~\ldots,~\Ee{}{\en}\ \R)\espace\espace\\
+\it \Ee{}{\esel}\ \R\\
+\it \ \ \ \ \ =\ \Ee{}{\eone}\ \R[\xxi\mapsto\itt{sel}_i\ (\Ee{}{\enot}\ \R)~|~\oin]\espace\espace\\
+\it \Ee{}{\econ}\ \R\ =\ \itt{inc}_i\ (\Ee{}{\makebox{\tt e}}\ \R)\espace\espace\\
+\it \Ee{}{\ecase}\ \R\\
+\it \ \ \ \ \ \ =\ \itt{choose}~(\ba[t]{l}\Ee{}{\enot}\ \R,\\
+\it \Ee{}{\eone}\ \R[\xone\mapsto\itt{outc}_{\rm1}\ (\Ee{}{\enot}\ \R)],\\
+\it \ \ \ ~\vdots\\
+\it \Ee{}{\en}\ \R[\xn\mapsto\itt{outc}_n\ (\Ee{}{\enot}\ \R)])\ea\espace\espace\\
+\it \Ee{}{\elam}\ \R\ =\ \itt{mkfun}\ (\L{}x.\Ee{}{\makebox{\tt e}}\ \R[\evar\mapsto{}x])\espace\espace\\
+\it \Ee{}{\eapp}\ \R\ =\ \itt{apply}\ (\Ee{}{\eone}\ \R)\ (\Ee{}{\etwo}\ \R)\espace\espace\\
+\it \Ee{}{\efix}\ \R\ =\ (\itt{fix}\o\itt{apply})\ (\Ee{}{\makebox{\tt e}}\ \R)
+\eeqs
 }
@@
 $\itt{lift}~f$ for some $f$.  (This can be made even more explicit by
-introducing a unary sum type, whence "\L" becomes a constructor \cite{Dav94}.)
+introducing a unary sum type, whence ${\it \L}$ becomes a constructor \cite{Dav94}.)
 Recursive type definitions give rise to recursive domain
@@
 \fig{tb}{Standard type and expression semantics.}{fig:ssem}{
-"""
-\Te{\Ss}{\tint} = \Int = \lift{\bf{Z}} \espace\espace
-\Te{\Ss}{\tprod} = \Te{\Ss}{\tone}\ \times\ \ldots\ \times\ \Te{\Ss}{\tn}\espace\espace
-\Te{\Ss}{\tsum} = \Te{\Ss}{\tone}\ \plus\ \ldots\ \plus\ \Te{\Ss}{\tn}\espace\espace
-\Te{\Ss}{\tfun} = \liftp{\Te{\Ss}{\tone} \fto \Te{\Ss}{\ttwo}}
-
-%	\itt{mkunit}^{\Ss}\ \R			 = 	()\espace\espace
-%	\itt{mkint}^{\Ss}_n\ \R			 = 	\itt{lift}\ n\espace\espace
-	\itt{plus}^{\Ss}\ (x,y) 		 = 	x+y\espace\espace
-	\itt{tuple}^{\Ss}\ (x_1,\ldots,x_n)	 = 	(x_1,\ldots,x_n)\espace\espace
-	\itt{sel}_i^{\,\Ss}\ (x_1,\ldots,x_n)	 = 	x_i\espace\espace
-	\itt{inc}_i^{\Ss} 			 = 	\itt{in}_i\o\itt{lift}\espace\espace
-	\itt{outc}_i^{\Ss}\ x			 = 	\itt{drop}\o\itt{out}_i\espace\espace
-	\ba{lll}\itt{choose}^{\Ss}\ (\bot,&\ x_1,\ldots,x_n)		 = \bot
-	\itt{choose}^{\Ss}\ (\itt{in}_i~v,&\ x_1,\ldots,x_n)	 = x_i\ea\espace\espace
-	\itt{mkfun}^{\Ss}		 = 	\itt{lift}\espace\espace
-	\itt{apply}^{\Ss}		 = 	\itt{drop}\espace\espace
-	\itt{fix}^{\Ss}  		 = 	\itt{lfp} \reason{Least fixed point}
-"""
+\beqqs
+\it \Te{\Ss}{\tint}\ =\ \Int\ =\ \lift{\bf{Z}}\ \espace\espace\\
+\it \Te{\Ss}{\tprod}\ =\ \Te{\Ss}{\tone}\ \times\ \ldots\ \times\ \Te{\Ss}{\tn}\espace\espace\\
+\it \Te{\Ss}{\tsum}\ =\ \Te{\Ss}{\tone}\ \plus\ \ldots\ \plus\ \Te{\Ss}{\tn}\espace\espace\\
+\it \Te{\Ss}{\tfun}\ =\ \liftp{\Te{\Ss}{\tone}\ \fto\ \Te{\Ss}{\ttwo}}\\
+\it \\
+\it %\ \itt{mkunit}^{\Ss}\ \R\ =\ ()\espace\espace\\
+\it %\ \itt{mkint}^{\Ss}_n\ \R\ =\ \itt{lift}\ n\espace\espace\\
+\it \itt{plus}^{\Ss}\ (x,y)\ =\ x+y\espace\espace\\
+\it \itt{tuple}^{\Ss}\ (x_{\rm1},\ldots,x_n)\ =\ (x_{\rm1},\ldots,x_n)\espace\espace\\
+\it \itt{sel}_i^{\,\Ss}\ (x_{\rm1},\ldots,x_n)\ =\ x_i\espace\espace\\
+\it \itt{inc}_i^{\Ss}\ =\ \itt{in}_i\o\itt{lift}\espace\espace\\
+\it \itt{outc}_i^{\Ss}\ x\ =\ \itt{drop}\o\itt{out}_i\espace\espace\\
+\it \ba{lll}\itt{choose}^{\Ss}\ (\bot,&\it \ x_{\rm1},\ldots,x_n)\ =\ \bot\\
+\it \itt{choose}^{\Ss}\ (\itt{in}_i~v,&\it \ x_{\rm1},\ldots,x_n)\ =\ x_i\ea\espace\espace\\
+\it \itt{mkfun}^{\Ss}\ =\ \itt{lift}\espace\espace\\
+\it \itt{apply}^{\Ss}\ =\ \itt{drop}\espace\espace\\
+\it \itt{fix}^{\Ss}\ =\ \itt{lfp}\ \reason{Least\ fixed\ point}
+\eeqqs
 }
@@
 applicable {\it forward\/} parts.  For example, for expressions of
-function type ${@T@\ @->@\ @U@}$, with
-$\liftp{T\fto{}U}=\Te{\Ss}{@T@\ @->@\ @U@}$, we distinguish two
+function type ${\mbox{\tt T}\ \mbox{\tt ->}\ \mbox{\tt U}}$, with
+$\liftp{T\fto{}U}=\Te{\Ss}{\mbox{\tt T}\ \mbox{\tt ->}\ \mbox{\tt U}}$, we distinguish two
 degrees of definedness---between values of the form
@@
 domain $\lone=\{\bot, \itt{lift}~()\}$---the {\it data domain\/} corresponding
-to type ${@T@\ @->@\ @U@}$.  There are two distinct projections on $\lone$,
+to type ${\mbox{\tt T}\ \mbox{\tt ->}\ \mbox{\tt U}}$.  There are two distinct projections on $\lone$,
 namely $\ID$ and $\BOT$: denoted functions are 
@@
 is $\lone\times(T\fto{}U)$.  The factorisation function
-${\it{}fac}_{@@@T->U@@@}\in\liftp{T\fto{}U}\fto(\lone\times(T\fto{}U))$
+${\it{}fac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }\in\liftp{T\fto{}U}\fto(\lone\times(T\fto{}U))$
 is an embedding defined by
-${\it{}fac}_{@@@T->U@@@}~\bot=(\bot,\bot)$, and
-${\it{}fac}_{@@@T->U@@@}~(\itt{lift}~f)=(\lunit,~f)$ for all $f$.
+${\it{}fac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~\bot=(\bot,\bot)$, and
+${\it{}fac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\itt{lift}~f)=(\lunit,~f)$ for all $f$.
 Every embedding {\it determines\/} a corresponding projection; 
@@
 projection is given by
-${\it{}unfac}_{@@@T->U@@@}~(\bot,f)=\bot$ and
-${\it{}unfac}_{@@@T->U@@@}~(\lunit,~f)=\itt{lift}~f$.
-
-Generally, the data domain corresponding to type @T@ is $\De{}{@T@}$,
+${\it{}unfac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\bot,f)=\bot$ and
+${\it{}unfac}_{\subtypespace{\makebox{\subtypesize\tt T->U}} }~(\lunit,~f)=\itt{lift}~f$.
+
+Generally, the data domain corresponding to type \mbox{\tt T} is $\De{}{\mbox{\tt T}}$,
 where $\cD{}$ is defined exactly like $\cT{\Ss}$ except that function
@@
 The function $\data$ from values in
-$\Te{\Ss}{@T@}$ to their data parts in $\De{}{@T@}$ is a projection:
+$\Te{\Ss}{\mbox{\tt T}}$ to their data parts in $\De{}{\mbox{\tt T}}$ is a projection:
 it is like the identity except that values from function spaces are
 mapped into $\one$.  The projection $\data$ is defined in terms of the
-structure of @T@ as given in Figure~\ref{fig:fact}.
+structure of \mbox{\tt T} as given in Figure~\ref{fig:fact}.
 \fig{tb}{Factorisation and unfactorisation functions.}{fig:fact}{
-""
-	{\it{}data}_{\stint} = {\it{}id}_{\stint}\ ,\espace
-	{\it{}data}_{\stprod} = {\it{}data}_{\stone} \times \ldots \times {\it{}data}_{\stn} ,\espace
-	{\it{}data}_{\stsum} = {\it{}data}_{\stone} \plus \ldots \plus {\it{}data}_{\stn} ,\espace
-	{\it{}data}_{\stfun} = {\liftp{\Lx.\unit}} .
-""
-
-"""
-	{\it{}for}_{\stint} = \Lx.\unit\ ,\espace
-	{\it{}for}_{\stprod} = {\it{}for}_{\stone}\ \times\ \ldots\ \times\ {\it{}for}_{\stn}\ ,\espace
-        \ba{lll}{\it{}for}_{\stsum}\ \bot	&\ = \bot
-	{\it{}for}_{\stsum}\ (\itt{in}_i~v) &\ = (\bot,\ldots,\bot,v,\bot_,\ldots,\bot)	&\ \ \reason{$v$ in $\ith$ position}\ ,\ea\espace
-	{\it{}for}_{\stfun} = \itt{drop}\ .
-"""
-
-"""
-	\itt{unfac}_{\stint} (n,()) = n ,\espace
-	\itt{unfac}_{\stprod} ((d_1,...,d_n), (f_1,...,f_n)) = (v_1,...,v_n)
-	~~~{\rm{}where} 
-	~~~~~~v_i=\itt{unfac}_{\sti} (d_i,f_i), \oin ,\espace
-	\itt{unfac}_{\stsum} ((d_1,...,d_n), (f_1,...,f_n)) = (v_1,...,v_n)
-	~~~{\rm{}where} 
-	~~~~~~v_i=\itt{unfac}_{\sti} (d_i,f_i), \oin ,\espace
-	\itt{unfac}_{\stfun} (\bot,f) = \bot,
-	\itt{unfac}_{\stfun} (\lunit, f) = \itt{lift} f . 
-"""
+\beqs
+\it {\it{}data}_{\subtypespace\stint}\ =\ {\it{}id}_{\subtypespace\stint}\ ,\espace\\
+\it {\it{}data}_{\subtypespace\stprod}\ =\ {\it{}data}_{\subtypespace\stone}\ \times\ \ldots\ \times\ {\it{}data}_{\subtypespace\stn}\ ,\espace\\
+\it {\it{}data}_{\subtypespace\stsum}\ =\ {\it{}data}_{\subtypespace\stone}\ \plus\ \ldots\ \plus\ {\it{}data}_{\subtypespace\stn}\ ,\espace\\
+\it {\it{}data}_{\subtypespace\stfun}\ =\ {\liftp{\Lx.\unit}}\ .
+\eeqs
+
+\beqqs
+\it {\it{}for}_{\subtypespace\stint}\ =\ \Lx.\unit\ ,\espace\\
+\it {\it{}for}_{\subtypespace\stprod}\ =\ {\it{}for}_{\subtypespace\stone}\ \times\ \ldots\ \times\ {\it{}for}_{\subtypespace\stn}\ ,\espace\\
+\it \ba{lll}{\it{}for}_{\subtypespace\stsum}\ \bot&\it \ =\ \bot\\
+\it {\it{}for}_{\subtypespace\stsum}\ (\itt{in}_i~v)&\it \ =\ (\bot,\ldots,\bot,v,\bot_,\ldots,\bot)&\it \ \ \reason{$v$\ in\ $\ith$\ position}\ ,\ea\espace\\
+\it {\it{}for}_{\subtypespace\stfun}\ =\ \itt{drop}\ .
+\eeqqs
+\beqqs
+\it \itt{unfac}_{\subtypespace\stint}\ (n,())\ =\ n\ ,\espace\\
+\it \itt{unfac}_{\subtypespace\stprod}\ ((d_{\rm1},\ldots ,d_n),\ (f_{\rm1},\ldots ,f_n))\ =\ (v_{\rm1},\ldots ,v_n)\\
+\it ~~~{\rm{}where}\\
+\it ~~~~~~v_i=\itt{unfac}_{\subtypespace\sti}\ (d_i,f_i),\ \oin\ ,\espace\\
+\it \itt{unfac}_{\subtypespace\stsum}\ ((d_{\rm1},\ldots ,d_n),\ (f_{\rm1},\ldots ,f_n))\ =\ (v_{\rm1},\ldots ,v_n)\\
+\it ~~~{\rm{}where}\\
+\it ~~~~~~v_i=\itt{unfac}_{\subtypespace\sti}\ (d_i,f_i),\ \oin\ ,\espace\\
+\it \itt{unfac}_{\subtypespace\stfun}\ (\bot,f)\ =\ \bot,\\
+\it \itt{unfac}_{\subtypespace\stfun}\ (\lunit,\ f)\ =\ \itt{lift}\ f\ .
+\eeqqs
 }
-The definition of ${\it{}data}_{\stfun}$ uses function lifting, 
+The definition of ${\it{}data}_{\subtypespace\stfun}$ uses function lifting, 
 defined by $\lf~\bot=\bot$
@@
 
-In the same style as  $\De{}{@T@}$ and $\data$ we define
-$\Tde{\Ss}{@T@}$ to give the forward domain for @T@, and $\for$ to be
-the function mapping values in $\Te{\Ss}{@T@}$ to their forward parts
-in $\Tde{\Ss}{@T@}$, as follows.  Roughly, $\cTd{\Ss}$ is like
+In the same style as  $\De{}{\mbox{\tt T}}$ and $\data$ we define
+$\Tde{\Ss}{\mbox{\tt T}}$ to give the forward domain for \mbox{\tt T}, and $\for$ to be
+the function mapping values in $\Te{\Ss}{\mbox{\tt T}}$ to their forward parts
+in $\Tde{\Ss}{\mbox{\tt T}}$, as follows.  Roughly, $\cTd{\Ss}$ is like
 $\cT{\Ss}$ with all lifting removed and sum replaced by product, so
-""
-	\Tde{\Ss}{\tint} = \one ,\espace
-	\Tde{\Ss}{\tprod} = \Tde{\Ss}{\tone} \times \ldots \times \Tde{\Ss}{\tn} ,\espace
-	\Tde{\Ss}{\tsum} = \Tde{\Ss}{\tone} \times \ldots \times \Tde{\Ss}{\tn} ,\espace
-	\Tde{\Ss}{\tfun} = \Te{\Ss}{\tone} \fto \Te{\Ss}{\ttwo} .
-""
+\beqs
+\it \Tde{\Ss}{\tint}\ =\ \one\ ,\espace\\
+\it \Tde{\Ss}{\tprod}\ =\ \Tde{\Ss}{\tone}\ \times\ \ldots\ \times\ \Tde{\Ss}{\tn}\ ,\espace\\
+\it \Tde{\Ss}{\tsum}\ =\ \Tde{\Ss}{\tone}\ \times\ \ldots\ \times\ \Tde{\Ss}{\tn}\ ,\espace\\
+\it \Tde{\Ss}{\tfun}\ =\ \Te{\Ss}{\tone}\ \fto\ \Te{\Ss}{\ttwo}\ .
+\eeqs
 The mapping from standard values to their forward parts is defined in
@@
 We write $\fac$ for $\Lx.(\data~x,\ \for~x)$.  Then
-$\De{}{@T@}\times\Tde{\Ss}{@T@}$ is a factorisation of
-$\Te{\Ss}{@T@}$, and
-""
-	\fac \in \Te{\Ss}{@T@} \fto (\De{}{@T@} \times \Tde{\Ss}{@T@})
-""
+$\De{}{\mbox{\tt T}}\times\Tde{\Ss}{\mbox{\tt T}}$ is a factorisation of
+$\Te{\Ss}{\mbox{\tt T}}$, and
+\beqs
+\it \fac\ \in\ \Te{\Ss}{\makebox{\tt T}}\ \fto\ (\De{}{\makebox{\tt T}}\ \times\ \Tde{\Ss}{\makebox{\tt T}})
+\eeqs
 is an embedding which, as mentioned, determines the corresponding projection
@@
 $\Int$ is $\Int\times\one$; more generally, the factorisation of any
-domain $D$ corresponding to a type not containing @->@ is just
-$D\times\one$.  For @FunList@ the standard domain is
-"""
-	\mu{}X . \lone+(\liftp{\itt{Int}\fto\itt{Int}}\x{}X) ,
-"""
+domain $D$ corresponding to a type not containing \mbox{\tt ->} is just
+$D\times\one$.  For \mbox{\tt FunList} the standard domain is
+\beqqs
+\it \mu{}X\ .\ \lone+(\liftp{\itt{Int}\fto\itt{Int}}\x{}X)\ ,
+\eeqqs
 that is, lists of functions from integers to integers. In factored form
 the domain is
-"""
-	(\mu{}X . \lone+(\lone\x{}X)) \x (\mu{}Y . (\itt{Int}\fto\itt{Int})\x{}Y) ,
-"""
+\beqqs
+\it (\mu{}X\ .\ \lone+(\lone\x{}X))\ \x\ (\mu{}Y\ .\ (\itt{Int}\fto\itt{Int})\x{}Y)\ ,
+\eeqqs
 that is, pairs consisting of a list of values from $\{\bot,\ \lbot\}$, and
 an infinite tuple of functions from integers to integers.  Then the 
-standard value of the list @cons (\x.x-1, cons (\x.x-2, nil ()))@ is
-"""
-	\itt{incons} (\itt{lift} (\L{}x.x-1), \itt{incons} (\itt{lift} (\L{}x.x-2), \itt{innil} ())) ,
-"""
+standard value of the list \mbox{\tt cons\ ({\char'134}x.x-1,\ cons\ ({\char'134}x.x-2,\ nil\ ()))} is
+\beqqs
+\it \itt{incons}\ (\itt{lift}\ (\L{}x.x-{\rm1}),\ \itt{incons}\ (\itt{lift}\ (\L{}x.x-{\rm2}),\ \itt{innil}\ ()))\ ,
+\eeqqs
 the data value is
-"""
-	\itt{incons} (\lunit, \itt{incons} (\lunit, \itt{innil} ())) ,
-"""
+\beqqs
+\it \itt{incons}\ (\lunit,\ \itt{incons}\ (\lunit,\ \itt{innil}\ ()))\ ,
+\eeqqs
 and the forward value is 
@@
 their forward parts are {\it logically related\/} \cite{Abr90} by
-$\K$.  At each type @T@ the relation is
-$\Re{}{@T@}~((\A,\K),\cdot,\cdot)$, defined by
-""
-	{\Re{}{@T@}~((\A,\K),v,v')} =&\mit(\A~d=\A~d')\ \wedge\ \Rde{}{@T@}~(\K,f,f')
-		&\mbox{\rm{}where}
-		&\ \ \ \ba[t]{lll}(d,f)  &\:= \fac~v
-				\mit(d',f')	&\:= \fac~v' ,\ea
-""
-where $\Rde{}{@T@}$ is defined by
-""
-	\Rde{}{\tint}~((),(),()) = \itt{True}\ ,\espace
-	\Rde{}{\tprod} = \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace
-	\Rde{}{\tsum} = \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace
-	\Rde{}{\tfun} = \Re{}{\tone}\ \fto\ \Re{}{\ttwo}\ .
-""
+$\K$.  At each type \mbox{\tt T} the relation is
+$\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$, defined by
+\beqs
+\it {\Re{}{\makebox{\tt T}}~((\A,\K),v,v')}\ =&\it \mit(\A~d=\A~d')\ \wedge\ \Rde{}{\makebox{\tt T}}~(\K,f,f')\\
+\it &\it \mbox{\rm{}where}\\
+\it &\it \ \ \ \ba[t]{lll}(d,f)&\it \:=\ \fac~v\\
+\it \mit(d',f')&\it \:=\ \fac~v'\ ,\ea
+\eeqs
+where $\Rde{}{\mbox{\tt T}}$ is defined by
+\beqs
+\it \Rde{}{\tint}~((),(),())\ =\ \itt{True}\ ,\espace\\
+\it \Rde{}{\tprod}\ =\ \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace\\
+\it \Rde{}{\tsum}\ =\ \Rde{}{\tone}\ \x\ \ldots\ \x\ \Rde{}{\tn}\ ,\espace\\
+\it \Rde{}{\tfun}\ =\ \Re{}{\tone}\ \fto\ \Re{}{\ttwo}\ .
+\eeqs
 Here $\x$ and $\fto$ are the standard operations on {\it ternary\/}
@@
 $\cT{\Ps}$ must be
-""
-	\Te{\Ps}{@T@} = \proj{\De{}{@T@}}\ \times\ \Tde{\Ps}{@T@}\ ,
-""
-where $\proj{\De{}{@T@}}$ is the lattice of projections on domain
-$\De{}{@T@}$, and $\cTd{\Ps}$ is defined exactly like $\cTd{\Ss}$ with
+\beqs
+\it \Te{\Ps}{\makebox{\tt T}}\ =\ \proj{\De{}{\makebox{\tt T}}}\ \times\ \Tde{\Ps}{\makebox{\tt T}}\ ,
+\eeqs
+where $\proj{\De{}{\mbox{\tt T}}}$ is the lattice of projections on domain
+$\De{}{\mbox{\tt T}}$, and $\cTd{\Ps}$ is defined exactly like $\cTd{\Ss}$ with
 superscript $\Ps$ everywhere replacing superscript $\Ss$.
@@
 \fig{tb}{Projection semantics.}{fig:fsem}{
-""
-%	\itt{mkunit}^{\Ps}\ \R			 = 	()
-%\espace\espace
-%	\itt{mkint}^{\Ps}_n\ \R			 = 	\itt{lift}\ n
-%\espace\espace
-	\itt{plus}^{\Ps}\ ((\A,()),(\B,())) = \left\{\ba{ll}(\ID,()),	&\ \ \ \ \mbox{\rm if $\A=\ID$ and $\B=\ID$}
-			(\BOT,()),	&\ \ \ \ \mbox{\rm otherwise}\ea\right.\espace
-	\itt{tuple}^{\Ps}\ ((\A_1,\K_1),\ldots,(\A_n,\K_n)) = ((\A_1\x\ldots\x\A_n),(\K_1,\ldots,\K_n))\espace
-	\itt{sel}_i^{\,\Ps}\ ((\A_1\x\ldots\x\A_n),(\K_1,\ldots,\K_n))  = (\A_i,\K_i)\espace
-	\itt{inc}_i^{\Ps}\ (\A,\K)	 = 	(C_i\ \A,\ (\top,\ldots,\top,\K,\top,\ldots,\top))\espace
-	\itt{outc}_i^{\Ps}\ (\A,(\K_1,\ldots,\K_n))	 = (\OUTC_i\ \A,\ \K_i)\espace
-	\itt{choose}^{\Ps}\ ((\A,\K),x_1,\ldots,x_n) = \left\{\ba{ll}\bot,		&\ \ \ \ \mbox{\rm if}\ \A\not\se\glb_{\oin} C_i~\BOT
-			x_1\glb\ldots\glb{}x_n,	&\ \ \ \ \mbox{\rm otherwise}\ea\right.\espace
-	\itt{mkfun}^{\Ps}\ f			 = 	(\ID,f)\espace
-	\itt{apply}^{\Ps}\ (\A,f) = \left\{\ba{lll}\bot,	&\ \ \ \ \mbox{\rm if}\ \A=\BOT
-						f,	&\ \ \ \ \mbox{\rm if}\ \A=\ID \ea\right.\espace
-	\itt{fix}^{\Ps} = 	\itt{gfp} \reason{Greatest fixed point}
-""
+\beqs
+\it %\ \itt{mkunit}^{\Ps}\ \R\ =\ ()\\
+\it %\espace\espace\\
+\it %\ \itt{mkint}^{\Ps}_n\ \R\ =\ \itt{lift}\ n\\
+\it %\espace\espace\\
+\it \itt{plus}^{\Ps}\ ((\A,()),(\B,()))\ =\ \left\{\ba{ll}(\ID,()),&\it \ \ \ \ \mbox{\rm\ if\ $\A=\ID$\ and\ $\B=\ID$}\\
+\it (\BOT,()),&\it \ \ \ \ \mbox{\rm\ otherwise}\ea\right.\espace\\
+\it \itt{tuple}^{\Ps}\ ((\A_{\rm1},\K_{\rm1}),\ldots,(\A_n,\K_n))\ =\ ((\A_{\rm1}\x\ldots\x\A_n),(\K_{\rm1},\ldots,\K_n))\espace\\
+\it \itt{sel}_i^{\,\Ps}\ ((\A_{\rm1}\x\ldots\x\A_n),(\K_{\rm1},\ldots,\K_n))\ =\ (\A_i,\K_i)\espace\\
+\it \itt{inc}_i^{\Ps}\ (\A,\K)\ =\ (C_i\ \A,\ (\top,\ldots,\top,\K,\top,\ldots,\top))\espace\\
+\it \itt{outc}_i^{\Ps}\ (\A,(\K_{\rm1},\ldots,\K_n))\ =\ (\OUTC_i\ \A,\ \K_i)\espace\\
+\it \itt{choose}^{\Ps}\ ((\A,\K),x_{\rm1},\ldots,x_n)\ =\ \left\{\ba{ll}\bot,&\it \ \ \ \ \mbox{\rm\ if}\ \A\not\se\glb_{\oin}\ C_i~\BOT\\
+\it x_{\rm1}\glb\ldots\glb{}x_n,&\it \ \ \ \ \mbox{\rm\ otherwise}\ea\right.\espace\\
+\it \itt{mkfun}^{\Ps}\ f\ =\ (\ID,f)\espace\\
+\it \itt{apply}^{\Ps}\ (\A,f)\ =\ \left\{\ba{lll}\bot,&\it \ \ \ \ \mbox{\rm\ if}\ \A=\BOT\\
+\it f,&\it \ \ \ \ \mbox{\rm\ if}\ \A=\ID\ \ea\right.\espace\\
+\it \itt{fix}^{\Ps}\ =\ \itt{gfp}\ \reason{Greatest\ fixed\ point}
+\eeqs
 }
@@
 
-\ \\ \noindent {\bf The Central Result.} For expression @e@ of type
-@T@ and free-variable environment type @E@, the functions
-$\Ee{\Ps}{@e@}$, $\Ee{\Ss}{@e@}$, and $\Ee{\Ss}{@e@}$ are logically
+\ \\ \noindent {\bf The Central Result.} For expression \mbox{\tt e} of type
+\mbox{\tt T} and free-variable environment type \mbox{\tt E}, the functions
+$\Ee{\Ps}{\mbox{\tt e}}$, $\Ee{\Ss}{\mbox{\tt e}}$, and $\Ee{\Ss}{\mbox{\tt e}}$ are logically
 related by $\cR{}$, that is
-""
-	(\Re{}{@E@}\fto\Re{}{@T@})\ (\Ee{\Ps}{@e@},\ \Ee{\Ss}{@e@},\ \Ee{\Ss}{@e@})\ .
-""
-Further, $\Re{}{@T@}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
-$\K$.\ \ %$\Box$
+\beqs
+\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\makebox{\tt T}})\ (\Ee{\Ps}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}})\ .
+\eeqs
+Further, $\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
+$\K$.\ \ $\Box$
 
@@
 and $\it plus^{\Ps}$ are correctly related we need to show that
-"""
-	((\Re{}{\tint}\x\Re{}{\tint})\fto\Re{}{\tint}) (plus^{\Ps},plus^{\Ss},plus^{\Ss}) .
-"""
+\beqqs
+\it ((\Re{}{\tint}\x\Re{}{\tint})\fto\Re{}{\tint})\ (plus^{\Ps},plus^{\Ss},plus^{\Ss})\ .
+\eeqqs
 By the definitions of $\fto$ and $\x$ on ternary relations, this is
 equivalent to 
-""
-	(\Re{}{\tint}\x\Re{}{\tint}) (x,y,z) \limp \Re{}{\tint}\
-		\ (plus^{\Ps} x, plus^{\Ss} y, plus^{\Ss} z) ,
-""
+\beqs
+\it (\Re{}{\tint}\x\Re{}{\tint})\ (x,y,z)\ \limp\ \Re{}{\tint}\ (plus^{\Ps}\ x,\ plus^{\Ss}\ y,\ plus^{\Ss}\ z)\ ,
+\eeqs
 for all $x$, $y$, and $z$.  In turn, this is equivalent to
-"""
-	&\ \Re{}{\tint} (x,y,z) \land \Re{}{\tint} (x',y',z')
-\limp	&\ \Re{}{\tint} (plus^{\Ps} (x,x'), plus^{\Ss} (y,y'), plus^{\Ss} (z,z'))
-"""
+\beqqs
+\it &\it \ \Re{}{\tint}\ (x,y,z)\ \land\ \Re{}{\tint}\ (x',y',z')\\
+\it \limp&\it \ \Re{}{\tint}\ (plus^{\Ps}\ (x,x'),\ plus^{\Ss}\ (y,y'),\ plus^{\Ss}\ (z,z'))
+\eeqqs
 for all $x$, $x'$, $y$, $y'$, $z$, and $z'$; equivalently, for
@@
 $\Te{\Ss}{\tint}$ we have
-""
-	\mit(\ID y=\ID z) \land (\ID y'=\ID z') \limp \mit\ID (y+y')=\ID (z+z') ,
-""
+\beqs
+\it \mit(\ID\ y=\ID\ z)\ \land\ (\ID\ y'=\ID\ z')\ \limp\ \mit\ID\ (y+y')=\ID\ (z+z')\ ,
+\eeqs
 which trivially holds, and for $\A_1\neq\ID$ or $\A_2\neq\ID$ that
-""
-	\mit(\A_1 y=\A_1 z) \land (\A_2 y'=\A_2 z') \limp \mit\BOT (y+y')=\BOT (z+z') ,
-""
+\beqs
+\it \mit(\A_{\rm1}\ y=\A_{\rm1}\ z)\ \land\ (\A_{\rm2}\ y'=\A_{\rm2}\ z')\ \limp\ \mit\BOT\ (y+y')=\BOT\ (z+z')\ ,
+\eeqs
 which is also trivial.  For the inductive case $\esum$ with 
-environment type @E@ we need to show that if
-""
-	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\eone}, \Ee{\Ss}{\eone}, \Ee{\Ss}{\eone}) 
-""
+environment type \mbox{\tt E} we need to show that if
+\beqs
+\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (\Ee{\Ps}{\eone},\ \Ee{\Ss}{\eone},\ \Ee{\Ss}{\eone})
+\eeqs
 and
-""
-	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\etwo}, \Ee{\Ss}{\etwo}, \Ee{\Ss}{\etwo}) 
-""
+\beqs
+\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (\Ee{\Ps}{\etwo},\ \Ee{\Ss}{\etwo},\ \Ee{\Ss}{\etwo})
+\eeqs
 then
-"""
-	(\Re{}{@E@}\fto\Re{}{\tint}) (\Ee{\Ps}{\esum},\Ee{\Ss}{\esum},\Ee{\Ss}{\esum}) .
-"""
-Suppose $\Re{}{@E@}~(\R^{\Ps},\R^{\Ss},\R'^{\Ss})$.  Then by 
+\beqqs
+\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\tint})\ (\Ee{\Ps}{\esum},\Ee{\Ss}{\esum},\Ee{\Ss}{\esum})\ .
+\eeqqs
+Suppose $\Re{}{\mbox{\tt E}}~(\R^{\Ps},\R^{\Ss},\R'^{\Ss})$.  Then by 
 the induction hypothesis we have that
@@
 Now
-"""
-	\Re{}{\tint} (\Ee{\Ps}{\esum} \R^{\Ps}, \Ee{\Ss}{\esum}~\R^{\Ss}, \Ee{\Ss}{\esum}~\R'^{\Ss})
-"""
+\beqqs
+\it \Re{}{\tint}\ (\Ee{\Ps}{\esum}\ \R^{\Ps},\ \Ee{\Ss}{\esum}~\R^{\Ss},\ \Ee{\Ss}{\esum}~\R'^{\Ss})
+\eeqqs
 iff
-"""
-	\Re{}{\tint} (&plus^{\Ps} ((\A_1,()),(\A_2,())), 
-		      &plus^{\Ss} \mit((v_1,()),(v_2,())), 
-		      &plus^{\Ss} \mit((v'_1,()),(v'_2,()))) ,
-"""
+\beqqs
+\it \Re{}{\tint}\ (&\it plus^{\Ps}\ ((\A_{\rm1},()),(\A_{\rm2},())),\\
+\it &\it plus^{\Ss}\ \mit((v_{\rm1},()),(v_{\rm2},())),\\
+\it &\it plus^{\Ss}\ \mit((v'_{\rm1},()),(v'_{\rm2},())))\ ,
+\eeqqs
 which holds since $\it plus^{\Ps}$ and $\it plus^{\Ps}$, and their
@@
 
-At each type @T@ we require a finite abstraction of $\Te{\Ps}{@T@}$.
-This abstract domain is ${\it{}FProj}_{\stt}\times{\it{}FFor}_{\stt}$,
-where ${\it{}FProj}_{\stt}$ is a finite abstraction of the lattice
-$\proj{\De{}{@T@}}$, and ${\it{}FFor}_{\stt}$ is a finite abstraction
-of $\Tde{\Ps}{@T@}$.  The definition of ${\it{}FProj}_{\stt}$ is based
+At each type \mbox{\tt T} we require a finite abstraction of $\Te{\Ps}{\mbox{\tt T}}$.
+This abstract domain is ${\it{}FProj}_{\subtypespace\stt}\times{\it{}FFor}_{\subtypespace\stt}$,
+where ${\it{}FProj}_{\subtypespace\stt}$ is a finite abstraction of the lattice
+$\proj{\De{}{\mbox{\tt T}}}$, and ${\it{}FFor}_{\subtypespace\stt}$ is a finite abstraction
+of $\Tde{\Ps}{\mbox{\tt T}}$.  The definition of ${\it{}FProj}_{\subtypespace\stt}$ is based
 on Launchbury's \cite{Lau91a}.  A projection $\G$ is in
-${\it{}FProj}_{\stt}$ if $\G~\mbox{\bf proj}~@T@$ can be inferred from
+${\it{}FProj}_{\subtypespace\stt}$ if $\G~\mbox{\bf proj}~\mbox{\tt T}$ can be inferred from
 the rules given in Figure~\ref{fig:infrules}.
@@
 \fig{tb}{Inference rules for finite projection domains.}{fig:infrules}{
-""
-\begin{array}{lll}\
-\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tint}\ \ \ \ \ \ \ \ \ \ & \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tint}\
-\espace\espace
-\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tfun} & \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tfun}\end{array}\
-\espace\espace
-\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tsum}\
-\espace\espace
-\infero{\G_1\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{(C_1\ \G_1)\glb\cdots\glb(C_n\ \G_n)\ \ \mbox{\bf{}proj}\ \ \tsum}\
-\espace\espace
-\infero{\G_1\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{\G_1\x\ldots\x\G_n\ \ \mbox{\bf{}proj}\ \ \tprod}\
-\espace\espace
-\infer{\ba{ll}&\G_1\ \ \mbox{\bf{}proj}\ \ \aone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \an
-\vdash & P(\G_1,...,\G_n)\ \ \mbox{\bf{}proj}\ \ \ti@(@\aone@,...,@\an@)@\ea}{\mu(\G_1,\ldots,\G_n).P(\G_1,\ldots,\G_n)\ \ \mbox{\bf{}proj}\ \ \ai}\
-\ \ \ \ \ [\mbox{\rm{}where}\ \ai@=@\ti@(@\aone@,...,@\an@)@]
-""
+\beqs
+\it \begin{array}{lll}\axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tint}\ \ \ \ \ \ \ \ \ \ &\it \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tint}\espace\espace\\
+\it \axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tfun}&\it \axm{\ID\ \ \mbox{\bf{}proj}\ \ \tfun}\end{array}\espace\espace\\
+\it \axm{\BOT\ \ \mbox{\bf{}proj}\ \ \tsum}\espace\espace\\
+\it \infero{\G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{(C_{\rm1}\ \G_{\rm1})\glb\cdots\glb(C_n\ \G_n)\ \ \mbox{\bf{}proj}\ \ \tsum}\espace\espace\\
+\it \infero{\G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \tone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \tn}{\G_{\rm1}\x\ldots\x\G_n\ \ \mbox{\bf{}proj}\ \ \tprod}\espace\espace\\
+\it \infer{\ba{ll}&\it \G_{\rm1}\ \ \mbox{\bf{}proj}\ \ \aone\ \ \cdots\ \ \G_n\ \ \mbox{\bf{}proj}\ \ \an\\
+\it \vdash&\it P(\G_{\rm1},\ldots ,\G_n)\ \ \mbox{\bf{}proj}\ \ \ti\makebox{\tt (}\aone\makebox{\tt ,...,}\an\makebox{\tt )}\ea}{\mu(\G_{\rm1},\ldots,\G_n).P(\G_{\rm1},\ldots,\G_n)\ \ \mbox{\bf{}proj}\ \ \ai}\ \ \ \ \ [\mbox{\rm{}where}\ \ai\makebox{\tt =}\ti\makebox{\tt (}\aone\makebox{\tt ,...,}\an\makebox{\tt )}]
+\eeqs
 }
@@
 data structure in the same
-way.  Thus ${\it{}FProj}_{@@@FunList@@@}$ comprises $\BOT$ and
+way.  Thus ${\it{}FProj}_{\subtypespace{\makebox{\subtypesize\tt FunList}} }$ comprises $\BOT$ and
 $\SPINE~\A$ for $\A$ ranging over $\BOT$ and $\ID$, where
-""
-	\SPINE\ \A = \mu\G.(\itt{NIL}\ \ID)\ \glb\ (\itt{CONS}\ (\A\ \x\ \G))\ ,
-""
+\beqs
+\it \SPINE\ \A\ =\ \mu\G.(\itt{NIL}\ \ID)\ \glb\ (\itt{CONS}\ (\A\ \x\ \G))\ ,
+\eeqs
 so $\SPINE\ \BOT$ specifies static spines and dynamic elements, and
@@
 constructor is isomorphic to lifting.  Similarly,
-${\it{}FProj}_{@@@FunTree@@@}$
+${\it{}FProj}_{\subtypespace{\makebox{\subtypesize\tt FunTree}} }$
 comprises $\BOT$ and $\LBR\ \A$ for $\A$ ranging over $\BOT$ and $\ID$, where
-""
-	\LBR\ \A = \mu\G.(\itt{LEAF}\ \A)\ \glb\ (\itt{BRANCH}\ (\G\ \x\ \G))\ ,
-""
+\beqs
+\it \LBR\ \A\ =\ \mu\G.(\itt{LEAF}\ \A)\ \glb\ (\itt{BRANCH}\ (\G\ \x\ \G))\ ,
+\eeqs
 so $\LBR\ \A$ specifies static branches and leaves, and $\A$ of the
@@
 
-Value $\K$ is in ${\it{}FFor}_{\stt}$ if $\K~\bfabsf~@T@$ can be
+Value $\K$ is in ${\it{}FFor}_{\subtypespace\stt}$ if $\K~\bfabsf~\mbox{\tt T}$ can be
 inferred from the following.
 First, there is only one forward value at type $\tint$.
-""
-	\axm{() {} \bfabsf {} \tint} .
-""
+\beqs
+\it \axm{()\ {}\ \bfabsf\ {}\ \tint}\ .
+\eeqs
 For products and sums,
-""
-\infer{\K_1 {} \bfabsf {} \tone { } \cdots { } \K_n {} \bfabsf {} \tn}\
-{(\K_1,\ldots,\K_n) {} \bfabsf {} \tprod} ,
-""
+\beqs
+\it \infer{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tone\ {\ }\ \cdots\ {\ }\ \K_n\ {}\ \bfabsf\ {}\ \tn}{(\K_{\rm1},\ldots,\K_n)\ {}\ \bfabsf\ {}\ \tprod}\ ,
+\eeqs
 %\espace\espace
-""
-\infero{\K_1 {} \bfabsf {} \tone { } \cdots { } \K_n {} \bfabsf {} \tn}\
-{(\K_1,\ldots,\K_n) { } \bfabsf {} \tsum} .
-""
+\beqs
+\it \infero{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tone\ {\ }\ \cdots\ {\ }\ \K_n\ {}\ \bfabsf\ {}\ \tn}{(\K_{\rm1},\ldots,\K_n)\ {\ }\ \bfabsf\ {}\ \tsum}\ .
+\eeqs
 Function spaces consist of a set of step functions closed under lub.
-""
-\infero{\ba{ll}\A_1\in{\it{}FProj}_{\stone} { { } } \K_1 \bfabsf \tone
-\A_2\in{\it{}FProj}_{\sttwo} { { } } \K_2 \bfabsf \ttwo\ea}\
-{\it{}step ((\A_1,\K_1),(\A_2,\K_2)) {} \bfabsf {} \tfunp} ,
-""
+\beqs
+\it \infero{\ba{ll}\A_{\rm1}\in{\it{}FProj}_{\subtypespace\stone}\ {\ {\ }\ }\ \K_{\rm1}\ \bfabsf\ \tone\\
+\it \A_{\rm2}\in{\it{}FProj}_{\subtypespace\sttwo}\ {\ {\ }\ }\ \K_{\rm2}\ \bfabsf\ \ttwo\ea}{\it{}step\ ((\A_{\rm1},\K_{\rm1}),(\A_{\rm2},\K_{\rm2}))\ {}\ \bfabsf\ {}\ \tfunp}\ ,
+\eeqs
 where
-""
-	step (v_1,v_2) x = v_2,		&{\rm if} v_1\we{}x 
-	step (v_1,v_2) x = \bot,	&{\rm otherwise} ,
-""
+\beqs
+\it step\ (v_{\rm1},v_{\rm2})\ x\ =\ v_{\rm2},&\it {\rm\ if}\ v_{\rm1}\we{}x\\
+\it step\ (v_{\rm1},v_{\rm2})\ x\ =\ \bot,&\it {\rm\ otherwise}\ ,
+\eeqs
 and
-""
-\infer{\K_1 {} \bfabsf {} \tfunp { { } } \K_2 {} \bfabsf {} \tfunp}\
-{(\K_1\lub\K_2) {} \bfabsf {} \tfunp} .
-""
+\beqs
+\it \infer{\K_{\rm1}\ {}\ \bfabsf\ {}\ \tfunp\ {\ {\ }\ }\ \K_{\rm2}\ {}\ \bfabsf\ {}\ \tfunp}{(\K_{\rm1}\lub\K_{\rm2})\ {}\ \bfabsf\ {}\ \tfunp}\ .
+\eeqs
 This gives the full space of monotonic functions on the abstract
@@
 same value.  Given type definitions $\tdefn$, which will be written
-$@A@_i@ = T@_i@(A@_1@,...,A@_n@)@,\ \oin$, if from assumptions
+$\mbox{\tt A}_i\mbox{\tt \ =\ T}_i\mbox{\tt (A}_1\mbox{\tt ,...,A}_n\mbox{\tt )},\ \oin$, if from assumptions
 $\K_i~\bfabsf~\ai$ for $\oin$ it may be deduced that
 $P_i(\K_1,\ldots,\K_n)~~\bfabsf~~\ti(\aone\ldots\an)$ for $\oin$, then
-""
-	\mu(\K_1,...,\K_n).(P_1(\K_1,...,\K_n),...,P_n(\K_1,...,\K_n))
-""
-is a tuple $(\K_1,...,\K_n)$ such that $\K_i\ \bfabsf\ @A@_i$
+\beqs
+\it \mu(\K_{\rm1},\ldots ,\K_n).(P_{\rm1}(\K_{\rm1},\ldots ,\K_n),\ldots ,P_n(\K_{\rm1},\ldots ,\K_n))
+\eeqs
+is a tuple $(\K_1,...,\K_n)$ such that $\K_i\ \bfabsf\ \mbox{\tt A}_i$
 for $\oin$.
-For all @T@ the set ${\it{}FProj}_{\stt}\times{\it{}FFor}_{\stt}$ is a 
-finite lattice containing the top and bottom elements of $\Te{\Ps}{@T@}$.
-
-For @T@ not containing @->@ the domain $\Tde{\Ps}{@T@}$ is isomorphic
-to $\one$, so ${\it{}FFor}_{\stint}$ is $\one$.  For @Int@\ @->@\ @Int@
+For all \mbox{\tt T} the set ${\it{}FProj}_{\subtypespace\stt}\times{\it{}FFor}_{\subtypespace\stt}$ is a 
+finite lattice containing the top and bottom elements of $\Te{\Ps}{\mbox{\tt T}}$.
+
+For \mbox{\tt T} not containing \mbox{\tt ->} the domain $\Tde{\Ps}{\mbox{\tt T}}$ is isomorphic
+to $\one$, so ${\it{}FFor}_{\subtypespace\stint}$ is $\one$.  For \mbox{\tt Int}\ \mbox{\tt ->}\ \mbox{\tt Int}
 we have
-$\Tde{\Ps}{@Int@\ @->@\ @Int@}=(\proj{\De{}{\tint}}\x\one)\fto(\proj{\De{}{\tint}}\x\one)$,
-so ${\it{}FFor}_{@@@Int->Int@@@}$ is
-$({\it{}FProj}_{\stint}\x\one)\fto({\it{}FProj}_{\stint}\x\one)$.  A
-data structure of recursive type @A=T(A)@ may be thought of as some
-(possibly infinite) number of elements of @T(())@.  For example, the
+$\Tde{\Ps}{\mbox{\tt Int}\ \mbox{\tt ->}\ \mbox{\tt Int}}=(\proj{\De{}{\tint}}\x\one)\fto(\proj{\De{}{\tint}}\x\one)$,
+so ${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt Int->Int}} }$ is
+$({\it{}FProj}_{\subtypespace\stint}\x\one)\fto({\it{}FProj}_{\subtypespace\stint}\x\one)$.  A
+data structure of recursive type \mbox{\tt A=T(A)} may be thought of as some
+(possibly infinite) number of elements of \mbox{\tt T(())}.  For example, the
 value $\itt{cons}\ (f,\ \itt{cons}\ (g,\ \itt{nil}\ ()))$ in the
-standard domain for @FunList@ decomposes into $\itt{cons}\ (f,())$,
+standard domain for \mbox{\tt FunList} decomposes into $\itt{cons}\ (f,())$,
 $\itt{cons}\ (g,())$, and $\itt{nil}\ ()$.  The (implicit) abstraction
@@
 each element.  Thus
-${\it{}FFor}_{@@@A=T(A)@@@}={\it{}FFor}_{@@@T(())@@@}$, so
-${\it{}FFor}_{@@@FunList@@@}\iso{\it{}FFor}_{@@@FunTree@@@}\iso{\it{}FFor}_{@@@Int->Int@@@}$.
+${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt A=T(A)}} }={\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt T(())}} }$, so
+${\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt FunList}} }\iso{\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt FunTree}} }\iso{\it{}FFor}_{\subtypespace{\makebox{\subtypesize\tt Int->Int}} }$.
 
@@
 
-For all closed expressions @e@ the abstract value
-$\Ee{\Ps}{@e@}~[\,]$ of @e@ is of the form $(\ID,\K)$, showing
+For all closed expressions \mbox{\tt e} the abstract value
+$\Ee{\Ps}{\mbox{\tt e}}~[\,]$ of \mbox{\tt e} is of the form $(\ID,\K)$, showing
 that closed expressions are always entirely static.  For expressions of
 function type the abstract forward value $\K$ is a function from
-abstract arguments of @e@ to abstract results.
+abstract arguments of \mbox{\tt e} to abstract results.
 
 First we consider functions on lists.
-Let expression @length@ denote the usual length function:
-@@@
-     fix (\length . \xs . case xs of
-                            nil u  -> 0
-                            cons p -> let (z,zs) = p in
-                                        1 + length zs)
-@@@
-The abstract forward value of @length@ is of the form
+Let expression \mbox{\tt length} denote the usual length function:
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}length\ .\ {\char'134}xs\ .\ case\ xs\ of}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ 0}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 1\ +\ length\ zs)}\vspace{-0.0em}
+\end{flushleft}
+The abstract forward value of \mbox{\tt length} is of the form
 $\L(\A,\K).(\T~\A,())$, where $\T$ maps $\SPINE~\BOT$ and $\SPINE~\ID$
 to $\ID$, and $\BOT$ to $\BOT$.  This reveals that the result of
-@length@ is independent of the values of list elements, and gives
+\mbox{\tt length} is independent of the values of list elements, and gives
 a static result when the argument has a static spine.  It is worth
-pointing out that the result of @length@ is static---determined---for
+pointing out that the result of \mbox{\tt length} is static---determined---for
 infinite-length arguments with static spines; staticness and definedness
@@
 
-Let @append@ stand for the expression denoting the usual function for
+Let \mbox{\tt append} stand for the expression denoting the usual function for
 appending two lists:
-@@@
-     fix (\append . \xs . \ys . case xs of
-                                  nil u  -> ys
-                                  cons p -> let (z,zs) = p in
-		                              cons (z, 
-						    append zs ys))
-@@@
-The abstract value of @append@ is
-""
-(\ID, \L(\A_{@@@xs@@@},\K_{@@@xs@@@}).
-(\ID, \L(\A_{@@@ys@@@},\K_{@@@ys@@@}).
-(\A_{@@@xs@@@}\glb\A_{@@@ys@@@}, \K_{@@@xs@@@}\glb\K_{@@@ys@@@}))) .
-""
-This reveals that partial applications of @append@ are static up to
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}append\ .\ {\char'134}xs\ .\ {\char'134}ys\ .\ case\ xs\ of}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ ys}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
+\makebox{\progsize\tt 		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ (z,\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt 						\ \ \ \ append\ zs\ ys))}\vspace{-0.0em}
+\end{flushleft}
+The abstract value of \mbox{\tt append} is
+\beqs
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt xs}} },\K_{\subtypespace{\makebox{\subtypesize\tt xs}} }).\\
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt ys}} },\K_{\subtypespace{\makebox{\subtypesize\tt ys}} }).\\
+\it (\A_{\subtypespace{\makebox{\subtypesize\tt xs}} }\glb\A_{\subtypespace{\makebox{\subtypesize\tt ys}} },\ \K_{\subtypespace{\makebox{\subtypesize\tt xs}} }\glb\K_{\subtypespace{\makebox{\subtypesize\tt ys}} })))\ .
+\eeqs
+This reveals that partial applications of \mbox{\tt append} are static up to
 WHNF, and the abstract value of the result is the greatest lower bound
 of the two arguments.  In general, the abstract value of a closed
-expression of the form $@\x@_1@.\x@_2\ldots@.e@$ will reveal that all
+expression of the form $\mbox{\tt {\char'134}x}_1\mbox{\tt .{\char'134}x}_2\ldots\mbox{\tt .e}$ will reveal that all
 partial applications of the expression are static up to WHNF.
 
-Let @reverse1@ stand for the expression denoting the naive reverse
+Let \mbox{\tt reverse1} stand for the expression denoting the naive reverse
 function:
-@@@
-     fix (\reverse1 . \xs . case xs of
-                              nil u  -> nil ()
-                              cons p -> let (z,zs) = p in
-                                          append (reverse zs) 
-			                         (cons (z, nil ())))
-@@@
-The abstract forward value of @reverse1@ is the identity, so the
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}reverse1\ .\ {\char'134}xs\ .\ case\ xs\ of}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ nil\ ()}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (z,zs)\ =\ p\ in}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ append\ (reverse\ zs)\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt 			\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (cons\ (z,\ nil\ ())))}\vspace{-0.0em}
+\end{flushleft}
+The abstract forward value of \mbox{\tt reverse1} is the identity, so the
 abstraction of the elements of a list doesn't change by reversing the
@@
 
-Let @compose@ stand for @\f.\g.\x.f (g x)@.  The abstract value of @compose@ is
-""
-(\ID, \L(\A_{@@@f@@@},\K_{@@@f@@@}).
-(\ID, \L(\A_{@@@g@@@},\K_{@@@g@@@}).
-(\ID, \left\{\ba{lll}\bot, &\mbox{\rm if}\ \A_{@@@f@@@}=\BOT\mbox{\rm or} \A_{@@@g@@@}=\BOT
-	\K_{@@@f@@@}\o\K_{@@@g@@@},&\mbox{\rm otherwise}\ea\right\}))) .
-""
+Let \mbox{\tt compose} stand for \mbox{\tt {\char'134}f.{\char'134}g.{\char'134}x.f\ (g\ x)}.  The abstract value of \mbox{\tt compose} is
+\beqs
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt f}} },\K_{\subtypespace{\makebox{\subtypesize\tt f}} }).\\
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt g}} },\K_{\subtypespace{\makebox{\subtypesize\tt g}} }).\\
+\it (\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt f}} }=\BOT\mbox{\rm\ or}\ \A_{\subtypespace{\makebox{\subtypesize\tt g}} }=\BOT\\
+\it \K_{\subtypespace{\makebox{\subtypesize\tt f}} }\o\K_{\subtypespace{\makebox{\subtypesize\tt g}} },&\it \mbox{\rm\ otherwise}\ea\right\})))\ .
+\eeqs
 Thus, the
@@
 
-Let @listcomp@ stand for the expression denoting the function
+Let \mbox{\tt listcomp} stand for the expression denoting the function
 that composes lists of functions:
-@@@
-     fix (\listcomp . \fs . case fs of
-                              nil u  -> \x.x
-                              cons p -> let (g,gs) = p in 
-                                          compose g (listcomp gs))
-@@@
-The abstract value of @listcomp@ is
-""
-	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
-	(\ID,\ \left\{\ba{lll}\
-\bot,		&\mbox{\rm if}\ \A_{@@@fs@@@}\neq\SPINE\ \ID
-\glb_{i\geq0}(\K_{@@@fs@@@})^i,	&\mbox{\rm otherwise}\ea\right\})) .
-""
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}listcomp\ .\ {\char'134}fs\ .\ case\ fs\ of}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ nil\ u\ \ ->\ {\char'134}x.x}\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ cons\ p\ ->\ let\ (g,gs)\ =\ p\ in\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ compose\ g\ (listcomp\ gs))}\vspace{-0.0em}
+\end{flushleft}
+The abstract value of \mbox{\tt listcomp} is
+\beqs
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).(\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\SPINE\ \ID\\
+\it \glb_{i\geq{\rm0}}(\K_{\subtypespace{\makebox{\subtypesize\tt fs}} })^i,&\it \mbox{\rm\ otherwise}\ea\right\}))\ .
+\eeqs
 Since the abstract values of lists contain no information about the
@@
 
-Let @flatten@ stand for the expression denoting the function that
+Let \mbox{\tt flatten} stand for the expression denoting the function that
 flattens trees into lists:
-@@@
-     fix (\flatten . \t . case t of
-	                    leaf l   -> cons (l, nil ())
-	                    branch p -> let (t1,t2) = p in
-			                  append (flatten t1) 
-					         (flatten t2))
-@@@
-The abstract value of @flatten@ is
-""
-	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
-	(\T\ \A,\ \left\{\ba{lll}\
-\bot,		&\mbox{\rm if}\ \T\ \A_{@@@fs@@@}\neq\LBR\ \ID
-\K_{@@@fs@@@},	&\mbox{\rm otherwise}\ea\right\})) ,
-""
+\begin{flushleft}
+\vspace{-0.0em}\makebox{\progsize\tt \ \ \ \ \ fix\ ({\char'134}flatten\ .\ {\char'134}t\ .\ case\ t\ of}\vspace{-0.0em}\\
+\makebox{\progsize\tt 	\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ leaf\ l\ \ \ ->\ cons\ (l,\ nil\ ())}\vspace{-0.0em}\\
+\makebox{\progsize\tt 	\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ branch\ p\ ->\ let\ (t1,t2)\ =\ p\ in}\vspace{-0.0em}\\
+\makebox{\progsize\tt 			\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ append\ (flatten\ t1)\ }\vspace{-0.0em}\\
+\makebox{\progsize\tt 					\ \ \ \ \ \ \ \ \ (flatten\ t2))}\vspace{-0.0em}
+\end{flushleft}
+The abstract value of \mbox{\tt flatten} is
+\beqs
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).(\T\ \A,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \T\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\LBR\ \ID\\
+\it \K_{\subtypespace{\makebox{\subtypesize\tt fs}} },&\it \mbox{\rm\ otherwise}\ea\right\}))\ ,
+\eeqs
 where $\T$ maps $\BOT$ to $\BOT$, and $\LBR~\A$ to $\SPINE~\A$ for $\A$
@@
 
-The function denoted by @compose listcomp flatten@ that composes
+The function denoted by \mbox{\tt compose\ listcomp\ flatten} that composes
 trees of functions has abstract value
-""
-	(\ID,\ \L(\A_{@@@fs@@@},\K_{@@@fs@@@}).\
-	(\ID,\ \left\{\ba{lll}\
-\bot,		&\mbox{\rm if}\ \A_{@@@fs@@@}\neq\LBR\ \ID
-\glb_{i\geq0}(\K_{@@@fs@@@})^i,	&\mbox{\rm otherwise}\ea\right\}))\ .
-""
+\beqs
+\it (\ID,\ \L(\A_{\subtypespace{\makebox{\subtypesize\tt fs}} },\K_{\subtypespace{\makebox{\subtypesize\tt fs}} }).(\ID,\ \left\{\ba{lll}\bot,&\it \mbox{\rm\ if}\ \A_{\subtypespace{\makebox{\subtypesize\tt fs}} }\neq\LBR\ \ID\\
+\it \glb_{i\geq{\rm0}}(\K_{\subtypespace{\makebox{\subtypesize\tt fs}} })^i,&\it \mbox{\rm\ otherwise}\ea\right\}))\ .
+\eeqs
 Similarly to the case for lists, the abstract values of trees contain
@@
 $f$.  More generally, for function types embedded within data
-structures, e.g.\ $@(@\tint@,@\tfun@)@$, abstract values
+structures, e.g.\ $\mbox{\tt (}\tint\mbox{\tt ,}\tfun\mbox{\tt )}$, abstract values
 $((\A,()),(\BOT,f))$ would be identified over all $f$.
@@
 In the following we restrict attention to {\it denotable\/} values: at
-each type @T@ those values that can be expressed as $\Ee{}{@e@}~[\,]$
-for some @e@.  For every value $v\in\Te{\Ss}{@T@}$ there is a best
-abstraction---a greatest value $(\A,\K)\in\Te{\Ps}{@T@}$ such that
-$\Re{}{@T@}((\A,\K),v,v)$.  For types $\tone$ and $\ttwo$ not
-containing @->@ we have
+each type \mbox{\tt T} those values that can be expressed as $\Ee{}{\mbox{\tt e}}~[\,]$
+for some \mbox{\tt e}.  For every value $v\in\Te{\Ss}{\mbox{\tt T}}$ there is a best
+abstraction---a greatest value $(\A,\K)\in\Te{\Ps}{\mbox{\tt T}}$ such that
+$\Re{}{\mbox{\tt T}}((\A,\K),v,v)$.  For types $\tone$ and $\ttwo$ not
+containing \mbox{\tt ->} we have
 $\Tde{\Ps}{\tfun}\iso(\proj{\De{}{\tone}}\x\one)\fto(\proj{\De{}{\ttwo}}\x\one)$,
-and if $\K\in\Tde{\Ps}{@T@}$ is greatest such that
+and if $\K\in\Tde{\Ps}{\mbox{\tt T}}$ is greatest such that
 $\Rde{}{\tfun}~(\K,f,f)$ for denotable $f$, then $\K$ maps $(\ID,())$
@@
 
-We have successfully generalised Launchbury's polymorphic
-partial-projection-based demand analysis to higher-order, using abstract
+We have successfully generalised Launchbury's monomorphic
+projection-based binding-time analysis to higher-order, using abstract
 domains smaller than those typically used in BHA strictness analysis.
 
-%The next step would be to generalise to handle Hindley-Milner
-%polymorphism.  This has been done with good results at first order for
-%binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
-%at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
-%use of these theories will give a reasonably straightforward
-%generalisation to polymorphism.  In addition to making the analysis
-%more widely applicable, this should also greatly reduce the run-time
-%cost of analysis, as it does for BHA analysis.
-
-%Though not developed here, using the same approach it is possible to
-%give a strictness analysis technique, again closely related to Hunt's
-%PER-based strictness analysis technique \cite{Hun91}.
+The next step would be to generalise to handle Hindley-Milner
+polymorphism.  This has been done with good results at first order for
+binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
+at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
+use of these theories will give a reasonably straightforward
+generalisation to polymorphism.  In addition to making the analysis
+more widely applicable, this should also greatly reduce the run-time
+cost of analysis, as it does for BHA analysis.
+
+Though not developed here, using the same approach it is possible to
+give a strictness analysis technique, again closely related to Hunt's
+PER-based strictness analysis technique \cite{Hun91}.
 
@@
 
+\end{article}
 \end{document}

### 2017-04-17 pubs/demand_analysis/new.tex

--- 2017-04-17 pubs/demand_analysis/jlsc.tex
@@
-%% Kluwer Small Journal Article Sample File
-%% June 20, 1994
-
-%%   Use one of these two commands: 
-\documentstyle{smjrnl} % For Computer Modern Fonts, or
-%\documentstyle[smjfonts]{smjrnl} % for PostScript fonts
+\documentclass[11pt]{article}
+
+\usepackage{geometry}
+\geometry{textwidth=7.5in,
+  textheight=9in}
+
+%\usepackage{draftwatermark}
+%\SetWatermarkText{DRAFT}
+%\SetWatermarkScale{6}
+%\SetWatermarkColor{yellow}
 
 \begin{document}
-\begin{article}
-
-%%%%% To be entered at Kluwers: =====>>
-
-\journame{Lisp and Symbolic Computation}
-\volnumber{?}
-\issuenumber{?}
-\issuemonth{?}
-\volyear{1995}
-
-%% Do not delete either of the following two commands.
-%% Please supply facing curly brackets for the part you
-%% are not using for this article.
-%\received{May 1, 1991}\revised{}
-
-\authorrunninghead{K. Davis}
-\titlerunninghead{PERs from Projections for Binding-time Analysis}
+
+\title{PERs from Partial Projections for Higher-order Polymorphic Demand Analysis}
+\author{Kei Davis\\
+  Los Alamos National Laboratory\\
+  kei at lanl.gov}
+\maketitle
 
@@
                        %% issue, if desired.
-
-%% <<== End of commands to be entered at Kluwers
-
-%%  Authors, start here ==>>
 
@@
 \com{\tyip}{\theta_{\infty{}i+{\rm1}}}
-\com{\th}{\theta}
+%\com{\th}{\theta}
 
@@
 
-\title{PERs from Projections for Binding-time Analysis}
-
-\authors{Kei Davis}
-\email{kei@crl.nmsu.edu}
-
-\affil{Center for Non-Linear Studies \\
-Mail Stop B258 \\
-Los Alamos National Laboratory \\
-Los Alamos, NM  87545, U.S.A.}
-
-%\editor{Dennis Kibler}
-
-\abstract{{\it First-order\/} projection-based binding-time analysis
-has proven genuinely useful in partial evaluation
-\cite{Lau91a}.  There have been three notable generalisations
-of projection-based analysis to higher order.  The first lacked a
-formal basis \cite{Mog89}; the second used structures strictly more
-general than projections, namely {\it partial equivalence relations\/}
-(PERs) \cite{HS91}; the third involved a complex construction that
-gave rise to impractically large abstract domains \cite{Dav93b}.  This
-paper presents a technique free of these shortcomings: it is simple,
-entirely projection-based, satisfies a formal correctness condition,
-and gives rise to reasonably small abstract domains.  Though the
-technique is cast in terms of projections, there is also an
-interpretation in terms of PERs.  The principal limitation of the
-technique is the restriction to {\it monomorphic\/} typing.}
-
-\keywords{Binding-time analysis, abstract interpretation, projections,
-PERs.}
+\begin{abstract}
+  {\it First-order\/} projection-based strictness analysis (properly,
+  \emph{demand analysis}) was first introduced by Wadler and Hughes in
+  1987~\cite{WH87}.  This development was restricted to first-order,
+  monomorphic languages and provided no proof of soundness.  Subsequent
+  developments extended the theory to other analyses, approaches to higher
+  order, and polymorphism.  Yet other developments were based on
+  generalization of projections, specifically partial equivalence relations
+  and partial projections, the latter notably simplifying a difficulty with
+  projection-based strictness dating back to Wadler and Hughes.  However, to
+  date no development has combined the desirable contributions of all of these
+  approaches.  We present the first practical higher-order, polymorphic,
+  partial-projection/partial-equivalence-relation based demand analysis,
+  together with a proof of satisfaction of the fundamental safety condition
+  for such analyses.  Such an analysis can be used both to improve serial code
+  generation, as well as enable safe automatic parallelization, for lazy or
+  non-strict functional languages.
+
+% has proven genuinely useful in partial evaluation
+% \cite{Lau91a}.  There have been three notable generalisations
+% of projection-based analysis to higher order.  The first lacked a
+% formal basis \cite{Mog89}; the second used structures strictly more
+% general than projections, namely {\it partial equivalence relations\/}
+% (PERs) \cite{HS91}; the third involved a complex construction that
+% gave rise to impractically large abstract domains \cite{Dav93b}.  This
+% paper presents a technique free of these shortcomings: it is simple,
+% entirely projection-based, satisfies a formal correctness condition,
+% and gives rise to reasonably small abstract domains.  Though the
+% technique is cast in terms of projections, there is also an
+% interpretation in terms of PERs.  The principal limitation of the
+% technique is the restriction to {\it monomorphic\/} typing.
+\end{abstract}
+
+%\keywords{Binding-time analysis, abstract interpretation, projections, PERs.}
 
@@
 
-We take as given that binding-time analysis is essential for good
-partial evaluation, and we do not address the issue of annotating
-programs according to the results of analysis.  Numerous binding-time
-analysis techniques have been proposed and implemented; we greatly
-narrow the field of discussion by restricting attention to those for
-which there is a formally stated notion of correctness that the
-technique has been proven to satisfy.
+We take as given that demand analysis is essential for good serial code
+generation, and for identifying where parallelism may be safely introduced.
+Numerous such analysis techniques have been proposed and implemented; we
+greatly narrow the field of discussion by restricting attention to those for
+which there is a formally stated notion of correctness that the technique has
+been proven to satisfy.
 
@@
 Further, $\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
-$\K$.\ \ $\Box$
+$\K$.\ \ %$\Box$
 
@@
 
-We have successfully generalised Launchbury's monomorphic
-projection-based binding-time analysis to higher-order, using abstract
+We have successfully generalised Launchbury's polymorphic
+partial-projection-based demand analysis to higher-order, using abstract
 domains smaller than those typically used in BHA strictness analysis.
 
-The next step would be to generalise to handle Hindley-Milner
-polymorphism.  This has been done with good results at first order for
-binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
-at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
-use of these theories will give a reasonably straightforward
-generalisation to polymorphism.  In addition to making the analysis
-more widely applicable, this should also greatly reduce the run-time
-cost of analysis, as it does for BHA analysis.
-
-Though not developed here, using the same approach it is possible to
-give a strictness analysis technique, again closely related to Hunt's
-PER-based strictness analysis technique \cite{Hun91}.
+%The next step would be to generalise to handle Hindley-Milner
+%polymorphism.  This has been done with good results at first order for
+%binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
+%at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
+%use of these theories will give a reasonably straightforward
+%generalisation to polymorphism.  In addition to making the analysis
+%more widely applicable, this should also greatly reduce the run-time
+%cost of analysis, as it does for BHA analysis.
+
+%Though not developed here, using the same approach it is possible to
+%give a strictness analysis technique, again closely related to Hunt's
+%PER-based strictness analysis technique \cite{Hun91}.
 
@@
 
-\end{article}
 \end{document}

### 2017-05-01 pubs/demand_analysis/new.verb.tex
sync before branch

@@
   partial equivalence relations and partial projections, the latter notably
-  simplifying a difficulty with projection-based strictness analysis dating
-  back to Wadler and Hughes.  However, to date no development has combined the
-  desirable attributes of all of these approaches.  We present the first
-  practical, higher-order, polymorphic,
-  partial-projection/partial-equivalence-relation based demand analysis,
-  together with a proof of satisfaction of the fundamental safety condition
-  for such analyses.  Such an analysis can be used both to improve serial code
-  generation as well as enable safe automatic parallelization for lazy or
-  non-strict functional languages.
+  smoothing over a complicating construction in projection-based strictness
+  analysis dating back to Wadler and Hughes.  However, to date no development
+  has combined the desirable attributes of all of these approaches, or given
+  the details where unboxed types are explicit.  We present the first
+  practical, higher-order, polymorphic, forward demand analysis based on
+  partial projections, incorporating the notions of partial equivalence
+  relations and logical relations, together with a proof of satisfaction of
+  the fundamental correctness condition for such analyses.  Such an analysis
+  can be used both to improve serial code generation, as well as enable safe
+  automatic parallelization, for lazy or non-strict functional languages.
 

### 2017-05-18 pubs/tfp16-draft/paper.tex
works?

@@
 \\
-Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
+Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $f\ x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
              &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
@@
 update-in-place mechanism via a uniqueness type system.
+
+Mercury~\cite{bone12:autoparallelism_thesis}.  ``Overall, we have attacked and solved a number of issues that are critical to making automatic parallelism a realistic option for developers.''
 

### 2017-05-18 codegen/CodeGen.hs

@@ -214,18 +214,29 @@ cgUBa _ at _ = error $ "CodeGen.cgUBa: not expecting Atom - " ++ show at
 --   all objects produce a (S)HO
 -- for CG, objects are heap allocated only by let
 
+culParFun :: [(String, Definition,  CFun)] -> ([Definition], [CFun])
+culParFun (("fun_par", d, f):xs) =
+    let (ss, ds, fs) = unzip3 xs
+    in  (ds, fs)
+culParFun ((_, d, f):xs) =
+    let (ds, fs) = culParFun xs
+    in (d:ds, f:fs)
+culParFun [] = ([],[])
+
 data YPN = Yes | Possible | No -- could use Maybe Bool but seems obscure
            deriving(Eq, Show)
 
--- the Language.C isn't particularly convenient to pattern match on so attach a name
 cgObjs :: [Obj InfoTab] -> [String] -> ([Definition], [CFun])
 cgObjs objs runtimeGlobals =
    let tlnames = runtimeGlobals ++ map (name . omd) objs
        env = zip tlnames $ repeat SO
        funcs = cgos env objs
-       (names, forwards, funs) = unzip3 funcs
+       -- dirty hack:  cull the funs for par
+       -- is this a general approach/should be generalized?
+       (forwards, funs) = culParFun funcs
    in (forwards, funs)
 
+-- the Language.C isn't particularly convenient to pattern match on so attach a name
 cgos :: Env -> [Obj InfoTab] -> [(String, Definition,  CFun)]
 cgos env = concatMap (cgo env)
 

### 2017-05-18 runtime/stg.c

@@ -140,3 +140,4 @@ void initStg(int argc, char *argv[]) {
   }
 
 }
+

### 2017-05-18 runtime/stg.h

@@ -140,6 +140,7 @@ void showStgValPretty(LogLevel priority, PtrOrLiteral v);
 void checkStgHeap();
 void showIT(InfoTab *);
 
+
 // Codegen.hs currently uses STGJUMP(), STGJUMP0(f), and STGRETURN0() to
 // exit functions
 

### 2017-05-18 runtime/stgutils.c

@@ -295,3 +295,31 @@ CInfoTab it_stgCallCont __attribute__((aligned(8))) =
     .cLayoutInfo.bm.bitmap.mask = 0x0,   // shouldn't be using this
     .cLayoutInfo.bm.bitmap.size = 0,  // shouldn't be using this
   };
+
+// forall t27,t29.t27 -> t29 -> t29
+// par(self, x, y)
+FnPtr fun_par()
+{
+    LOG(LOG_INFO, "par here thread=%d\n", myThreadID());
+    
+    PtrOrLiteral *argp = &stgGetStackArgp(myThreadID())->payload[0];
+    
+    (void) argp;
+    // suppress warning
+    ;
+    // y
+    stgCurVal[myThreadID()] = argp[2];
+    // boxed EAtom, stgCurVal updates itself
+    STGJUMP();
+}
+
+InfoTab it_par __attribute__((aligned(OBJ_ALIGN))) = {.name ="par", .entryCode =
+                                                      &stg_funcall, .objType =
+                                                      FUN,
+                                                      .layoutInfo.payloadSize =
+                                                      0,
+                                                      .layoutInfo.boxedCount =0,
+                                                      .layoutInfo.unboxedCount =
+                                                      0, .funFields.arity =2,
+                                                      .funFields.trueEntryCode =
+                                                      fun_par};

### 2017-05-18 runtime/stgutils.h

@@ -43,6 +43,10 @@ extern Obj sho_stg_case_not_exhaustiveN;
 
 void stgCaseToPopMe(Cont *contp);
 
+FnPtr fun_par();
+extern InfoTab it_par;
+
+
 #if USE_ARGTYPE
 #define HOTOPL(HO) ((PtrOrLiteral) {.argType = HEAPOBJ, .op = (HO) })
 #define INTTOPL(L) ((PtrOrLiteral) {.argType = INT,     .i = L   })

### 2017-05-18 runtime/gc.c
Observation:  the only SHOs that need to be roots for GC are THUNKs that have become INDIRECTS (or BLACKHOLEs later?)

@@ -231,7 +231,9 @@ void gc(void) {
 
   // all SHO's
   for (int i = 0; i < stgStatObjCount; i++) {
-    processObj(stgStatObj[i]);
+    if (// getObjType(stgStatObj[i]) == THUNK ||
+	getObjType(stgStatObj[i]) == INDIRECT ||
+	getObjType(stgStatObj[i]) == BLACKHOLE) processObj(stgStatObj[i]);
   }
 
   //Cont. stack

### 2017-05-18 runtime/gc.c
ditto

@@ -231,9 +231,10 @@ void gc(void) {
 
   // all SHO's
   for (int i = 0; i < stgStatObjCount; i++) {
+    ObjType ot = getObjType(stgStatObj[i]);
     if (// getObjType(stgStatObj[i]) == THUNK ||
-	getObjType(stgStatObj[i]) == INDIRECT ||
-	getObjType(stgStatObj[i]) == BLACKHOLE) processObj(stgStatObj[i]);
+	ot == INDIRECT ||
+	ot == BLACKHOLE) processObj(stgStatObj[i]);
   }
 
   //Cont. stack

### 2017-07-11 codegen/Analysis/DSL.hs
Fixed Fact generation, simple typechecking seems to work

@@ -111,27 +111,43 @@ builtin s = fromString s =: "builtin"
 
 builtins = map builtin
   [ "builtin"
-  -- , "add#"
-  -- , "mul#"
-  -- , "div#"
+  , "add#"
+  , "mul#"
+  , "div#"
   ]
 
-subDef, facDef, broken :: ValDef ()
-subDef = "sub#" =: "builtin"
+appendDef, foldrDef, subDef, facDef, broken :: ValDef ()
+subDef = "sub# a b" =: "builtin a b"
 
 
 facDef = "fac a#" =: match "a#" "x#"
          [ "0#" --> "1#"
-         , "_"  --> letrec [] .> "fac" .$ "sub# 1# x#"
+         , "_"  --> letrec [] .> "mul# a#" .$ "fac" .$ "sub# 1# x#"
          ]
 
-broken = "broken" =: "fac True"
+appendDef = "append xs ys" =: match "xs" "z"
+            [ "Nil" --> "ys"
+            , "Cons h t" --> "Cons h" .$ "append t ys"
+            ]
+
+foldrDef = "foldr f s l" =: match "l" "z"
+           [ "Nil" --> "s"
+           , "Cons h t" --> "f h" .$ "foldr f s t"
+           ]
+
+broken = "broken x" =: match "x" "z"
+         [ "Nil" --> "Nil"
+         , "Cons h t" --> "Cons" .$ "broken h" .$ "broken t"
+         ]
 
 prog = Prog
   ( builtins ++
-    [ facDef
+    [
+      facDef
     , subDef
-    , broken      
+    , foldrDef
+--    , broken
+--    , appendDef
     ]
   , [ listDef
     , boolDef

### 2017-07-11 codegen/Analysis/Language.hs

@@ -429,9 +429,10 @@ uniqExpr e = scoped $ case e of
     LetRec <$> uniqVDefs bnds <*> uniqExpr exp <*> pure m
   Apply f e m ->
     Apply <$> uniqExpr f <*> uniqExpr e <*> pure m
-p
+
 uniqClause :: Clause a -> UniqState (Clause (a :-> Uniquify))
 uniqClause c = scoped $ case c of
-  LitMatch l e m    -> LitMatch l <$> uniqExpr e                      <*> pure m
-  ConMatch c vs e m -> ConMatch c <$> mapM newScope vs <*> uniqExpr e <*> pure m
-  Default e m       -> Default    <$> uniqExpr e                      <*> pure m
+  LitMatch l e m    -> LitMatch l <$> uniqExpr e <*> pure m
+  ConMatch c vs e m -> ConMatch <$> setUniq c <*>
+                       mapM newScope vs <*> uniqExpr e <*> pure m
+  Default e m       -> Default <$> uniqExpr e <*> pure m

### 2017-07-11 codegen/Analysis/Type.hs

@@ -24,6 +24,8 @@ import qualified Data.Set          as S
 
 import           Control.Monad.RWS
 
+import Text.Show.Pretty
+
 import Debug.Trace
 
 type Fact        = (ID, Type)
@@ -69,7 +71,7 @@ addGetAssum id@(ID name uniq) = do
   let ty = TVar $ 't': show i ++ '_' : showUniqName id
   modifyAssums $ M.insertWith (++) id [ty]
   return ty
-  
+
 withMonotypes mts act = do
   mtys <- gets monotys
   modifyMonotys (S.union (S.fromList mts))
@@ -92,11 +94,12 @@ getAllAssums ids = gets $ M.filterWithKey (\k _ -> k `elem` ids) . assumps
 type UT f a = Unique (Typed f) a
 
 typecheck :: Unique Prog a ->
-  ([UT ValDef a], Constraints, Sub, Sub)
+  ([UT ValDef a], Constraints, Sub, Sub, Bool)
 typecheck (Prog (vdefs, ddefs)) =
   if M.null $ assumps endState
   then typed
-  else typed -- error $ "Assumptions aren't empty?\n" ++ show (assumps endState) ++
+  else typed
+       -- error $ "Assumptions aren't empty?\n" ++ show (assumps endState) ++
        -- '\n':show newVDefs
   where facts = M.unions $ map makeFacts ddefs
         newDDefs = map coerceDef ddefs
@@ -104,17 +107,16 @@ typecheck (Prog (vdefs, ddefs)) =
         initState = CGS M.empty S.empty IM.empty 0
         (newVDefs, endState, constraints) =
           runRWS (constrainVDefs vdefs) facts initState
-        subst = case traceShowId newVDefs of
-                  [] -> solve constraints
-                  x -> solve constraints
+        subst = case traceWith ppShow (newVDefs, constraints) of
+                  ([], _) -> solve constraints
+                  x  -> solve constraints
         fixed = fixSub subst
         tprog = applyAll fixed newVDefs
-        typed = (newVDefs, constraints, subst, fixed)
+        typed = (newVDefs, constraints, subst, fixed, subst == fixed)
 
 makeFacts :: Unique DataDef a -> Facts
 makeFacts (DDef t@(TApp tcon vs) cons) = M.fromList $ map mkFact cons
-  where mkFact dc = (conName dc, theType)
-        theType = TForall vars t
+  where mkFact (DCon id tys _) = (id, TForall vars $ foldr TFun t tys)
[…]

### 2017-07-11 codegen/Analysis/Type.hs
some comments and cleanup in typechecker

@@ -159,6 +159,7 @@ constrainVDef (VDef name rhs _) = do
   pure $ VDef name newRhs (getType newRhs)
 
 
+-- | Generate constraints/assumptions from an Expr type
 constrainExpr :: Unique Expr a -> CGM (Unique (Typed Expr) a)
 constrainExpr e = case e of
   Lit v _ -> pure $ Lit v (getType v)
@@ -215,7 +216,8 @@ constrainExpr e = case e of
   Apply f e _ -> do
     newF <- constrainExpr f
     newE <- constrainExpr e
-    atyp <- freshTVar -- We want to have a type for every expression,
+    atyp <- freshTVar -- We want to have a type for every expression, so make a
+                      -- fresh one for the result of this application
     let constr = getType newF :=: TFun (getType newE) atyp
     tell $ S.singleton constr
     pure $ Apply newF newE atyp
@@ -231,6 +233,10 @@ constrainExpr e = case e of
     pure $ LetRec newBinds newRes (getType newRes)
 
 
+-- | Helper function for generating Constraints from Assumptions and a Map of
+-- types inferred for some set of variables.  The Constraints generated are on
+-- types associated only with variables that appear in both the Assumptions and
+-- the Map.
 constrainAssumsWith
   :: (Type -> Type -> Constraint) -- How to make a constraint
   -> Assumptions -- Assumptions about some variables
@@ -242,6 +248,7 @@ constrainAssumsWith comb assums tymap =
   tymap assums
 
 
+-- | Generate assumptions and constrains about a Clause.
 constrainClause :: Unique Clause a -> CGM (Unique (Typed Clause) a)
 constrainClause c = case c of
   LitMatch lit consq _ -> do
@@ -249,21 +256,23 @@ constrainClause c = case c of
     let ty = (getType lit, getType newConsq)
     pure $ LitMatch lit newConsq ty
 
-  ConMatch con args consq _ -> do
-    fact     <- asks getFact
-    newConsq <- constrainExpr consq
-    assums   <- getAllAssums args
-    argTys   <- mapM addGetAssum args
-    let
-        resTy   = finalResTy fact
+  ConMatch con args consq _ ->
+    let getFact facts = fromMaybe err $ M.lookup con facts
+        err = error $ "Why isn't there a fact for " ++ show con
+    in do
+      fact     <- asks getFact
+      newConsq <- constrainExpr consq
+      assums   <- getAllAssums args
[…]
